### 2021-04-14 mensch-computer-interaktion/esa-1.tex
add mci esa template

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik \\
		Mensch-Computer-Interaktion
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 1: Personas und Storyboards}\\[0.5em]
		\large{Sommersemester 2021}\\[0.25em]
		\large{Abgabetermin 18.04.2021}
	\end{center}

	\vfill

	\begin{flushright}
		Maximilian Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\tableofcontents

\vfill

\section{Aufgabenstellung}

Folgende Aufgabenstellung wurde im Moodle-Kurs bekannt gegeben:

\begin{quote}
	Sie haben die Aufgabe eine Smartphone-App zu konzipieren, die OSMI-Studierende beim Studium unterstützt.
	Die App soll insbesondere die Kommunikation, die gegenseitige Unterstützung und das gemeinsame Bearbeiten
	von Aufgaben und Projekten unterstützen. Denken Sie auch darüber nach, wie die App in der aktuellen
	SARS-CoV-2-Situation helfen kann (also beispielsweise auch Studierende, die normalerweise in Präsenz studieren).
	\\[1em]
	(a) Erstellen Sie zwei Personas der (potentiellen) Zielgruppe.

	(b) Erstellen Sie für jede Persona zwei Storyboards (also insgesamt vier) zur Nutzung der App. Storyboards
	werden in der Regel von Hand gezeichnet (auf Schönheit kommt es in diesem Zusammenhang nicht an). Scannen
	oder fotografieren Sie die Skizzen und beschreiben Sie ggf. kurz wie die Skizzen verstanden werden sollen.
	\\[1em]
	Anmerkungen:

	Abzugeben ist ein PDF-Dokument, das Ihre Ausführungen enthält. Bitte beachten Sie dazu die Hinweise zu den
	Einsendeaufgaben (siehe oben). Der zeitliche Umfang dieser Einsendeaufgabe wird auf 6 Stunden geschätzt.
	Ihre Ausarbeitung sollte ca. 7-9 Seiten (A4) umfassen (eine Seite für Titelblatt inkl. Aufgabenstellung,
	je eine Seite pro Persona, und je eine pro Storyboard sowie ggf. zusätzliche Seiten für die Erläuterungen).
	Die Lösung, die Sie zur Deadline abgeben, sollte eine aus Ihrer Sicht endgültige Lösung sein. Falls Sie
	Fragen zur Aufgabenstellung haben, stellen Sie diese bitte im Vorfeld!
\end{quote}

\newpage

\section{Persona: Mark}

\begin{tabularx}{0.8\textwidth} { 
	| >{\raggedright\arraybackslash}X 
	| >{\centering\arraybackslash}X 
	| >{\raggedleft\arraybackslash}X | }
	\hline
	item 11 & item 12 & item 13 \\
	\hline
	item 21  & item 22  & item 23  \\
	\hline
\end{tabularx}

Name

"quote"

OSMI Studenten, die berufsbegleitend studieren
Alter
Beruf höchstem Bildungsabschluss (Realschulabschluss/ Abitur/ Ausbildung/Studium)
Familienstand,
Kinder?
Motivation für Entscheidung für OSMI?
Hobbies
Wohnort (könnte z.B. auch auf dem Land wohnen oder halt in Berlin)
OSMI-Student seit wann?
wie verbringe ich den Tag?
Konsumverhalten?
Bedürfnisse
Wünsche
Werte
Ängste
Introvertiert
Extrovertiert

\newpage

\section{Persona: XYZ}

\newpage

\section{Storyboard 1: Mark}

\newpage

\section{Storyboard 2: Mark}

\newpage

\section{Storyboard 3: XYZ}

\newpage

\section{Storyboard 4: XYZ}

\newpage

\end{document}

### 2021-04-15 algorithmen-und-datenstrukturen/esa-1.tex
add aud esa 1

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{color}

\lstset{
  language=Python,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  breaklines=true,
  breakatwhitespace=true,
  tabsize=4
}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik \\
		Algorithmen und Datenstrukturen
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 1}\\[0.5em]
		\large{Sommersemester 2021}\\[0.25em]
		\large{Abgabetermin 18.04.2021}
	\end{center}

	\vfill

	\begin{flushright}
		Maximilian Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\newpage

\section*{1. Zweitkleinstes Element einer Folge}

Das zweitkleinste Element einer Folge von n $\geq$ 2 Zahlen soll bestimmt werden.

\subsection*{a) Algorithmus in Pseudocode}

\begin{lstlisting}
def second_minimum(list):
    second = list[0]
    minimum = list[0]

    for n in list[1:]:
        if n > minimum:
            second = n
            break

    for n in list[1:]:
        if n < minimum:
            second = minimum
            minimum = n

    return second
\end{lstlisting}

\subsection*{b) Laufzeit-Analyse}

Der Algorithmus braucht im \textbf{Best-Case n} Vergleiche, liegt also dementsprechend in $\Omega(n)$. Der Best-Case tritt ein,
wenn direkt das zweite Element größer als das erste ist, da dann die erste Schleife nach dem ersten Schritt abgebrochen wird und
die 2. Schleife immer genau $n - 1$ vergleiche ausführt.

Er braucht im \textbf{Worst-Case 2(n - 1)} Vergleiche und liegt daher in $O(n)$. Der Worst-Case kommt zustande wenn wir z.B. eine List der
Länge n betrachten, die n mal das gleiche Element enthält. Dann benötigen wir beim der ersten und der zweiten Schleife $n - 1$ Vergleiche.

\end{document}

### 2021-04-15 algorithmen-und-datenstrukturen/esa-1.tex
complete exercise 2 of the aud assignment

@@
 \usepackage[T1]{fontenc}
+\usepackage{amsmath}
 \usepackage{listings}
@@
 
-\section*{1. Zweitkleinstes Element einer Folge}
+\section{Zweitkleinstes Element einer Folge}
 
@@
 
-\subsection*{a) Algorithmus in Pseudocode}
+\subsection{Algorithmus in Pseudocode}
 
@@
 
-\subsection*{b) Laufzeit-Analyse}
+\subsection{Laufzeit-Analyse}
 
@@
 
+\section{Asymptotische Notation}
+
+Gegeben sei die Funktion $f(n) = 2n^2 + 3n\log_2n - 72$
+
+\subsection{Beweis von $f(n) \in O(n^2)$}
+
+\begin{align*}
+	f(n) & = 2n^2 + 3n\log_2n - 72 \\
+	     & \leq 2n^2 + 3n\log_2n   \\
+	     & \leq 2n^2 + 3n^2        \\
+	     & = 5n^2
+\end{align*}
+
+Somit können wir sagen, dass mit $c \geq 5$ und $n_0 = 1$ die Behauptung $f(n) \in O(n^2)$ gilt
+
+\subsection{Beweis von $f(n) \in \Omega(n^2)$}
+
+\begin{align*}
+	f(n) & = 2n^2 + 3n\log_2n - 72 \\
+		 & \geq 2n^2 - 72 \\
+	     & \geq n^2
+\end{align*}
+
+Nun können wir $n_0$ als Schnittpunkt der beiden Funktionen $2n^2 - 72$ und $n^2$ berechnen.
+
+\begin{align*}
+	2n^2 - 72 & = n^2 \mid - 2n^2  \\
+		 - 72 & = -n^2 \mid * -1 \\
+		   72 & = n^2 \\
+		    n & = \sqrt{72}
+\end{align*}
+
+Also, mit $c = 1$ und $n_0 = \sqrt{72}$ gilt $f(n) \in \Omega(n^2)$
+
+\subsection{Gilt $f(n) \in \Theta(n^2)$ ?}
+
+$\Theta(g)$ ist im Skript mit der \emph{Definition 2.5} als $\{ \, f \mid f \in O(g) \wedge f \in \Omega(g) \, \}$ definiert.\\[0.5em]
+Somit wissen wir, dass $f(n) \in \Theta(n^2)$, da wir in 2.1 und 2.2 gezeigt haben, dass $f \in O(g)$ und $f \in \Omega(g)$ gelten.
+
 \end{document}

### 2021-04-15 algorithmen-und-datenstrukturen/esa-1.tex
format doc and add next chapter
finish aud exercise 3.2

@@
 
-Also, mit $c = 1$ und $n_0 = \sqrt{72}$ gilt $f(n) \in \Omega(n^2)$
+Also, mit $c = 1$ und $n_0 = \lceil\sqrt{72}\rceil = 9$ gilt $f(n) \in \Omega(n^2)$
 
@@
 
+\section{Average-Case-Aufwand der binären Suche}
+
+\subsection{Durchschnittliche Anzahl der Vergleiche für einen Hit}
+
+\vspace{1em}
+
+\begin{center}
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|}
+		\hline
+		Element    & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
+		\hline
+		Vergleiche & 3 & 2 & 3 & 4 & 1 & 3 & 4 & 2 & 3 & 4 \\
+		\hline
+	\end{tabular}
+\end{center}
+
+\vspace{1em}
+
+Macht in Summe 29 Vergleiche und somit $\frac{29}{10} = 2.9$ Vergleiche im Durchschnitt.
+
+\subsection{Summenformel für Vergleiche bei $2^k - 1$ Elementen}
+
+\begin{center}
+	\vspace{1em}
+
+	Beispiel für $k = 3$
+
+	\vspace{1em}
+
+	\begin{tabular}{|c|c|c|c|c|c|c|}
+		\hline
+		&  &  & 3 & & & \\
+		\hline
+		 & 1 & &  & & 5 & \\
+		\hline
+		0 &  & 2 &  & 4 &  & 6 \\
+		\hline
+	\end{tabular}
+
+	\vspace{2em}
+
+	Beispiel für $k = 4$
+
+	\vspace{1em}
+
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
+		\hline
+		   &   &   &   &   &   &   & 7 &   &   &    &    &    &    &    \\
+		\hline
+		   &   &   & 3 &   &   &   &   &   &   &    & 11 &    &    &    \\
+		\hline
+		   & 1 &   &   &   & 5 &   &   &   & 9 &    &    &    & 13 &    \\
+		\hline
+		0  &   & 2 &   & 4 &   & 6 &   & 8 &   & 10 &    & 12 &    & 14 \\
+		\hline
+	\end{tabular}
+\end{center}
+
+\newpage
+
+Es gibt bei $2^k - 1$ immer einen perfekten, gleichmäßigen Baum und immer genau $\log_2n$ bzw. $k$ Ebenen.
+Auf (einer 0 indizierten) Ebene $i$ haben wir den Baum $i$ Mal geteilt und vergleichen $(i + 1) * 2^i$ Elemente.
+Wenn wir nun alle Ebenen addieren möchten, um die gesamt Anzahl der verglichenen Elemente bekommen möchten,
+müssen wir lediglich alle Ebenen addieren. Also bei $k = 4$ wären wir bei $1 * 2^0 + 2 * 2^1 + 3 * 2^2 + 4 * 2^3$ Vergleichen.
+Dies lässt sich durch die gaußsche Summenformel eleganter (und allgemeingültiger) Zusammenfassen zu $\sum_{i=0}^{k - 1} (i + 1) * 2^i$.
+
 \end{document}

### 2021-04-15 algorithmen-und-datenstrukturen/assignment-1.tex
rename the aud assignment 1 document

--- 2021-04-15 algorithmen-und-datenstrukturen/esa-1.tex
@@
 Es gibt bei $2^k - 1$ immer einen perfekten, gleichmäßigen Baum und immer genau $\log_2n$ bzw. $k$ Ebenen.
-Auf (einer 0 indizierten) Ebene $i$ haben wir den Baum $i$ Mal geteilt und vergleichen $(i + 1) * 2^i$ Elemente.
-Wenn wir nun alle Ebenen addieren möchten, um die gesamt Anzahl der verglichenen Elemente bekommen möchten,
-müssen wir lediglich alle Ebenen addieren. Also bei $k = 4$ wären wir bei $1 * 2^0 + 2 * 2^1 + 3 * 2^2 + 4 * 2^3$ Vergleichen.
-Dies lässt sich durch die gaußsche Summenformel eleganter (und allgemeingültiger) Zusammenfassen zu $\sum_{i=0}^{k - 1} (i + 1) * 2^i$.
+Auf (einer 0 indizierten) Ebene $i$ haben wir den Baum $i$ Mal geteilt und haben $(i + 1) 2^i$ Vergleiche
+auf dieser Ebene. Wenn wir nun alle Ebenen addieren möchten, um die gesamt Anzahl der Vergleiche zu bekommen,
+müssen wir lediglich alle Ebenen addieren. Also bei $k = 4$ wären wir bei $1 * 2^0 + 2 * 2^1 + 3 * 2^2 + 4 * 2^3$
+Vergleichen. Dies lässt sich durch die gaußsche Summenformel eleganter (und allgemeingültiger) Zusammenfassen
+zu $\sum_{i=0}^{k - 1} (i + 1) * 2^i$.
+
+\subsection{Laufzeit-Analyse}
 

### 2021-04-15 mensch-computer-interaktion/assignment-1.tex
rename the mci assignment

--- 2021-04-14 mensch-computer-interaktion/esa-1.tex
@@
 
-\begin{tabularx}{0.8\textwidth} { 
-	| >{\raggedright\arraybackslash}X 
-	| >{\centering\arraybackslash}X 
-	| >{\raggedleft\arraybackslash}X | }
-	\hline
-	item 11 & item 12 & item 13 \\
-	\hline
-	item 21  & item 22  & item 23  \\
-	\hline
-\end{tabularx}
-
-Name
-
-"quote"
-
-OSMI Studenten, die berufsbegleitend studieren
-Alter
-Beruf höchstem Bildungsabschluss (Realschulabschluss/ Abitur/ Ausbildung/Studium)
-Familienstand,
-Kinder?
-Motivation für Entscheidung für OSMI?
-Hobbies
-Wohnort (könnte z.B. auch auf dem Land wohnen oder halt in Berlin)
-OSMI-Student seit wann?
-wie verbringe ich den Tag?
-Konsumverhalten?
-Bedürfnisse
-Wünsche
-Werte
-Ängste
-Introvertiert
-Extrovertiert
+Name					Marc Zahn
+Alter					28
+Beruf					Hochschul-Verwaltungsangestellter
+Bildungsabschluss		Abitur
+Familienstand			Verheiratet
+Kinder					Ja
+Motto					""
+Motivation für OSMI		Möchte nun Umlernen, Kontakt zum Studium durch Job an der Hochschule
+Hobbies					Alte Filme sammeln, Kochen, Wandern
+Wohnort					Brandenburg a.d.H.
+OSMI-Student seit		WS 2019
+Alltag?					Kinder Aufwecken, Frühstück für Familie machen, Arbeiten, Studieren
+Bedürfnisse				Erfolg, Soziale Anerkennung, Sicherheit für Familie
+Wünsche					Neuer Job in Wunschbereich
+Werte					Treue, Zuverlässigkeit, Ehrlichkeit
+Ängste					Angst zu Versagen, Unsichere Zukunft
+Persönlichkeit			Introvertiert
 
@@
 
+Name					Rosa Beetz
+Alter					23
+Beruf					Screen Desginerin
+Bildungsabschluss		Ausbildung
+Familienstand			Ledig
+Kinder					Nein
+Motto					""
+Motivation für OSMI		Möchte sich in Richtung Informatik weiterentwickeln und sich selbst etwas Beweisen
+Hobbies					Netflix, Feiern, Kaffee
+Wohnort					Berlin Wedding
+OSMI-Student seit		SS 2021
+Alltag					Arbeit, Freunde treffen \& Netflix
+Bedürfnisse				Sozialer Kontakt, Anerkennung
+Wünsche					Mehr Geld, Anerkennung im Job
+Werte					Freundlichkeit, Ehrgeiz
+Ängste					Angst zu Versagen
+Persönlichkeit			Extrovertiert
+
 \newpage
@@
 
+Als OSMI Student möchte ich Fahrgemeinschaften gründen
+
 \newpage
@@
 \section{Storyboard 2: Mark}
+
+Als OSMI Student möchte live mit anderen Studenten in einem Whiteboard oder einer PDF zusammenarbeiten
 
@@
 
+Als OSMI Student möchte ich mit anderen Studenten Videotelefonieren
+
 \newpage

### 2021-04-15 softwaretechnik/assignment-1.tex
rename the swt assignment

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik \\
		Softwaretechnik
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 1: Requirements Engineering}\\[0.5em]
		\large{Sommersemester 2021}\\[0.25em]
		\large{Abgabetermin 18.04.2021}
	\end{center}

	\vfill

	\begin{flushright}
		Maximilian Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\tableofcontents

\newpage

\section{Aufgabenstellung}

\section{Projekt}

Als Projekt für dieses Semester verwende ich ein privates Freizeit-Projekt von mir – einen kleinen,
minimalistischen Tiling-Window-Manager für Linux bzw. X11.

\subsection{Zusammenfassung der Domain}

Für den Fall, dass Sie noch keinen Kontakt mit dieser Domain, hatten gehe ich nun kurz darauf ein,
welche Aufgaben ein Tiling-Window-Manager (kurz. \emph{TWM} oder nur \emph{WM}) typischer Weise übernimmt.
Die generelle Aufgabe eines WM's besteht hauptsächlich darin sich um die Kommunikation mit dem
Window-Server (i.d.R. X11 oder Wayland auf Linux, Quartz Compositor auf MacOS) zu kümmern, und die
tatsächliche Anordnung der Fenster auf dem Bildschirm zu regeln (in Ebenen, Kacheln etc. – die Möglichkeiten
sind nahe zu unbegrenzt). Typische Window-Manager von z.B. MacOS können diverse, für uns als Endnutzer
mittlerweile als üblich angesehene, Anordnungen wie Floating, Split-Screen und Fullscreen realisieren. \par
Ein Tiling-Window-Manager hat nun die Besonderheit, dass er anders als von Windows, MacOS oder diversen Linux
Desktop-Umgebungen bekannt, die Aufteilung der Fenster automatisch und bestmöglich regelt. Typischerweise
hat ein TWM eine feste Konfiguration mit Layouts, in die er die Fenster einsortieren kann (z.B. ``Master
and Stack``). Somit muss sich der Anwender eines solchen TWM's (zumindest initial) nicht selber um die
Anordnung kümmern.

\subsection{Projekt-Ziel}

Das Problem von bestehenden Window-Managern ist in der Regel deren Alter (bspw. ist das Projekt ``Toms
Window Manager`` im Jahr 1987 entstanden). Aufgrund des Alters sind viele dieser Window-Manager noch in
C, C++ oder einer vergleichbaren Sprache verfasst worden und mit der Zeit immer weiter gewachsen. Dies
macht, nicht von den Entwicklern vorhergesehene, grundlegende Änderungen – zumindest für mich – äußert
schwer.

% Da ich erstens C & C++ nicht behersche und die Projektgröße sowas ich ohnehin, die meisten Standardfunktionalitäten nicht 
% "Rust TWM Library" um selber mit rust knowledge seine twm zusammen zu schrauben.

\section{User Stories}

\end{document}

### 2021-04-15 algorithmen-und-datenstrukturen/assignment-1.tex
add the proof for 4.2

@@
 Vergleichen. Dies lässt sich durch die gaußsche Summenformel eleganter (und allgemeingültiger) Zusammenfassen
-zu $\sum_{i=0}^{k - 1} (i + 1) * 2^i$.
+zu $\sum_{i=0}^{k - 1} (i + 1) * 2^i$. Um jetzt auf die durchschnittlichen Vergleiche zu kommen muss nun einfach
+die Gesamtanzahl durch die Anzahl der Elemente geteilt werden. Also entweder $\sum_{i=0}^{k - 1} \frac{(i + 1) * 2^i}{2^k - 1}$
+$\frac{\sum_{i=0}^{k - 1} (i + 1) * 2^i}{2^k - 1}$
 
@@
 
+TBD
+
+\section{Analyse einer rekursiven Funktion}
+
+\subsection{Für welche n terminiert die Rekursion, für welche nicht?}
+
+Die Funktion terminiert nur für gerade postive Zahlen und $n = 0$. Bei negativen positiven Zahlen, verfehlen wir den
+Basis-Fall immer um genau 1 und landen danach in einer Endlosschleife. Bei negativen Zahlen, sind wir schon initial
+``unter`` dem Basis-Fall.
+
+\subsection{Geben Sie F als geschlossene nicht-rekursive Formel an und beweisen Sie Ihre Formel durch Induktion.}
+
+Geschlossene Formel für $F$:
+
+$$F(n) = \frac{n}{2} + \frac{n^2}{4}$$
+
+\begin{flushleft}
+	Induktionsbeweis:
+\end{flushleft}
+
+\begin{align*}
+	Induktionsvoraussetzung & = \frac{n}{2} + \frac{n^2}{4} = F(n)                           & n \in \{ x \mid x \in N_0 \wedge x \; mod \; 2 = 0 \} \\
+	Induktionsbehauptung    & = \frac{n + 2}{2} + \frac{(n + 2)^2}{4} = F(n + 2)                                                                     \\
+	Induktionsanfang        & = \frac{0}{2} + \frac{0^2}{4} = F(0) \Leftrightarrow 0 = 0                                                             \\
+	Induktionsschritt       & = \frac{n + 2}{2} + \frac{(n + 2)^2}{4}                                                                                \\
+	                        & = \frac{n}{2} + \frac{n^2 + 4n + 4}{4} + \frac{2}{2}                                                                   \\
+	                        & = \frac{n}{2} + \frac{n^2}{4} + \frac{4n + 4}{4} + \frac{2}{2}                                                         \\
+	                        & = F(n) + (n + 2) = F(n + 2)                                                                                            \\
+\end{align*}
+
 \end{document}

### 2021-04-15 algorithmen-und-datenstrukturen/assignment-1.tex
format aud assignment

@@
 
-\section{Zweitkleinstes Element einer Folge}
-
-Das zweitkleinste Element einer Folge von n $\geq$ 2 Zahlen soll bestimmt werden.
+\section{Zweitkleinstes Element einer Folge  von $n \geq 2$ Zahlen}
 
@@
 
+\begin{flushleft}
 	Somit können wir sagen, dass mit $c \geq 5$ und $n_0 = 1$ die Behauptung $f(n) \in O(n^2)$ gilt
+\end{flushleft}
 
@@
 
+\begin{flushleft}
 	Nun können wir $n_0$ als Schnittpunkt der beiden Funktionen $2n^2 - 72$ und $n^2$ berechnen.
+\end{flushleft}
 
@@
 
+\begin{flushleft}
 	Also, mit $c = 1$ und $n_0 = \lceil\sqrt{72}\rceil = 9$ gilt $f(n) \in \Omega(n^2)$
+\end{flushleft}
 
@@
 
+\newpage
+
 \subsection{Laufzeit-Analyse}
@@
 TBD
+
+\newpage
 
@@
 
+\subsection{Stellen Sie eine Rekursionsgleichung für $T(n)$ auf und geben Sie eine geschlossene Formel für $T(n)$ an.}
+
 \end{document}

### 2021-04-15 algorithmen-und-datenstrukturen/assignment-1.tex
add the solution for 4.3 to the aud assignment
format the aud assignment

@@
 
-Er braucht im \textbf{Worst-Case 2(n - 1)} Vergleiche und liegt daher in $O(n)$. Der Worst-Case kommt zustande wenn wir z.B. eine List der
-Länge n betrachten, die n mal das gleiche Element enthält. Dann benötigen wir beim der ersten und der zweiten Schleife $n - 1$ Vergleiche.
+Er braucht im \textbf{Worst-Case 2(n - 1)} Vergleiche und liegt daher in $O(n)$. Der Worst-Case kommt zustande wenn wir z.B. eine
+List der Länge n betrachten, die n mal das gleiche Element enthält. Dann benötigen wir beim durchlaufen der ersten und der zweiten
+Schleife $n - 1$ Vergleiche.
 
@@
 
+\vspace{1em}
+
 \begin{flushleft}
@@
 
-$\Theta(g)$ ist im Skript mit der \emph{Definition 2.5} als $\{ \, f \mid f \in O(g) \wedge f \in \Omega(g) \, \}$ definiert.\\[0.5em]
+\begin{flushleft}
+	$\Theta(g)$ ist im Skript mit der \emph{Definition 2.5} als $\{ \, f \mid f \in O(g) \wedge f \in \Omega(g) \, \}$ definiert.
+
+	\hspace{0.5em}
+
 	Somit wissen wir, dass $f(n) \in \Theta(n^2)$, da wir in 2.1 und 2.2 gezeigt haben, dass $f \in O(g)$ und $f \in \Omega(g)$ gelten.
+\end{flushleft}
 
@@
 
+Rekursionsgleichung:
+
+\begin{align*}
+	T(0) & = 0 \\
+	T(n) & = \, T(n - 2) + 2
+\end{align*}
+
+\vspace{1em}
+
+\begin{flushleft}
+	Geschlossene Formel für $T(n)$:
+\end{flushleft}
+
+$$T(n) = n$$
+
 \end{document}

### 2021-04-16 softwaretechnik/assignment-1.tex
refactor section 2 and add section 3 to the swt assignment

@@
 \usepackage[T1]{fontenc}
+\usepackage{hyperref}
+
+\hypersetup{
+    colorlinks=true,
+    linkcolor=black,
+    filecolor=black,      
+    urlcolor=black,
+}
 
@@
 		Fachbereich Informatik \\
-		Softwaretechnik
+		Softwaretechnik \\
+		Prof. Dr-Ing. Martin Schafföner
 	\end{flushleft}
@@
 
+\vspace{100px}
+\begin{center}
+	\large{To be inserted!}
+\end{center}
+\vspace{100px}
+
 \section{Projekt}
 
-Als Projekt für dieses Semester verwende ich ein privates Freizeit-Projekt von mir – einen kleinen,
-minimalistischen Tiling-Window-Manager für Linux bzw. X11.
+Als Projekt für dieses Semester verwende ich eine Idee die mir schon seit längerem im Kopf rumgeistert
+– einen kleinen, minimalistischen \emph{Tiling Window Manager} für Linux bzw. X11, der deutlich flexibler
+und entwicklerfreundlicher ist, als bestehende, vergleichbare Software.
 
@@
 
-Für den Fall, dass Sie noch keinen Kontakt mit dieser Domain, hatten gehe ich nun kurz darauf ein,
-welche Aufgaben ein Tiling-Window-Manager (kurz. \emph{TWM} oder nur \emph{WM}) typischer Weise übernimmt.
-Die generelle Aufgabe eines WM's besteht hauptsächlich darin sich um die Kommunikation mit dem
-Window-Server (i.d.R. X11 oder Wayland auf Linux, Quartz Compositor auf MacOS) zu kümmern, und die
+Für den Fall, dass Sie noch keinen Kontakt mit dieser Domain hatten, gehe ich im folgenden kurz darauf ein,
+welche Aufgaben ein \emph{Tiling Window Manager} (kurz. \emph{TWM} oder nur \emph{WM}) typischer Weise
+übernimmt. Die generelle Aufgabe eines WM's besteht hauptsächlich darin sich um die Kommunikation mit dem
+\emph{Window-Server} (i.d.R. X11 oder Wayland auf Linux, Quartz Compositor auf MacOS) zu kümmern, und die
 tatsächliche Anordnung der Fenster auf dem Bildschirm zu regeln (in Ebenen, Kacheln etc. – die Möglichkeiten
-sind nahe zu unbegrenzt). Typische Window-Manager von z.B. MacOS können diverse, für uns als Endnutzer
+sind nahe zu unbegrenzt). Typische \emph{Window Manager} von z.B. MacOS können diverse, für uns als Endnutzer
 mittlerweile als üblich angesehene, Anordnungen wie Floating, Split-Screen und Fullscreen realisieren. \par
-Ein Tiling-Window-Manager hat nun die Besonderheit, dass er anders als von Windows, MacOS oder diversen Linux
-Desktop-Umgebungen bekannt, die Aufteilung der Fenster automatisch und bestmöglich regelt. Typischerweise
-hat ein TWM eine feste Konfiguration mit Layouts, in die er die Fenster einsortieren kann (z.B. ``Master
-and Stack``). Somit muss sich der Anwender eines solchen TWM's (zumindest initial) nicht selber um die
-Anordnung kümmern.
+Ein \emph{Tiling Window Managers} hat nun die Besonderheit, dass er anders als von Windows, MacOS oder diversen
+Linux Desktop-Umgebungen bekannt, die Aufteilung der Fenster automatisch und bestmöglich regelt. Typischerweise
+hat ein \emph{TWM} eine feste Konfiguration mit Layouts, in die er die Fenster einsortieren kann (z.B. ``Master
+and Stack``). Somit muss sich der Anwender eines solchen \emph{TWM's} (zumindest initial) nicht selber um die
+Anordnung seiner Fenster kümmern. Desweiteren zählt die Möglichkeit Keybindings zur Navigation oder Veränderung
+der Aufteilung zu definieren zu den typischen Features eines \emph{Tiling Window Managers}.
 
@@
 
-Das Problem von bestehenden Window-Managern ist in der Regel deren Alter (bspw. ist das Projekt ``Toms
-Window Manager`` im Jahr 1987 entstanden). Aufgrund des Alters sind viele dieser Window-Manager noch in
-C, C++ oder einer vergleichbaren Sprache verfasst worden und mit der Zeit immer weiter gewachsen. Dies
-macht, nicht von den Entwicklern vorhergesehene, grundlegende Änderungen – zumindest für mich – äußert
-schwer.
+Das Problem von bestehenden \emph{Window Managern} ist in der Regel deren Alter (bspw. ist das Projekt ``Toms
+Window Manager`` im Jahr 1987 entstanden). Aufgrund des Alters sind viele dieser \emph{Window Manager} noch in
+C, C++ oder einer vergleichbaren Sprache verfasst worden und mit der Zeit immer weiter gewachsen. Dies macht es
+deutlich schwerer als es sein müsste, grundlegende Änderungen, die nicht von den Entwicklern vorhergesehen waren,
+vorzunehmen. \par
+Ziel ist es einen \emph{minimalen Window Manager} zu entwickeln, der \emph{durch eigenen Programm-Code konfiguriert}
+und erweitert werden kann, zu entwickeln. Dieses Konzept ist schon öfter Implementiert worden (als Inspiration für
+dieses Modell dienen die Projekte \emph{XMonad\footnotemark} und \emph{DWM\footnotemark}), allerdings sind diese
+Projekte leider meistens schlecht Dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}.
 
-% Da ich erstens C & C++ nicht behersche und die Projektgröße sowas ich ohnehin, die meisten Standardfunktionalitäten nicht 
-% "Rust TWM Library" um selber mit rust knowledge seine twm zusammen zu schrauben.
+\footnotetext[1]{XMonad wurde 2007 erstmals veröffentlicht. Dieses wurde in Haskell geschrieben und wird auch damit
+	konfiguriert. Siehe \url{https://xmonad.org/} für weitere Informationen.}
+\footnotetext[2]{DWM ist ein äußerst kleiner (unter 2000 Zeilen Source-Code) Window Manager geschrieben in C. Er
+	hat keine Konfigurationsdatei und wird durch Patchen des Codes konfiguriert. Siehe \url{https://dwm.suckless.org/}
+	für weitere Informationen.}
 
-\section{User Stories}
+\section{Erfassung der Anforderungen}
+
+Zur Erfassung der Anforderungen verwende ich das Projektmanagement Feature von GitHub. Dieses hat zumindest im
+kleinen Rahmen, diverse Vorteile gegenüber herkömmlicher Software für Agile Projekte (z.B. Jira von Atlassian).
+Unter anderem folgende:
+
+\begin{itemize}
+	\item Übersichtlicher durch kleineren Umfang und somit leichter zu Bedienen
+	\item Direkte Integration mit dem Entwicklungsprozess (Issues können z.B.
+	      geschlossen werden sobald ein Feature merged wurde)
+	\item Nutzer der Software können direkt Issues anlegen
+\end{itemize}
+
+Diese Entscheidung hat keine Endgültigkeit. Wenn es ein echtes Team mit einem klaren Vorgehensmodell gibt, können
+andere Tools besser geeignet sein. Außerdem sei erwähnt, dass Jira mit diversen Integrationen erweitert werden
+kann, um die oben beschriebene Funktionalität nach zu rüsten. Zur geordneten Erfassung von Anforderungen, deren
+priorisierung, kategorisierung und Abnahme reicht GitHub Projects allerdings vorerst vollkommen aus. Es gibt die
+Möglichkeit Sprint oder Kanban Boards anzulegen, diese zu automatisieren oder komplett eigene Prozesse zu
+integrieren.
+
+\vspace{100px}
+\begin{center}
+	\large{Bild Vom Fertigen Board einfügen!}
+\end{center}
+\vspace{100px}
 

### 2021-04-16 softwaretechnik/assignment-1.tex
fix spelling

@@
 
-Als Projekt für dieses Semester verwende ich eine Idee die mir schon seit längerem im Kopf rumgeistert
+Als Projekt für dieses Semester verwende ich eine Idee, die mir schon seit längerem im Kopf rumgeistert
 – einen kleinen, minimalistischen \emph{Tiling Window Manager} für Linux bzw. X11, der deutlich flexibler
-und entwicklerfreundlicher ist, als bestehende, vergleichbare Software.
+und entwicklerfreundlicher ist als bestehende, vergleichbare Software.
 
@@
 and Stack``). Somit muss sich der Anwender eines solchen \emph{TWM's} (zumindest initial) nicht selber um die
-Anordnung seiner Fenster kümmern. Desweiteren zählt die Möglichkeit Keybindings zur Navigation oder Veränderung
-der Aufteilung zu definieren zu den typischen Features eines \emph{Tiling Window Managers}.
+Anordnung seiner Fenster kümmern. Desweiteren zählt die Möglichkeit, Keybindings zur Navigation oder Veränderung
+der Aufteilung zu definieren, zu den typischen Features eines \emph{Tiling Window Managers}.
 
@@
 Ziel ist es einen \emph{minimalen Window Manager} zu entwickeln, der \emph{durch eigenen Programm-Code konfiguriert}
-und erweitert werden kann, zu entwickeln. Dieses Konzept ist schon öfter Implementiert worden (als Inspiration für
-dieses Modell dienen die Projekte \emph{XMonad\footnotemark} und \emph{DWM\footnotemark}), allerdings sind diese
-Projekte leider meistens schlecht Dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}.
+und erweitert werden kann. Dieses Konzept ist schon öfter Implementiert worden (als Inspiration für dieses Modell
+dienen die Projekte \emph{XMonad\footnotemark} und \emph{DWM\footnotemark}), allerdings sind diese Projekte leider
+meistens schlecht dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}.
 
@@
 Zur Erfassung der Anforderungen verwende ich das Projektmanagement Feature von GitHub. Dieses hat zumindest im
-kleinen Rahmen, diverse Vorteile gegenüber herkömmlicher Software für Agile Projekte (z.B. Jira von Atlassian).
-Unter anderem folgende:
+kleinen Rahmen diverse Vorteile gegenüber herkömmlicher Software für Agile Projekte (z.B. Jira von Atlassian).
+Unter anderem Folgende:
 
@@
 	\item Direkte Integration mit dem Entwicklungsprozess (Issues können z.B.
-	      geschlossen werden sobald ein Feature merged wurde)
+	      geschlossen werden sobald ein Feature gemerged wurde)
 	\item Nutzer der Software können direkt Issues anlegen
@@
 andere Tools besser geeignet sein. Außerdem sei erwähnt, dass Jira mit diversen Integrationen erweitert werden
-kann, um die oben beschriebene Funktionalität nach zu rüsten. Zur geordneten Erfassung von Anforderungen, deren
-priorisierung, kategorisierung und Abnahme reicht GitHub Projects allerdings vorerst vollkommen aus. Es gibt die
+kann, um die oben beschriebene Funktionalität nachzurüsten. Zur geordneten Erfassung von Anforderungen, deren
+Priorisierung, Kategorisierung und Abnahme, reicht GitHub Projects allerdings vorerst vollkommen aus. Es gibt die
 Möglichkeit Sprint oder Kanban Boards anzulegen, diese zu automatisieren oder komplett eigene Prozesse zu

### 2021-04-16 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
rename the swt assignment and add a figure

--- 2021-04-16 softwaretechnik/assignment-1.tex
@@
 \usepackage{hyperref}
+\usepackage{graphicx}
+
+\graphicspath{ {./images/} }
 
@@
 
-\footnotetext[1]{XMonad wurde 2007 erstmals veröffentlicht. Dieses wurde in Haskell geschrieben und wird auch damit
+\footnotetext[1]{wurde 2007 erstmals veröffentlicht. Wurde in Haskell geschrieben und wird auch damit
 	konfiguriert. Siehe \url{https://xmonad.org/} für weitere Informationen.}
-\footnotetext[2]{DWM ist ein äußerst kleiner (unter 2000 Zeilen Source-Code) Window Manager geschrieben in C. Er
+\footnotetext[2]{ist ein äußerst kleiner (unter 2000 Zeilen Source-Code) Window Manager geschrieben in C. Er
 	hat keine Konfigurationsdatei und wird durch Patchen des Codes konfiguriert. Siehe \url{https://dwm.suckless.org/}
@@
 kann, um die oben beschriebene Funktionalität nachzurüsten. Zur geordneten Erfassung von Anforderungen, deren
-Priorisierung, Kategorisierung und Abnahme, reicht GitHub Projects allerdings vorerst vollkommen aus. Es gibt die
+Priorisierung, Kategorisierung und Abnahme, reicht \emph{GitHub Projects} allerdings vorerst vollkommen aus. Es gibt die
 Möglichkeit Sprint oder Kanban Boards anzulegen, diese zu automatisieren oder komplett eigene Prozesse zu
@@
 
-\vspace{100px}
-\begin{center}
-	\large{Bild Vom Fertigen Board einfügen!}
-\end{center}
-\vspace{100px}
+\vspace{1em}
+
+\begin{figure}[h]
+	\includegraphics[width=0.6\textwidth]{github-projects}
+	\centering
+	\caption{\emph{GitHub Projects} Board}
+\end{figure}
+
+\newpage
+
+\section{Anforderungen}
 

### 2021-04-17 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
added project targets to swt assignment
fix title pages

@@
 		Online Studiengang Medieninformatik \\
-		Fachbereich Informatik \\
+		Fachbereich Informatik und Medien \\
 		Softwaretechnik \\
@@
 dienen die Projekte \emph{XMonad\footnotemark} und \emph{DWM\footnotemark}), allerdings sind diese Projekte leider
-meistens schlecht dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}.
+meistens schlecht dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}. Zu den typischen Anwendern dieser
+\emph{Window Manager} gehören Entwickler und Power User.
 
@@
 kann, um die oben beschriebene Funktionalität nachzurüsten. Zur geordneten Erfassung von Anforderungen, deren
-Priorisierung, Kategorisierung und Abnahme, reicht \emph{GitHub Projects} allerdings vorerst vollkommen aus. Es gibt die
-Möglichkeit Sprint oder Kanban Boards anzulegen, diese zu automatisieren oder komplett eigene Prozesse zu
-integrieren.
+Priorisierung, Kategorisierung und Abnahme, reicht \emph{GitHub Projects} allerdings vorerst vollkommen aus. Es
+gibt die Möglichkeit Sprint oder Kanban Boards anzulegen, diese zu automatisieren oder komplett eigene Prozesse
+zu integrieren.
 
-\vspace{1em}
+\vspace{0.5em}
 

### 2021-04-17 mensch-computer-interaktion/assignment-1.tex
use tabularx to format tables

--- 2021-04-14 mensch-computer-interaktion/esa-1.tex
@@
 \usepackage[T1]{fontenc}
+\usepackage{tabularx}
 
@@
 		Online Studiengang Medieninformatik \\
-		Fachbereich Informatik \\
+		Fachbereich Informatik und Medien \\
 		Mensch-Computer-Interaktion
@@
 
-\section{Persona: Mark}
+\section{Persona: Mark Zahn}
 
-\begin{tabularx}{0.8\textwidth} { 
-	| >{\raggedright\arraybackslash}X 
-	| >{\centering\arraybackslash}X 
-	| >{\raggedleft\arraybackslash}X | }
+\begin{tabularx}{\textwidth}{|l|X|}
 	\hline
-	item 11 & item 12 & item 13 \\
+	Name                & Marc Zahn                                           \\
 	\hline
-	item 21  & item 22  & item 23  \\
+	Alter               & 28                                                  \\
+	\hline
+	Beruf               & Hochschul-Verwaltungsangestellter                   \\
+	\hline
+	Bildungsabschluss   & Abitur                                              \\
+	\hline
+	Familienstand       & Verheiratet                                         \\
+	\hline
+	Kinder              & Ja                                                  \\
+	\hline
+	Motto               & ""                                                  \\
+	\hline
+	Motivation für OSMI & Möchte nun Umlernen, bekam Kontakt zum Studium als
+	Umschulungsmöglichkeit, durch Job an der Hochschule                       \\
+	\hline
+	Hobbies             & Alte Filme sammeln, Kochen, Wandern                 \\
+	\hline
+	Wohnort             & Brandenburg an der Havel                            \\
+	\hline
+	OSMI-Student seit   & WS 2019                                             \\
+	\hline
+	Bedürfnisse         & Erfolg, Soziale Anerkennung, Sicherheit für Familie \\
+	\hline
+	Wünsche             & Neuer Job in Wunschbereich                          \\
+	\hline
+	Werte               & Treue, Zuverlässigkeit, Ehrlichkeit                 \\
+	\hline
+	Ängste              & Angst zu Versagen, Unsichere Zukunft                \\
+	\hline
+	Persönlichkeit      & Introvertiert                                       \\
 	\hline
@@
 
-Name
-
-"quote"
-
-OSMI Studenten, die berufsbegleitend studieren
-Alter
-Beruf höchstem Bildungsabschluss (Realschulabschluss/ Abitur/ Ausbildung/Studium)
-Familienstand,
-Kinder?
-Motivation für Entscheidung für OSMI?
-Hobbies
-Wohnort (könnte z.B. auch auf dem Land wohnen oder halt in Berlin)
-OSMI-Student seit wann?
-wie verbringe ich den Tag?
-Konsumverhalten?
-Bedürfnisse
-Wünsche
-Werte
-Ängste
-Introvertiert
-Extrovertiert
-
 \newpage
 
-\section{Persona: XYZ}
+\section{Persona: Rosa Beetz}
+
+\begin{tabularx}{\textwidth}{|l|X|}
+	\hline
+	Name                & Rosa Beetz                                          \\
+	\hline
+	Alter               & 23                                                  \\
+	\hline
+	Beruf               & Screen Desginerin                                   \\
+	\hline
+	Bildungsabschluss   & Ausbildung                                          \\
+	\hline
+	Familienstand       & Ledig                                               \\
+	\hline
+	Kinder              & Nein                                                \\
+	\hline
+	Motto               & ""                                                  \\
+	\hline
+	Motivation für OSMI & Möchte sich in Richtung Informatik weiterentwickeln
+	und sich selbst etwas Beweisen                                            \\
+	\hline
+	Hobbies             & Netflix, Feiern, Kaffee                             \\
+	\hline
+	Wohnort             & Berlin Wedding                                      \\
+	\hline
+	OSMI-Student seit   & SS 2021                                             \\
+	\hline
+	Bedürfnisse         & Sozialer Kontakt, Anerkennung                       \\
+	\hline
+	Wünsche             & Mehr Geld, Anerkennung im Job                       \\
+	\hline
+	Werte               & Freundlichkeit, Ehrgeiz                             \\
+	\hline
+	Ängste              & Angst zu Versagen                                   \\
+	\hline
+	Persönlichkeit      & Extrovertiert                                       \\
+	\hline
+\end{tabularx}
+
+% Name					Rosa Beetz
+% Alter					23
+% Beruf					Screen Desginerin
+% Bildungsabschluss		Ausbildung
+% Familienstand			Ledig
+% Kinder					Nein
+% Motto					""
+% Motivation für OSMI
 
@@
 
-\newpage
-
-\section{Storyboard 3: XYZ}
+Als OSMI Student möchte live mit anderen Studenten in einem Whiteboard oder einer PDF zusammenarbeiten
 
@@
 
-\section{Storyboard 4: XYZ}
+\section{Storyboard 3: Rosa Beetz}
+
+Als OSMI Student möchte ich Fahrgemeinschaften gründen
+
+\newpage
+
+\section{Storyboard 4: Rosa Beetz}
+
+Als OSMI Student möchte ich mit anderen Studenten Videotelefonieren
 

### 2021-04-17 mensch-computer-interaktion/assignment-1.tex
add descriptive quotes to each persona

@@
 \usepackage{tabularx}
+\usepackage{graphicx}
+
+\graphicspath{ {./images/} }
 
@@
 
+\begin{figure}
+
+\end{figure}
+
+\begin{quote}
+	\large{``Ich verbringe meine Zeit gerne mit meiner Familie. Wenn ich Abends noch studiere, dann versuche
+		ich möglichst effizient durch den Stoff zu kommen. Bisher habe ich eher selten in Lerngruppen gearbeitet,
+		da mir der Austausch mit anderen Studenten immer sehr Zeitaufwendig vorkam.``}
+\end{quote}
+
+\begin{center}
 	\begin{tabularx}{\textwidth}{|l|X|}
@@
 		Kinder              & Ja                                                  \\
-	\hline
-	Motto               & ""                                                  \\
 		\hline
@@
 	\end{tabularx}
+\end{center}
 
@@
 
+\begin{quote}
+	\large{``Ich bin gerne mit anderen Menschen zusammen und finde, dass das Studium
+		zusammen viel mehr Spaß macht und man besser lernt. Deshalb habe ich mit einigen
+		Kommilitonen eine Lerngruppe gegründet.``}
+\end{quote}
+
+\begin{center}
 	\begin{tabularx}{\textwidth}{|l|X|}
@@
 		Kinder              & Nein                                                \\
-	\hline
-	Motto               & ""                                                  \\
 		\hline

### 2021-04-17 mensch-computer-interaktion/assignments/personas-and-storyboards/personas-and-storyboards.tex
move the mci assignment into a seperate dir

--- 2021-04-17 mensch-computer-interaktion/assignment-1.tex
@@
 
-\begin{figure}
-
+\begin{figure}[h]
+	\includegraphics[width=0.25\textwidth]{mark}
+	\centering
+	\caption{Portrait eines Mannes von Roman Holoschchuk auf Unsplash}
 \end{figure}
@@
 	\begin{tabularx}{\textwidth}{|l|X|}
+		\hline
+		\textbf{Demographische Daten} &                                                     \\
 		\hline
@@
 		\hline
-		Motivation für OSMI & Möchte nun Umlernen, bekam Kontakt zum Studium als
-		Umschulungsmöglichkeit, durch Job an der Hochschule                       \\
+		Wohnort                       & Brandenburg an der Havel                            \\
+		\hline
+		\textbf{Persönlichkeit}       &                                                     \\
+		\hline
+		Motivation für OSMI           & Möchte nun Umlernen, bekam Kontakt zum OSMI als
+		Umschulungsmöglichkeit, durch seinen Job an einer Hochschule                        \\
 		\hline
 		Hobbies                       & Alte Filme sammeln, Kochen, Wandern                 \\
-		\hline
-		Wohnort             & Brandenburg an der Havel                            \\
 		\hline
@@
 		\hline
-		Persönlichkeit      & Introvertiert                                       \\
+		Persönlichkeitstyp            & Introvertiert                                       \\
 		\hline
@@
 \section{Persona: Rosa Beetz}
+
+\begin{figure}[h]
+	\includegraphics[width=0.25\textwidth]{rosa}
+	\centering
+	\caption{Portrait einer Frau von Svyatoslav Romanov auf Unsplash}
+\end{figure}
 
@@
 		\hline
+		Wohnort                       & Berlin Wedding                                      \\
+		\hline
+		\textbf{Persönlichkeit}       &                                                     \\
+		\hline
 		Motivation für OSMI           & Möchte sich in Richtung Informatik weiterentwickeln
@@
 		Hobbies                       & Netflix, Feiern, Kaffee                             \\
-		\hline
-		Wohnort             & Berlin Wedding                                      \\
 		\hline
@@
 		\hline
-		Persönlichkeit      & Extrovertiert                                       \\
+		Persönlichkeitstyp            & Extrovertiert                                       \\
 		\hline
@@
 \end{center}
-
-% Name					Rosa Beetz
-% Alter					23
-% Beruf					Screen Desginerin
-% Bildungsabschluss		Ausbildung
-% Familienstand			Ledig
-% Kinder					Nein
-% Motto					""
-% Motivation für OSMI
 
@@
 
-Als OSMI Student möchte live mit anderen Studenten in einem Whiteboard oder einer PDF zusammenarbeiten
+Als OSMI Student möchte ich unkompliziert mit anderen Studenten in einer PDF zusammenarbeiten
 

### 2021-04-17 mensch-computer-interaktion/assignments/personas-and-storyboards/personas-and-storyboards.tex
add placeholders to insert the storyboards into

@@
 
-\section{Storyboard 1: Mark}
+\section{Storyboard 1: Mark Zahn}
+
+\begin{figure}[h]
+	\includegraphics[angle=90,width=\textwidth]{mark}
+	\centering
+	\caption{Storyboard Mark – Als OSMI Studen möchte ich xyz}
+\end{figure}
 
@@
 
-\section{Storyboard 2: Mark}
+\section{Storyboard 2: Mark Zahn}
 
-Als OSMI Student möchte ich unkompliziert mit anderen Studenten in einer PDF zusammenarbeiten
+\begin{figure}[h]
+	\includegraphics[angle=90,width=\textwidth]{mark}
+	\centering
+	\caption{Storyboard Mark – Als OSMI Student möchte ich unkompliziert mit anderen Studenten in einer PDF zusammenarbeiten, um den Zeitaufwand der Zusammenarbeit zu minimieren.}
+\end{figure}
 
@@
 
-Als OSMI Student möchte ich Fahrgemeinschaften gründen
+\begin{figure}[h]
+	\includegraphics[angle=90,width=\textwidth]{rosa}
+	\centering
+	\caption{Storyboard Rosa – Als OSMI Student möchte ich Fahrgemeinschaften gründen, um mir Fahrtkosten zu sparen.}
+\end{figure}
 
@@
 
-Als OSMI Student möchte ich mit anderen Studenten Videotelefonieren
+\begin{figure}[h]
+	\includegraphics[angle=90,width=\textwidth]{rosa}
+	\centering
+	\caption{Storyboard Rosa – Als OSMI Student möchte ich mit anderen Studenten Videotelefonieren, um eine zugängliche Plattform für unsere digitale Lerngruppe zu haben.}
+\end{figure}
 

### 2021-04-17 mensch-computer-interaktion/assignments/personas-and-storyboards/personas-and-storyboards.tex
update captions
fix the section hierachy

@@
 		Fachbereich Informatik und Medien \\
-		Mensch-Computer-Interaktion
+		Mensch-Computer-Interaktion \\
+		Prof. Dr. Martin Christof Kindsmüller
 	\end{flushleft}
@@
 
-\section{Persona: Mark Zahn}
+\section{Personas}
+
+\subsection{Mark Zahn}
 
@@
 	\centering
-	\caption{Portrait eines Mannes von Roman Holoschchuk auf Unsplash}
+	\caption{Portrait von ``Mark Zahn`` (Roman Holoschchuk, unsplash.com)}
 \end{figure}
@@
 
-\section{Persona: Rosa Beetz}
+\subsection{Rosa Beetz}
 
@@
 	\centering
-	\caption{Portrait einer Frau von Svyatoslav Romanov auf Unsplash}
+	\caption{Portrait von ``Rosa Beetz`` (Svyatoslav Romanov, unsplash.com)}
 \end{figure}
@@
 
-\section{Storyboard 1: Mark Zahn}
+\section{Storyboards}
+
+\subsection{Storyboard 1 für Mark Zahn}
 
@@
 
-\section{Storyboard 2: Mark Zahn}
+\subsection{Storyboard 2 für Mark Zahn}
 
@@
 
-\section{Storyboard 3: Rosa Beetz}
+\subsection{Storyboard 3 für Rosa Beetz}
 
@@
 
-\section{Storyboard 4: Rosa Beetz}
+\subsection{Storyboard 4 für Rosa Beetz}
 
@@
 
+\section{Einsatzmöglichkeiten für Präsenzstudenten}
+
+Foo baz
+
 \end{document}

### 2021-04-18 kommunikationsnetze-1/assignment-1.tex
add kn1 assignment 1

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\usepackage{enumerate}

\renewcommand\thesubsection{\alph{subsection})}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik und Medien \\
		Kommunikationsnetze 1 \\
		Prof. Dr.-Ing. habil. Michael Syrjakow
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 1}\\[0.5em]
		\large{Sommersemester 2021}\\[0.25em]
		\large{Abgabetermin 18.04.2021}
	\end{center}

	\vfill

	\begin{flushright}
		Maximilian Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\newpage

\section{Datensicherungsschicht}

\subsection{Bestimmen Sie die Quer- und Längsparität (gerade Parität) bei dem zum übertragenden Datenblock}

\begin{center}
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
		\hline
		0                    & 1                    & 1                    & 0                    & 1                    & 1                    & 0                    & 0                    & \cellcolor{gray!25}0 \\
		\hline
		1                    & 1                    & 0                    & 0                    & 0                    & 1                    & 1                    & 0                    & \cellcolor{gray!25}0 \\
		\hline
		0                    & 0                    & 0                    & 1                    & 1                    & 1                    & 0                    & 0                    & \cellcolor{gray!25}1 \\
		\hline
		\cellcolor{gray!25}1 & \cellcolor{gray!25}0 & \cellcolor{gray!25}1 & \cellcolor{gray!25}1 & \cellcolor{gray!25}0 & \cellcolor{gray!25}1 & \cellcolor{gray!25}1 & \cellcolor{gray!25}0 & \cellcolor{gray!25}1 \\
		\hline
	\end{tabular}
\end{center}

\subsection{An die zu übertragende Bitfolge \texttt{1100011011} (Nutzdaten) soll zur Fehlersicherung ein 4 Bit langer Kontrollblock (Frame Check Sequence) angehängt werden. Das Generatorpolynom sei \texttt{10011}. Berechnen Sie den Kontrollblock mittels Polynomdivision.}

\begin{verbatim}
    1 1 0 0 0 1 1 0 1 1 0 0 0 0
    1 0 0 1 1
    ---------
    0 1 0 1 1 1
      1 0 0 1 1
      ---------
      0 0 1 0 0 1 0
          1 0 0 1 1
          ---------
          0 0 0 0 1 1 1 0 0
                  1 0 0 1 1
                  ---------
                  0 1 1 1 1 0
                    1 0 0 1 1
                    ---------
                    0 1 1 0 1 0
                      1 0 0 1 1
                      ---------
                      0 1 0 0 1
\end{verbatim}

Dadurch ergibt sich der Kontrollblock \texttt{1001}

\subsection{Wie groß ist die MTU bei Ethernet?}

Bei Ethernet II 1500 Byte und bei IEEE 802.3 1492 Byte.

\subsection{Welches Vielfachzugriffsverfahren wird von Ethernet benutzt?}

\emph{Carrier Sense Multiple Access/Collision Detection} (kurz \emph{CSMCA/CD})

\section{Internet-Protokoll}

\subsection{Wie wird das Endlos-Kreisen von IP-Paketen im Netz verhindert?}

Durch die Header TTL bei IPv4 bzw. Hop-Limit bei IPv6 werden die maximale Anzahl der Weiterleitungen über Router festgelegt.
Er wird bei jeder Weiterleitungen durch einen Router um 1 verringert – erreicht er den Wert 0 wird das betroffene IP-Paket verworfen.

\subsection{Welche verschiedenen Klassen von IP-Adressen gibt es?}

Bis 1993 wurden IP-Adressen in 5 Klassen eingeteilt:

\begin{center}
	\begin{tabular}{|l|l|l|l|l|}
		\hline
		\textbf{Klasse} & \textbf{Beginnt mit} & \textbf{Subnetzmaske}  & \textbf{Netzeanzahl} & \textbf{Hostanzahl} \\
		\hline
		A               & 0-128            & \texttt{255.0.0.0}     & $2^7$                & $2^{24} - 2$        \\
		\hline
		B               & 129-191          & \texttt{255.255.0.0}   & $2^14$               & $2^{16} - 2$        \\
		\hline
		C               & 192-223          & \texttt{255.255.255.0} & $2^21$               & $2^{8} - 2$         \\
		\hline
		D               & 224-239          & Nicht vorhanden        &                      &                     \\
		\hline
		E               & 240-255          & Nicht vorhanden        &                      &                     \\
		\hline
	\end{tabular}
\end{center}

\subsection{Wie viele Hosts können in einem Class C Netz maximal installiert werden?}

$2^8 - 2 = 254$

\subsection{Zu welcher Klasse gehört die IP-Adresse 129.3.1.13? Ist das eine Netz oder Host-Adresse?}

Da diese Adresse mit 129 beginnt, gehört sie zur Klasse B und hat die Subnetzmasken 255.255.0.0.
Dementsprechend ist der für diese Frage relevante Teil \texttt{1.13}. In diesen beiden Bytes sind
weder alle Bits auf 0 noch alle Bits auf 1, deshalb handelt es sich um eine Host-Adresse.

\subsection{Kann die IP-Adresse 192.168.128.4 im Internet vorkommen}

Nein, Adressen die mit \texttt{192.168} Anfangen sind immer privat.

\subsection{Ist 192.256.20.132 eine gültige IP-Adresse?}

Nein \texttt{256} kann man nicht in einem Byte codieren.

\subsection{Was bedeutet Subnetting?}

Durch Subnetting lassen sich größere Netze in kleinere aufteilen, in dem der Host-Anteil eines zugewiesenen IP-Adressbereichs
noch weiter in Subnetz-ID und Host-ID unterteilt wird. Somit können Adressbereiche effizienter ausgenutzt werden.

\subsection{Sie benötigen im Netz 198.200.40.0 mehrere Subnetze, wobei in jedem Subnetz bis maximal 13 Hosts installiert werden sollen. Definieren Sie alle möglichen Subnetzmasken}

\begin{center}
	\begin{tabular}{|l|l|l|}
		\hline
		\textbf{Subnetzmaske}    & \textbf{Subnetze} & \textbf{Hosts} \\
		\hline
		\texttt{255.255.255.240} & $2^4$             & $2^4 - 2$   \\
		\hline
		\texttt{255.255.255.224} & $2^3$             & $2^5 - 2$   \\
		\hline
		\texttt{255.255.255.192} & $2^2$             & $2^6 - 2$    \\
		\hline
		\texttt{255.255.255.128} & $2^1$             & $2^7 - 2$    \\
		\hline
	\end{tabular}
\end{center}

\end{document}

### 2021-04-21 algorithmen-und-datenstrukturen/assignment-1.tex
fix aud assignment title page
fix title pages
fix aud spelling

@@
 		Online Studiengang Medieninformatik \\
-		Fachbereich Informatik \\
-		Algorithmen und Datenstrukturen
+		Fachbereich Informatik und Medien \\
+		Algorithmen und Datenstrukturen \\
+		Prof. Dr. rer. nat. Ulrich Baum
 	\end{flushleft}
@@
 
-Die Funktion terminiert nur für gerade postive Zahlen und $n = 0$. Bei negativen positiven Zahlen, verfehlen wir den
+Die Funktion terminiert nur für gerade postive Zahlen und $n = 0$. Bei ungeraden positiven Zahlen, verfehlen wir den
 Basis-Fall immer um genau 1 und landen danach in einer Endlosschleife. Bei negativen Zahlen, sind wir schon initial

### 2021-04-23 mensch-computer-interaktion/assignments/personas-and-storyboards/personas-and-storyboards.tex
fix rosas persona
add missing escape
fix mci userstories
refactor personas

@@
 \usepackage{graphicx}
+\usepackage{fancyhdr}
 
 \graphicspath{ {./images/} }
+
+\pagestyle{fancyplain}
+\fancyhf{}
+\lhead{\fancyplain{}{Maximilian Schulke} }
+\rhead{\fancyplain{}{\today}}
+\cfoot{\fancyplain{}{\thepage}}
 
@@
 		\large{Sommersemester 2021}\\[0.25em]
-		\large{Abgabetermin 18.04.2021}
+		\large{Abgabetermin 23.04.2021}
 	\end{center}
@@
 
-\subsection{Mark Zahn}
+\subsection{Mark Schneider - Der Effiziente}
 
@@
 	\centering
-	\caption{Portrait von ``Mark Zahn`` (Roman Holoschchuk, unsplash.com)}
+	\caption{Mark Schneider – thispersondoesnotexist.com}
 \end{figure}
 
+\vspace{-1em}
+
 \begin{quote}
-	\large{``Ich verbringe meine Zeit gerne mit meiner Familie. Wenn ich Abends noch studiere, dann versuche
+	``Ich verbringe meine Zeit gerne mit meiner Familie. Wenn ich Abends noch studiere, dann versuche
 	ich möglichst effizient durch den Stoff zu kommen. Bisher habe ich eher selten in Lerngruppen gearbeitet,
-		da mir der Austausch mit anderen Studenten immer sehr Zeitaufwendig vorkam.``}
+	da mir der Austausch mit anderen Studenten immer sehr Zeitaufwendig vorkam.`` - Mark
 \end{quote}
+
+\vspace{0em}
+
+\begin{flushleft}
+	\textbf{Verhaltensweisen} \\
+	Überspringt meistens die Skripte und versucht direkt die Aufgaben zu bearbeiten um schnell durchzukommen.
+	Lernt gerne durch Videos, und schaut diese auf 1.5-facher Geschwindigkeit. Versucht mit so wenig Einsatz
+	wie möglich durch die Klausur zu kommen.\\
+	\vspace{1em}
+
+	\textbf{Motivation für OSMI} \\
+	Möchte einen Karrieresprung machen, um mehr Geld für seine Familie zu verdienen - er bekam Kontakt zum
+	OSMI als Weiterbildungsprogramm in seinem Job an einer Hochschule. \\
+
+	\vspace{1em}
+
+	\textbf{Motivation für Lerngruppen} \\
+	Mark hat durch seine relativ begrenzten Zeiteinsatz öfter Probleme und ist dankbar wenn er unkompliziert mit
+	anderen seine Probleme besprechen und lösen kann.
+\end{flushleft}
+
+\vspace{1em}
 
@@
 		\hline
-		\textbf{Demographische Daten} &                                                     \\
+		\textbf{Eckdaten}  &                                     \\
 		\hline
-		Name                          & Marc Zahn                                           \\
+		Name               & Marc Schneider                      \\
 		\hline
@@
 		\hline
-		\textbf{Persönlichkeit}       &                                                     \\
-		\hline
-		Motivation für OSMI           & Möchte nun Umlernen, bekam Kontakt zum OSMI als
-		Umschulungsmöglichkeit, durch seinen Job an einer Hochschule                        \\
-		\hline
 		Hobbies            & Alte Filme sammeln, Kochen, Wandern \\
@@
 		OSMI-Student seit  & WS 2019                             \\
-		\hline
-		Bedürfnisse                   & Erfolg, Soziale Anerkennung, Sicherheit für Familie \\
-		\hline
-		Wünsche                       & Neuer Job in Wunschbereich                          \\
-		\hline
-		Werte                         & Treue, Zuverlässigkeit, Ehrlichkeit                 \\
-		\hline
-		Ängste                        & Angst zu Versagen, Unsichere Zukunft                \\
 		\hline
@@
 
-\subsection{Rosa Beetz}
+\subsection{Rosa Haas - Die Motivierte}
 
@@
 	\centering
-	\caption{Portrait von ``Rosa Beetz`` (Svyatoslav Romanov, unsplash.com)}
+	\caption{Rosa Haas – thispersondoesnotexist.com}
 \end{figure}
@@
 \begin{quote}
-	\large{``Ich bin gerne mit anderen Menschen zusammen und finde, dass das Studium
+	``Ich bin gerne mit anderen Menschen zusammen und finde, dass das Studium
 	zusammen viel mehr Spaß macht und man besser lernt. Deshalb habe ich mit einigen
-		Kommilitonen eine Lerngruppe gegründet.``}
+	Kommilitonen eine Lerngruppe gegründet.`` - Rosa
 \end{quote}
+
+\vspace{0em}
+
+\begin{flushleft}
+	\textbf{Verhaltensweisen} \\
+	Fängt bereits sehr früh an für Klausuren zu lernen, da sie gerne die Situation unter Kontrolle hat.
+	Ist sehr interessiert an Informatik und schaut auch gerne über den Modulinhalt hinaus.
+	\vspace{1em}
+
+	\textbf{Motivation für OSMI} \\
+	Möchte sich in Richtung Informatik weiterentwickeln und sich selbst etwas Beweisen. Möchte letztendlich
+	auch mehr Geld durch das Studium verdienen, aber die neugier war ausschalggebend für die Enschreibung. \\
+	\vspace{1em}
+
+	\textbf{Motivation für Lerngruppen} \\
+	Rosa trifft sich gerne mit anderen zum lernen und tauscht sich gerne über den Stoff aus, da sie den
+	Kontakt zu anderen beim Online-Studiunm etwas vermisst.
+\end{flushleft}
+
+\vspace{1em}
 
@@
 		\hline
-		\textbf{Demographische Daten} &                                                     \\
+		\textbf{Eckdaten}  &                                              \\
 		\hline
-		Name                          & Rosa Beetz                                          \\
+		Name               & Rosa Haas                                    \\
 		\hline
@@
 		\hline
-		\textbf{Persönlichkeit}       &                                                     \\
-		\hline
-		Motivation für OSMI           & Möchte sich in Richtung Informatik weiterentwickeln
-		und sich selbst etwas Beweisen                                                      \\
-		\hline
-		Hobbies                       & Netflix, Feiern, Kaffee                             \\
+		Hobbies            & Netflix, Feiern, guter Kaffee \& Kreativität \\
 		\hline
 		OSMI-Student seit  & SS 2021                                      \\
-		\hline
-		Bedürfnisse                   & Sozialer Kontakt, Anerkennung                       \\
-		\hline
-		Wünsche                       & Mehr Geld, Anerkennung im Job                       \\
-		\hline
-		Werte                         & Freundlichkeit, Ehrgeiz                             \\
-		\hline
-		Ängste                        & Angst zu Versagen                                   \\
 		\hline
@@
 
-\subsection{Storyboard 1 für Mark Zahn}
+\subsection{Storyboard 1 für Mark}
 
 \begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{mark}
+	\includegraphics[width=\textwidth]{storyboard-1-mark}
 	\centering
-	\caption{Storyboard Mark – Als OSMI Studen möchte ich xyz}
+	\caption{Storyboard Mark – Als OSMI Student möchte ich über Lerngruppen-Termine benachrichtigt werden, da ich diese sonst u.U. verpasse.}
 \end{figure}
@@
 
-\subsection{Storyboard 2 für Mark Zahn}
+\subsection{Storyboard 2 für Mark}
 
 \begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{mark}
+	\includegraphics[width=\textwidth]{storyboard-2-mark}
 	\centering
@@
 
-\subsection{Storyboard 3 für Rosa Beetz}
+\subsection{Storyboard 3 für Rosa}
 
 \begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{rosa}
+	\includegraphics[width=\textwidth]{storyboard-3-rosa}
 	\centering
-	\caption{Storyboard Rosa – Als OSMI Student möchte ich Fahrgemeinschaften gründen, um mir Fahrtkosten zu sparen.}
+	\caption{Storyboard Rosa – Als OSMI Student möchte ich Fragen in ein Forum posten, um bei Problemen schnell Hilfe von anderen Studenten zu bekommen.}
 \end{figure}
@@
 
-\subsection{Storyboard 4 für Rosa Beetz}
+\subsection{Storyboard 4 für Rosa}
 
 \begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{rosa}
+	\includegraphics[width=\textwidth]{storyboard-4-rosa}
 	\centering
-	\caption{Storyboard Rosa – Als OSMI Student möchte ich mit anderen Studenten Videotelefonieren, um eine zugängliche Plattform für unsere digitale Lerngruppe zu haben.}
+	\caption{Storyboard Rosa – Als OSMI Student möchte ich Fahrgemeinschaften gründen, um mir Fahrtkosten zu sparen und die Umwelt zu schonen.}
 \end{figure}
 
-\newpage
-
-\section{Einsatzmöglichkeiten für Präsenzstudenten}
-
-Foo baz
-
 \end{document}

### 2021-04-25 algorithmen-und-datenstrukturen/assignment-1.tex
add solution for 3c in the aud assignment

@@
 
+\renewcommand\thesubsection{\alph{subsection})}
+
 \begin{document}
@@
 		\large{Sommersemester 2021}\\[0.25em]
-		\large{Abgabetermin 18.04.2021}
+		\large{Abgabetermin 25.04.2021}
 	\end{center}
@@
 
-	Somit wissen wir, dass $f(n) \in \Theta(n^2)$, da wir in 2.1 und 2.2 gezeigt haben, dass $f \in O(g)$ und $f \in \Omega(g)$ gelten.
+	Somit wissen wir, dass $f(n) \in \Theta(n^2)$, da wir in 2 a) und 2 b) gezeigt haben, dass $f \in O(g)$ und $f \in \Omega(g)$ gelten.
 \end{flushleft}
@@
 die Gesamtanzahl durch die Anzahl der Elemente geteilt werden. Also entweder $\sum_{i=0}^{k - 1} \frac{(i + 1) * 2^i}{2^k - 1}$
-$\frac{\sum_{i=0}^{k - 1} (i + 1) * 2^i}{2^k - 1}$
+oder alternativ $\frac{1}{2^k - 1}\sum_{i=0}^{k - 1} (i + 1) * 2^i$
 
-\newpage
+\subsection{Laufzeit-Analyse für Average-Case und vergleich mit Worst-Case}
 
-\subsection{Laufzeit-Analyse}
+\vspace{1em}
 
-TBD
+$$\frac{1}{2^k - 1}\sum_{i=0}^{k - 1} (i + 1) * 2^i = \frac{1}{2^k - 1}\sum_{i=1}^k i * 2^{i - 1} = k + \frac{k}{2^k - 1} - 1$$
 
-\newpage
+\begin{center}
+	Von \texttt{http://www.mcs.sdsmt.edu/ecorwin/cs251/binavg/binavg.html}
+\end{center}
+
+\vspace{1em}
+
+\begin{flushleft}
+	Der Worst-Case der binären Suche ist $O(\log_2 n)$. Wenn wir in der o.g. geschlossenen Formel $k$
+	durch $\lceil\log_2 n\rceil$ ersetzen erhalten wir:
+\end{flushleft}
+
+\vspace{0em}
+
+$$\lceil\log_2 n\rceil + \frac{\lceil\log_2 n\rceil}{n} - 1$$
+
+\vspace{0em}
+
+\begin{flushleft}
+	Wenn wir diese Gleichung nun asymptotisch betrachten entfällt $\frac{\lceil\log_2 n\rceil}{n}$
+\end{flushleft}
+
+\vspace{0em}
+
+$$\lim_{n\to\infty} \lceil\log_2 n\rceil + \frac{\lceil\log_2 n\rceil}{n} - 1 = \lceil\log_2 n\rceil + 0 - 1 = \lceil\log_2 n\rceil - 1$$
+
+\vspace{0em}
+
+\begin{flushleft}
+	Somit können wir sagen, dass der Average-Case der binären Suche asymptotisch gesehen nur um 1 Schritt
+	besser ist, als der Wort-Case $\log_2n$
+\end{flushleft}
 

### 2021-04-25 algorithmen-und-datenstrukturen/assignment-1.tex
fix spelling issues in the aud assignment

@@
 Der Algorithmus braucht im \textbf{Best-Case n} Vergleiche, liegt also dementsprechend in $\Omega(n)$. Der Best-Case tritt ein,
-wenn direkt das zweite Element größer als das erste ist, da dann die erste Schleife nach dem ersten Schritt abgebrochen wird und
-die 2. Schleife immer genau $n - 1$ vergleiche ausführt.
+wenn direkt das zweite Element größer als das Erste ist, da dann die erste Schleife nach dem ersten Schritt abgebrochen wird und
+die 2. Schleife immer genau $n - 1$ Vergleiche ausführt.
 
-Er braucht im \textbf{Worst-Case 2(n - 1)} Vergleiche und liegt daher in $O(n)$. Der Worst-Case kommt zustande wenn wir z.B. eine
-List der Länge n betrachten, die n mal das gleiche Element enthält. Dann benötigen wir beim durchlaufen der ersten und der zweiten
+Er braucht im \textbf{Worst-Case 2(n - 1)} Vergleiche und liegt daher in $O(n)$. Der Worst-Case kommt zustande, wenn wir z.B. eine
+List der Länge n betrachten, die n mal das gleiche Element enthält. Dann benötigen wir bei dem Durchlaufen der ersten und der zweiten
 Schleife $n - 1$ Vergleiche.
@@
 \begin{flushleft}
-	Somit können wir sagen, dass mit $c \geq 5$ und $n_0 = 1$ die Behauptung $f(n) \in O(n^2)$ gilt
+	Somit können wir sagen, dass mit $c \geq 5$ und $n_0 = 1$ die Behauptung $f(n) \in O(n^2)$ gilt.
 \end{flushleft}
@@
 \begin{flushleft}
-	Also, mit $c = 1$ und $n_0 = \lceil\sqrt{72}\rceil = 9$ gilt $f(n) \in \Omega(n^2)$
+	Also mit $c = 1$ und $n_0 = \lceil\sqrt{72}\rceil = 9$ gilt $f(n) \in \Omega(n^2)$
 \end{flushleft}
@@
 
-	\hspace{0.5em}
+	\vspace{0.5em}
 
@@
 
-	\vspace{1em}
+	\vspace{0.5em}
 
@@
 
-	\vspace{2em}
+	\vspace{1.5em}
 
@@
 Auf (einer 0 indizierten) Ebene $i$ haben wir den Baum $i$ Mal geteilt und haben $(i + 1) 2^i$ Vergleiche
-auf dieser Ebene. Wenn wir nun alle Ebenen addieren möchten, um die gesamt Anzahl der Vergleiche zu bekommen,
+auf dieser Ebene. Wenn wir nun alle Ebenen addieren möchten, um die Gesamtanzahl der Vergleiche zu bekommen,
 müssen wir lediglich alle Ebenen addieren. Also bei $k = 4$ wären wir bei $1 * 2^0 + 2 * 2^1 + 3 * 2^2 + 4 * 2^3$
-Vergleichen. Dies lässt sich durch die gaußsche Summenformel eleganter (und allgemeingültiger) Zusammenfassen
-zu $\sum_{i=0}^{k - 1} (i + 1) * 2^i$. Um jetzt auf die durchschnittlichen Vergleiche zu kommen muss nun einfach
+Vergleichen. Dies lässt sich durch die Gaußsche Summenformel eleganter (und allgemeingültiger) Zusammenfassen
+zu $\sum_{i=0}^{k - 1} (i + 1) * 2^i$. Um jetzt auf die durchschnittlichen Vergleiche zu kommen, muss nun einfach
 die Gesamtanzahl durch die Anzahl der Elemente geteilt werden. Also entweder $\sum_{i=0}^{k - 1} \frac{(i + 1) * 2^i}{2^k - 1}$
@@
 
-\subsection{Laufzeit-Analyse für Average-Case und vergleich mit Worst-Case}
+\subsection{Laufzeit-Analyse für Average-Case und Vergleich mit Worst-Case}
 
@@
 	Somit können wir sagen, dass der Average-Case der binären Suche asymptotisch gesehen nur um 1 Schritt
-	besser ist, als der Wort-Case $\log_2n$
+	besser ist als der Wort-Case $\log_2n$
 \end{flushleft}
@@
 
-Die Funktion terminiert nur für gerade postive Zahlen und $n = 0$. Bei ungeraden positiven Zahlen, verfehlen wir den
-Basis-Fall immer um genau 1 und landen danach in einer Endlosschleife. Bei negativen Zahlen, sind wir schon initial
+Die Funktion terminiert nur für gerade postive Zahlen und $n = 0$. Bei ungeraden positiven Zahlen verfehlen wir den
+Basis-Fall immer um genau 1 und landen danach in einer Endlosschleife. Bei negativen Zahlen sind wir schon initial
 ``unter`` dem Basis-Fall.

### 2021-04-25 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
add a product backlog screenshot to the swt assignment

@@
 
-\vspace{100px}
-\begin{center}
-	\large{To be inserted!}
-\end{center}
-\vspace{100px}
+Folgende Aufgabenstellung wurde im Moodle-Kurs bekannt gegeben:
+
+\begin{quote}
+	Wählen Sie sich ein kleines Entwicklungsprojekt für das gesamte Studienmodul Softwaretechnik und
+	verwenden Sie dieses Projekt für alle Lerneinheiten. Beispielsweise ein Tool zur Erfassung von
+	Requirements oder ein anderes Tool, das Ihnen im Bereich der Softwaretechnik helfen kann, oder
+	ein Tool, welches Sie schon immer mal bauen wollten. Im Verlauf des Moduls können Sie alle Themen
+	wie beispielsweise Modellierung oder Testen anhand dieses Projekts praktisch erproben.
+
+	Ermitteln und dokumentieren Sie die Requirements für ihr Projekt!
+
+	\begin{itemize}
+		\item Formulieren Sie die Anforderungen als User Stories in der Sprache des Benutzers!
+		\item Formulieren Sie Abnahmeszenarien!
+		\item Reichern Sie weitere Informationen an, um die Qualitätskriterien für gute Anforderungen zu erfüllen!
+	\end{itemize}
+
+	Stellen Sie weiterhin dar, wie, d.h. mit welchen Werkzeugen bzw. in welcher Struktur, Sie diese Requirements verwalten!
+
+	Die Abgabe erfolgt in Form genau einer PDF-Datei.
+Falls sie Werkzeuge einsetzen, die keinen PDF-Export erlauben, fertigen Sie Screenshots an, die Sie mit der übrigen Beschreibung in eine PDF-Datei einfügen.
+
+	Anmerkungen:
+	\begin{itemize}
+		\item Der zeitliche Umfang dieser Einsendeaufgabe wird mit durchschnittlich 4 Stunden abgeschätzt
+		\item Eine Nachbearbeitung der abgegeben Lösung nach der Deadline ist nicht vorgesehen. Falls Sie Fragen zur Aufgabenstellung haben, fragen Sie diese also bitte im Vorfeld!
+	\end{itemize}
+\end{quote}
 
@@
 
-\newpage
-
 \section{Anforderungen}
 
+\begin{figure}[h]
+	\includegraphics[width=\textwidth]{product-backlog}
+	\centering
+	\caption{\emph{Product Backlog}}
+\end{figure}
+
 \end{document}

### 2021-04-26 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
add the product backlog and a detailed view of some requirements to the swt assignment 1

@@
 \usepackage[T1]{fontenc}
+\usepackage{tabularx}
 \usepackage{hyperref}
 \usepackage{graphicx}
+\usepackage{float}
 
@@
 		\large{Sommersemester 2021}\\[0.25em]
-		\large{Abgabetermin 25.04.2021}
+		\large{Abgabetermin 27.04.2021}
 	\end{center}
@@
 
-\begin{figure}[h]
-	\includegraphics[width=0.6\textwidth]{github-projects}
+\begin{figure}[ht]
+	\includegraphics[width=1\textwidth]{github-projects}
 	\centering
@@
 
+\vspace{0em}
+\newpage
+
 \section{Anforderungen}
 
-\begin{figure}[h]
-	\includegraphics[width=\textwidth]{product-backlog}
+Um die Lesbarkeit zu gewährleisten, habe ich hier eine Kopie des Backlogs aus GitHub Projects eingefügt.
+Einen Screenshot des Interfaces finden Sie weiter unten (Abbildung \ref{fig:backlog} auf Seite \pageref{fig:backlog}).
+
+\vspace{1em}
+
+\begin{tabularx}{\textwidth}{|c|X|}
+	\hline
+	\textbf{ID} & \textbf{Story}                                                                                                                                                          \\
+	\hline
+	\#10        & Als Benutzer möchte ich typische Operationen für Fenster ausführen können (z.B. Beenden oder Verschieben), damit ich Fenster einfach über mein Programm verwalten kann. \\
+	\hline
+	\#18        & Als Benutzer möchte ich Keyboard-Shortcuts definieren, um meine Maus so selten wie möglich zu verwenden.                                                                \\
+	\hline
+	\#7         & Als Benutzer möchte ich Layouts definieren, damit meine Fenster bestmöglich angeordnet werden.                                                                          \\
+	\hline
+	\#20        & Als Benutzer möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können.                                \\
+	\hline
+	\#16        & Als Benutzer möchte ich ein Fenster in Fullscreen anzeigen können, damit ich ablenkungsfrei arbeiten kann.                                                              \\
+	\hline
+	\#19        & Als Benutzer möchte ich Fenster zwischen verschiedenen Workspaces verschieben, um die Kontrolle über die angezeigten Fenster je Workspace zu haben.                     \\
+	\hline
+	\#17        & Als Benutzer möchte ich Border für Fenster fetlegen können, um zu sehen welches Fenster aktiv ist.                                                                      \\
+	\hline
+	\#9         & Als Benutzer möchte ich eine Statusleiste einbinden können, um den Status, wie z.B. den momentanen Workspace, angezeigt zu kriegen.                                     \\
+	\hline
+	\#6         & Als Benutzer möchte ich Regeln definieren, um Fenster einer Anwendung in einen Workspace einzusortieren.                                                                \\
+\end{tabularx}
+
+\begin{tabularx}{\textwidth}{|c|X|}
+	\#8  & Als Benutzer möchte austauschbare Layouts, um eigene Layouts implementieren und verwenden zu können.                                                        \\
+	\hline
+	\#12 & Als Benutzer möchte ich eine gute Dokumentation der Schnittstellen-Funktionen, damit ich ohne weitreichendes Fachwissen eine Konfiguration erstellen kann.  \\
+	\hline
+	\#11 & Als Benutzer möchte ich auf Metadaten von Fenstern zugreifen können, um in Abhängigkeit der Metadaten (z.B. Initiale Fenstergröße) Operationen auszuführen. \\
+	\hline
+	\#4  & Als Benutzer möchte ich eine Liste der angeschlossenen Bildschirme erhalten, um Regeln pro Bildschirm festzulegen.                                          \\
+	\hline
+	\#13 & Als Benutzer möchte ich einfach Zugriff auf Community Erweiterungen haben, um so wenig wie möglich zu programmieren.                                        \\
+	\hline
+	\#5  & Als Benutzer möchte ich beliebig viele Workspaces haben, um bestimmte Fenster in bestimmten Layouts anzuzeigen.                                             \\
+	\hline
+	\#15 & Als Benutzer möchte ich einfach mit Maus-Events arbeiten, um die Maus als Alternative zur Tastatur verwenden.                                               \\
+	\hline
+	\#14 & Als Benutzer möchte ich Hooks haben, damit ich die Funktionsweise grundlegend beeinflussen kann ohne die Schnittstelle zu Patchen.                          \\
+	\hline
+\end{tabularx}
+
+\begin{figure}[ht]
+	\includegraphics[width=1\textwidth]{product-backlog}
 	\centering
 	\caption{\emph{Product Backlog}}
+	\label{fig:backlog}
+\end{figure}
+
+\vspace{1em}
+
+Die Priorisierung der Anforderungen erfolgt relativ über die Reihenfolge im oben gezeigten Product-Backlog.
+Innerhalb von GitHub Projects können die einzelnen Stories noch über Flags kategorisiert werden (z.B. als
+Funktional oder als Epic)
+
+\subsection{Detaillierte Betrachtung einzelner Anforderungen}
+
+\subsubsection{Details zur User Story \#10}
+
+\begin{tabularx}{\textwidth}{r X}
+	\textbf{ID}                 & \#10                                                                                                                                                                    \\
+	\\
+	\textbf{Story}              & Als Benutzer möchte ich typische Operationen für Fenster ausführen können (z.B.
+Beenden oder Verschieben), damit ich Fenster einfach über mein Programm verwalten kann.
+\\
+	\\
+	\textbf{Typ}                & Funktional                                                                                                                                                              \\
+	\\
+	\textbf{Akzeptanzkriterien} & Szenario 1 - Fenster schließen
+	\begin{quote}
+		Gegeben einer gültigen Fenster-Referenz (z.B. durch ID)
+
+		Wenn der Nutzer dieses Fenster schließen will
+
+		Dann schließt sich dieses Fenster
+	\end{quote}
+	Szenario 2 - Fenster verschieben
+	\begin{quote}
+		Gegeben einer gültigen Fenster-Referenz (z.B. durch ID)
+
+		Wenn der Nutzer dieses Fenster an eine Stelle (X, Y) bewegen will
+
+		Dann render dieses Fenster auf dem Bildschirm an die Koordinaten
+	\end{quote}
+\end{tabularx}
+
+\begin{tabularx}{\textwidth}{r X}
+	\hspace*{7.75em} & Szenario 3 - Ungültiges Fenster
+	\begin{quote}
+		Gegeben einer ungültigen Fenster-Referenz (z.B. da das Programm abgestürzt ist)
+
+		Wenn der Nutzer versucht eine Operation (Beenden / Verschieben etc.) auszuführen
+
+		Dann gebe dem Nutzer eine Fehlermeldung zurück
+	\end{quote}
+\end{tabularx}
+
+\subsubsection{Details zur User Story \#20}
+
+\begin{tabularx}{\textwidth}{r X}
+	\textbf{ID}                 & \#20                                                                                                                                     \\
+	\\
+	\textbf{Story}              & Als Benutzer möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können. \\
+	\\
+	\textbf{Typ}                & Funktional                                                                                                                               \\
+	\\
+	\textbf{Akzeptanzkriterien} & Szenario 1 - Workspace wechseln
+	\begin{quote}
+		Gegeben einer gültigen Workspace-ID
+
+		Wenn der Nutzer zu dem Workspace wechseln möchte
+
+		Dann verstecke alle Fenster im momentanen Workspace und zeige alle Fenster aus dem Workspace mit der gegebenen ID.
+	\end{quote}
+	Szenario 2 - Zum letzten Workspace
+	\begin{quote}
+		Gegeben eines vorherigen Workspace-Wechsels
+
+		Wenn der Nutzer zum letzten Workspace möchte
+
+		Dann verstecke alle Fenster im momentanen Workspace und zeige alle Fenster aus dem vorherigen Workspace.
+	\end{quote}
+	Szenario 3 - Ungültiger Workspace
+	\begin{quote}
+		Gegeben einer ungültigen Workspace-ID
+
+		Wenn der Nutzer zu dem Workspace wechseln möchte
+
+		Dann bleibe auf dem momentanen Workspace und gebe dem Nutzer eine Fehlermeldung zurück.
+	\end{quote}
+\end{tabularx}
+
+\newpage
+
+\subsection{Weitere Merkmale von Anforderungen}
+
+Durch das Issue-Feature von GitHub werden viele Merkmale guter Anforderungen automatisch mit aufgenommen. Zum Beispiel werden
+in der Detail-Ansicht eines Issues (die als Story markiert werden kann) immer der Autor, die ID, das Erstellungsdatum, die Historie,
+ein Kommentarverlauf / bzw. eine Diskussion zu diesem Issue, der Status (Offen / Duplikat / Closed etc.) und ggf. die Person(en) die gerade
+an diesem Issue arbeitet/en festgehalten. \par
+Außerdem kann man projektspezifische Templates für Issues anlegen, um z.B. Akzeptanzkriterien / Abnahmeszenarien immer in einem
+gewissen Umfang und einer ähnlichen Form zu erfassen.
+
+\begin{figure}[H]
+	\includegraphics[width=0.8\textwidth]{issue-detail}
+	\centering
+	\caption{Detail-Ansicht einer Story}
 \end{figure}

### 2021-04-27 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
fix a typo
rename userrole for stories in swt assignment

@@
 	\hline
-	\#10        & Als Benutzer möchte ich typische Operationen für Fenster ausführen können (z.B. Beenden oder Verschieben), damit ich Fenster einfach über mein Programm verwalten kann. \\
+	\#10        & Als Konfigurierender möchte ich typische Operationen für Fenster ausführen können (z.B. Beenden oder Verschieben), damit ich Fenster einfach über mein Programm verwalten kann. \\
 	\hline
-	\#18        & Als Benutzer möchte ich Keyboard-Shortcuts definieren, um meine Maus so selten wie möglich zu verwenden.                                                                \\
+	\#18        & Als Konfigurierender möchte ich Keyboard-Shortcuts definieren, um meine Maus so selten wie möglich zu verwenden.                                                                \\
 	\hline
-	\#7         & Als Benutzer möchte ich Layouts definieren, damit meine Fenster bestmöglich angeordnet werden.                                                                          \\
+	\#7         & Als Konfigurierender möchte ich Layouts definieren, damit meine Fenster bestmöglich angeordnet werden.                                                                          \\
 	\hline
-	\#20        & Als Benutzer möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können.                                \\
+	\#20        & Als Konfigurierender möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können.                                \\
 	\hline
-	\#16        & Als Benutzer möchte ich ein Fenster in Fullscreen anzeigen können, damit ich ablenkungsfrei arbeiten kann.                                                              \\
+	\#16        & Als Konfigurierender möchte ich ein Fenster in Fullscreen anzeigen können, damit ich ablenkungsfrei arbeiten kann.                                                              \\
 	\hline
-	\#19        & Als Benutzer möchte ich Fenster zwischen verschiedenen Workspaces verschieben, um die Kontrolle über die angezeigten Fenster je Workspace zu haben.                     \\
+	\#19        & Als Konfigurierender möchte ich Fenster zwischen verschiedenen Workspaces verschieben, um die Kontrolle über die angezeigten Fenster je Workspace zu haben.                     \\
 	\hline
-	\#17        & Als Benutzer möchte ich Border für Fenster fetlegen können, um zu sehen welches Fenster aktiv ist.                                                                      \\
+	\#17        & Als Konfigurierender möchte ich Border für Fenster festlegen können, um zu sehen welches Fenster aktiv ist.                                                                     \\
 	\hline
-	\#9         & Als Benutzer möchte ich eine Statusleiste einbinden können, um den Status, wie z.B. den momentanen Workspace, angezeigt zu kriegen.                                     \\
+	\#9         & Als Konfigurierender möchte ich eine Statusleiste einbinden können, um den Status, wie z.B. den momentanen Workspace, angezeigt zu kriegen.                                     \\
 	\hline
-	\#6         & Als Benutzer möchte ich Regeln definieren, um Fenster einer Anwendung in einen Workspace einzusortieren.                                                                \\
+	\#6         & Als Konfigurierender möchte ich Regeln definieren, um Fenster einer Anwendung in einen Workspace einzusortieren.                                                                \\
 \end{tabularx}
@@
 \begin{tabularx}{\textwidth}{|c|X|}
-	\#8  & Als Benutzer möchte austauschbare Layouts, um eigene Layouts implementieren und verwenden zu können.                                                        \\
+	\#8  & Als Konfigurierender möchte austauschbare Layouts, um eigene Layouts implementieren und verwenden zu können.                                                        \\
 	\hline
-	\#12 & Als Benutzer möchte ich eine gute Dokumentation der Schnittstellen-Funktionen, damit ich ohne weitreichendes Fachwissen eine Konfiguration erstellen kann.  \\
+	\#12 & Als Konfigurierender möchte ich eine gute Dokumentation der Schnittstellen-Funktionen, damit ich ohne weitreichendes Fachwissen eine Konfiguration erstellen kann.  \\
 	\hline
-	\#11 & Als Benutzer möchte ich auf Metadaten von Fenstern zugreifen können, um in Abhängigkeit der Metadaten (z.B. Initiale Fenstergröße) Operationen auszuführen. \\
+	\#11 & Als Konfigurierender möchte ich auf Metadaten von Fenstern zugreifen können, um in Abhängigkeit der Metadaten (z.B. Initiale Fenstergröße) Operationen auszuführen. \\
 	\hline
-	\#4  & Als Benutzer möchte ich eine Liste der angeschlossenen Bildschirme erhalten, um Regeln pro Bildschirm festzulegen.                                          \\
+	\#4  & Als Konfigurierender möchte ich eine Liste der angeschlossenen Bildschirme erhalten, um Regeln pro Bildschirm festzulegen.                                          \\
 	\hline
-	\#13 & Als Benutzer möchte ich einfach Zugriff auf Community Erweiterungen haben, um so wenig wie möglich zu programmieren.                                        \\
+	\#13 & Als Konfigurierender möchte ich einfach Zugriff auf Community Erweiterungen haben, um so wenig wie möglich zu programmieren.                                        \\
 	\hline
-	\#5  & Als Benutzer möchte ich beliebig viele Workspaces haben, um bestimmte Fenster in bestimmten Layouts anzuzeigen.                                             \\
+	\#5  & Als Konfigurierender möchte ich beliebig viele Workspaces haben, um bestimmte Fenster in bestimmten Layouts anzuzeigen.                                             \\
 	\hline
-	\#15 & Als Benutzer möchte ich einfach mit Maus-Events arbeiten, um die Maus als Alternative zur Tastatur verwenden.                                               \\
+	\#15 & Als Konfigurierender möchte ich einfach mit Maus-Events arbeiten, um die Maus als Alternative zur Tastatur verwenden.                                               \\
 	\hline
-	\#14 & Als Benutzer möchte ich Hooks haben, damit ich die Funktionsweise grundlegend beeinflussen kann ohne die Schnittstelle zu Patchen.                          \\
+	\#14 & Als Konfigurierender möchte ich Hooks haben, damit ich die Funktionsweise grundlegend beeinflussen kann ohne die Schnittstelle zu Patchen.                          \\
 	\hline
@@
 	\\
-	\textbf{Story}              & Als Benutzer möchte ich typische Operationen für Fenster ausführen können (z.B.
+	\textbf{Story}              & Als Konfigurierender möchte ich typische Operationen für Fenster ausführen können (z.B.
 Beenden oder Verschieben), damit ich Fenster einfach über mein Programm verwalten kann.
@@
 	\\
-	\textbf{Story}              & Als Benutzer möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können. \\
+	\textbf{Story}              & Als Konfigurierender möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können. \\
 	\\

### 2021-04-27 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
add a note about the userroles

@@
 meistens schlecht dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}. Zu den typischen Anwendern dieser
-\emph{Window Manager} gehören Entwickler und Power User.
+\emph{Window Manager} gehören (meiner subjektiven Erfahrung nach) Software Entwickler, System Administratoren und
+Power User. Diese drei Kategorien fasse ich im folgenden zur Nutzerrolle ``Konfigurierender`` zusammen, da alle
+einen (für diesen Kontext) recht ähnlichen Kenntnisstand bzgl. der Domain haben.
 

### 2021-04-27 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
fix swt typos

@@
 Für den Fall, dass Sie noch keinen Kontakt mit dieser Domain hatten, gehe ich im folgenden kurz darauf ein,
-welche Aufgaben ein \emph{Tiling Window Manager} (kurz. \emph{TWM} oder nur \emph{WM}) typischer Weise
+welche Aufgaben ein \emph{Tiling Window Manager} (kurz. \emph{TWM} oder nur \emph{WM}) typischerweise
 übernimmt. Die generelle Aufgabe eines WM's besteht hauptsächlich darin sich um die Kommunikation mit dem
-\emph{Window-Server} (i.d.R. X11 oder Wayland auf Linux, Quartz Compositor auf MacOS) zu kümmern, und die
+\emph{Window-Server} (i.d.R. X11 oder Wayland auf Linux, Quartz Compositor auf MacOS) zu kümmern und die
 tatsächliche Anordnung der Fenster auf dem Bildschirm zu regeln (in Ebenen, Kacheln etc. – die Möglichkeiten
@@
 mittlerweile als üblich angesehene, Anordnungen wie Floating, Split-Screen und Fullscreen realisieren. \par
-Ein \emph{Tiling Window Managers} hat nun die Besonderheit, dass er anders als von Windows, MacOS oder diversen
+Ein \emph{Tiling Window Manager} hat nun die Besonderheit, dass er anders als von Windows, MacOS oder diversen
 Linux Desktop-Umgebungen bekannt, die Aufteilung der Fenster automatisch und bestmöglich regelt. Typischerweise
@@
 and Stack``). Somit muss sich der Anwender eines solchen \emph{TWM's} (zumindest initial) nicht selber um die
-Anordnung seiner Fenster kümmern. Desweiteren zählt die Möglichkeit, Keybindings zur Navigation oder Veränderung
+Anordnung seiner Fenster kümmern. Des Weiteren zählt die Möglichkeit, Keybindings zur Navigation oder Veränderung
 der Aufteilung zu definieren, zu den typischen Features eines \emph{Tiling Window Managers}.
@@
 C, C++ oder einer vergleichbaren Sprache verfasst worden und mit der Zeit immer weiter gewachsen. Dies macht es
-deutlich schwerer als es sein müsste, grundlegende Änderungen, die nicht von den Entwicklern vorhergesehen waren,
+deutlich schwerer, als es sein müsste, grundlegende Änderungen, die nicht von den Entwicklern vorhergesehen waren,
 vorzunehmen. \par
@@
 dienen die Projekte \emph{XMonad\footnotemark} und \emph{DWM\footnotemark}), allerdings sind diese Projekte leider
-meistens schlecht dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}. Zu den typischen Anwendern dieser
+meistens schlecht dokumentiert und haben i. d. R. unnötig komplexe \emph{API's}. Zu den typischen Anwendern dieser
 \emph{Window Manager} gehören (meiner subjektiven Erfahrung nach) Software Entwickler, System Administratoren und
-Power User. Diese drei Kategorien fasse ich im folgenden zur Nutzerrolle ``Konfigurierender`` zusammen, da alle
+Power User. Diese drei Kategorien fasse ich im Folgenden zur Nutzerrolle ``Konfigurierender`` zusammen, da alle
 einen (für diesen Kontext) recht ähnlichen Kenntnisstand bzgl. der Domain haben.

### 2021-05-14 algorithmen-und-datenstrukturen/assignment-2.tex
add inital version of the second aud assignment

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{color}

\lstset{
  language=Python,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  breaklines=true,
  breakatwhitespace=true,
  tabsize=4
}

\renewcommand\thesubsection{\alph{subsection})}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik und Medien \\
		Algorithmen und Datenstrukturen \\
		Prof. Dr. rer. nat. Ulrich Baum
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 2}\\[0.5em]
		\large{Sommersemester 2021}\\[0.25em]
		\large{Abgabetermin XX.05.2021}
	\end{center}

	\vfill

	\begin{flushright}
		Maximilian Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\newpage

\section{Finde die falsche Münze}

\subsection{Man kann das Problem lösen, indem man die erste Münze nacheinander mit
allen anderen vergleicht. Geben Sie den Worst-Case und Best-Case Aufwand dafür an.}

Falls der Best-Case entritt (die falsche Münze liegt an einer der ersten beiden
Positionen) haben wir nur einen einzigen Vergleich dementsprechen $O(1)$. Der Worst-
Case tritt dementsprechen im umgekehrten Fall ein, also wenn die falsche Münze ganz
am Ende der Folge liegt - dann benötigen wir $n - 1$ Vergleiche und liegen somit in
$O(n)$.

\subsection{Entwerfen und beschreiben Sie umgangssprachlich einen rekursiven Divide-And-Conquer
Algorithmus, der das Problem für große $n$ deutlich schneller löst.}

Am besten geht man hier mit einer modifizierten Version der binären Suche vor.
Da $n$ immer glatt durch 2 teilbar ist, haben wir immer eine identische Menge an Münzen
auf der linken und auf der rechten Seite, wenn wir die Ausgangsmenge in der exakten
Mitte teilen. Nun kann man diese beiden Mengen miteinander vergleichen. Falls die Anzahl
der Münzen in jeder Hälfte größer als 1 ist, muss dieser Vorgang wiederholt werden, bis
nurnoch 2 Münzen miteinander verglichen werden, dann ist die leichtere der beiden die
falsche Münze.

\subsection{Analysieren Sie den Worst-Case und Best-Case Aufwand dafür. Zählen Sie
die Anzahl der Wiegevorgänge dabei genau (nicht asymptotisch)}

Nun unterscheidet sich der o.g. Algorithmus von der normalen binären Suche dadurch, dass
wir keinen Vergleich mit einem einzelnen Element haben, mit dem wir vorzeitig abbrechen
könnten. Wir müssen also jedes mal, selbst wenn die falsche Münze in der Mitte liegt,
die Hälften immer ganz abarbeiten was bei 2k Elementen zu genau $log_2 n$ oder anders
ausgedrückt $k$ Vergleichen führt.

\section{Modifizierter Mergesort}

\subsection{Beschreiben Sie die Arbeitsweise von $merge4$ umgangssprachlich. Bestimmen
Sie die worst-case-Anzahl von Vergleichen von Vergleichen für $merge4$ auf 4 gleich
großen sortierten Folgen von je $m$ Elementen. Zählen Sie die Vergleiche genau (keine
O-Notation).}

Beim normalen $merge$ werden zwei in sich sortierten Listen zu einer großen sortierten
Liste zusammen gefügt, in dem immer das kleinere der beiden Elemente der beiden Teillisten
zu erst in die neue Liste geschrieben wird, dieser Vorgang kostet einen Vergleich.

Nun müssen bei $merge4$ nicht 2 sondern 4 Listen verglichen werden, es muss also bei jedem
Vorgang das minimum von 4 Elementen identifiziert werden, das kostet 3 Vergleiche. Sobald
die Anzahl aller Elemente der Teillisten 4 ist brauchen wir nur $3!$ Vergleiche um die restlichen
Elemente einzusortieren.

Also anders ausgedrückt braucht $merge4$ im Worst-Case $((m - 1) * 4 * 3) + 3!$ Vergleiche.
Der Worst-Case tritt ein, wenn die Elemente gleichmäßig auf die Teillisten verteilt ist, also
bei $m = 2$ die erste Liste $[1, 5]$, die zweite Liste $[2, 6]$, die dritte Liste $[3, 7]$
und die vierte Liste $[4, 8]$ ist. Diese Aufteilung verhindert das vorzeitige Abarbeiten
einer einzelenen Liste wodurch Vergleiche eingespart werden könnten.

\subsection{T(n) sei die worst-case Anzahl von Vergleichen des Sortierverfahrens auf
einer Inputfolge der Länge $n$. Nehmen Sie an, dass $n$ ein Vielfaches von 4 ist.
Stellen Sie die Rekursionsgleichung für T(n) auf.}

\begin{align*}
	         T(4) & = 0                                        \\
	         T(n) & = \, 4T(\frac{n}{4}) + merge4(\frac{n}{4}) \\
	\\
	    merge4(m) & = ((m - 1) * 4 * 3) + 3!
\end{align*}

\subsection{Lösen Sie die Rekursionsgleichung asymptotisch mit dem Master-Theorem}

\subsection{Wird diese Variante in der Praxis schneller sein als der Standard-Mergesort?}

Nein. Der Aufwand der einzelenen Merge-Vorgänge ist deutlich höher, viel mehr Elemente
gleichzeitig betrachtet werden müssen.

\section{Umdrehen von Listen}

\begin{lstlisting}
def reverse(head):
    first = copy(head)

    while first.next is not Null:
        old_head = copy(head)
        head = first.next
        first.next = head.next
        head.next = old_head

    return head
\end{lstlisting}

%3. Umdrehen von Listen (5 Punkte)Beschreiben Sie in Pseudocode einen Algorithmus, der eine gegebene einfach verkettete Listedestruktiv umkehrt, d.h. alle Verzeigerungen zwischen den Listenelementen umdreht, ohne dieElemente zu kopieren.1
%Input ist ein Zeiger auf das erste Element der Liste, Output ist ein Zeiger auf das erste Elementder umgekehrten Liste (=letztes Element der ursprünglichen Liste).4. Datenstrom-Analyse (5 Punkte)Sie sollen die Umsätze eines Web-Shops kontinutierlich überwachen und erhalten zu jedemVerkauf den Umsatz als Zahl übermittelt. Ihr Input ist also ein beliebig langer fortlaufenderStrom von Zahlen. Weiterhin wird vom Anwender beim Start des Algorithmus eine natürlicheZahlk >1gewählt.kbleibt während der Laufzeit unverändert.Das zu lösende Problem besteht aus zwei Teilproblemen:1.Jede neu eingehende Zahl einzeln geeignet verarbeiten, damit Sie die Anfragen in 2.beantworten können.2.Jederzeitauf Anfrage diekkleinsten allerbisher empfangenenUmsatzzahlen in irgendeinerReihenfolge auszugeben.(Sie können davon ausgehen, dass Anfragen erst gestellt werden, wenn schon mindestenskZahlen eingegangen sind.)Beispiel:k= 3, Zahlenfolge5 3 1 4 6 A 2 0 A(A steht für eine Anfrage zu diesem Zeitpunkt).Dann liefert die erste Anfrage3,1,4und die zweite Anfrage2,1,0.a.(3P) Entwerfen und beschreiben Sie umgangssprachlich einen Algorithmus, der die beidenTeilprobleme unter Verwendung einer Prioritätswarteschlange effizient löst.Die Laufzeit Ihres Algorithmus darf nur vonkabhängen, nicht aber von der Anzahlder empfangenen Umsatzzahlen! Brauchen Sie dafür eine Maximum- oder Minimum-Prioritätswarteschlange? Gehen Sie davon aus, dass die Proritätswarteschlange als Heap mitSpeicherung in einem Array implementiert ist und Ihnen die Zugriffsmethodeninsert()undremoveMax()bzw.removeMin()zur Verfügung stehen. Sie müssen also den Heapnicht selbst implementieren.b.(2P) Analysieren Sie den asymptotischen Aufwand Ihres Algorithmus für Teilproblem 1und 2 in Abhängigkeit vonk.

% 1(n-1)! * level!!

% 16(n-1)!
% 4 (n-1)!
% 1 (n-1)!

\end{document}

### 2021-05-15 softwaretechnik/assignments/uml/uml.tex
add the swt uml assignment document

--- 2021-04-27 softwaretechnik/assignments/requirements-engineering/requirements-engineering.tex
@@
 \usepackage{graphicx}
-\usepackage{float}
+\usepackage{plantuml}
 
@@
 
-Folgende Aufgabenstellung wurde im Moodle-Kurs bekannt gegeben:
-
-\begin{quote}
-	Wählen Sie sich ein kleines Entwicklungsprojekt für das gesamte Studienmodul Softwaretechnik und
-	verwenden Sie dieses Projekt für alle Lerneinheiten. Beispielsweise ein Tool zur Erfassung von
-	Requirements oder ein anderes Tool, das Ihnen im Bereich der Softwaretechnik helfen kann, oder
-	ein Tool, welches Sie schon immer mal bauen wollten. Im Verlauf des Moduls können Sie alle Themen
-	wie beispielsweise Modellierung oder Testen anhand dieses Projekts praktisch erproben.
-
-	Ermitteln und dokumentieren Sie die Requirements für ihr Projekt!
-
-	\begin{itemize}
-		\item Formulieren Sie die Anforderungen als User Stories in der Sprache des Benutzers!
-		\item Formulieren Sie Abnahmeszenarien!
-		\item Reichern Sie weitere Informationen an, um die Qualitätskriterien für gute Anforderungen zu erfüllen!
-	\end{itemize}
-
-	Stellen Sie weiterhin dar, wie, d.h. mit welchen Werkzeugen bzw. in welcher Struktur, Sie diese Requirements verwalten!
-
-	Die Abgabe erfolgt in Form genau einer PDF-Datei.
-Falls sie Werkzeuge einsetzen, die keinen PDF-Export erlauben, fertigen Sie Screenshots an, die Sie mit der übrigen Beschreibung in eine PDF-Datei einfügen.
-
-	Anmerkungen:
-	\begin{itemize}
-		\item Der zeitliche Umfang dieser Einsendeaufgabe wird mit durchschnittlich 4 Stunden abgeschätzt
-		\item Eine Nachbearbeitung der abgegeben Lösung nach der Deadline ist nicht vorgesehen. Falls Sie Fragen zur Aufgabenstellung haben, fragen Sie diese also bitte im Vorfeld!
-	\end{itemize}
-\end{quote}
-
-\section{Projekt}
-
-Als Projekt für dieses Semester verwende ich eine Idee, die mir schon seit längerem im Kopf rumgeistert
-– einen kleinen, minimalistischen \emph{Tiling Window Manager} für Linux bzw. X11, der deutlich flexibler
-und entwicklerfreundlicher ist als bestehende, vergleichbare Software.
-
-\subsection{Zusammenfassung der Domain}
-
-Für den Fall, dass Sie noch keinen Kontakt mit dieser Domain hatten, gehe ich im folgenden kurz darauf ein,
-welche Aufgaben ein \emph{Tiling Window Manager} (kurz. \emph{TWM} oder nur \emph{WM}) typischer Weise
-übernimmt. Die generelle Aufgabe eines WM's besteht hauptsächlich darin sich um die Kommunikation mit dem
-\emph{Window-Server} (i.d.R. X11 oder Wayland auf Linux, Quartz Compositor auf MacOS) zu kümmern, und die
-tatsächliche Anordnung der Fenster auf dem Bildschirm zu regeln (in Ebenen, Kacheln etc. – die Möglichkeiten
-sind nahe zu unbegrenzt). Typische \emph{Window Manager} von z.B. MacOS können diverse, für uns als Endnutzer
-mittlerweile als üblich angesehene, Anordnungen wie Floating, Split-Screen und Fullscreen realisieren. \par
-Ein \emph{Tiling Window Managers} hat nun die Besonderheit, dass er anders als von Windows, MacOS oder diversen
-Linux Desktop-Umgebungen bekannt, die Aufteilung der Fenster automatisch und bestmöglich regelt. Typischerweise
-hat ein \emph{TWM} eine feste Konfiguration mit Layouts, in die er die Fenster einsortieren kann (z.B. ``Master
-and Stack``). Somit muss sich der Anwender eines solchen \emph{TWM's} (zumindest initial) nicht selber um die
-Anordnung seiner Fenster kümmern. Desweiteren zählt die Möglichkeit, Keybindings zur Navigation oder Veränderung
-der Aufteilung zu definieren, zu den typischen Features eines \emph{Tiling Window Managers}.
-
-\subsection{Projekt-Ziel}
-
-Das Problem von bestehenden \emph{Window Managern} ist in der Regel deren Alter (bspw. ist das Projekt ``Toms
-Window Manager`` im Jahr 1987 entstanden). Aufgrund des Alters sind viele dieser \emph{Window Manager} noch in
-C, C++ oder einer vergleichbaren Sprache verfasst worden und mit der Zeit immer weiter gewachsen. Dies macht es
-deutlich schwerer als es sein müsste, grundlegende Änderungen, die nicht von den Entwicklern vorhergesehen waren,
-vorzunehmen. \par
-Ziel ist es einen \emph{minimalen Window Manager} zu entwickeln, der \emph{durch eigenen Programm-Code konfiguriert}
-und erweitert werden kann. Dieses Konzept ist schon öfter Implementiert worden (als Inspiration für dieses Modell
-dienen die Projekte \emph{XMonad\footnotemark} und \emph{DWM\footnotemark}), allerdings sind diese Projekte leider
-meistens schlecht dokumentiert und haben i.d.R. unnötig komplexe \emph{API's}. Zu den typischen Anwendern dieser
-\emph{Window Manager} gehören Entwickler und Power User.
-
-\footnotetext[1]{wurde 2007 erstmals veröffentlicht. Wurde in Haskell geschrieben und wird auch damit
-	konfiguriert. Siehe \url{https://xmonad.org/} für weitere Informationen.}
-\footnotetext[2]{ist ein äußerst kleiner (unter 2000 Zeilen Source-Code) Window Manager geschrieben in C. Er
-	hat keine Konfigurationsdatei und wird durch Patchen des Codes konfiguriert. Siehe \url{https://dwm.suckless.org/}
-	für weitere Informationen.}
-
-\section{Erfassung der Anforderungen}
-
-Zur Erfassung der Anforderungen verwende ich das Projektmanagement Feature von GitHub. Dieses hat zumindest im
-kleinen Rahmen diverse Vorteile gegenüber herkömmlicher Software für Agile Projekte (z.B. Jira von Atlassian).
-Unter anderem Folgende:
-
-\begin{itemize}
-	\item Übersichtlicher durch kleineren Umfang und somit leichter zu Bedienen
-	\item Direkte Integration mit dem Entwicklungsprozess (Issues können z.B.
-	      geschlossen werden sobald ein Feature gemerged wurde)
-	\item Nutzer der Software können direkt Issues anlegen
-\end{itemize}
-
-Diese Entscheidung hat keine Endgültigkeit. Wenn es ein echtes Team mit einem klaren Vorgehensmodell gibt, können
-andere Tools besser geeignet sein. Außerdem sei erwähnt, dass Jira mit diversen Integrationen erweitert werden
-kann, um die oben beschriebene Funktionalität nachzurüsten. Zur geordneten Erfassung von Anforderungen, deren
-Priorisierung, Kategorisierung und Abnahme, reicht \emph{GitHub Projects} allerdings vorerst vollkommen aus. Es
-gibt die Möglichkeit Sprint oder Kanban Boards anzulegen, diese zu automatisieren oder komplett eigene Prozesse
-zu integrieren.
-
-\vspace{0.5em}
-
-\begin{figure}[ht]
-	\includegraphics[width=1\textwidth]{github-projects}
-	\centering
-	\caption{\emph{GitHub Projects} Board}
-\end{figure}
-
-\vspace{0em}
-\newpage
-
-\section{Anforderungen}
-
-Um die Lesbarkeit zu gewährleisten, habe ich hier eine Kopie des Backlogs aus GitHub Projects eingefügt.
-Einen Screenshot des Interfaces finden Sie weiter unten (Abbildung \ref{fig:backlog} auf Seite \pageref{fig:backlog}).
-
-\vspace{1em}
-
-\begin{tabularx}{\textwidth}{|c|X|}
-	\hline
-	\textbf{ID} & \textbf{Story}                                                                                                                                                                  \\
-	\hline
-	\#10        & Als Konfigurierender möchte ich typische Operationen für Fenster ausführen können (z.B. Beenden oder Verschieben), damit ich Fenster einfach über mein Programm verwalten kann. \\
-	\hline
-	\#18        & Als Konfigurierender möchte ich Keyboard-Shortcuts definieren, um meine Maus so selten wie möglich zu verwenden.                                                                \\
-	\hline
-	\#7         & Als Konfigurierender möchte ich Layouts definieren, damit meine Fenster bestmöglich angeordnet werden.                                                                          \\
-	\hline
-	\#20        & Als Konfigurierender möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können.                                \\
-	\hline
-	\#16        & Als Konfigurierender möchte ich ein Fenster in Fullscreen anzeigen können, damit ich ablenkungsfrei arbeiten kann.                                                              \\
-	\hline
-	\#19        & Als Konfigurierender möchte ich Fenster zwischen verschiedenen Workspaces verschieben, um die Kontrolle über die angezeigten Fenster je Workspace zu haben.                     \\
-	\hline
-	\#17        & Als Konfigurierender möchte ich Border für Fenster festlegen können, um zu sehen welches Fenster aktiv ist.                                                                     \\
-	\hline
-	\#9         & Als Konfigurierender möchte ich eine Statusleiste einbinden können, um den Status, wie z.B. den momentanen Workspace, angezeigt zu kriegen.                                     \\
-	\hline
-	\#6         & Als Konfigurierender möchte ich Regeln definieren, um Fenster einer Anwendung in einen Workspace einzusortieren.                                                                \\
-\end{tabularx}
-
-\begin{tabularx}{\textwidth}{|c|X|}
-	\#8  & Als Konfigurierender möchte austauschbare Layouts, um eigene Layouts implementieren und verwenden zu können.                                                        \\
-	\hline
-	\#12 & Als Konfigurierender möchte ich eine gute Dokumentation der Schnittstellen-Funktionen, damit ich ohne weitreichendes Fachwissen eine Konfiguration erstellen kann.  \\
-	\hline
-	\#11 & Als Konfigurierender möchte ich auf Metadaten von Fenstern zugreifen können, um in Abhängigkeit der Metadaten (z.B. Initiale Fenstergröße) Operationen auszuführen. \\
-	\hline
-	\#4  & Als Konfigurierender möchte ich eine Liste der angeschlossenen Bildschirme erhalten, um Regeln pro Bildschirm festzulegen.                                          \\
-	\hline
-	\#13 & Als Konfigurierender möchte ich einfach Zugriff auf Community Erweiterungen haben, um so wenig wie möglich zu programmieren.                                        \\
-	\hline
-	\#5  & Als Konfigurierender möchte ich beliebig viele Workspaces haben, um bestimmte Fenster in bestimmten Layouts anzuzeigen.                                             \\
-	\hline
-	\#15 & Als Konfigurierender möchte ich einfach mit Maus-Events arbeiten, um die Maus als Alternative zur Tastatur verwenden.                                               \\
-	\hline
-	\#14 & Als Konfigurierender möchte ich Hooks haben, damit ich die Funktionsweise grundlegend beeinflussen kann ohne die Schnittstelle zu Patchen.                          \\
-	\hline
-\end{tabularx}
-
-\begin{figure}[ht]
-	\includegraphics[width=1\textwidth]{product-backlog}
-	\centering
-	\caption{\emph{Product Backlog}}
-	\label{fig:backlog}
-\end{figure}
-
-\vspace{1em}
-
-Die Priorisierung der Anforderungen erfolgt relativ über die Reihenfolge im oben gezeigten Product-Backlog.
-Innerhalb von GitHub Projects können die einzelnen Stories noch über Flags kategorisiert werden (z.B. als
-Funktional oder als Epic)
-
-\subsection{Detaillierte Betrachtung einzelner Anforderungen}
-
-\subsubsection{Details zur User Story \#10}
-
-\begin{tabularx}{\textwidth}{r X}
-	\textbf{ID}                 & \#10                                                                                                                                                                            \\
-	\\
-	\textbf{Story}              & Als Konfigurierender möchte ich typische Operationen für Fenster ausführen können (z.B.
-Beenden oder Verschieben), damit ich Fenster einfach über mein Programm verwalten kann.
-\\
-	\\
-	\textbf{Typ}                & Funktional                                                                                                                                                                      \\
-	\\
-	\textbf{Akzeptanzkriterien} & Szenario 1 - Fenster schließen
-	\begin{quote}
-		Gegeben einer gültigen Fenster-Referenz (z.B. durch ID)
-
-		Wenn der Nutzer dieses Fenster schließen will
-
-		Dann schließt sich dieses Fenster
-	\end{quote}
-	Szenario 2 - Fenster verschieben
-	\begin{quote}
-		Gegeben einer gültigen Fenster-Referenz (z.B. durch ID)
-
-		Wenn der Nutzer dieses Fenster an eine Stelle (X, Y) bewegen will
-
-		Dann render dieses Fenster auf dem Bildschirm an die Koordinaten
-	\end{quote}
-\end{tabularx}
-
-\begin{tabularx}{\textwidth}{r X}
-	\hspace*{7.75em} & Szenario 3 - Ungültiges Fenster
-	\begin{quote}
-		Gegeben einer ungültigen Fenster-Referenz (z.B. da das Programm abgestürzt ist)
-
-		Wenn der Nutzer versucht eine Operation (Beenden / Verschieben etc.) auszuführen
-
-		Dann gebe dem Nutzer eine Fehlermeldung zurück
-	\end{quote}
-\end{tabularx}
-
-\subsubsection{Details zur User Story \#20}
-
-\begin{tabularx}{\textwidth}{r X}
-	\textbf{ID}                 & \#20                                                                                                                                             \\
-	\\
-	\textbf{Story}              & Als Konfigurierender möchte ich zwischen verschiedenen Workspaces hin und her wechseln, um mehrere Workspaces auf einem Monitor haben zu können. \\
-	\\
-	\textbf{Typ}                & Funktional                                                                                                                                       \\
-	\\
-	\textbf{Akzeptanzkriterien} & Szenario 1 - Workspace wechseln
-	\begin{quote}
-		Gegeben einer gültigen Workspace-ID
-
-		Wenn der Nutzer zu dem Workspace wechseln möchte
-
-		Dann verstecke alle Fenster im momentanen Workspace und zeige alle Fenster aus dem Workspace mit der gegebenen ID.
-	\end{quote}
-	Szenario 2 - Zum letzten Workspace
-	\begin{quote}
-		Gegeben eines vorherigen Workspace-Wechsels
-
-		Wenn der Nutzer zum letzten Workspace möchte
-
-		Dann verstecke alle Fenster im momentanen Workspace und zeige alle Fenster aus dem vorherigen Workspace.
-	\end{quote}
-	Szenario 3 - Ungültiger Workspace
-	\begin{quote}
-		Gegeben einer ungültigen Workspace-ID
-
-		Wenn der Nutzer zu dem Workspace wechseln möchte
-
-		Dann bleibe auf dem momentanen Workspace und gebe dem Nutzer eine Fehlermeldung zurück.
-	\end{quote}
-\end{tabularx}
-
-\newpage
-
-\subsection{Weitere Merkmale von Anforderungen}
-
-Durch das Issue-Feature von GitHub werden viele Merkmale guter Anforderungen automatisch mit aufgenommen. Zum Beispiel werden
-in der Detail-Ansicht eines Issues (die als Story markiert werden kann) immer der Autor, die ID, das Erstellungsdatum, die Historie,
-ein Kommentarverlauf / bzw. eine Diskussion zu diesem Issue, der Status (Offen / Duplikat / Closed etc.) und ggf. die Person(en) die gerade
-an diesem Issue arbeitet/en festgehalten. \par
-Außerdem kann man projektspezifische Templates für Issues anlegen, um z.B. Akzeptanzkriterien / Abnahmeszenarien immer in einem
-gewissen Umfang und einer ähnlichen Form zu erfassen.
-
-\begin{figure}[H]
-	\includegraphics[width=0.8\textwidth]{issue-detail}
-	\centering
-	\caption{Detail-Ansicht einer Story}
-\end{figure}
+\begin{plantuml}
+  @startuml
+  Alice -> Bob: Hello
+  Alice <- Bob: Hi!
+  @enduml
+\end{plantuml}
 

### 2021-05-16 kommunikationsnetze-1/assignments/assignment-2/assignment-2.tex
add kn1 assignment 2

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\usepackage{graphicx}

\graphicspath{ {./images/} }

\renewcommand\thesubsection{\alph{subsection})}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik und Medien \\
		Kommunikationsnetze 1 \\
		Prof. Dr.-Ing. habil. Michael Syrjakow
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 2}\\[0.5em]
		\large{Sommersemester 2021}\\[0.25em]
		\large{Abgabetermin 21.05.2021}
	\end{center}

	\vfill

	\begin{flushright}
		Maximilian Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\newpage

\section*{Aufgabenstellung}

Eine Firma hat das Class C Netz 198.2.11.0 zugewiesen bekommen. Host A, Host B, Host C, R1 und R2 gehören
einer Firma, R ist ein Router im Internet. Die Firma benötigt drei Subnetze, wobei in den Subnetzen 2 und
3 jeweils nicht mehr als 10 Rechner angeschlossen werden sollen, alle anderen Rechner werden im Netz 1
angeschlossen.

\begin{figure}[ht]
	\includegraphics[width=0.75\textwidth]{plain}
	\centering
	\caption{Netzwerk}
\end{figure}

\section*{Lösung}

\subsection{Legen Sie die Adressen der Subnetze und deren Subnetzmasken bzw. Präfixe fest.}

\begin{tabular}{|c|c|c|}
	\hline
	Netz & Adresse        & Subnetzmaske    \\
	\hline
	1    & 198.2.11.0/24  & 255.255.255.0   \\
	\hline
	2    & 198.2.11.16/28 & 255.255.255.240 \\
	\hline
	3    & 198.2.11.32/28 & 255.255.255.240 \\
	\hline
\end{tabular}

\subsection{Legen Sie die Adressen in Netz 1 des Hosts A und der Router-Interfaces von R1 und R2 fest.}

\begin{tabular}{|c|c|}
	\hline
	Gerät  & Adresse       \\
	\hline
	R1     & 198.2.11.1/24 \\
	\hline
	R2     & 198.2.11.2/24 \\
	\hline
	Host A & 198.2.11.3/24 \\
	\hline
\end{tabular}

\subsection{Legen Sie die Adressen in Netz 2 des Hosts B und des Router-Interfaces von R2 fest.}

\begin{tabular}{|c|c|}
	\hline
	Gerät  & Adresse        \\
	\hline
	R2     & 198.2.11.17/28 \\
	\hline
	Host B & 198.2.11.18/28 \\
	\hline
\end{tabular}

\subsection{Legen Sie die Adressen in Netz 3 des Hosts C und des Router-Interfaces von R2 fest.}

\begin{tabular}{|c|c|}
	\hline
	Gerät  & Adresse        \\
	\hline
	R2     & 198.2.11.33/28 \\
	\hline
	Host C & 198.2.11.34/28 \\
	\hline
\end{tabular}

\subsection{Tragen Sie alle Adressen in die Zeichnung ein.}

\subsection{Legen Sie die Einträge in den Routing-Tabellen für Host A, Host B, Host C sowie R1 und R2 so
fest, dass von allen Rechnern aus das Internet sowie die Subnetze erreichbar sind.}

\vspace{1em}

\subsubsection*{R1}

\begin{tabular}{|c|c|}
	\hline
	default        & <Adresse von R> \\
	\hline
	198.2.11.16/28 & 198.2.11.2/24   \\
	\hline
	198.2.11.32/28 & 198.2.11.2/24   \\
	\hline
\end{tabular}

\subsubsection*{R2}

\begin{tabular}{|c|c|}
	\hline
	default        & 198.2.11.1/24 \\
	\hline
\end{tabular}

\subsubsection*{Host A}

\begin{tabular}{|c|c|}
	\hline
	default        & 198.2.11.1/24 \\
	\hline
	198.2.11.16/28 & 198.2.11.2/24 \\
	\hline
	198.2.11.32/28 & 198.2.11.2/24 \\
	\hline
\end{tabular}

\subsubsection*{Host B}

\begin{tabular}{|c|c|}
	\hline
	default & 198.2.11.17/28 \\
	\hline
\end{tabular}

\subsubsection*{Host C}

\begin{tabular}{|c|c|}
	\hline
	default & 198.2.11.33/28 \\
	\hline
\end{tabular}

\subsection{Welchen Eintrag braucht ein Router im Internet, damit das Netz und die Subnetze der Firma erreicht
werden können? Wie sieht die Subnetzmaske bzw. das Präfix aus?}

Jedes Paket was in das Firmennetz verschickt werden soll muss immer zu erst an R1 geroutet werden, da nur R1 die
interne Netzstruktur kennt. Daher muss der folgende Eintrag vorhanden sein:

\vspace{1em}

\begin{tabular}{|c|c|}
	\hline
	198.2.11.0/24 & 198.2.11.1/24 \\
	\hline
\end{tabular}

\end{document}

### 2021-05-21 algorithmen-und-datenstrukturen/assignment-2.tex
finish aud esa 2 exercise 2

@@
 
+\begin{align*}
+	         T(n) & = \, aT(\frac{n}{b}) + merge4(\frac{n}{b}) \\
+		merge4(m) & \in \Theta(m^d)                            \\
+				  & \Rightarrow a = 4                          \\
+				  & \Rightarrow b = 4                          \\
+				  & \Rightarrow c = 1                          \\
+				  & \Rightarrow d = \log_b a                   \\
+				  & \Rightarrow T(n) \in \Theta(n \log_2 n)
+\end{align*}
+
 \subsection{Wird diese Variante in der Praxis schneller sein als der Standard-Mergesort?}
@@
 Nein. Der Aufwand der einzelenen Merge-Vorgänge ist deutlich höher, viel mehr Elemente
-gleichzeitig betrachtet werden müssen.
+gleichzeitig betrachtet werden müssen. Grob betrachtet ist der Aufwand für $merge2$ für
+2 Elemente der Länge $m$ $= 2m - 1$ der Aufwand für $merge4$ war ja bereits oben als
+$((m - 1) * 4 * 3) + 3!$ definiert. Also bei 16 Elementen wäre der Aufwand von
+$merge2(8) = 2 * 8 - 1 = 15$ und der Aufwand von $merge4(4) = ((4 - 1) * 4 * 3) + 3! = 36 + 6 = 42$.
+Die größten Bestandteile der Laufzeit von $merge2$ bzw. $merge4$ sind bei $merge2 = 2m$
+und bei $merge4 = 12(m - 1)$ – somit wird $merge4$ in der Praxis eigentlich immer langsamer
+sein.
+
+\newpage
 
@@
 
-%3.
-Umdrehen von Listen (5 Punkte)Beschreiben Sie in Pseudocode einen Algorithmus, der eine gegebene einfach verkettete Listedestruktiv umkehrt, d.h.
-alle Verzeigerungen zwischen den Listenelementen umdreht, ohne dieElemente zu kopieren.1
-%Input ist ein Zeiger auf das erste Element der Liste, Output ist ein Zeiger auf das erste Elementder umgekehrten Liste (=letztes Element der ursprünglichen Liste).4.
-Datenstrom-Analyse (5 Punkte)Sie sollen die Umsätze eines Web-Shops kontinutierlich überwachen und erhalten zu jedemVerkauf den Umsatz als Zahl übermittelt.
-Ihr Input ist also ein beliebig langer fortlaufenderStrom von Zahlen.
-Weiterhin wird vom Anwender beim Start des Algorithmus eine natürlicheZahlk >1gewählt.kbleibt während der Laufzeit unverändert.Das zu lösende Problem besteht aus zwei Teilproblemen:1.Jede neu eingehende Zahl einzeln geeignet verarbeiten, damit Sie die Anfragen in 2.beantworten können.2.Jederzeitauf Anfrage diekkleinsten allerbisher empfangenenUmsatzzahlen in irgendeinerReihenfolge auszugeben.(Sie können davon ausgehen, dass Anfragen erst gestellt werden, wenn schon mindestenskZahlen eingegangen sind.)Beispiel:k= 3, Zahlenfolge5 3 1 4 6 A 2 0 A(A steht für eine Anfrage zu diesem Zeitpunkt).Dann liefert die erste Anfrage3,1,4und die zweite Anfrage2,1,0.a.(3P) Entwerfen und beschreiben Sie umgangssprachlich einen Algorithmus, der die beidenTeilprobleme unter Verwendung einer Prioritätswarteschlange effizient löst.Die Laufzeit Ihres Algorithmus darf nur vonkabhängen, nicht aber von der Anzahlder empfangenen Umsatzzahlen!
+\section{Datenstrom-Analyse}
+Sie sollen die Umsätze eines Web-Shops kontinutierlich überwachen und erhalten zu jedem
+Verkauf den Umsatz als Zahl übermittelt. Ihr Input ist also ein beliebig langer fortlaufender
+Strom von Zahlen. Weiterhin wird vom Anwender beim Start des Algorithmus eine natürliche Zahl
+$k > 1$ gewählt. $k$ bleibt während der Laufzeit unverändert. Das zu lösende Problem besteht
+aus zwei Teilproblemen:
+
+\begin{enumerate}
+	\item{Jede neu eingehende Zahl einzeln geeignet verarbeiten, damit Sie die Anfragen in 2.
+		beantworten können.}
+\end{enumerate}
+
+2.Jederzeitauf Anfrage diekkleinsten allerbisher empfangenenUmsatzzahlen in irgendeinerReihenfolge auszugeben.(Sie können davon ausgehen, dass Anfragen erst gestellt werden, wenn schon mindestenskZahlen eingegangen sind.)Beispiel:k= 3, Zahlenfolge5 3 1 4 6 A 2 0 A(A steht für eine Anfrage zu diesem Zeitpunkt).Dann liefert die erste Anfrage3,1,4und die zweite Anfrage2,1,0.a.(3P) Entwerfen und beschreiben Sie umgangssprachlich einen Algorithmus, der die beidenTeilprobleme unter Verwendung einer Prioritätswarteschlange effizient löst.Die Laufzeit Ihres Algorithmus darf nur vonkabhängen, nicht aber von der Anzahlder empfangenen Umsatzzahlen!
 Brauchen Sie dafür eine Maximum- oder Minimum-Prioritätswarteschlange?

### 2021-05-22 algorithmen-und-datenstrukturen/assignment-2.tex
add exercise 4 for aud assignment 2

@@
 		beantworten können.}
+	\item{Jederzeit auf Anfrage die $k$ kleinsten alle rbisher empfangenen Umsatzzahlen in
+		irgendeiner Reihenfolge auszugeben. (Sie können davon ausgehen, dass Anfragen erst
+		gestellt werden, wenn schon mindestens $k$ Zahlen eingegangen sind.)}
 \end{enumerate}
 
-2.Jederzeitauf Anfrage diekkleinsten allerbisher empfangenenUmsatzzahlen in irgendeinerReihenfolge auszugeben.(Sie können davon ausgehen, dass Anfragen erst gestellt werden, wenn schon mindestenskZahlen eingegangen sind.)Beispiel:k= 3, Zahlenfolge5 3 1 4 6 A 2 0 A(A steht für eine Anfrage zu diesem Zeitpunkt).Dann liefert die erste Anfrage3,1,4und die zweite Anfrage2,1,0.a.(3P) Entwerfen und beschreiben Sie umgangssprachlich einen Algorithmus, der die beidenTeilprobleme unter Verwendung einer Prioritätswarteschlange effizient löst.Die Laufzeit Ihres Algorithmus darf nur vonkabhängen, nicht aber von der Anzahlder empfangenen Umsatzzahlen!
-Brauchen Sie dafür eine Maximum- oder Minimum-Prioritätswarteschlange?
-Gehen Sie davon aus, dass die Proritätswarteschlange als Heap mitSpeicherung in einem Array implementiert ist und Ihnen die Zugriffsmethodeninsert()undremoveMax()bzw.removeMin()zur Verfügung stehen.
-Sie müssen also den Heapnicht selbst implementieren.b.(2P) Analysieren Sie den asymptotischen Aufwand Ihres Algorithmus für Teilproblem 1und 2 in Abhängigkeit vonk.
+\subsection{Entwerfen und beschreiben Sie umgangssprachlich einen Algorithmus, der die beiden
+Teilprobleme unter Verwendung einer Prioritätswarteschlange effizient löst. Die Laufzeit Ihres
+Algorithmus darf nur von $k$ abhängen, nicht aber von der Anzahl der empfangenen Umsatzzahlen!
+Brauchen Sie dafür eine Maximum- oder Minimum-Prioritätswarteschlange?}
 
-% 1(n-1)! * level!!
+Es muss lediglich eine einzige Maximum-Prioritätswarteschlange der Länge $k$ gepflegt werden.
+Sind noch weniger als $k$ Zahlen in der Prioritätswarteschlange, kann die Zahl in jedem Fall
+einfach über $insert()$ eingefügt werden.
 
-% 16(n-1)!
-% 4 (n-1)!
-% 1 (n-1)!
+Ansonsten muss die eintreffende Zahl mit der jeweils höchsten Zahl bzw. dem Root der Prioritätswarteschlange
+verglichen werden. Wenn die neue Zahl kleiner als die alte Root ist, qualifiziert sich die neue
+dafür in den $k$ kleinsten der bisher eingetroffenen Zahlen zu sein. Wenn die eintreffende Zahl
+qualifiziert ist, kann sie über $insert()$ eingefügt werden und anschließend kann die größte Zahl
+über $removeMax()$ entfernt werden.
+
+Sobald eine Anfrage eintrifft, die $k$ kleinsten Zahlen auszugeben, gibt es 2 Möglichkeiten.
+Entweder es wird eine Referenz zur Prioritätswarteschlange zurückgegeben, was zwar sehr effizient
+ist, aber unter Umständen ungewünschte Seiteneffekte nach sich ziehen könnte, falls jemand von
+''außen'' in die Prioritätswarteschlange schreibt. Alternativ, könnte eine ineffizientere, aber
+sichere Variante sein, eine Kopie der Inhalte der Prioritätswarteschlange zu erstellen und diese
+zurückzugegeben - so können ungewünschte Seiteneffekte vermieden werden, außerdem würden sich die
+Daten des Abfragenden nach der Abfrage nicht mehr ändern. Würde die Referenz zurückgegeben werden,
+müsste quasi nur ein einziges Mal eine Anfrage gestellt werden, da jedes Mal über die Referenz
+zugegriffen werden könnte.
+
+Im Zuge der Sicherheit würde ich zu der langsameren aber stabilieren Variante - der Kopie - raten.
+
+\subsection{Analysieren Sie den asymptotischen Aufwand Ihres Algorithmus für Teilproblem 1 und 2 in
+Abhängigkeit von $k$.}
+
+Der Vorgang einen Umsatz zu empfangen, liegt im BestCase in $\Omega(1)$ (falls die Zahl größer als das alte
+Maximum ist). Im WorstCase muss der neue Wert eingefügt und der alte Wert entfernt werden. Ein $insert()$
+in eine Prioritätswarteschlange kostet $log_2k$ und ein $removeMax()$ kostet $2log_2k$. Somit ist der 
+Gesamtaufwand im WorstCase ca. $3log_2k$ und liegt daher in $O(log_2k)$. Der AverageCase dürfte in der Praxis
+weit unter $log_2k$ liegen, da bei unendlich vielen Zahlen die Wahrscheinlichkeit zunehmend abnimmt, dass
+eine Zahl erfasst wird, die kleiner als die bisherigen Zahlen ist.
+
+Die Ausgabe der $k$ kleinsten Zahlen bei einer Anfrage zu tätigen, muss die $k$ Zahlen lange Prioritätswarteschlange
+kopiert werden - dies liegt in $\Theta(k)$.
+
+Um nun eine Zusammenfassende Auskunft über den tatsächlichen AverageCase zu treffen fehlen Eckdaten, wie
+der (bisherige) maximale Umsatz, häufigkeit der Abfragen und Wahrscheinlichkeit bestimmter Umsatzgrößen.
 

### 2021-05-22 algorithmen-und-datenstrukturen/assignment-2.tex
finish aud esa2

@@
 		\large{Sommersemester 2021}\\[0.25em]
-		\large{Abgabetermin XX.05.2021}
+		\large{Abgabetermin 23.05.2021}
 	\end{center}
@@
 Gesamtaufwand im WorstCase ca. $3log_2k$ und liegt daher in $O(log_2k)$. Der AverageCase dürfte in der Praxis
-weit unter $log_2k$ liegen, da bei unendlich vielen Zahlen die Wahrscheinlichkeit zunehmend abnimmt, dass
-eine Zahl erfasst wird, die kleiner als die bisherigen Zahlen ist.
+weit unter $log_2k$ liegen (natürlich total davon abhängig davon wie viele Zahlen aus welchem Zahlenbereich),
+da bei unendlich vielen Zahlen die Wahrscheinlichkeit zunehmend abnimmt, dass eine Zahl erfasst wird, die
+kleiner als die höchste der bisherigen Zahlen ist.
 

### 2021-05-22 softwaretechnik/presentations/architecture.tex
add inital version of the swt presentation

\documentclass{beamer}

\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage{graphicx}

\usecolortheme{seagull}
\usefonttheme{serif}

\title{Softwarearchitektur}
\subtitle{5 in 5 Zusammenfassung}
\author{Maximilian Schulke}

\begin{document}

\begin{frame}
	Softwaretechnik \hfill SS21
	\titlepage
	Prof. Dr. Martin Schafföner
\end{frame}

\begin{frame}{Übersicht}
	\tableofcontents
\end{frame}

\begin{frame}{Was ist Architektur?}

	- Ordentlich, sauberstrukturierte Komponenten
	- Patterns, kapselung, Abstraktion
	- Dienste Aufteilen => Kleinere Services die einzelne Aufgaben erledigen und bspw. über eine fest definierte REST / GraphQL API reden
\end{frame}

\begin{frame}{Wieso braucht man Architektur?}
	Software sollte nicht in Form eines Big Ball of Mud, also eines großen Schlammhaufens vorliegen.
	Die unstrukturierte Anhäufung von (oft funktionalen) Codeeinheiten wird unübersichtlich.

	Wird der Code komplex, kann er besser verstanden werden, wenn er strukturiert ist. Man findet sich besser zurecht.

	Code kann besser erweitert und gewartet werden wenn er strukturiert ist.

	Struktur im Softwaresystem spiegelt Domänen auf Aufgabenbereiche wieder. Diese können dann wiederum viel besser und unabhängig von den jeweiligen spezialisierten Personen bearbeitet werden.

	Korrektheit und Qualität lassen sich in einer guten Architektur besser sicherstellen.
\end{frame}

\begin{frame}{Ziele der Architektur}
	Wartbarkeit, erweiterbarkeit

	kOHÄSION / Kopplung

	Grundsätzlich ist eine schwache / lose Kopplung erstrebenswert. Es gibt mehrere Möglichkeiten dies zu erreichen, unter anderem:

	der Einsatz von Dependency Injection Frameworks

	der Einsatz von Design Patterns wie Factories
	die Kommunikation über ein Protokoll, z. B. durch Middleware / Messaging.
\end{frame}

\begin{frame}{Sample Page 2}{An Example of Lists}

	Architektur Ebenen? Vert vs Horiz
\end{frame}

\begin{frame}{Standardisierung}
	Die folgenden Punkte sollten innerhalb einer Architektur standardisiert sein:

	\begin{itemize}
		\item Codestyle
		\item Test und Buildframeworks
		\item CI/CD Pipeline
		\item uvm...
	\end{itemize}
\end{frame}

\begin{frame}{Architekturstile}{Grobe Stilübersicht}

	\begin{itemize}
		\item Kommunizierende Prozesse (z. B. Middleware oder Bussysteme)
		\item Call and Return (z. B. Prozedurale Programmierung)
		\item Datenorientierter Fluss (z. B. Batch Processing)
		\item Datenzentriert (z. B. Repositories oder Blackboards)
		\item Virtuelle Maschine (z. B. DSL oder Regelbasierte Sytsteme)
	\end{itemize}

	% Technologiearchitektur

	% Containerarchitektur? Komponenten Wrappen und fachfreie Logik (sicherheit, auth, log / monitoring, etc.) in container auslagern

	% Middleware architekturen (Im web beliebt)

	% P2P, Dezentralearchitektur
\end{frame}

\begin{frame}{Prinzipien der Architektur}

Die bisherigen Auflistungen von Werkzeugen und Stilmitteln, die dem Architekten zur Verfügung stehen, werden von verschiedenen Autoren wie beispielsweise von VOGEL ET AL. variiert, zusammengefasst und erweitert.

Beispielhaft sind hier 10 wichtige Prinzipien aufgeführt.

    Inkrementalität
    - durch Prototyping und schrittweises Wachstum
    Lose Kopplung
    - durch Gesetz von Demeter und Vermeidung von Zyklen
    Entwurf für Veränderung
    - Test was passiert, wenn Komponenten sich ändern oder wegfallen
    Hohe Kohäsion
    - gute fachliche Gruppierungen überleben auch Technologieshifts (z. B. wie vor langer Zeit von Corba nach JavaEE)
    Rückverfolgbarkeit
    - Änderungen, Fehler (Exceptions) und Datenflüsse transparent machen
    Abstraktion
    - Interfaces als explizite Schnittstellen, Subclassing, Polymorphie
    Information Hiding
    - Fassaden, Black Box Prinzip, Schichtenbildung
    Separation of Concerns
    - Analyse der Zuständigkeiten
    Modularität
    - Anwendung der Sprachmittel für Gruppierung
    Selbstdokumentation
    - Namensgebung, interne und externe Dokumentation
\end{frame}

\begin{frame}
		Middleware im Detail

		P2P (dezentrale Architektur) Im Detail
		    Sie organisieren sich selbst.

    Sie handeln mit ihren Ressourcen autonom.

    Dienste und Ressourcen können wandern.

    Peers können anbieten und nutzen.

    Verhalten, Zustand oder Durchsatz kann nicht vorhergesagt werden und erinnert eher an einen Schwarm.
\end{frame}

\begin{frame}{Frameworks}
	Es gibt frameworks, die ändern sich aber ständig.
\end{frame}

\begin{frame}
% https://vfhswt.eduloop.de/loop/Enterprise_Design_Patterns_ARC
\end{frame}

\begin{frame}{Architektur im Detail}{Domain-Specific-Language}
	DSL etc hier her!!
\end{frame}

\begin{frame}
	ADL statt UML manchmal
\end{frame}

% https://vfhswt.eduloop.de/loop/Nichtfunktionale_Anforderungen

% https://vfhswt.eduloop.de/loop/Was_machen_gute_Architekt/innen%3F

% https://vfhswt.eduloop.de/loop/Architektur_Ebenen_ARC
%	https://vfhswt.eduloop.de/loop/Packages_und_Namensgebung_ARC
\end{document}

### 2021-05-22 softwaretechnik/presentations/architecture.tex
add toc to the swt presentation

@@
 
+\AtBeginSection[]{
+	\begin{frame}{Übersicht}
+		\tableofcontents[currentsection]
+	\end{frame}
+}
+
 \begin{document}
@@
 
-\begin{frame}{Übersicht}
-	\tableofcontents
-\end{frame}
+\section{Was ist Architektur?}
 
@@
 
-\begin{frame}{Wieso braucht man Architektur?}
-	Software sollte nicht in Form eines Big Ball of Mud, also eines großen Schlammhaufens vorliegen.
-	Die unstrukturierte Anhäufung von (oft funktionalen) Codeeinheiten wird unübersichtlich.
+\section{Wieso braucht man Architektur?}
 
-	Wird der Code komplex, kann er besser verstanden werden, wenn er strukturiert ist. Man findet sich besser zurecht.
+\begin{frame}{Einleitung}{Wieso braucht man Architektur?}
 
-	Code kann besser erweitert und gewartet werden wenn er strukturiert ist.
+	\begin{itemize}
+		\item Vermeidung des "Big Ball of Mud".
+		\item Unstrukturierte Anhäufung von Code wird unübersichtlich.
+		\item Bei steigender Komplexität hilft Strukturierung.
+		\item Strukturierter Code ist leichter erweiterbar und wartbar.
+		\item Hilft die Software in Teilsysteme zu zerlegen die einzeln bearbeitet werden können.
+		\item Korrektheit und Qualität lässt sich leichter sicherstellen.
+	\end{itemize}
+\end{frame}
 
-	Struktur im Softwaresystem spiegelt Domänen auf Aufgabenbereiche wieder. Diese können dann wiederum viel besser und unabhängig von den jeweiligen spezialisierten Personen bearbeitet werden.
-
-	Korrektheit und Qualität lassen sich in einer guten Architektur besser sicherstellen.
-\end{frame}
+\section{Ziele der Architektur}
 
@@
 
-\begin{frame}{Sample Page 2}{An Example of Lists}
+\section{Architekturebenen}
+
+\begin{frame}{Architekturebenen}
 
 	Architektur Ebenen? Vert vs Horiz
+
+	Bild einfügen!!!
+
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{frame}
+
+\section{Standardisierung}
 
@@
 		\item CI/CD Pipeline
-		\item uvm...
+		\item ...
 	\end{itemize}
@@
 
+\section{Architekturstile}
+
 \begin{frame}{Architekturstile}{Grobe Stilübersicht}
-
 	\begin{itemize}
@@
 \end{frame}
+
+\section{Prinzipien der Architektur}
 
@@
 
+\section{Architekturen im Detail}
+
 \begin{frame}
@@
 
-\begin{frame}{Frameworks}
-	Es gibt frameworks, die ändern sich aber ständig.
-\end{frame}
-
-\begin{frame}
-% https://vfhswt.eduloop.de/loop/Enterprise_Design_Patterns_ARC
-\end{frame}
-
 \begin{frame}{Architektur im Detail}{Domain-Specific-Language}
@@
 	ADL statt UML manchmal
+\end{frame}
+
+\section{Frameworks}
+
+\begin{frame}{Frameworks}
+	Es gibt frameworks, die ändern sich aber ständig.
+\end{frame}
+
+\section{Enterprise Design Patterns}
+
+\begin{frame}
+% https://vfhswt.eduloop.de/loop/Enterprise_Design_Patterns_ARC
 \end{frame}

### 2021-05-23 algorithmen-und-datenstrukturen/assignments/assignment-2/assignment-2.tex
format aud assignment 2

--- 2021-05-22 algorithmen-und-datenstrukturen/assignment-2.tex
@@
 
-\subsection{Man kann das Problem lösen, indem man die erste Münze nacheinander mit
-allen anderen vergleicht. Geben Sie den Worst-Case und Best-Case Aufwand dafür an.}
+\subsection{Man kann das Problem lösen, indem man die erste Münze nacheinander
+mit allen anderen vergleicht. Geben Sie den Worst-Case und Best-Case Aufwand
+dafür an.}
 
 Falls der Best-Case entritt (die falsche Münze liegt an einer der ersten beiden
-Positionen) haben wir nur einen einzigen Vergleich dementsprechen $O(1)$. Der Worst-
-Case tritt dementsprechen im umgekehrten Fall ein, also wenn die falsche Münze ganz
-am Ende der Folge liegt - dann benötigen wir $n - 1$ Vergleiche und liegen somit in
-$O(n)$.
+Positionen) haben wir nur einen einzigen Vergleich dementsprechen $O(1)$. Der
+Worst- Case tritt dementsprechen im umgekehrten Fall ein, also wenn die falsche
+Münze ganz am Ende der Folge liegt - dann benötigen wir $n - 1$ Vergleiche und
+liegen somit in $O(n)$.
 
-\subsection{Entwerfen und beschreiben Sie umgangssprachlich einen rekursiven Divide-And-Conquer
-Algorithmus, der das Problem für große $n$ deutlich schneller löst.}
+\subsection{Entwerfen und beschreiben Sie umgangssprachlich einen rekursiven
+Divide-And-Conquer Algorithmus, der das Problem für große $n$ deutlich
+schneller löst.}
 
 Am besten geht man hier mit einer modifizierten Version der binären Suche vor.
-Da $n$ immer glatt durch 2 teilbar ist, haben wir immer eine identische Menge an Münzen
-auf der linken und auf der rechten Seite, wenn wir die Ausgangsmenge in der exakten
-Mitte teilen. Nun kann man diese beiden Mengen miteinander vergleichen. Falls die Anzahl
-der Münzen in jeder Hälfte größer als 1 ist, muss dieser Vorgang wiederholt werden, bis
-nurnoch 2 Münzen miteinander verglichen werden, dann ist die leichtere der beiden die
-falsche Münze.
+Da $n$ immer glatt durch 2 teilbar ist, haben wir immer eine identische Menge
+an Münzen auf der linken und auf der rechten Seite, wenn wir die Ausgangsmenge
+in der exakten Mitte teilen. Nun kann man diese beiden Mengen miteinander
+vergleichen. Falls die Anzahl der Münzen in jeder Hälfte größer als 1 ist, muss
+dieser Vorgang wiederholt werden, bis nurnoch 2 Münzen miteinander verglichen
+werden, dann ist die leichtere der beiden die falsche Münze.
 
-\subsection{Analysieren Sie den Worst-Case und Best-Case Aufwand dafür. Zählen Sie
-die Anzahl der Wiegevorgänge dabei genau (nicht asymptotisch)}
+\subsection{Analysieren Sie den Worst-Case und Best-Case Aufwand dafür. Zählen
+Sie die Anzahl der Wiegevorgänge dabei genau (nicht asymptotisch)}
 
-Nun unterscheidet sich der o.g. Algorithmus von der normalen binären Suche dadurch, dass
-wir keinen Vergleich mit einem einzelnen Element haben, mit dem wir vorzeitig abbrechen
-könnten. Wir müssen also jedes mal, selbst wenn die falsche Münze in der Mitte liegt,
-die Hälften immer ganz abarbeiten was bei 2k Elementen zu genau $log_2 n$ oder anders
-ausgedrückt $k$ Vergleichen führt.
+Nun unterscheidet sich der o.g. Algorithmus von der normalen binären Suche
+dadurch, dass wir keinen Vergleich mit einem einzelnen Element haben, mit dem
+wir vorzeitig abbrechen könnten. Wir müssen also jedes mal, selbst wenn die
+falsche Münze in der Mitte liegt, die Hälften immer ganz abarbeiten was bei 2k
+Elementen zu genau $log_2 n$ oder anders ausgedrückt $k$ Vergleichen führt.
 
@@
 
-\subsection{Beschreiben Sie die Arbeitsweise von $merge4$ umgangssprachlich. Bestimmen
-Sie die worst-case-Anzahl von Vergleichen von Vergleichen für $merge4$ auf 4 gleich
-großen sortierten Folgen von je $m$ Elementen. Zählen Sie die Vergleiche genau (keine
-O-Notation).}
+\subsection{Beschreiben Sie die Arbeitsweise von $merge4$ umgangssprachlich.
+Bestimmen Sie die worst-case-Anzahl von Vergleichen von Vergleichen für
+$merge4$ auf 4 gleich großen sortierten Folgen von je $m$ Elementen. Zählen Sie
+die Vergleiche genau (keine O-Notation).}
 
-Beim normalen $merge$ werden zwei in sich sortierten Listen zu einer großen sortierten
-Liste zusammen gefügt, in dem immer das kleinere der beiden Elemente der beiden Teillisten
-zu erst in die neue Liste geschrieben wird, dieser Vorgang kostet einen Vergleich.
+Beim normalen $merge$ werden zwei in sich sortierten Listen zu einer großen
+sortierten Liste zusammen gefügt, in dem immer das kleinere der beiden Elemente
+der beiden Teillisten zu erst in die neue Liste geschrieben wird, dieser
+Vorgang kostet einen Vergleich.
 
-Nun müssen bei $merge4$ nicht 2 sondern 4 Listen verglichen werden, es muss also bei jedem
-Vorgang das minimum von 4 Elementen identifiziert werden, das kostet 3 Vergleiche. Sobald
-die Anzahl aller Elemente der Teillisten 4 ist brauchen wir nur $3!$ Vergleiche um die restlichen
-Elemente einzusortieren.
+Nun müssen bei $merge4$ nicht 2 sondern 4 Listen verglichen werden, es muss
+also bei jedem Vorgang das minimum von 4 Elementen identifiziert werden, das
+kostet 3 Vergleiche. Sobald die Anzahl aller Elemente der Teillisten 4 ist
+brauchen wir nur $3!$ Vergleiche um die restlichen Elemente einzusortieren.
 
-Also anders ausgedrückt braucht $merge4$ im Worst-Case $((m - 1) * 4 * 3) + 3!$ Vergleiche.
-Der Worst-Case tritt ein, wenn die Elemente gleichmäßig auf die Teillisten verteilt ist, also
-bei $m = 2$ die erste Liste $[1, 5]$, die zweite Liste $[2, 6]$, die dritte Liste $[3, 7]$
-und die vierte Liste $[4, 8]$ ist. Diese Aufteilung verhindert das vorzeitige Abarbeiten
-einer einzelenen Liste wodurch Vergleiche eingespart werden könnten.
+Also anders ausgedrückt braucht $merge4$ im Worst-Case $((m - 1) * 4 * 3) + 3!$
+Vergleiche.  Der Worst-Case tritt ein, wenn die Elemente gleichmäßig auf die
+Teillisten verteilt ist, also bei $m = 2$ die erste Liste $[1, 5]$, die zweite
+Liste $[2, 6]$, die dritte Liste $[3, 7]$ und die vierte Liste $[4, 8]$ ist.
+Diese Aufteilung verhindert das vorzeitige Abarbeiten einer einzelenen Liste
+wodurch Vergleiche eingespart werden könnten.
 
-\subsection{T(n) sei die worst-case Anzahl von Vergleichen des Sortierverfahrens auf
-einer Inputfolge der Länge $n$. Nehmen Sie an, dass $n$ ein Vielfaches von 4 ist.
-Stellen Sie die Rekursionsgleichung für T(n) auf.}
+\subsection{T(n) sei die worst-case Anzahl von Vergleichen des
+Sortierverfahrens auf einer Inputfolge der Länge $n$. Nehmen Sie an, dass $n$
+ein Vielfaches von 4 ist.  Stellen Sie die Rekursionsgleichung für T(n) auf.}
 
@@
 
-\subsection{Lösen Sie die Rekursionsgleichung asymptotisch mit dem Master-Theorem}
+\subsection{Lösen Sie die Rekursionsgleichung asymptotisch mit dem
+Master-Theorem}
 
@@
 
-\subsection{Wird diese Variante in der Praxis schneller sein als der Standard-Mergesort?}
+\subsection{Wird diese Variante in der Praxis schneller sein als der
+Standard-Mergesort?}
 
-Nein. Der Aufwand der einzelenen Merge-Vorgänge ist deutlich höher, viel mehr Elemente
-gleichzeitig betrachtet werden müssen. Grob betrachtet ist der Aufwand für $merge2$ für
-2 Elemente der Länge $m$ $= 2m - 1$ der Aufwand für $merge4$ war ja bereits oben als
-$((m - 1) * 4 * 3) + 3!$ definiert. Also bei 16 Elementen wäre der Aufwand von
-$merge2(8) = 2 * 8 - 1 = 15$ und der Aufwand von $merge4(4) = ((4 - 1) * 4 * 3) + 3! = 36 + 6 = 42$.
-Die größten Bestandteile der Laufzeit von $merge2$ bzw. $merge4$ sind bei $merge2 = 2m$
-und bei $merge4 = 12(m - 1)$ – somit wird $merge4$ in der Praxis eigentlich immer langsamer
-sein.
+Nein. Der Aufwand der einzelenen Merge-Vorgänge ist deutlich höher, da viel mehr
+Elemente gleichzeitig betrachtet werden müssen. Grob betrachtet ist der Aufwand
+für $merge2$ für 2 Elemente der Länge $m$ $= 2m - 1$, der Aufwand für $merge4$
+war ja bereits oben als $((m - 1) * 4 * 3) + 3!$ definiert. Also bei 16
+Elementen wäre der Aufwand von $merge2(8) = 2 * 8 - 1 = 15$ und der Aufwand von
+$merge4(4) = ((4 - 1) * 4 * 3) + 3! = 36 + 6 = 42$.  Die größten Bestandteile
+der Laufzeit von $merge2$ bzw. $merge4$ sind bei $merge2 = 2m$ und bei $merge4
+= 12(m - 1)$ – somit wird $merge4$ in der Praxis immer langsamer sein.
 
@@
 \section{Datenstrom-Analyse}
-Sie sollen die Umsätze eines Web-Shops kontinutierlich überwachen und erhalten zu jedem
-Verkauf den Umsatz als Zahl übermittelt. Ihr Input ist also ein beliebig langer fortlaufender
-Strom von Zahlen. Weiterhin wird vom Anwender beim Start des Algorithmus eine natürliche Zahl
-$k > 1$ gewählt. $k$ bleibt während der Laufzeit unverändert. Das zu lösende Problem besteht
-aus zwei Teilproblemen:
+Sie sollen die Umsätze eines Web-Shops kontinutierlich überwachen und erhalten
+zu jedem Verkauf den Umsatz als Zahl übermittelt. Ihr Input ist also ein
+beliebig langer fortlaufender Strom von Zahlen. Weiterhin wird vom Anwender
+beim Start des Algorithmus eine natürliche Zahl $k > 1$ gewählt. $k$ bleibt
+während der Laufzeit unverändert. Das zu lösende Problem besteht aus zwei
+Teilproblemen:
 
 \begin{enumerate}
-	\item{Jede neu eingehende Zahl einzeln geeignet verarbeiten, damit Sie die Anfragen in 2.
-		beantworten können.}
-	\item{Jederzeit auf Anfrage die $k$ kleinsten alle rbisher empfangenen Umsatzzahlen in
-		irgendeiner Reihenfolge auszugeben. (Sie können davon ausgehen, dass Anfragen erst
-		gestellt werden, wenn schon mindestens $k$ Zahlen eingegangen sind.)}
+	\item{Jede neu eingehende Zahl einzeln geeignet verarbeiten,
+		damit Sie die Anfragen in 2. beantworten können.}
+	\item{Jederzeit auf Anfrage die $k$ kleinsten alle rbisher empfangenen
+		Umsatzzahlen irgendeiner Reihenfolge auszugeben. (Sie können davon
+		ausgehen, dass Anfragen erst gestellt werden, wenn schon mindestens $k$
+		Zahlen eingegangen sind.)}
 \end{enumerate}
 
-\subsection{Entwerfen und beschreiben Sie umgangssprachlich einen Algorithmus, der die beiden
-Teilprobleme unter Verwendung einer Prioritätswarteschlange effizient löst. Die Laufzeit Ihres
-Algorithmus darf nur von $k$ abhängen, nicht aber von der Anzahl der empfangenen Umsatzzahlen!
-Brauchen Sie dafür eine Maximum- oder Minimum-Prioritätswarteschlange?}
+\subsection{Entwerfen und beschreiben Sie umgangssprachlich einen Algorithmus,
+der die beiden Teilprobleme unter Verwendung einer Prioritätswarteschlange
+effizient löst. Die Laufzeit Ihres Algorithmus darf nur von $k$ abhängen,
+nicht aber von der Anzahl der empfangenen Umsatzzahlen!  Brauchen Sie dafür
+eine Maximum- oder Minimum-Prioritätswarteschlange?}
 
-Es muss lediglich eine einzige Maximum-Prioritätswarteschlange der Länge $k$ gepflegt werden.
-Sind noch weniger als $k$ Zahlen in der Prioritätswarteschlange, kann die Zahl in jedem Fall
-einfach über $insert()$ eingefügt werden.
+Es muss lediglich eine einzige Maximum-Prioritätswarteschlange der Länge $k$
+gepflegt werden.  Sind noch weniger als $k$ Zahlen in der
+Prioritätswarteschlange, kann die Zahl in jedem Fall einfach über $insert()$
+eingefügt werden.
 
-Ansonsten muss die eintreffende Zahl mit der jeweils höchsten Zahl bzw. dem Root der Prioritätswarteschlange
-verglichen werden. Wenn die neue Zahl kleiner als die alte Root ist, qualifiziert sich die neue
-dafür in den $k$ kleinsten der bisher eingetroffenen Zahlen zu sein. Wenn die eintreffende Zahl
-qualifiziert ist, kann sie über $insert()$ eingefügt werden und anschließend kann die größte Zahl
-über $removeMax()$ entfernt werden.
+Ansonsten muss die eintreffende Zahl mit der jeweils höchsten Zahl bzw. dem
+Root der Prioritätswarteschlange verglichen werden. Wenn die neue Zahl kleiner
+als die alte Root ist, qualifiziert sich die neue dafür in den $k$ kleinsten
+der bisher eingetroffenen Zahlen zu sein. Wenn die eintreffende Zahl
+qualifiziert ist, kann sie über $insert()$ eingefügt werden und anschließend
+kann die größte Zahl über $removeMax()$ entfernt werden.
 
-Sobald eine Anfrage eintrifft, die $k$ kleinsten Zahlen auszugeben, gibt es 2 Möglichkeiten.
-Entweder es wird eine Referenz zur Prioritätswarteschlange zurückgegeben, was zwar sehr effizient
-ist, aber unter Umständen ungewünschte Seiteneffekte nach sich ziehen könnte, falls jemand von
-''außen'' in die Prioritätswarteschlange schreibt. Alternativ, könnte eine ineffizientere, aber
-sichere Variante sein, eine Kopie der Inhalte der Prioritätswarteschlange zu erstellen und diese
-zurückzugegeben - so können ungewünschte Seiteneffekte vermieden werden, außerdem würden sich die
-Daten des Abfragenden nach der Abfrage nicht mehr ändern. Würde die Referenz zurückgegeben werden,
-müsste quasi nur ein einziges Mal eine Anfrage gestellt werden, da jedes Mal über die Referenz
-zugegriffen werden könnte.
+Sobald eine Anfrage eintrifft, die $k$ kleinsten Zahlen auszugeben, gibt es 2
+Möglichkeiten.  Entweder es wird eine Referenz zur Prioritätswarteschlange
+zurückgegeben, was zwar sehr effizient ist, aber unter Umständen ungewünschte
+Seiteneffekte nach sich ziehen könnte, falls jemand von ''außen'' in die
+Prioritätswarteschlange schreibt. Alternativ, könnte eine ineffizientere, aber
+sichere Variante sein, eine Kopie der Inhalte der Prioritätswarteschlange zu
+erstellen und diese zurückzugegeben - so können ungewünschte Seiteneffekte
+vermieden werden, außerdem würden sich die Daten des Abfragenden nach der
+Abfrage nicht mehr ändern. Würde die Referenz zurückgegeben werden, müsste
+quasi nur ein einziges Mal eine Anfrage gestellt werden, da jedes Mal über die
+Referenz zugegriffen werden könnte.
 
-Im Zuge der Sicherheit würde ich zu der langsameren aber stabilieren Variante - der Kopie - raten.
+Im Zuge der Sicherheit würde ich zu der langsameren aber stabilieren Variante -
+der Kopie - raten.
 
-\subsection{Analysieren Sie den asymptotischen Aufwand Ihres Algorithmus für Teilproblem 1 und 2 in
-Abhängigkeit von $k$.}
+\subsection{Analysieren Sie den asymptotischen Aufwand Ihres Algorithmus für
+Teilproblem 1 und 2 in Abhängigkeit von $k$.}
 
-Der Vorgang einen Umsatz zu empfangen, liegt im BestCase in $\Omega(1)$ (falls die Zahl größer als das alte
-Maximum ist). Im WorstCase muss der neue Wert eingefügt und der alte Wert entfernt werden. Ein $insert()$
-in eine Prioritätswarteschlange kostet $log_2k$ und ein $removeMax()$ kostet $2log_2k$. Somit ist der 
-Gesamtaufwand im WorstCase ca. $3log_2k$ und liegt daher in $O(log_2k)$. Der AverageCase dürfte in der Praxis
-weit unter $log_2k$ liegen (natürlich total davon abhängig davon wie viele Zahlen aus welchem Zahlenbereich),
-da bei unendlich vielen Zahlen die Wahrscheinlichkeit zunehmend abnimmt, dass eine Zahl erfasst wird, die
-kleiner als die höchste der bisherigen Zahlen ist.
+Der Vorgang einen Umsatz zu empfangen, liegt im BestCase in $\Omega(1)$ (falls
+die Zahl größer als das alte Maximum ist). Im WorstCase muss der neue Wert
+eingefügt und der alte Wert entfernt werden. Ein $insert()$ in eine
+Prioritätswarteschlange kostet $log_2k$ und ein $removeMax()$ kostet $2log_2k$.
+Somit ist der Gesamtaufwand im WorstCase ca. $3log_2k$ und liegt daher in
+$O(log_2k)$. Der AverageCase dürfte in der Praxis weit unter $log_2k$ liegen
+(natürlich total davon abhängig davon wie viele Zahlen aus welchem
+Zahlenbereich), da bei unendlich vielen Zahlen die Wahrscheinlichkeit zunehmend
+abnimmt, dass eine Zahl erfasst wird, die kleiner als die höchste der
+bisherigen Zahlen ist.
 
-Die Ausgabe der $k$ kleinsten Zahlen bei einer Anfrage zu tätigen, muss die $k$ Zahlen lange Prioritätswarteschlange
-kopiert werden - dies liegt in $\Theta(k)$.
+Die Ausgabe der $k$ kleinsten Zahlen bei einer Anfrage zu tätigen, muss die $k$
+Zahlen lange Prioritätswarteschlange kopiert werden - dies liegt in
+$\Theta(k)$.
 
-Um nun eine Zusammenfassende Auskunft über den tatsächlichen AverageCase zu treffen fehlen Eckdaten, wie
-der (bisherige) maximale Umsatz, häufigkeit der Abfragen und Wahrscheinlichkeit bestimmter Umsatzgrößen.
+Um nun eine Zusammenfassende Auskunft über den tatsächlichen AverageCase zu
+treffen fehlen Eckdaten, wie der (bisherige) maximale Umsatz, häufigkeit der
+Abfragen und Wahrscheinlichkeit bestimmter Umsatzgrößen.
 

### 2021-05-25 softwaretechnik/presentations/architecture.tex
enhance the swt architecture presentation

@@
 \begin{frame}{Was ist Architektur?}
-
-	- Ordentlich, sauberstrukturierte Komponenten
-	- Patterns, kapselung, Abstraktion
-	- Dienste Aufteilen => Kleinere Services die einzelne Aufgaben erledigen und bspw. über eine fest definierte REST / GraphQL API reden
+	Architektur ist die ...
+	\begin{itemize}
+		\item{strukturierte Aufteilung von Software in Teilprobleme / Domänen}
+		\item{Anwendung von Prinzipien wie Patterns, Kaspelung und
+			Abstraktion}
+		\item{
+			Dienste Aufteilen => Kleinere Services die einzelne Aufgaben
+			erledigen und bspw. über eine fest definierte REST / GraphQL API
+			reden}
+	\end{itemize}
 \end{frame}

### 2021-05-27 softwaretechnik/presentations/architecture.tex
Update architecture presentation slides

@@
 	\begin{itemize}
-		\item{strukturierte Aufteilung von Software in Teilprobleme / Domänen}
-		\item{Anwendung von Prinzipien wie Patterns, Kaspelung und
+		\item{strukturierte Aufteilung von Software in Teilprobleme}
+		\item{systematische Anwendung von Prinzipien wie Patterns, Kapselung und
 			Abstraktion}
-		\item{
-			Dienste Aufteilen => Kleinere Services die einzelne Aufgaben
-			erledigen und bspw. über eine fest definierte REST / GraphQL API
-			reden}
+		\item{Definition der Kommunikation und Verteilung der
+			Systemkomponenten}
 	\end{itemize}
@@
 
-\begin{frame}{Einleitung}{Wieso braucht man Architektur?}
-
+\begin{frame}{Wieso braucht man Architektur?}
 	\begin{itemize}
-		\item Vermeidung des "Big Ball of Mud".
-		\item Unstrukturierte Anhäufung von Code wird unübersichtlich.
-		\item Bei steigender Komplexität hilft Strukturierung.
-		\item Strukturierter Code ist leichter erweiterbar und wartbar.
-		\item Hilft die Software in Teilsysteme zu zerlegen die einzeln bearbeitet werden können.
-		\item Korrektheit und Qualität lässt sich leichter sicherstellen.
+		\item{Vermeidung des "Big Ball of Mud".}
+		\item{Unstrukturierte Anhäufung von Code wird unübersichtlich.}
+		\item{Bei steigender Komplexität hilft gute Strukturierung eine
+			Übersicht zu behalten.}
+		\item{Strukturierter Code ist leichter erweiterbar und wartbar.}
+		\item{Hilft die Software in Teilsysteme zu zerlegen die unabhängig
+			voneinander bearbeitet werden können.}
+		\item{Korrektheit und Qualität lässt sich leichter sicherstellen.}
 	\end{itemize}
@@
 \begin{frame}{Ziele der Architektur}
-	Wartbarkeit, erweiterbarkeit
+	\begin{itemize}
+		\item{Wartbarkeit, Erweiterbarkeit erhöhen}
+		\item{Kohäsion erhöhen / Kopplung minimieren}
+	\end{itemize}
+\end{frame}
 
-	kOHÄSION / Kopplung
+\begin{frame}{Ziele der Architektur}
+	Grundsätzlich ist eine schwache / lose Kopplung erstrebenswert.
+	Es gibt mehrere Möglichkeiten dies zu erreichen, unter anderem:
 
-	Grundsätzlich ist eine schwache / lose Kopplung erstrebenswert. Es gibt mehrere Möglichkeiten dies zu erreichen, unter anderem:
-
-	der Einsatz von Dependency Injection Frameworks
-
-	der Einsatz von Design Patterns wie Factories
-	die Kommunikation über ein Protokoll, z. B. durch Middleware / Messaging.
+	\begin{itemize}
+		\item{Einsatz von Dependency Injection Frameworks}
+		\item{Einsatz von Design Patterns (wie z. B. Factories)}
+		\item{Kommunkation über Protokolle (oder z. B. Middleware oder
+			Messaging)}
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{Architekturebenen}
-
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 	Architektur Ebenen? Vert vs Horiz
@@
 	Bild einfügen!!!
-
 	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 		\item Virtuelle Maschine (z. B. DSL oder Regelbasierte Sytsteme)
+		\item Dezentrale Netze (z. B. P2P)
 	\end{itemize}
@@
 	% Containerarchitektur? Komponenten Wrappen und fachfreie Logik (sicherheit, auth, log / monitoring, etc.) in container auslagern
-
-	% Middleware architekturen (Im web beliebt)
-
-	% P2P, Dezentralearchitektur
 \end{frame}
@@
 \begin{frame}{Prinzipien der Architektur}
-
-Die bisherigen Auflistungen von Werkzeugen und Stilmitteln, die dem Architekten zur Verfügung stehen, werden von verschiedenen Autoren wie beispielsweise von VOGEL ET AL.
+	% Die bisherigen Auflistungen von Werkzeugen und Stilmitteln, die dem Architekten zur Verfügung stehen, werden von verschiedenen Autoren wie beispielsweise von VOGEL ET AL.
 variiert, zusammengefasst und erweitert.
 
-Beispielhaft sind hier 10 wichtige Prinzipien aufgeführt.
+	% Beispielhaft sind hier 10 wichtige Prinzipien aufgeführt.
 
-    Inkrementalität
-    - durch Prototyping und schrittweises Wachstum
-    Lose Kopplung
-    - durch Gesetz von Demeter und Vermeidung von Zyklen
-    Entwurf für Veränderung
-    - Test was passiert, wenn Komponenten sich ändern oder wegfallen
-    Hohe Kohäsion
-    - gute fachliche Gruppierungen überleben auch Technologieshifts (z. B. wie vor langer Zeit von Corba nach JavaEE)
-    Rückverfolgbarkeit
-    - Änderungen, Fehler (Exceptions) und Datenflüsse transparent machen
-    Abstraktion
-    - Interfaces als explizite Schnittstellen, Subclassing, Polymorphie
-    Information Hiding
-    - Fassaden, Black Box Prinzip, Schichtenbildung
-    Separation of Concerns
-    - Analyse der Zuständigkeiten
-    Modularität
-    - Anwendung der Sprachmittel für Gruppierung
-    Selbstdokumentation
-    - Namensgebung, interne und externe Dokumentation
+	\begin{itemize}
+		\item{Inkrementalität} % durch Prototyping und schrittweises Wachstum.}
+		\item{Lose Kopplung} % durch Gesetz von Demeter und Vermeidung von Zyklen}
+		\item{Hohe Kohäsion} % gute fachliche Gruppierungen überleben auch Technologieshifts (z. B. wie vor langer Zeit von Corba nach JavaEE)}
+		\item{Entwurf für Veränderung} % Test was passiert, wenn Komponenten sich ändern oder wegfallen}
+		\item{Rückverfolgbarkeit} % Änderungen, Fehler (Exceptions) und Datenflüsse transparent machen}
+		\item{Abstraktion} % Interfaces als explizite Schnittstellen, Subclassing, Polymorphie}
+		\item{Information Hiding} % Fassaden, Black Box Prinzip, Schichtenbildung}
+		\item{Separation of Concerns} % Analyse der Zuständigkeiten}
+		\item{Modularität} % Anwendung der Sprachmittel für Gruppierung}
+		\item{Selbstdokumentation} % Namensgebung, interne und externe Dokumentation}
+	\end{itemize}
 \end{frame}
@@
 
-\section{Enterprise Design Patterns}
-
-\begin{frame}
 % https://vfhswt.eduloop.de/loop/Enterprise_Design_Patterns_ARC
-\end{frame}
-
 % https://vfhswt.eduloop.de/loop/Nichtfunktionale_Anforderungen
-
 % https://vfhswt.eduloop.de/loop/Was_machen_gute_Architekt/innen%3F
-
 % https://vfhswt.eduloop.de/loop/Architektur_Ebenen_ARC

### 2021-05-27 softwaretechnik/presentations/architecture.tex
finish swt architecture presentation

@@
 			Übersicht zu behalten.}
-		\item{Strukturierter Code ist leichter erweiterbar und wartbar.}
+		\item{Strukturierter Code ist leichter erweiterbar und wartbarer.}
 		\item{Hilft die Software in Teilsysteme zu zerlegen die unabhängig
@@
 	\begin{itemize}
+		\item{Identifikation von potenziellen Risiken}
 		\item{Wartbarkeit, Erweiterbarkeit erhöhen}
 		\item{Kohäsion erhöhen / Kopplung minimieren}
+		\item{Identifikation von Möglichkeiten zur Wiederverwendung von
+			Softwarekomponenten}
+		\item{...}
 	\end{itemize}
@@
 \begin{frame}{Architekturebenen}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	Architektur Ebenen? Vert vs Horiz
-
-	Bild einfügen!!!
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\begin{itemize}
+		\item{Ziel (Gesamtziel der Anwendung)}
+		\item{Firma (Firmenspezifische Ziele)}
+		\item{Anwendung / System (Systeme innerhalb einer Firma)}
+		\item{Softwareschicht (Architektur der Teilsoftware z. B. MVC ..)}
+		\item{Package / Komponentendiagramme (detailierte Ausarbeitung)}
+		\item{Klassen (Implementierungsschicht)}
+	\end{itemize}
 \end{frame}
@@
 
-\begin{frame}{Architekturstile}{Grobe Stilübersicht}
+\begin{frame}{Architekturstile}
 	\begin{itemize}
@@
 		\item Call and Return (z. B. Prozedurale Programmierung)
-		\item Datenorientierter Fluss (z. B. Batch Processing)
+		\item Datenorientierter Fluss (z. B. Batch Processing / Streams)
 		\item Datenzentriert (z. B. Repositories oder Blackboards)
@@
 	\end{itemize}
-
-	% Technologiearchitektur
-
-	% Containerarchitektur? Komponenten Wrappen und fachfreie Logik (sicherheit, auth, log / monitoring, etc.) in container auslagern
 \end{frame}
@@
 \begin{frame}{Prinzipien der Architektur}
-	% Die bisherigen Auflistungen von Werkzeugen und Stilmitteln, die dem Architekten zur Verfügung stehen, werden von verschiedenen Autoren wie beispielsweise von VOGEL ET AL.
-variiert, zusammengefasst und erweitert.
-
-	% Beispielhaft sind hier 10 wichtige Prinzipien aufgeführt.
-
 	\begin{itemize}
@@
 
-\section{Architekturen im Detail}
-
-\begin{frame}
-	Middleware im Detail
-
-	P2P (dezentrale Architektur) Im Detail
-	Sie organisieren sich selbst.
-
-    Sie handeln mit ihren Ressourcen autonom.
-
-    Dienste und Ressourcen können wandern.
-
-    Peers können anbieten und nutzen.
-
-    Verhalten, Zustand oder Durchsatz kann nicht vorhergesagt werden und erinnert eher an einen Schwarm.
-\end{frame}
-
-\begin{frame}{Architektur im Detail}{Domain-Specific-Language}
-	DSL etc hier her!!
-\end{frame}
-
-\begin{frame}
-	ADL statt UML manchmal
-\end{frame}
-
 \section{Frameworks}
@@
 \begin{frame}{Frameworks}
-	Es gibt frameworks, die ändern sich aber ständig.
+	Frameworks liefern generische projektübergreifende Lösungen für wiederkehrende Teilprobleme.
+
+	\vspace{0.5em}
+
+	Bekannte Frameworks:
+
+	\begin{itemize}
+		\item JavaEE
+		\item Spring
+		\item Django
+		\item Rails
+		\item .NET
+		\item ...
+	\end{itemize}
 \end{frame}
@@
 % https://vfhswt.eduloop.de/loop/Packages_und_Namensgebung_ARC
+
+\section{Zusammenfassung}
+
+\begin{frame}{Zusammenfassung}
+	Gute Architektur hilft Software..
+	\begin{itemize}
+		\item zu vereinheitlichen / zu standardisieren (u. a. Code, Tools..)
+		\item testbarer zu machen und die qualität sicherzustellen
+		\item erweiterbar und wartbar zu halten
+		\item übersichtlich zu halten
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Zusammenfassung}
+	Danke fürs zuhören! \textbf{Gibt es Fragen?}
+\end{frame}
+
 \end{document}

### 2021-05-27 softwaretechnik/assignments/uml/uml.tex
update graphic path for swt esa2
Add the assignment definition for swt assignment 2

@@
 \usepackage{graphicx}
-\usepackage{plantuml}
 
-\graphicspath{ {./images/} }
+\graphicspath{ {./diagrams/} }
 
@@
 
-\begin{plantuml}
-  @startuml
-  Alice -> Bob: Hello
-  Alice <- Bob: Hi!
-  @enduml
-\end{plantuml}
+Im Moodle-Kurs wurde folgende Aufgabenstellung bekannt gegeben:
+
+\begin{quote}
+	Die Aufgabenstellung benötigt Ihre bisherigen Ergebnisse aus der ESA1.
+	Bitte entwerfen Sie für Ihre in ESA1 entworfene Problemstellung zwei
+	verschiedene UML-Diagramme, die sich jeweils an verschiede
+	Interessenvertreter bzw. Projektteilnehmer richten sollen. Die Lösung MUSS
+	ein Klassendiagramm und ein Aktivitätsdiagramm enthalten. Begründen Sie,
+	welches weitere UML-Diagramm sie als nächstes erstellen würden. Beim
+	Klassendiagramm können Sie auf die detaillierte Darstellung von Methoden
+	verzichten, wichtige Attribute sollen jedoch enthalten sein. Achten Sie auf
+	richtige Verwendung von Assoziation, Aggregation und Komposition sowie auf
+	Multiplizitäten.
+\end{quote}
+
+\section{Klassendiagramm}
+
+\section{Aktivitätsdiagramm}
+
+% Er liebt Aktivitätsdiagramme für dynamische Diagramme
+
+% Interaktionsdiagramm
+
+% Klassen + Paket + Deployment + Activity + Zustand + Interaktion
 

### 2021-05-29 mensch-computer-interaktion/assignments/interaction-types/interaction-types.tex
Add the first draft version for mci assignment 2

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{fancyhdr}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Maximilian Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\hypersetup{colorlinks=true, linkcolor=black, filecolor=black, urlcolor=black}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik und Medien \\
		Mensch-Computer-Interaktion \\
		Prof. Dr. Martin Christof Kindsmüller
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 2: Interaktionsformen}\\[0.5em]
		\large{Sommersemester 2021}\\[0.25em]
		\large{Abgabetermin 01.06.2021}
	\end{center}

	\vfill

	\begin{flushright}
		Maximilian Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\tableofcontents

\vfill

\section{Aufgabenstellung}

Folgende Aufgabenstellung wurde im Moodle-Kurs bekannt gegeben:

\begin{quote}
	Führen Sie ein kleines Forschungsprojekt zur Positionierungsgeschwindigkeit
	und zur Positionierungsgenauigkeit verschiedener Zeigegeräte (Pointing
	Devices) durch und dokumentieren Sie Ihre Ergebnisse. Überprüfen Sie die
	Gültigkeit des Fitts's Law indem Sie verschiedene Zeigegeräte vergleichend
	testen. Verwenden Sie dazu mindestens zwei unterschiedliche Zeigegeräte,
	also beispielsweise zwei verschiedene Mäuse, Maus versus Trackpad oder
	Trackball oder Maus versus Touch-Interaktion mit einem Bildschirm. Ich
	empfehle Ihnen dazu die Fitts's Law Demonstration von Marcin Wichary zu
	verwenden. Abzugeben ist ein Forschungsbericht in dem die verwendeten
	Zeigegeräte und die Ergebnisse dokumentiert werden. Treffen Sie auch eine
	Aussage, ob Sie die vom Fitts's Law postulierten Effekte belegen konnten
	oder nicht. Falls Sie die Effekte mit einem Ihrer Zeigegeräte nicht
	replizieren konnten stellen Sie bitte Vermutungen an, woran das liegen
	könnte.
\end{quote}

\newpage

\section{Eingabegeräte}

Als Eingabegeräte habe ich mir meine alltags Maus (Logitech MX Master 3), das
eingebaute Trackpad meines ThinkPads T490 und den Touchscreen meines iPads
ausgesucht. Natürlich gibt es Unterschiede in der Benutzungsfrequenz dieser
Geräte, dennoch habe ich jedes der Eingabegeräte bereits über einen längeren
Zeitraum zur alltäglichen Arbeit eingesetzt und bin an diese gewöhnt.

\subsection{Maus: Logitech MX Master 3}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.5\textwidth]{maus.jpg}
	\caption{Logitech MX Master 3}
\end{figure}

\subsection{Trackpad: ThinkPad T490}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.5\textwidth]{trackpad}
	\caption{Trackpad vom ThinkPad T490}
\end{figure}

\subsection{Touch: iPad Pro 2020}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.5\textwidth]{ipad}
	\caption{iPad Pro 2020}
\end{figure}

\section{Auswertung einzelner Eingabegeräte}

Im folgenden sind die Ergebnisse des 4. Experiments (von der Seite
\href{https://www.aresluna.org/fitts}{\texttt{aresluna.org/fitts}}) aufgelistet
und erläutert.

\subsection{Maus: Logitech MX Master 3}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.75\textwidth]{experiment-4-maus}
	\caption{Auswertung der Maus in Experiment 4}
\end{figure}

Auffälligkeit: Roter Kreis am schnellsten. Würde ich aber auf Muscle Memory
zurückfüren

\subsection{Trackpad: ThinkPad T490 Trackpad}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.75\textwidth]{experiment-4-trackpad}
	\caption{Auswertung des Trackpads in Experiment 4}
\end{figure}

Bei einem Eingabegerät, mit dem nicht so regelmäßig gearbeitet wird,
gibt es hier auch weniger Abweichungen von der Formel.

\subsection{Touch: iPad Pro 2020}

%\begin{figure}[h!]
%	\centering
%	\includegraphics[width=\textwidth]{experiment-4-ipad}
%	\caption{Auswertung der Touch-Eingabe in Experiment 4}
%\end{figure}

Interessantestes Ergebnis!

\section{Vergleich}

\subsection{Maus vs Trackpad}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.75\textwidth]{experiment-5-maus-vs-trackpad}
	\caption{Auswertung des 5. Experiments (Maus in Blau, Trackpad in Rot)}
\end{figure}

Hier lässt sich klar erkennen, dass die Maus das überlegene Eingabegerät ist.
Zwar ist dieses Ergebnis durch meine generelle Tendenz zur Maus etwas beeinflusst,
aber dennoch ist der Zeitunterschied erheblich – teilweise hat sich die Eingabezeit
durch die Verwendung des Trackpads verdoppelt. Die Maus (in Blau) hat ziemlich
konstant die Bestzeiten für die jeweiligen $\frac{D}{W}$ Werte in diesem Test
erzielt und ist somit auf der $y$ Achse eigentlich durchgehend unten angesiedelt.

Die Werte für das Trackpad (in Rot) schwanken stark, was unter anderem dadurch
zu begründen sein dürfte, dass mit diesem wenig alltags Arbeit erledigt wird
und somit die Bedienung etwas ungesteuerter und zufälliger abläuft, als bei der
Maus.

\section{Bewertung der Ergebnisse}

Im großen und ganzen hat sich Fitts' Law bewahrheitet. Geräte, die oft verwendet
werden, haben zwar eine grundsätzlich geringere Interaktionszeit, aber die
Abstufung der Zeiten innerhalb eines Experiments liegen so vor, wie von Fitts'
Law beschrieben.

Interessanterweise konnte bei dem Experiment 4 aber beobachtet werden, dass bei
dieser Problemstellung Touchgeräte eine deutlich schnellere Interaktion erlauben.
Dies hängt vermutlich damit zusammen, dass kein Cursor auf dem Bildschirm bewegt
werden muss, sondern die Hand bzw.\ die Finger in echt auf das Ziel gerichtet
werden.

Abgesehen davon konnte unter den getesteten Zeigegeräten (Maus und Trackpad),
auch die Aussage validiert werden, dass die Maus das tendenziell bessere
Eingabegerät ist. Natürlich gibt es immer Einzelfälle, in denen dies nicht
zutreffen mag.

Ein Interessanter nächster Schritt könnte nun der Vergleich zwischen
Zeigegeräten und Tastaturbasierten Lösungen sein. Zum Beispiel gibt es die
Möglichkeit außschlieslich grafische Oberflöchen nur durch Shortcuts zu bedienen,
die dann abhängig von der jeweiligen Oberfläche generiert werden (siehe
\href{https://github.com/philc/vimium}{\texttt{github.com/philic/vimium}})

% Interessanter weise konnte fitts law bei den pointer basierten eingabegeräten bestätigtwerden, aber bei dem touchpad sieht das ganze etwas anders aus

% Text muss geändert werden!!

\end{document}

### 2021-06-05 mensch-computer-interaktion/assignments/interaction-types/interaction-types.tex
Finish the mci assignment 2

@@
 
-Als Eingabegeräte habe ich mir meine alltags Maus (Logitech MX Master 3), das
+Als Eingabegeräte habe ich mir meine Alltagsmaus (Logitech MX Master 3), das
 eingebaute Trackpad meines ThinkPads T490 und den Touchscreen meines iPads
@@
 
-Auffälligkeit: Roter Kreis am schnellsten. Würde ich aber auf Muscle Memory
-zurückfüren
+Auch wenn bei diesem Ergebnis auffällig ist, dass der rote Kreis am schnellsten
+ist, lässt sich das wahrscheinlich auf die Benutzungsfrequenz zurückführen.
+Alle anderen vorhergesagten Aussagen von Fitts' Law sind eingetreten ($T(Blau) <
+T(Gruen) < T(Gelb)$).
 
@@
 
-Bei einem Eingabegerät, mit dem nicht so regelmäßig gearbeitet wird,
-gibt es hier auch weniger Abweichungen von der Formel.
+Da dieses das wahrscheinlich am wenigsten genutzte Eingabegerät der Testgeräte ist,
+ist dieses Experiment sehr ``neutral'' im Vergleich zu den anderen
+Einzel-Experimenten.
 
@@
 
-%\begin{figure}[h!]
-%	\centering
-%	\includegraphics[width=\textwidth]{experiment-4-ipad}
-%	\caption{Auswertung der Touch-Eingabe in Experiment 4}
-%\end{figure}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.75\textwidth]{experiment-4-ipad}
+	\caption{Auswertung der Touch-Eingabe in Experiment 4}
+\end{figure}
 
-Interessantestes Ergebnis!
+Hieran lässt sich gut erkennen, dass bei dieser Art von Interaktionproblemen,
+Touchgeräte einen klaren Vorteil gegenüber Zeigegeräten haben. Fitts' Law
+schlägt in gewisser Weise fehl, da 3 der 4 Kreise eine sehr ähnliche Zeit,
+aber total unterschiedliche Größen und Entfernungen haben. Dementsprechend
+lässt sich die Vermutung aufstellen, dass bei einer relativ kleinen
+Eingabefläche, die größe der Elemente mehr Auswirkung auf die Interaktionszeit
+hat als die Entfernungen.
 
@@
 
-Hier lässt sich klar erkennen, dass die Maus das überlegene Eingabegerät ist.
-Zwar ist dieses Ergebnis durch meine generelle Tendenz zur Maus etwas beeinflusst,
-aber dennoch ist der Zeitunterschied erheblich – teilweise hat sich die Eingabezeit
-durch die Verwendung des Trackpads verdoppelt. Die Maus (in Blau) hat ziemlich
-konstant die Bestzeiten für die jeweiligen $\frac{D}{W}$ Werte in diesem Test
-erzielt und ist somit auf der $y$ Achse eigentlich durchgehend unten angesiedelt.
+An diesem Experiment lässt sich klar erkennen, dass die Maus das überlegene
+Eingabegerät ist. Zwar ist dieses Ergebnis durch meine generelle Tendenz zur
+Maus etwas beeinflusst, aber dennoch ist der Zeitunterschied erheblich –
+teilweise hat sich die Eingabezeit durch die Verwendung des Trackpads
+verdoppelt. Die Maus (in Blau) hat ziemlich konstant die Bestzeiten für die
+jeweiligen $\frac{D}{W}$ Werte in diesem Test erzielt und ist somit auf der $y$
+Achse eigentlich durchgehend unten angesiedelt.
 
 Die Werte für das Trackpad (in Rot) schwanken stark, was unter anderem dadurch
-zu begründen sein dürfte, dass mit diesem wenig alltags Arbeit erledigt wird
-und somit die Bedienung etwas ungesteuerter und zufälliger abläuft, als bei der
+zu begründen sein dürfte, dass mit diesem wenig Alltagsarbeit erledigt wird
+und somit die Bedienung etwas ungesteuerter und zufälliger abläuft als bei der
 Maus.
@@
 
-Im großen und ganzen hat sich Fitts' Law bewahrheitet. Geräte, die oft verwendet
+Im Großen und Ganzen hat sich Fitts' Law bewahrheitet. Geräte, die oft verwendet
 werden, haben zwar eine grundsätzlich geringere Interaktionszeit, aber die
@@
 Ein Interessanter nächster Schritt könnte nun der Vergleich zwischen
-Zeigegeräten und Tastaturbasierten Lösungen sein. Zum Beispiel gibt es die
-Möglichkeit außschlieslich grafische Oberflöchen nur durch Shortcuts zu bedienen,
+Zeigegeräten und tastaturbasierten Lösungen sein. Zum Beispiel gibt es die
+Möglichkeit außschlieslich grafische Oberflächen nur durch Shortcuts zu bedienen,
 die dann abhängig von der jeweiligen Oberfläche generiert werden (siehe
@@
 
-% Interessanter weise konnte fitts law bei den pointer basierten eingabegeräten bestätigtwerden, aber bei dem touchpad sieht das ganze etwas anders aus
-
-% Text muss geändert werden!!
-
 \end{document}

### 2021-06-06 softwaretechnik/assignments/uml/uml.tex
Include diagrams in the second swt assignment
Finish swt assignment 2

@@
 
-\graphicspath{ {./diagrams/} }
+\graphicspath{{./diagrams/}}
 
-\hypersetup{
-    colorlinks=true,
-    linkcolor=black,
-    filecolor=black,
-    urlcolor=black,
-}
+\hypersetup{colorlinks=true, linkcolor=black, filecolor=black, urlcolor=black}
 
@@
 	\begin{center}
-		\Large{Einsendeaufgabe 1: Requirements Engineering}\\[0.5em]
+		\Large{Einsendeaufgabe 2: UML}\\[0.5em]
 		\large{Sommersemester 2021}\\[0.25em]
-		\large{Abgabetermin 27.04.2021}
+		\large{Abgabetermin 06.06.2021}
 	\end{center}
@@
 
-\newpage
+\vfill
 
@@
 
+\newpage
+
 \section{Klassendiagramm}
+
+\vfill
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.55\textwidth]{class}
+	\caption{Klassendiagramm}
+\end{figure}
+
+\vfill
+
+Dieses Klassendiagramm könnte eine gute Dokumentationsgrundlage darstellen und
+einem Konfigurierenden einen guten Überblick über die grobe Struktur der
+Software und die Rolle seiner Konfiguration geben.
+
+\newpage
 
@@
 
-% Er liebt Aktivitätsdiagramme für dynamische Diagramme
+\vfill
 
-% Interaktionsdiagramm
+\begin{figure}[h]
+	\centering
+	\includegraphics[height=0.7\textheight]{activity}
+	\caption{Aktivitätsdiagramm}
+\end{figure}
 
-% Klassen + Paket + Deployment + Activity + Zustand + Interaktion
+\vfill
+
+Durch dieses Aktivitätsdiagramm können Entwickler eine Übersicht über den
+Ablauf des Main-Threads des Window Managers erhalten und dies als Gundlage
+für eine Implementierung verwenden.
+
+\section{Empfehlung für ein weiteres Diagramm}
+
+Als weiteres Diagramm könnte ein Paketdiagramm interessant sein, da dies für
+die Entwickler und die Konfigurierenden einen Mehrwert bietet.
 

### 2021-06-06 mensch-computer-interaktion/assignments/personas-and-storyboards/personas-and-storyboards.tex
use landscape pages for storyboards in the mci assignment
rebuild mci assignment 1
Refactor the whole mci assignment 2..

--- 2021-04-17 mensch-computer-interaktion/assignments/personas-and-storyboards/personas-and-storyboards.tex
@@
 \usepackage{graphicx}
+\usepackage{fancyhdr}
+\usepackage{pdfpages}
 
-\graphicspath{ {./images/} }
+\graphicspath{{./images/}}
+
+\pagestyle{fancyplain}
+\fancyhf{}
+\lhead{\fancyplain{}{Maximilian Schulke} }
+\rhead{\fancyplain{}{\today}}
+\cfoot{\fancyplain{}{\thepage}}
 
@@
 		\large{Sommersemester 2021}\\[0.25em]
-		\large{Abgabetermin 18.04.2021}
+		\large{Abgabetermin 01.05.2021}
 	\end{center}
@@
 
-\tableofcontents
-
-\vfill
-
 \section{Aufgabenstellung}
@@
 \begin{quote}
-	Sie haben die Aufgabe eine Smartphone-App zu konzipieren, die OSMI-Studierende beim Studium unterstützt.
-	Die App soll insbesondere die Kommunikation, die gegenseitige Unterstützung und das gemeinsame Bearbeiten
-	von Aufgaben und Projekten unterstützen. Denken Sie auch darüber nach, wie die App in der aktuellen
-	SARS-CoV-2-Situation helfen kann (also beispielsweise auch Studierende, die normalerweise in Präsenz studieren).
+	Sie haben die Aufgabe eine Smartphone-App zu konzipieren, die
+	OSMI-Studierende beim Studium unterstützt.  Die App soll insbesondere die
+	Kommunikation, die gegenseitige Unterstützung und das gemeinsame Bearbeiten
+	von Aufgaben und Projekten unterstützen. Denken Sie auch darüber nach, wie
+	die App in der aktuellen SARS-CoV-2-Situation helfen kann (also
+	beispielsweise auch Studierende, die normalerweise in Präsenz studieren).
 	\\[1em]
@@
 
-	(b) Erstellen Sie für jede Persona zwei Storyboards (also insgesamt vier) zur Nutzung der App. Storyboards
-	werden in der Regel von Hand gezeichnet (auf Schönheit kommt es in diesem Zusammenhang nicht an). Scannen
-	oder fotografieren Sie die Skizzen und beschreiben Sie ggf. kurz wie die Skizzen verstanden werden sollen.
+	(b) Erstellen Sie für jede Persona zwei Storyboards (also insgesamt vier)
+	zur Nutzung der App. Storyboards werden in der Regel von Hand gezeichnet
+	(auf Schönheit kommt es in diesem Zusammenhang nicht an). Scannen oder
+	fotografieren Sie die Skizzen und beschreiben Sie ggf.\ kurz wie die Skizzen
+	verstanden werden sollen.
 	\\[1em]
@@
 
-	Abzugeben ist ein PDF-Dokument, das Ihre Ausführungen enthält. Bitte beachten Sie dazu die Hinweise zu den
-	Einsendeaufgaben (siehe oben). Der zeitliche Umfang dieser Einsendeaufgabe wird auf 6 Stunden geschätzt.
-	Ihre Ausarbeitung sollte ca. 7-9 Seiten (A4) umfassen (eine Seite für Titelblatt inkl. Aufgabenstellung,
-	je eine Seite pro Persona, und je eine pro Storyboard sowie ggf. zusätzliche Seiten für die Erläuterungen).
-	Die Lösung, die Sie zur Deadline abgeben, sollte eine aus Ihrer Sicht endgültige Lösung sein. Falls Sie
-	Fragen zur Aufgabenstellung haben, stellen Sie diese bitte im Vorfeld!
+	Abzugeben ist ein PDF-Dokument, das Ihre Ausführungen enthält. Bitte
+	beachten Sie dazu die Hinweise zu den Einsendeaufgaben (siehe oben). Der
+	zeitliche Umfang dieser Einsendeaufgabe wird auf 6 Stunden geschätzt.  Ihre
+	Ausarbeitung sollte ca. 7-9 Seiten (A4) umfassen (eine Seite für Titelblatt
+	inkl. Aufgabenstellung, je eine Seite pro Persona, und je eine pro
+	Storyboard sowie ggf.\ zusätzliche Seiten für die Erläuterungen).  Die
+	Lösung, die Sie zur Deadline abgeben, sollte eine aus Ihrer Sicht
+	endgültige Lösung sein. Falls Sie Fragen zur Aufgabenstellung haben,
+	stellen Sie diese bitte im Vorfeld!
 \end{quote}
@@
 
-\section{Personas}
-
-\subsection{Mark Zahn}
-
-\begin{figure}[h]
-	\includegraphics[width=0.25\textwidth]{mark}
-	\centering
-	\caption{Portrait von ``Mark Zahn`` (Roman Holoschchuk, unsplash.com)}
-\end{figure}
-
-\begin{quote}
-	\large{``Ich verbringe meine Zeit gerne mit meiner Familie. Wenn ich Abends noch studiere, dann versuche
-		ich möglichst effizient durch den Stoff zu kommen. Bisher habe ich eher selten in Lerngruppen gearbeitet,
-		da mir der Austausch mit anderen Studenten immer sehr Zeitaufwendig vorkam.``}
-\end{quote}
-
-\begin{center}
-	\begin{tabularx}{\textwidth}{|l|X|}
-		\hline
-		\textbf{Demographische Daten} &                                                     \\
-		\hline
-		Name                          & Marc Zahn                                           \\
-		\hline
-		Alter                         & 28                                                  \\
-		\hline
-		Beruf                         & Hochschul-Verwaltungsangestellter                   \\
-		\hline
-		Bildungsabschluss             & Abitur                                              \\
-		\hline
-		Familienstand                 & Verheiratet                                         \\
-		\hline
-		Kinder                        & Ja                                                  \\
-		\hline
-		Wohnort                       & Brandenburg an der Havel                            \\
-		\hline
-		\textbf{Persönlichkeit}       &                                                     \\
-		\hline
-		Motivation für OSMI           & Möchte nun Umlernen, bekam Kontakt zum OSMI als
-		Umschulungsmöglichkeit, durch seinen Job an einer Hochschule                        \\
-		\hline
-		Hobbies                       & Alte Filme sammeln, Kochen, Wandern                 \\
-		\hline
-		OSMI-Student seit             & WS 2019                                             \\
-		\hline
-		Bedürfnisse                   & Erfolg, Soziale Anerkennung, Sicherheit für Familie \\
-		\hline
-		Wünsche                       & Neuer Job in Wunschbereich                          \\
-		\hline
-		Werte                         & Treue, Zuverlässigkeit, Ehrlichkeit                 \\
-		\hline
-		Ängste                        & Angst zu Versagen, Unsichere Zukunft                \\
-		\hline
-		Persönlichkeitstyp            & Introvertiert                                       \\
-		\hline
-	\end{tabularx}
-\end{center}
+\includepdf[pages=-,landscape]{./images/persona-mark.jpg}
 
@@
 
-\subsection{Rosa Beetz}
-
-\begin{figure}[h]
-	\includegraphics[width=0.25\textwidth]{rosa}
-	\centering
-	\caption{Portrait von ``Rosa Beetz`` (Svyatoslav Romanov, unsplash.com)}
-\end{figure}
-
-\begin{quote}
-	\large{``Ich bin gerne mit anderen Menschen zusammen und finde, dass das Studium
-		zusammen viel mehr Spaß macht und man besser lernt. Deshalb habe ich mit einigen
-		Kommilitonen eine Lerngruppe gegründet.``}
-\end{quote}
-
-\begin{center}
-	\begin{tabularx}{\textwidth}{|l|X|}
-		\hline
-		\textbf{Demographische Daten} &                                                     \\
-		\hline
-		Name                          & Rosa Beetz                                          \\
-		\hline
-		Alter                         & 23                                                  \\
-		\hline
-		Beruf                         & Screen Desginerin                                   \\
-		\hline
-		Bildungsabschluss             & Ausbildung                                          \\
-		\hline
-		Familienstand                 & Ledig                                               \\
-		\hline
-		Kinder                        & Nein                                                \\
-		\hline
-		Wohnort                       & Berlin Wedding                                      \\
-		\hline
-		\textbf{Persönlichkeit}       &                                                     \\
-		\hline
-		Motivation für OSMI           & Möchte sich in Richtung Informatik weiterentwickeln
-		und sich selbst etwas Beweisen                                                      \\
-		\hline
-		Hobbies                       & Netflix, Feiern, Kaffee                             \\
-		\hline
-		OSMI-Student seit             & SS 2021                                             \\
-		\hline
-		Bedürfnisse                   & Sozialer Kontakt, Anerkennung                       \\
-		\hline
-		Wünsche                       & Mehr Geld, Anerkennung im Job                       \\
-		\hline
-		Werte                         & Freundlichkeit, Ehrgeiz                             \\
-		\hline
-		Ängste                        & Angst zu Versagen                                   \\
-		\hline
-		Persönlichkeitstyp            & Extrovertiert                                       \\
-		\hline
-	\end{tabularx}
-\end{center}
+\includepdf[pages=-,landscape]{./images/persona-rosa.jpg}
 
@@
 
-\section{Storyboards}
-
-\subsection{Storyboard 1 für Mark Zahn}
-
-\begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{mark}
-	\centering
-	\caption{Storyboard Mark – Als OSMI Studen möchte ich xyz}
-\end{figure}
+\includepdf[pages=-,landscape]{./images/storyboard-1-mark.jpg}
 
@@
 
-\subsection{Storyboard 2 für Mark Zahn}
-
-\begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{mark}
-	\centering
-	\caption{Storyboard Mark – Als OSMI Student möchte ich unkompliziert mit anderen Studenten in einer PDF zusammenarbeiten, um den Zeitaufwand der Zusammenarbeit zu minimieren.}
-\end{figure}
+\includepdf[pages=-,landscape]{./images/storyboard-2-mark.jpg}
 
@@
 
-\subsection{Storyboard 3 für Rosa Beetz}
-
-\begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{rosa}
-	\centering
-	\caption{Storyboard Rosa – Als OSMI Student möchte ich Fahrgemeinschaften gründen, um mir Fahrtkosten zu sparen.}
-\end{figure}
+\includepdf[pages=-,landscape]{./images/storyboard-3-rosa.jpg}
 
@@
 
-\subsection{Storyboard 4 für Rosa Beetz}
-
-\begin{figure}[h]
-	\includegraphics[angle=90,width=\textwidth]{rosa}
-	\centering
-	\caption{Storyboard Rosa – Als OSMI Student möchte ich mit anderen Studenten Videotelefonieren, um eine zugängliche Plattform für unsere digitale Lerngruppe zu haben.}
-\end{figure}
-
-\newpage
-
-\section{Einsatzmöglichkeiten für Präsenzstudenten}
-
-Foo baz
+\includepdf[pages=-,landscape]{./images/storyboard-4-rosa.jpg}
 

### 2021-06-22 mensch-computer-interaktion/assignments/interaction-types/interaction-types.tex
Fix the submission date
Remove table of contents and update submission date
Refactor the second mci assignment

@@
 		\large{Sommersemester 2021}\\[0.25em]
-		\large{Abgabetermin 01.06.2021}
+		\large{Abgabetermin 07.06.2021}
 	\end{center}
@@
 \end{titlepage}
-
-\tableofcontents
-
-\vfill
 
@@
 Als Eingabegeräte habe ich mir meine Alltagsmaus (Logitech MX Master 3), das
-eingebaute Trackpad meines ThinkPads T490 und den Touchscreen meines iPads
-ausgesucht. Natürlich gibt es Unterschiede in der Benutzungsfrequenz dieser
-Geräte, dennoch habe ich jedes der Eingabegeräte bereits über einen längeren
-Zeitraum zur alltäglichen Arbeit eingesetzt und bin an diese gewöhnt.
+eingebaute Trackpad meines 14'' ThinkPads T490 und den Touchscreen und Apple
+Pencil meines 11'' iPads ausgesucht. Natürlich gibt es Unterschiede in der
+Benutzungsfrequenz dieser Geräte, dennoch habe ich jedes der Eingabegeräte
+bereits über einen längeren Zeitraum zur alltäglichen Arbeit eingesetzt und bin
+an diese gewöhnt.
 
@@
 
-\subsection{Trackpad: ThinkPad T490}
+\subsection{Trackpad: 14'' ThinkPad T490}
 
@@
 
-\subsection{Touch: iPad Pro 2020}
+\subsection{Touch \& Pencil: iPad Pro 11'' 2020}
 
@@
 
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.75\textwidth]{experiment-4-maus}
-	\caption{Auswertung der Maus in Experiment 4}
-\end{figure}
+\vspace{0.5em}
 
-Auch wenn bei diesem Ergebnis auffällig ist, dass der rote Kreis am schnellsten
-ist, lässt sich das wahrscheinlich auf die Benutzungsfrequenz zurückführen.
-Alle anderen vorhergesagten Aussagen von Fitts' Law sind eingetreten ($T(Blau) <
-T(Gruen) < T(Gelb)$).
+\begin{tabularx}{\textwidth}{|X|X|X|X|X|}
+	\hline
+	Durchführung & Blauer Kreis & Roter Kreis & Grüner Kreis & Gelber Kreis \\
+	\hline
+	1            & $601ms$      & $592ms$     & $820ms$      & $950ms$      \\
+	\hline
+	2            & $542ms$      & $655ms$     & $644ms$      & $871ms$      \\
+	\hline
+	3            & $515ms$      & $567ms$     & $711ms$      & $813ms$      \\
+	\hline
+	\hline
+	Durchschnitt & $553ms$      & $604ms$     & $725ms$      & $878ms$      \\
+	\hline
+\end{tabularx}
+\vspace{0.5em}
 
-\subsection{Trackpad: ThinkPad T490 Trackpad}
+Alle vorhergesagten Aussagen von Fitts' Law sind eingetreten ($T(Blau) < T(Rot)
+< T(Gruen) < T(Gelb)$) – die einzige Auffälligkeit ist der schon klare Unterschied
+von ca. $125ms$ zwischen dem roten und grünen Kreis, die laut Fitts' Law näher bei
+einander sein sollten, da sie die gleichen $\frac{D}{W}$ Werte haben.
 
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.75\textwidth]{experiment-4-trackpad}
-	\caption{Auswertung des Trackpads in Experiment 4}
-\end{figure}
+\subsection{Trackpad: ThinkPad T490 (14'') Trackpad}
 
-Da dieses das wahrscheinlich am wenigsten genutzte Eingabegerät der Testgeräte ist,
-ist dieses Experiment sehr ``neutral'' im Vergleich zu den anderen
-Einzel-Experimenten.
+\vspace{0.5em}
 
-\subsection{Touch: iPad Pro 2020}
+\begin{tabularx}{\textwidth}{|X|X|X|X|X|}
+	\hline
+	Durchführung & Blauer Kreis & Roter Kreis & Grüner Kreis & Gelber Kreis \\
+	\hline
+	1            & $607ms$      & $775ms$     & $867ms$      & $1285ms$     \\
+	\hline
+	2            & $1101ms$     & $994ms$     & $1282ms$     & $1759ms$     \\
+	\hline
+	3            & $762ms$      & $876ms$     & $957ms$      & $1422ms$     \\
+	\hline
+	\hline
+	Durchschnitt & $823ms$      & $881ms$     & $1035ms$     & $1488ms$     \\
+	\hline
+\end{tabularx}
 
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.75\textwidth]{experiment-4-ipad}
-	\caption{Auswertung der Touch-Eingabe in Experiment 4}
-\end{figure}
+\vspace{0.5em}
+
+Das Trackpad hat mit Abstand die höchste durchschnittliche Eingabezeit – es ist
+zwischen 50\% und 150\% langsamer als die Durchschnittswerte der anderen
+Eingabegeräte. Dies könnte unter anderem durch die, im Vergleich zu den anderen
+Eingabegeräten / Methoden, seltene Verwendung der Fall sein. Wie bei 3.1 ist auch
+hier der Abstand zwischen dem roten und grünen Kreis mit ca. $150ms$ deutlich
+spürbar.
+
+\subsection{Touch: iPad Pro (11'') 2020}
+
+\vspace{0.5em}
+
+\begin{tabularx}{\textwidth}{|X|X|X|X|X|}
+	\hline
+	Durchführung & Blauer Kreis & Roter Kreis & Grüner Kreis & Gelber Kreis \\
+	\hline
+	1            & $399ms$      & $372ms$     & $407ms$      & $551ms$      \\
+	\hline
+	2            & $406ms$      & $411ms$     & $475ms$      & $658ms$      \\
+	\hline
+	3            & $401ms$      & $494ms$     & $373ms$      & $565ms$      \\
+	\hline
+	\hline
+	Durchschnitt & $403ms$      & $426ms$     & $418ms$      & $591ms$      \\
+	\hline
+\end{tabularx}
+
+\vspace{0.5em}
 
@@
 lässt sich die Vermutung aufstellen, dass bei einer relativ kleinen
-Eingabefläche, die größe der Elemente mehr Auswirkung auf die Interaktionszeit
-hat als die Entfernungen.
+Eingabefläche (in diesem Fall das 11'' Display des iPads), die Größe der
+Elemente mehr Auswirkung auf die Interaktionszeit hat als die Entfernungen.
+
+\subsection{Pencil: iPad Pro (11'') 2020}
+
+\vspace{0.5em}
+
+\begin{tabularx}{\textwidth}{|X|X|X|X|X|}
+	\hline
+	Durchführung & Blauer Kreis & Roter Kreis & Grüner Kreis & Gelber Kreis \\
+	\hline
+	1            & $548ms$      & $578ms$     & $606ms$      & $706ms$      \\
+	\hline
+	2            & $554ms$      & $675ms$     & $656ms$      & $654ms$      \\
+	\hline
+	3            & $417ms$      & $667ms$     & $698ms$      & $633ms$      \\
+	\hline
+	\hline
+	Durchschnitt & $506ms$      & $640ms$     & $653ms$      & $664ms$      \\
+	\hline
+\end{tabularx}
+
+\vspace{0.5em}
+
+Auch hier tritt das Phänomen auf, das in 3.3 beschrieben wurde.
+Interessanterweise sind hier die Zeiten aller Kreise nahezu identisch, bis auf
+die des blauen Kreises, welche rund 25\% unter den anderen liegt.
 

### 2021-10-13 mensch-computer-interaktion/notes/notes.tex
Add my mci exam note paper

\documentclass{article}

\usepackage[a4paper, margin=1em]{geometry}
\usepackage[dvipsnames]{xcolor}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{multicol}
\usepackage{plex-mono}

\begin{document}

\ttfamily

\begin{multicols}{3}
	\begin{flushleft}
		\begin{tiny}
			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Bestimmung des
				Kontextes}}}

			\textbf{\# Setze Kontext, Nutzungskontext, Gebrauchtstglkt.\ in
				Bezug}

			Der Nutzungskontext ist wichtig zur Ermittlung der
			Gebrauchstauglichkeit.

			\textbf{\# Welche Daten bei Ermittlung des Kontextes
				berücksichtigen?}

			Ob User bei hellem Licht oder in der Dunktelheit das System nutzen,
			Ob User während der Benutzung im Stress ist. Die technische
			Umgebung des Systems. Also alle technischen Hilfs- und
			Arbeitsmittel des Benutzers

			\textbf{\# Unterschied zw. Arbeitsmittel und Kontext}
			Kontext umfasst technische, physische,
			gesellschaftliche Umgebung; Arbeitsmittel
			sind alle Werkzeuge

			\textbf{\# Einführung neuer beruflicher Software verändert
				Arbeitsprozesse}
			Richtig, da Einf.\ einer brfl genutzt Software m.Zielen
			verbunden ist, die normalerws. Auswirkungen auf die
			Arbeitsproz.hat (Einsparungen, Qualiverbesserung, Änderung d.\
			Kommunikationsstrukturen, Controlling o.ä.)

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Menschengerechte
				Arbeitsgestaltung}}}

			\textbf{\# Begriffe Grundmodell Hacker}
			1.Ausführbar
			2.Schädigungslos
			3.Beeinträchtigungsfrei
			4.Persönlichkeitsfördernd
			6 Humankriterien
			1.Benutzerorientierung
			2.Anforderungsvielfalt
			3.Ganzheitlichkeit und Bedeutsamkeit
			4.Handlungsspielräume
			5.Rückmeldungen
			6.Entwicklungsmöglichkeiten

			\textbf{\# Ganzheitlichkeit und Bedeutsamkeit}
			Bedeutung von Aufgabe kann erkannt und in Gesamtablauf eingeordnet
			werden. Ergebnis / Arbeitsfortschritt kann selbst beurteilt werden
			=> steigert Zufriedenh. Mitarbeiter sollte Sinnhaftigkeit und
			Bedeutsamkeit einer Aufg.\ verstehen.

			\textbf{\# Unterschied Belastung und Beanspruchung?}
			\textit{Belastung:} Äußere Bedingungen und Anforderungen;
			\textit{Beanspruchung:} Reaktion des Individuums auf Belastung
			Überforderung kann zu innerer Anspannung, Angst, Erschöpfung
			führen; Unterforderung zu Langeweile, Unlust und dazu, dass Arbeit
			schlecht gemacht

			\textbf{\# Ziel menschengerechter Arbeitsgestaltung?}
			Gestaltung der ergonomischen Maßnahmen (Arbeitspl.\
			u.Arbeitsmittel) Gestaltung der organisatorischen und sozialen
			Maßnahmen (Arbeitszeiten, Hierarch, Informatnsfls in Firma

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Grundsätze der
				Dialoggestaltung}}}

			\textbf{\# Definition Dialog}
			Interaktion zw. Benutzer und interaktiven System
			in Form einer Folge von Handlungen des Benutzers (Eingaben) und
			Antworten des interaktiven Systems (Ausgaben), um ein Ziel zu
			erreichen.

			\textbf{\# Definition Benutzungsschnittstelle}

			„Alle Bestandteile eines interaktiven Systems (Software oder
			Hardware), die Informationen und Steuerelemente zur Verfügung
			stellen, die für den Benutzer notwendig sind, um eine bestimmte
			Arbeitsaufgabe mit dem interaktiven System zu erledigen.“

			\textbf{\# Was ist der Nutzen von Gestaltungsgrundsätzen für
				Dialoge?}

			- gut gestalt. Dialoge haben großen Anteil daran, wie gut
			  Gebrauchsttglkt.
			- Die Gestaltungsgrundsätze geben konkrete Kriterien wie Dialoge
			  gebrauchstauglich gestaltet werden.

			\textbf{\# Die 7 Grundsätze der Dialoggestaltung nach
				ISO 9241-110}

			1. Aufgabenangemessenheit (Möglichst einfach gestaltet)
			2. Selbstbeschreibungsfahigkeit (Wo befinde ich mich / was muss ich
			   tun)
			3. Erwartungskonformität (Dialog ist intuitiv nutzbar - beispiel:
			   Überweisungsträger)
			4. Lernförderlichkeit (Tutorials + direkt erkennbar was getan
			   werden soll)
			5. Steuerbarkeit (Nutzer kann Aktionen selber abbrechen/starten)
			6. Fehlertoleranz (Kann mit minimalen Aufwand die meisten Fehler
			   beheben)
			7. Individualisierbarkeit (Nutzer kann die System ändern)

			\textbf{\# 5 Erweit.\ der Grundsätze der Dialoggestaltung
				nach Sarodn.\ und Brau?}
			1. Prozessangemessenheit
			2. System- und Datensicherheit
			3. Wahrnehmungssteuerung
			4. Interkulturelle Aspekte
			5. Joy of use

			\textbf{\# Shneidermans 8 goldene Regeln des
				Dialogdesigns?}

			1. Streben nach Konsistenz
			2. Universelle Gebrauchstauglichkeit
			3. Biete informative Rückmeldungen
			4. In sich geschlossene Dialoge (Dialoge sollten eine klar
			   erkennbare Struktur haben)
			5. Verhindere Fehler
			6. Erlaube einfache Rücksetzungen (Undo)
			7. Unterstütze das eigene Kontrollbedürfnis (Nutzer steuert
			   Software, nicht anders herum)
			8. Reduziere die Belastung des Arbeitsgedächtnisses

			\textbf{\# Welche Shneiderman/ISO Grunds.werden verl wenn
				Softwareanw.autonom korrigiert}

			- Lernförderlichkeit – DIN ISO 9241-110
			- Steuerbarkeit – DIN ISO 9241-110
			- Fehlertoleranz – DIN ISO 9241-110
			- Biete informative Rückmeldungen – Shneiderman
			- Erlaube einfache Rücksetzung – Shneiderman
			- Unterstütze das eigene Kontrollbedürfnis – Shneiderman

			\textbf{\# 3 Arten der Konsistenz / Begriffe
				Erwartungskonformität?}
			- Äußere (OS)
			- Metaphorische (zu real Welt.)
			- Innere (in app)

			\textbf{\# Vorteile von Styleguides in Bez.\ auf die
				Dialoggrundsätze?}

			Einheitliches Design => ``Erwartungskonformität''

			\textbf{\# ``Fehlererkennung'' Beispiel}

			- System validiert die Eingaben (z.B. Überprüfung eines
			  PLZ-Feldes auf korrekte Syntax und Länge)

			\textbf{\# ``Fehlervermeidung'' Beispiel}

			- Kennzeichnung von Pflichtfeldern oder
			- Auswahloptionen verwenden statt Freitexteingaben

			\textbf{\# ``Fehlerkorrektur'' Beispiel}

			- System gibt Korrekturvorschläge (z.B. Rechtschreibung)

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Interaktionsformen}}}

			\textbf{\# Welches Gesetz beschr. Zeit zur Positionierung in Abh.\
				von Entf.\ und Zielgröße?}
			Fitts’ Law

			\textbf{\# Womit die Interaktionszeiten für den Wechsel von Maus
				und Tastatur abschätzen?}
			GOMS

			\textbf{\# Gesetz das sagt, dass Zeiten f.Erkennung/Auswahl
				abhängig von Anz. Menüpkte?}
			Hicks’ Law

			\textbf{\# Welche Interaktionsformen für Touch-Systeme sehr gut?}

			- Interakt.\ via Icons oder via Kontextmenü
			- Modale Dialoge

			\textbf{\# Worum geht es im GOMS Modell?}

			- durschnittliche Zeiten für typische Interaktionen
			- Anhand dieser Zeiten kann komplette Interaktion Ausgewertet

			\textbf{\# Nennen Sie mögl Interaktionsformen mit Tastatur}

			- Kommandosys.
			- FN Keys
			- Shrtcts
			- Cursorblock \& Pfeiltstn

			\textbf{\# Was sind Vor- und Nachteile von Kommandosystemen}

			Vorteile:
			- Nur Tastatur
			- History
			- Experten
			- Scriptable
			- Automatisierbar
			Nachteile:
			- Hoher Lernaufwand
			- System führt die Befehle ohne vis Rückmeldung aus ``Trial and
			  Error''
			- Tippfehler sind leicht möglich.

			\textbf{\# Erläutern Sie die Vor und Nachteile von Funktionstasten}

			Vorteile:
			- Kann Interaktionszeiten stark reduzieren
			Nachteile:
			- Eventuell ungewohnt wenn die Belegung nicht individualisiert
			  werden kann

			\textbf{\# Nennen Sie Vor- und Nachteile von Shortcuts}

			Vorteile:
			- Geschwindigkeit
			- Gut für Blinde und Sehbehinderte
			Nachteile:
			- Verwirrend für Anfänger
			- ggf.\ unergonomisch bei schlechter Wahl der Kombi

			\textbf{\# Vor- und Nachteile von Cursorblock + Pfeiltasten}

			Vorteile:
			- Schnelle Navigation durch Oberfächen jeglicher Art
			- Deutlich effizienter als wechsel zur Maus
			Nachteile:
			- Software muss Tasten unterstützen

			\textbf{\# Mögliche Interaktionsarten mit Zeigegeräten}

			- Menüs
			- Ribbons
			- Kontextmenüs
			- Mausgesten
			- Dialogfenstern mit Erweiterung

			\textbf{\# Wofür steht WIMP?}

			- Windows
			- Icons
			- Menus
			- Pointing Device

			\textbf{\# Was sind Vor- und Nachteile von WIMP Interfaces?}

			Vorteile:
			- sehr hohe Lernförderlichkeit
			- Objekte und Aktionen sind direkt erkennbar
			- direkte Reaktion auf Aktionen des Nutzers
			- Handlungen sind nah zur realen Welt
			  (statt Kommandos - Beispiel: Datei verschieben)
			Nachteile:
			- Langsam und ineffizient im Vergleich zu anderen
			  Interaktionsformen
			- ggf.\ hohe Frequenz der Wechsel zwischen Tastatur und Maus
			- bei komplexer Software müssen Menüs und Icons müssen gesucht
			  werden

			\textbf{\# Wann bieten sich Touch-Anwendungen an?}

			kurze Interaktionsdauer

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Interaktionsdesign}}}

			\textbf{\# Nenne charakteristische Eigensch.\ von dargestellten
				Infos (nach ISO 9241-12)}

			- Klarheit
			- Unterscheidbarkeit
			- Kompaktheit
			- Konsistenz
			- Erkennbarkeit
			- Lesbarkeit
			- Verständlichkeit

			\textbf{\# Welche Interaktionselemente gibt es?}

			1. Radiobuttons und Checkboxen
			2. Dropdown- und Auswahllisten
			3. Eingabefelder
			4. Schaltflächen

			\textbf{\# Was sind Masken? Was sind Strukturblöcke?}

			- Eingabemaske (=Maske) stammt aus Zeit in der
			  Computer grüne Bildschirme nur 25 Zeilen je 80 Chars anzeigen
			  konnten
			- Strukturblock = Gruppierung von Controls -> abgegrenzte
			  Informations- oder Interaktionsmöglichkeit

			\textbf{\# Fluchtlinien und wie sollte verwenden?}

			- FL = die Linien zu Kanten und Seiten von Interaktionsel
			- sollten auf ein Mindestmaß reduziert werden

			\textbf{\# Was bei Abständen zwisch Interaktionsel beachten?}

			- Gesetz der Nähe für vertikale und horiz Abstände
			- Abstände zu beschreibenden Texten berücksichtigen

			\textbf{\# Was ist bei der Gruppierung von Informationen zu
				beachten?}

			- Fachlich zusammenhängende Interaktionselemente sollten
			  entsprechend der erwarteten Reihenfolge gruppiert werden

			\textbf{\# Wie sind Styleguides aufgebaut?}

			1. Einleitung: Version, Datum und Autor, Zweck + Einsatzgeb,
			   Zielpublikum
			2. Konzept: Produktvision, UX-Ziele, Accesblty,
			   Designprnzpien
			3. Interaktnsmuster
			4. Struktur: Navigtnkonzept, Seitenlayout, Raster
			5. Visuelle Gestaltung: Col, Font, Grafiken, Interaktionsel in
			   exaktem Design
			6. Kommunikationsstil: Stil Texte, Wortwahl, Stil Bilder,
			   Einsatz von Audio, Video
			7. Weiteres: Beispiele, Werkzeuge

			\textbf{\# 2 Maßnahmen um ein System gut bedienbar für
				Farbfehlsichtige zu machen}

			- Nicht nur Farben zur Kodierung von Hinweisen/Information 
			  -> Symbole/Schraffuren/Kontrast
			- UIs mit Usability-Tests mit Farbfehlsichtigen testen

			\textbf{\# Nenne 4 Usability-Aspekte die für Senioren zu
				berücksichtigen sind + begründen}

			- Große Schrift und Schriftgröße auf einfache Weise noch stärker
			  vergrößerbar, da die Sehfähigkeit oft eingeschränkt ist
			- Große Interaktionsflächen (Buttons) und große Abstände zwischen
			  den Interaktionsflächen, da die motorischen Fähigkeiten oft
			  eingeschränkt sind
			- Einfache Wortwahl, einfache Sätze
			- einfache Navigation (wenige Menüpunkte), da kognitive Leistungen
			  eingeschr.\ sein können

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Usability
				Engineering}}}

			\textbf{\# Bedeutung Usability Engineering?}

			- beschreibt diesen Prozess, wie parallel zum Software Engineering
			  auch die Gebrauchstauglichkeit berücksichtigt werden kann
			- ``Engineering'' = strukturierte, methodische, prozess- und
			  phasenorientierte Vorgehensweise

			\textbf{\# 7 Schrt. Phasenmod nach Sarodnick und Brau}

			1. Analyse der Arbeit und des Arbeitsumfeldes
			2. Analyse der Benutzergruppen
	 		3. Bestimmung von Anforderungen
			4. Entscheidung über Funktionalität und Ableitung eines Handlungs-
			   und Bedienkonzeptes
			5. Entwicklungsbegleitende Evaluation und Verbesserung des Systems
			6. Einführung und Schulung
			7. Weiterentwicklung

			\textbf{\# 4 Phsn Usb. Engin\. nach Sarodnick u. Brau}

			1. Analysephase (Protokoll, Personas und gfx drstlng der Prozesse)
			2. Konzeptphase (visuelles Rohgerüst als Papier-Prototyp(en))
			3. Entwicklungsphase (fertiges System)
			4. Einführungsphase (Usability-Tests, Fragebögen,
			   Interviewergebnisse)

			\textbf{\# Arbeitsschritte der Analysephase}

			- Arbeitsanalyse (Arbeitsanforderungen, -bedingungen, -prozesse)
			- Prozessanalyse
			- Systemanalyse
			- Erhebnung von Nutzeranforderungen

			\textbf{\# Arbeitsschritte der Konzeptphase}

			- Arbeitsgestaltung und Prozessdefinition
			- Entscheidung über Systemfunktionalitäten
			- Konzepterstellung

			\textbf{\# Wo ist Phasenmodell v.Sarodnick/Brau
				m.Vorgehensm.d.Softwaretechnik verknüpft?}

			Beim Unterpunkt „Systemintegration“ verlassen wir im
			Projektmanagement das Phasenmodell des Usability Engineering und
			wechseln in das Vorgehensmodell der Softwaretechnik. Nach den
			Regeln der Softwaretechnik wird das System erstellt und nachdem das
			lauffähige System vorliegt, erfolgt wieder der Wechsel in das
			Phasenmodell des Usability Engineering

			\textbf{\# Mögliche Reaktionen von User auf neues System + Gründe?}

			- gerade Nutzer im betrieblichen Umfeld reagieren auf Neuerungen
			  oftmals nicht positiv
			- viele Nutzer reagieren skeptisch und mit geringer Akzeptanz

			Gründe:
			Unterbrechung der Routine, Neulernen, Entwertung der Qualifikation,
			Statusverlust, Doppelbelastung während der Einführung, Verlust von
			Freiräumen, Angst vor Arbeitsplatzverlust

			\textbf{\# Was kann Akzeptanz der User ggü.\ neu
			eingeführten/weiterentw. System erhöhen?}

			- gute Nutzerbeteiligung (optimal in allen Phasen)

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Heuristiken}}}

			\textbf{\# Nenne die Kriterien der heuristischen Evaluation (nach
				Nielsen)}

			1. Sichtbarkeit des Systemzustands
			2. Übereinstimmung von Systemzustand und Realwelt
			3. Freiheit der Benutzersteuerung, ``Notausgang''
			4. Konsistenz und Einhaltung von Konventionen und Standards
			5. Fehlerverhinderung
			6. Wiedererkennen vor Erinnerung
			7. Flexibilitt und Nutzungseffektivität (Abkürzungen für geübte
			   Benutzer)
			8. Aesthetisches und minimalistisches Design
			9. Unterstuetzung beim Erkennen, Deuten und Beheben von Fehlern
			10. Hilfe und Dokumentation

			\textbf{\# Definieren sie Heuristik}

			- die Kunst, mit begrenztem Wissen und wenig Zeit zu guten Lösungen
			  zu kommen

			\textbf{\# Definition heuristische Evaluation?}

			- Verfahren zur Problemlösung durch analytische Betrachtung und
			  systematisches Probieren

			\textbf{\# Grober Ablauf einer heuristischen Evaluation?}

			- Usability-Experten bewerten die Software mit anerkannten
			  Gestaltungsgrundsätze und geben konkrete Verbesserungsvorschl.
			- Dialoge werden einzeln betrachtet und jeder Dialog für sich wird
			  gegen die Heuristiken geprüft - Hierbei werden keine Personas
			  oder Use-Cases eingesetzt

			\textbf{\# Was sind typische Heuristiken?}

			- Shneidermans 8 goldene Regeln
			- Usability-Prinzipien von Nielsen
			- Heuristiken nach Sarodnick und Brau

			\textbf{\# Wie haengen die 8 goldenen Regeln mit der heuristischen
				Evaluation zusammen?}

			- die 8 goldenen Regeln sind eine Heuristik Allgemein anerkannte
			  Kriterien, mit denen Dialoge geprüft und verbessert und evaluiert
			  werden können.
			- Wiederholte anlegen gleich. Kriterien (8 Reglen) können für
			  Statistiken genutzt werden => Fließen wieder in Evaluation ein.

			\textbf{\# Erläutern Sie die Begriffe "Formative Evaluation" und
				``Summative Evaluation''}

			Formative Evaluation:
			- Evaluationsmethoden, die während der Entwicklung eines Systems
			  verwendet werden
			- Ergebnisse fließen direkt in den weiteren Entwicklungsprozess ein
			  -> Wie kann bei der Entwicklung das optimale Ergebnis erzielt
			  werden?

			Summutative Evaluation:
			- Evaluationsmethoden, die nach der Entwicklung eines Systems
			  verwendet werden und das System bewerten
			- Summativ = abschließend, zusammenfassend -> Wie gut
			  Gebrauchstauglichkeit? + Umfang erreichter Ziele

			\textbf{\# Was ist der Vorteil von formativer Evaluation?}

			- beantwortet Frage: Wie kann bei der Entwicklung das optimale
			  Ergebnis erzielt werden? -> es wird effizienter entwickelt
			- bezüglich Zeiten und Kosten ist es viel besser während des
			  Entwicklungsprozesses zu evaluieren, als am Ende die fertige
			  Entwicklung zu betrachten

			\textbf{\# Ordnen sie dem Prinzip ``Benutzerbasierte Evaluation''
				Methoden zu}

			- Befragung (Fragebogen, Interview) - Usability-Test - Eye-Tracking
			- Beobachtung - Thinking Aloud

			\textbf{\# Ordnen Sie dem Prinzip Menschzentrierte Gestaltung
				Methoden zu}

			- Informationsanalyse (Datenanalyse, Dokumentensichtung)
			- Befragung (Fragebogen, Interview)
			- Personas
			- Usability-Test
			- Storyboards
			- Zukunftswerkstätten
			- Fokusgruppen

			\textbf{\# Ordnen Sie dem Prinzip ``Theoriebasierte Evaluation''
				Methoden zu}

			- Informationsanalyse (Datenanalyse, Dokumentensichtung)
			- Inspektionsmethoden (Cognitive Walkthrough, Heuristische
			  Evaluation)
			- Personas
			- Styleguide

			\textbf{\# Welche Maßnahmen gehören zu Cognitive Walkthrough?}

			- Testziele festlegen
			- Typische Szenarien bestimmen
			- Zielgruppe definieren (z.B. Personas/Storyboards erstellen)
			- Testdurchführung
			- Ergebnisse bewerten

			\textbf{\# Welche Maßnahmen gehören zu einem Usability Test?}

			- Testziele festlegen
			- Testablauf festlegen (Testkonzept)
			- Auswerteverfahren festlegen
			- Probanden auswählen
			- Probedurchlauf vornehmen
			- Testdurchführung

			\textbf{\# Nennen Sie jeweils 3 Vor- und Nachteile von
				Beobachtungen}

			Vorteile:
			- sehr gut geeignet um das Arbeitsgebiet des Benutzers kennen zu
			  lernen
			- im Rahmen einer Arbeits- und Aufgabenanalyse entsteht ein gutes
			  Bild von der realen Benutzung einer Software
			- besonders für Routineaufgaben geeignet (für die Optimierung
			  v.Arbeitsabläufen)

			Nachteile:
			- reine Beobachtung ist für Evaluation eher ungeeignet, besser
			  Kombination aus Interview und Beobachtung 
			- verdeckte Beobachtungen sind für eine Evaluation eher ungeeignet
			- Mitarbeiter verhalten sich bei angekündigten Beobachtungen
			  zurückhaltender

			\textbf{\# Was versteht man unter menschzentrierter Gestaltung?}

			- die Benutzer werden in den Entwicklungsprozess einbezogen

			\textbf{\# Wie User bei menschzentr.Gestalt.in Prozess der
				Sys.entw.\ einbeziehen?}

			Wie wird beteiligt?:
			- aktive Mitentscheidung (über Funktionsumfänge etc.)
			- Aktive Partizipation: Die Benutzer werden in den frühen
			  Gestaltungsphasen direkt gestaltend tätig (Arbeitsprozesse
			  abbilden, Designentwürfe vorbringen)

			Wann wird beteiligt?:
			- während des gesamten Entwicklungsprozesses oder
			- während eines Teils des Entwicklungsprozesses oder
			- zu ausgewählten Zeitpunkten in Workshops

			Woran wird beteiligt?
			- Beteiligung kann sich auf alle Einzelbereiche der Software
			  (Prozesse, Funktionalitäten, Schnittstellen etc.) beziehen

			\textbf{\# Was haben „Cognitive Walkthrough“ und „Heuristische
				Evaluation“ gemeinsam?}

			- beides Inspektionsmethoden (= Expertenbasierte Methoden = keine
			  Einbeziehung der tatsaechlichen Benutzer) und/oder
			- beide sind relativ schnell durchzufuehren und kostenguenstig

			\textbf{\# App f.Projektzeiten auf Gestaltunssgr\. untersuchen,
				worauf muss man achten?}

			- Aufgabenangemessenheit: kann man Projektzeiten mit wenigen Klicks
			  erfassen
			- Selbstbeschreibungsfähigkeit: Hat jeder App-Screen eine
			  klare Überschrift/ Seitenbenennung
			- Erwartungskonformität: Ist die
			  innere Konsistenz der Buttons bzgl. Anordnung und Aussehen geben
			- Lernförderlichkeit: Ist ein Link zu einem Tutorial/Video
			  vorgesehen
			- Individualisierbarkeit: Können verschiedene Projekte entsprechend
			  der Reihen- folge in der App verschoben werden
			- Steuerbarkeit: Sind alle Interaktionselemente leicht zugänglich
			  angebracht
			- Fehlertoleranz: sind möglichst wenige Input-Felder vorhanden und
			  wurden wo immer möglich stattdessen Auswahlmenüs verwendet

			\textbf{Beispiel-Heuristiken:}

			\textbf{Aufgabenangemessenheit:} kann man Projektzeiten mit
				wenigen Klicks erfassen
			\textbf{Selbstbeschreibungsfähigkeit} Hat jeder App-Screen eine
				klare Überschrift/Seitenbenennung
			\textbf{Erwartungskonformität} Ist die innere Konsistenz der
				Buttons bzgl. Anordnung und Aussehen geben
			\textbf{Lernförderlichkeit} Ist ein Link zu einem Tutorial/Video
				vorgesehen
			\textbf{Individualisierbarkeit} Können verschiedene Projekte
				entsprechend der Reihenfolge in der App verschoben werden
			\textbf{Steuerbarkeit} Nutzer kann aktionen abbrechen / undo /
				multistep formular (zwischen schritten springen)
			\textbf{Fehlertoleranz} Formular geht nicht kaputt bei falscher
				eingabe / Falsche eingaben werden verhindert

			\textbf{6 Usability-Maßnahmen:}

			\textbf{1. Beobachtung, Interviews} (mit Zielgruppe)
			\textbf{Ergebnis:} Nutzungskontextes, derzeitige Nutzung,
				Stärken / Schwächen.
			\textbf{2. Usability-Tests} (Eye-Tracking / lautes Denken zu
				wichtigen Aufgaben)
			\textbf{Ergebnis:} Usability-Probleme
			\textbf{3. Kreativtechniken} (Fokusgruppen / Card-Sorting zur
				Ideenfindung / prüfen von Konzeptideen)
			\textbf{Ergebnis:} Neue Ideen / Absichern
			\textbf{4. Papier-Prototypen} (möglichst viele verschiedene)
			\textbf{Ergebnis:} Vergleich Ideen / Konzepte -> Optimales
				Konzept
			\textbf{5. Funktionalen Prototypen + Cognitive Walkthrough}
			\textbf{Ergebnis:} Interaktionspfade + weitere Optimierung
			\textbf{6. Test \& Beobachtung} beim Umgang mit funkt.
				Prototypen
			\textbf{Ergebnis:} Weitere Optimierungen

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# 6-Ebenen-Modell}}}

			Es untersucht das Planen und Ausführen einer Aktion auf
			verschiedenen Ebenen, von der globalen Betrachtung (Intention =
			Tätigkeiten und Bewertung) bis hin zu der Motorik. Damit
			ermöglicht das Modell sehr gut die Analyse der GUI von der
			planenden Idee über die pragmatische, semantische, syntaktische
			und lexikalische Ebene bis zur motorischen Handlung. Es kann
			damit sehr gut erkannt werden, ob eine GUI auf allen Denk- und
			Handlungsebenen optimiert wurde Bsp: Zur Überprüfung der
			lexikalischen Ebene muss in einem Druckdialog jedes Eingabefeld
			auf alle möglichen Zeichen hin untersucht werden

			\textcolor{LimeGreen}{\textbf{\scriptsize{\# Beispiele}}}

			\textbf{\# Usability-Maßn.\ in Reihenfolge, um Website d.\
				Hochschule bez.Usability zu verbessern}

			1. Beobachtung u.Interviews m.d.Zielgruppe (Interessenten,
			Studierende, Angestellte) => \textit{Verstehen d.Nutzungskontextes
			u.d.derzeitigen Nutz- ung sowie Ermittlung v.Stärken u.Schwächen.}

			2. Durchführung v.Usability-Tests (auch mittels Eye-
			Tracking od.„lautem Denken“) zu wichtigen Aufg.\
			wie z.B. d.Immatrikulation,dem Auffinden v.Adress-
			daten, Ansprechpartner in d.Fachbereichen,Öffnungs-
			zeiten d.Bibliothek etc.). => \textit{Erm.d.Usability-Probleme.}

			3.(Kreativtechniken) Durchf.d.Methode „Fokusgr.“
			od.„Zukunftswerkstatt“od.„Card-Sorting“ zur Ideen-
			findung bzw.zur Überpr.v.Konzeptideen => \textit{Absicherung
			Bish.Ideen u.Findung neuer Ideen}

			4. Erstellen v.(möglichst vielen alternativen) Papier-Prototypen =>
			\textit{Vergl.\ untersch.Konz.z.Erm.eines optimalen Informations-
			und Bedienkonzepts.}

			5.Erstellung eines funktionalen Prototypen u.Durchf.eines Cognitive
			Walkthrough => \textit{Erm.d.Interaktionspfade aus Sicht
			d.Zielgruppe z.\ weiteren Optimierung d.Prototypen}

			6. Test + Beob.\ v.Probanden beim Umgang m.d.funktionalen Protot.
			(ggf.Nutzung v.Eye-Tracking) => \textit{Optimierung d.Protot.}

			Maximilian Schulke, 20215853
		\end{tiny}
	\end{flushleft}
\end{multicols}

\end{document}

### 2021-11-16 computergrafik/implementation/device/src/main.rs
Cleanup the dir

@@ -1,77 +1,0 @@
-use anyhow::anyhow;
-use async_std::net::TcpListener;
-//use async_tungstenite::tungstenite::Message;
-use async_std::sync::*;
-use futures::prelude::*;
-use gstreamer::prelude::*;
-use http_types::{Response, StatusCode};
-
-//use gstreamer::{parse_bin_from_description, ElementFlags, MessageType};
-use std::io::Read;
-//use std::io::Write;
-//use std::sync::Arc;
-//
-//
-
-const WEBRTC_PIPELINE: &str = "webrtcbin name=webrtcbin stun-server=stun://stun.l.google.com:19302 VIDEO_SRC ! videorate ! videoscale \
-                              ! video/x-raw,width=640,height=360,framerate=15/1 ! videoconvert ! queue max-size-buffers=1 \
-                              ! x264enc bitrate=600 speed-preset=ultrafast tune=zerolatency key-int-max=15 \
-                              ! video/x-h264,profile=constrained-baseline ! queue max-size-time=100000000 ! h264parse \
-                              ! rtph264pay config-interval=-1 name=payloader aggregate-mode=zero-latency \
-                              ! application/x-rtp,media=video,encoding-name=H264,payload=96 \
-                              ! webrtcbin. autoaudiosrc is-live=1 ! queue max-size-buffers=1 leaky=downstream \
-                              ! audioconvert ! audioresample ! opusenc ! rtpopuspay pt=97 \
-                              ! webrtcbin. ";
-
-//const TEST_PIPELINE: &str = "videotestsrc pattern=ball is-live=1 ! video/x-raw,width=320,height=240,framerate=1/1 ! jpegenc ! appsink name=appsink drop=true max-buffers=1";
-
-#[async_std::main]
-async fn main() -> anyhow::Result<()> {
-    gstreamer::init().unwrap();
-
-    let pipeline = gstreamer::parse_launch(WEBRTC_PIPELINE)
-        .unwrap()
-        .downcast::<gstreamer::Pipeline>()
-        .unwrap();
-
-    let appsink = pipeline.by_name("appsink").unwrap();
-    let appsink = appsink.dynamic_cast::<gstreamer_app::AppSink>().unwrap();
-    let image_stream = Arc::new(Mutex::new(appsink.stream()));
-    pipeline.set_state(gstreamer::State::Playing).unwrap();
-
-    let listener = TcpListener::bind("127.0.0.1:1234").await?;
[…]

### 2021-12-01 computergrafik/echtzeit-video-streaming.tex
Reformat the article and add a camera section

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xcolor}

\hypersetup{colorlinks=true, allcolors=black}

\title{Ansätze zum Echtzeit-Video-Streaming im Web}
\author{
	Maximilian Schulke \textit{(Matrikel-Nr. 20215853)}\\
	Technische Hochschule Brandenburg \\
	B.Sc. Medieninformatik \\
	Computergrafik
}

\begin{document}

\maketitle
betreut durch Prof.\ Dr.\ rer.\ nat.\ Reiner Creutzburg\\
Wintersemester 2021\\

\newpage

\tableofcontents

\begin{abstract}
	Das abstract schreibe ich zu letzt!
\end{abstract}

\newpage

\markboth{Hausarbeit Computergrafik – Maximilian Schulke}{}
\section{Einleitung / Motivation}
% \IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden Vernetzung der Arbeitswelt
in den letzten Jahren \textcolor{red}{quelle suchen corona zoom} werden auch
digitale Meeting-Systeme immer relevanter und müssen immer mehr Nutzer in
Echtzeit mit einander verbinden um einen reibungslosen Arbeitsalltag zu
gewährleisten. Dies impliziert natürlich auch dass eine \textcolor{red}{quelle
suchen ab wann gute Qualität} Verbindungsqualität gegeben sein muss, damit die
Systeme nutzbar bleiben.

\textcolor{red}{\& sicherheitsaspekt}
Aber wie können wir skalierbare Meeting-Systeme realisieren ohne große
Datenmengen über einen Streaming-Server zu schicken der diese an alle anderen
broadcasted? Die Entwicklung der letzten Jahre deuten immer mehr darauf hin,
dass \textit{Peer-To-Peer} basierte Lösungensansätze aufgrund der besseren
Performance und Skalierbarkeit, in der Regel die bessere Wahl darstellen
\textcolor{red}{quelle suchen}. Natürlich spielen zur Auswahl der Architektur
noch weitere Parameter eine wichtige Rolle (z. B. die maximale Bandbreite und
Rechenleistung der Endgeräte), aber mit immer großer werdenden
Heimnetz-Leitungen und zunehmender Rechenleistung der Endgeräte stellt dies
meistens kein Problem mehr da. \textcolor{red}{quelle suchen}.

Die Problematik der Echtzeit-Kommunikation im Web beschäftigt auch
das \textit{W3C} seit 2011 im Zuge der Standardisierung des seit diesem Jahr
zum Web-Standard erklärten Protokoll \textit{WebRTC}. \textcolor{red}{quelle
suchen}.

Es ist also (immer noch) eine sehr aktuelle Thematik in der Informatik
Echtzeit- oder \textcolor{blue}{Nahe-Zu-Echtzeit-Kommunikation} zuverlässig zu
bewältigen. Die Aufgabe dieser Arbeit soll sein, einen Überblick über den
Stand der Architekturmuster, Protokolle und möglicher Problematiken bei der
Implementierung von eigenen Echtzeit-Video-Streaming-Diensten geben,
\textcolor{orange}{diese anhand eines Experiments durchsprechen und auswerten
usw.}

\section{Historie der Echzeit-Übertragung}

Um zu verstehen wie sich die Protokolle hin zum heutigen Stand entwickelt
haben, ist es besonders interessant nachzuvollziehen wie die ersten Schritte
der IEFT oder auch \textit{Internet Engineering Task Force} bezüglich der
Echtzeit-Kommunikation aussahen, welche Probleme erkannt und behoben wurden und
welche Protokolle heute der Standard sind.

\subsection{1996: RTP \& RTCP – RFC 1889}

Am weitesten reicht das \textit{Real-Time Transport Protocol} oder auch
\textit{RTP} zurück. Es wurde erstmals 1996 von der IEFT standardisiert und
stellt seit dem einen Grundbaustein der datenformatsaganostischen
Echtzeit-Übertragung da. Es kann für diverse
Echtzeit-Übertragungs-Problematiken dienlich sein, da jegliche Binärdaten
verschickt werden können; Somit gibt es keinen ``Lock-In'' auf bestimmte Audio-
oder Video-Codecs. \textcolor{orange}{aussage prüfen und quelle}.

\textit{RTCP} ist das mit \textit{RTP} einhergehende Kontrollprotokoll. Es
wird primär dazu verwendet, die Übertragungsparameter der Sender zu
beeinflussen – z. B. durch ein Feedback zur Übertragungsqualität oder ein
Abmelden der Session. Desweiteren bietet es eine persistente ID für die
\textit{RTP}-Mitglieder, die über Programm-Neustarts hinweg zur Identifikation
von Mitgliedern und der zuordnung von Datenströmen verwendet werden können.
\textcolor{red}{cite rfc1889 kapitel 6 / 6.1 bzw. S. 15-17}

Das Protokoll siedelt sich im TCP/IP-Stack über \textit{UDP} an
\textcolor{red}{cite rfc1889 introduction}. Es fügt wichtige Informationen zu
UDP-Datagrammen hinzu: Im wesentlichen eine Sequenznummer um die
Sende-Reihenfolge zu codieren und einen Payload-Type, der den Codec des
Segments angibt. Somit kann auch bei nicht sequenziell übertragenden
Datagrammen die Ursprüngliche Reihenfolge rekonstruiert werden und es können
bei verlorenen Segmenten Interpolationsalgorithmen verwendet werden.
\textcolor{red}{quelle / beispiele suchen}. Der Payload-Type ist essenziell um
ohne Session-Aushandlung zu kommunizieren wie der Empfänger die Daten zu
decodieren hat, um eine Sinnvolle nachricht zu erhalten.

In der ersten Version aus 1996 gab es einige Probleme bezüglich
\textcolor{red}{Probleme suchen}. Diese wurden 2003 in dem RFC3550
überarbeitet - somit wurde das RFC1889 durch die neuere Version 3550 obsolet.
In der aktuelleren Version wurden einige Änderungen eingearbeit: Im
wesentlichen ``RTCP Packet Send and Receive Rules'' ``Layered Encodings''
``Congestion Control'' ``Security Considerations'' ``IANA Considerations''
\textcolor{orange}{Dummer text}.

Eine wichtige Voraussetzung zur Verwendung von \textit{RTP} ist ein externer
\textit{Signaling-Server}, den alle Beteiligten zur Session-Aushandlung
verwenden. Ein Standard hierfür ist im \textit{RTP-Framework} selbst nicht
definiert, eine beliebte Wahl für ein Protokoll zur Session-Aushandlung ist
allerdings das \textit{Session Initiation Protokoll} (oder kurz \textit(SIP)).
% https://datatracker.ietf.org/doc/html/rfc3550#section-3

\subsection{2004:\ SRTP (RFC 3711)}

Im März 2004 wurden \textit{SRTP} und \textit{SRTCP} vorgestellt – die
verschlüsselten Version von \textit{RTP} bzw. \textit{RTCP}. Diese bauen
weitestgehend auf dem \textit{Advanced Encryption Standard} (kurz.
\textit{AES}) auf \textcolor{red}{(siehe RFC3711 Seite 19).}. Der RFC
beschäftigt sich weitestgehend mit den kryptografischen Aspekten des
Protokolls.

Eine weitverbreitete und offene Implementierung für \textit{SRTP / SRTCP}
wird von der US-Amerikanischen Telekomunikations-Gesellschaft Cisco
bereitgestellt. Diese hat den Namen \textit{libsrtp} und ist öffentlich auf der
Entwickler-Plattform GitHub einsehbar (siehe
\href{https://github.com/cisco/libsrtp}{github.com/cisco/libsrtp}).

% Anwendung in VoIP\!
% https://de.wikipedia.org/wiki/ZRTP

\subsection{2005: RTMP von Adobe}

% https://www.adobe.com/content/dam/acom/en/devnet/rtmp/pdf/rtmp_specification_1.0.pdf
\textit{RTMP} ist ein von der Firma \textit{Adobe Systems Incorporated}
spezifiziertes Protokoll zur Echtzeit-Übertragung von Multimedia-Streams.
\textit{RTMP} wurde laut Spezifikation\textcolor{red}{siehe RTMP spec 1.0},
entwickelt um über dem Transport-Protokoll \textit{TCP} verwendet zu werden.

Das Protokoll wurde dazu entwickelt um im Kontext eines Flash-Players verwendet
zu werden. Dies führt dazu, dass es heutzutage nurnoch bedingt Anwendung
findet, da mittlerweile viele Browser ihren Flash-Player-Support eingestellt
haben (\textcolor{red}{siehe Chrome und Firefox}) % https://support.mozilla.org/en-US/kb/end-support-adobe-flash

Außerdem spricht die hohe Latenz von bis zu 30 Sekunden, die durch die
Verwendung von \textit{TCP} zustandekommt (\textcolor{red}{siehe
restram.io/streaming-protocls}), gegen den Einsatz von \textit{RTMP} in einem
Echtzeit-Übertragungs-Kontext.

Desweiteren gibt es noch Protokollvarianten die HTTP bzw. HTTPS als zugrundeliegendes
Protokoll verwenden.\textcolor{red}{Siehe XYZ Quelle suchen}

\subsection{2010: WebRTC (RFC 8825)}

\textit{WebRTC} ist eine Peer-To-Peer-Technologie die über mehrere Protokolle
und Audio- und Video-Codecs performante und generische
Echtzeit-Kommunikations-Kanäle zwischen Nutzern (oder auch \textit{Peers})
realisiert. Sie

WebRTC stellt eine direkte Verbindung zwischen zwei Endgeräten her und
verwendet einen mix aus RTP, RTCP, SDP, (ICE Candidates) und einem signaling
server. Es unterstützt beliebte video codecs wie V8 / V9 und ist in der zukunft 
auf AV1 angelegt. Der standard audio codec ist opus.

Durch die direkte verbindung wird eine sogennante sub second latency erreicht.
Dies beschreibt im grunde \textcolor{red}{XYZ}.

% Hat mit 1996 RTP und dann 2005 mit RTMP angefangen, wann kam SRTP?\ JSEP Global IP Solutions \& Google WebRTC kam 2010, Microsoft wollte CU-RTC

% > Datenströme sind verpflichtend zu verschlüsseln. Dazu werden Verbindungen über DTLS verschlüsselt und Audio- und Videokommunikation zusätzlich durch SRTP geschützt.[16]
% https://datatracker.ietf.org/doc/html/rfc8834#section-4.1

% https://www.w3.org/2021/01/pressrelease-webrtc-rec.html.en

Die ersten \textit{Requests for Comments} oder auch \textit{RFC} zu
den grundlegenden Protokollen \textit{RTP} und
\textit{\textcolor{blue}{Protokoll X}}, auf denen die heutigen Protokolle
weitestgehend aufbauen, wurden bereits in den späten 1990er Jahren
veröffentlicht und seit dem immer weiterentwickelt. \textcolor{red}{quelle
anhängen RFC35XX}.

\section{Architekturmuster}

In diesem Kapitel werden zwei der bekanntesten Architekturmuster in der
Echtzeit-Übertragung vorgestellt und verglichen. Es geht primär um die
verteilte Peer-To-Peer Architektur und die zentralisierte Relay / Broadcast
Architektur.

\subsection{Peer-To-Peer}

Auf Grund der hohen Anforderungen an möglichst niedrige Übertragungslatenzen
bietet eine Peer-To-Peer-Architektur klare Vorteile durch den stark verkürzten
Weg, den die Pakete zurücklegen müssen bis sie bei dem Empfänger ankommen.

Deutlich komplizierter wird allerdings die aushandlung bzw.\ initialisierung
einer Verbindung zwischen zwei peers, da diese sich nicht wie bei der typischen
client-server-architektur eine fixe adresse zur verbindung haben. Dieses
problem wird typischerweise über einen signaling server gelöst.

\subsubsection{Signaling Server}
Ein signaling server ist allen peers bekannt und dient als
kommunikationsplattform, damit sich die beiden (sich gegenseitig initial
unbekannten) peers gegenseitig vorstellen können.

Signaling server sind interessanterweise keine feste anforderung für peer to
peer muster. Wenn alle peers immer statische addressen hätten, könnten sie auch
offline ihre ips / ice candidates / sdp offers etc.\ austauschen und so eine
session aufbauen. Wichtig ist leidlgich eine initiale out of band
kommunikation.

Der signaling server kann außerdem noch nach verbindungsaufbau dazu verwendet
werden um die bestehende verbindung zu optimieren. Peers können neue ICE
candidates vorschlagen um die Übertragung an neue gegebenheiten im internet
(z.B. ausfall eines routers) anzupassen. (siehe mdn)

% https://datatracker.ietf.org/wg/wish/about/

\subsubsection{Holepunching}
Holepunching beschreibt den prozess lokale firewalls und nats zu durchbrechen
um eine direkte verbindung zwischen zwei endgeräten herzustellen.

% https://datatracker.ietf.org/doc/html/rfc8834#section-10
% https://de.wikipedia.org/wiki/Multicast
% peer to peer wäre möglich mit ip multicast mit vielen usern ohne pakete zu
% duplizieren

\subsubsection{IP-Multicast}

Je mehr Nutzer / Endgeräte an einer peer-to-peer-übertragung teilnehmen, desto
größer wird die Belastung der Bandbreite bei den einzelnen Teilnehmern. Jedes
Paket muss nicht einmal, sondern n-mal verschickt werden (wobei n die anzahl
der teilnehmer ist). Dies kann bei labilen oder einfach schwachen Netzwerken
entweder zur verminderung der übertragungsqualität führen, oder in besonders
schlimmen fällen sogar das restliche netzwerk eines einzelnen Teilnehmers
negativ beeinflussen, da dieser die meiste Bandbreite für die wiederholte
übertragung gleicher pakete verwendet.

Eine theoretische Lösung für dieses Problem, ist die Verwendung von
IP-Multicast-Adressen. Diese erlauben es einem Gerät ein IP-Paket einmalig zu
übertragen, und dieses von Multicast-Routern im internet multiplizieren zu
lassen. (Siehe grafik)

Vergleichs grafik einfügen..

Dieser ansatz hat klare vorteile: Das gesamte - lokale \& globale – netzwerk
wird geschont. So würde die Anzahl möglicher Teilnehmer deutlich steigen, da
ein Netzwerk-Bottleneck erst bei einem zu großen Download-Volumen des
empfangenen Contents eintreten würde.

\subsection{Relay}
% https://webrtc.org/getting-started/turn-server

\section{Protokolle}
\subsection{RTP}

\subsection{RTCP}
\subsection{RTSP}
\subsection{SDP}
% https://datatracker.ietf.org/doc/html/rfc4566#section-3.1
\subsection{SID}
% https://datatracker.ietf.org/doc/html/rfc3261
% \subsection{STUN}
\subsection{WebRTC}

\section{Implementierung eines Kamera-Live-Streams}

In diesem Kapitel wird ein experimentelles Kamera-System entwickelt, das die
oben erläuterten Protokolle / Technologien verwendet.

\subsection{Ziel}
Ziel dieses Experiments ist es eine möglichst einfach ein funktionierendes
System zu entwickeln und mögliche Probleme, benötigten Zeitaufwand usw.\ zu
ermitteln.  Desweiteren ist ein ``Performance'' Vergleich mit anderen
Live-Stream systemen interessant. Es gilt also die frage zu klären, wie viel
arbeit benötigt ein grundsätzlich funktionierendes Software-Produkt das auf
Echtzeit-Übertragung beruht.

\subsection{Umfang}

Das zu implementierende Kamera-System wird lediglich eine einzige Kernfunktion
aufweisen: Sobald die Kamera an ist, streamt sie via WebRTC, mit ausreichend
FPS (mindestens 15 im durchschnitt) in ausreichender Qualität (720 x\ 480),
ihre Aufnahmen an ein User Interface. Dieses wird im browser laufen, muss aber
ausreichend Alternativen für andere Plattformen aufweisen (Nativ, Smartphone
usw.). Das Interface und die Kamera starten die Aushandlung des
WebRTC-Streams über den Signaling Server. D.h.\ sie versenden SDP (Offer /
Answer) und tauschen ihre ICE-Candidates aus.

\subsection{Architektur}

Die Architektur ist sehr simpel. Sie besteht lediglich aus 3 Kernelementen:

\subsubsection{Kamera}

Die Hauptanforderung an die Hardware auf der die Kamera-Software läuft ist eine
Linux-Installation (mit den entsprechenden installierten Paketen für die
Abhangigkeiten) und eine angeschlossene Kamera die von Linux erkannt wird.

Für dieses Experiment wurde ein Einplatinencomputer der Marke Raspberry PI in
der 4. Version mit 8GB RAM – erweitert durch ein Drittanbieter-Kamera-Modul –
verwendet. Dieser ist allerdings absolut austauschbar, da jeder Rechner der
die o.g. Hauptanforderung erfüllt eine geeignete Umgebung darstellt – so kann
die Kamera-Software auch auf einem Laptop ausgeführt werden, falls keine
Hardware verfügbar ist. Dies wurde zur veranschaulichung mit einem ThinkPad
T490 und einer aktuellen NixOS Installation getestet.

Auf der Einplatinencomputer ist das mitgelieferte Betriebssystem ``Raspberry PI
OS'' und die entsprechenden Software-Abhängigkeiten installiert. (Siehe GitHub
für Dependencies.)

\subsubsection{Signaling-Server}

Der Signaling-Server ist ein WebSocket-Secure-Server, der zwei
WebSocket-Verbindungen miteinander verknüpft. Er verwaltet sogenannte Räume.
Ein Raum besteht aus 0 bis 2 Clients und dient dazu diese miteinander zu
verbinden. Auf dem Server kann es mehrere Räume gleichzeitig geben – dadurch
könnte dieser Signaling-Server theoretisch auch noch für weitere
WebRTC-Anwendungen verwendet werden. Ein Raum hat eine ID, diese ist eine
256-Bit-Entropie. Clients werden durch eine 128-Bit-Entropie identifiziert.

So können sich zwei Clients durch ein Out-Of-Band kommuniziertes Secret (die
Raum-ID) über diesen in Kontakt treten. Diese könnte beispielsweise, würde es
sich bei der entwickelten Kamera um ein echtes Produkt handeln, bei der
Herstellung generiert werden, ausgedruckt und neben die Kamera in die
Verpackung gelegt werden.

\subsubsection{Interface}

Das Interface verbindet sich mit dem Signaling Server und nimmt eine Raum-ID
entgegen. Mit dieser Raum-ID wird dann auf dem Signaling Server entweder ein
neuer Raum erstellt, falls noch keiner mit der ID existiert, oder es wird dem
bestehenden Raum beigetreten. Nach dem ein weiterer User beigetreten ist, fängt
der in \textcolor{red}{Kapitel XY} erklärte Aufbau eines WebRTC-Streams zu dem
Nutzer an.

\subsection{Signaling Server}

Als erstes wurde der Signaling Server entwickelt, da dieser keine
Abhängigkeiten an seine Clients hat. Die Kamera-Software und Interface setzen
beide jeweils den laufenden Signaling Server vorraus um korrekt zu
funktioneren.

\subsubsection*{Asynchronität}

Eine logische Anforderungen an den Server ist das gleichzeitige verarbeiten
mehrerer Verbindungen – ansonsten könnte immer nur ein Client alleine mit dem
Server Verbunden sein. Dies würde die Signaling-Funktionalität eines Raumes
unbrauchbar machen.

Die asynchrone Programmierung ist ein Konzept zur Lösung dieser Problemklasse.
Da langlebige Verbindungen in der Regel einen großteil der Zeit ungenutzt sind,
ist es naheliegend Verbindungen ohne neue Ereignisse keine CPU-Zeit zu geben
um andere Verbindungen in der Zeit abzuarbeiten, in der auf Ereignisse gewartet
wird.

HIER ZIETIEREN WAS DAS ZEUG HÄLT / ERKLÄRGRAFIKEN NUTZEN

\subsubsection{Verbindungsaufbau}

Um eine WebSocket-Secure-Verbindung aufzubauen, muss zu erst eine TCP- und dann
darüber eine TLS-Verbindung zu dem Client aufgebaut werden. Über diese wird
dann zu erst in HTTP kommuniziert (siehe WebSocket verbindungsaufbau Quelle
suchen) und nach einer erfolgreichen Nachricht des Servers mit dem Status-Code
101 (Switching Protocols) wird die über den gleichen Transport-Weg (TLS)
nach dem WebSocket-Protokoll kommuniziert.

Grafik einfügen.

\subsubsection{Spezifikation des Signaling-Protokolls}

Das Signaling-Protokoll ist in 2 Nachrichten-Typen unterteilt:
Server-Nachrichten und Peer-Nachrichten. Clients dürfen nur Peer-Nachrichten
senden, ansonsten wird die Verbindung aufgrund eines Protokoll-Verstoßes
geschlossen. Der Server verschickt nur Server-Nachrichten.

Alle Protokoll-Nachrichten werden in JSON kodiert und dann über den
WebSocket-Nachrichten-Typ Binär an den Empfänger geschickt werden.

\vspace{0.5em}

\textbf{Server-Nachrichten}

\begin{itemize}
	\item Hello \{Client-ID\}
	\item Joined
	\item Error
	\item Room/Join \{Client-ID\}
	\item Room/Leave \{Client-ID\}
	\item Room/Signal \{Signal\}
\end{itemize}

\vspace{0.5em}

\textbf{Peer-Nachrichten}

\begin{itemize}
	\item JoinOrCreate \{Raum-ID\}
	\item Signal \{Signal\}
\end{itemize}

\vspace{0.5em}

Nach einer erfolgreich aufgebauten Verbindung mit dem Server schickt dieser ein
\textit{Hello} mit der zugewiesenen Client-ID.\ Darauf hin muss der Client ein
\textit{JoinOrCreate} senden um einem Raum beizutreten. Bevor der Client die
Nachricht \textit{Joined} vom Server erhält, darf dieser keine Signale
verschicken.

Diagramme einfügen!

Eine \textit{Room/Join} Nachricht wird an ggf.\ andere Clients im Raum
verschickt, sobald ein Client diesen betritt. Diese Nachricht kann auf dem
Client als anlass genutzt werden eine SDP-Offer zu erstellen, da nun sicher
ist, das ein anderer Client im Raum ist. Wenn beide Clients sich an dieses
Schema halten, schickt immer der Client, der sich zu erst Verbunden hat die
Einladung und der andere die Answer.

Analog wird eine \textit{Room/Leave} Nachricht verschickt wenn ein Client einen
Raum verlässt.

Die \textit{Room/Signal} Nachricht wird an den Empfänger eines Signals
geschickt, nachdem der Sender des Signals eine \textit{Signal} Nachricht an den
Server geschickt hat.

\subsubsection{Raum-Verwaltung}

Die Raum-Verwaltung fällt unter anderem in die Kategorie der
Synchronisationsprobleme. Es gibt unter Umständen $n$ Client-Verbindungen, die
den Status von einem Raum erfragen wollen, und ggf.\ einen Anlegen möchten. Dies
soll für alle Verbindungen öffentlich geschehen, Clients dürfen sich aber dabei
nicht gegenseitig überschreiben.

% Mutual exclusion -> gegenseitiger ausschluss -> nur einer darf

Um zwei asynchron laufende Programmteile zum sequenziellen Zugriff auf
gemeinsamen Speicher zu zwingen, gibt es Möglichkeit einen Semaphor bzw. Mutex
zu verwenden. Dieser blockiert den Teil des Programms, der gerade auf den
gemeinsamen Speicher zugreifen möchte solange, bis ein ggf.\ anderer Zugriff
beendet ist. Um Deadlocks bzw. Inperformante Code-Abschnitte zu vermeiden, ist
es ratsam einen Mutex nicht über länger andauernde Operationen hinweg zu
locken, da sonst für diese Zeit alle anderen Teile des Programms, die gerade
auf den Speicher zugreifen möchten blockiert sind.

Als konkrete Lösung für das bestehende Problem bietet sich eine HashMap, deren
zugriff von einem Mutex kontrolliert wird, an. Jede Client-Verbindung, die
aktuelle Raum-Informationen braucht (was hauptsächlich bei dem erstellen /
betreten von Räumen der Fall ist), muss nun vorher erst den mutex locken.

\subsubsection{Kommunikation zwischen mehreren Client-Verbindungen}

Bis zu diesem Punkt ist der Aufbau der Verbindung und das zu implementierende
Protokoll klar, allerdings fehlt noch das Verknüpfen von Client-Verbindungen um
Signale weiterzuleiten.

Da bereits Räume verwaltet werden, liegt es nahe dort einen Kommunikationskanal
einzubetten. Eine Implementierung für diesen Kommunikationskanal stellen z.B.
sogenannte Channels dar. Diese erlauben Inter-Thread- (und im asynchronen
Kontext: Inter-Task-) Kommunikation über ein einfaches
Sender-Empfänger-Prinzip. Ein Channel ist unterteilt in eben diese beiden
Teile: Der Sender darf Nachrichten schreiben, der Empfänger darf sie aus dem
Channel lesen.

In diesem Fall ist allerdings zusätzlich noch bidirektionale Kommunikation
gefragt, da beide Clients Signals des jeweils anderen erhalten sollen. Dafür
bieten sich sogenannte Broadcast-Channel an. Sie sind dafür ausgelegt, dass von
mehreren Stellen in diese geschrieben und gelesen wird. Sie funktioneren analog
zu dem Broadcast aus der Netzwerktechnik.

Diagramm einfügen..

%      C
%    s r r
%  s   r   r
% o    o     o

%      C
%    r s r
%  r   s   r
% o    o     o

%      C
%    r r s
%  r   r   s
% o    o     o

Somit kriegt jeder Client bei betreten eines Raumes Zugriff auf einen Sender
und einen Empfänger dieses Broadcast-Channels. In diesen schreibt er bei erhalt
einer Peer-Nachricht vom Typ \textit{Signal} das Signal. Eine Client-Verbindung
die gerade keine CPU-Zeit erhält, da ansonsten keine Ereignisse aufgetreten
sind, wird wieder aktiv sobald eine neue Nachricht über den Channel da ist.
Diese kann dann als Server-Nachricht vom Typ \textit{Room/Signal} mit dem
Signal aus dem Channel an den Empfänger gesendet werden.

\subsection{Interface}

Um den Signaling-Server für ein erstes nutzbares Zwischenergebnis zu verwenden,
wurde eine einfache browserbasierte Benutzer-Oberfläche implementiert, die
analog zu sehr bekannten Videokonferenz-Systemen wie z.B. Zoom oder Google Meet
funktioniert. Bei diesem Experiment wurde der Fokus primär auf die
Bildübertragung gelegt – eine Audioübertragung könnte aber ohne großen Aufwand
implementiert werden.

Ein Benutzer muss über dieses Interface einem beliebigen Raum auf dem
Signaling-Server beitreten können. Sobald ein weiterer Benutzer beitritt soll
der erste eine Nachricht erhalten, dass Jemand seinem Raum beigetreten ist.
Letztendlich soll die Aushandlung der Video-Übertragung zwischen den beiden
Benutzern ohne weitere Interaktion des Benutzers stattfinden.

\subsubsection{WebSocket-Verbindung}

Als Vorkehrungen für die singalisierungs Kommunikation mit dem Signaling-Server
muss eine WebSocket-Secure Verbindung mit dem Server aufgebaut werden. Dies
gelingt ebenfalls über eine Web-API:\
https://developer.mozilla.org/de/docs/Web/API/WebSocket

Nun meldet sich der Server nach dem Verbindungsaufbau als erstes mit der
\textit{Hello}-Nachricht und weist uns eine ID zu. Nach dieser ersten Nachricht
können wir die eigentliche Anwendung starten.

\subsubsection{Erstellen / Beitritt eines Raumes}

Um festzustellen mit wem sich der Benutzer verbinden möchte, muss dieser
eine Raum-Id angeben. Diese ist als 64 Zeichen langen Hex-String einzugeben.

Nach erfolgreicher Eingabe einer Raum-Id wird diese mit der Nachricht
\textit{JoinOrCreate} an den Server geschickt, dieser verarbeitet die Anfrage
und schickt dem Benutzer eine \textit{Joined}-Nachricht. Ab diesem Zeitpunkt
wird eine \textit{RTCPeerConnection} erstellt, da ggf.\ bereits ein anderer
Nutzer im Raum war, der dem neu dazugekommenen eine SDP-Offer schicken wird.
Um diese Zeitnah zu verarbeiten, findet die Initialisierung vor der ersten
Signalisierungs-Nachricht statt.

% https://webrtc.org/getting-started/media-devices

\subsubsection{Verbindungsaufbau}
\label{browser-webrtc}

Moderne Browser wie Firefox, Chrome, Safari etc.\ stellen eine Implementierung
des WebRTC-Stacks bereit, die durch eine übersichtliche API leicht zu
integrieren ist. % Siehe https://caniuse.com/?search=webrtc

Um eine vollständige WebRTC-Verbindung aufzubauen ist bei beiden Nutzern eine
RTCPeerConnection aufzubauen. Auf einer RTCPeerConnection ist eine
Event-basierte API verfügbar. (Siehe tabelle)

% table mit der API und der quelle der API:

Insbesondere sind für den Verbindungsaufbau die Events onnegotiationneeded und
onicecandidate von Interesse. Sobald eine Verbindung erstellt wurde und ein
lokaler Stream zu dieser hinzugefügt wurde, wird das onnegotiationneeded Event
ausgelöst. An diesem Punkt ist nun eine eigene SDP-Offer zu erstellen und über
den singalisierungs Server an den Empfänger zu versenden. Der Empfänger
verarbeitet diese anschließend und antwortet mit einer SDP-Answer. Nach dem
beide Teilnehmer die SDP-Nachrichten füe ihren Kommunikationspartner und sich
selbst gesetzt haben (siehe setRemoteDescription und setLocalDescription),
werden vom Browser onicecandidate-Events ausgelöst um eine direkte Verbindung
zwischen den beiden Browsern auszuhandeln.

% Signalisierungsablauf zwischen Interface und Kamera visualisiert:

Nach einem leeren ICE-Kandidaten ist die aushandlung abgeschlossen und der
Stream kann nun konsumiert werden.

\subsubsection{Verbindungsabbau}

Nach dem der Signalisierungs-Server eine \textit{Room/Leave}-Nachricht gesendet
hat, kann die RTCPeerConnection ohne negative Auswirkungen abgebaut werden. Im
Interface wird die Anzeige des Video-Streams des anderen Teilnehmers
zurückgesetzt und die RTCPeerConnection geschlossen. Somit ist das Interface
wieder bereit einem weiteren Raum beizutreten und erneut eine WebRTC Verbindung
herzustellen.

% Screenshots des Demos Interfaces anfügen

\subsection{Kamera}

Zu letzt ist die Kamera zu entwickeln..

\subsubsection{GStreamer}

Anders als in einer Browser-Umgebung bietet Linux abgesehen von
Hardware-Treibern keine wirklichen Abstraktionen für die Verwendung einer
Kamera und automatischer Bildverarbeitung. Dies ist in einer nativen Umgebung
die Aufgabe von weiteren Programmen und Bibliotheken.

Hier wird das weitverbreitete Multimedia-Framework GStreamer relevant. Dieses
abstrahiert sämtliche Zugriffe auf medienbezogene Hardware (in diesem Fall die
Kamera), beherscht die meistgenutzten Media-Codecs für Audio und Video und kann
über Plugins mit weiteren Funktionalitäten erweitert werden: Wie zum Beispiel
WebRTC.\

Somit muss die folgende Liste an Abhängigkeiten installiert werden, damit die
Kamera-Software korrekt funktioniert:

\begin{itemize}
	\item libgstreamer1.0dev oder so..
	\item Paketnamen raussuchen!!!
\end{itemize}

(Aufgelistet in Debian-Paket-Namen, da Raspberry PI Os Debian basiert ist)

GStreamer ist um das Konzept von Pipelines herum aufgebaut: Es

\subsubsection{Architektur der Kamera-Software}

Die Kamera-Software ist grundsätzlich in 2 Schritte unterteilt, die endlos
wiederholt werden:

\begin{enumerate}
	\item Verbindungsaufbau zum Signalisierungsserver
	\item Verbindungsaufbau via WebRTC
\end{enumerate}

Im Regelfall dauert der 1. Schritt einige Millisekunden. Nach dem die
Verbindung zum Signalisierungsserver korrekt aufgebaut wurde (und die Kamera
dem festgelegten Raum beitreten konnte) wird eine GStreamer-Pipeline
initialisiert und auf den Beitritt eines Benutzers in den Raum erwartet. Nach
dem die \textit{Room/Join}-Nachricht eingetroffen wird die Pipeline auf Playing
gesetzt.

GStreamer unterstützt nach erfolgreicher Installation oben genannter Plugins
eine kompatible API zu der der Browser. Somit läuft der gesamte
WebRTC-Spezifische Verbindungsaufbau nach der exakt gleichen Logik ab wie unter
\nameref{browser-webrtc} erläutert.

\subsubsection{}

\subsubsection{Installation des Betriebssystems}

\subsubsection{Automatischer Start der Software}

\section{Benchmarks}

- CPU Load
- Netzwerkauslastung

\section{Auswertung}

- Raspberry Pi Kamera langsam..\ unerwünschte Qualität (evtl 32 bit?)

\section{Nächste Schritte}

\large{IPV6 only => NAT holepunching entfällt}

\section{Abbildungsverzeichnis}
abc
def

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,references}

\newpage

\begin{appendix}

foo

\end{appendix}

\end{document}

### 2021-12-04 computergrafik/echtzeit-video-streaming.tex
Add figures

@@
-\documentclass{article}
+\documentclass{IEEEtran}
 
-\usepackage[a4paper]{geometry}
+\usepackage[a4paper, top=3cm, left=2cm, right=2cm, bottom=2cm]{geometry}
 \usepackage[ngerman]{babel}
@@
 \usepackage{hyperref}
+\usepackage{graphicx}
 \usepackage{xcolor}
+
+\graphicspath{ {./images/} }
 
@@
 \title{Ansätze zum Echtzeit-Video-Streaming im Web}
+
 \author{
-	Maximilian Schulke \textit{(Matrikel-Nr. 20215853)}\\
+	\IEEEauthorblockN{Maximilian Schulke \textit{(Matrikel-Nr. 20215853)}}\\
+	\IEEEauthorblockA{
 		Technische Hochschule Brandenburg \\
@@
 	}
+}
 
 \begin{document}
+\begin{onecolumn}
+
+\IEEEspecialpapernotice{
+	betreut durch Prof.\ Dr.\ rer.\ nat.\ Reiner Creutzburg\\
+	Wintersemester 2021\\
+	Abgabetermin \today
+}
 
 \maketitle
-betreut durch Prof.\ Dr.\ rer.\ nat.\ Reiner Creutzburg\\
-Wintersemester 2021\\
-
-\newpage
-
-\tableofcontents
 
@@
 
-\newpage
+\tableofcontents
 
+\listoffigures
+\listoftables
+
+\end{onecolumn}
+
+\begin{twocolumn}
 \markboth{Hausarbeit Computergrafik – Maximilian Schulke}{}
+
 \section{Einleitung / Motivation}
-% \IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden Vernetzung der Arbeitswelt
+
+\IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden Vernetzung der Arbeitswelt
 in den letzten Jahren \textcolor{red}{quelle suchen corona zoom} werden auch
@@
 FPS (mindestens 15 im durchschnitt) in ausreichender Qualität (720 x\ 480),
-ihre Aufnahmen an ein User Interface. Dieses wird im browser laufen, muss aber
+ihre Aufnahmen an ein User Interface. Dieses wird im Browser laufen, muss aber
 ausreichend Alternativen für andere Plattformen aufweisen (Nativ, Smartphone
@@
 
-Die Architektur ist sehr simpel. Sie besteht lediglich aus 3 Kernelementen:
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{diagram-signaling}
+	\centering
+	\caption{Signaling zwischen der Kamera und dem Interface (Angelehnt an IEEE
+	PAPER!!!!!!!!!!!)}
+\end{figure}
+
+Die angedachte Architektur wurde möglichst einfach gehalten. Sie besteht
+lediglich aus 3 Elementen:
 
 \subsubsection{Kamera}
+
+Für dieses Experiment wurde ein Einplatinencomputer der Marke Raspberry PI in
+der 4. Version mit 8GB RAM – erweitert durch ein Drittanbieter-Kamera-Modul –
+verwendet. Dieser ist allerdings absolut austauschbar, da jeder Rechner der
+die u.g. Hauptanforderung erfüllt eine geeignete Umgebung darstellt – so kann
+die Kamera-Software auch auf einem Laptop ausgeführt werden, falls keine
+Hardware verfügbar ist. Dies wurde zur veranschaulichung mit einem ThinkPad
+T490 und einer aktuellen NixOS Installation getestet.
 
@@
 
-Für dieses Experiment wurde ein Einplatinencomputer der Marke Raspberry PI in
-der 4. Version mit 8GB RAM – erweitert durch ein Drittanbieter-Kamera-Modul –
-verwendet. Dieser ist allerdings absolut austauschbar, da jeder Rechner der
-die o.g. Hauptanforderung erfüllt eine geeignete Umgebung darstellt – so kann
-die Kamera-Software auch auf einem Laptop ausgeführt werden, falls keine
-Hardware verfügbar ist. Dies wurde zur veranschaulichung mit einem ThinkPad
-T490 und einer aktuellen NixOS Installation getestet.
-
 Auf der Einplatinencomputer ist das mitgelieferte Betriebssystem ``Raspberry PI
-OS'' und die entsprechenden Software-Abhängigkeiten installiert. (Siehe GitHub
-für Dependencies.)
+OS'' und die entsprechenden Software-Abhängigkeiten installiert. 
+\textcolor{red}{Siehe GitHub für Dependencies.}
 
@@
 
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{diagram-asynchronität}
+	\centering
+	\caption{Asynchrone Abarbeitung einer IO intensiven Aufgabe}
+\end{figure}
+
 \subsubsection{Verbindungsaufbau}
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{diagram-signaling-server-tls}
+	\centering
+	\caption{Signaling-Server mit mehreren TLS Verbindungen}
+\end{figure}
 
@@
 
-Grafik einfügen.
+\textcolor{red}{Grafik einfügen}
 
@@
 
-Diagramme einfügen!
+\textcolor{red}{Diagramme einfügen!}
 
@@
 
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{diagram-broadcast}
+	\centering
+	\caption{Broadcast-Channel Konzept}
+\end{figure}
+
 Somit kriegt jeder Client bei betreten eines Raumes Zugriff auf einen Sender
@@
 Signal aus dem Channel an den Empfänger gesendet werden.
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{diagram-websocket-verknüpfung}
+	\centering
+	\caption{Kommunikation von Nutzern über einen Broadcast-Channel}
+\end{figure}
 
@@
 
-\subsubsection{Verbindungsaufbau}
-\label{browser-webrtc}
+\subsubsection{Verbindungsaufbau}\label{browser-webrtc}
 
@@
 % Signalisierungsablauf zwischen Interface und Kamera visualisiert:
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{cited-webrtc-connection-establishment}
+	\centering
+	\caption{WebRTC-Verbindungsaufbau (Siehe QUELLE)}
+\end{figure}
 

### 2021-12-04 computergrafik/echtzeit-video-streaming.tex
Add a section about gstreamer

@@
 
-Anders als in einer Browser-Umgebung bietet Linux abgesehen von
-Hardware-Treibern keine wirklichen Abstraktionen für die Verwendung einer
-Kamera und automatischer Bildverarbeitung. Dies ist in einer nativen Umgebung
-die Aufgabe von weiteren Programmen und Bibliotheken.
+Anders als in einer Browser-Umgebung bietet Linux abgesehen von der
+bereitstellung von Hardware-Treibern keine guten Abstraktionen für die
+Verwendung einer Kamera und automatischer Bildverarbeitung. Dies ist in einer
+nativen Umgebung die Aufgabe von weiteren Programmen und Bibliotheken.
 
@@
 Kamera), beherscht die meistgenutzten Media-Codecs für Audio und Video und kann
-über Plugins mit weiteren Funktionalitäten erweitert werden: Wie zum Beispiel
-WebRTC.\
+über Plugins mit weiteren Funktionalitäten erweitert werden.
+
+GStreamer ist dafür ausgelegt das erstellen von Audio und/oder Video basierten
+Anwendung zu erleichtern. Es unterstützt weiterhin, durch die
+Pipeline-Architektur, jegliche Art von Datenströmen.
+
+% Siehe https://gstreamer.freedesktop.org/documentation/application-development/introduction/gstreamer.html?
+% Siehe https://gstreamer.freedesktop.org/documentation/application-development/introduction/gstreamer.html?
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{gstreamer-overview.png}
+	\centering
+	\caption{GStreamer Architektur QUELLE von GSTREAMER DOCS}
+\end{figure}
+
+Zu den Kern-Features von GStreamer gehören eine API für diverse
+Multimedia-Anwendungen, eine Pipeline- und Plugin-Architektur, Mechanismen für
+die Aushandlung von Datenformaten zwischen Pipeline-Elementen, Synchronisation
+von Datenströmen. (Siehe Abbildung X und QUELLE!!)
+
+Das Framework baut auf einem ausgereiften Plugin-System mit über 250 Plugins
+auf, welches den Codec-Support erweitert und weitere Komponenten zur Erstellung
+von komplexeren Pipelines bereitstellt. (Siehe QUELLE)
+
+Diese Plugins können in die folgenden Kategorien unterteilt werden:
+
+\begin{itemize}
+	\item Protokolle
+	\item Datenquellen
+	\item Formate
+	\item Codecs
+	\item Filter \& Effekte
+	\item Ausgabe
+\end{itemize}
+
+% Siehe https://gstreamer.freedesktop.org/documentation/application-development/introduction/gstreamer.html?
+
+Für dieses Experiment ist das GStreamer-Plugin für WebRTC besonders von
+bedeutung.
 
@@
 
-GStreamer ist um das Konzept von Pipelines herum aufgebaut: Es
-
 \subsubsection{Architektur der Kamera-Software}
 
-Die Kamera-Software ist grundsätzlich in 2 Schritte unterteilt, die endlos
-wiederholt werden:
+Die Kamera-Software ist grundsätzlich in 2 Zustände unterteilt, die sich endlos
+abwechseln:
 
@@
 	\item Verbindungsaufbau zum Signalisierungsserver
-	\item Verbindungsaufbau via WebRTC
+	\item Streaming via WebRTC
 \end{enumerate}
@@
 eine kompatible API zu der der Browser. Somit läuft der gesamte
-WebRTC-Spezifische Verbindungsaufbau nach der exakt gleichen Logik ab wie unter
-\nameref{browser-webrtc} erläutert.
+WebRTC-Spezifische Verbindungsaufbau nach der exakt gleichen Logik ab wie
+unter\ \ref{browser-webrtc} erläutert.
 
-\subsubsection{}
+\subsubsection{Vorbereitung der Laufzeitumgebung}
 
-\subsubsection{Installation des Betriebssystems}
+Die Installation des Betriebssystems für die Kamera-Hardware
 
-\subsubsection{Automatischer Start der Software}
+\subsubsection{Automatischer Start der Kamera-Software}
 

### 2021-12-04 computergrafik/echtzeit-video-streaming.tex
Start the conclousion section

@@
 
-- Raspberry Pi Kamera langsam..\ unerwünschte Qualität (evtl 32 bit?)
+Der Signaling-Server hatte den größten Implementierungsaufwand der 3
+Komponenten, hat sich dafür aber als sehr stabil und vielseitig einsetzbar
+erwiesen. Hinsichtlich der Performance ist die Implementierung ebenfalls ein
+Erfolg. Auf einem System mit einem Intel i7-8565U-Chip übersteigt der Server
+mit 10 Benutzern nicht ansatzweise 1\% der CPU-Kapazität. Falls nicht eine
+große Anzahl an Nutzern gleichzeitig Signale über diesen Server schickt,
+pendelt sich dieser aufgrund seiner asynchronen Architektur im Ruhemodus bei
+einer CPU-Last von 0.0\% ein, da er nur aktiv wird, wenn auch Nachrichten
+eintreffen.
+
+Die Benutzeroberfläche ist rein funktional geblieben und bietet kein gutes
+Nutzererlebnis. Abgesehen von der schlechten Nutzbarkeit, ist die erwartete
+Funktionalität vollständig und mit angemessenem Aufwand implementierbar
+gewesen. Die Browser-APIs bieten eine exzellente Grundlage um reale
+Anwendungen in kurzer Zeit mit einer Live-Stream-Funktionalität zu entwickeln.
+Desweiteren funtkioniert auch die Videoübertragung vom Browser über eine
+WebRTC-Verbindung sehr gut, die Bildübertragungsraten waren i.d.R. über den
+angestrebten 15 Bildern pro Sekunde.
+
+Die Implementierung einer Kamera-Software ist im Vergleich zu den beiden
+anderen Komponenten verhältnismäßig aufwändig gewesen. GStreamer bringt
+ebenfalls vergleichbare WebRTC-Abstraktionen mit sich, wie moderne Browser,
+allerdings ist GStreamer erheblich instabiler und unzuverlässiger was den
+Stream-Aufbau angeht als die Browser. Auch die Bildübertragungsrate ist leider
+erst bei sehr niedriger Qualität (160px$\times$90px) nah an dem angestrebten
+Wert. Sobald die Qualität diese überschritt, nimmt die Bildübertragungsrate
+drastisch ab und somit wird der Stream ab einer gewissen Bildqualität
+unbrauchbar. Dies könnte ggf.\ auf das 32-Bit-Betriebssystem der Hardware
+zurückzuführen zu sein.
 
@@
 
-\large{IPV6 only => NAT holepunching entfällt}
+Die nächsten Schritte um dieses Experiment weiterzuführen wären definitiv die
+Fehlersuche an der Kamera-Software um die Übertragungsqualität zu steigern.
 
-\section{Abbildungsverzeichnis}
-abc
-def
+Außerdem sollte die Kamera-Software nicht erneut dem Raum auf dem
+Signaling-Server beitreten sobald der Stream abbricht, dies zu beheben könnte
+eine (auch wenn nur geringe) Netzwerkentlastung bringen und würde zusätzlich
+den Signaling-Server schonen.
 
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,references}
+Auch die Sicherheit von Räumen könnte durch ausgreiftere
+Authentifizierungsverfahren drastisch erhöht werden.
 
-\newpage
+Als letzte Verbesserung wäre eine Implementierung von Audio-Übertragung von der
+Kamera zu der Benutzeroberfläche interessant um auch diesen Aspekt der
+Echtzeitübertragung im Web auszustesten.
 
+% Mögliche Optimierungen:
+% - Verbesserung der Video-Qualität: Evtl kein 32bit os
+% - Nicht ständig signalisieren, einmal joinen als kamera
+% - Kamera interface verbessern
+% \large{IPV6 only => NAT holepunching entfällt}
+
+%\bibliographystyle{IEEEtran}
+%\bibliography{IEEEabrv,references}
+
+\end{twocolumn}
+
+\begin{onecolumn}
 \begin{appendix}
 
-foo
-
 \end{appendix}
-
+\end{onecolumn}
 \end{document}

### 2021-12-05 computergrafik/echtzeit-video-streaming.tex
Finish the introduction

@@
 
-\IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden Vernetzung der Arbeitswelt
-in den letzten Jahren \textcolor{red}{quelle suchen corona zoom} werden auch
-digitale Meeting-Systeme immer relevanter und müssen immer mehr Nutzer in
-Echtzeit mit einander verbinden um einen reibungslosen Arbeitsalltag zu
-gewährleisten. Dies impliziert natürlich auch dass eine \textcolor{red}{quelle
-suchen ab wann gute Qualität} Verbindungsqualität gegeben sein muss, damit die
-Systeme nutzbar bleiben.
+\IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden dezentralisierung und
+der damit einhergehenden Vernetzung der Arbeitswelt in den letzten
+Jahren~\cite{CoronaHomeOffice} werden auch digitale Meeting-Systeme immer
+relevanter und müssen immer mehr Nutzer in Echtzeit mit einander verbinden um
+einen reibungslosen Arbeitsalltag zu gewährleisten. Zu begin der
+Corona-Pandemie stieg der Netzwerktraffic für VoIP und Videokonferenzen im März
+2020 um 210\%-285\%~\cite{NetzwerkStatistik}.
 
-\textcolor{red}{\& sicherheitsaspekt}
-Aber wie können wir skalierbare Meeting-Systeme realisieren ohne große
-Datenmengen über einen Streaming-Server zu schicken der diese an alle anderen
-broadcasted? Die Entwicklung der letzten Jahre deuten immer mehr darauf hin,
-dass \textit{Peer-To-Peer} basierte Lösungensansätze aufgrund der besseren
-Performance und Skalierbarkeit, in der Regel die bessere Wahl darstellen
-\textcolor{red}{quelle suchen}. Natürlich spielen zur Auswahl der Architektur
-noch weitere Parameter eine wichtige Rolle (z. B. die maximale Bandbreite und
-Rechenleistung der Endgeräte), aber mit immer großer werdenden
-Heimnetz-Leitungen und zunehmender Rechenleistung der Endgeräte stellt dies
-meistens kein Problem mehr da. \textcolor{red}{quelle suchen}.
+Damit die Konferenzsysteme ihren Einsatzzweck erfüllen können, ist es
+erforderlich, dass eine ausreichend gute \textit{Quality of Service (QoS)}
+gegeben ist.
+
+Aber wie können skalierbare Meeting-Systeme realisiert werden ohne große
+Datenmengen über einen zentralisierten Streaming-Server zu schicken der diese
+an alle Teilnehmer broadcasted? Die Entwicklung der letzten Jahre deuten immer
+mehr darauf hin, dass \textit{Peer-To-Peer} basierte Lösungensansätze aufgrund
+der besseren Performance und Skalierbarkeit, in der Regel die bessere Wahl
+darstellen~\cite{PeerToPeerVersusCentralized}. \textit{Peer-To-Peer} halten
+unvorhersehbare Anstiege in der Nutzung~\cite{NetzwerkStatistik} deutlich
+besser aus, da die Echtzeit-Datenverarbeitung nicht bei dem
+Konferenzsystem-Anbieter stattfindet, sondern bei den Teilnehmern.
+
+Natürlich spielen zur Auswahl der Architektur noch weitere Parameter eine
+wichtige Rolle (z. B. die maximale Bandbreite und Rechenleistung der
+Endgeräte), aber mit immer größer werdenden Heimnetz-Leitungen und zunehmender
+Rechenleistung der Endgeräte stellt dies mittlerweile für einfache
+Videokonferenzen kein Problem mehr da.~\cite{VideokonferenzNetzwerklast}
 
@@
 das \textit{W3C} seit 2011 im Zuge der Standardisierung des seit diesem Jahr
-zum Web-Standard erklärten Protokoll \textit{WebRTC}. \textcolor{red}{quelle
-suchen}.
+zum Web-Standard erklärten Protokoll \textit{WebRTC}.~\cite{WebRTCStandardW3C}
 
 Es ist also (immer noch) eine sehr aktuelle Thematik in der Informatik
-Echtzeit- oder \textcolor{blue}{Nahe-Zu-Echtzeit-Kommunikation} zuverlässig zu
-bewältigen. Die Aufgabe dieser Arbeit soll sein, einen Überblick über den
-Stand der Architekturmuster, Protokolle und möglicher Problematiken bei der
-Implementierung von eigenen Echtzeit-Video-Streaming-Diensten geben,
-\textcolor{orange}{diese anhand eines Experiments durchsprechen und auswerten
-usw.}
+Echtzeit- oder Nahe-Zu-Echtzeit-Kommunikation zuverlässig zu bewältigen. Die
+
+Aufgabe dieser Arbeit soll sein, einen Überblick über den Stand der
+Architekturmuster, Protokolle und möglicher Problematiken bei der
+Implementierung von eigenen Echtzeit-Video-Streaming-Diensten geben. Dafür
+wird im Rahmen eines Experiments ein Kamera-System implementiert. Die
+Herangehensweise und Ansätze für die Implementierung werden erläutert und 
+anschließend wird das Ergebnis hinsichtlich der \textit{Quality of Service}
+im Zusammenhang mit dem Arbeitsaufwand ausgewertet.
 
@@
 
-%\bibliographystyle{IEEEtran}
-%\bibliography{IEEEabrv,references}
+\nocite{*}
+
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv, references}
 

### 2021-12-05 computergrafik/echtzeit-video-streaming.tex
Cite rfcs were applicable

@@
 verschickt werden können; Somit gibt es keinen ``Lock-In'' auf bestimmte Audio-
-oder Video-Codecs. \textcolor{orange}{aussage prüfen und quelle}.
+oder Video-Codecs.
 
-\textit{RTCP} ist das mit \textit{RTP} einhergehende Kontrollprotokoll. Es
-wird primär dazu verwendet, die Übertragungsparameter der Sender zu
-beeinflussen – z. B. durch ein Feedback zur Übertragungsqualität oder ein
-Abmelden der Session. Desweiteren bietet es eine persistente ID für die
-\textit{RTP}-Mitglieder, die über Programm-Neustarts hinweg zur Identifikation
-von Mitgliedern und der zuordnung von Datenströmen verwendet werden können.
-\textcolor{red}{cite rfc1889 kapitel 6 / 6.1 bzw. S. 15-17}
+\textit{Realtime-Transport-Control-Protocol} oder auch \textit{RTCP} ist das
+mit \textit{RTP} einhergehende Kontrollprotokoll. Es wird primär dazu
+verwendet, die Übertragungsparameter der Sender zu beeinflussen – z. B. durch
+ein Feedback zur Übertragungsqualität oder ein Abmelden der Session.
+Desweiteren bietet es eine persistente ID für die \textit{RTP}-Mitglieder, die
+über Programm-Neustarts hinweg zur Identifikation von Mitgliedern und der
+zuordnung von Datenströmen verwendet werden können.~\cite{RFC1889}
+% kapitel 6 / 6.1 bzw. S. 15-17
 
-Das Protokoll siedelt sich im TCP/IP-Stack über \textit{UDP} an
-\textcolor{red}{cite rfc1889 introduction}. Es fügt wichtige Informationen zu
-UDP-Datagrammen hinzu: Im wesentlichen eine Sequenznummer um die
-Sende-Reihenfolge zu codieren und einen Payload-Type, der den Codec des
-Segments angibt. Somit kann auch bei nicht sequenziell übertragenden
-Datagrammen die Ursprüngliche Reihenfolge rekonstruiert werden und es können
-bei verlorenen Segmenten Interpolationsalgorithmen verwendet werden.
-\textcolor{red}{quelle / beispiele suchen}. Der Payload-Type ist essenziell um
+\textit{RTP und RTCP} siedeln sich im TCP/IP-Stack über \textit{UDP}
+an.~\cite{RFC1889}
+
+\textit{RTP} fügt wichtige Informationen zu UDP-Datagrammen hinzu: Im
+wesentlichen eine Sequenznummer um die Sende-Reihenfolge zu codieren und einen
+Payload-Type, der den Codec des Segments angibt. Somit kann auch bei nicht
+sequenziell übertragenden Datagrammen die Ursprüngliche Reihenfolge
+rekonstruiert werden und es können bei verlorenen Segmenten
+Interpolationsalgorithmen verwendet werden. Der Payload-Type ist essenziell um
 ohne Session-Aushandlung zu kommunizieren wie der Empfänger die Daten zu
-decodieren hat, um eine Sinnvolle nachricht zu erhalten.
+decodieren hat, um eine sinnvolle Nachricht zu erhalten.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TODO
 In der ersten Version aus 1996 gab es einige Probleme bezüglich
-\textcolor{red}{Probleme suchen}. Diese wurden 2003 in dem RFC3550
-überarbeitet - somit wurde das RFC1889 durch die neuere Version 3550 obsolet.
+\textcolor{orange}{Probleme suchen}. Diese wurden 2003 in dem RFC3550
+überarbeitet - somit wurde der RFC1889 durch die neuere Version 3550 obsolet.
 In der aktuelleren Version wurden einige Änderungen eingearbeit: Im
@@
 \textcolor{orange}{Dummer text}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TODO
 
@@
 definiert, eine beliebte Wahl für ein Protokoll zur Session-Aushandlung ist
-allerdings das \textit{Session Initiation Protokoll} (oder kurz \textit(SIP)).
-% https://datatracker.ietf.org/doc/html/rfc3550#section-3
+allerdings das \textit{Session Initiation Protokoll} oder kurz
+\textit{SIP}.~\cite{RFC3550, RFC3261}
 

### 2021-12-05 computergrafik/echtzeit-video-streaming.tex
Cleanup figures and references

@@
-\documentclass{IEEEtran}
+\documentclass{article}
 
-\usepackage[a4paper, top=3cm, left=2cm, right=2cm, bottom=2cm]{geometry}
+\usepackage[a4paper]{geometry}
 \usepackage[ngerman]{babel}
@@
 \usepackage[T1]{fontenc}
+\usepackage{listings}
 \usepackage{hyperref}
@@
 \usepackage{xcolor}
+\usepackage{float}
+
+\lstset{
+    aboveskip=3mm,
+    belowskip=3mm,
+    showstringspaces=false,
+    columns=flexible,
+    basicstyle={\scriptsize\ttfamily},
+    numbers=none,
+    breaklines=true,
+    breakatwhitespace=true,
+    tabsize=4
+}
 
@@
 
+\newcommand{\figuresource}[1]{
+	\begin{center}Quelle: #1\end{center}
+}
+
 \title{Ansätze zum Echtzeit-Video-Streaming im Web}
@@
 \author{
-	\IEEEauthorblockN{Maximilian Schulke \textit{(Matrikel-Nr. 20215853)}}\\
-	\IEEEauthorblockA{
+	%\IEEEauthorblockN{Maximilian Schulke \textit{(Matrikel-Nr. 20215853)}}\\
+	%\IEEEauthorblockA{
 		Technische Hochschule Brandenburg \\
@@
 		Computergrafik
-	}
+	%}
 }
@@
 \begin{document}
+
 \begin{onecolumn}
+	\begin{titlepage}
+		\pagenumbering{gobble}
 
-\IEEEspecialpapernotice{
+	%	\IEEEspecialpapernotice{
 			betreut durch Prof.\ Dr.\ rer.\ nat.\ Reiner Creutzburg\\
@@
 			Abgabetermin \today
-}
+	%	}
 
@@
 		\listoftables
-
+	\end{titlepage}
 \end{onecolumn}
 
-\begin{twocolumn}
+\pagenumbering{arabic}
+\begin{onecolumn}
 \markboth{Hausarbeit Computergrafik – Maximilian Schulke}{}
@@
 
-\IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden dezentralisierung und
+%\IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden dezentralisierung und
 der damit einhergehenden Vernetzung der Arbeitswelt in den letzten
@@
 
-\subsection{1996: RTP \& RTCP – RFC 1889}
+\subsection{1996: RTP \& RTCP}
 
@@
 
-\subsection{2004:\ SRTP (RFC 3711)}
+\subsection{2004:\ SRTP}
 
@@
 weitestgehend auf dem \textit{Advanced Encryption Standard} (kurz.
-\textit{AES}) auf \textcolor{red}{(siehe RFC3711 Seite 19).}. Der RFC
-beschäftigt sich weitestgehend mit den kryptografischen Aspekten des
-Protokolls.
+\textit{AES}) auf.~\cite{RFC3711} Der RFC beschäftigt sich weitestgehend mit
+den kryptografischen Aspekten des Protokolls und ist in dieser Hinsicht für
+dies vorliegende Arbeit nur bedingt interessant. Dennoch bildet
+\textit{SRTP} eine wichtige Grundlage für sichere Echtzeitübertragung und
+wird im WebRTC-Standard verwendet.
 
@@
 bereitgestellt. Diese hat den Namen \textit{libsrtp} und ist öffentlich auf der
-Entwickler-Plattform GitHub einsehbar (siehe
+Entwickler-Plattform GitHub öffentlich einsehbar (siehe
 \href{https://github.com/cisco/libsrtp}{github.com/cisco/libsrtp}).
@@
 
-\subsection{2005: RTMP von Adobe}
+\subsection{2005: RTMP}
 
@@
 
-\subsection{2010: WebRTC (RFC 8825)}
+\subsection{2006: DTLS}
+
+Das \textit{Datagram-Transport-Layer-Security} Protokoll bietet starke
+Sicherheitsgarantien (equivalent zu \textit{TLS}) für Datagram basierte
+Protokolle wie z. B. \textit{UDP}.~\cite{RFC4347}
+
+\subsection{2010: WebRTC}
 
@@
 WebRTC stellt eine direkte Verbindung zwischen zwei Endgeräten her und
-verwendet einen mix aus RTP, RTCP, SDP, (ICE Candidates) und einem signaling
-server. Es unterstützt beliebte video codecs wie V8 / V9 und ist in der zukunft 
-auf AV1 angelegt. Der standard audio codec ist opus.
-
-Durch die direkte verbindung wird eine sogennante sub second latency erreicht.
-Dies beschreibt im grunde \textcolor{red}{XYZ}.
+verwendet einen mix aus SRTP, DTLS, SDP, (ICE Candidates) und einem
+Signaling-Channel. Es unterstützt beliebte video codecs wie V8 / V9 und ist in
+der zukunft auf AV1 angelegt. Der standard audio codec ist opus.
 
@@
 
-\section{Protokolle}
-\subsection{RTP}
-
-\subsection{RTCP}
-\subsection{RTSP}
-\subsection{SDP}
+% \section{Protokolle}
+% \subsection{RTP}
+% \subsection{RTCP}
+% \subsection{RTSP}
+% \subsection{SDP}
 % https://datatracker.ietf.org/doc/html/rfc4566#section-3.1
-\subsection{SID}
+% \subsection{SID}
 % https://datatracker.ietf.org/doc/html/rfc3261
 % \subsection{STUN}
-\subsection{WebRTC}
+% \subsection{WebRTC}
 
@@
 
-Das zu implementierende Kamera-System wird lediglich eine einzige Kernfunktion
-aufweisen: Sobald die Kamera an ist, streamt sie via WebRTC, mit ausreichend
-FPS (mindestens 15 im durchschnitt) in ausreichender Qualität (720 x\ 480),
-ihre Aufnahmen an ein User Interface. Dieses wird im Browser laufen, muss aber
-ausreichend Alternativen für andere Plattformen aufweisen (Nativ, Smartphone
+Das zu implementierende Kamera-System wird lediglich eine einzige
+Kernfunktion aufweisen: Sobald die Kamera an ist, streamt sie via
+WebRTC, mit ausreichend FPS (mindestens 15 im durchschnitt) in
+ausreichender Qualität (720$\times$480), ihre Aufnahmen an ein User
+Interface. Dieses wird im Browser laufen, muss aber ausreichend
+Alternativen für andere Plattformen aufweisen (Nativ, Smartphone
 usw.). Das Interface und die Kamera starten die Aushandlung des
-WebRTC-Streams über den Signaling Server. D.h.\ sie versenden SDP (Offer /
-Answer) und tauschen ihre ICE-Candidates aus.
+WebRTC-Streams über den Signaling Server. D.h.\ sie versenden SDP
+(Offer / Answer) und tauschen ihre ICE-Candidates aus.
 
@@
 	\centering
-	\caption{Signaling zwischen der Kamera und dem Interface (Angelehnt an IEEE
-	PAPER!!!!!!!!!!!)}
+	\caption{Signaling zwischen der Kamera und dem Interface.}
+	\figuresource{Schulke 2021, nach~\cite{WebRTC}}
 \end{figure}
@@
 	\caption{Asynchrone Abarbeitung einer IO intensiven Aufgabe}
+	\figuresource{Schulke 2021}
 \end{figure}
@@
 \subsubsection{Verbindungsaufbau}
-
-\begin{figure}[ht]
-	\includegraphics[width=0.5\textwidth]{diagram-signaling-server-tls}
-	\centering
-	\caption{Signaling-Server mit mehreren TLS Verbindungen}
-\end{figure}
 
@@
 
-\textcolor{red}{Diagramme einfügen!}
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{signaling-protocol}
+	\centering
+	\caption{Anwendungsbeispiel des Singaling-Protokolls}
+	\figuresource{Schulke 2021}
+\end{figure}
 
@@
 
-Diagramm einfügen..
-
-%      C
-%    s r r
-%  s   r   r
-% o    o     o
-
-%      C
-%    r s r
-%  r   s   r
-% o    o     o
-
-%      C
-%    r r s
-%  r   r   s
-% o    o     o
+\textcolor{orange}{Diagramm einfügen..}
 
@@
 	\caption{Broadcast-Channel Konzept}
+	\figuresource{Schulke 2021}
 \end{figure}
@@
 	\caption{Kommunikation von Nutzern über einen Broadcast-Channel}
+	\figuresource{Schulke 2021}
 \end{figure}
@@
 Moderne Browser wie Firefox, Chrome, Safari etc.\ stellen eine Implementierung
-des WebRTC-Stacks bereit, die durch eine übersichtliche API leicht zu
-integrieren ist. % Siehe https://caniuse.com/?search=webrtc
+des WebRTC-Standards bereit (siehe Abb.~\ref{fig:webrtc-support}), die
+durch eine übersichtliche API leicht zu integrieren ist. % Siehe https://caniuse.com/?search=webrtc
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{webrtc-support}
+	\centering
+	\caption[WebRTC Support-Matrix~\cite{WebRTCSupport}]{WebRTC Support-Matrix}\label{fig:webrtc-support}
+	\figuresource{\cite{WebRTCSupport}}
+\end{figure}
 
@@
 	\centering
-	\caption{WebRTC-Verbindungsaufbau (Siehe QUELLE)}
+	\caption[WebRTC-Verbindungsaufbau~\cite{WebRTC}]{WebRTC-Verbindungsaufbau}
+	\figuresource{\cite{WebRTC}}
 \end{figure}
@@
 	\centering
-	\caption{GStreamer Architektur QUELLE von GSTREAMER DOCS}
+	\caption[GStreamer Architektur~\cite{GStreamerManualIntro}]{GStreamer Architektur}\label{fig:gstreamer-arch}
+	\figuresource{\cite{GStreamerManualIntro}}
 \end{figure}
@@
 die Aushandlung von Datenformaten zwischen Pipeline-Elementen, Synchronisation
-von Datenströmen. (Siehe Abbildung X und QUELLE!!)
+von Datenströmen. (siehe Abb.~\ref{fig:gstreamer-arch}
+und~\cite{GStreamerManualIntro})
 
@@
 auf, welches den Codec-Support erweitert und weitere Komponenten zur Erstellung
-von komplexeren Pipelines bereitstellt. (Siehe QUELLE)
+von komplexeren Pipelines bereitstellt.
 
@@
 
+Siehe~\cite{GStreamerManualIntro}.
+
 % Siehe https://gstreamer.freedesktop.org/documentation/application-development/introduction/gstreamer.html?
@@
 \begin{itemize}
-	\item libgstreamer1.0dev oder so..
-	\item Paketnamen raussuchen!!!
+	\item libgstreamer-gl1.0-0
+	\item libgstreamer-opencv1.0-0
+	\item libgstreamer-plugins-bad1.0-0
+	\item libgstreamer-plugins-bad1.0-dev
+	\item libgstreamer-plugins-base1.0-0
+	\item libgstreamer-plugins-base1.0-dev
+	\item libgstreamer1.0-0
+	\item libgstreamer1.0-dev
+	\item libgstrtspserver-1.0-0
 \end{itemize}
-
-(Aufgelistet in Debian-Paket-Namen, da Raspberry PI Os Debian basiert ist)
 
@@
 \end{enumerate}
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{diagram-states}
+	\centering
+	\caption{Zustandsdiagramm der Kamera-Software}
+	\figuresource{Schulke 2021}
+\end{figure}
 
@@
 
-\end{twocolumn}
+\end{onecolumn}
 
@@
 
+\section*{Anhang}
+
+\begin{scriptsize}
+	\begin{flushleft}
+		\texttt{\textbf{signaling/main.rs}}
+		\lstinputlisting{implementation/signaling/src/main.rs}
+		\texttt{\textbf{signaling/lib.rs}}
+		\lstinputlisting{implementation/signaling/src/lib.rs}
+		\texttt{\textbf{signaling/tls.rs}}
+		\lstinputlisting{implementation/signaling/src/tls.rs}
+		\texttt{\textbf{signaling/rooms.rs}}
+		\lstinputlisting{implementation/signaling/src/rooms.rs}
+		\texttt{\textbf{signaling/state.rs}}
+		\lstinputlisting{implementation/signaling/src/state.rs}
+		\texttt{\textbf{signaling/signals.rs}}
+		\lstinputlisting{implementation/signaling/src/signals.rs}
+		\texttt{\textbf{camera/main.rs}}
+		\lstinputlisting{implementation/camera/src/main.rs}
+		\texttt{\textbf{camera/webrtc.rs}}
+		\lstinputlisting{implementation/camera/src/webrtc.rs}
+		\texttt{\textbf{camera/signaling.rs}}
+		\lstinputlisting{implementation/camera/src/signaling.rs}
+
+		\texttt{\textbf{interface/index.html}}
+		\lstinputlisting{implementation/interface/index.html}
+	\end{flushleft}
+\end{scriptsize}
+
+\begin{figure*}[p]
+	\includegraphics[width=\textwidth]{photo-board}
+	\centering
+	\caption{Rapsberry PI 4}
+\end{figure*}
+
+\begin{figure*}[p]
+	\includegraphics[width=\textwidth]{photo-cam}
+	\centering
+	\caption{Camera Modul}
+\end{figure*}
+
+\begin{figure*}[p]
+	\includegraphics[width=\textwidth]{photo-installed}
+	\centering
+	\caption{Installierte Kamera}
+\end{figure*}
+
 \end{appendix}
 \end{onecolumn}
+
+% TODO: Jede Abbildung erwähnen
+% TODO: Jede Section auf Rechtschreibung prüfen
+% TODO: Zitate checken!
+% TODO: Fehlende Tabellen einbauen
+
 \end{document}

### 2021-12-07 computergrafik/echtzeit-video-streaming.tex
Write the abstract

@@
 		\begin{abstract}
-			Das abstract schreibe ich zu letzt!
+			Durch die steigende dezentralisierung der Welt gewinnt die
+			Echtzeit-Übertragung von Daten exponentiell an Relevanz. Die
+			vorliegende Arbeit beschäftigt sich mit verschiedenen Protokollen
+			zur Lösung dieses Problems und führt eine experimentelle
+			Implementierung eines \textit{Peer-To-Peer} Kamera-Systems durch
+			und wertet diese anschließend hinsichtlich der Qualität der
+			Echtzeit-Übertragung, verbrauchten Ressourcen und ihres
+			Implementierungsaufwandes aus.
 		\end{abstract}
@@
 \pagenumbering{arabic}
+
 \begin{onecolumn}

### 2021-12-07 computergrafik/echtzeit-video-streaming.tex
Rewrite the RTMP section

@@
 Entwickler-Plattform GitHub öffentlich einsehbar (siehe
-\href{https://github.com/cisco/libsrtp}{github.com/cisco/libsrtp}).
+\url{https://github.com/cisco/libsrtp}).
 
@@
 
-% https://www.adobe.com/content/dam/acom/en/devnet/rtmp/pdf/rtmp_specification_1.0.pdf
 \textit{RTMP} ist ein von der Firma \textit{Adobe Systems Incorporated}
 spezifiziertes Protokoll zur Echtzeit-Übertragung von Multimedia-Streams.
-\textit{RTMP} wurde laut Spezifikation\textcolor{red}{siehe RTMP spec 1.0},
-entwickelt um über dem Transport-Protokoll \textit{TCP} verwendet zu werden.
+\textit{RTMP} wurde laut Spezifikation~\cite{RTMPSpec} entwickelt um
+über dem Transport-Protokoll \textit{TCP} verwendet zu werden.
 
 Das Protokoll wurde dazu entwickelt um im Kontext eines Flash-Players verwendet
-zu werden. Dies führt dazu, dass es heutzutage nurnoch bedingt Anwendung
-findet, da mittlerweile viele Browser ihren Flash-Player-Support eingestellt
-haben (\textcolor{red}{siehe Chrome und Firefox}) % https://support.mozilla.org/en-US/kb/end-support-adobe-flash
+zu werden. Dies führt dazu, dass es heutzutage nurnoch bedingt in Browsern
+Anwendung findet, da mittlerweile viele Browser ihren Flash-Player-Support
+eingestellt haben.~\cite{RTMPFlashSupport}
 
 Außerdem spricht die hohe Latenz von bis zu 30 Sekunden, die durch die
-Verwendung von \textit{TCP} zustandekommt (\textcolor{red}{siehe
-restram.io/streaming-protocls}), gegen den Einsatz von \textit{RTMP} in einem
-Echtzeit-Übertragungs-Kontext.
+Verwendung von \textit{TCP} zustandekommt~\cite{StreamingProtokolle} gegen den
+Einsatz von \textit{RTMP} in einem Echtzeit-Übertragungs-Kontext.
 
-Desweiteren gibt es noch Protokollvarianten die HTTP bzw. HTTPS als zugrundeliegendes
-Protokoll verwenden.\textcolor{red}{Siehe XYZ Quelle suchen}
+Desweiteren gibt nach~\cite{RTMP} noch weitere Protokollvarianten:
+
+\begin{itemize}
+	\item \textit{RTMPT} (Tunneled) verwendet \textit{HTTP} bzw. \textit{HTTPS}
+		als Transport-Protokoll.
+	\item \textit{RTMPE} (Encrypted)
+	\item \textit{RTMPTE} (Tunneled und Encrypted)
+	\item \textit{RTMPS} verwendet \textit{SSL}
+	\item \textit{RTMFP} verwendet \textit{UDP} als Transport-Protokoll
+		anstelle von \textit{TCP}
+\end{itemize}
+
+Auch wenn \textit{RTMP} aufgrund der hohen Latenz für Videokonferenzen
+ungeeignet ist, findet es heutzutage immernoch Anwendungen in
+unidirektionalen Datenübertragungen an eine große Anzahl von Teilnehmern, da es
+sich gut skalieren lässt.~\cite{RTMP}
 
@@
 
-Das \textit{Datagram-Transport-Layer-Security} Protokoll bietet starke
+Das \textit{Datagram-Transport-Layer-Security}-Protokoll bietet starke
 Sicherheitsgarantien (equivalent zu \textit{TLS}) für Datagram basierte
 Protokolle wie z. B. \textit{UDP}.~\cite{RFC4347}
+
+Es bildet heutzutage die Grundlage für sichere, verbindungslose Kommunikation.
+QUelle suchen.
 

### 2021-12-07 computergrafik/echtzeit-video-streaming.tex
Write about the relay architecture

@@
 
+\textcolor{red}{mehr schreiben}
+
 % https://datatracker.ietf.org/doc/html/rfc8834#section-10
@@
 
-Je mehr Nutzer / Endgeräte an einer peer-to-peer-übertragung teilnehmen, desto
+Je mehr Nutzer / Endgeräte an einer Peer-To-Peer-Übertragung teilnehmen, desto
 größer wird die Belastung der Bandbreite bei den einzelnen Teilnehmern. Jedes
-Paket muss nicht einmal, sondern n-mal verschickt werden (wobei n die anzahl
-der teilnehmer ist). Dies kann bei labilen oder einfach schwachen Netzwerken
-entweder zur verminderung der übertragungsqualität führen, oder in besonders
-schlimmen fällen sogar das restliche netzwerk eines einzelnen Teilnehmers
-negativ beeinflussen, da dieser die meiste Bandbreite für die wiederholte
-übertragung gleicher pakete verwendet.
+Paket muss nicht einmal, sondern $N$-mal verschickt werden (wobei $N$ die
+Anzahl der Teilnehmer ist). Dies kann bei labilen oder einfach schwachen
+Netzwerken entweder zur Verminderung der Übertragungsqualität führen, oder in
+besonders schlimmen Fällen sogar das restliche Netzwerk eines einzelnen
+Teilnehmers negativ beeinflussen, da dieser die meiste Bandbreite für die
+wiederholte Übertragung gleicher pakete verwendet.
 
@@
 
-Dieser ansatz hat klare vorteile: Das gesamte - lokale \& globale – netzwerk
+Dieser Ansatz hat klare Vorteile: Das gesamte - lokale \& globale – Netzwerk
 wird geschont. So würde die Anzahl möglicher Teilnehmer deutlich steigen, da
@@
 
+Leider ist dies aus verschiedenen Gründen bisher nur in der Theorie möglich.
+Der Hauptgrund ist die mangelnde Unterstützung für Multicast-Pakete bei einem
+großteil der Router (Statistik suchen / Quelle suchen). Desweiteren würden
+Heimnetzwerke somit in Backbonenetzwerken deutlich mehr Netzwerklast auslösen
+können als bei typischen Heimnetz-Tarifen (Statistik suchen, Stand 2021 bis
+1gbit) zulässig wäre.
+
 \subsection{Relay}
+
+Die zentralisierte Architektur ist eine der besten Alternativen zum
+Peer-To-Peer-Ansatz. Sie wird bei diversen Streaming-Plattformen erfolgreich
+bei einer großen Anzahl von Nutzern betrieben. \textcolor{red}{Siehe twitch,
+youtube live etc}. Der wohl grundlegenste Unterschied liegt darin, dass
+zwischen den Teilnehmern eine zentralisierte Instanz geschaltet ist, die den
+Stream vervielfacht um so die Netzwerkauslastung bei einzelnen Teilnehmern zu
+minimieren. (Siehe Abbildung X!!!)
+
+% grafik einfügen
+
+Diese Architektur lässt im Gegensatz zu Peer-To-Peer-Architekturen deutlich
+mehr Teilnehmer zu, da nur der Relay-Server die Last skalieren muss und nicht
+ein Teilnehmer. Wenn ein Live-Stream auf einer Streaming-Plattform
+beispielsweise einige Hundert Zuschauer hat und dies via Peer-To-Peer laufen
+würde, wäre dies bereits eine Größe bei der typische Endgeräte und
+Heimnetzwerke definitiv überlastet wären (\textcolor{red}{siehe xyz}). Folglich
+finden zentralisierte Architekturen auch bei Mobilgeräte anwendung, da
+Mobilgeräte oft (abgesehen von einer stärker limiterten Bandbreite) lediglich
+über ein fixes Datenvolumen verfügen. (\textcolor{red}{siehe xyz})
+
+Zu den Vorteilen zählt die einfache Skalierbarkeit, die (weitestgehende)
+Unabhängigkeit von Teilnehmer Netzwerken und die Kontrolle über den verteilten
+Datenstrom.
+
+Die Nachteile hingegen zeichnen sich für die Teilnehmer in einer erhöhten
+Latenz ab, was ein Relay für Live-Videokonferenzen in der Regel uninteressanter
+macht. Desweiteren fallen Hosting-Kosten in direkter Abhängig von der
+Nutzeranzahl ab, da jeder Datenstrom vom Server verarbeitet wird.
+
 % https://webrtc.org/getting-started/turn-server

### 2021-12-07 computergrafik/echtzeit-video-streaming.tex
Cleanup small stuff

@@
 
-In diesem Kapitel wird ein experimentelles Kamera-System entwickelt, das die
-oben erläuterten Protokolle / Technologien verwendet.
+\textcolor{red}{In diesem Kapitel wird ein experimentelles Kamera-System
+entwickelt, das die oben erläuterten Protokolle / Technologien verwendet.}
 
@@
 
-Zu letzt ist die Kamera zu entwickeln..
+\textcolor{red}{Zu letzt ist die Kamera zu entwickeln..}
 
@@
 
-Siehe~\cite{GStreamerManualIntro}.
+Siehe~\cite{GStreamerManualIntro}
 
@@
 	\item libgstreamer-plugins-bad1.0-0
-	\item libgstreamer-plugins-bad1.0-dev
 	\item libgstreamer-plugins-base1.0-0
-	\item libgstreamer-plugins-base1.0-dev
 	\item libgstreamer1.0-0
-	\item libgstreamer1.0-dev
 	\item libgstrtspserver-1.0-0
@@
 - Netzwerkauslastung
+- Zeitaufwand für Impl
 
@@
 große Anzahl an Nutzern gleichzeitig Signale über diesen Server schickt,
-pendelt sich dieser aufgrund seiner asynchronen Architektur im Ruhemodus bei
-einer CPU-Last von 0.0\% ein, da er nur aktiv wird, wenn auch Nachrichten
+pendelt sich dieser aufgrund seiner asynchronen Architektur bei einer
+CPU-Last von 0.0\% ein, da er nur aktiv wird, wenn auch Nachrichten
 eintreffen.
@@
 Desweiteren funtkioniert auch die Videoübertragung vom Browser über eine
-WebRTC-Verbindung sehr gut, die Bildübertragungsraten waren i.d.R. über den
-angestrebten 15 Bildern pro Sekunde.
+WebRTC-Verbindung sehr gut, die Bildübertragungsraten waren in der
+Regel über den angestrebten 15 Bildern pro Sekunde.
 
@@
 
-Die nächsten Schritte um dieses Experiment weiterzuführen wären definitiv die
-Fehlersuche an der Kamera-Software um die Übertragungsqualität zu steigern.
+Einer der nächsten Schritte um dieses Experiment weiterzuführen wäre
+definitiv die Fehlersuche an der Kamera-Software um die
+Übertragungsqualität zu steigern.
 
@@
 		\texttt{\textbf{signaling/main.rs}}
-		\lstinputlisting{implementation/signaling/src/main.rs}
+		%\lstinputlisting{implementation/signaling/src/main.rs}
 		\texttt{\textbf{signaling/lib.rs}}
-		\lstinputlisting{implementation/signaling/src/lib.rs}
+		%\lstinputlisting{implementation/signaling/src/lib.rs}
 		\texttt{\textbf{signaling/tls.rs}}
-		\lstinputlisting{implementation/signaling/src/tls.rs}
+		%\lstinputlisting{implementation/signaling/src/tls.rs}
 		\texttt{\textbf{signaling/rooms.rs}}
-		\lstinputlisting{implementation/signaling/src/rooms.rs}
+		%\lstinputlisting{implementation/signaling/src/rooms.rs}
 		\texttt{\textbf{signaling/state.rs}}
-		\lstinputlisting{implementation/signaling/src/state.rs}
+		%\lstinputlisting{implementation/signaling/src/state.rs}
 		\texttt{\textbf{signaling/signals.rs}}
-		\lstinputlisting{implementation/signaling/src/signals.rs}
+		%\lstinputlisting{implementation/signaling/src/signals.rs}
 		\texttt{\textbf{camera/main.rs}}
-		\lstinputlisting{implementation/camera/src/main.rs}
+		%\lstinputlisting{implementation/camera/src/main.rs}
 		\texttt{\textbf{camera/webrtc.rs}}
-		\lstinputlisting{implementation/camera/src/webrtc.rs}
+		%\lstinputlisting{implementation/camera/src/webrtc.rs}
 		\texttt{\textbf{camera/signaling.rs}}
-		\lstinputlisting{implementation/camera/src/signaling.rs}
-
+		%\lstinputlisting{implementation/camera/src/signaling.rs}
 		\texttt{\textbf{interface/index.html}}
-		\lstinputlisting{implementation/interface/index.html}
+		%\lstinputlisting{implementation/interface/index.html}
 	\end{flushleft}
@@
 \begin{figure*}[p]
-	\includegraphics[width=\textwidth]{photo-board}
+	%\includegraphics[width=0.5\textwidth]{photo-board}
 	\centering
@@
 \begin{figure*}[p]
-	\includegraphics[width=\textwidth]{photo-cam}
+	%\includegraphics[width=0.5\textwidth]{photo-cam}
 	\centering
@@
 \begin{figure*}[p]
-	\includegraphics[width=\textwidth]{photo-installed}
+	%\includegraphics[width=0.5\textwidth]{photo-installed}
 	\centering
@@
 % TODO: Fehlende Tabellen einbauen
+% TODO: Centricular zitieren
 

### 2021-12-11 computergrafik/echtzeit-video-streaming.tex
Finish the webrtc / dtls section and format the paper

@@
 
-\title{Ansätze zum Echtzeit-Video-Streaming im Web}
+\title{Möglichkeiten zum Echtzeit-Video-Streaming im Web}
 
 \author{
-	%\IEEEauthorblockN{Maximilian Schulke \textit{(Matrikel-Nr. 20215853)}}\\
-	%\IEEEauthorblockA{
-		Technische Hochschule Brandenburg \\
-		B.Sc. Medieninformatik \\
-		Computergrafik
-	%}
+	\vspace{0.5em}
+	Maximilian Schulke \textit{(Matrikel-Nr. 20215853)}\\
+	\small{Technische Hochschule Brandenburg}\\
+	\small{B.Sc. Medieninformatik}\\
+	\small{Computergrafik}\\
+	\small{Wintersemester 2021}\\
+	\small{betreut durch Prof.\ Dr.\ rer.\ nat.\ Reiner Creutzburg}\\
 }
@@
 		\pagenumbering{gobble}
-
-	%	\IEEEspecialpapernotice{
-			betreut durch Prof.\ Dr.\ rer.\ nat.\ Reiner Creutzburg\\
-			Wintersemester 2021\\
-			Abgabetermin \today
-	%	}
 
@@
 
-%\IEEEPARstart{M}{\MakeLowercase{it}} der zunehmenden dezentralisierung und
-der damit einhergehenden Vernetzung der Arbeitswelt in den letzten
-Jahren~\cite{CoronaHomeOffice} werden auch digitale Meeting-Systeme immer
-relevanter und müssen immer mehr Nutzer in Echtzeit mit einander verbinden um
-einen reibungslosen Arbeitsalltag zu gewährleisten. Zu begin der
-Corona-Pandemie stieg der Netzwerktraffic für VoIP und Videokonferenzen im März
-2020 um 210\%-285\%~\cite{NetzwerkStatistik}.
+Mit der zunehmenden dezentralisierung und der damit einhergehenden Vernetzung
+der Arbeitswelt in den letzten Jahren~\cite{CoronaHomeOffice} werden auch
+digitale Meeting-Systeme immer relevanter und müssen immer mehr Nutzer in
+Echtzeit mit einander verbinden um einen reibungslosen Arbeitsalltag zu
+gewährleisten. Zu begin der Corona-Pandemie stieg der Netzwerktraffic für
+\textit{Voice over IP} (kurz \textit{VoIP}) und Videokonferenzen im März 2020
+um 210\%-285\%~\cite{NetzwerkStatistik}.
 
@@
 Datenmengen über einen zentralisierten Streaming-Server zu schicken der diese
-an alle Teilnehmer broadcasted? Die Entwicklung der letzten Jahre deuten immer
-mehr darauf hin, dass \textit{Peer-To-Peer} basierte Lösungensansätze aufgrund
-der besseren Performance und Skalierbarkeit, in der Regel die bessere Wahl
-darstellen~\cite{PeerToPeerVersusCentralized}. \textit{Peer-To-Peer} halten
-unvorhersehbare Anstiege in der Nutzung~\cite{NetzwerkStatistik} deutlich
-besser aus, da die Echtzeit-Datenverarbeitung nicht bei dem
-Konferenzsystem-Anbieter stattfindet, sondern bei den Teilnehmern.
+an alle Teilnehmer weiterleitet? Die Entwicklungen der letzten Jahre deuten
+immer mehr darauf hin, dass \textit{Peer-To-Peer} basierte Lösungensansätze
+aufgrund der besseren Performance und Skalierbarkeit, in der Regel die bessere
+Wahl darstellen~\cite{PeerToPeerVersusCentralized}.
+\textit{Peer-To-Peer}-Architekturen halten einen unvorhersehbaren Anstieg in
+der Nutzung deutlich besser aus, da die Echtzeit-Datenverarbeitung nicht bei
+dem Konferenzsystem-Anbieter stattfindet, sondern bei den Teilnehmern.
 
@@
 
-Es ist also (immer noch) eine sehr aktuelle Thematik in der Informatik
-Echtzeit- oder Nahe-Zu-Echtzeit-Kommunikation zuverlässig zu bewältigen. Die
+Es ist also immer noch eine sehr aktuelle Thematik in der Informatik
+Echtzeit- oder Nahe-Zu-Echtzeit-Kommunikation zuverlässig zu bewältigen.
 
-Aufgabe dieser Arbeit soll sein, einen Überblick über den Stand der
+Die Aufgabe dieser Arbeit soll sein, einen Überblick über den Stand der
 Architekturmuster, Protokolle und möglicher Problematiken bei der
-Implementierung von eigenen Echtzeit-Video-Streaming-Diensten geben. Dafür
-wird im Rahmen eines Experiments ein Kamera-System implementiert. Die
+Implementierung von eigenen Echtzeit-Video-Streaming-Diensten geben. Dafür wird
+im Rahmen eines Experiments ein Kamera-System implementiert. Die
 Herangehensweise und Ansätze für die Implementierung werden erläutert und
-anschließend wird das Ergebnis hinsichtlich der \textit{Quality of Service}
-im Zusammenhang mit dem Arbeitsaufwand ausgewertet.
+anschließend wird das Ergebnis hinsichtlich der \textit{QoS} im Zusammenhang
+mit dem Arbeitsaufwand ausgewertet.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TODO
-In der ersten Version aus 1996 gab es einige Probleme bezüglich
-\textcolor{orange}{Probleme suchen}. Diese wurden 2003 in dem RFC3550
-überarbeitet - somit wurde der RFC1889 durch die neuere Version 3550 obsolet.
-In der aktuelleren Version wurden einige Änderungen eingearbeit: Im
-wesentlichen ``RTCP Packet Send and Receive Rules'' ``Layered Encodings''
-``Congestion Control'' ``Security Considerations'' ``IANA Considerations''
-\textcolor{orange}{Dummer text}.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TODO
+Die ersten Version aus 1996 wurde 2003 durch die überarbeitet Version,
+spezifiziert in dem RFC3550, abgelöst.~\cite{RFC3550}
 
@@
 Sicherheitsgarantien (equivalent zu \textit{TLS}) für Datagram basierte
-Protokolle wie z. B. \textit{UDP}.~\cite{RFC4347}
-
-Es bildet heutzutage die Grundlage für sichere, verbindungslose Kommunikation.
-QUelle suchen.
+Protokolle wie z. B. \textit{UDP}.~\cite{RFC4347} Es bildet heutzutage die
+Grundlage für sichere, verbindungslose Kommunikation.
 
@@
 Echtzeit-Kommunikations-Kanäle zwischen Nutzern (oder auch \textit{Peers})
-realisiert. Sie
+realisiert. Dafür stellt eine \textit{WebRTC}-Implementierung durch einen
+komplexen Signaling-Ablauf eine direkte Verbindung zu anderen Endgeräten
+her.~\cite{MDNSignaling}
 
-WebRTC stellt eine direkte Verbindung zwischen zwei Endgeräten her und
-verwendet einen mix aus SRTP, DTLS, SDP, (ICE Candidates) und einem
-Signaling-Channel. Es unterstützt beliebte video codecs wie V8 / V9 und ist in
-der zukunft auf AV1 angelegt. Der standard audio codec ist opus.
+\subsubsection{Signaling}
 
-% Hat mit 1996 RTP und dann 2005 mit RTMP angefangen, wann kam SRTP?\ JSEP Global IP Solutions \& Google WebRTC kam 2010, Microsoft wollte CU-RTC
+Während der Signaling-Phase generiert zunächst einer der Teilnehmer eine
+\textit{SDP-Offer}. Diese wird über einen, nicht näher im Standard
+beschreibenen, Signaling-Channel an den anderen Teilnehmer gesendet.
+Dieser verarbeitet die \textit{SDP-Offer} und generiert eine
+\textit{SDP-Answer} und sendet diese über den gleichen Signaling-Channel an den
+Absender zurück.
 
-% > Datenströme sind verpflichtend zu verschlüsseln. Dazu werden Verbindungen über DTLS verschlüsselt und Audio- und Videokommunikation zusätzlich durch SRTP geschützt.[16]
-% https://datatracker.ietf.org/doc/html/rfc8834#section-4.1
+Nach erfolgreichem austauschen der \textit{SDP}-Nachrichten,
+verwendet generiert jeder Teilnehmer mithilfe des \textit{Interactive
+Connectivity Establishment}-Frameworks \textit{ICE}-Candidates und versendet
+sie über den Signaling-Channel. Diese werden verwendet um einen eindeutigen
+Weg durch das Internet zum anderen Teilnehmer zu finden. Dieser Vorgang ist
+notwendig, da durch die Einführung \textit{Network-Address-Translation}, in den
+meisten Netzwerken Endgeräte keine globale eindeutige IPv4-Adresse
+haben.~\cite{MDNSessionLifetime}
 
-% https://www.w3.org/2021/01/pressrelease-webrtc-rec.html.en
+\subsubsection{Übertragung}
 
-Die ersten \textit{Requests for Comments} oder auch \textit{RFC} zu
-den grundlegenden Protokollen \textit{RTP} und
-\textit{\textcolor{blue}{Protokoll X}}, auf denen die heutigen Protokolle
-weitestgehend aufbauen, wurden bereits in den späten 1990er Jahren
-veröffentlicht und seit dem immer weiterentwickelt. \textcolor{red}{quelle
-anhängen RFC35XX}.
+Sobald die Signaling-Phase erfolgreich beendet wurde und beide Teilnehmer sich
+im Internet gefunden haben, wird eine direkte \textit{SRTP}-Verbindung
+hergestellt.~\cite{RFC8834} Diese wird über \textit{DTLS}
+verschlüsselt.~\cite{RFC8827}
+
+Anschließend können die zu übertragenden Daten in den entsprechenden Codecs
+über diese Verbindung versendet werden.
 
@@
 client-server-architektur eine fixe adresse zur verbindung haben. Dieses
-problem wird typischerweise über einen signaling server gelöst.
+problem wird typischerweise über einen Signaling-Server gelöst.
 
-\subsubsection{Signaling Server}
-Ein signaling server ist allen peers bekannt und dient als
-kommunikationsplattform, damit sich die beiden (sich gegenseitig initial
-unbekannten) peers gegenseitig vorstellen können.
+\subsubsection{Signaling-Server}
+Ein Signaling-Server ist allen Teilnehmern bekannt und dient als
+Kommunikationsplattform. Somit können sich zwei, sich initial unbekannte
+Teilnehmer, gegenseitig vorstellen.
 
-Signaling server sind interessanterweise keine feste anforderung für peer to
-peer muster. Wenn alle peers immer statische addressen hätten, könnten sie auch
-offline ihre ips / ice candidates / sdp offers etc.\ austauschen und so eine
-session aufbauen. Wichtig ist leidlgich eine initiale out of band
-kommunikation.
+Signaling-Server sind interessanterweise keine feste Anforderung für
+Peer-To-Peer-Architekturen. Wenn alle Teilnehmer statische Addressen hätten,
+könnten sie auch Out-Of-Band ihre IP-Adressen und ggf.\ weitere Details zum
+Verbindungsaufbau austauschen, um eine Verbindung aufzubauen.
 
-Der signaling server kann außerdem noch nach verbindungsaufbau dazu verwendet
-werden um die bestehende verbindung zu optimieren. Peers können neue ICE
-candidates vorschlagen um die Übertragung an neue gegebenheiten im internet
-(z.B. ausfall eines routers) anzupassen. (siehe mdn)
+Der Signaling-Server kann weiterhin noch nach Verbindungsaufbau dazu verwendet
+werden um die bestehende Verbindung zu optimieren. Peers können neue
+ICE-Candidates vorschlagen um die Übertragung an neue Gegebenheiten anzupassen.
 
-% https://datatracker.ietf.org/wg/wish/about/
-
-\subsubsection{Holepunching}
-Holepunching beschreibt den prozess lokale firewalls und nats zu durchbrechen
-um eine direkte verbindung zwischen zwei endgeräten herzustellen.
-
-\textcolor{red}{mehr schreiben}
-
-% https://datatracker.ietf.org/doc/html/rfc8834#section-10
-% https://de.wikipedia.org/wiki/Multicast
-% peer to peer wäre möglich mit ip multicast mit vielen usern ohne pakete zu
-% duplizieren
-
-\subsubsection{IP-Multicast}
+\subsubsection{Theoretische Optimierung durch IP-Multicast}
 
@@
 Teilnehmers negativ beeinflussen, da dieser die meiste Bandbreite für die
-wiederholte Übertragung gleicher pakete verwendet.
+wiederholte Übertragung gleicher Pakete verwendet.
 
@@
 übertragen, und dieses von Multicast-Routern im internet multiplizieren zu
-lassen. (Siehe grafik)
+lassen. \textcolor{red}{(Siehe grafik)}
 
@@
 Der Hauptgrund ist die mangelnde Unterstützung für Multicast-Pakete bei einem
-großteil der Router (Statistik suchen / Quelle suchen). Desweiteren würden
-Heimnetzwerke somit in Backbonenetzwerken deutlich mehr Netzwerklast auslösen
-können als bei typischen Heimnetz-Tarifen (Statistik suchen, Stand 2021 bis
-1gbit) zulässig wäre.
+großteil der Router \textcolor{red}{(Statistik suchen / Quelle suchen)}.
+Desweiteren würden Heimnetzwerke somit in Backbonenetzwerken deutlich mehr
+Netzwerklast auslösen können als bei typischen Heimnetz-Tarifen
+\textcolor{red}{(Statistik suchen, Stand 2021 bis 1gbit)} zulässig wäre.
 
@@
 
-% https://webrtc.org/getting-started/turn-server
-
-% \section{Protokolle}
-% \subsection{RTP}
-% \subsection{RTCP}
-% \subsection{RTSP}
-% \subsection{SDP}
-% https://datatracker.ietf.org/doc/html/rfc4566#section-3.1
-% \subsection{SID}
-% https://datatracker.ietf.org/doc/html/rfc3261
-% \subsection{STUN}
-% \subsection{WebRTC}
-
 \section{Implementierung eines Kamera-Live-Streams}
@@
 usw.). Das Interface und die Kamera starten die Aushandlung des
-WebRTC-Streams über den Signaling Server. D.h.\ sie versenden SDP
+WebRTC-Streams über den Signaling-Server. D.h.\ sie versenden SDP
 (Offer / Answer) und tauschen ihre ICE-Candidates aus.
@@
 Auf der Einplatinencomputer ist das mitgelieferte Betriebssystem ``Raspberry PI
-OS'' und die entsprechenden Software-Abhängigkeiten installiert. 
-\textcolor{red}{Siehe GitHub für Dependencies.}
+OS'' und die entsprechenden Software-Abhängigkeiten (siehe~\ref{items:deps})
+installiert.
 
@@
 WebSocket-Verbindungen miteinander verknüpft. Er verwaltet sogenannte Räume.
-Ein Raum besteht aus 0 bis 2 Clients und dient dazu diese miteinander zu
+Ein Raum besteht aus 0 bis 2 Teilnehmer und dient dazu diese miteinander zu
 verbinden. Auf dem Server kann es mehrere Räume gleichzeitig geben – dadurch
@@
 WebRTC-Anwendungen verwendet werden. Ein Raum hat eine ID, diese ist eine
-256-Bit-Entropie. Clients werden durch eine 128-Bit-Entropie identifiziert.
+256-Bit-Entropie. Teilnehmer werden durch eine 128-Bit-Entropie identifiziert.
 
-So können sich zwei Clients durch ein Out-Of-Band kommuniziertes Secret (die
+So können sich zwei Teilnehmer durch ein Out-Of-Band kommuniziertes Secret (die
 Raum-ID) über diesen in Kontakt treten. Diese könnte beispielsweise, würde es
@@
 
-Das Interface verbindet sich mit dem Signaling Server und nimmt eine Raum-ID
-entgegen. Mit dieser Raum-ID wird dann auf dem Signaling Server entweder ein
+Das Interface verbindet sich mit dem Signaling-Server und nimmt eine Raum-ID
+entgegen. Mit dieser Raum-ID wird dann auf dem Signaling-Server entweder ein
 neuer Raum erstellt, falls noch keiner mit der ID existiert, oder es wird dem
 bestehenden Raum beigetreten. Nach dem ein weiterer User beigetreten ist, fängt
-der in \textcolor{red}{Kapitel XY} erklärte Aufbau eines WebRTC-Streams zu dem
+der unter~\ref{sec:signaling-spec} erklärte Aufbau eines WebRTC-Streams zu dem
 Nutzer an.
 
-\subsection{Signaling Server}
+\subsection{Signaling-Server}
 
-Als erstes wurde der Signaling Server entwickelt, da dieser keine
-Abhängigkeiten an seine Clients hat. Die Kamera-Software und Interface setzen
-beide jeweils den laufenden Signaling Server vorraus um korrekt zu
+Als erstes wurde der Signaling-Server entwickelt, da dieser keine
+Abhängigkeiten an seine Teilnehmer hat. Die Kamera-Software und Interface setzen
+beide jeweils den laufenden Signaling-Server vorraus um korrekt zu
 funktioneren.
@@
 wird.
-
-HIER ZIETIEREN WAS DAS ZEUG HÄLT / ERKLÄRGRAFIKEN NUTZEN
 
@@
 
-\textcolor{red}{Grafik einfügen}
-
-\subsubsection{Spezifikation des Signaling-Protokolls}
+\subsubsection{Spezifikation des Signaling-Protokolls}\label{sec:signaling-spec}
 
@@
 
-\begin{itemize}
+\begin{itemize}\label{items:deps}
 	\item libgstreamer-gl1.0-0
@@
 
-\begin{figure}[ht]
-	\includegraphics[width=0.5\textwidth]{diagram-states}
-	\centering
-	\caption{Zustandsdiagramm der Kamera-Software}
-	\figuresource{Schulke 2021}
-\end{figure}
-
 Im Regelfall dauert der 1. Schritt einige Millisekunden. Nach dem die
@@
 unter\ \ref{browser-webrtc} erläutert.
+
+Bei der Einbindung von \textit{GStreamer} wurde~\cite{GStreamerExamples} als
+Referenz verwendet.
 

### 2022-03-03 computergrafik/echtzeit-video-streaming.tex
Update computer graphics

@@
 		\listoffigures
-		\listoftables
 	\end{titlepage}
@@
 einer Verbindung zwischen zwei peers, da diese sich nicht wie bei der typischen
-client-server-architektur eine fixe adresse zur verbindung haben. Dieses
-problem wird typischerweise über einen Signaling-Server gelöst.
+Client-Server-Architektur eine fixe adresse zur verbindung haben. Dieses
+Problem wird typischerweise über einen Signaling-Server gelöst.
 
@@
 IP-Multicast-Adressen. Diese erlauben es einem Gerät ein IP-Paket einmalig zu
-übertragen, und dieses von Multicast-Routern im internet multiplizieren zu
-lassen. \textcolor{red}{(Siehe grafik)}
-
-Vergleichs grafik einfügen..
+übertragen, und dieses von Multicast-Routern im Internet multiplizieren zu
+lassen.
 
@@
 Leider ist dies aus verschiedenen Gründen bisher nur in der Theorie möglich.
-Der Hauptgrund ist die mangelnde Unterstützung für Multicast-Pakete bei einem
-großteil der Router \textcolor{red}{(Statistik suchen / Quelle suchen)}.
-Desweiteren würden Heimnetzwerke somit in Backbonenetzwerken deutlich mehr
-Netzwerklast auslösen können als bei typischen Heimnetz-Tarifen
-\textcolor{red}{(Statistik suchen, Stand 2021 bis 1gbit)} zulässig wäre.
+Der Hauptgrund ist die mangelnde Unterstützung für IPv4-Multicast-Pakete bei
+einem Großteil der Router. Desweiteren würden Heimnetzwerke somit in
+Backbonenetzwerken deutlich mehr Netzwerklast auslösen können als vorgesehen
+ist.
 
@@
 Peer-To-Peer-Ansatz. Sie wird bei diversen Streaming-Plattformen erfolgreich
-bei einer großen Anzahl von Nutzern betrieben. \textcolor{red}{Siehe twitch,
-youtube live etc}. Der wohl grundlegenste Unterschied liegt darin, dass
-zwischen den Teilnehmern eine zentralisierte Instanz geschaltet ist, die den
-Stream vervielfacht um so die Netzwerkauslastung bei einzelnen Teilnehmern zu
-minimieren. (Siehe Abbildung X!!!)
-
-% grafik einfügen
+bei einer großen Anzahl von Nutzern betrieben.~\cite{YoutubeRTMP, TwitchRTMP}
+Der wohl grundlegenste Unterschied liegt darin, dass zwischen den Teilnehmern
+eine zentralisierte Instanz geschaltet ist, die den Stream vervielfacht um so
+die Netzwerkauslastung bei einzelnen Teilnehmern zu minimieren.
 
@@
 würde, wäre dies bereits eine Größe bei der typische Endgeräte und
-Heimnetzwerke definitiv überlastet wären (\textcolor{red}{siehe xyz}). Folglich
-finden zentralisierte Architekturen auch bei Mobilgeräte anwendung, da
-Mobilgeräte oft (abgesehen von einer stärker limiterten Bandbreite) lediglich
-über ein fixes Datenvolumen verfügen. (\textcolor{red}{siehe xyz})
+Heimnetzwerke definitiv überlastet wären.
+
+Folglich finden zentralisierte Architekturen auch bei Mobilgeräte
+anwendung~\cite{YoutubeRTMP, TwitchRTMP}, da Mobilgeräte oft (abgesehen von
+einer stärker limiterten Bandbreite) lediglich über ein fixes Datenvolumen
+verfügen.
 
@@
 
-\textcolor{red}{In diesem Kapitel wird ein experimentelles Kamera-System
-entwickelt, das die oben erläuterten Protokolle / Technologien verwendet.}
+In diesem Kapitel wird ein experimentelles Kamera-System entwickelt. Dieses
+baut auf einem WebSocket-Secure basierten Signaling-Protokoll, \textit{WebRTC}
+und \textit{GStreamer} auf.
 
 \subsection{Ziel}
+
 Ziel dieses Experiments ist es eine möglichst einfach ein funktionierendes
-System zu entwickeln und mögliche Probleme, benötigten Zeitaufwand usw.\ zu
-ermitteln.  Desweiteren ist ein ``Performance'' Vergleich mit anderen
-Live-Stream systemen interessant. Es gilt also die frage zu klären, wie viel
-arbeit benötigt ein grundsätzlich funktionierendes Software-Produkt das auf
-Echtzeit-Übertragung beruht.
+Kamera-System zu entwickeln und mögliche Probleme, benötigten Zeitaufwand etc.\
+zu ermitteln. Zukünftig könnte noch ein Performance-Vergleich mit anderen
+Echtzeit-Systemen interessant sein.
+
+Es gilt also die Frage zu klären, wie viel Arbeit ein simples, aber
+funktionierendes Software-Produkt das auf Echtzeit-Übertragung beruht benötigt
+und wie gut die Übertragungsqualität anschließend ist.
 
@@
 WebRTC, mit ausreichend FPS (mindestens 15 im durchschnitt) in
-ausreichender Qualität (720$\times$480), ihre Aufnahmen an ein User
-Interface. Dieses wird im Browser laufen, muss aber ausreichend
+ausreichender Qualität (720$\times$480), ihre Aufnahmen an eine
+Benutzeroberfläche. Diese wird im Browser laufen, muss aber ausreichend
 Alternativen für andere Plattformen aufweisen (Nativ, Smartphone
-usw.). Das Interface und die Kamera starten die Aushandlung des
-WebRTC-Streams über den Signaling-Server. D.h.\ sie versenden SDP
-(Offer / Answer) und tauschen ihre ICE-Candidates aus.
+usw.). Die Benutzeroberfläche und die Kamera starten die Aushandlung des
+WebRTC-Streams über den Signaling-Server. D.h.\ sie versenden
+\textit{SDP}-Nachrichten und tauschen ihre \textit{ICE}-Candidates aus
+(siehe~\figurename~\ref{fig:signaling}).
 
@@
 	\centering
-	\caption{Signaling zwischen der Kamera und dem Interface.}
+	\caption{Signaling zwischen der Kamera und der Benutzeroberfläche.}\label{fig:signaling}
 	\figuresource{Schulke 2021, nach~\cite{WebRTC}}
@@
 
-Die angedachte Architektur wurde möglichst einfach gehalten. Sie besteht
-lediglich aus 3 Elementen:
+Die geplante Architektur ist besteht lediglich aus 3 Elementen:
 
@@
 
-\subsubsection{Interface}
+\subsubsection{Benutzeroberfläche}
 
-Das Interface verbindet sich mit dem Signaling-Server und nimmt eine Raum-ID
-entgegen. Mit dieser Raum-ID wird dann auf dem Signaling-Server entweder ein
-neuer Raum erstellt, falls noch keiner mit der ID existiert, oder es wird dem
-bestehenden Raum beigetreten. Nach dem ein weiterer User beigetreten ist, fängt
-der unter~\ref{sec:signaling-spec} erklärte Aufbau eines WebRTC-Streams zu dem
-Nutzer an.
+Die Benutzeroberfläche verbindet sich mit dem Signaling-Server und nimmt eine
+Raum-ID entgegen. Mit dieser Raum-ID wird dann auf dem Signaling-Server
+entweder ein neuer Raum erstellt, falls noch keiner mit der ID existiert, oder
+es wird dem bestehenden Raum beigetreten. Nach dem ein weiterer Teilnehmer
+beigetreten ist, fängt der unter~\ref{sec:signaling-spec} erklärte Aufbau eines
+WebRTC-Streams zu dem Nutzer an.
 
@@
 Als erstes wurde der Signaling-Server entwickelt, da dieser keine
-Abhängigkeiten an seine Teilnehmer hat. Die Kamera-Software und Interface setzen
-beide jeweils den laufenden Signaling-Server vorraus um korrekt zu
-funktioneren.
+Abhängigkeiten an seine Teilnehmer hat. Die Kamera-Software und
+Benutzeroberfläche setzen beide jeweils den laufenden Signaling-Server vorraus
+um korrekt zu funktioneren.
 
@@
 Eine logische Anforderungen an den Server ist das gleichzeitige verarbeiten
-mehrerer Verbindungen – ansonsten könnte immer nur ein Client alleine mit dem
-Server Verbunden sein. Dies würde die Signaling-Funktionalität eines Raumes
+mehrerer Verbindungen – ansonsten könnte immer nur ein Teilnehmer alleine mit
+dem Server Verbunden sein. Dies würde die Signaling-Funktionalität eines Raumes
 unbrauchbar machen.
@@
 
-\begin{figure}[ht]
-	\includegraphics[width=0.5\textwidth]{diagram-asynchronität}
-	\centering
-	\caption{Asynchrone Abarbeitung einer IO intensiven Aufgabe}
-	\figuresource{Schulke 2021}
-\end{figure}
-
 \subsubsection{Verbindungsaufbau}
@@
 Um eine WebSocket-Secure-Verbindung aufzubauen, muss zu erst eine TCP- und dann
-darüber eine TLS-Verbindung zu dem Client aufgebaut werden. Über diese wird
-dann zu erst in HTTP kommuniziert (siehe WebSocket verbindungsaufbau Quelle
-suchen) und nach einer erfolgreichen Nachricht des Servers mit dem Status-Code
-101 (Switching Protocols) wird die über den gleichen Transport-Weg (TLS)
-nach dem WebSocket-Protokoll kommuniziert.
+darüber eine TLS-Verbindung zu dem Teilnehmer aufgebaut werden. Über diese wird
+dann zu erst in HTTP kommuniziert und nach einer erfolgreichen Nachricht des
+Servers mit dem Status-Code 101 (Switching Protocols) wird die über den
+gleichen Transport-Weg (TLS) nach dem WebSocket-Protokoll kommuniziert.
 
@@
 Das Signaling-Protokoll ist in 2 Nachrichten-Typen unterteilt:
-Server-Nachrichten und Peer-Nachrichten. Clients dürfen nur Peer-Nachrichten
+Server-Nachrichten und Peer-Nachrichten. Teilnehmer dürfen nur Peer-Nachrichten
 senden, ansonsten wird die Verbindung aufgrund eines Protokoll-Verstoßes
@@
 \begin{itemize}
-	\item Hello \{Client-ID\}
+	\item Hello \{Teilnehmer-ID\}
 	\item Joined
 	\item Error
-	\item Room/Join \{Client-ID\}
-	\item Room/Leave \{Client-ID\}
+	\item Room/Join \{Teilnehmer-ID\}
+	\item Room/Leave \{Teilnehmer-ID\}
 	\item Room/Signal \{Signal\}
@@
 Nach einer erfolgreich aufgebauten Verbindung mit dem Server schickt dieser ein
-\textit{Hello} mit der zugewiesenen Client-ID.\ Darauf hin muss der Client ein
-\textit{JoinOrCreate} senden um einem Raum beizutreten. Bevor der Client die
-Nachricht \textit{Joined} vom Server erhält, darf dieser keine Signale
-verschicken.
+\textit{Hello} mit der zugewiesenen Teilnehmer-ID.\ Darauf hin muss der
+Teilnehmer ein \textit{JoinOrCreate} senden um einem Raum beizutreten. Bevor
+der Teilnehmer die Nachricht \textit{Joined} vom Server erhält, darf dieser keine
+Signale verschicken.
 
-\begin{figure}[ht]
-	\includegraphics[width=0.5\textwidth]{signaling-protocol}
-	\centering
-	\caption{Anwendungsbeispiel des Singaling-Protokolls}
-	\figuresource{Schulke 2021}
-\end{figure}
-
-Eine \textit{Room/Join} Nachricht wird an ggf.\ andere Clients im Raum
-verschickt, sobald ein Client diesen betritt. Diese Nachricht kann auf dem
-Client als anlass genutzt werden eine SDP-Offer zu erstellen, da nun sicher
-ist, das ein anderer Client im Raum ist. Wenn beide Clients sich an dieses
-Schema halten, schickt immer der Client, der sich zu erst Verbunden hat die
+Eine \textit{Room/Join} Nachricht wird an ggf.\ andere Teilnehmer im Raum
+verschickt, sobald ein Teilnehmer diesen betritt. Diese Nachricht kann auf dem
+Teilnehmer als Anlass genutzt werden eine SDP-Offer zu erstellen, da nun sicher
+ist, das ein anderer Teilnehmer im Raum ist. Wenn beide Teilnehmer sich an dieses
+Schema halten, schickt immer der Teilnehmer, der sich zu erst Verbunden hat die
 Einladung und der andere die Answer.
 
-Analog wird eine \textit{Room/Leave} Nachricht verschickt wenn ein Client einen
-Raum verlässt.
+Analog wird eine \textit{Room/Leave} Nachricht verschickt wenn ein Teilnehmer
+einen Raum verlässt.
 
@@
 
+Siehe \figurename~\ref{fig:signaling-protocol-flow} für einen kompletten
+WebRTC-Verbindungsaufbau.
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{signaling-protocol}
+	\centering
+	\caption{Anwendungsbeispiel des Singaling-Protokolls}\label{fig:signaling-protocol-flow}
+	\figuresource{Schulke 2021}
+\end{figure}
+
 \subsubsection{Raum-Verwaltung}
@@
 Die Raum-Verwaltung fällt unter anderem in die Kategorie der
-Synchronisationsprobleme. Es gibt unter Umständen $n$ Client-Verbindungen, die
-den Status von einem Raum erfragen wollen, und ggf.\ einen Anlegen möchten. Dies
-soll für alle Verbindungen öffentlich geschehen, Clients dürfen sich aber dabei
-nicht gegenseitig überschreiben.
-
-% Mutual exclusion -> gegenseitiger ausschluss -> nur einer darf
+Synchronisationsprobleme. Es gibt unter Umständen $n$ Teilnehmer-Verbindungen,
+die den Status von einem Raum erfragen wollen, und ggf.\ einen Anlegen möchten.
+Dies soll für alle Verbindungen öffentlich geschehen, Teilnehmer dürfen sich
+aber dabei nicht gegenseitig überschreiben.
 
@@
 Als konkrete Lösung für das bestehende Problem bietet sich eine HashMap, deren
-zugriff von einem Mutex kontrolliert wird, an. Jede Client-Verbindung, die
+zugriff von einem Mutex kontrolliert wird, an. Jede Teilnehmer-Verbindung, die
 aktuelle Raum-Informationen braucht (was hauptsächlich bei dem erstellen /
@@
 
-\subsubsection{Kommunikation zwischen mehreren Client-Verbindungen}
+\subsubsection{Kommunikation zwischen mehreren Teilnehmer-Verbindungen}
 
 Bis zu diesem Punkt ist der Aufbau der Verbindung und das zu implementierende
-Protokoll klar, allerdings fehlt noch das Verknüpfen von Client-Verbindungen um
-Signale weiterzuleiten.
+Protokoll klar, allerdings fehlt noch das Verknüpfen von
+Teilnehmer-Verbindungen um Signale weiterzuleiten.
 
@@
 In diesem Fall ist allerdings zusätzlich noch bidirektionale Kommunikation
-gefragt, da beide Clients Signals des jeweils anderen erhalten sollen. Dafür
-bieten sich sogenannte Broadcast-Channel an. Sie sind dafür ausgelegt, dass von
-mehreren Stellen in diese geschrieben und gelesen wird. Sie funktioneren analog
-zu dem Broadcast aus der Netzwerktechnik.
-
-\textcolor{orange}{Diagramm einfügen..}
+gefragt, da beide Teilnehmer die Signale des jeweils anderen erhalten sollen.
+Dafür bieten sich sogenannte Broadcast-Channel an. Sie sind dafür ausgelegt,
+dass von mehreren Stellen in diese geschrieben und gelesen wird. Sie
+funktioneren analog zu dem Broadcast aus der Netzwerktechnik (siehe
+\figurename~\ref{fig:broadcast}).
 
@@
 	\centering
-	\caption{Broadcast-Channel Konzept}
+	\caption{Broadcast-Channel Konzept}\label{fig:broadcast-channel}
 	\figuresource{Schulke 2021}
@@
 
-Somit kriegt jeder Client bei betreten eines Raumes Zugriff auf einen Sender
-und einen Empfänger dieses Broadcast-Channels. In diesen schreibt er bei erhalt
-einer Peer-Nachricht vom Typ \textit{Signal} das Signal. Eine Client-Verbindung
-die gerade keine CPU-Zeit erhält, da ansonsten keine Ereignisse aufgetreten
-sind, wird wieder aktiv sobald eine neue Nachricht über den Channel da ist.
-Diese kann dann als Server-Nachricht vom Typ \textit{Room/Signal} mit dem
-Signal aus dem Channel an den Empfänger gesendet werden.
+Somit kriegt jeder Teilnehmer bei betreten eines Raumes Zugriff auf einen
+Sender und einen Empfänger dieses Broadcast-Channels. In diesen schreibt er bei
+erhalt einer Peer-Nachricht vom Typ \textit{Signal} das Signal (siehe
+\figurename~\ref{fig:broadcast-usage}). Eine Teilnehmer-Verbindung die gerade
+keine CPU-Zeit erhält, da ansonsten keine Ereignisse aufgetreten sind, wird
+wieder aktiv sobald eine neue Nachricht über den Channel da ist.  Diese kann
+dann als Server-Nachricht vom Typ \textit{Room/Signal} mit dem Signal aus dem
+Channel an den Empfänger gesendet werden.
 
-\begin{figure}[ht]
+\begin{figure}[ht!]
 	\includegraphics[width=0.5\textwidth]{diagram-websocket-verknüpfung}
 	\centering
-	\caption{Kommunikation von Nutzern über einen Broadcast-Channel}
+	\caption{Kommunikation von Nutzern über einen Broadcast-Channel}\label{fig:broadcast-usage}
 	\figuresource{Schulke 2021}
@@
 
-\subsection{Interface}
+\subsection{Benutzeroberfläche}
 
 Um den Signaling-Server für ein erstes nutzbares Zwischenergebnis zu verwenden,
-wurde eine einfache browserbasierte Benutzer-Oberfläche implementiert, die
+wurde eine einfache browserbasierte Benutzeroberfläche implementiert, die
 analog zu sehr bekannten Videokonferenz-Systemen wie z.B. Zoom oder Google Meet
@@
 
-Ein Benutzer muss über dieses Interface einem beliebigen Raum auf dem
+Ein Benutzer muss über diese Benutzeroberfläche einem beliebigen Raum auf dem
 Signaling-Server beitreten können. Sobald ein weiterer Benutzer beitritt soll
@@
 muss eine WebSocket-Secure Verbindung mit dem Server aufgebaut werden. Dies
-gelingt ebenfalls über eine Web-API:\
-https://developer.mozilla.org/de/docs/Web/API/WebSocket
+gelingt ebenfalls über eine Web-API.\
 
@@
 
-% https://webrtc.org/getting-started/media-devices
-
 \subsubsection{Verbindungsaufbau}\label{browser-webrtc}
@@
 Moderne Browser wie Firefox, Chrome, Safari etc.\ stellen eine Implementierung
-des WebRTC-Standards bereit (siehe Abb.~\ref{fig:webrtc-support}), die
-durch eine übersichtliche API leicht zu integrieren ist. % Siehe https://caniuse.com/?search=webrtc
+des WebRTC-Standards bereit (siehe \figurename~\ref{fig:webrtc-support}), die
+durch eine übersichtliche API leicht zu integrieren ist.
 
@@
 
-% Signalisierungsablauf zwischen Interface und Kamera visualisiert:
+% Signalisierungsablauf zwischen Benutzeroberflächen und Kamera visualisiert:
 
@@
 Nach dem der Signalisierungs-Server eine \textit{Room/Leave}-Nachricht gesendet
-hat, kann die RTCPeerConnection ohne negative Auswirkungen abgebaut werden. Im
-Interface wird die Anzeige des Video-Streams des anderen Teilnehmers
-zurückgesetzt und die RTCPeerConnection geschlossen. Somit ist das Interface
-wieder bereit einem weiteren Raum beizutreten und erneut eine WebRTC Verbindung
-herzustellen.
+hat, kann die RTCPeerConnection ohne negative Auswirkungen abgebaut werden. In
+der Benutzeroberfläche wird die Anzeige des Video-Streams des anderen
+Teilnehmers zurückgesetzt und die RTCPeerConnection geschlossen. Somit ist die
+Benutzeroberfläche wieder bereit einem weiteren Raum beizutreten und erneut
+eine WebRTC Verbindung herzustellen.
 
-% Screenshots des Demos Interfaces anfügen
+% Screenshots der Demo Benutzeroberfläche anfügen
 
 \subsection{Kamera}
-
-\textcolor{red}{Zu letzt ist die Kamera zu entwickeln..}
 
@@
 Kamera), beherscht die meistgenutzten Media-Codecs für Audio und Video und kann
-über Plugins mit weiteren Funktionalitäten erweitert werden.
+über Plugins mit weiteren Funktionalitäten erweitert werden (siehe
+\figurename~\ref{fig:gstreamer-arch}).
 
@@
 die Aushandlung von Datenformaten zwischen Pipeline-Elementen, Synchronisation
-von Datenströmen. (siehe Abb.~\ref{fig:gstreamer-arch}
-und~\cite{GStreamerManualIntro})
+von Datenströmen. (siehe \figurename~\ref{fig:gstreamer-arch},~\cite{GStreamerManualIntro})
 
@@
 
-Siehe~\cite{GStreamerManualIntro}
-
-% Siehe https://gstreamer.freedesktop.org/documentation/application-development/introduction/gstreamer.html?
+Siehe~\cite{GStreamerManualIntro}\\
 
@@
 
-Die Installation des Betriebssystems für die Kamera-Hardware
+Sobald die Kamera-Software auf einem herkömlichen Desktop-System lauffähig ist,
+gilt es die Laufzeitumgebung auf der Kamera-Hardware vorzubereiten.
+
+Hierfür muss zunächst das Betriebssystem der Hardware, Raspberry PI OS,
+installiert werden. Anschließend sind die unter~\ref{items:deps} aufgezählten
+Software-Abhängigkeiten zu installieren. Sobald diese vorhanden sind, kann die
+Kamera-Software auf anderen Desktop-Systemen für die entsprechende
+System-Architektur (\textit{armv7-unknown-linux-gnueabihf}) kompiliert werden
+und anschließend auf den Einplatinencomputer übertragen werden.
 
@@
 
-\section{Benchmarks}
+Da die Kamera-Hardware nicht permanent Stromzufuhr haben wird, ist es hilfreich
+einen Betriebssystem-Service für anzulegen, der für die automatische Ausführung
+der Software verantwortlich ist. Somit muss nicht bei jedem Neustart händisch
+die Software ausgeführt werden.
 
-- CPU Load
-- Netzwerkauslastung
-- Zeitaufwand für Impl
+Unter Raspberry PI OS übernimmt diese aufgabe der Init-Service Systemd.
+
+% \section{Benchmarks}
+% - CPU Load
+% - Netzwerkauslastung
+% - Zeitaufwand für Impl
 
@@
 Erfolg. Auf einem System mit einem Intel i7-8565U-Chip übersteigt der Server
-mit 10 Benutzern nicht ansatzweise 1\% der CPU-Kapazität. Falls nicht eine
-große Anzahl an Nutzern gleichzeitig Signale über diesen Server schickt,
+mit 50 Benutzern nicht annähernd 1\% der CPU-Kapazität. Falls nicht eine
+große Anzahl an Nutzern gleichzeitig viele Signale über diesen Server schickt,
 pendelt sich dieser aufgrund seiner asynchronen Architektur bei einer
-CPU-Last von 0.0\% ein, da er nur aktiv wird, wenn auch Nachrichten
+CPU-Last von 0.0\% ein, da dieser nur aktiv wird, wenn auch Nachrichten
 eintreffen.
@@
 
-Einer der nächsten Schritte um dieses Experiment weiterzuführen wäre
-definitiv die Fehlersuche an der Kamera-Software um die
-Übertragungsqualität zu steigern.
+Einer der nächsten Schritte um dieses Experiment weiterzuführen wäre definitiv
+die Fehlersuche an der Kamera-Software um die Übertragungsqualität zu steigern.
 

### 2022-03-03 einführung-in-die-wissenschaftliche-projektarbeit/extended-abstract/extended-abstract.tex
Add ewp contents

\documentclass{article}

\usepackage[style=apa]{biblatex}
\usepackage[a4paper, left=2cm, top=2cm, right=2cm, bottom=2cm]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{csquotes}
\usepackage{doc}

\hypersetup{colorlinks=true, allcolors=black}

\title{Die Sicherheit von energiesparenden 2,4-GHz-Kommunikationsprotokollen für vermaschte Netzwerke}
\author{\vspace{-0.1cm}\normalsize{Friedemann R. Pruß, Maximilian Schulke}\\\small{prussf@th-brandenburg.de 20215742, schulke@th-brandenburg.de 20215853}}
\date{\small{\today}\vspace{-0.1cm}}

\bibliography{references}

\begin{document}
\newgeometry{top=0.4cm, left=2cm, right=2cm, bottom=2cm}
\maketitle
\thispagestyle{empty}
\pagestyle{empty}

Die vorliegende Arbeit beleuchtet die Möglichkeiten zur sicheren Kommunikation
mit energiesparenden 2,4-GHz-Protokollen im Smart-Home-Kontext. Dazu werden zwei
weit verbreitete WPAN-Mesh-Protokolle – Thread und Bluetooth Mesh –
hinsichtlich ihrer Sicherheit verglichen. Vermaschte Netzwerke bieten im Smart
Home viele Vorteile, wie z.B. eine hohe Ausfallsicherheit und leichte
Erweiterbarkeit. Gegen die Verwendung von WAN-Protokollen wie LoRa, SigFox und
NB-IoT für Smart-Home-Netzwerke sprechen die hohen Kosten des initialen Aufbaus
eines WAN und die häufig niedrigeren Datenraten. Außerdem ist die Anzahl
möglicher Erweiterungen für das Netzwerk bei 2,4-GHz-Protokollen
deutlich höher \parencite{ThreadMeshVsOtherWirelessIEEE}.

Die Threadgroup, zu deren Mitgliedern Firmen wie Apple, Amazon und Google
gehören, bezeichnet Thread als Nachfolger von Zigbee
\parencite{ThePromiseOfThread}. Des Weiteren
setzt der IoT-Standard Matter für die Datenübertragung auf Thread \parencite{Matter}.
Matter wurde von der Connectivity Standards Alliance (ehem. Zigbee Alliance)
entwickelt, zu deren Mitgliedern unter anderen auch die vorher genannten Firmen
gehören \parencite{Matter}.

Das Forschungsergebnis wurde erreicht, in dem im Rahmen einer Survey und
Literaturanalyse die Spezifikationen der Protokolle und bereits getätigte
Sicherheitsanalysen ausgewertet wurden. Auch nach ausführlicher Recherche ließ
sich, trotz der bereits großen und immer weiter zunehmenden Bekanntheit der
beiden Protokolle, keine Publikation finden, in der ein detaillierter
Sicherheitsvergleich stattgefunden hat. In der Regel wurde eines der beiden
Protokolle alleinstehend betrachtet
\parencite{ThreadApplicationIEEE,BluetoothMeshIntro}, oder es fand ein
allgemeiner konzeptioneller Vergleich statt – zum Beispiel hinsichtlich
Architektur, Protokollaufbau, Funktionsumfang etc
\parencite{ComparativeAnalysisIEEE,ThreadMeshVsOtherWirelessIEEE}. Des Weiteren
sind die Spezifikationen für beide Protokolle öffentlich beziehungsweise
kostenlos einsehbar \parencite{BluetoothSpec,ThreadSpec}. Zu Bluetooth bzw.
Bluetooth Mesh gibt es weitreichende Sicherheitsanalysen und gut dokumentierte
Sicherheitslücken \parencite{BluetoothLowEnergyAttackOxford, BluetoothIssues}.
In weiteren Publikationen wurden Angriffe auf Thread durchgeführt
\parencite{ThreadEMAttack} und mögliche Schwachstellen untersucht
\parencite{ThreadSecurityCSIAC}. Die vorliegende Arbeit grenzt sich durch einen
sicherheitsbezogenen Vergleich von den oben gennanten Arbeiten ab.

In den letzten Jahren wurden mehrere, teils kritische, Sicherheitslücken in der
Bluetooth Spezifikation entdeckt \parencite{BluetoothIssues}. Diese
beeinträchtigen zum Großteil die Sicherheit von Direktverbindungen - also
Bluetooth BR/EDR und Bluetooth Low Energy. Zu den schwerwiegendsten
Schwachstellen in der Bluetooth Core Spezifikation zählen die sogenannten
``Bluetooth Impersonation Attacks''. Diese macht es möglich, dass sich ein
Angreifer als bereits authentifiziertes Gerät ausgibt und somit die komplette
Authentifizierung bei einem Verbindungsaufbau umgehen kann. Dies ermöglicht
Man-In-The-Middle-Attacken \parencite{BluetoothLowEnergyAttackOxford}.

Bluetooth Mesh verwendet als Übertragungsprotokoll Bluetooth Low Energy und
anders als beim Bluetooth-Low-Energy-Pairing werden elliptische 256-Bit-Kurven
und Out-of-Band-Authentifizierung verwendet, um das Hinzufügen von
Netzwerkknoten abzusichern \parencite{BluetoothSpec}. Allerdings wurden
diesbezüglich im Jahr 2020 die Sicherheitswarnungen CVE-2020-26556,
CVE–2020-26557, CVE-2020-26559, CVE-2020-26560 veröffentlicht
\parencite{BluetoothIssues}.

Thread ist im Gegensatz zu Bluetooth Mesh ein offenes Protokoll, da es kaum
einen Einfluss auf die Anwendungsebene des Netzwerks hat. Im Jahr 2020 wurde
eine Sicherheitsanalyse von Thread anhand der 10 relevantesten
Sicherheitsbedenken dem Open Web Application Security Project (OWASP) für
IoT-Netzwerke durchgeführt \parencite{ThreadSecurityCSIAC}.

Diese Analyse zeigt, dass Thread den Fokus primär auf die Verschlüsselung und
die Sicherung der Übertragung legt, aber Risiken durch die fehlende
Spezifikation der Anwendungsebene entstehen könnten. Hersteller müssen selbst
für die Sicherheit auf der Anwendungsebene sorgen. Ein großer Vorteil der
offenen Spezifikation ist die Möglichkeit angemessene und erforschte
Kryptografie einzusetzen.

Die aktuell bekannten Sicherheitslücken von Bluetooth Mesh betreffen das
Hinzufügen von neuen Netzwerkknoten \parencite{BluetoothIssues} und vermindern
somit die Integrität des Netzwerks. Bislang wurden bei Thread, trotz ausgiebiger
Untersuchungen \parencite{ThreadSecurityCSIAC, ThreadSecurityEmbeddedCom}
noch keine vergleichbaren Schwachstellen gefunden. Auf der anderen Seite tritt
Thread die Verantwortung zu Absicherung der Anwendungsschicht an den Hersteller
ab, ist somit zwar flexibler, könnte aber mögliche Sicherheitslücken durch
nicht abgesicherte Geräte öffnen.

Eine eindeutige Empfehlung für oder gegen eines der beiden Protokolle nur auf
Basis der Sicherheit kann nicht klar gegeben werden, da beide Protokolle
unterschiedliche Vorteile und Sicherheitsrisiken mit sich bringen. Eine
Entscheidung muss vom Implementierungskontext abhängig gemacht werden.

\restoregeometry

\printbibliography[heading=none]

\end{document}

### 2022-03-03 einführung-in-die-wissenschaftliche-projektarbeit/konzept/konzept-exposè.tex
Add ewp contents

\documentclass{article}

\usepackage[style=apa]{biblatex}
\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{csquotes}
\usepackage{doc}

\hypersetup{colorlinks=true, allcolors=black}

\newcommand{\titlecontent}{Drahtlose Low Power Kommunikationsprotokolle und ihre Sicherheit}

\title{\titlecontent}
\author{Friedemann Richard Pruß, Maximilian Schulke}

\bibliography{references}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		Technische Hochschule Brandenburg \\
		Online Studiengang Medieninformatik \\
		Fachbereich Informatik und Medien \\
		Einführung in die wissenschaftliche Projekarbeit\\
		Prof.\ Dr.\ rer.\ nat.\ Martin Christof Kindsmüller
	\end{flushleft}

	\vfill

	\begin{center}
		\large{Konzept:}\\
		\Huge{
			Drahtlose Low Power Kommunikationsprotokolle\\
			und ihre Sicherheit
		}\\[0.5em]
		\large{Wintersemester 2021}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{flushright}
		Friedemann Richard Pruß (Matrikel-Nr. 20215742)\\
		Maximilian Schulke (Matrikel-Nr. 20215853)
	\end{flushright}
\end{titlepage}

\newpage

\tableofcontents

\section{Thematische Fokussierung}

In einer vernetzten Welt mit unzähligen IoT-Geräten wird sichere Kommunikation
mit sogenannten Low-Power-Protokollen immer wichtiger. Smart-Devices wie zum
Beispiel Sicherheitskameras, Smart-Lampen oder allgemeiner – jegliche
Embedded-Hardware, lässt sich mittlerweile so gut wie überall wiederfinden.

Diese Geräte zeichnen sich dadurch aus, dass sie oft stark begrenzte Ressourcen
haben. In der Regel ist ein niedriger Stromverbrauch gewünscht, da viele Geräte
mit Akku- oder Batterie betrieben werden. Bei diesen ist folglich die
Rechenleistung eingeschränkt, da mit einer leistungsstarken CPU die
Batterielaufzeit enorm verkürzt würde.

Da Kommunikation dennoch meist eine elementare Aufgabe dieser Geräte ist, gibt
es zahlreiche Entwicklungen, um die Problematik energieeffizient zu lösen.
Low-Power-Protokolle benötigen weniger Strom als vergleichbare Protokolle aus
dem Desktop-Umfeld. Weiterhin ist die Vertraulichkeit \& Sicherheit der
kommunizierten Daten unabdinglich, da gerade im Smart-Home-Bereich ein Angriff
fatale Folgen haben könnte.

Die entstehende Arbeit soll in dem oben genannten Kontext das Thema der
sicheren Kommunikation genauer beleuchten. Dazu werden zwei der am weitesten
verbreiteten Protokolle, namentlich Thread und Bluetooth-Low-Energy (kurz
\textit{BLE}) hinsichtlich ihrer Sicherheit verglichen. Insbesondere ist die
Frage interessant, ob das jüngere Thread-Protokoll die bekannten
sicherheitsrelevanten Problematiken von Bluetooth-Low-Energy behebt.

\section{Motivation \& Abgrenzung}

Auch nach ausführlicher Recherche ließ sich, trotz der bereits großen und immer
weiter zunehmenden Bekanntheit der beiden Protokolle, keine Publikation finden,
in der ein detaillierter Sicherheitsvergleich stattgefunden hat. In der Regel
wurde eines der beiden Protokolle alleinstehend betrachtet
\parencite{ThreadInteroperabilityIEEE, BluetoothMeshIntroBLTW:online}, oder es
fand ein allgemeiner konzeptioneller Vergleich statt – zum Beispiel
hinsichtlich Architektur, Protkoll-Aufbau, Funktionsumfang etc.
\parencite{ThreadMeshVsOtherWirelessIEEE, ComparativeAnalysisIEEE,
ThreadVsBluetoothEnterpriseIotInsights:online}. Desweiteren sind die
Protokollspezifikationen für beide Protokolle öffentlich einsehbar
\parencite{ThreadSpec:online, BluetoothSpec:online}.

\textit{Antonioli, Tippenhauer und Rasmussen} haben im Jahr 2020 bereits eine
sehr detaillierte Arbeit über die Sicherheitsprobleme und Angriffsmöglichkeiten
auf Bluetooth-Low-Energy veröffentlicht.
\parencite{BluetoothLowEnergyAttackOxford}

In verschiedensten Arbeiten werden diverse Anwendungsszenarien erläutert und
die Vorteile der jeweiligen Architektur für die betrachtete Problemstellung
herausgestellt. \parencite{ThreadApplicationIEEE,
ThreadApplicationSmartBuildingsIEEE}

Die geplante Arbeit grenzt sich durch die Fokussierung auf den
Sicherheitsaspekt von den oben genannten Arbeiten ab. Im besten Fall soll diese
eine konkrete Empfehlung für oder gegen die Benutzung eines der Protokolle
geben – eventuell auch einen Ausblick über eine mögliche zukünftige
Entwicklung. Dafür werden mögliche Sicherheitslücken und -mechanismen von
Thread und Bluetooth-Low-Energy untersucht. Insbesondere werden sich diese
Untersuchungen auf mögliche Angriffe und die Mechanismen der Protokolle, um
sich vor eben diesen zu schützen, beziehen.
\parencite{BluetoothPracticalAttacksICACCS, ThreadSecurityCSIAC:online} Zu
typischen Angriffen zählen beispielsweise Man-In-The-Middle-Attacken.
\parencite{GeneralManInTheMiddle}

Andere Arbeiten stellen Eigenschaften und Problematiken, teils durch größer
angelegte Feldversuche, heraus. Anhand dieser experimentellen Untersuchungen
kann man Chancen und Risiken der verglichenen Protokolle gegeneinander abwägen.
Daraus soll die angestrebte Empfehlung abgeleitet werden.

\section{Recherchebericht}

Um einen Einstieg in das Thema zu finden, wurde graue und nicht-zitierfähige
Literatur gesichtet, wie zum Beispiel Blogposts, Videos und Wikipedia-Artikel.
Dadurch konnte ein grober Überblick über die zu bearbeitende Thematik –
\textit{Sicherheit im Smart-Home} – gewonnen werden.

Es wurde deutlich, dass momentan ein Umbruch der Kommunikationsprotokolle im
IoT-Bereich stattfindet. Da dies große Relevanz für die Zukunft der Branche
hat, lag es nahe, das Thema Sicherheit mit den am weitest verbreiteten
Kommunikationsprotokollen zu verbinden. Dies führte letztendlich zu dem Thema
der Arbeit: \textit{\titlecontent}.

% Mindmap + Notizen oder so.

% Thread vs Bluetooth -> wieso? Thread sehr neu, Bluetooth sehr bekannt und
% arbeit zu bluetooth.

Nachdem die Themenfindung abgeschlossen war, begann die fokussierte Recherche
bei den großen wissenschaftlichen Datenbanken. Dabei zeigten sich die
Datenbanken der IEEE, ACM und die Gesellschaft für Informatik besonders
ergiebig für das gewählte Thema.  Zuerst wurden die Publikationen nach
Schlagwörtern, Titeln und Aktualität vorgefiltert. Nachdem eine grobe
Vorauswahl bestand, hat sich das Überfliegen des Abstracts und der
Zusammenfassung als effiziente Möglichkeit zur Überprüfung der Relevanz einer
Arbeit erwiesen.

Dadurch fanden sich besonders interessante und relevante Quellen wie zum
Beispiel \textcite{BluetoothLowEnergyAttackOxford} oder
\textcite{ThreadMeshVsOtherWirelessIEEE}. Diese und weitere sehr
vielversprechende Arbeiten boten einen guten Anhaltspunkt um über deren
Literaturliste weitere themennahe Publikationen zu finden. Nach der ersten
Kollektion von Quellen, war es ein sinnvoller Schritt diese nach Kategorie (zum
Beispiel \textit{Bluetooth}, \textit{Thread} und \textit{Vergleiche}) und
Relevanz zu sortieren. Priorisiert wurde nach Aktualität, thematischer Nähe und
wissenschaftlicher Wertigkeit.

Jede relevante Publikation wurde in \BibTeX\ erfasst. Nachdem eine erste Liste
vorhanden war, lag es nahe, diese erneut auf Vollständigkeit und Verfügbarkeit –
insbesondere bei online Quellen – zu überprüfen. Dieses Vorgehen wurde so lange
wiederholt, bis sich für den Umfang der zu erstellenden Arbeit genügend
hochwertige Quellen ergeben haben.

%Bei der suche nach bestimmten werden ähnliche vorgeschlagen. Nach relevanz und
%zitathäufgigkeit filtern.

\newpage

\section{Zeitplan}

\begin{enumerate}
	\item[KW43]    – Recherche \& mediale vorbereitung Präsentation
	\item[KW44]    – Recherche \& Präsentation
	\item[KW45-48] – Inhaltliche Ausarbeitung
	\item[KW49]    – Schreiben des Extended Abstract
	\item[KW50-02] – Entwicklung des wissenschaftlichen Posters
\end{enumerate}

\section{Vorläufige Literaturliste}
\printbibliography[heading=none]
\nocite{*}

%Infos                         | Quelle
%Bluetooth ist unsicher weil.. | Quelle X

%Argumentation
%- Sicherheitslücke X gefunden in Jahr Y von Z
%- Thread bietet neuerung für Angriff

\end{document}

### 2022-03-03 einführung-in-die-wissenschaftliche-projektarbeit/poster/poster.tex
Add ewp contents

\documentclass{article}

\usepackage[style=apa]{biblatex}
\usepackage[a4paper, left=1.5cm, right=1.5cm]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{csquotes}
\usepackage{doc}

\hypersetup{colorlinks=true, allcolors=black}

\bibliography{references}
\pagestyle{empty}

\begin{document}

Sichere Kommunikation ist eine Grundvoraussetzung im Smart Home. Mesh Netzwerke
bieten viele Vorteile, wie eine hohe Ausfallsicherheit und leichte
Erweiterbarkeit. Bluetooth Mesh und Thread sind zwei sehr unterschiedliche Low
Power Mesh-Protokolle mit hoher Relevanz für die IoT-Industrie
\parencite{ThreadMeshVsOtherWirelessIEEE, HardwareSecurity, BLEMeshEmbeddedComputingCom}.

Im Gegensatz zu Thread baut Bluetooth Mesh auf dem weit verbreiteten Protokolle
Bluetooth Low Energy auf. Thread verwendet, ähnlich wie sein Vorgänger Zigbee
IP, IEEE 802.15.4 für die drahtlose Datenübertragung \parencite{ThreadSpec, BluetoothSpec}.
Diese Arbeit grenzt sich von vorangegangenen Arbeiten durch den
sicherheitsbezogenen Vergleich der beiden Protokolle ab. Wir präsentieren die
Ergebnisse unserer Literaturanalyse von Sicherheitsuntersuchungen und
Spezifikationen.

\textbf{On-Mesh Commissioning:} Der Commissioner kann dem Joiner festgelegte
Credentials bereitstellen um dem Netzwerk beizutreten. Er kann diese
Credentials auf die vom Hersteller vergebene EUI64 beschränken
\parencite{ThreadSpec}.

\textbf{External Commissioning:} Der authentifizierte Commissioner stellt die
Identität des Joiners sicher. Anschließend weist er den Border-Router an, dem
Joiner alle Informationen zum Beitritt des Netzwerks bereitzustellen
\parencite{ThreadSpec}.

Thread ist ein offenes Protokoll, da es im Vergleich zu Bluetooth Mesh, kaum
einen Einfluss auf die Anwendungsebene des Netzwerks hat.

Im Jahr 2020 wurde eine Sicherheitsanalyse von Thread anhand der 10
relevantesten Sicherheitsbedenken der OWASP für IoT-Netzwerke durchgeführt
\parencite{ThreadSecurityCSIAC}. Diese Analyse zeigt, dass Thread den Fokus
primär auf die Verschlüsselung und die Sicherung der Übertragung legt, aber
Risiken durch die fehlende Spezifikation der Anwendungsebene entstehen
könnten. Hersteller müssen selbst für die Sicherheit auf der Anwendungsebene
sorgen. Das macht es möglich, bereits gut erforschte
Verschlüsselungsprotokolle, wie z. B. DTLS, zu verwenden und diese im
Bedarfsfall auszu-tauschen. Diese Freiheit birgt aber auch Risiken, da sich der
Nutzer auf die Sicher-heitsmaßnahmen des Hersteller verlassen muss.

Dinu und Kizhvatov (\citeyear{ThreadEMAttack}) haben herausgefunden, dass durch
eine Verkettung von Schwachstellen und einer differenziellen
elektromagnetischen Analyse Rückschlüsse auf den Datenverkehr bei Thread zu
ziehen ist, und Netzwerkschlüssel abgegriffen werden können. Dies lohnt sich
allerdings aufgrund des hohen Aufwands i.d.R. bei Smart-Home-Netzwerken nicht,
sondern stellt eher eine Gefahr für kommerzielle Anwendungen dar
\parencite{ThreadEMAttack}.

In den letzten Jahren wurden mehrere, teils kritische, Sicherheitslücken in der
Bluetooth Spezifikation entdeckt \parencite{BluetoothIssues}.
Diese beeinträchtigen zum Großteil die Sicherheit von Direktverbindungen - also
Bluetooth BR/EDR und Bluetooth Low Energy (BLE). Zu den schwerwiegendsten
Schwachstellen in der Bluetooth Core Spezifikation zählen die sogenannten
“Bluetooth Impersonation Attacks”. Diese macht es möglich, dass sich ein
Angreifer als bereits authentifiziertes Gerät ausgibt und somit die komplette
Authentifizierung bei einem Verbindungsaufbau umgehen kann. Dies ermöglicht
Man-In-The-Middle-Attacken \parencite{BluetoothLowEnergyAttackOxford}, durch
die Daten ausgelesen, Daten manipuliert oder gefälschte Nachrichten in Umlauf
gebracht werden können.

Bluetooth Mesh verwendet als Übertragungsprotokoll BLE und verwendet, anders
als beim BLE-Pairing, elliptische 256-Bit-Kurven und
Out-of-Band-Authentifizierung um das Hinzufügen von Netzwerkknoten abzusichern
\parencite{BluetoothSpec}.
Allerdings wurden diesbezüglich im Jahr 2020 die Sicherheitswarnungen
CVE-2020-26556, CVE–2020-26557, CVE-2020-26559, CVE-2020-26560 veröffentlicht
\parencite{BluetoothIssues}.

Die aktuell bekannten Sicherheitslücken von Bluetooth Mesh betreffen das
Hinzufügen von neuen Netzwerkknoten \parencite{BluetoothIssues} und vermindern
somit die Integrität des Netzwerks. Bislang wurden bei Thread, trotz
ausgiebiger Untersuchungen \parencite{ThreadSecurityCSIAC, ThreadSecurityEmbeddedCom},
noch keine vergleichbaren Schwachstellen gefunden. Auf der anderen Seite tritt
Thread die Verantwortung zu Absicherung der Anwendungsschicht an den Hersteller
ab, ist somit zwar flexibler, könnte aber mögliche Sicherheitslücken durch
nicht abgesicherte Geräte öffnen. Eine eindeutige Empfehlung für oder gegen
eines der beiden Protokolle nur auf Basis der Sicherheit kann nicht klar
gegeben werden, da beide Protokolle unterschiedliche Vorteile und
Sicherheitsrisiken mit sich bringen. Eine Entscheidung muss vom
Implementierung-Kontext abhängig gemacht werden.

\printbibliography[heading=none]

\end{document}

### 2022-05-21 grundlagen-der-kryptografie/assignments/assignment-6/assignment-6.tex
Add assignment 6

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts} 
\usepackage{listings}
\usepackage{color}

\renewcommand\thesubsubsection{\textnormal{(\alph{subsubsection})}}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg\\
		Online Studiengang IT Sicherheit\\
		Fachbereich Informatik und Medien\\
		Algorithmen und Datenstrukturen\\
		Prof.\ Dr.\ rer.\ nat.\ Ulrich Baum
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 6}\\[0.5em]
		\large{Sommersemester 2022}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{flushright}
		Mara Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\newpage

\section*{Einsendeaufgabe 6}
\stepcounter{section}
\stepcounter{section}
\stepcounter{section}
\stepcounter{section}
\stepcounter{section}
\stepcounter{section}

\subsection{Zyklische Gruppen}

\subsubsection{}

\begin{gather*}
	O = \{1, 2, 3, 4, 6, 12\}\\
	\vspace{0.5em}\\
	\forall o \in O: o\, |\, 12
\end{gather*}

\subsubsection{}

\begin{align*}
	a^4 \neq a^6 \neq 1 \Rightarrow a\, \textnormal{ist ein Generator für}\, Z_{13}^*,\, \textnormal{da gilt}:
\end{align*}
\begin{align*}
	\forall o_n \in O: \exists n \in Z_{13}^* \land \textnormal{|<}n\textnormal{>|} = o_n: \forall o_m \in \{ o_m \in O: o_n \,|\, o_m \}: n^{o_m}\, = 1
\end{align*}

Anders ausgedrückt gilt für alle Elemente der Ordnung $n$ (also $a^n = 1$) dass
bei einem Vielfachen der Ordnung $nk$ mit $k \in \mathbb{N}: a^{nk} = 1$. Somit
lässt sich aus dem Fakt das $a^{nk} = 1$ ist nicht ableiten ob $a$ die Ordnung
$n$ oder $nk$ (mit $k \neq 1$) hat.

Ist allerdings $a^{nk} \neq 1$ schließt man somit alle Teiler von $nk$ aus.
Also konkreter: Gilt $a^4 \neq 1$ kann $a$ nicht die Ordnung 1, 2 oder 4
haben. Wenn nun zusätzlich $a^6 \neq 1$ gilt kann $a$ auch nicht die Ordnung 1,
2, 3 oder 6 haben. Da jedes Element eine Ordnung haben muss bleibt nun nurnoch
die Ordnung 12 über und wir sind sicher dass $a$ ein Generator für $Z_{13}^*$
ist.

\subsubsection{}

\begin{align*}
	g = xy
\end{align*}

\end{document}

### 2022-05-31 hardware-sicherheit/fido2.tex
Update fido2 paper

\documentclass[journal]{IEEEtran}

\usepackage[a4paper, left=2cm, right=2cm, top=2cm, bottom=3cm]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{xcolor}

\hypersetup{colorlinks=true, allcolors=black}

\title{Implementierung eines Atuhentifizierungs-Servers mit FIDO2 Support}
\author{
	\IEEEauthorblockN{Mara Schulke \textit{(Matrikel-Nr. 20215853)}}\\
	\IEEEauthorblockA{
		Technische Hochschule Brandenburg \\
		B.Sc. IT Sicherheit \\
		Hardware Sicherheit
	}
}

\begin{document}

\markboth{Hausarbeit Hardware Sicherheit – Mara Schulke}{}
\IEEEspecialpapernotice{
	betreut durch Prof.\ Dr.\ Oliver Stecklina\\
	Sommersemester 2022\\
	Abgabetermin \today
}

\maketitle

\begin{abstract}
\end{abstract}

\section{Einleitung}
\IEEEPARstart{A}{\MakeLowercase{uthentifizierung}} ist eines der größten
Probleme die durch verteilte Systeme entstehen. Es gibt zahlreiche
Möglichkeiten die Identität einer Gegenseite sicherzustellen allerdings weisen
viele von ihnen Schwachstellen hinsichtlich Man-In-The-Middle-Attacken und
basieren auf der Annahme, dass das System des Nutzers nicht kompromitiert
wurde. \textcolor{red}{tbd}

\section{Der FIDO2 Standard}

FIDO2 steht für \textit{F}ast \textit{ID}entity \textit{O}nline 2 und ist ein von
der FIDO Alliance entwickleter, offener und lizenzfreier Standard für
Hardware-Token gestützte Authentifizierung. (\textcolor{red}{quote})

Ein Hardware-Token (kann auch in Form eines Trusted-Plattform-Moduls oder als
Teil des Betriebssystems implementiert sein) ist ein physischer Speicher für
die FIDO Schlüsselpaare eines Nutzers. Kernmerkmale die FIDO2 von
herkömmlicher asymmetrischer Kryptografie unterscheiden sind beispielsweise die
Isolation der privaten Schlüssel auf dem Hardware-Token, die Notwendigkeit
einer Nutzerinteraktion zum Verwenden eines privaten Schlüssels und die
Generierung von einem Schlüsselpaar pro Online-Dienst.

Durch all diese Eigenschaften werden Schlsselverluste unwahrscheinlicher und
weniger Sicherheitskritisch, da selbst bei einem hypothetischen
Schlüsselverlustes der Schaden immer auf einen Online-Dienst begrenzt ist.
Die größte Schwachstelle ist allerdings der physische Diebstahl des
Hardware-Tokens, da dessen Besitz ausreicht für Impersonation-Attacken. Eine
Absicherung dagegen kann eine biometrische Authentifizierung erfolgen bevor ein
privater Schlüssel verwendet werden kann - wie beispielsweise bei FaceID.
\textcolor{red}{tbd}

\subsection{Welche Probleme löst FIDO2?}

Die Notwendigkeit für einen solchen Standard hat sich in den letzten Jahren
immer stärker gezeigt, da die klassische Knowledge-Based-Authentication
(\textcolor{red}{KBA})
durch zunehmende Rechenleistung und effizientere Angriffe immer unsicherer und
unhandlicher für Nutzer wird. Die minimale Passwortlängen steigen
dementsprechend an und führen zur Wiederverwendung von gleichen Login-Daten für
mehrere Online-Dienste. Bekannte Lösungen sind die Verwendung von sog.
Passwort-Managern um lange und zufällige Passwörter für verschiedenste
Online-Dienste zu verwenden ohne, dass sich Nutzer diese merken müssen. Solche
Passwort-Manager sind zwar eine Lösung für die sichere Aufbewahrung von langen
Passwörtern, können aber nicht die durch \textcolor{red}{KBA} eröffneten
Angriffsvektoren wie z.B. Man-In-The-Middle-Attacken. So bald ein Angreifer
in den Besitz des geheimen Wissens (in diesem Fall das Passwort) gelangt kann
dieser uneingeschränkt und unbegegrenzt oft auf das Zielsystem zugreifen, bis
der Nutzer seine Daten ändert (vorausgesetzt, der Angreifer hat dies noch nicht
getan).

Durch den Wechsel von \textcolor{red}{KBA} auf Zero-Knowledge-Proof basierte
Authentifizierungsmethoden lassen sich ganze Angriffsvektoren ausschließen, da
ein kompromitierter Server oder eine kompromitierte Verbindung niemals das
geheime Wissen des Nutzers einem Angreifer zugänglich machen. Das heißt, dass
sich ein Angreifer im Falle einer kompromitierten Verbindung maximal in die
Sitzung des Nutzers einschleichen könnte, allerdings bei der nächsten
Authentifizierung nicht erneut die Identität des Nutzers beweisen könnte und
somit den Zugriff verlieren würde.

Im Falle von FIDO2 kennt nichtmal der Nutzer selber seine Schlüssel da diese
auf einem Trusted-Plattform-Modul \textcolor{red}{TPM} oder einem externen
Hardware-Token gespeichert wird und der Beweis der Identität durch die Signatur
einer vom Authentifizierungs-Server ausgestellten Challange erfolgt die das
\textcolor{red}{TPM} oder der Hardware-Token intern durchführen und dem Nutzer
nur die Signatur zurückgeben. So stellt selbst ein kompromitiertes Nutzersystem
nur eine temporäre Schwachstelle dar.

\section{Relevante Protokolle: CTAP \& WebAuthn}

Der FIDO2 Standard umfasst hauptsächlich die beiden Protokolle \textit{CTAP}
und \textit{WebAuthn}. Diese unterteilen den gesammten
Authentifizierungsvorgang in
zwei Bereiche:

\textit{Client zu Authenticator} also die Kommunikation zwischen dem
Nutzersystem und dem Hardware-Token oder TPM und \textit{Client zu Server}, die
Kommunikation zwischen dem Nutzersystem und dem Online-Dienst bei dem sich der
Nutzer authentifizieren möchte.

\textcolor{red}{grafik}

\subsection{Gemeinsame Begriffsdefinitionen}

\subsubsection{Relying Party}

Ein Online-Dienst der den FIDO2 Standard zur Nutzerauthentifizierung verwendet.

\subsubsection{Authenticator}

Ein externer Hardware-Token oder ein Teil des Betriebssystems (z.B. über ein
TPM implementiert) der FIDO2 Schlüsselpaare verwaltet.

\subsubsection{Credential}

Ein FIDO2 Schlüssel

\subsubsection{CBOR}

Concise Binary Object Representation, ein Binär-Format zur Darstellung von JSON Objekten.

\subsection{CTAP}

Das Client-To-Authenticator-Protocol kurz \textcolor{red}{CTAP} ist Teil
des FIDO2 Standards und beschreibt den Ablauf der Kommunikation zwischen einem
Nutzersystem und dem Hardware-Token beziehungsweise dem Authenticator.

Neben einer Spezifikation für den Transportlayer / die Nachrichtenstruktur
besteht das Protokoll primär aus der sogenannten ``Authenticator API'' - diese
beschreibt Operationen die ein Nutzersystem auf einem Authenticator ausführen kann.

Spezifiziert sind die folgenden 6 Operationen:

\begin{enumerate}
	\item \textit{authenticatorMakeCredential} - Schlüsselpaar für eine
		``Relying Party'' erstellen
	\item \textit{authenticatorGetAssertion} - Signatur einer Challange
	\item \textit{authenticatorGetNextAssertion} - Nächste Signatur der
		Challange erhalten bei mehreren Schlüsselpaaren
	\item \textit{authenticatorGetInfo} - Informationen über die Fähigkeiten
		des Authenticators
	\item \textit{authenticatorClientPIN} - Setzt den Authenticator-PIN
	\item \textit{authenticatorReset} - Zurücksetzen auf Werkseinstellungen
\end{enumerate}

Die im Standard beschriebenen Transportlayer umfassen USB, NFC oder Bluetooth
Low Energy. Für jeden dieser Transportlayer gibt es eigene Mechanismen zum
sicheren Verbindungsaufbau und zur Nachrichtenstruktur.

\subsection{WebAuthn}

\section{Implementierung des Authetifizierungs-Servers}

\section{Technische Dokumentation}

Der Server öffnet den TCP Port 8080 und erwartet eine externe TLS
Terminierung. Tokens können dem Server über den \texttt{Authorization} Header
mitgegeben werden und der \texttt{Content-Type} aller Anfragen und Antworten
ist ausschließlich \texttt{application/json}.

\subsection{/auth/signup - Nutzer erstellen}

\begin{itemize}
	\setlength{\leftskip}{1.5cm}
	\setlength{\itemsep}{0pt}
	\item[Methode:] POST
	\item[Token:] -
	\item[Eingabe:] Credentials \{ email, password \}
	\item[Ausgabe:] UserDetails \{ token, verified, keys \}
	\item[Beschreibung:] Erstellt einen unverifizierten Nutzer ohne FIDO2 Keys.
		Gibt einen Verifizierungscode in den Server-Logs aus (könnte in einem
		echten Szenario per E-Mail verschickt werden).
\end{itemize}

\subsection{/auth/verify - Nutzer verifizieren}

\begin{itemize}
	\setlength{\leftskip}{1.5cm}
	\setlength{\itemsep}{0pt}
	\item[Methode:] POST
	\item[Token:] Notwendig
	\item[Eingabe:] Verification \{ code \}
	\item[Ausgabe:] -
	\item[Beschreibung:] Verifiziert einen Nutzer falls der Code mit dem bei
		der Registrierung generierten Code übereinstimmt.
\end{itemize}

\subsection{/auth/login - Nutzer anmelden}

\begin{itemize}
	\setlength{\leftskip}{1.5cm}
	\setlength{\itemsep}{0pt}
	\item[Methode:] POST
	\item[Token:] -
	\item[Eingabe:] Credentials \{ email, password \}
	\item[Ausgabe:] UserDetails \{ token, verified, keys \} | \textcolor{red}{webauthn challange}
	\item[Beschreibung:] Gibt dem Nutzer entweder seine UserDetails zurück oder
		stellt eine WebAuthn Authetifizierungs-Challange die der Nutzer
		signiert bei dem Endpunkt \textit{/auth/fido2/login} enreichen muss
		falls ein FIDO2 Schlüssel hinterlegt wurde.
\end{itemize}

\subsection{/auth/fido2/login - Nutzer mit WebAuthn anmelden}

\begin{itemize}
	\setlength{\leftskip}{1.5cm}
	\setlength{\itemsep}{0pt}
	\item[Methode:] POST
	\item[Token:] -
	\item[Eingabe:] \textcolor{red}{challange}
	\item[Ausgabe:] UserDetails \{ token, verified, keys \}
	\item[Beschreibung:] Validiert die WebAuthn Challange des Nutzers mit den
		hinterlegten FIDO2 Schlüsseln und gibt bei erfolgreicher Validierung
		dem Nutzer seine UserDetails zurück.
\end{itemize}

\subsection{/auth/fido2/challenges - WebAuthn Registrierungs Challange}

\begin{itemize}
	\setlength{\leftskip}{1.5cm}
	\setlength{\itemsep}{0pt}
	\item[Methode:] POST
	\item[Token:] Notwendig
	\item[Eingabe:] -
	\item[Ausgabe:] \textcolor{red}{ccr}
	\item[Beschreibung:] Startet einen Registrierungsprozess für einen FIDO2
		Schlüssel. Setzt Serverseitig den ``Key-Registration-State'' eines
		Nutzers und gibt eine Registrierungs-Challange zurück.
\end{itemize}

\subsection{/auth/fido2/keys - WebAuthn Registrierung abschließen}

\begin{itemize}
	\setlength{\leftskip}{1.5cm}
	\setlength{\itemsep}{0pt}
	\item[Methode:] POST
	\item[Token:] Notwendig
	\item[Eingabe:] \textcolor{red}{PublicKeyCredential}
	\item[Ausgabe:] Key \{ id \}
	\item[Beschreibung:] Nimmt die CTAP Ausgabe der
		``authenticatorMakeCredential'' Operation an und ordnet diesen
		Schlüssel dem Nutzer zu.
\end{itemize}

\subsection{/auth/fido2/keys/:id - WebAuthn Schlüssel entfernen}

\begin{itemize}
	\setlength{\leftskip}{1.5cm}
	\setlength{\itemsep}{0pt}
	\item[Methode:] DELETE
	\item[Token:] Notwendig
	\item[Eingabe:] \textit{/:id}
	\item[Ausgabe:] -
	\item[Beschreibung:] Entfernt einen FIDO2 Schlüssel anhand seiner ID.\@
\end{itemize}

    %app.at("/auth/login").post(auth::login);
    %app.at("/auth/signup").post(auth::signup);
    %app.at("/auth/verify")
        %.with(middleware::authenticated)
        %.post(auth::verify);

    %app.at("/auth/fido2/login").post(auth::fido2::login);
    %app.at("/auth/fido2/challenges")
        %.with(middleware::authenticated)
        %.post(auth::fido2::create_challenge);
    %app.at("/auth/fido2/keys")
        %.with(middleware::authenticated)
        %.post(auth::fido2::store_key);
    %app.at("/auth/fido2/keys/:id")
        %.with(middleware::authenticated)
        %.delete(auth::fido2::remove_key);

\section{Auswertung}

\section{Abbildungsverzeichnis}

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,references}

\end{document}

### 2022-05-31 it-forensik/sichere-datenlöschung.tex
Add initial slides for the id forensics class

\documentclass{beamer}

\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usecolortheme{seagull}
\usefonttheme{serif}

\title{Sichere Löschung von Festplatten und Datenspeichern}
\author{Mara Schulke\\\tiny{Matrikelnr. 20215853, SS22 B.Sc. IT Security, THB}}

\begin{document}

\begin{frame}
	IT Forensik
	\vspace{1em}
	\titlepage
\end{frame}

\begin{frame}{Übersicht}
	\tableofcontents
\end{frame}

\section{Gefahren durch falsche Löschung}
\begin{frame}{Gefahren durch falsche Löschung}
	\begin{itemize}
		\item Häufige Gerätewechsel bei mangelnder Aufbereitung des Speichermediums
			bergen die Gefahr dass ein Angreifer diese nach Verkauf / Entsorgung
			rekonstruieren kann
		\item Selbst normale Endanwender können mittels Software wie z.B.
			EaseUS schlecht gelöschte Speichermedien wiederherstellen 
	\end{itemize}
\end{frame}

\subsection{Wie funktioniert die Rekonstruktion von Datenseichern?}
\begin{frame}{Wie funktioniert die Rekonstruktion von Datenseichern?}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

\section{Typsiche Speichertypen}
\begin{frame}{Typsiche Speichertypen}
	\begin{itemize}
		\item HDD (Hard Disk Drive)
		\item SSD (Solid State Drive)
		\item SSHD (Solid State Hybrid Drive)
	\end{itemize}
\end{frame}

\subsection{Funktionsweise einer HDD}
\begin{frame}{Funktionsweise einer HDD}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

\subsection{Sichere Löschung einer HDD}
\begin{frame}{Sichere Löschung einer HDD}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

\subsection{Funktionsweise einer SSD}
\begin{frame}{Funktionsweise einer SSD}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

\subsection{Sichere Löschung einer SSD}
\begin{frame}{Sichere Löschung einer SSD}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

\subsection{Funktionsweise einer SSHD}
\begin{frame}{Funktionsweise einer SSHD}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

\subsection{Sichere Löschung einer SSHD}
\begin{frame}{Sichere Löschung einer SSHD}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

% https://www.bsi.bund.de/DE/Themen/Verbraucherinnen-und-Verbraucher/Informationen-und-Empfehlungen/Cyber-Sicherheitsempfehlungen/Daten-sichern-verschluesseln-und-loeschen/Daten-endgueltig-loeschen/daten-endgueltig-loeschen_node.html
% https://www.experte.de/it-sicherheit/festplatte-loeschen
% https://www.easeus.de/festplatte-wiederherstellen/formatierte-festplatte-wiederherstellen-freeware.html

\end{document}

### 2022-06-02 hardware-sicherheit/fido2.tex
Finish the webauthn / ctap page and begin to document the server

@@
 dementsprechend an und führen zur Wiederverwendung von gleichen Login-Daten für
-mehrere Online-Dienste. Bekannte Lösungen sind die Verwendung von sog.
+mehrere Online-Dienste. Bekannte Lösungen sind die Verwendung von sogenannten
 Passwort-Managern um lange und zufällige Passwörter für verschiedenste
@@
 
+\textcolor{red}{CTAP Spec references}
+
 Die im Standard beschriebenen Transportlayer umfassen USB, NFC oder Bluetooth
@@
 sicheren Verbindungsaufbau und zur Nachrichtenstruktur.
+\textcolor{red}{CTAP Spec references}
+
+Die JSON Objekte werden innerhalb von \textit{CTAP} mittels dem Datenformat
+\textit{CBOR} dargestellt um eine effizientere Codierung zu erhalten und
+dennoch die kompatiblität mit dem weitverbreiteten Standard für
+Datenserialisierung zu garantieren.
+\textcolor{red}{CBOR references}
 
@@
 
+Ein Server der das WebAuthn Protokoll implementiert enthält im wesentlichen
+zwei grundlegende Operationen die jeweils aus mehreren Schritten bestehen.
+Diese sind die Registrierung von Sicherheitsschlüsseln (Credentials) und deren
+zuordnung zu einem Nutzeraccount im System der Relying Party und die
+Austellung beziehungsweise Validierung von WebAuthn Challanges die bei
+erfolgreicher Validierung einem Nutzer zugriff auf das System der Relying Party
+gibt.
+
+Ein wichtiger Punkt im Zusammenspiel der Spezifikationen und folgich der
+Implementierungen der beiden Protokolle \textit{CTAP} und \textit{WebAuthn} ist
+die kompatiblität der Datenttypen. So lässt sich die spezifizierte Ausgabe
+einer WebAuthn Registrierungs Challange ohne veränderung mit der
+``authenticatorMakeCredential'' Operation an den Authenticator weitergeben.
+
+Dies hat zur Folge dass die Client-Implementierung selbst keine
+protokollrelevanten kryptografischen Berechnungen durchführen muss und effektiv
+der Server der Relying Party mit dem Authenticator des Nutzers kommuniziert.
+Die Client-Implementierung leitet lediglich jeweils die Ausgaben weiter.
+
+\textcolor{red}{weubauthn spec references}
+
+\subsubsection{Registrierung}
+
+Um die Registrierung eines neuen Schlüssels durchzuführen muss die Relying
+Party auf dem Server \texttt{PublicKeyCredentialCreationOptions} für den Nutzer
+generieren und an den Client zurücksenden. Diese enthalten Informationen wie
+zum Beispiel erlaubte kryptografische Algorithmen, Informationen über den
+Nutzer und die Relying Party, ausgeschlossene Authenticator etc.
+
+Diese Parameter nimmt der Client entgegen und gibt diese dann anschließend an
+die \textit{CTAP}-Operation ``authenticatorMakeCredential'' weiter. Der
+Authenticator überprüft anschließend ob er ein Schlüsselpaar erstellen kann
+dass den Anforderungen des Servers entspricht und gibt entweder einen Fehler
+oder den öffentlichen Schlüssel an den Client zurück. Dieser wird nun vom
+Client an den Server weitergeben und die Registrierung wurde erfolgreich
+abgeschlossen.
+
+\textcolor{red}{weubauthn spec references}
+
+\subsubsection{Authentifizierung}
+
+Für die Authentifizierung eines Nutzers mittels WebAuthn muss zu erst die
+Relying Party eine Challange für den Client generieren. Diese muss zufällig und
+nur einmal gültig sein um Replay-Attacken zu verhindern. Nach dem der Client
+seine Challange erhalten hat kann er diese über die CTAP-Operation
+``authenticatorGetAssertion'' signieren lassen und die signierte Challange
+wieder an den Server zurückgeben. Anschließend überprüft dieser ob einer der
+für den Nutzer hinterlegten Schlüssel verwendet wurde um die Challange zu
+signieren. War dies der Fall gilt die Identität des Nutzers als verifiziert und
+der Server kann dem Nutzer Zugriff auf geschützte Ressourcen geben.
+
+\textcolor{red}{weubauthn spec references}
+
+\vspace{0.25em}
+\rule{0.45\textwidth}{0.4pt}
+\vspace{0.5em}
+
+Je nach Kommunikationsprotokoll zwischen dem Client und dem Server muss sich
+der Server den Status der Registrierung bzw. Authentifizierung über einzelne
+Nachrichten hinweg merken. So sind (auf Nachrichtenebene) verbindungslose
+Protokolle wie bspw. HTTPS davon betroffen. Eine bidirektionaler Transport
+würde dieses Problem umgehen, da sich der Server nur den Status für die Dauer
+der Verbindung merken müsste.
+
 \section{Implementierung des Authetifizierungs-Servers}
+
+\subsection{Zielsetzung}
+
+Der Umfang der Proof-of-Concept Implementierung ist es eine funktionale,
+flüchtige User-Verwaltung die optional FIDO2 Schlüssel als zweiten
+Login-Faktor unterstützt. Als API wird ein HTTP Server geöffnet.
+
+\subsection{Authentifizierungsmechanismus}
+
+Nutzer ohne FIDO2 Schlüssel können sich bei dem Server mittels der Kombination
+aus E-Mail und ihrem Passwort authentifizieren und erhalten sofort einen Token
+zurück.
+
+Die Nutzer die einen FIDO2 Schlüssel registriert haben erhalten nach
+initialer Angabe ihrer E-Mail und ihres Passworts eine Aufforderung ihren FIDO2
+Schlüssel zu verwenden um letztendlich auch einen Token zu erhalten.
+
+Der Token ist ein JSON-Web-Token der eine User-Id enthält und ein Ablaufdatum.
+Alle Tokens werden mit dem symmetrischen Algorithmus HMAC (Hash-based Message
+Authentication Code) signiert.
+
+Alle geschützten beziehungsweise nutzerbezogenen Ressourcen die dieser Server
+vorhält dürfen nur durch die Mitgabe eines Tokens abgerufen werden. Dieser muss
+mit dem \texttt{Authorization}-Header gesetzt werden werden.
+
+\subsection{Speicherung der Nutzer-Passwörter}
+
+Da bekanntermaßen in Klartext gespeicherte Passwörter eine verherende
+Sicherheitslücke darstellen speichert der Server nur den Bcrypt-Hash der
+Passwörter die durch einen sogenannten Salt randomisiert werden.
+
+Sowohl der Schlüssel zur signierung von Tokens als auch der Salt für Bcrypt
+müssen in einem realen Anwendungsszenario geheim gehalten werden.
 
@@
 
-    %app.at("/auth/login").post(auth::login);
-    %app.at("/auth/signup").post(auth::signup);
-    %app.at("/auth/verify")
-        %.with(middleware::authenticated)
-        %.post(auth::verify);
-
-    %app.at("/auth/fido2/login").post(auth::fido2::login);
-    %app.at("/auth/fido2/challenges")
-        %.with(middleware::authenticated)
-        %.post(auth::fido2::create_challenge);
-    %app.at("/auth/fido2/keys")
-        %.with(middleware::authenticated)
-        %.post(auth::fido2::store_key);
-    %app.at("/auth/fido2/keys/:id")
-        %.with(middleware::authenticated)
-        %.delete(auth::fido2::remove_key);
-
 \section{Auswertung}

### 2022-06-02 hardware-sicherheit/fido2.tex
Add more text regarding the implementation details

@@
 Der Umfang der Proof-of-Concept Implementierung ist es eine funktionale,
-flüchtige User-Verwaltung die optional FIDO2 Schlüssel als zweiten
+flüchtige Nutzer-Verwaltung die optional FIDO2 Schlüssel als zweiten
 Login-Faktor unterstützt. Als API wird ein HTTP Server geöffnet.
@@
 
-Der Token ist ein JSON-Web-Token der eine User-Id enthält und ein Ablaufdatum.
-Alle Tokens werden mit dem symmetrischen Algorithmus HMAC (Hash-based Message
-Authentication Code) signiert.
+Der Token ist ein JSON-Web-Token der eine Nutzer-Id enthält und ein
+Ablaufdatum. Alle Tokens werden mit dem symmetrischen Algorithmus HMAC
+(Hash-based Message Authentication Code) signiert.
 
@@
 müssen in einem realen Anwendungsszenario geheim gehalten werden.
+
+\subsection{Datenstrukturen zur flüchtigen Verwaltung von Nutzern}
+
+Ein Nutzer erhält bei der Registrierung eine eindeutige Id, einen
+Verifizierungscode (der in einem realen Anwendungsszenario ggf. per E-Mail
+versendet werden könnte) und eine leere Liste an FIDO2 Schlüsseln.
+
+Die Nutzer werden in einer HashMap im RAM vorgehalten, eine Datenbankanbindung
+des Servers zur Nutzerverwaltung wären bei einem Produktivsystem unabdingbar.
+
+Um die konsistzenz der Nutzerverwaltung innerhalb des RAM sicherzustellen
+befindet sich die Nutzer-HashMap in einem Mutex um gleichzeitige ggf.
+inkonsistente Schreibvorgänge zu unterbinden.
+
+Bei jeder eingehenden Anfrage wird die Datenstrukturen zur Nutzerverwaltung der
+Anfrage zugewiesen, diese hat nun die Garantie dass sie die einzige Anfrage ist
+die in diesem Moment schreiben kann. Sollten mehrere Anfragen gleichzeitig
+eingehen muss eine warten bis die andere abgearbeitet wurde. Dies ist zwar
+absolut inperformant bei großen Anfragemengen aber bei einem
+Proof-of-Concept-Server mit sehr wenigen Nutzern vertretbar.
+
+\subsection{Implementierung von WebAuthn}
+
+Da die Erstellung von den durch WebAuthn definierten Datenstrukturen und die
+kryptografischen Algorithmen nicht trivial zu implementieren sind gibt es für
+gängige Sprachen die eine Anwendung im Web-Kontext finden Bibliotheken die das
+WebAuthn-Protokoll abstrahiert bereitstellen.
+
+Für Rust ist ``webauthn-rs'' eine der umfangreicheren WebAuthn-Bibliotheken 
+\textcolor{red}{https://github.com/kanidm/webauthn-rs}
 

### 2022-06-04 hardware-sicherheit/fido2.tex
Fix typos and do minor restructuring

@@
 
-Durch all diese Eigenschaften werden Schlsselverluste unwahrscheinlicher und
+Durch all diese Eigenschaften werden Schlüsselverluste unwahrscheinlicher und
 weniger Sicherheitskritisch, da selbst bei einem hypothetischen
@@
 Die größte Schwachstelle ist allerdings der physische Diebstahl des
-Hardware-Tokens, da dessen Besitz ausreicht für Impersonation-Attacken. Eine
+Hardware-Tokens, da dessen Besitz ausreicht für Impersonation-Attacken. Als
 Absicherung dagegen kann eine biometrische Authentifizierung erfolgen bevor ein
-privater Schlüssel verwendet werden kann - wie beispielsweise bei FaceID.
+privater Schlüssel verwendet werden kann - wie beispielsweise bei FaceID.\
+
 \textcolor{red}{tbd}
@@
 Hardware-Token gespeichert wird und der Beweis der Identität durch die Signatur
-einer vom Authentifizierungs-Server ausgestellten Challange erfolgt die das
+einer vom Authentifizierungs-Server ausgestellten Challenge erfolgt die das
 \textcolor{red}{TPM} oder der Hardware-Token intern durchführen und dem Nutzer
@@
 		``Relying Party'' erstellen
-	\item \textit{authenticatorGetAssertion} - Signatur einer Challange
+	\item \textit{authenticatorGetAssertion} - Signatur einer Challenge
 	\item \textit{authenticatorGetNextAssertion} - Nächste Signatur der
-		Challange erhalten bei mehreren Schlüsselpaaren
+		Challenge erhalten bei mehreren Schlüsselpaaren
 	\item \textit{authenticatorGetInfo} - Informationen über die Fähigkeiten
@@
 zuordnung zu einem Nutzeraccount im System der Relying Party und die
-Austellung beziehungsweise Validierung von WebAuthn Challanges die bei
+Austellung beziehungsweise Validierung von WebAuthn Challenges die bei
 erfolgreicher Validierung einem Nutzer zugriff auf das System der Relying Party
@@
 die kompatiblität der Datenttypen. So lässt sich die spezifizierte Ausgabe
-einer WebAuthn Registrierungs Challange ohne veränderung mit der
+einer WebAuthn Registrierungs Challenge ohne veränderung mit der
 ``authenticatorMakeCredential'' Operation an den Authenticator weitergeben.
@@
 Für die Authentifizierung eines Nutzers mittels WebAuthn muss zu erst die
-Relying Party eine Challange für den Client generieren. Diese muss zufällig und
+Relying Party eine Challenge für den Client generieren. Diese muss zufällig und
 nur einmal gültig sein um Replay-Attacken zu verhindern. Nach dem der Client
-seine Challange erhalten hat kann er diese über die CTAP-Operation
-``authenticatorGetAssertion'' signieren lassen und die signierte Challange
+seine Challenge erhalten hat kann er diese über die CTAP-Operation
+``authenticatorGetAssertion'' signieren lassen und die signierte Challenge
 wieder an den Server zurückgeben. Anschließend überprüft dieser ob einer der
-für den Nutzer hinterlegten Schlüssel verwendet wurde um die Challange zu
+für den Nutzer hinterlegten Schlüssel verwendet wurde um die Challenge zu
 signieren. War dies der Fall gilt die Identität des Nutzers als verifiziert und
@@
 
-Der Umfang der Proof-of-Concept Implementierung ist es eine funktionale,
+Ziel der Proof-of-Concept Implementierung ist es eine funktionale,
 flüchtige Nutzer-Verwaltung die optional FIDO2 Schlüssel als zweiten
-Login-Faktor unterstützt. Als API wird ein HTTP Server geöffnet.
+Login-Faktor unterstützt. Diese soll eine API anbieten auf der eine
+Client-Implementierung aufsetzen kann.
 
@@
 Ein Nutzer erhält bei der Registrierung eine eindeutige Id, einen
-Verifizierungscode (der in einem realen Anwendungsszenario ggf. per E-Mail
+Verifizierungscode (der in einem realen Anwendungsszenario ggf.\ per E-Mail
 versendet werden könnte) und eine leere Liste an FIDO2 Schlüsseln.

### 2022-06-04 hardware-sicherheit/fido2.tex
Add images and a section about the client implementation

@@
 \usepackage{hyperref}
+\usepackage{graphicx}
 \usepackage{lipsum}
 \usepackage{xcolor}
+
+\graphicspath{ {./images/} }
 
@@
 
-\textcolor{red}{grafik}
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{ctap-webauthn-registration.png}
+	\centering
+	\caption{Unterteilung des Kommunikationsablaufs in CTAP und WebAuthn am
+	beispiel einer Schlüssel-Registrierung}\label{fig:ctap-webauthn}
+\end{figure}
 
@@
 
-Für Rust ist ``webauthn-rs'' eine der umfangreicheren WebAuthn-Bibliotheken 
-\textcolor{red}{https://github.com/kanidm/webauthn-rs}
+Für Rust ist eine umfangreicherere dieser Bibliotheken ``webauthn-rs'',
+verfügbar auf GitHub unter der Mozilla-Public-License-2.0:
 
-\section{Technische Dokumentation}
+\url{https://github.com/kanidm/webauthn-rs}
+
+Die Bibliothek lässt sich durch Informationen über den Server konfigurieren:
+Es wird eine Relying Party Id und eine registrierbare Domain erwartet um die
+Identität des Servers für die Schlüsselerstellung wiederverwenden zu können.
+
+Nach erfolgreicher Initialisierung gibt es im wesentlichen 4 relevante
+Funktionen mit denen sich der WebAuthn-Flow komplett implementieren lässt:
+
+\subsubsection{start\_securitykey\_registration}
+Gibt die Public-Key-Credential-Creation-Options und einen
+Schlüssel-Registrierungsstatus für den momentanen Nutzer zurück. Dieser muss
+persistiert werden.
+
+\subsubsection{finish\_securitykey\_registration}
+Nimmt einen Schlüssel-Registrierungsstatus und den generierten Schlüssel und
+validiert die Gültigkeit des generierten Schlüssels.
+
+\subsubsection{start\_securitykey\_authentication}
+Gibt eine Challenge und einen Schlüssel-Authetifizierungsstatus für den
+momentanen Nutzer zurück. Dieser muss muss persistiert werden.
+
+\subsubsection{finish\_securitykey\_authentication}
+Nimmt eine Schlüssel-Authetifizierungsstatus und eine signierte Challenge und
+validiert diese gegeninander.
+
+Somit sind Serverseitig alle Voraussetzungen geschaffen um eine API
+bereitzustellen die WebAuthn unterstützt. Detailierte Informationen zu den
+bereitgestellten Endpunkten finden sich unter Abschnitt~\ref{docs}.
+
+\section{Implementierung der Nutzer-Oberfläche}
+
+\subsection{Zielsetzung}
+
+Ziel der Client-Implementierung ist es beispielhaftes Nutzer-Oberfläche zu
+schaffen das die unter Abschnitt~\ref{docs} beschriebene API verwendet um ein
+Nutzer-Login mit FIDO2 zu ermöglichen.
+
+\subsection{Konzept der Oberfläche}
+
+Die Oberfläche ist eine sehr kleine JavaScript Anwendung für die API.\ Folglich
+orientieren sich die Aktionen die der Nutzer in diesem ausführen kann direkt an
+dieser. In der Abbildung~\ref{fig:ui-flow} sind der Ablauf aller möglichen
+Nutzeraktionen und die dazugehörigen API Endpunkte abgebildet.
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{ui-flow.png}
+	\centering
+	\caption{Abfolge möglicher Nutzer-Aktionen}\label{fig:ui-flow}
+\end{figure}
+
+\subsection{Anwendungs Architektur}
+
+Da der Umfang der Nutzer-Aktionen sich stark in Grenzen hält gibt es keinen
+Grund eines der komplexeren Frameworks einzusetzen die typischerweise für
+Web-Oberflächen eingesetzt werden. Die einzige externe Bibliothek ist eine
+moderene und stabilere Base64 Implementierung als die Browser-Native
+Alternative.
+
+Das Grundkonzept zum rendering der Oberfläche ist einen Anwendungszustand durch
+eine pure (also Seiten-Effekt freie) Funktion in HTML zu übersetzen.
+Nutzeraktionen können diesen Zustand durch Interaktion verändern und lösen
+somit einen neuen Render-Vorgang aus. Diese bzw. ähnliche Architekturen findet
+sich in Frameworks wie React oder Elm wieder.
+
+\textcolor{red}{refs}
+
+Der Zustand speichert den Token, Verifizierungsstatus, Ladezustand der
+Anwendung und die Liste aller registrierten Schlüssel.
+
+\subsection{Interaktionen}
+
+Die definierten möglichen Interaktionen die der Nutzer tätigen kann sind:
+
+\begin{enumerate}
+	\item signup
+	\item verify
+	\item login
+	\item logout
+	\item addKey
+	\item removeKey
+\end{enumerate}
+
+Diese werden durch die entsprechenden Bedienfelder der Oberfläche ausgelöst.
+
+\section{Technische Dokumentation}\label{docs}
 
@@
 	\item[Eingabe:] Credentials \{ email, password \}
-	\item[Ausgabe:] UserDetails \{ token, verified, keys \} | \textcolor{red}{webauthn challange}
+	\item[Ausgabe:] UserDetails \{ token, verified, keys \} | \textcolor{red}{webauthn challenge}
 	\item[Beschreibung:] Gibt dem Nutzer entweder seine UserDetails zurück oder
-		stellt eine WebAuthn Authetifizierungs-Challange die der Nutzer
+		stellt eine WebAuthn Authetifizierungs-Challenge die der Nutzer
 		signiert bei dem Endpunkt \textit{/auth/fido2/login} enreichen muss
@@
 	\item[Token:] -
-	\item[Eingabe:] \textcolor{red}{challange}
+	\item[Eingabe:] \textcolor{red}{challenge}
 	\item[Ausgabe:] UserDetails \{ token, verified, keys \}
-	\item[Beschreibung:] Validiert die WebAuthn Challange des Nutzers mit den
+	\item[Beschreibung:] Validiert die WebAuthn Challenge des Nutzers mit den
 		hinterlegten FIDO2 Schlüsseln und gibt bei erfolgreicher Validierung
@@
 
-\subsection{/auth/fido2/challenges - WebAuthn Registrierungs Challange}
+\subsection{/auth/fido2/challenges - WebAuthn Registrierungs Challenge}
 
@@
 		Schlüssel. Setzt Serverseitig den ``Key-Registration-State'' eines
-		Nutzers und gibt eine Registrierungs-Challange zurück.
+		Nutzers und gibt eine Registrierungs-Challenge zurück.
 \end{itemize}
@@
 
-\section{Abbildungsverzeichnis}
+\subsection{Nutzerbarkeit}
+
+Nutzbarkeit abgesehen von der Verifizierungsmethode, die aus dem Terminal
+abgelesen werden muss ist gut.
+
+\subsection{Risiken}
+
+- Fehlender Zwang FIDO2 zu verwenden
+
+\subsection{Verbesserungsvorschläge}
+
+- FIDO2 Key als hauptfaktor, backup codes und long lived access
+
+\listoffigures
 
@@
 
+\section*{Anhang}
+
+\subsection*{Quellcode der Implementierung}
+
+\url{https://github.com/mara214/fido2-auth}
+
 \end{document}

### 2022-06-04 hardware-sicherheit/fido2.tex
Add a section for analysis of the risks

@@
 
+\subsection{Implementierung von CTAP}
+
+Da die Nutzer-Oberfläche in JavaScript geschrieben ist und dazu gedacht ist in
+einem Browser-Kontext ausgeführt zu werden kann die Browser-Implementierung für
+CTAP verwendet werden was die komplexität der Anwendung drastisch veringert, da
+diese Implementierung sehr stark abstrahiert ist.
+
+Im wesentlichen werden nur zwei der bereitgestellten Funktionen in der gesamten
+Nutzer-Oberfläche verwendet:
+
+\begin{enumerate}
+	\item navigator.credential.create (führt die Operation
+		``authenticatorMakeCredential'' aus)
+	\item navigator.credential.get (führt die Operation
+		``authenticatorGetAssertion'' aus)
+\end{enumerate}
+
+Da die Parameter dieser Funktionen von dem Server bereitgestellt werden ist die
+Implementierung der Kommunikation mit dem Authenticator also sehr trivial.
+
 \section{Technische Dokumentation}\label{docs}
@@
 
-\subsection{Nutzerbarkeit}
+\subsection{Nutzbarkeit}
 
-Nutzbarkeit abgesehen von der Verifizierungsmethode, die aus dem Terminal
-abgelesen werden muss ist gut.
+Das System ist unter Einschränkungen voll funktional und bietet eine
+verhältnismäßig gute Nutzbarkeit. Durch die Verwendung eines flüchtigen
+Speichermediums (RAM) für die Nutzerverwaltung sind nach jedem Neustart alle
+Anwendungsdaten zurückgesetzt. Dies ist für Test-Zwecke absolut in Ordnung, und
+ließe sich für eine reale Anwendung leicht durch die Implementierung einer
+Datenbankanbindung beheben.
+
+Die Nutzer-Oberfläche ist sehr schlicht, lässt aber den Nutzer übersichtlich
+alle definierten Interaktionen ausführen. Auch hier ließe sich die
+ausgestaltung der Oberfläche für eine reale Anwedung mit der entsprechenden
+Fachexpertise verbessern.
+
+Die Nutzerauthentifizierung an sich verläuft zuverlässig, schnell und
+reibungslos. Neben der allgemeinen Dateneingabe (E-Mail und Passwort) besteht
+die signierung der Challenge lediglich aus einem Tippen auf den Authenticator.
+Dies allein bietet ein deutlich verbessertes Erlebnis im vergleich zu
+SMS-Tokens oder One-Time-Passwords als zweiten Faktor (vorrausgesetzt der
+Authenticator befindet sich in der Nähe des Systems oder ist sogar bereits
+angeschlossen).
 
@@
 
-- Fehlender Zwang FIDO2 zu verwenden
+Die Hauptrisiken des momentanen Systems ist die fehlende Passwortvalidierung
+und der fehlende Zwang einen FIDO Schlüssel bei der Registrierung zu
+hinterlegen. So ist es momentan möglich einen Nutzer mit einem schwachen
+Passwort anzulegen und keinen Schlüssel zu registrieren. Dies ließe sich
+gegebenenfalls auch durch Anpassung des Authentifizierungs bzw.
+Registrierungsprozesses beheben.
 
-\subsection{Verbesserungsvorschläge}
+Unter der Annahme dass eine Schlüssel-Registrierung verpflichtend sei wäre ein
+Angriff auf einen Nutzeraccount nurnoch durch physikalische Entwendung des
+Hardware Tokens möglich und dementsprechend unwahrscheinlich. Somit kann unter
+dieser Annahme das System als sicher angesehen werden.
 
-- FIDO2 Key als hauptfaktor, backup codes und long lived access
+\subsection{Mögliche Verbesserungen}
+
+Als abschließende Verbesserung wäre ein Umstellen auf
+Single-Faktor-Authentifizierung mit dem Hardware-Token hinsichtlich des
+Nutzererlebnises empfehlenswert (Sicherheitsimplikationen außer Acht gelassen).
+
+Eine alternative Möglichkeit könnten weiterhin zwei Faktoren sein, allerdings
+unter der Bedingung, dass das Passwort lediglich zur einrichtung neuer Systeme
+abgefragt wird und sonst ein Single-Faktor-Login mit dem Hardware-Token möglich
+ist.
+
+Desweiteren ist es bei Hardware Tokens - deren physikalischer Verlust nicht
+gänzlich auszuschließen ist - ratsam einen Backup Token zu besitzen und diesen
+bei jedem Online-Dienst als Backup Token zu hinterlegen um einen
+Schlüsselverlust abfangen zu können. Eine bekannte Alternative zu einem zweiten
+Hardware Token wären auch sogenannte Backup Codes, die bei Schlüsselverlust
+einmaligen Zugriff gewähren.
+
+All diese ausgereiften Konzepte werden in der momentanen
+Proof-of-Concept-Implementierung noch außer Acht gelassen, ließen sich
+allerdings leicht mit geringem zusätzlichen Aufwand hinzufügen.
 
@@
 
+\subsection*{Zur Entwicklung verwendeter Authenticator}
+
+\url{https://store.google.com/de/product/titan_security_key}
+
 \end{document}

### 2022-06-04 hardware-sicherheit/fido2.tex
Fix types in the api docs

@@
 Um die Registrierung eines neuen Schlüssels durchzuführen muss die Relying
-Party auf dem Server \texttt{PublicKeyCredentialCreationOptions} für den Nutzer
+Party auf dem Server \texttt{CredentialCreationOptions} für den Nutzer
 generieren und an den Client zurücksenden. Diese enthalten Informationen wie
@@
 \subsubsection{start\_securitykey\_registration}
-Gibt die Public-Key-Credential-Creation-Options und einen
-Schlüssel-Registrierungsstatus für den momentanen Nutzer zurück. Dieser muss
-persistiert werden.
+Gibt die Credential-Creation-Options und einen Schlüssel-Registrierungsstatus
+für den momentanen Nutzer zurück. Dieser muss persistiert werden.
 
@@
 	\item[Eingabe:] Credentials \{ email, password \}
-	\item[Ausgabe:] UserDetails \{ token, verified, keys \} | \textcolor{red}{webauthn challenge}
+	\item[Ausgabe:] UserDetails \{ token, verified, keys \} | CredentialRequestOptions 
 	\item[Beschreibung:] Gibt dem Nutzer entweder seine UserDetails zurück oder
@@
 	\item[Token:] -
-	\item[Eingabe:] \textcolor{red}{challenge}
+	\item[Eingabe:] PublicKeyCredential
 	\item[Ausgabe:] UserDetails \{ token, verified, keys \}
@@
 	\item[Eingabe:] -
-	\item[Ausgabe:] \textcolor{red}{ccr}
+	\item[Ausgabe:] CredentialCreationOptions 
 	\item[Beschreibung:] Startet einen Registrierungsprozess für einen FIDO2
 		Schlüssel. Setzt Serverseitig den ``Key-Registration-State'' eines
-		Nutzers und gibt eine Registrierungs-Challenge zurück.
+		Nutzers und gibt die Registrierungs-Optionen inklusive einer
+		Registrierungs-Challenge zurück.
 \end{itemize}
@@
 	\item[Token:] Notwendig
-	\item[Eingabe:] \textcolor{red}{PublicKeyCredential}
+	\item[Eingabe:] RegisterPublicKeyCredential 
 	\item[Ausgabe:] Key \{ id \}
@@
 reibungslos. Neben der allgemeinen Dateneingabe (E-Mail und Passwort) besteht
-die signierung der Challenge lediglich aus einem Tippen auf den Authenticator.
+die Signierung der Challenge lediglich aus einem Tippen auf den Authenticator.
 Dies allein bietet ein deutlich verbessertes Erlebnis im vergleich zu

### 2022-06-04 hardware-sicherheit/fido2.tex
Add an abstract

@@
 \begin{abstract}
+	Durch die Implementierung von FIDO2 können bestehende Systeme hinsichtlich
+	ihrer Nutzbarkeit und Sicherheit stark verbessert werden und neue Systeme
+	ganz neue Authentifizierungskonzepte implementieren. Die steigende Relevanz
+	zeigt sich durch die Unterstützung großer Firmen. \textcolor{red}{ref}
+	Durch die Verwendung von ausgereiften Abstraktionen über die Protokolle
+	\textit{CTAP} und \textit{WebAuthn} sind schnelle und zuverlässige
+	Implementierungen des Standards leicht realisierbar.
 \end{abstract}
@@
 basieren auf der Annahme, dass das System des Nutzers nicht kompromitiert
-wurde. \textcolor{red}{tbd}
+wurde. Als eine mögliche Lösung für viele dieser Probleme hat sich FIDO
+innerhalb der letzten Jahre zu einem beliebten Standard entwickelt der von
+vielen größeren Anbietern wie Facebook, Google, Apple, Mozilla, Microsoft etc.
+bereits untersützt beziehungsweise implementiert wird. \textcolor{red}{refs}
 

### 2022-06-04 hardware-sicherheit/fido2.tex
Remove reminders

--- 2022-06-04 hardware-sicherheit/fido2.tex
@@
 
-\title{Implementierung eines Atuhentifizierungs-Servers mit FIDO2 Support}
+\title{Proof-of-Concept-Implementierung einer Anwendung mit FIDO2 Support}
 \author{
@@
 \begin{abstract}
+	Durch die Implementierung von FIDO2 können bestehende Systeme hinsichtlich
+	ihrer Nutzbarkeit und Sicherheit stark verbessert werden und neue Systeme
+	grundlegend neue Authentifizierungskonzepte implementieren. Die steigende
+	Relevanz zeigt sich durch die Unterstützung großer Firmen.
+	\textcolor{red}{ref} Durch die Verwendung von ausgereiften Abstraktionen
+	über die Protokolle \textit{CTAP} und \textit{WebAuthn} sind schnelle und
+	zuverlässige Implementierungen des Standards leicht realisierbar.
 \end{abstract}
@@
 
-\textcolor{red}{tbd}
-
 \subsection{Welche Probleme löst FIDO2?}
@@
 Die Notwendigkeit für einen solchen Standard hat sich in den letzten Jahren
-immer stärker gezeigt, da die klassische Knowledge-Based-Authentication
-(\textcolor{red}{KBA})
+immer stärker gezeigt, da die klassische Knowledge-Based-Authentication (KBA)
 durch zunehmende Rechenleistung und effizientere Angriffe immer unsicherer und
@@
 Passwort-Manager sind zwar eine Lösung für die sichere Aufbewahrung von langen
-Passwörtern, können aber nicht die durch \textcolor{red}{KBA} eröffneten
-Angriffsvektoren wie z.B. Man-In-The-Middle-Attacken. So bald ein Angreifer
-in den Besitz des geheimen Wissens (in diesem Fall das Passwort) gelangt kann
-dieser uneingeschränkt und unbegegrenzt oft auf das Zielsystem zugreifen, bis
-der Nutzer seine Daten ändert (vorausgesetzt, der Angreifer hat dies noch nicht
+Passwörtern, können aber nicht die durch KBA eröffneten Angriffsvektoren wie
+z.B. Man-In-The-Middle-Attacken. So bald ein Angreifer in den Besitz des
+geheimen Wissens (in diesem Fall das Passwort) gelangt kann dieser
+uneingeschränkt und unbegegrenzt oft auf das Zielsystem zugreifen, bis der
+Nutzer seine Daten ändert (vorausgesetzt, der Angreifer hat dies noch nicht
 getan).
 
-Durch den Wechsel von \textcolor{red}{KBA} auf Zero-Knowledge-Proof basierte
+Durch den Wechsel von KBA auf Zero-Knowledge-Proof basierte
 Authentifizierungsmethoden lassen sich ganze Angriffsvektoren ausschließen, da
@@
 Im Falle von FIDO2 kennt nichtmal der Nutzer selber seine Schlüssel da diese
-auf einem Trusted-Plattform-Modul \textcolor{red}{TPM} oder einem externen
-Hardware-Token gespeichert wird und der Beweis der Identität durch die Signatur
-einer vom Authentifizierungs-Server ausgestellten Challenge erfolgt die das
-\textcolor{red}{TPM} oder der Hardware-Token intern durchführen und dem Nutzer
-nur die Signatur zurückgeben. So stellt selbst ein kompromitiertes Nutzersystem
-nur eine temporäre Schwachstelle dar.
+auf einem Trusted-Plattform-Modul (TPM) oder einem externen Hardware-Token
+gespeichert wird und der Beweis der Identität durch die Signatur einer vom
+Authentifizierungs-Server ausgestellten Challenge erfolgt die das TPM oder der
+Hardware-Token intern durchführen und dem Nutzer nur die Signatur zurückgeben.
+So stellt selbst ein kompromitiertes Nutzersystem nur eine temporäre
+Schwachstelle dar.
 

### 2022-06-04 hardware-sicherheit/fido2.tex
Update introduction text

@@
 innerhalb der letzten Jahre zu einem beliebten Standard entwickelt der von
-vielen größeren Anbietern wie Facebook, Google, Apple, Mozilla, Microsoft etc.
-bereits untersützt beziehungsweise implementiert wird. \textcolor{red}{refs}
+vielen großen Anbietern wie Facebook, Amazon, Apple, Google, Mozilla,
+Microsoft etc.\ bereits untersützt beziehungsweise implementiert wird.
+\textcolor{red}{refs}
 
@@
 der FIDO Alliance entwickleter, offener und lizenzfreier Standard für
-Hardware-Token gestützte Authentifizierung. (\textcolor{red}{quote})
+Hardware-Token gestützte Authentifizierung.
 
-Ein Hardware-Token (kann auch in Form eines Trusted-Plattform-Moduls oder als
-Teil des Betriebssystems implementiert sein) ist ein physischer Speicher für
-die FIDO Schlüsselpaare eines Nutzers. Kernmerkmale die FIDO2 von
+Ein Hardware-Token (kann auch in Form eines Trusted-Plattform-Moduls (TPM) oder
+als Teil des Betriebssystems implementiert sein) ist ein physischer Speicher
+für die FIDO Schlüsselpaare eines Nutzers. Kernmerkmale die FIDO2 von
 herkömmlicher asymmetrischer Kryptografie unterscheiden sind beispielsweise die
@@
 Die Notwendigkeit für einen solchen Standard hat sich in den letzten Jahren
-immer stärker gezeigt, da die klassische Knowledge-Based-Authentication (KBA)
-durch zunehmende Rechenleistung und effizientere Angriffe immer unsicherer und
+immer stärker gezeigt, da die klassische Passwort-Authentifizierung durch
+zunehmende Rechenleistung und effizientere Angriffe immer unsicherer und
 unhandlicher für Nutzer wird. Die minimale Passwortlängen steigen
@@
 
-Durch den Wechsel von KBA auf Zero-Knowledge-Proof basierte
-Authentifizierungsmethoden lassen sich ganze Angriffsvektoren ausschließen, da
-ein kompromitierter Server oder eine kompromitierte Verbindung niemals das
+Durch den Wechsel von Passwort-Authentifizierung auf Zero-Knowledge-Proof
+basierte Authentifizierungsmethoden (in diesem Fall
+Public-Key-Authentifizierung) lassen sich ganze Angriffsvektoren ausschließen,
+da ein kompromitierter Server oder eine kompromitierte Verbindung niemals das
 geheime Wissen des Nutzers einem Angreifer zugänglich machen. Das heißt, dass
@@
 Im Falle von FIDO2 kennt nichtmal der Nutzer selber seine Schlüssel da diese
-auf einem Trusted-Plattform-Modul (TPM) oder einem externen Hardware-Token
+auf einem Trusted-Plattform-Modul oder einem externen Hardware-Token
 gespeichert wird und der Beweis der Identität durch die Signatur einer vom
@@
 und \textit{WebAuthn}. Diese unterteilen den gesammten
-Authentifizierungsvorgang in
-zwei Bereiche:
+Authentifizierungsvorgang in zwei Bereiche:
 
 \textit{Client zu Authenticator} also die Kommunikation zwischen dem
-Nutzersystem und dem Hardware-Token oder TPM und \textit{Client zu Server}, die
+Nutzersystem und dem Hardware-Token und \textit{Client zu Server}, die
 Kommunikation zwischen dem Nutzersystem und dem Online-Dienst bei dem sich der

### 2022-06-04 hardware-sicherheit/fido2.tex
Add proper references

@@
 	grundlegend neue Authentifizierungskonzepte implementieren. Die steigende
-	Relevanz zeigt sich durch die Unterstützung großer Firmen.
-	\textcolor{red}{ref} Durch die Verwendung von ausgereiften Abstraktionen
-	über die Protokolle \textit{CTAP} und \textit{WebAuthn} sind schnelle und
-	zuverlässige Implementierungen des Standards leicht realisierbar.
+	Relevanz zeigt sich durch die Unterstützung großer
+	Firmen.~\cite{fidomembers} Durch die Verwendung von ausgereiften
+	Abstraktionen über die Protokolle \textit{CTAP} und \textit{WebAuthn} sind
+	schnelle und zuverlässige Implementierungen des Standards leicht
+	realisierbar.
 \end{abstract}
@@
 vielen großen Anbietern wie Facebook, Amazon, Apple, Google, Mozilla,
-Microsoft etc.\ bereits untersützt beziehungsweise implementiert wird.
-\textcolor{red}{refs}
+Microsoft etc.\ bereits untersützt beziehungsweise implementiert
+wird.~\cite{fidomembers}
 
@@
 einer Nutzerinteraktion zum Verwenden eines privaten Schlüssels und die
-Generierung von einem Schlüsselpaar pro Online-Dienst.
+Generierung von einem Schlüsselpaar pro Online-Dienst.~\cite{ctapspec}
 
@@
 weniger Sicherheitskritisch, da selbst bei einem hypothetischen
-Schlüsselverlustes der Schaden immer auf einen Online-Dienst begrenzt ist.
-Die größte Schwachstelle ist allerdings der physische Diebstahl des
+Schlüsselverlustes der Schaden immer auf einen Online-Dienst begrenzt ist. Die
+größte Schwachstelle ist allerdings der physische Diebstahl des
 Hardware-Tokens, da dessen Besitz ausreicht für Impersonation-Attacken. Als
@@
 Passwort-Manager sind zwar eine Lösung für die sichere Aufbewahrung von langen
-Passwörtern, können aber nicht die durch KBA eröffneten Angriffsvektoren wie
-z.B. Man-In-The-Middle-Attacken. So bald ein Angreifer in den Besitz des
-geheimen Wissens (in diesem Fall das Passwort) gelangt kann dieser
-uneingeschränkt und unbegegrenzt oft auf das Zielsystem zugreifen, bis der
-Nutzer seine Daten ändert (vorausgesetzt, der Angreifer hat dies noch nicht
-getan).
+Passwörtern, können aber nicht die grundsätzlich durch
+Passwort-Authentifizierung eröffneten Angriffsvektoren wie z.B.
+Man-In-The-Middle-Attacken. So bald ein Angreifer in den Besitz des geheimen
+Wissens (in diesem Fall das Passwort) gelangt kann dieser uneingeschränkt und
+unbegegrenzt oft auf das Zielsystem zugreifen, bis der Nutzer seine Daten
+ändert (vorausgesetzt, der Angreifer hat dies noch nicht getan).
 
@@
 Im Falle von FIDO2 kennt nichtmal der Nutzer selber seine Schlüssel da diese
-auf einem Trusted-Plattform-Modul oder einem externen Hardware-Token
-gespeichert wird und der Beweis der Identität durch die Signatur einer vom
-Authentifizierungs-Server ausgestellten Challenge erfolgt die das TPM oder der
-Hardware-Token intern durchführen und dem Nutzer nur die Signatur zurückgeben.
-So stellt selbst ein kompromitiertes Nutzersystem nur eine temporäre
-Schwachstelle dar.
+auf einem externen Hardware-Token oder TPM gespeichert werden und der Beweis
+der Identität durch die Signatur einer vom Authentifizierungs-Server
+ausgestellten Challenge erfolgt die das TPM oder der Hardware-Token intern
+durchführen und dem Nutzer nur die Signatur zurückgeben. So stellt selbst ein
+kompromitiertes Nutzersystem nur eine temporäre Schwachstelle dar.
 
@@
 Kommunikation zwischen dem Nutzersystem und dem Online-Dienst bei dem sich der
-Nutzer authentifizieren möchte.
+Nutzer authentifizieren möchte.~\cite{ctapspec, webauthnspec}
 
@@
 
-Concise Binary Object Representation, ein Binär-Format zur Darstellung von JSON Objekten.
+Concise Binary Object Representation, ein Binär-Format zur Darstellung von JSON
+Objekten
 
@@
 
-Das Client-To-Authenticator-Protocol kurz \textcolor{red}{CTAP} ist Teil
-des FIDO2 Standards und beschreibt den Ablauf der Kommunikation zwischen einem
-Nutzersystem und dem Hardware-Token beziehungsweise dem Authenticator.
+Das Client-To-Authenticator-Protocol kurz CTAP ist Teil des FIDO2 Standards und
+beschreibt den Ablauf der Kommunikation zwischen einem Nutzersystem und dem
+Hardware-Token beziehungsweise dem Authenticator.
 
@@
 besteht das Protokoll primär aus der sogenannten ``Authenticator API'' - diese
-beschreibt Operationen die ein Nutzersystem auf einem Authenticator ausführen kann.
+beschreibt Operationen die ein Nutzersystem auf einem Authenticator ausführen
+kann.~\cite{ctapspec}
 
@@
 
-\textcolor{red}{CTAP Spec references}
+Siehe~\cite{ctapspec} Abschnitt 5.
 
@@
 Low Energy. Für jeden dieser Transportlayer gibt es eigene Mechanismen zum
-sicheren Verbindungsaufbau und zur Nachrichtenstruktur.
-\textcolor{red}{CTAP Spec references}
+sicheren Verbindungsaufbau und zur Nachrichtenstruktur.~\cite{ctapspec}
 
@@
 dennoch die kompatiblität mit dem weitverbreiteten Standard für
-Datenserialisierung zu garantieren.
-\textcolor{red}{CBOR references}
+Datenserialisierung zu garantieren.~\cite{cborspec}
 
@@
 einer WebAuthn Registrierungs Challenge ohne veränderung mit der
-``authenticatorMakeCredential'' Operation an den Authenticator weitergeben.
+``authenticatorMakeCredential'' Operation an den Authenticator
+weitergeben.~\cite{ctapspec, webauthnspec}
 
@@
 der Server der Relying Party mit dem Authenticator des Nutzers kommuniziert.
-Die Client-Implementierung leitet lediglich jeweils die Ausgaben weiter.
-
-\textcolor{red}{weubauthn spec references}
+Die Client-Implementierung leitet lediglich jeweils die Ausgaben
+weiter.~\cite{webauthnspec}
 
@@
 zum Beispiel erlaubte kryptografische Algorithmen, Informationen über den
-Nutzer und die Relying Party, ausgeschlossene Authenticator etc.
+Nutzer und die Relying Party, ausgeschlossene Authenticator
+etc.~\cite{webauthnspec}
 
@@
 Client an den Server weitergeben und die Registrierung wurde erfolgreich
-abgeschlossen.
-
-\textcolor{red}{weubauthn spec references}
+abgeschlossen.~\cite{ctapspec, webauthnspec}
 
@@
 seine Challenge erhalten hat kann er diese über die CTAP-Operation
-``authenticatorGetAssertion'' signieren lassen und die signierte Challenge
-wieder an den Server zurückgeben. Anschließend überprüft dieser ob einer der
-für den Nutzer hinterlegten Schlüssel verwendet wurde um die Challenge zu
-signieren. War dies der Fall gilt die Identität des Nutzers als verifiziert und
-der Server kann dem Nutzer Zugriff auf geschützte Ressourcen geben.
-
-\textcolor{red}{weubauthn spec references}
+``authenticatorGetAssertion'' signieren lassen~\cite{ctapspec} und die
+signierte Challenge wieder an den Server zurückgeben. Anschließend überprüft
+dieser ob einer der für den Nutzer hinterlegten Schlüssel verwendet wurde um
+die Challenge zu signieren.~\cite{webauthnspec} War dies der Fall gilt die
+Identität des Nutzers als verifiziert und der Server kann dem Nutzer Zugriff
+auf geschützte Ressourcen geben.
 
@@
 
-\textcolor{red}{refs}
+\textcolor{red}{res react elm}
 
@@
 
+\nocite{*}
+
 \bibliographystyle{IEEEtran}

### 2022-06-06 hardware-sicherheit/fido2.tex
Add more references
Fix typos and finish paper

@@
 \IEEEPARstart{A}{\MakeLowercase{uthentifizierung}} ist eines der größten
-Probleme die durch verteilte Systeme entstehen. Es gibt zahlreiche
-Möglichkeiten die Identität einer Gegenseite sicherzustellen allerdings weisen
+Probleme, die durch verteilte Systeme entstehen. Es gibt zahlreiche
+Möglichkeiten, die Identität einer Gegenseite sicherzustellen allerdings weisen
 viele von ihnen Schwachstellen hinsichtlich Man-In-The-Middle-Attacken und
-basieren auf der Annahme, dass das System des Nutzers nicht kompromitiert
+basieren auf der Annahme, dass das System des Nutzers nicht kompromittiert 
 wurde. Als eine mögliche Lösung für viele dieser Probleme hat sich FIDO
-innerhalb der letzten Jahre zu einem beliebten Standard entwickelt der von
+innerhalb der letzten Jahre zu einem beliebten Standard entwickelt, der von
 vielen großen Anbietern wie Facebook, Amazon, Apple, Google, Mozilla,
-Microsoft etc.\ bereits untersützt beziehungsweise implementiert
+Microsoft etc.\ bereits unterstützt beziehungsweise implementiert
 wird.~\cite{fidomembers}
@@
 
-FIDO2 steht für \textit{F}ast \textit{ID}entity \textit{O}nline 2 und ist ein von
-der FIDO Alliance entwickleter, offener und lizenzfreier Standard für
+FIDO2 steht für \textit{F}ast \textit{ID}entity \textit{O}nline 2 und ist ein
+von der FIDO Alliance entwickleter, offener und lizenzfreier Standard für
 Hardware-Token gestützte Authentifizierung.
@@
 als Teil des Betriebssystems implementiert sein) ist ein physischer Speicher
-für die FIDO Schlüsselpaare eines Nutzers. Kernmerkmale die FIDO2 von
-herkömmlicher asymmetrischer Kryptografie unterscheiden sind beispielsweise die
-Isolation der privaten Schlüssel auf dem Hardware-Token, die Notwendigkeit
+für die FIDO Schlüsselpaare eines Nutzers. Kernmerkmale, die FIDO2 von
+herkömmlicher asymmetrischer Kryptografie unterscheiden, sind beispielsweise
+die Isolation der privaten Schlüssel auf dem Hardware-Token, die Notwendigkeit
 einer Nutzerinteraktion zum Verwenden eines privaten Schlüssels und die
@@
 Durch all diese Eigenschaften werden Schlüsselverluste unwahrscheinlicher und
-weniger Sicherheitskritisch, da selbst bei einem hypothetischen
+weniger sicherheitskritisch, da selbst bei einem hypothetischen
 Schlüsselverlustes der Schaden immer auf einen Online-Dienst begrenzt ist. Die
@@
 Hardware-Tokens, da dessen Besitz ausreicht für Impersonation-Attacken. Als
-Absicherung dagegen kann eine biometrische Authentifizierung erfolgen bevor ein
-privater Schlüssel verwendet werden kann - wie beispielsweise bei FaceID.\
+Absicherung dagegen kann eine biometrische Authentifizierung erfolgen, bevor
+ein privater Schlüssel verwendet werden kann - wie beispielsweise bei FaceID.\
 
@@
 zunehmende Rechenleistung und effizientere Angriffe immer unsicherer und
-unhandlicher für Nutzer wird. Die minimale Passwortlängen steigen
+unhandlicher für Nutzer wird. Die minimalen Passwortlängen steigen
 dementsprechend an und führen zur Wiederverwendung von gleichen Login-Daten für
@@
 Passwort-Managern um lange und zufällige Passwörter für verschiedenste
-Online-Dienste zu verwenden ohne, dass sich Nutzer diese merken müssen. Solche
+Online-Dienste zu verwenden, ohne, dass sich Nutzer diese merken müssen. Solche
 Passwort-Manager sind zwar eine Lösung für die sichere Aufbewahrung von langen
 Passwörtern, können aber nicht die grundsätzlich durch
-Passwort-Authentifizierung eröffneten Angriffsvektoren wie z.B.
+Passwort-Authentifizierung eröffneten Angriffsvektoren wie zum Beispiel
 Man-In-The-Middle-Attacken. So bald ein Angreifer in den Besitz des geheimen
-Wissens (in diesem Fall das Passwort) gelangt kann dieser uneingeschränkt und
-unbegegrenzt oft auf das Zielsystem zugreifen, bis der Nutzer seine Daten
+Wissens (in diesem Fall das Passwort) gelangt, kann dieser uneingeschränkt und
+unbegrenzt oft auf das Zielsystem zugreifen, bis der Nutzer seine Daten
 ändert (vorausgesetzt, der Angreifer hat dies noch nicht getan).
@@
 Public-Key-Authentifizierung) lassen sich ganze Angriffsvektoren ausschließen,
-da ein kompromitierter Server oder eine kompromitierte Verbindung niemals das
+da ein kompromittierter Server oder eine kompromitierte Verbindung niemals das
 geheime Wissen des Nutzers einem Angreifer zugänglich machen. Das heißt, dass
-sich ein Angreifer im Falle einer kompromitierten Verbindung maximal in die
+sich ein Angreifer im Falle einer kompromittierten Verbindung maximal in die
 Sitzung des Nutzers einschleichen könnte, allerdings bei der nächsten
@@
 
-Im Falle von FIDO2 kennt nichtmal der Nutzer selber seine Schlüssel da diese
-auf einem externen Hardware-Token oder TPM gespeichert werden und der Beweis
-der Identität durch die Signatur einer vom Authentifizierungs-Server
-ausgestellten Challenge erfolgt die das TPM oder der Hardware-Token intern
+Im Falle von FIDO2 kennt nicht einmal der Nutzer selber seine Schlüssel, da
+diese auf einem externen Hardware-Token oder TPM gespeichert werden und der
+Beweis der Identität durch die Signatur einer vom Authentifizierungs-Server
+ausgestellten Challenge erfolgt, die das TPM oder der Hardware-Token intern
 durchführen und dem Nutzer nur die Signatur zurückgeben. So stellt selbst ein
-kompromitiertes Nutzersystem nur eine temporäre Schwachstelle dar.
+kompromittiertes Nutzersystem nur eine temporäre Schwachstelle dar.
 
@@
 Der FIDO2 Standard umfasst hauptsächlich die beiden Protokolle \textit{CTAP}
-und \textit{WebAuthn}. Diese unterteilen den gesammten
+und \textit{WebAuthn}. Diese unterteilen den gesamten 
 Authentifizierungsvorgang in zwei Bereiche:
@@
 Nutzersystem und dem Hardware-Token und \textit{Client zu Server}, die
-Kommunikation zwischen dem Nutzersystem und dem Online-Dienst bei dem sich der
+Kommunikation zwischen dem Nutzersystem und dem Online-Dienst, bei dem sich der
 Nutzer authentifizieren möchte.~\cite{ctapspec, webauthnspec}
@@
 
-Ein Online-Dienst der den FIDO2 Standard zur Nutzerauthentifizierung verwendet.
+Ein Online-Dienst, der den FIDO2 Standard zur Nutzerauthentifizierung
+verwendet.
 
@@
 
-Ein externer Hardware-Token oder ein Teil des Betriebssystems (z.B. über ein
-TPM implementiert) der FIDO2 Schlüsselpaare verwaltet.
+Ein externer Hardware-Token oder ein Teil des Betriebssystems (zum Beispiel
+über ein TPM implementiert) der FIDO2 Schlüsselpaare verwaltet.
 
@@
 
-Ein FIDO2 Schlüssel
+Ein FIDO2 Schlüssel.
 
@@
 Concise Binary Object Representation, ein Binär-Format zur Darstellung von JSON
-Objekten
+Objekten.
 
@@
 besteht das Protokoll primär aus der sogenannten ``Authenticator API'' - diese
-beschreibt Operationen die ein Nutzersystem auf einem Authenticator ausführen
+beschreibt Operationen, die ein Nutzersystem auf einem Authenticator ausführen
 kann.~\cite{ctapspec}
@@
 	\item \textit{authenticatorGetInfo} - Informationen über die Fähigkeiten
-		des Authenticators
+		des Authenticators abfragen
 	\item \textit{authenticatorClientPIN} - Setzt den Authenticator-PIN
@@
 \textit{CBOR} dargestellt um eine effizientere Codierung zu erhalten und
-dennoch die kompatiblität mit dem weitverbreiteten Standard für
-Datenserialisierung zu garantieren.~\cite{cborspec}
+dennoch die Kompatibilität mit dem weitverbreiteten Standard für
+Datenserialisierung zu garantieren.~\cite{ctap, cborspec}
 
@@
 
-Ein Server der das WebAuthn Protokoll implementiert enthält im wesentlichen
-zwei grundlegende Operationen die jeweils aus mehreren Schritten bestehen.
+Ein Server, der das WebAuthn Protokoll implementiert, enthält im wesentlichen
+zwei grundlegende Operationen, die jeweils aus mehreren Schritten bestehen.
 Diese sind die Registrierung von Sicherheitsschlüsseln (Credentials) und deren
-zuordnung zu einem Nutzeraccount im System der Relying Party und die
-Austellung beziehungsweise Validierung von WebAuthn Challenges die bei
+Zuordnung zu einem Nutzeraccount im System der Relying Party und die
+Ausstellung beziehungsweise Validierung von WebAuthn Challenges die bei
 erfolgreicher Validierung einem Nutzer zugriff auf das System der Relying Party
@@
 
-Ein wichtiger Punkt im Zusammenspiel der Spezifikationen und folgich der
+Ein wichtiger Punkt im Zusammenspiel der Spezifikationen und folglich der
 Implementierungen der beiden Protokolle \textit{CTAP} und \textit{WebAuthn} ist
-die kompatiblität der Datenttypen. So lässt sich die spezifizierte Ausgabe
-einer WebAuthn Registrierungs Challenge ohne veränderung mit der
+die Kompatiblität der Datentypen. So lässt sich die spezifizierte Ausgabe
+einer WebAuthn Registrierungs Challenge ohne Veränderung mit der
 ``authenticatorMakeCredential'' Operation an den Authenticator
@@
 
-Dies hat zur Folge dass die Client-Implementierung selbst keine
+Dies hat zur Folge, dass die Client-Implementierung selbst keine
 protokollrelevanten kryptografischen Berechnungen durchführen muss und effektiv
@@
 
-Um die Registrierung eines neuen Schlüssels durchzuführen muss die Relying
+Um die Registrierung eines neuen Schlüssels durchzuführen, muss die Relying
 Party auf dem Server \texttt{CredentialCreationOptions} für den Nutzer
@@
 die \textit{CTAP}-Operation ``authenticatorMakeCredential'' weiter. Der
-Authenticator überprüft anschließend ob er ein Schlüsselpaar erstellen kann
+Authenticator überprüft anschließend, ob er ein Schlüsselpaar erstellen kann,
 dass den Anforderungen des Servers entspricht und gibt entweder einen Fehler
@@
 
-Für die Authentifizierung eines Nutzers mittels WebAuthn muss zu erst die
+Für die Authentifizierung eines Nutzers mittels WebAuthn muss zuerst die
 Relying Party eine Challenge für den Client generieren. Diese muss zufällig und
-nur einmal gültig sein um Replay-Attacken zu verhindern. Nach dem der Client
-seine Challenge erhalten hat kann er diese über die CTAP-Operation
+nur einmal gültig sein, um Replay-Attacken zu verhindern. Nachdem der Client
+seine Challenge erhalten hat, kann er diese über die CTAP-Operation
 ``authenticatorGetAssertion'' signieren lassen~\cite{ctapspec} und die
 signierte Challenge wieder an den Server zurückgeben. Anschließend überprüft
-dieser ob einer der für den Nutzer hinterlegten Schlüssel verwendet wurde um
-die Challenge zu signieren.~\cite{webauthnspec} War dies der Fall gilt die
+dieser, ob einer der für den Nutzer hinterlegten Schlüssel verwendet wurde, um
+die Challenge zu signieren.~\cite{webauthnspec} War dies der Fall, gilt die
 Identität des Nutzers als verifiziert und der Server kann dem Nutzer Zugriff
@@
 Nachrichten hinweg merken. So sind (auf Nachrichtenebene) verbindungslose
-Protokolle wie bspw. HTTPS davon betroffen. Eine bidirektionaler Transport
+Protokolle wie bspw. HTTPS davon betroffen. Ein bidirektionaler Transport
 würde dieses Problem umgehen, da sich der Server nur den Status für die Dauer
@@
 Ziel der Proof-of-Concept Implementierung ist es eine funktionale,
-flüchtige Nutzer-Verwaltung die optional FIDO2 Schlüssel als zweiten
-Login-Faktor unterstützt. Diese soll eine API anbieten auf der eine
+flüchtige Nutzer-Verwaltung, die optional FIDO2 Schlüssel als zweiten
+Login-Faktor unterstützt. Diese soll eine API anbieten, auf der eine
 Client-Implementierung aufsetzen kann.
@@
 
-Die Nutzer die einen FIDO2 Schlüssel registriert haben erhalten nach
-initialer Angabe ihrer E-Mail und ihres Passworts eine Aufforderung ihren FIDO2
-Schlüssel zu verwenden um letztendlich auch einen Token zu erhalten.
+Die Nutzer, die einen FIDO2 Schlüssel registriert haben, erhalten nach
+initialer Angabe ihrer E-Mail und ihres Passworts eine Aufforderung, ihren
+FIDO2 Schlüssel zu verwenden, um letztendlich auch einen Token zu erhalten.
 
@@
 
-Alle geschützten beziehungsweise nutzerbezogenen Ressourcen die dieser Server
-vorhält dürfen nur durch die Mitgabe eines Tokens abgerufen werden. Dieser muss
-mit dem \texttt{Authorization}-Header gesetzt werden werden.
+Alle geschützten beziehungsweise nutzerbezogenen Ressourcen, die dieser Server
+vorhält, dürfen nur durch die Mitgabe eines Tokens abgerufen werden. Dieser
+muss mit dem \texttt{Authorization}-Header gesetzt werden.
 
@@
 
-Da bekanntermaßen in Klartext gespeicherte Passwörter eine verherende
-Sicherheitslücke darstellen speichert der Server nur den Bcrypt-Hash der
-Passwörter die durch einen sogenannten Salt randomisiert werden.
+Da bekanntermaßen in Klartext gespeicherte Passwörter eine verheerende
+Sicherheitslücke darstellen, speichert der Server nur den Bcrypt-Hash der
+Passwörter, die durch einen sogenannten Salt randomisiert werden.
 
-Sowohl der Schlüssel zur signierung von Tokens als auch der Salt für Bcrypt
+Sowohl der Schlüssel zur Signierung von Tokens als auch der Salt für Bcrypt
 müssen in einem realen Anwendungsszenario geheim gehalten werden.
@@
 
-Um die konsistzenz der Nutzerverwaltung innerhalb des RAM sicherzustellen
-befindet sich die Nutzer-HashMap in einem Mutex um gleichzeitige ggf.
+Um die Konsistzenz der Nutzerverwaltung innerhalb des RAM sicherzustellen,
+befindet sich die Nutzer-HashMap in einem Mutex, um gleichzeitige ggf.
 inkonsistente Schreibvorgänge zu unterbinden.
@@
 Bei jeder eingehenden Anfrage wird die Datenstrukturen zur Nutzerverwaltung der
-Anfrage zugewiesen, diese hat nun die Garantie dass sie die einzige Anfrage ist
-die in diesem Moment schreiben kann. Sollten mehrere Anfragen gleichzeitig
-eingehen muss eine warten bis die andere abgearbeitet wurde. Dies ist zwar
-absolut inperformant bei großen Anfragemengen aber bei einem
+Anfrage zugewiesen, diese hat nun die Garantie, dass sie die einzige Anfrage
+ist, die in diesem Moment schreiben kann. Sollten mehrere Anfragen gleichzeitig
+eingehen, können diese nur sequenziell abgearbeitet werden. Dies ist zwar
+absolut inperformant bei großen Anfragemengen, aber bei einem
 Proof-of-Concept-Server mit sehr wenigen Nutzern vertretbar.
@@
 Da die Erstellung von den durch WebAuthn definierten Datenstrukturen und die
-kryptografischen Algorithmen nicht trivial zu implementieren sind gibt es für
-gängige Sprachen die eine Anwendung im Web-Kontext finden Bibliotheken die das
-WebAuthn-Protokoll abstrahiert bereitstellen.
+kryptografischen Algorithmen nicht trivial zu implementieren sind, gibt es für
+gängige Sprachen, die eine Anwendung im Web-Kontext finden, Bibliotheken die
+das WebAuthn-Protokoll abstrahiert bereitstellen.
 
-Für Rust ist eine umfangreicherere dieser Bibliotheken ``webauthn-rs'',
+Für Rust ist eine umfangreichere dieser Bibliotheken ``webauthn-rs'',
 verfügbar auf GitHub unter der Mozilla-Public-License-2.0:
@@
 Die Bibliothek lässt sich durch Informationen über den Server konfigurieren:
-Es wird eine Relying Party Id und eine registrierbare Domain erwartet um die
+Es wird eine Relying Party Id und eine registrierbare Domain erwartet, um die
 Identität des Servers für die Schlüsselerstellung wiederverwenden zu können.
@@
 Nach erfolgreicher Initialisierung gibt es im wesentlichen 4 relevante
-Funktionen mit denen sich der WebAuthn-Flow komplett implementieren lässt:
+Funktionen, mit denen sich der WebAuthn-Flow komplett implementieren lässt:
 
@@
 Gibt eine Challenge und einen Schlüssel-Authetifizierungsstatus für den
-momentanen Nutzer zurück. Dieser muss muss persistiert werden.
+momentanen Nutzer zurück. Dieser muss persistiert werden.
 
@@
 Nimmt eine Schlüssel-Authetifizierungsstatus und eine signierte Challenge und
-validiert diese gegeninander.
+validiert diese gegeneinander.
 
-Somit sind Serverseitig alle Voraussetzungen geschaffen um eine API
-bereitzustellen die WebAuthn unterstützt. Detailierte Informationen zu den
+Somit sind serverseitig alle Voraussetzungen geschaffen, um eine API
+bereitzustellen, die WebAuthn unterstützt. Detaillierte Informationen zu den
 bereitgestellten Endpunkten finden sich unter Abschnitt~\ref{docs}.
@@
 
-Ziel der Client-Implementierung ist es beispielhaftes Nutzer-Oberfläche zu
-schaffen das die unter Abschnitt~\ref{docs} beschriebene API verwendet um ein
-Nutzer-Login mit FIDO2 zu ermöglichen.
+Ziel ist es, eine beispielhafte Nutzer-Oberfläche zu schaffen, das die unter
+Abschnitt~\ref{docs} beschriebene API verwendet, um einen Nutzer-Login mit
+FIDO2 zu ermöglichen.
 
@@
 Die Oberfläche ist eine sehr kleine JavaScript Anwendung für die API.\ Folglich
-orientieren sich die Aktionen die der Nutzer in diesem ausführen kann direkt an
-dieser. In der Abbildung~\ref{fig:ui-flow} sind der Ablauf aller möglichen
-Nutzeraktionen und die dazugehörigen API Endpunkte abgebildet.
+orientieren sich die Aktionen, die der Nutzer in diesem ausführen kann, direkt
+an dieser. In der Abbildung~\ref{fig:ui-flow} sind der Ablauf aller möglichen
+Nutzeraktionen und die dazugehörigen API-Endpunkte abgebildet.
 
@@
 
-Da der Umfang der Nutzer-Aktionen sich stark in Grenzen hält gibt es keinen
-Grund eines der komplexeren Frameworks einzusetzen die typischerweise für
+Da der Umfang der Nutzer-Aktionen sich stark in Grenzen hält, gibt es keinen
+Grund, eines der komplexeren Frameworks einzusetzen, die typischerweise für
 Web-Oberflächen eingesetzt werden. Die einzige externe Bibliothek ist eine
-moderene und stabilere Base64 Implementierung als die Browser-Native
+modernere und stabilere Base64 Implementierung als die Browser-native
 Alternative.
 
-Das Grundkonzept zum rendering der Oberfläche ist einen Anwendungszustand durch
-eine pure (also Seiten-Effekt freie) Funktion in HTML zu übersetzen.
+Das Grundkonzept zum Rendering der Oberfläche ist, einen Anwendungszustand
+durch eine pure (also Seiten-Effekt freie) Funktion in HTML zu übersetzen.
 Nutzeraktionen können diesen Zustand durch Interaktion verändern und lösen
-somit einen neuen Render-Vorgang aus. Diese bzw. ähnliche Architekturen findet
-sich in Frameworks wie React oder Elm wieder.
-
-\textcolor{red}{res react elm}
+somit einen neuen Rendervorgang aus. Diese bzw. ähnliche Architekturen findet
+sich in Frameworks wie React oder Elm wieder.~\cite{elm}
 
@@
 
-Die definierten möglichen Interaktionen die der Nutzer tätigen kann sind:
+Die definierten möglichen Interaktionen, die der Nutzer tätigen kann, sind:
 
@@
 
-Da die Nutzer-Oberfläche in JavaScript geschrieben ist und dazu gedacht ist in
-einem Browser-Kontext ausgeführt zu werden kann die Browser-Implementierung für
-CTAP verwendet werden was die komplexität der Anwendung drastisch veringert, da
-diese Implementierung sehr stark abstrahiert ist.
+Da die Nutzer-Oberfläche in JavaScript geschrieben ist und dazu gedacht ist, in
+einem Browser-Kontext ausgeführt zu werden, kann die Browser-Implementierung
+für CTAP verwendet werden, was die Komplexität der Anwendung drastisch
+verringert, da diese Implementierung sehr stark abstrahiert ist.
 
-Im wesentlichen werden nur zwei der bereitgestellten Funktionen in der gesamten
+Im Wesentlichen werden nur zwei der bereitgestellten Funktionen in der gesamten
 Nutzer-Oberfläche verwendet:
@@
 
-Da die Parameter dieser Funktionen von dem Server bereitgestellt werden ist die
-Implementierung der Kommunikation mit dem Authenticator also sehr trivial.
+Da die Parameter dieser Funktionen von dem Server bereitgestellt werden, ist
+die Implementierung der Kommunikation mit dem Authenticator also sehr trivial.
 
@@
 	\item[Ausgabe:] -
-	\item[Beschreibung:] Verifiziert einen Nutzer falls der Code mit dem bei
+	\item[Beschreibung:] Verifiziert einen Nutzer, falls der Code mit dem bei
 		der Registrierung generierten Code übereinstimmt.
@@
 	\item[Eingabe:] Credentials \{ email, password \}
-	\item[Ausgabe:] UserDetails \{ token, verified, keys \} | CredentialRequestOptions 
+	\item[Ausgabe:] UserDetails \{ token, verified, keys \} |
+		CredentialRequestOptions 
 	\item[Beschreibung:] Gibt dem Nutzer entweder seine UserDetails zurück oder
 		stellt eine WebAuthn Authetifizierungs-Challenge die der Nutzer
-		signiert bei dem Endpunkt \textit{/auth/fido2/login} enreichen muss
+		signiert bei dem Endpunkt \textit{/auth/fido2/login} enreichen muss,
 		falls ein FIDO2 Schlüssel hinterlegt wurde.
@@
 	\item[Beschreibung:] Startet einen Registrierungsprozess für einen FIDO2
-		Schlüssel. Setzt Serverseitig den ``Key-Registration-State'' eines
+		Schlüssel. Setzt serverseitig den ``Key-Registration-State'' eines
 		Nutzers und gibt die Registrierungs-Optionen inklusive einer
@@
 Speichermediums (RAM) für die Nutzerverwaltung sind nach jedem Neustart alle
-Anwendungsdaten zurückgesetzt. Dies ist für Test-Zwecke absolut in Ordnung, und
+Anwendungsdaten zurückgesetzt. Dies ist für Test-Zwecke absolut in Ordnung und
 ließe sich für eine reale Anwendung leicht durch die Implementierung einer
@@
 alle definierten Interaktionen ausführen. Auch hier ließe sich die
-ausgestaltung der Oberfläche für eine reale Anwedung mit der entsprechenden
+Ausgestaltung der Oberfläche für eine reale Anwendung mit der entsprechenden
 Fachexpertise verbessern.
@@
 die Signierung der Challenge lediglich aus einem Tippen auf den Authenticator.
-Dies allein bietet ein deutlich verbessertes Erlebnis im vergleich zu
-SMS-Tokens oder One-Time-Passwords als zweiten Faktor (vorrausgesetzt der
+Dies allein bietet ein deutlich verbessertes Erlebnis im Vergleich zu
+SMS-Tokens oder One-Time-Passwords als zweiten Faktor (vorausgesetzt der
 Authenticator befindet sich in der Nähe des Systems oder ist sogar bereits
@@
 Die Hauptrisiken des momentanen Systems ist die fehlende Passwortvalidierung
-und der fehlende Zwang einen FIDO Schlüssel bei der Registrierung zu
-hinterlegen. So ist es momentan möglich einen Nutzer mit einem schwachen
+und der fehlende Zwang, einen FIDO Schlüssel bei der Registrierung zu
+hinterlegen. So ist es momentan möglich, einen Nutzer mit einem schwachen
 Passwort anzulegen und keinen Schlüssel zu registrieren. Dies ließe sich
@@
 
-Unter der Annahme dass eine Schlüssel-Registrierung verpflichtend sei wäre ein
-Angriff auf einen Nutzeraccount nurnoch durch physikalische Entwendung des
-Hardware Tokens möglich und dementsprechend unwahrscheinlich. Somit kann unter
+Unter der Annahme, dass eine Schlüssel-Registrierung verpflichtend sei, wäre
+ein Angriff auf einen Nutzeraccount nur noch durch physikalische Entwendung des
+Hardware-Tokens möglich und dementsprechend unwahrscheinlich. Somit kann unter
 dieser Annahme das System als sicher angesehen werden.
@@
 Single-Faktor-Authentifizierung mit dem Hardware-Token hinsichtlich des
-Nutzererlebnises empfehlenswert (Sicherheitsimplikationen außer Acht gelassen).
+Nutzer-Erlebnises empfehlenswert (Sicherheitsimplikationen außer Acht
+gelassen).
 
-Eine alternative Möglichkeit könnten weiterhin zwei Faktoren sein, allerdings
-unter der Bedingung, dass das Passwort lediglich zur einrichtung neuer Systeme
-abgefragt wird und sonst ein Single-Faktor-Login mit dem Hardware-Token möglich
-ist.
+Eine alternative Möglichkeit könnten weiterhin die Verwendung von zwei Faktoren
+sein, allerdings unter der Bedingung, dass das Passwort lediglich zur
+einrichtung neuer Systeme abgefragt wird und sonst ein Single-Faktor-Login mit
+dem Hardware-Token möglich ist.
 
-Desweiteren ist es bei Hardware Tokens - deren physikalischer Verlust nicht
-gänzlich auszuschließen ist - ratsam einen Backup Token zu besitzen und diesen
-bei jedem Online-Dienst als Backup Token zu hinterlegen um einen
+Des Weiteren ist es bei Hardware Tokens - deren physikalischer Verlust nicht
+gänzlich auszuschließen ist - ratsam, einen Backup Token zu besitzen und diesen
+bei jedem Online-Dienst als Backup Token zu hinterlegen, um einen
 Schlüsselverlust abfangen zu können. Eine bekannte Alternative zu einem zweiten
@@
 \listoffigures
-
-\nocite{*}
 

### 2022-06-06 grundlagen-der-programmierung-1/02/HelloApp.java
Change my name :)

@@ -1,7 +1,7 @@
 /**
  * HelloApp writes a string to stdout.
  * 
- * @author Maximilian Schulke schulke@th-brandenburg.de
+ * @author Mara Schulke schulke@th-brandenburg.de
  * @version 1.01
  */
 public class HelloApp {

### 2022-06-06 grundlagen-der-programmierung-1/02/HelloGUI.java

@@ -4,7 +4,7 @@ import javax.swing.*;
 /**
  * HelloGUI opens a window with a given title.
  * 
- * @author Maximilian Schulke schulke@th-brandenburg.de
+ * @author Mara Schulke schulke@th-brandenburg.de
  * @version 1.01
  * @see <a href="https://docs.oracle.com/javase/10/docs/api/javax/swing/JFrame.html">JFrame</a>
  */

### 2022-06-06 grundlagen-der-programmierung-1/03/DataTypes.java

@@ -1,5 +1,5 @@
 /**
- * @author Maximilian Schulke schulke@th-brandenburg.de
+ * @author Mara Schulke schulke@th-brandenburg.de
  * @version 1.0.0
  * @see Java Documentation
  */

### 2022-06-06 grundlagen-der-programmierung-1/04/Operatoren.java

@@ -1,5 +1,5 @@
 /**
- * @author schulke@th-brandenburg.de
+ * @author Mara Schulke schulke@th-brandenburg.de
  * @version 1.0.0
  * @see Java Documentation
  */

### 2022-06-06 grundlagen-der-programmierung-1/05/Berechnungen.java

@@ -1,5 +1,5 @@
 /**
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  * @version 1.0.0
  * @see Java Documentation
  */

### 2022-06-06 grundlagen-der-programmierung-1/05/TestBerechnen.java

@@ -1,6 +1,6 @@
 /**
  * Testet die Klasse Berechnen
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  * @version 1.0.0
  * @see Berechnen
  */

### 2022-06-06 grundlagen-der-programmierung-1/06/Kalkulator.java

@@ -2,7 +2,7 @@ import java.util.Scanner;
 import java.util.Locale;
 
 /**
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  * @version 1.0.0
  */
 public class Kalkulator {

### 2022-06-06 grundlagen-der-programmierung-1/07/Converter.java

@@ -2,7 +2,7 @@
  * Ein Converter stellt methoden zur Konvertierung von 
  * String nach Int und Int nach String zu verfügung.
  * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  * @version 1.0.0
  */
 public class Converter {

### 2022-06-06 grundlagen-der-programmierung-1/07/StringChange.java

@@ -1,7 +1,7 @@
 /**
  * StringChange experimentiert mit dem Verhalten von String und char[]
  * 
- * @author Maximilian Schulke <schulke@th-brandeburg>
+ * @author Mara Schulke <schulke@th-brandeburg>
  * @version 1.0.0
  */
 public class StringChange {

### 2022-06-06 grundlagen-der-programmierung-1/08/Labyrinth.java

@@ -5,7 +5,7 @@ import java.util.Arrays;
  * rekursiv einen Lösungweg zu finden.
  * 
  * @version 1.0.0
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  */
 public class Labyrinth {
 	/**

### 2022-06-06 grundlagen-der-programmierung-1/08/Utilities.java

@@ -1,7 +1,7 @@
 /**
  * Beinhaltet nützliche Funktionen.
  * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  * @version 1.0.0
  */
 public class Utilities {

### 2022-06-06 grundlagen-der-programmierung-1/09/FileAttribute.java

@@ -6,7 +6,7 @@ import java.util.Scanner;
  * FileAttribute ist ein Hilfsprogramm um Datei- und Ordner-Infos auszugeben.
  * Pfade können entweder über stdin oder als cli parameter übergeben werden. 
  * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  * @version 1.0.0
  */
 public class FileAttribute {

### 2022-06-06 grundlagen-der-programmierung-1/10/TelefonListe.java

@@ -11,7 +11,7 @@ import java.util.Arrays;
 /**
  * Die Klasse TelefonListe kann Telefonlisten-Einträge verwalten, speichern und laden. 
  * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
+ * @author Mara Schulke <schulke@th-brandenburg.de>
  * @version 1.0.0
  */
 public class TelefonListe {

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le03/Circle.java

@@ -1,0 +1,90 @@
+package com.maraschulke.gdp2.le03;
+
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt oder Umfang
+ * eines Kreises zu berechnen und auf die Konsole auszugeben.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0 16.08.2020
+ */
+public class Circle {
+
+	/**
+	 * Haelt den Radius des Kreises.
+	 */
+	private double r;
+
+	/**
+	 * Der Konstruktor der Klasse
+	 * 
+	 * @param radius Der Radius des Kreises
+	 */
+	public Circle (double radius) {
+		r = radius;
+	}
+
+	/**
+	 * Berechnet den Umfang anhand des gesetzten Radius.
+	 * 
+	 * @return Umfang
+	 */
+	public double circumference() {
+		return 2 * Math.PI * r;
+	}
+
+	/**
+	 * Berechnet den Flaecheninhalt anhand des gesetzten Radius.
+	 * 
+	 * @return Flaecheninhalt
+	 */
+	public double area() {
+		return Math.PI * Math.pow(r, 2);
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String.
+	 *
+	 * @return String der Klasse
+	 */
+	public String toString() {
+		return "Circle [ Radius " + getValue() + ", Area " + area() + ", Circumference " + circumference() + " ]";
+	}
+
+	/**
+	 * Gibt den Radius, den Umfang und den Flaecheninhalt auf die Konsole aus.
+	 */
+	public void show() {
+		System.out.println(toString());
+	}
+
+	/**
+	 * Ueberschreibt den Radius des Kreises.
+	 * 
+	 * @param radius Neuer Radius
+	 */
+	public void setValue(double radius) {
+		r = radius;
+	}
+
+	/**
+	 * Gibt den Radius des Kreises zurueck.
+	 * 
+	 * @return Radius des Kreises
+	 */
+	public double getValue() {
+		return r;
+	}
+
+	/**
+	 * Testet die Klasse Circle.
+	 * 
+	 * @param args CLI-Argumente
+	 */
+	public static void main(String[] args) {
+		Circle c = new Circle(13.5);
+		c.show();
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Circle.java

@@ -1,0 +1,64 @@
+package com.maraschulke.gdp2.le04;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Kreises zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Circle extends Ellipse {
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param x X-Position
+	 * @param y Y-Position
+	 * @param radius Radius
+	 */
+	public Circle(double x, double y, double radius) {
+		super(x, y, radius, radius);
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @return Umfang
+	 */
+	public double circumference() {
+		return perimeter();
+	}
+
+
+	/**
+	 * Getter für Radius
+	 * 
+	 * @return Radius
+	 */
+	public double getRadius() {
+		return getRadiusX();
+	}
+
+
+	/**
+	 * Setter für Radius
+	 * 
+	 * @param radius
+	 */
+	public void setRadius(double radius) {
+		setRadiusX(radius);
+		setRadiusY(radius);
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return super.toString().replaceAll("Ellipse", "Circle").replaceAll("Perimeter", "Circumference");
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Ellipse.java

@@ -1,0 +1,162 @@
+package com.maraschulke.gdp2.le04;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang einer Ellipsezu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Ellipse {
+
+	/**
+	 * Die X-Position
+	 */
+	private double x;
+
+	/**
+	 * Die Y-Position
+	 */
+	private double y;
+
+	/**
+	 * Der Radius in der horizontalen
+	 */
+	private double radiusX;
+
+	/**
+	 * Der Radius in der vertikalen
+	 */
+	private double radiusY;
+
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param x X-Position
+	 * @param y Y-Position
+	 * @param rx Radius in der horizontalen
+	 * @param ry Radius in der vertikalen
+	 */
+	public Ellipse(double x, double y, double rx, double ry) {
+		setX(x);
+		setY(y);
+		setRadiusX(rx);
+		setRadiusY(ry);
+	}
+
+
+	/**
+	 * Berechnet den Flächeninhalt
+	 * 
+	 * @return Flächeninhalt
+	 */
+	public double area() {
+		return radiusX * radiusY * Math.PI;
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @return Umfang
+	 */
+	public double perimeter() {
+		if (radiusX == 0 && radiusY == 0) return 0;
+
+		double h = Math.pow(radiusX - radiusY, 2) / Math.pow(radiusX + radiusY, 2);
+
+		return Math.PI * (radiusX + radiusY) * (1 + 3 * h / (10 + Math.sqrt(4 - 3 * h)));
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return "Ellipse [ Pos [ " + getX() + ", " + getY() + " ], RadiusX " + getRadiusX() + ", RadiusY " + getRadiusY() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
+	}
+
+
+	/**
+	 * Getter für X
+	 * 
+	 * @return X
+	 */
+	public double getX() {
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Line.java

@@ -1,0 +1,152 @@
+package com.maraschulke.gdp2.le04;
+
+
+/**
+ * Diese Klasse kann dazu verwendet werden
+ * die Länge einer Linie zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Line {
+
+	/**
+	 * X-Wert der Start-Position
+	 */
+	private double ax;
+
+	/**
+	 * Y-Wert der Start-Position
+	 */
+	private double ay;
+
+	/**
+	 * X-Wert der End-Position
+	 */
+	private double bx;
+
+	/**
+	 * Y-Wert der End-Position
+	 */
+	private double by;
+
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param ax X-Wert der Start-Position
+	 * @param ay Y-Wert der Start-Position
+	 * @param bx X-Wert der End-Position
+	 * @param by Y-Wert der End-Position
+	 */
+	public Line(double ax, double ay, double bx, double by) {
+		setAX(ax);
+		setAY(ay);
+		setBX(bx);
+		setBY(by);
+	}
+
+
+	/**
+	 * Berechnet die Länge
+	 * 
+	 * @return Länge
+	 */
+	public double length() {
+		double x = bx - ax;
+		double y = by - ay;
+
+		return Math.sqrt(Math.pow(x, 2) + Math.pow(y, 2));
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return "Line [ Pos [ " + getAX() + ", " + getAY() + " ], Dest [ " + getBX() + ", " + getBY() + " ], Length " + length() + " ]";
+	}
+
+
+	/**
+	 * Getter für AX
+	 * 
+	 * @return AX
+	 */
+	public double getAX() {
+		return ax;
+	}
+
+
+	/**
+	 * Setter für AX
+	 * 
+	 * @param AX
+	 */
+	public void setAX(double ax) {
+		this.ax = ax;
+	}
+
+
+	/**
+	 * Getter für AY
+	 * 
+	 * @return AY
+	 */
+	public double getAY() {
+		return ay;
+	}
+
+
+	/**
+	 * Setter für AY
+	 * 
+	 * @param AY
+	 */
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Polygon.java

@@ -1,0 +1,182 @@
+package com.maraschulke.gdp2.le04;
+
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Polygons zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Polygon {
+
+	/**
+	 * X-Positon
+	 */
+	private double x;
+
+	/**
+	 * Y-Positon
+	 */
+	private double y;
+
+	/**
+	 * Liste der Punkte
+	 */
+	private double[][] points;
+
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param x X-Position
+	 * @param y Y-Position
+	 * @param points Liste von Punkten
+	 */
+	public Polygon(double x, double y, double[][] points) {
+		setX(x);
+		setY(y);
+		setPoints(points);
+	}
+
+
+	/**
+	 * Gibt den nächsten Punkt für einen belibigen Index zurück
+	 * 
+	 * @param index
+	 * @return Punkt
+	 */
+	private double[] nextPoint(int index) {
+		return points[index == points.length - 1 ? 0 : index + 1];
+	}
+
+
+	/**
+	 * Berechnet den Flächeninhalt
+	 * 
+	 * @see https://www.mathopenref.com/coordpolygonarea.html
+	 * @return Flächeninhalt
+	 */
+	public double area() {
+		double accum = 0.0;
+
+		for (int i = 0; i < points.length; i++) {
+			double[] current = points[i];
+			double[] next = nextPoint(i);
+
+			double x1 = current[0];
+			double y1 = current[1];
+			double x2 = next[0];
+			double y2 = next[1];
+
+			accum += (x1 * y2 - y1 * x2);
+		}
+
+		return Math.abs(accum / 2);
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @see https://www.mathopenref.com/polygonperimeter.html
+	 * @return Umfang
+	 */
+	public double perimeter() {
+		double accum = 0;
+
+		for (int i = 0; i < points.length; i++) {
+			double[] current = points[i];
+			double[] next = nextPoint(i);
+
+			double[] vec = {
+				(next[0] - current[0]),
+				(next[1] - current[1])
+			};
+
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Rectangle.java

@@ -1,0 +1,158 @@
+package com.maraschulke.gdp2.le04;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Rechtecks zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Rectangle {
+
+	/**
+	 * X-Position
+	 */
+	private double x;
+
+	/**
+	 * Y-Position
+	 */
+	private double y;
+
+	/**
+	 * Länge A
+	 */
+	private double a;
+
+	/**
+	 * Länge B
+	 */
+	private double b;
+
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param x X-Position
+	 * @param y Y-Position
+	 * @param a Länge A
+	 * @param b Länge B
+	 */
+	public Rectangle(double x, double y, double a, double b) {
+		setX(x);
+		setY(y);
+		setA(a);
+		setB(b);
+	}
+
+
+	/**
+	 * Berechnet den Flächeninhalt
+	 * 
+	 * @return Flächeninhalt
+	 */
+	public double area() {
+		return a * b;
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @return Umfang
+	 */
+	public double perimeter() {
+		return a + a + b + b;
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return "Rectangle [ Pos [ " + getX() + ", " + getY() + " ], A " + getA() + ", B " + getB() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
+	}
+
+
+	/**
+	 * Getter für X
+	 * 
+	 * @return X
+	 */
+	public double getX() {
+		return x;
+	}
+
+
+	/**
+	 * Setter für X
+	 * 
+	 * @param X
+	 */
+	public void setX(double x) {
+		this.x = x;
+	}
+
+
+	/**
+	 * Getter für Y
+	 * 
+	 * @return Y
+	 */
+	public double getY() {
+		return y;
+	}
+
+
+	/**
+	 * Setter für Y
+	 * 
+	 * @param Y
+	 */
+	public void setY(double y) {
+		this.y = y;
+	}
+
+
+	/**
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Square.java

@@ -1,0 +1,33 @@
+package com.maraschulke.gdp2.le04;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Vierecks zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Square extends Rectangle {
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param x X-Position
+	 * @param y Y-Position
+	 * @param a Länger einer Seite
+	 */
+	public Square(double x, double y, double a) {
+		super(x, y, a, a);
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString()  {
+		return super.toString().replaceAll("Rectangle", "Square");
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Test.java

@@ -1,0 +1,32 @@
+package com.maraschulke.gdp2.le04;
+
+/**
+ * Diese Klasse kann dazu verwenden die anderen
+ * Klassen in diesem Package zu testen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Test {
+
+	/**
+	 * Die main methode
+	 * 
+	 * @param args
+	 */
+	public static void main(String[] args) {
+		Ellipse e = new Ellipse(0, 0, 10, 10);
+		System.out.println(e);
+
+		Circle c = new Circle(0, 0, 15);
+		System.out.println(c);
+
+		Polygon p = new Polygon(0, 0, new double[][] {
+			{0, 0},
+			{0, 4},
+			{4, 4}
+		});
+		System.out.println(p);
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le04/Triangle.java

@@ -1,0 +1,42 @@
+package com.maraschulke.gdp2.le04;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Dreiecks zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Triangle extends Polygon {
+
+	/**
+	 * Der Konstruktor
+	 *
+	 * @param x X-Position
+	 * @param y Y-Position
+	 * @param ax X-Position von Punkt A
+	 * @param ay Y-Position von Punkt A
+	 * @param bx X-Position von Punkt B
+	 * @param by Y-Position von Punkt B
+	 * @param cx X-Position von Punkt C
+	 * @param cy Y-Position von Punkt C
+	 */
+	public Triangle(double x, double y, double ax, double ay, double bx, double by, double cx, double cy) {
+		super(x, y, new double[][] {
+			{ax, ay},
+			{bx, by},
+			{cx, cy}
+		});
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 *
+	 * @return String-Darstellung
+	 */
+	public String toString()  {
+		return super.toString().replaceAll("Polygon", "Triangle");
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Circle.java

@@ -1,0 +1,62 @@
+package com.maraschulke.gdp2.le05;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flächeninhalt
+ * oder Umfang eines Kreises zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Circle extends Ellipse {	
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param pos Position
+	 * @param radius Radius
+	 */
+	public Circle(Point pos, double radius) {
+		super(pos, radius, radius);
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @return Umfang
+	 */
+	public double circumference() {
+		return perimeter();
+	}
+
+
+	/**
+	 * Getter für Radius
+	 * 
+	 * @return Radius
+	 */
+	public double getRadius() {
+		return getRadiusX();
+	}
+
+
+	/**
+	 * Setter für Radius
+	 * 
+	 * @param radius
+	 */
+	public void setRadius(double radius) {
+		setRadiusX(radius);
+		setRadiusY(radius);
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return super.toString().replaceAll("Ellipse", "Circle").replaceAll("Perimeter", "Circumference");
+	}
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Drawing.java

@@ -1,0 +1,72 @@
+package com.maraschulke.gdp2.le05;
+
+import java.util.ArrayList;
+
+
+/**
+ * Stellt eine Zeichnung dar.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Drawing {
+
+	/**
+	 * Die Liste der Formen in dieser Zeichnung.
+	 */
+	private ArrayList<Figure> figures;
+
+
+	/**
+	 * Der Konstruktor.
+	 * 
+	 * @param figs Initiale Formen
+	 */
+	public Drawing(Figure[] figs) {
+		this.figures = new ArrayList<Figure>();
+
+		for (Figure f : figs) {
+			add(f);
+		}
+	}
+
+
+	/**
+	 * Fügt eine Form zur Zeichnung hinzu.
+	 * 
+	 * @param f Form
+	 */
+	public void add(Figure f) {
+		figures.add(f);
+	}
+
+
+	/**
+	 * Entfernt eine Form von der Zeichnung.
+	 * 
+	 * @param f Form
+	 */
+	public void remove(Figure f) {
+		figures.remove(f);
+	}
+
+
+	/**
+	 * Verändert die Position einer Form mit dem gegebenen index
+	 * 
+ 	 * @param index Index der Form
+	 * @param pos Neue Position
+	 */
+	public void move(int index, Point pos) {
+		figures.get(index).setPos(pos);
+	}
+
+
+	/**
+	 * Löscht alle Formen der Zeichnung.
+	 */
+	public void clear() {
+		figures.clear();
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Ellipse.java

@@ -1,0 +1,110 @@
+package com.maraschulke.gdp2.le05;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flächeninhalt
+ * oder Umfang einer Ellipse zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Ellipse extends Figure implements Sizeable {
+
+	/**
+	 * Der Radius in der horizontalen
+	 */
+	private double radiusX;
+
+	/**
+	 * Der Radius in der vertikalen
+	 */
+	private double radiusY;
+
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param pos Position
+	 * @param rx Radius in der horizontalen
+	 * @param ry Radius in der vertikalen
+	 */
+	public Ellipse(Point pos, double rx, double ry) {
+		super(pos);
+		setRadiusX(rx);
+		setRadiusY(ry);
+	}
+
+
+	/**
+	 * Berechnet den Flächeninhalt
+	 * 
+	 * @return Flächeninhalt
+	 */
+	public double area() {
+		return radiusX * radiusY * Math.PI;
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @return Umfang
+	 */
+	public double perimeter() {
+		if (radiusX == 0 && radiusY == 0) return 0;
+
+		double h = Math.pow(radiusX - radiusY, 2) / Math.pow(radiusX + radiusY, 2);
+
+		return Math.PI * (radiusX + radiusY) * (1 + 3 * h / (10 + Math.sqrt(4 - 3 * h)));
+	}
+
+
+	/**
+	 * Getter für RadiusX
+	 * 
+	 * @return Radius X
+	 */
+	public double getRadiusX() {
+		return radiusX;
+	}
+
+
+	/**
+	 * Setter für RadiusX
+	 * 
+	 * @param Radius X
+	 */
+	public void setRadiusX(double radiusX) {
+		this.radiusX = radiusX;
+	}
+
+
+	/**
+	 * Getter für RadiusY
+	 * 
+	 * @return Radius Y
+	 */
+	public double getRadiusY() {
+		return radiusY;
+	}
+
+
+	/**
+	 * Setter für RadiusY
+	 * 
+	 * @param Radius Y
+	 */
+	public void setRadiusY(double radiusY) {
+		this.radiusY = radiusY;
+	}
+
+
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Figure.java

@@ -1,0 +1,46 @@
+package com.maraschulke.gdp2.le05;
+
+/**
+ * Stellt eine abstrakte Form dar.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+abstract class Figure {
+
+	/**
+	 * Die Position der Form.
+	 */
+	private Point pos;
+
+
+	/**
+	 * Der Konstruktor.
+	 * 
+	 * @param pos Initiale Position
+	 */
+	public Figure(Point pos) {
+		setPos(pos);
+	}
+
+
+	/**
+	 * Getter der Position
+	 * 
+	 * @return Position
+	 */
+	public Point getPos() {
+		return pos;
+	}
+
+
+	/**
+	 * Setter der Position
+	 * 
+	 * @param pos Position
+	 */
+	public void setPos(Point pos) {
+		this.pos = pos;
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Line.java

@@ -1,0 +1,72 @@
+package com.maraschulke.gdp2.le05;
+
+/**
+ * Diese Klasse kann dazu verwendet werden
+ * die Länge einer Linie zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Line extends Figure {
+
+	/**
+	 * End-Position
+	 */
+	private Point dest;
+
+
+	/**
+	 * Der Konstruktor.
+	 * 
+	 * @param start Start-Position
+	 * @param dest End-Position
+	 */
+	public Line(Point start, Point dest) {
+		super(start);
+		setDest(dest);
+	}
+
+
+	/**
+	 * Berechnet die Länge
+	 * 
+	 * @return Länge
+	 */
+	public double length() {
+		double xdiff = getPos().xDistanceTo(getDest());
+		double ydiff = getPos().yDistanceTo(getDest());
+
+		return Math.sqrt(Math.pow(xdiff, 2) + Math.pow(ydiff, 2));
+	}
+
+
+	/**
+	 * Getter der End-Position
+	 * 
+	 * @return End-Position
+	 */
+	public Point getDest() {
+		return dest;
+	}
+
+
+	/**
+	 * Setter der End-Position
+	 * 
+	 * @param dest Neue End-Position
+	 */
+	public void setDest(Point dest) {
+		this.dest = dest;
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return "Line [ Pos " + getPos() + ", Dest " + getDest() + ", Length " + length() + " ]";
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Point.java

@@ -1,0 +1,105 @@
+package com.maraschulke.gdp2.le05;
+
+/**
+ * Ein 2D-Punkt
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Point {
+
+	/**
+	 * Die X Position
+	 */
+	private double x;
+
+	/**
+	 * Die Y-Position
+	 */
+	private double y;
+
+
+	/**
+	 * Der Konstruktor.
+	 * 
+	 * @param x Initialer X-Wert
+	 * @param y Initialer Y-Wert
+	 */
+	public Point (double x, double y) {
+		this.setX(x);
+		this.setY(y);
+	}
+
+
+	/**
+	 * Berechnet die Distanz auf der X-Achse zu einem anderen Punkt
+	 * 
+	 * @param p Zu vergleichender Punkt
+	 * @return Distanz
+	 */
+	public double xDistanceTo(Point p) {
+		return p.getX() - this.getX();
+	}
+
+
+	/**
+	 * Berechnet die Distanz auf der Y-Achse zu einem anderen Punkt
+	 * 
+	 * @param p Zu vergleichender Punkt
+	 * @return Distanz
+	 */
+	public double yDistanceTo(Point p) {
+		return p.getY() - this.getY();
+	}
+
+
+	/**
+	 * Getter für X
+	 * 
+	 * @return X
+	 */
+	public double getX() {
+		return x;
+	}
+
+
+	/**
+	 * Setter für X
+	 * 
+	 * @param X
+	 */
+	public void setX(double x) {
+		this.x = x;
+	}
+
+
+	/**
+	 * Getter für Y
+	 * 
+	 * @return Y
+	 */
+	public double getY() {
+		return y;
+	}
+
+
+	/**
+	 * Setter für Y
+	 * 
+	 * @param Y
+	 */
+	public void setY(double y) {
+		this.y = y;
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return "[" + getX() + "; " + getY() + "]";
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Polygon.java

@@ -1,0 +1,123 @@
+package com.maraschulke.gdp2.le05;
+
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Polygons zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Polygon extends Figure implements Sizeable {
+
+	/**
+	 * Liste der Punkte
+	 */
+	private Point[] points;
+
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param pos Position
+	 * @param points Liste der Punkte
+	 */
+	public Polygon(Point pos, Point[] points) {
+		super(pos);
+		setPoints(points);
+	}
+
+
+	/**
+	 * Gibt den nächsten Punkt für einen belibigen Index zurück
+	 * 
+	 * @param index
+	 * @return Punkt
+	 */
+	private Point nextPoint(int index) {
+		return points[index == points.length - 1 ? 0 : index + 1];
+	}
+
+
+	/**
+	 * Berechnet den Flächeninhalt
+	 * 
+	 * @see https://www.mathopenref.com/coordpolygonarea.html
+	 * @return Flächeninhalt
+	 */
+	public double area() {
+		double accum = 0.0;
+
+		for (int i = 0; i < points.length; i++) {
+			Point current = points[i];
+			Point next = nextPoint(i);
+
+			accum += (current.getX() * next.getY() - current.getY() * next.getX());
+		}
+
+		return Math.abs(accum / 2);
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @see https://www.mathopenref.com/polygonperimeter.html
+	 * @return Umfang
+	 */
+	public double perimeter() {
+		double accum = 0;
+
+		for (int i = 0; i < points.length; i++) {
+			Point current = points[i];
+			Point next = nextPoint(i);
+			
+			Line vec = new Line(
+				new Point(0, 0),
+				new Point(current.xDistanceTo(next), current.yDistanceTo(next))
+			);
+			
+			accum += vec.length();
+		}
+
+		return accum;
+	}
+
+
+	/**
+	 * Getter für Punkte
+	 * 
+	 * @return Punkte
+	 */
+	public Point[] getPoints() {
+		return points;
+	}
+
+
+	/**
+	 * Setter für Punkte
+	 * 
+	 * @param Punkte
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Rectangle.java

@@ -1,0 +1,106 @@
+package com.maraschulke.gdp2.le05;
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Rechtecks zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Rectangle extends Figure implements Sizeable {
+
+	/**
+	 * Länge A
+	 */
+	private double a;
+
+	/**
+	 * Länge B
+	 */
+	private double b;
+
+	
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param pos Position
+	 * @param a Länge A
+	 * @param b Länge B
+	 */
+	public Rectangle(Point pos, double a, double b) {
+		super(pos);
+		setA(a);
+		setB(b);
+	}
+
+
+	/**
+	 * Berechnet den Flächeninhalt
+	 * 
+	 * @return Flächeninhalt
+	 */
+	public double area() {
+		return a * b;
+	}
+
+
+	/**
+	 * Berechnet den Umfang
+	 * 
+	 * @return Umfang
+	 */
+	public double perimeter() {
+		return a + a + b + b;
+	}
+
+
+	/**
+	 * Getter für A
+	 * 
+	 * @return A
+	 */
+	public double getA() {
+		return a;
+	}
+
+
+	/**
+	 * Setter für A
+	 * 
+	 * @param A
+	 */
+	public void setA(double a) {
+		this.a = a;
+	}
+
+
+	/**
+	 * Getter für B
+	 * 
+	 * @return B
+	 */
+	public double getB() {
+		return b;
+	}
+
+
+	/**
+	 * Setter für B
+	 * 
+	 * @param B
+	 */
+	public void setB(double b) {
+		this.b = b;
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString() {
+		return "Rectangle [ Pos " + getPos() + ", A " + getA() + ", B " + getB() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Sizeable.java

@@ -1,0 +1,15 @@
+package com.maraschulke.gdp2.le05;
+
+
+/**
+ * Ein Objekt das einen Flächeninhalt und Umfang hat.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public interface Sizeable {
+
+	public double area();
+	public double perimeter();
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Square.java

@@ -1,0 +1,33 @@
+package com.maraschulke.gdp2.le05;
+
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Vierecks zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Square extends Rectangle {
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param pos Position
+	 * @param a Seitenlänge
+	 */
+	public Square(Point pos, double a) {
+		super(pos, a, a);
+	}
+
+
+	/**
+	 * Formatiert die Instanz als String
+	 * 
+	 * @return String-Darstellung
+	 */
+	public String toString()  {
+		return super.toString().replaceAll("Rectangle", "Square");
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Test.java

@@ -1,0 +1,36 @@
+package com.maraschulke.gdp2.le05;
+
+
+/**
+ * Diese Klasse kann dazu verwenden die anderen
+ * Klassen in diesem Package zu testen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Test {
+
+	/**
+	 * Die main methode
+	 * 
+	 * @param args
+	 */
+	public static void main(String[] args) {
+		Ellipse e = new Ellipse(new Point(0, 0), 10, 10);
+		System.out.println(e);
+
+		Circle c = new Circle(new Point(0, 0), 15);
+		System.out.println(c);
+
+		Polygon p = new Polygon(new Point(0, 0), new Point[] {
+			new Point(0, 0),
+			new Point(0, 4),
+			new Point(4, 4)
+		});
+		System.out.println(p);
+
+		Line l = new Line(new Point(0, 0), new Point(10, 10));
+		System.out.println(l);
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le05/Triangle.java

@@ -1,0 +1,34 @@
+package com.maraschulke.gdp2.le05;
+
+
+/**
+ * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
+ * oder Umfang eines Dreiecks zu berechnen.
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class Triangle extends Polygon {
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param pos Position
+	 * @param a Punkt A
+	 * @param b Punkt B
+	 * @param c Punkt C
+	 */
+	public Triangle(Point pos, Point a, Point b, Point c) {
+		super(pos, new Point[] {a, b, c});
+	}
+	
+	/**
+	 * Formatiert die Instanz als String
+	 *
+	 * @return String-Darstellung
+	 */
+	public String toString()  {
+		return super.toString().replaceAll("Polygon", "Triangle");
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le06/CircularQueue.java

@@ -1,0 +1,239 @@
+package com.maraschulke.gdp2.le06;
+
+import java.util.ArrayList;
+
+
+/**
+ * Eine Ringspeichers Queue-Implementation
+ * 
+ * @author Mara Schulke <schulke@th-brandenbug.de>
+ * @version 1.0.0
+ */
+public class CircularQueue<T> implements IQueue<T> {
+
+	/**
+	 * Die Maximalgröße
+	 */
+	private int size;
+
+	/**
+	 * Index des ersten Elementes
+	 */
+	public int head;
+
+	/**
+	 * Index des letzten Elementes
+	 */
+	public int tail;
+
+	/**
+	 * Liste mit allen Elementen
+	 */
+	private ArrayList<T> queue;
+
+
+	/**
+	 * Der Konstruktor
+	 * 
+	 * @param size Maximalgröße
+	 */
+	public CircularQueue(int size) {
+		head = 0;
+		tail = 0;
+		queue = new ArrayList<T>();
+		this.size = size;
+
+		for (int i = 0; i < size; i++) {
+			queue.add(null);
+		}
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public void push(T el) throws QueueException {
+		if (isFull()) {
+			head = (tail + 1) % size;
+		}
+
+		queue.set(tail, el);
+		tail = (tail + 1) % size;
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public T pop() throws QueueException {
+		if (isEmpty()) {
+			throw new QueueException("Queue is empty!");
+		}
+
+		T el = queue.set(head, null);
+		head = (head + 1) % size;
+
+		return el;
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public T head() throws QueueException {
+		if (queue.isEmpty()) {
+			throw new QueueException("Queue is empty!");
+		}
+
+		return queue.get(head);
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public T tail() throws QueueException {
+		if (queue.isEmpty()) {
+			throw new QueueException("Queue is empty!");
+		}
+
+		return queue.get(tail);
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public int length() {
+		int emptySlots = queue.stream().filter(el -> el == null).toArray().length;
+
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le06/IQueue.java

@@ -1,0 +1,86 @@
+package com.maraschulke.gdp2.le06;
+
+
+/**
+ * Ein generisches Interface für alle Queue ähnlichen
+ * Datenstrukturen
+ * 
+ * @author Mara Schulke <schulke@th-brandenbug.de>
+ * @version 1.0.0
+ */
+public interface IQueue<T> {
+
+	/**
+	 * Fügt ein Element ans Ende an
+	 * 
+	 * @param el Element
+	 * @throws QueueException
+	 */
+	public void push(T el) throws QueueException;
+
+	/**
+	 * Entfernt das erste Element
+	 * 
+	 * @return Entferntes Element
+	 * @throws QueueException
+	 */
+	public T pop() throws QueueException;
+
+	/**
+	 * Gibt das erste Element zurück
+	 * 
+	 * @return Erstes Element
+	 * @throws QueueException
+	 */
+	public T head() throws QueueException;
+
+	/**
+	 * Gibt das letztes Element zurück
+	 * 
+	 * @return Letztes Element
+	 * @throws QueueException
+	 */
+	public T tail() throws QueueException;
+
+	/**
+	 * Gibt die Länge der Queue zurück
+	 * 
+	 * @return Länge
+	 */
+	public int length();	
+
+	/**
+	 * Entfernt alle Element
+	 */
+	public void clear();
+
+	/**
+	 * Gibt an ob die Queue voll ist
+	 * 
+	 * @return Ist die Liste voll
+	 */
+	public boolean isFull();
+
+	/**
+	 * Gibt an ob die Queue leer ist
+	 * 
+	 * @return Ist die Queue leer?
+	 */
+	public boolean isEmpty();
+
+	/**
+	 * Gibt die Maximalgröße zurück
+	 * 
+	 * @return Maximalgröße
+	 */
+	public int getSize();
+
+	/**
+	 * Setzt die Maximalgröße
+	 * 
+	 * @param size Maximalgröße
+	 * @throws QueueException
+	 */
+	public void setSize(int size) throws QueueException;
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le06/LinearQueue.java

@@ -1,0 +1,191 @@
+package com.maraschulke.gdp2.le06;
+
+import java.util.ArrayList;
+
+
+/**
+ * Eine lineare Queue-Implementation
+ * 
+ * @author Mara Schulke <schulke@th-brandenbug.de>
+ * @version 1.0.0
+ */
+public class LinearQueue<T> implements IQueue<T> {
+
+	/**
+	 * Maximalgröße der Queue
+	 */
+	private int size;
+
+	/**
+	 * Enthält die Elemente der Queue
+	 */
+	private ArrayList<T> queue;
+
+
+	/**
+	 * Der Konstruktor
+	 */
+	public LinearQueue() {
+		size = Integer.MAX_VALUE;
+		queue = new ArrayList<>();
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public void push(T el) throws QueueException {
+		if (isFull()) {
+			throw new QueueException("The queue is full!");
+		}
+
+		queue.add(el);	
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public T pop() throws QueueException {
+		if (queue.isEmpty()) {			
+			throw new QueueException("The queue is empty!");
+		}
+
+		T el = queue.get(0);
+		queue.remove(0);
+		return el;
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public T head() throws QueueException {
+		if (queue.isEmpty()) {
+			throw new QueueException("The queue is empty!");
+		}
+
+		return queue.get(0);
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public T tail() throws QueueException {
+		if (queue.isEmpty()) {
+			throw new QueueException("The queue is empty!");
+		}
+
+		return queue.get(queue.size() - 1);
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public int length() {
+		return queue.size();
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public boolean isFull() {
+		return queue.size() >= this.size;
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public boolean isEmpty() {
+		return queue.isEmpty();
+	}
+
+
+	/**
+	 * @see IQueue
+	 */
+	@Override
+	public void clear() {
+		queue.clear();
+	}
+
+
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le06/QueueException.java

@@ -1,0 +1,14 @@
+package com.maraschulke.gdp2.le06;
+
+
+/**
+ * Eine QueueException fasst alle Ausnahmen die mit
+ * dem Queues in Zusammenhang stehen zusammen
+ * 
+ * @author Mara Schulke <schulke@th-brandenbug.de>
+ * @version 1.0.0
+ */
+public class QueueException extends Exception {
+	public QueueException() {super();}
+	public QueueException(String message) {super(message);}
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le07/GUIBsp.java

@@ -1,0 +1,139 @@
+package com.maraschulke.gdp2.le07;
+
+import javafx.application.Application;
+import javafx.geometry.Insets;
+import javafx.scene.Scene;
+import javafx.scene.control.TextField;
+import javafx.scene.input.MouseEvent;
+import javafx.scene.control.Label;
+import javafx.scene.control.Button;
+import javafx.scene.layout.FlowPane;
+import javafx.scene.layout.VBox;
+import javafx.stage.Stage;
+
+
+/**
+ * Eine Beispiel GUI
+ * 
+ * @version 1.0.0
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ */
+public class GUIBsp extends Application {
+
+	/**
+	 * Das Hello Label
+	 */
+	Label hello;
+
+	/**
+	 * Das Name Label
+	 */
+	Label name;
+
+	/**
+	 * Das Textfeld für die Eingabe
+	 */
+	TextField input;
+
+	/**
+	 * Das Label für das Textfeld
+	 */
+	Label label;
+
+	/**
+	 * Der Button um zu bestätigen
+	 */
+	Button ok;
+
+	/**
+	 * Der Button um abzubrechen
+	 */
+	Button cancel;
+
+
+	/**
+	 * Baut die GUI auf einer Stage
+	 * 
+	 * @param primaryStage Stage
+	 */
+	@Override
+	public void start(Stage primaryStage) throws Exception {
+		// Setze Window-Titel
+		primaryStage.setTitle("GUIBsp");
+
+		// Erstelle Labels
+		hello = new Label();
+		hello.setText("Hallo");
+		name = new Label();
+
+		// Erstelle Layouts und füge Ausgabe-Labels hinzu
+		VBox layout = new VBox();
+		layout.setPadding(new Insets(25, 25, 25, 25));
+		layout.setSpacing(10);
+
+		FlowPane outputPane = new FlowPane();
+		outputPane.setHgap(5);
+		outputPane.getChildren().addAll(hello, name);
+
+		layout.getChildren().add(outputPane);
+
+		// Erstelle Eingabe-Textfeld
+		label = new Label();
+		label.setText("Name:");
+		input = new TextField();
+
+		FlowPane inputPane = new FlowPane();
+		inputPane.setHgap(10);
+		inputPane.getChildren().addAll(label, input);
+
+		layout.getChildren().add(inputPane);
+
+		// Erstelle Buttons
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le09/GUIPaint.java

@@ -1,0 +1,112 @@
+package com.maraschulke.gdp2.le09;
+
+import javafx.application.Application;
+import javafx.stage.Stage;
+import javafx.scene.Group;
+import javafx.scene.Scene;
+import javafx.scene.paint.Color;
+import javafx.scene.shape.*;
+
+/**
+ * Die GUIPaint Anwendung
+ * 
+ * @author Mara Schulke <schulke@th-brandenburg.de>
+ * @version 1.0.0
+ */
+public class GUIPaint extends Application {
+
+	/**
+	 * Variablen zum tracken der Rechteck-Position.
+	 */
+	public double x0,x1,y0,y1;
+
+	/**
+	 * Die Start Methode zeichnet das Haus und registriert
+	 * die EventHandler.
+	 * 
+	 * @param primaryStage Die Bühne
+	 */
+	@Override
+	public void start(Stage primaryStage) {
+		try {
+			Group root = new Group();
+			Scene scene = new Scene(root, 600, 400);
+
+			// Definiert Window-Verhalten
+			primaryStage.setScene(scene);
+			primaryStage.setTitle("GUIPaint");
+			primaryStage.setMinWidth(600);
+			primaryStage.setMinHeight(400);
+			primaryStage.show();
+
+			// Hintergrund
+			Rectangle sky = new Rectangle(0, 000, 600, 300);
+			sky.setFill(Color.LIGHTBLUE);
+
+			Rectangle earth = new Rectangle(0, 300, 600, 100);
+			earth.setFill(Color.GREEN);
+
+			Group house = new Group();
+
+			// Das Haus
+			Polygon roof = new Polygon(200.0, 150.0, 300.0, 100.0, 400.0, 150.0);
+			roof.setFill(Color.ORANGE);
+
+			Rectangle body = new Rectangle(200, 150, 200, 150);
+			body.setFill(Color.BEIGE);
+
+			Rectangle window1 = new Rectangle(237.5, 175, 25, 40);
+			window1.setFill(Color.WHITE);
+
+			Rectangle window2 = new Rectangle(237.5, 225, 25, 40);
+			window2.setFill(Color.WHITE);
+
+			Rectangle door = new Rectangle(285, 250, 30, 50);
+			door.setFill(Color.SADDLEBROWN);
+
+			Rectangle window3 = new Rectangle(337.5, 175, 25, 40);
+			window3.setFill(Color.WHITE);
+
+			Rectangle window4 = new Rectangle(337.5, 225, 25, 40);
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/Controller.java

@@ -1,0 +1,130 @@
+package com.maraschulke.gdp2.le10;
+
+import com.maraschulke.gdp2.le10.model.*;
+
+import javafx.fxml.FXML;
+import javafx.collections.ListChangeListener;
+import javafx.event.ActionEvent;
+import javafx.scene.input.MouseEvent;
+import javafx.scene.input.MouseButton;
+import javafx.scene.layout.BorderPane;
+import javafx.scene.control.ChoiceBox;
+import javafx.scene.control.ColorPicker;
+import javafx.scene.control.Label;
+import javafx.scene.layout.Pane;
+
+
+public class Controller {
+
+	private Drawing drawing = new Drawing(new Figure[] {});  
+	private Figure figure;
+	private Point start;
+
+	@FXML private BorderPane layout;
+	@FXML private Pane root;
+	@FXML private ChoiceBox<String> typeBox;
+	@FXML private ColorPicker colorPicker;
+	@FXML private Label mousePosition;
+	@FXML private Label area;
+	@FXML private Label perimeter;
+
+
+	@FXML public void initialize() {
+		drawing.getList().addListener(new ListChangeListener<Figure>() {
+			@Override
+	        public void onChanged(ListChangeListener.Change<? extends Figure> c) {
+	            if (c.next()){
+	            	root.getChildren().clear();
+	            	c.getList().stream().forEach(f -> {
+	            		root.getChildren().add(f.intoShape());
+	            	});
+            	}
+	        }
+	    });
+	}
+
+
+	@FXML protected void onMouseDown(MouseEvent ev) {
+		if (!ev.isPrimaryButtonDown()) return;
+
+		start = new Point(ev.getX(), ev.getY());
+		figure = FigureType.fromString(typeBox.getValue()).intoFigure(start, colorPicker.getValue());
+		drawing.add(figure);
+	}
+
+
+	@FXML protected void onMouseDragged(MouseEvent ev) {
+		if (!ev.isPrimaryButtonDown()) return;
+
+		onMouseMoved(ev);
+
+		Point current = new Point(ev.getX(), ev.getY());
+
+		switch (FigureType.fromString(typeBox.getValue())) {
+			case CIRCLE:
+				((Circle) figure).setRadius(Math.abs(new Line(start, current).length()));
+				break;
+			case ELLIPSE:
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/Main.java

@@ -1,0 +1,29 @@
+package com.maraschulke.gdp2.le10;
+
+import javafx.application.Application;
+import javafx.fxml.FXMLLoader;
+import javafx.scene.Parent;
+import javafx.scene.Scene;
+import javafx.stage.Stage;
+
+public class Main extends Application {
+
+	@Override
+	public void start(Stage stage) throws Exception {
+		Parent root = FXMLLoader.load(getClass().getResource("view.fxml"));
+
+		Scene scene = new Scene(root, 800, 600);
+
+		stage.setTitle("Drawing App");
+		stage.setScene(scene);
+		stage.setMinWidth(800);
+		stage.setMinHeight(600);
+		stage.show();
+	}
+
+
+	public static void main(String[] args) {
+		launch(args);
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Circle.java

@@ -1,0 +1,42 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+
+public class Circle extends Ellipse {	
+
+	public Circle(Point pos) {
+		super(pos);
+	}
+
+
+	public Circle(Point pos, Paint color) {
+		super(pos, color);
+	}
+
+
+	public Circle(Point pos, double radius) {
+		super(pos, radius, radius);
+	}
+
+
+	public double circumference() {
+		return perimeter();
+	}
+
+
+	public String toString() {
+		return super.toString().replaceAll("Ellipse", "Circle");
+	}
+
+
+	public double getRadius() {
+		return getRadiusX();
+	}
+
+
+	public void setRadius(double radius) {
+		setRadiusX(radius);
+		setRadiusY(radius);
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Drawing.java

@@ -1,0 +1,45 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.collections.FXCollections;
+import javafx.collections.ObservableList;
+
+
+public class Drawing {
+
+	private ObservableList<Figure> figures;
+
+
+	public Drawing(Figure[] figs) {
+		this.figures = FXCollections.observableArrayList();
+
+		for (Figure f : figs) {
+			add(f);
+		}
+	}
+
+
+	public void add(Figure f) {
+		figures.add(f);
+	}
+
+
+	public void remove(Figure f) {
+		figures.remove(f);
+	}
+
+
+	public void update(Figure f) {
+		figures.set(figures.indexOf(f), f);
+	}
+
+
+	public void clear() {
+		figures.clear();
+	}
+
+
+	public ObservableList<Figure> getList() {
+		return figures;
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Ellipse.java

@@ -1,0 +1,77 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+import javafx.scene.shape.Shape;
+
+public class Ellipse extends Figure implements Sizeable {
+
+	private double radiusX;
+	private double radiusY;
+
+
+	public Ellipse(Point pos) {
+		super(pos);
+		setRadiusX(0);
+		setRadiusY(0);
+	}
+	
+
+	public Ellipse(Point pos, Paint color) {
+		this(pos);
+		setColor(color);
+	}
+
+
+	public Ellipse(Point pos, double rx, double ry) {
+		super(pos);
+		setRadiusX(rx);
+		setRadiusY(ry);
+	}
+
+
+	public double area() {
+		return radiusX * radiusY * Math.PI;
+	}
+
+
+	public double perimeter() {
+		if (radiusX == 0 && radiusY == 0) return 0;
+
+		double h = Math.pow(radiusX - radiusY, 2) / Math.pow(radiusX + radiusY, 2);
+
+		return Math.PI * (radiusX + radiusY) * (1 + 3 * h / (10 + Math.sqrt(4 - 3 * h)));
+	}
+
+
+	public String toString() {
+		return "Ellipse [ Pos " + getPos() + ", Color " + getColor() + ", RadiusX " + getRadiusX() + ", RadiusY " + getRadiusY() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
+	}
+
+
+	public double getRadiusX() {
+		return radiusX;
+	}
+
+
+	public void setRadiusX(double radiusX) {
+		this.radiusX = radiusX;
+	}
+
+
+	public double getRadiusY() {
+		return radiusY;
+	}
+
+
+	public void setRadiusY(double radiusY) {
+		this.radiusY = radiusY;
+	}
+
+
+	public Shape intoShape() {
+		Shape shape = new javafx.scene.shape.Ellipse(getPos().getX(), getPos().getY(), getRadiusX(), getRadiusY());
+		shape.setFill(getColor());
+		return shape;
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Figure.java

@@ -1,0 +1,43 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+import javafx.scene.shape.Shape;
+
+public abstract class Figure {
+
+	private Point pos;
+	private Paint color;
+
+	public Figure(Point pos) {
+		setPos(pos);
+	}
+
+	public Figure(Point pos, Paint color) {
+		this(pos);
+		setColor(color);
+	}
+
+
+	abstract public Shape intoShape();
+
+
+	public Point getPos() {
+		return pos;
+	}
+
+
+	public void setPos(Point pos) {
+		this.pos = pos;
+	}
+
+
+	public Paint getColor() {
+		return color;
+	}
+
+
+	public void setColor(Paint color) {
+		this.color = color;
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/FigureType.java

@@ -1,0 +1,68 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+
+public enum FigureType {
+	CIRCLE,
+	ELLIPSE,
+	LINE,
+	RECTANGLE,
+	SQUARE,
+	POLYGON,
+	TRIANGLE;
+
+	public static FigureType fromString(String id) {
+		switch (id.toLowerCase()) {
+			case "circle":
+				return CIRCLE;
+			case "ellipse":
+				return ELLIPSE;
+			case "line":
+				return LINE;
+			case "rectangle":
+				return RECTANGLE;
+			case "square":
+				return SQUARE;
+			case "polygon":
+				return POLYGON;
+			case "triangle":
+				return TRIANGLE;
+		}
+
+		return null;
+	}
+
+
+	public static FigureType fromFigure(Figure f) {
+		if (f instanceof Circle) return CIRCLE;
+		else if (f instanceof Ellipse) return ELLIPSE;
+		else if (f instanceof Line) return LINE;
+		else if (f instanceof Square) return SQUARE;
+		else if (f instanceof Rectangle) return RECTANGLE;
+		else if (f instanceof Triangle) return TRIANGLE;
+		else if (f instanceof Polygon) return POLYGON;
+		else return null;
+	}
+
+
+	public Figure intoFigure(Point pos, Paint color) {
+		switch (this) {
+			case CIRCLE:
+				return new Circle(pos, color);
+			case ELLIPSE:
+				return new Ellipse(pos, color);
+			case LINE:
+				return new Line(pos, color);
+			case RECTANGLE:
+				return new Rectangle(pos, color);
+			case SQUARE:
+				return new Square(pos, color);
+			case POLYGON:
+				return new Polygon(pos, color);
+			case TRIANGLE:
+				return new Triangle(pos, color);
+		}
+
+		return null;
+	}
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Line.java

@@ -1,0 +1,58 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+import javafx.scene.shape.Shape;
+
+public class Line extends Figure {
+
+	private Point dest;
+
+
+	public Line(Point pos) {
+		super(pos);
+		setDest(new Point(pos.getX(), pos.getY()));
+	}
+
+	
+	public Line(Point pos, Paint color) {
+		this(pos);
+		setColor(color);
+	}
+
+
+	public Line(Point start, Point dest) {
+		super(start);
+		setDest(dest);
+	}
+
+
+	public double length() {
+		double xdiff = getPos().xDistanceTo(getDest());
+		double ydiff = getPos().yDistanceTo(getDest());
+
+		return Math.sqrt(Math.pow(xdiff, 2) + Math.pow(ydiff, 2));
+	}
+
+
+	public String toString() {
+		return "Line [ Pos " + getPos() + ", Color " + getColor() + ", Dest " + getDest() + ", Length " + length() + " ]";
+	}
+
+
+	public Point getDest() {
+		return dest;
+	}
+
+
+	public void setDest(Point dest) {
+		this.dest = dest;
+	}
+
+
+	public Shape intoShape() {
+		Shape shape = new javafx.scene.shape.Line(getPos().getX(), getPos().getY(), getDest().getX(), getDest().getY());
+		shape.setStroke(getColor());
+		return shape;
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Point.java

@@ -1,0 +1,49 @@
+package com.maraschulke.gdp2.le10.model;
+
+public class Point {
+
+	private double x;
+	private double y;
+
+
+	public Point (double x, double y) {
+		this.setX(x);
+		this.setY(y);
+	}
+
+
+	public double xDistanceTo(Point p) {
+		return p.getX() - this.getX();
+	}
+
+
+	public double yDistanceTo(Point p) {
+		return p.getY() - this.getY();
+	}
+
+
+	public double getX() {
+		return x;
+	}
+
+
+	public void setX(double x) {
+		this.x = x;
+	}
+
+
+	public double getY() {
+		return y;
+	}
+
+
+	public void setY(double y) {
+		this.y = y;
+	}
+
+
+	public String toString() {
+		return "[" + getX() + "; " + getY() + "]";
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Polygon.java

@@ -1,0 +1,112 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+import javafx.scene.shape.Shape;
+
+public class Polygon extends Figure implements Sizeable {
+
+	private Point[] points;
+
+
+	public Polygon(Point pos) {
+		super(pos);
+		setPoints(new Point[] {});
+	}
+	
+
+	public Polygon(Point pos, Paint color) {
+		this(pos);
+		setColor(color);
+	}
+
+
+	public Polygon(Point pos, Point[] points) {
+		super(pos);
+		setPoints(points);
+	}
+
+
+	private Point nextPoint(int index) {
+		return points[index == points.length - 1 ? 0 : index + 1];
+	}
+
+
+	/**
+	 * @see https://www.mathopenref.com/coordpolygonarea.html
+	 * @return Area of the Polygon
+	 */
+	public double area() {
+		double accum = 0.0;
+
+		for (int i = 0; i < points.length; i++) {
+			Point current = points[i];
+			Point next = nextPoint(i);
+
+			accum += (current.getX() * next.getY() - current.getY() * next.getX());
+		}
+
+		return Math.abs(accum / 2);
+	}
+
+
+	/**
+	 * @see https://www.mathopenref.com/polygonperimeter.html
+	 * @return Perimeter of the Polygon
+	 */
+	public double perimeter() {
+		double accum = 0;
+
+		for (int i = 0; i < points.length; i++) {
+			Point current = points[i];
+			Point next = nextPoint(i);
+			
+			Line vec = new Line(
+				new Point(0, 0),
+				new Point(current.xDistanceTo(next), current.yDistanceTo(next))
+			);
+			
+			accum += vec.length();
+		}
+
+		return accum;
+	}
+
+
+	public String toString() {
+		String s = "Polygon [ Pos " + getPos() + ", Color " + getColor() + ", Points [ ";
+
+		for (Point p : points) {
+			s += p + ", ";
+		}
+
+		s += " ], Area " + area() + ", Perimeter " + perimeter() + " ]";
+		return s;
+	}
+
+
+	public Point[] getPoints() {
+		return points;
+	}
+
+
+	public void setPoints(Point[] points) {
+		this.points = points;
+	}
+
+
+	public Shape intoShape() {
+		double[] coords = new double[] {};
+		int i = 0;
+
+		for (Point point : points) {
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Rectangle.java

@@ -1,0 +1,73 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+import javafx.scene.shape.Shape;
+
+public class Rectangle extends Figure implements Sizeable {
+
+	private double a;
+	private double b;
+
+
+	public Rectangle(Point pos) {
+		super(pos);
+		setA(0);
+		setB(0);
+	}
+
+	
+	public Rectangle(Point pos, Paint color) {
+		this(pos);
+		setColor(color);
+	}
+	
+
+	public Rectangle(Point pos, double a, double b) {
+		super(pos);
+		setA(a);
+		setB(b);
+	}
+
+
+	public double area() {
+		return a * b;
+	}
+
+
+	public double perimeter() {
+		return a + a + b + b;
+	}
+
+
+	public double getA() {
+		return a;
+	}
+
+
+	public void setA(double newA) {
+		a = newA;
+	}
+
+
+	public double getB() {
+		return b;
+	}
+
+
+	public void setB(double newB) {
+		b = newB;
+	}
+
+
+	public String toString() {
+		return "Rectangle [ Pos " + getPos() + ", Color " + getColor() + ", A " + getA() + ", B " + getB() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
+	}
+
+
+	public Shape intoShape() {
+		Shape shape = new javafx.scene.shape.Rectangle(getPos().getX(), getPos().getY(), getA(), getB());
+		shape.setFill(getColor());
+		return shape;
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Sizeable.java

@@ -1,0 +1,8 @@
+package com.maraschulke.gdp2.le10.model;
+
+public interface Sizeable {
+
+	public double area();
+	public double perimeter();
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Square.java

@@ -1,0 +1,26 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+
+public class Square extends Rectangle {
+
+	public Square(Point pos) {
+		super(pos);
+	}
+	
+	
+	public Square(Point pos, Paint color) {
+		super(pos, color);
+	}
+
+
+	public Square(Point pos, double a) {
+		super(pos, a, a);
+	}
+
+
+	public String toString() {
+		return super.toString().replaceAll("Rectangle", "Square");
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maraschulke/gdp2/le10/model/Triangle.java

@@ -1,0 +1,26 @@
+package com.maraschulke.gdp2.le10.model;
+
+import javafx.scene.paint.Paint;
+
+public class Triangle extends Polygon {
+
+	public Triangle(Point pos) {
+		super(pos);
+	}
+
+
+	public Triangle(Point pos, Paint color) {
+		super(pos, color);
+	}
+
+
+	public Triangle(Point pos, Point a, Point b, Point c) {
+		super(pos, new Point[] {a, b, c});
+	}
+
+
+	public String toString() {
+		return super.toString().replaceAll("Polygon", "Triangle");
+	}
+
+}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le03/Circle.java

@@ -1,90 +1,0 @@
-package com.maximilianschulke.gdp2.le03;
-
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt oder Umfang
- * eines Kreises zu berechnen und auf die Konsole auszugeben.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0 16.08.2020
- */
-public class Circle {
-
-	/**
-	 * Haelt den Radius des Kreises.
-	 */
-	private double r;
-
-	/**
-	 * Der Konstruktor der Klasse
-	 * 
-	 * @param radius Der Radius des Kreises
-	 */
-	public Circle (double radius) {
-		r = radius;
-	}
-
-	/**
-	 * Berechnet den Umfang anhand des gesetzten Radius.
-	 * 
-	 * @return Umfang
-	 */
-	public double circumference() {
-		return 2 * Math.PI * r;
-	}
-
-	/**
-	 * Berechnet den Flaecheninhalt anhand des gesetzten Radius.
-	 * 
-	 * @return Flaecheninhalt
-	 */
-	public double area() {
-		return Math.PI * Math.pow(r, 2);
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String.
-	 *
-	 * @return String der Klasse
-	 */
-	public String toString() {
-		return "Circle [ Radius " + getValue() + ", Area " + area() + ", Circumference " + circumference() + " ]";
-	}
-
-	/**
-	 * Gibt den Radius, den Umfang und den Flaecheninhalt auf die Konsole aus.
-	 */
-	public void show() {
-		System.out.println(toString());
-	}
-
-	/**
-	 * Ueberschreibt den Radius des Kreises.
-	 * 
-	 * @param radius Neuer Radius
-	 */
-	public void setValue(double radius) {
-		r = radius;
-	}
-
-	/**
-	 * Gibt den Radius des Kreises zurueck.
-	 * 
-	 * @return Radius des Kreises
-	 */
-	public double getValue() {
-		return r;
-	}
-
-	/**
-	 * Testet die Klasse Circle.
-	 * 
-	 * @param args CLI-Argumente
-	 */
-	public static void main(String[] args) {
-		Circle c = new Circle(13.5);
-		c.show();
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Circle.java

@@ -1,64 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Kreises zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Circle extends Ellipse {
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param x X-Position
-	 * @param y Y-Position
-	 * @param radius Radius
-	 */
-	public Circle(double x, double y, double radius) {
-		super(x, y, radius, radius);
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @return Umfang
-	 */
-	public double circumference() {
-		return perimeter();
-	}
-
-
-	/**
-	 * Getter für Radius
-	 * 
-	 * @return Radius
-	 */
-	public double getRadius() {
-		return getRadiusX();
-	}
-
-
-	/**
-	 * Setter für Radius
-	 * 
-	 * @param radius
-	 */
-	public void setRadius(double radius) {
-		setRadiusX(radius);
-		setRadiusY(radius);
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return super.toString().replaceAll("Ellipse", "Circle").replaceAll("Perimeter", "Circumference");
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Ellipse.java

@@ -1,162 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang einer Ellipsezu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Ellipse {
-
-	/**
-	 * Die X-Position
-	 */
-	private double x;
-
-	/**
-	 * Die Y-Position
-	 */
-	private double y;
-
-	/**
-	 * Der Radius in der horizontalen
-	 */
-	private double radiusX;
-
-	/**
-	 * Der Radius in der vertikalen
-	 */
-	private double radiusY;
-
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param x X-Position
-	 * @param y Y-Position
-	 * @param rx Radius in der horizontalen
-	 * @param ry Radius in der vertikalen
-	 */
-	public Ellipse(double x, double y, double rx, double ry) {
-		setX(x);
-		setY(y);
-		setRadiusX(rx);
-		setRadiusY(ry);
-	}
-
-
-	/**
-	 * Berechnet den Flächeninhalt
-	 * 
-	 * @return Flächeninhalt
-	 */
-	public double area() {
-		return radiusX * radiusY * Math.PI;
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @return Umfang
-	 */
-	public double perimeter() {
-		if (radiusX == 0 && radiusY == 0) return 0;
-
-		double h = Math.pow(radiusX - radiusY, 2) / Math.pow(radiusX + radiusY, 2);
-
-		return Math.PI * (radiusX + radiusY) * (1 + 3 * h / (10 + Math.sqrt(4 - 3 * h)));
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return "Ellipse [ Pos [ " + getX() + ", " + getY() + " ], RadiusX " + getRadiusX() + ", RadiusY " + getRadiusY() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
-	}
-
-
-	/**
-	 * Getter für X
-	 * 
-	 * @return X
-	 */
-	public double getX() {
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Line.java

@@ -1,152 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-
-/**
- * Diese Klasse kann dazu verwendet werden
- * die Länge einer Linie zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Line {
-
-	/**
-	 * X-Wert der Start-Position
-	 */
-	private double ax;
-
-	/**
-	 * Y-Wert der Start-Position
-	 */
-	private double ay;
-
-	/**
-	 * X-Wert der End-Position
-	 */
-	private double bx;
-
-	/**
-	 * Y-Wert der End-Position
-	 */
-	private double by;
-
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param ax X-Wert der Start-Position
-	 * @param ay Y-Wert der Start-Position
-	 * @param bx X-Wert der End-Position
-	 * @param by Y-Wert der End-Position
-	 */
-	public Line(double ax, double ay, double bx, double by) {
-		setAX(ax);
-		setAY(ay);
-		setBX(bx);
-		setBY(by);
-	}
-
-
-	/**
-	 * Berechnet die Länge
-	 * 
-	 * @return Länge
-	 */
-	public double length() {
-		double x = bx - ax;
-		double y = by - ay;
-
-		return Math.sqrt(Math.pow(x, 2) + Math.pow(y, 2));
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return "Line [ Pos [ " + getAX() + ", " + getAY() + " ], Dest [ " + getBX() + ", " + getBY() + " ], Length " + length() + " ]";
-	}
-
-
-	/**
-	 * Getter für AX
-	 * 
-	 * @return AX
-	 */
-	public double getAX() {
-		return ax;
-	}
-
-
-	/**
-	 * Setter für AX
-	 * 
-	 * @param AX
-	 */
-	public void setAX(double ax) {
-		this.ax = ax;
-	}
-
-
-	/**
-	 * Getter für AY
-	 * 
-	 * @return AY
-	 */
-	public double getAY() {
-		return ay;
-	}
-
-
-	/**
-	 * Setter für AY
-	 * 
-	 * @param AY
-	 */
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Polygon.java

@@ -1,182 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Polygons zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Polygon {
-
-	/**
-	 * X-Positon
-	 */
-	private double x;
-
-	/**
-	 * Y-Positon
-	 */
-	private double y;
-
-	/**
-	 * Liste der Punkte
-	 */
-	private double[][] points;
-
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param x X-Position
-	 * @param y Y-Position
-	 * @param points Liste von Punkten
-	 */
-	public Polygon(double x, double y, double[][] points) {
-		setX(x);
-		setY(y);
-		setPoints(points);
-	}
-
-
-	/**
-	 * Gibt den nächsten Punkt für einen belibigen Index zurück
-	 * 
-	 * @param index
-	 * @return Punkt
-	 */
-	private double[] nextPoint(int index) {
-		return points[index == points.length - 1 ? 0 : index + 1];
-	}
-
-
-	/**
-	 * Berechnet den Flächeninhalt
-	 * 
-	 * @see https://www.mathopenref.com/coordpolygonarea.html
-	 * @return Flächeninhalt
-	 */
-	public double area() {
-		double accum = 0.0;
-
-		for (int i = 0; i < points.length; i++) {
-			double[] current = points[i];
-			double[] next = nextPoint(i);
-
-			double x1 = current[0];
-			double y1 = current[1];
-			double x2 = next[0];
-			double y2 = next[1];
-
-			accum += (x1 * y2 - y1 * x2);
-		}
-
-		return Math.abs(accum / 2);
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @see https://www.mathopenref.com/polygonperimeter.html
-	 * @return Umfang
-	 */
-	public double perimeter() {
-		double accum = 0;
-
-		for (int i = 0; i < points.length; i++) {
-			double[] current = points[i];
-			double[] next = nextPoint(i);
-
-			double[] vec = {
-				(next[0] - current[0]),
-				(next[1] - current[1])
-			};
-
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Rectangle.java

@@ -1,158 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Rechtecks zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Rectangle {
-
-	/**
-	 * X-Position
-	 */
-	private double x;
-
-	/**
-	 * Y-Position
-	 */
-	private double y;
-
-	/**
-	 * Länge A
-	 */
-	private double a;
-
-	/**
-	 * Länge B
-	 */
-	private double b;
-
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param x X-Position
-	 * @param y Y-Position
-	 * @param a Länge A
-	 * @param b Länge B
-	 */
-	public Rectangle(double x, double y, double a, double b) {
-		setX(x);
-		setY(y);
-		setA(a);
-		setB(b);
-	}
-
-
-	/**
-	 * Berechnet den Flächeninhalt
-	 * 
-	 * @return Flächeninhalt
-	 */
-	public double area() {
-		return a * b;
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @return Umfang
-	 */
-	public double perimeter() {
-		return a + a + b + b;
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return "Rectangle [ Pos [ " + getX() + ", " + getY() + " ], A " + getA() + ", B " + getB() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
-	}
-
-
-	/**
-	 * Getter für X
-	 * 
-	 * @return X
-	 */
-	public double getX() {
-		return x;
-	}
-
-
-	/**
-	 * Setter für X
-	 * 
-	 * @param X
-	 */
-	public void setX(double x) {
-		this.x = x;
-	}
-
-
-	/**
-	 * Getter für Y
-	 * 
-	 * @return Y
-	 */
-	public double getY() {
-		return y;
-	}
-
-
-	/**
-	 * Setter für Y
-	 * 
-	 * @param Y
-	 */
-	public void setY(double y) {
-		this.y = y;
-	}
-
-
-	/**
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Square.java

@@ -1,33 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Vierecks zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Square extends Rectangle {
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param x X-Position
-	 * @param y Y-Position
-	 * @param a Länger einer Seite
-	 */
-	public Square(double x, double y, double a) {
-		super(x, y, a, a);
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString()  {
-		return super.toString().replaceAll("Rectangle", "Square");
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Test.java

@@ -1,32 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-/**
- * Diese Klasse kann dazu verwenden die anderen
- * Klassen in diesem Package zu testen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Test {
-
-	/**
-	 * Die main methode
-	 * 
-	 * @param args
-	 */
-	public static void main(String[] args) {
-		Ellipse e = new Ellipse(0, 0, 10, 10);
-		System.out.println(e);
-
-		Circle c = new Circle(0, 0, 15);
-		System.out.println(c);
-
-		Polygon p = new Polygon(0, 0, new double[][] {
-			{0, 0},
-			{0, 4},
-			{4, 4}
-		});
-		System.out.println(p);
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le04/Triangle.java

@@ -1,42 +1,0 @@
-package com.maximilianschulke.gdp2.le04;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Dreiecks zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Triangle extends Polygon {
-
-	/**
-	 * Der Konstruktor
-	 *
-	 * @param x X-Position
-	 * @param y Y-Position
-	 * @param ax X-Position von Punkt A
-	 * @param ay Y-Position von Punkt A
-	 * @param bx X-Position von Punkt B
-	 * @param by Y-Position von Punkt B
-	 * @param cx X-Position von Punkt C
-	 * @param cy Y-Position von Punkt C
-	 */
-	public Triangle(double x, double y, double ax, double ay, double bx, double by, double cx, double cy) {
-		super(x, y, new double[][] {
-			{ax, ay},
-			{bx, by},
-			{cx, cy}
-		});
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 *
-	 * @return String-Darstellung
-	 */
-	public String toString()  {
-		return super.toString().replaceAll("Polygon", "Triangle");
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Circle.java

@@ -1,62 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flächeninhalt
- * oder Umfang eines Kreises zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Circle extends Ellipse {	
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param pos Position
-	 * @param radius Radius
-	 */
-	public Circle(Point pos, double radius) {
-		super(pos, radius, radius);
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @return Umfang
-	 */
-	public double circumference() {
-		return perimeter();
-	}
-
-
-	/**
-	 * Getter für Radius
-	 * 
-	 * @return Radius
-	 */
-	public double getRadius() {
-		return getRadiusX();
-	}
-
-
-	/**
-	 * Setter für Radius
-	 * 
-	 * @param radius
-	 */
-	public void setRadius(double radius) {
-		setRadiusX(radius);
-		setRadiusY(radius);
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return super.toString().replaceAll("Ellipse", "Circle").replaceAll("Perimeter", "Circumference");
-	}
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Drawing.java

@@ -1,72 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-import java.util.ArrayList;
-
-
-/**
- * Stellt eine Zeichnung dar.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Drawing {
-
-	/**
-	 * Die Liste der Formen in dieser Zeichnung.
-	 */
-	private ArrayList<Figure> figures;
-
-
-	/**
-	 * Der Konstruktor.
-	 * 
-	 * @param figs Initiale Formen
-	 */
-	public Drawing(Figure[] figs) {
-		this.figures = new ArrayList<Figure>();
-
-		for (Figure f : figs) {
-			add(f);
-		}
-	}
-
-
-	/**
-	 * Fügt eine Form zur Zeichnung hinzu.
-	 * 
-	 * @param f Form
-	 */
-	public void add(Figure f) {
-		figures.add(f);
-	}
-
-
-	/**
-	 * Entfernt eine Form von der Zeichnung.
-	 * 
-	 * @param f Form
-	 */
-	public void remove(Figure f) {
-		figures.remove(f);
-	}
-
-
-	/**
-	 * Verändert die Position einer Form mit dem gegebenen index
-	 * 
- 	 * @param index Index der Form
-	 * @param pos Neue Position
-	 */
-	public void move(int index, Point pos) {
-		figures.get(index).setPos(pos);
-	}
-
-
-	/**
-	 * Löscht alle Formen der Zeichnung.
-	 */
-	public void clear() {
-		figures.clear();
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Ellipse.java

@@ -1,110 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flächeninhalt
- * oder Umfang einer Ellipse zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Ellipse extends Figure implements Sizeable {
-
-	/**
-	 * Der Radius in der horizontalen
-	 */
-	private double radiusX;
-
-	/**
-	 * Der Radius in der vertikalen
-	 */
-	private double radiusY;
-
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param pos Position
-	 * @param rx Radius in der horizontalen
-	 * @param ry Radius in der vertikalen
-	 */
-	public Ellipse(Point pos, double rx, double ry) {
-		super(pos);
-		setRadiusX(rx);
-		setRadiusY(ry);
-	}
-
-
-	/**
-	 * Berechnet den Flächeninhalt
-	 * 
-	 * @return Flächeninhalt
-	 */
-	public double area() {
-		return radiusX * radiusY * Math.PI;
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @return Umfang
-	 */
-	public double perimeter() {
-		if (radiusX == 0 && radiusY == 0) return 0;
-
-		double h = Math.pow(radiusX - radiusY, 2) / Math.pow(radiusX + radiusY, 2);
-
-		return Math.PI * (radiusX + radiusY) * (1 + 3 * h / (10 + Math.sqrt(4 - 3 * h)));
-	}
-
-
-	/**
-	 * Getter für RadiusX
-	 * 
-	 * @return Radius X
-	 */
-	public double getRadiusX() {
-		return radiusX;
-	}
-
-
-	/**
-	 * Setter für RadiusX
-	 * 
-	 * @param Radius X
-	 */
-	public void setRadiusX(double radiusX) {
-		this.radiusX = radiusX;
-	}
-
-
-	/**
-	 * Getter für RadiusY
-	 * 
-	 * @return Radius Y
-	 */
-	public double getRadiusY() {
-		return radiusY;
-	}
-
-
-	/**
-	 * Setter für RadiusY
-	 * 
-	 * @param Radius Y
-	 */
-	public void setRadiusY(double radiusY) {
-		this.radiusY = radiusY;
-	}
-
-
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Figure.java

@@ -1,46 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-/**
- * Stellt eine abstrakte Form dar.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-abstract class Figure {
-
-	/**
-	 * Die Position der Form.
-	 */
-	private Point pos;
-
-
-	/**
-	 * Der Konstruktor.
-	 * 
-	 * @param pos Initiale Position
-	 */
-	public Figure(Point pos) {
-		setPos(pos);
-	}
-
-
-	/**
-	 * Getter der Position
-	 * 
-	 * @return Position
-	 */
-	public Point getPos() {
-		return pos;
-	}
-
-
-	/**
-	 * Setter der Position
-	 * 
-	 * @param pos Position
-	 */
-	public void setPos(Point pos) {
-		this.pos = pos;
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Line.java

@@ -1,72 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-/**
- * Diese Klasse kann dazu verwendet werden
- * die Länge einer Linie zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Line extends Figure {
-
-	/**
-	 * End-Position
-	 */
-	private Point dest;
-
-
-	/**
-	 * Der Konstruktor.
-	 * 
-	 * @param start Start-Position
-	 * @param dest End-Position
-	 */
-	public Line(Point start, Point dest) {
-		super(start);
-		setDest(dest);
-	}
-
-
-	/**
-	 * Berechnet die Länge
-	 * 
-	 * @return Länge
-	 */
-	public double length() {
-		double xdiff = getPos().xDistanceTo(getDest());
-		double ydiff = getPos().yDistanceTo(getDest());
-
-		return Math.sqrt(Math.pow(xdiff, 2) + Math.pow(ydiff, 2));
-	}
-
-
-	/**
-	 * Getter der End-Position
-	 * 
-	 * @return End-Position
-	 */
-	public Point getDest() {
-		return dest;
-	}
-
-
-	/**
-	 * Setter der End-Position
-	 * 
-	 * @param dest Neue End-Position
-	 */
-	public void setDest(Point dest) {
-		this.dest = dest;
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return "Line [ Pos " + getPos() + ", Dest " + getDest() + ", Length " + length() + " ]";
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Point.java

@@ -1,105 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-/**
- * Ein 2D-Punkt
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Point {
-
-	/**
-	 * Die X Position
-	 */
-	private double x;
-
-	/**
-	 * Die Y-Position
-	 */
-	private double y;
-
-
-	/**
-	 * Der Konstruktor.
-	 * 
-	 * @param x Initialer X-Wert
-	 * @param y Initialer Y-Wert
-	 */
-	public Point (double x, double y) {
-		this.setX(x);
-		this.setY(y);
-	}
-
-
-	/**
-	 * Berechnet die Distanz auf der X-Achse zu einem anderen Punkt
-	 * 
-	 * @param p Zu vergleichender Punkt
-	 * @return Distanz
-	 */
-	public double xDistanceTo(Point p) {
-		return p.getX() - this.getX();
-	}
-
-
-	/**
-	 * Berechnet die Distanz auf der Y-Achse zu einem anderen Punkt
-	 * 
-	 * @param p Zu vergleichender Punkt
-	 * @return Distanz
-	 */
-	public double yDistanceTo(Point p) {
-		return p.getY() - this.getY();
-	}
-
-
-	/**
-	 * Getter für X
-	 * 
-	 * @return X
-	 */
-	public double getX() {
-		return x;
-	}
-
-
-	/**
-	 * Setter für X
-	 * 
-	 * @param X
-	 */
-	public void setX(double x) {
-		this.x = x;
-	}
-
-
-	/**
-	 * Getter für Y
-	 * 
-	 * @return Y
-	 */
-	public double getY() {
-		return y;
-	}
-
-
-	/**
-	 * Setter für Y
-	 * 
-	 * @param Y
-	 */
-	public void setY(double y) {
-		this.y = y;
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return "[" + getX() + "; " + getY() + "]";
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Polygon.java

@@ -1,123 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Polygons zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Polygon extends Figure implements Sizeable {
-
-	/**
-	 * Liste der Punkte
-	 */
-	private Point[] points;
-
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param pos Position
-	 * @param points Liste der Punkte
-	 */
-	public Polygon(Point pos, Point[] points) {
-		super(pos);
-		setPoints(points);
-	}
-
-
-	/**
-	 * Gibt den nächsten Punkt für einen belibigen Index zurück
-	 * 
-	 * @param index
-	 * @return Punkt
-	 */
-	private Point nextPoint(int index) {
-		return points[index == points.length - 1 ? 0 : index + 1];
-	}
-
-
-	/**
-	 * Berechnet den Flächeninhalt
-	 * 
-	 * @see https://www.mathopenref.com/coordpolygonarea.html
-	 * @return Flächeninhalt
-	 */
-	public double area() {
-		double accum = 0.0;
-
-		for (int i = 0; i < points.length; i++) {
-			Point current = points[i];
-			Point next = nextPoint(i);
-
-			accum += (current.getX() * next.getY() - current.getY() * next.getX());
-		}
-
-		return Math.abs(accum / 2);
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @see https://www.mathopenref.com/polygonperimeter.html
-	 * @return Umfang
-	 */
-	public double perimeter() {
-		double accum = 0;
-
-		for (int i = 0; i < points.length; i++) {
-			Point current = points[i];
-			Point next = nextPoint(i);
-			
-			Line vec = new Line(
-				new Point(0, 0),
-				new Point(current.xDistanceTo(next), current.yDistanceTo(next))
-			);
-			
-			accum += vec.length();
-		}
-
-		return accum;
-	}
-
-
-	/**
-	 * Getter für Punkte
-	 * 
-	 * @return Punkte
-	 */
-	public Point[] getPoints() {
-		return points;
-	}
-
-
-	/**
-	 * Setter für Punkte
-	 * 
-	 * @param Punkte
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Rectangle.java

@@ -1,106 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Rechtecks zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Rectangle extends Figure implements Sizeable {
-
-	/**
-	 * Länge A
-	 */
-	private double a;
-
-	/**
-	 * Länge B
-	 */
-	private double b;
-
-	
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param pos Position
-	 * @param a Länge A
-	 * @param b Länge B
-	 */
-	public Rectangle(Point pos, double a, double b) {
-		super(pos);
-		setA(a);
-		setB(b);
-	}
-
-
-	/**
-	 * Berechnet den Flächeninhalt
-	 * 
-	 * @return Flächeninhalt
-	 */
-	public double area() {
-		return a * b;
-	}
-
-
-	/**
-	 * Berechnet den Umfang
-	 * 
-	 * @return Umfang
-	 */
-	public double perimeter() {
-		return a + a + b + b;
-	}
-
-
-	/**
-	 * Getter für A
-	 * 
-	 * @return A
-	 */
-	public double getA() {
-		return a;
-	}
-
-
-	/**
-	 * Setter für A
-	 * 
-	 * @param A
-	 */
-	public void setA(double a) {
-		this.a = a;
-	}
-
-
-	/**
-	 * Getter für B
-	 * 
-	 * @return B
-	 */
-	public double getB() {
-		return b;
-	}
-
-
-	/**
-	 * Setter für B
-	 * 
-	 * @param B
-	 */
-	public void setB(double b) {
-		this.b = b;
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString() {
-		return "Rectangle [ Pos " + getPos() + ", A " + getA() + ", B " + getB() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Sizeable.java

@@ -1,15 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-
-/**
- * Ein Objekt das einen Flächeninhalt und Umfang hat.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public interface Sizeable {
-
-	public double area();
-	public double perimeter();
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Square.java

@@ -1,33 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Vierecks zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Square extends Rectangle {
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param pos Position
-	 * @param a Seitenlänge
-	 */
-	public Square(Point pos, double a) {
-		super(pos, a, a);
-	}
-
-
-	/**
-	 * Formatiert die Instanz als String
-	 * 
-	 * @return String-Darstellung
-	 */
-	public String toString()  {
-		return super.toString().replaceAll("Rectangle", "Square");
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Test.java

@@ -1,36 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-
-/**
- * Diese Klasse kann dazu verwenden die anderen
- * Klassen in diesem Package zu testen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Test {
-
-	/**
-	 * Die main methode
-	 * 
-	 * @param args
-	 */
-	public static void main(String[] args) {
-		Ellipse e = new Ellipse(new Point(0, 0), 10, 10);
-		System.out.println(e);
-
-		Circle c = new Circle(new Point(0, 0), 15);
-		System.out.println(c);
-
-		Polygon p = new Polygon(new Point(0, 0), new Point[] {
-			new Point(0, 0),
-			new Point(0, 4),
-			new Point(4, 4)
-		});
-		System.out.println(p);
-
-		Line l = new Line(new Point(0, 0), new Point(10, 10));
-		System.out.println(l);
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le05/Triangle.java

@@ -1,34 +1,0 @@
-package com.maximilianschulke.gdp2.le05;
-
-
-/**
- * Diese Klasse kann dazu verwendet werden den Flaecheninhalt
- * oder Umfang eines Dreiecks zu berechnen.
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class Triangle extends Polygon {
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param pos Position
-	 * @param a Punkt A
-	 * @param b Punkt B
-	 * @param c Punkt C
-	 */
-	public Triangle(Point pos, Point a, Point b, Point c) {
-		super(pos, new Point[] {a, b, c});
-	}
-	
-	/**
-	 * Formatiert die Instanz als String
-	 *
-	 * @return String-Darstellung
-	 */
-	public String toString()  {
-		return super.toString().replaceAll("Polygon", "Triangle");
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le06/CircularQueue.java

@@ -1,239 +1,0 @@
-package com.maximilianschulke.gdp2.le06;
-
-import java.util.ArrayList;
-
-
-/**
- * Eine Ringspeichers Queue-Implementation
- * 
- * @author Maximilian Schulke <schulke@th-brandenbug.de>
- * @version 1.0.0
- */
-public class CircularQueue<T> implements IQueue<T> {
-
-	/**
-	 * Die Maximalgröße
-	 */
-	private int size;
-
-	/**
-	 * Index des ersten Elementes
-	 */
-	public int head;
-
-	/**
-	 * Index des letzten Elementes
-	 */
-	public int tail;
-
-	/**
-	 * Liste mit allen Elementen
-	 */
-	private ArrayList<T> queue;
-
-
-	/**
-	 * Der Konstruktor
-	 * 
-	 * @param size Maximalgröße
-	 */
-	public CircularQueue(int size) {
-		head = 0;
-		tail = 0;
-		queue = new ArrayList<T>();
-		this.size = size;
-
-		for (int i = 0; i < size; i++) {
-			queue.add(null);
-		}
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public void push(T el) throws QueueException {
-		if (isFull()) {
-			head = (tail + 1) % size;
-		}
-
-		queue.set(tail, el);
-		tail = (tail + 1) % size;
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public T pop() throws QueueException {
-		if (isEmpty()) {
-			throw new QueueException("Queue is empty!");
-		}
-
-		T el = queue.set(head, null);
-		head = (head + 1) % size;
-
-		return el;
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public T head() throws QueueException {
-		if (queue.isEmpty()) {
-			throw new QueueException("Queue is empty!");
-		}
-
-		return queue.get(head);
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public T tail() throws QueueException {
-		if (queue.isEmpty()) {
-			throw new QueueException("Queue is empty!");
-		}
-
-		return queue.get(tail);
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public int length() {
-		int emptySlots = queue.stream().filter(el -> el == null).toArray().length;
-
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le06/IQueue.java

@@ -1,86 +1,0 @@
-package com.maximilianschulke.gdp2.le06;
-
-
-/**
- * Ein generisches Interface für alle Queue ähnlichen
- * Datenstrukturen
- * 
- * @author Maximilian Schulke <schulke@th-brandenbug.de>
- * @version 1.0.0
- */
-public interface IQueue<T> {
-
-	/**
-	 * Fügt ein Element ans Ende an
-	 * 
-	 * @param el Element
-	 * @throws QueueException
-	 */
-	public void push(T el) throws QueueException;
-
-	/**
-	 * Entfernt das erste Element
-	 * 
-	 * @return Entferntes Element
-	 * @throws QueueException
-	 */
-	public T pop() throws QueueException;
-
-	/**
-	 * Gibt das erste Element zurück
-	 * 
-	 * @return Erstes Element
-	 * @throws QueueException
-	 */
-	public T head() throws QueueException;
-
-	/**
-	 * Gibt das letztes Element zurück
-	 * 
-	 * @return Letztes Element
-	 * @throws QueueException
-	 */
-	public T tail() throws QueueException;
-
-	/**
-	 * Gibt die Länge der Queue zurück
-	 * 
-	 * @return Länge
-	 */
-	public int length();	
-
-	/**
-	 * Entfernt alle Element
-	 */
-	public void clear();
-
-	/**
-	 * Gibt an ob die Queue voll ist
-	 * 
-	 * @return Ist die Liste voll
-	 */
-	public boolean isFull();
-
-	/**
-	 * Gibt an ob die Queue leer ist
-	 * 
-	 * @return Ist die Queue leer?
-	 */
-	public boolean isEmpty();
-
-	/**
-	 * Gibt die Maximalgröße zurück
-	 * 
-	 * @return Maximalgröße
-	 */
-	public int getSize();
-
-	/**
-	 * Setzt die Maximalgröße
-	 * 
-	 * @param size Maximalgröße
-	 * @throws QueueException
-	 */
-	public void setSize(int size) throws QueueException;
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le06/LinearQueue.java

@@ -1,191 +1,0 @@
-package com.maximilianschulke.gdp2.le06;
-
-import java.util.ArrayList;
-
-
-/**
- * Eine lineare Queue-Implementation
- * 
- * @author Maximilian Schulke <schulke@th-brandenbug.de>
- * @version 1.0.0
- */
-public class LinearQueue<T> implements IQueue<T> {
-
-	/**
-	 * Maximalgröße der Queue
-	 */
-	private int size;
-
-	/**
-	 * Enthält die Elemente der Queue
-	 */
-	private ArrayList<T> queue;
-
-
-	/**
-	 * Der Konstruktor
-	 */
-	public LinearQueue() {
-		size = Integer.MAX_VALUE;
-		queue = new ArrayList<>();
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public void push(T el) throws QueueException {
-		if (isFull()) {
-			throw new QueueException("The queue is full!");
-		}
-
-		queue.add(el);	
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public T pop() throws QueueException {
-		if (queue.isEmpty()) {			
-			throw new QueueException("The queue is empty!");
-		}
-
-		T el = queue.get(0);
-		queue.remove(0);
-		return el;
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public T head() throws QueueException {
-		if (queue.isEmpty()) {
-			throw new QueueException("The queue is empty!");
-		}
-
-		return queue.get(0);
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public T tail() throws QueueException {
-		if (queue.isEmpty()) {
-			throw new QueueException("The queue is empty!");
-		}
-
-		return queue.get(queue.size() - 1);
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public int length() {
-		return queue.size();
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public boolean isFull() {
-		return queue.size() >= this.size;
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public boolean isEmpty() {
-		return queue.isEmpty();
-	}
-
-
-	/**
-	 * @see IQueue
-	 */
-	@Override
-	public void clear() {
-		queue.clear();
-	}
-
-
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le06/QueueException.java

@@ -1,14 +1,0 @@
-package com.maximilianschulke.gdp2.le06;
-
-
-/**
- * Eine QueueException fasst alle Ausnahmen die mit
- * dem Queues in Zusammenhang stehen zusammen
- * 
- * @author Maximilian Schulke <schulke@th-brandenbug.de>
- * @version 1.0.0
- */
-public class QueueException extends Exception {
-	public QueueException() {super();}
-	public QueueException(String message) {super(message);}
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le07/GUIBsp.java

@@ -1,139 +1,0 @@
-package com.maximilianschulke.gdp2.le07;
-
-import javafx.application.Application;
-import javafx.geometry.Insets;
-import javafx.scene.Scene;
-import javafx.scene.control.TextField;
-import javafx.scene.input.MouseEvent;
-import javafx.scene.control.Label;
-import javafx.scene.control.Button;
-import javafx.scene.layout.FlowPane;
-import javafx.scene.layout.VBox;
-import javafx.stage.Stage;
-
-
-/**
- * Eine Beispiel GUI
- * 
- * @version 1.0.0
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- */
-public class GUIBsp extends Application {
-
-	/**
-	 * Das Hello Label
-	 */
-	Label hello;
-
-	/**
-	 * Das Name Label
-	 */
-	Label name;
-
-	/**
-	 * Das Textfeld für die Eingabe
-	 */
-	TextField input;
-
-	/**
-	 * Das Label für das Textfeld
-	 */
-	Label label;
-
-	/**
-	 * Der Button um zu bestätigen
-	 */
-	Button ok;
-
-	/**
-	 * Der Button um abzubrechen
-	 */
-	Button cancel;
-
-
-	/**
-	 * Baut die GUI auf einer Stage
-	 * 
-	 * @param primaryStage Stage
-	 */
-	@Override
-	public void start(Stage primaryStage) throws Exception {
-		// Setze Window-Titel
-		primaryStage.setTitle("GUIBsp");
-
-		// Erstelle Labels
-		hello = new Label();
-		hello.setText("Hallo");
-		name = new Label();
-
-		// Erstelle Layouts und füge Ausgabe-Labels hinzu
-		VBox layout = new VBox();
-		layout.setPadding(new Insets(25, 25, 25, 25));
-		layout.setSpacing(10);
-
-		FlowPane outputPane = new FlowPane();
-		outputPane.setHgap(5);
-		outputPane.getChildren().addAll(hello, name);
-
-		layout.getChildren().add(outputPane);
-
-		// Erstelle Eingabe-Textfeld
-		label = new Label();
-		label.setText("Name:");
-		input = new TextField();
-
-		FlowPane inputPane = new FlowPane();
-		inputPane.setHgap(10);
-		inputPane.getChildren().addAll(label, input);
-
-		layout.getChildren().add(inputPane);
-
-		// Erstelle Buttons
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le09/GUIPaint.java

@@ -1,112 +1,0 @@
-package com.maximilianschulke.gdp2.le09;
-
-import javafx.application.Application;
-import javafx.stage.Stage;
-import javafx.scene.Group;
-import javafx.scene.Scene;
-import javafx.scene.paint.Color;
-import javafx.scene.shape.*;
-
-/**
- * Die GUIPaint Anwendung
- * 
- * @author Maximilian Schulke <schulke@th-brandenburg.de>
- * @version 1.0.0
- */
-public class GUIPaint extends Application {
-
-	/**
-	 * Variablen zum tracken der Rechteck-Position.
-	 */
-	public double x0,x1,y0,y1;
-
-	/**
-	 * Die Start Methode zeichnet das Haus und registriert
-	 * die EventHandler.
-	 * 
-	 * @param primaryStage Die Bühne
-	 */
-	@Override
-	public void start(Stage primaryStage) {
-		try {
-			Group root = new Group();
-			Scene scene = new Scene(root, 600, 400);
-
-			// Definiert Window-Verhalten
-			primaryStage.setScene(scene);
-			primaryStage.setTitle("GUIPaint");
-			primaryStage.setMinWidth(600);
-			primaryStage.setMinHeight(400);
-			primaryStage.show();
-
-			// Hintergrund
-			Rectangle sky = new Rectangle(0, 000, 600, 300);
-			sky.setFill(Color.LIGHTBLUE);
-
-			Rectangle earth = new Rectangle(0, 300, 600, 100);
-			earth.setFill(Color.GREEN);
-
-			Group house = new Group();
-
-			// Das Haus
-			Polygon roof = new Polygon(200.0, 150.0, 300.0, 100.0, 400.0, 150.0);
-			roof.setFill(Color.ORANGE);
-
-			Rectangle body = new Rectangle(200, 150, 200, 150);
-			body.setFill(Color.BEIGE);
-
-			Rectangle window1 = new Rectangle(237.5, 175, 25, 40);
-			window1.setFill(Color.WHITE);
-
-			Rectangle window2 = new Rectangle(237.5, 225, 25, 40);
-			window2.setFill(Color.WHITE);
-
-			Rectangle door = new Rectangle(285, 250, 30, 50);
-			door.setFill(Color.SADDLEBROWN);
-
-			Rectangle window3 = new Rectangle(337.5, 175, 25, 40);
-			window3.setFill(Color.WHITE);
-
-			Rectangle window4 = new Rectangle(337.5, 225, 25, 40);
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/Controller.java

@@ -1,130 +1,0 @@
-package com.maximilianschulke.gdp2.le10;
-
-import com.maximilianschulke.gdp2.le10.model.*;
-
-import javafx.fxml.FXML;
-import javafx.collections.ListChangeListener;
-import javafx.event.ActionEvent;
-import javafx.scene.input.MouseEvent;
-import javafx.scene.input.MouseButton;
-import javafx.scene.layout.BorderPane;
-import javafx.scene.control.ChoiceBox;
-import javafx.scene.control.ColorPicker;
-import javafx.scene.control.Label;
-import javafx.scene.layout.Pane;
-
-
-public class Controller {
-
-	private Drawing drawing = new Drawing(new Figure[] {});  
-	private Figure figure;
-	private Point start;
-
-	@FXML private BorderPane layout;
-	@FXML private Pane root;
-	@FXML private ChoiceBox<String> typeBox;
-	@FXML private ColorPicker colorPicker;
-	@FXML private Label mousePosition;
-	@FXML private Label area;
-	@FXML private Label perimeter;
-
-
-	@FXML public void initialize() {
-		drawing.getList().addListener(new ListChangeListener<Figure>() {
-			@Override
-	        public void onChanged(ListChangeListener.Change<? extends Figure> c) {
-	            if (c.next()){
-	            	root.getChildren().clear();
-	            	c.getList().stream().forEach(f -> {
-	            		root.getChildren().add(f.intoShape());
-	            	});
-            	}
-	        }
-	    });
-	}
-
-
-	@FXML protected void onMouseDown(MouseEvent ev) {
-		if (!ev.isPrimaryButtonDown()) return;
-
-		start = new Point(ev.getX(), ev.getY());
-		figure = FigureType.fromString(typeBox.getValue()).intoFigure(start, colorPicker.getValue());
-		drawing.add(figure);
-	}
-
-
-	@FXML protected void onMouseDragged(MouseEvent ev) {
-		if (!ev.isPrimaryButtonDown()) return;
-
-		onMouseMoved(ev);
-
-		Point current = new Point(ev.getX(), ev.getY());
-
-		switch (FigureType.fromString(typeBox.getValue())) {
-			case CIRCLE:
-				((Circle) figure).setRadius(Math.abs(new Line(start, current).length()));
-				break;
-			case ELLIPSE:
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/Main.java

@@ -1,29 +1,0 @@
-package com.maximilianschulke.gdp2.le10;
-
-import javafx.application.Application;
-import javafx.fxml.FXMLLoader;
-import javafx.scene.Parent;
-import javafx.scene.Scene;
-import javafx.stage.Stage;
-
-public class Main extends Application {
-
-	@Override
-	public void start(Stage stage) throws Exception {
-		Parent root = FXMLLoader.load(getClass().getResource("view.fxml"));
-
-		Scene scene = new Scene(root, 800, 600);
-
-		stage.setTitle("Drawing App");
-		stage.setScene(scene);
-		stage.setMinWidth(800);
-		stage.setMinHeight(600);
-		stage.show();
-	}
-
-
-	public static void main(String[] args) {
-		launch(args);
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Circle.java

@@ -1,42 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-
-public class Circle extends Ellipse {	
-
-	public Circle(Point pos) {
-		super(pos);
-	}
-
-
-	public Circle(Point pos, Paint color) {
-		super(pos, color);
-	}
-
-
-	public Circle(Point pos, double radius) {
-		super(pos, radius, radius);
-	}
-
-
-	public double circumference() {
-		return perimeter();
-	}
-
-
-	public String toString() {
-		return super.toString().replaceAll("Ellipse", "Circle");
-	}
-
-
-	public double getRadius() {
-		return getRadiusX();
-	}
-
-
-	public void setRadius(double radius) {
-		setRadiusX(radius);
-		setRadiusY(radius);
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Drawing.java

@@ -1,45 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.collections.FXCollections;
-import javafx.collections.ObservableList;
-
-
-public class Drawing {
-
-	private ObservableList<Figure> figures;
-
-
-	public Drawing(Figure[] figs) {
-		this.figures = FXCollections.observableArrayList();
-
-		for (Figure f : figs) {
-			add(f);
-		}
-	}
-
-
-	public void add(Figure f) {
-		figures.add(f);
-	}
-
-
-	public void remove(Figure f) {
-		figures.remove(f);
-	}
-
-
-	public void update(Figure f) {
-		figures.set(figures.indexOf(f), f);
-	}
-
-
-	public void clear() {
-		figures.clear();
-	}
-
-
-	public ObservableList<Figure> getList() {
-		return figures;
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Ellipse.java

@@ -1,77 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-import javafx.scene.shape.Shape;
-
-public class Ellipse extends Figure implements Sizeable {
-
-	private double radiusX;
-	private double radiusY;
-
-
-	public Ellipse(Point pos) {
-		super(pos);
-		setRadiusX(0);
-		setRadiusY(0);
-	}
-	
-
-	public Ellipse(Point pos, Paint color) {
-		this(pos);
-		setColor(color);
-	}
-
-
-	public Ellipse(Point pos, double rx, double ry) {
-		super(pos);
-		setRadiusX(rx);
-		setRadiusY(ry);
-	}
-
-
-	public double area() {
-		return radiusX * radiusY * Math.PI;
-	}
-
-
-	public double perimeter() {
-		if (radiusX == 0 && radiusY == 0) return 0;
-
-		double h = Math.pow(radiusX - radiusY, 2) / Math.pow(radiusX + radiusY, 2);
-
-		return Math.PI * (radiusX + radiusY) * (1 + 3 * h / (10 + Math.sqrt(4 - 3 * h)));
-	}
-
-
-	public String toString() {
-		return "Ellipse [ Pos " + getPos() + ", Color " + getColor() + ", RadiusX " + getRadiusX() + ", RadiusY " + getRadiusY() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
-	}
-
-
-	public double getRadiusX() {
-		return radiusX;
-	}
-
-
-	public void setRadiusX(double radiusX) {
-		this.radiusX = radiusX;
-	}
-
-
-	public double getRadiusY() {
-		return radiusY;
-	}
-
-
-	public void setRadiusY(double radiusY) {
-		this.radiusY = radiusY;
-	}
-
-
-	public Shape intoShape() {
-		Shape shape = new javafx.scene.shape.Ellipse(getPos().getX(), getPos().getY(), getRadiusX(), getRadiusY());
-		shape.setFill(getColor());
-		return shape;
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Figure.java

@@ -1,43 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-import javafx.scene.shape.Shape;
-
-public abstract class Figure {
-
-	private Point pos;
-	private Paint color;
-
-	public Figure(Point pos) {
-		setPos(pos);
-	}
-
-	public Figure(Point pos, Paint color) {
-		this(pos);
-		setColor(color);
-	}
-
-
-	abstract public Shape intoShape();
-
-
-	public Point getPos() {
-		return pos;
-	}
-
-
-	public void setPos(Point pos) {
-		this.pos = pos;
-	}
-
-
-	public Paint getColor() {
-		return color;
-	}
-
-
-	public void setColor(Paint color) {
-		this.color = color;
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/FigureType.java

@@ -1,68 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-
-public enum FigureType {
-	CIRCLE,
-	ELLIPSE,
-	LINE,
-	RECTANGLE,
-	SQUARE,
-	POLYGON,
-	TRIANGLE;
-
-	public static FigureType fromString(String id) {
-		switch (id.toLowerCase()) {
-			case "circle":
-				return CIRCLE;
-			case "ellipse":
-				return ELLIPSE;
-			case "line":
-				return LINE;
-			case "rectangle":
-				return RECTANGLE;
-			case "square":
-				return SQUARE;
-			case "polygon":
-				return POLYGON;
-			case "triangle":
-				return TRIANGLE;
-		}
-
-		return null;
-	}
-
-
-	public static FigureType fromFigure(Figure f) {
-		if (f instanceof Circle) return CIRCLE;
-		else if (f instanceof Ellipse) return ELLIPSE;
-		else if (f instanceof Line) return LINE;
-		else if (f instanceof Square) return SQUARE;
-		else if (f instanceof Rectangle) return RECTANGLE;
-		else if (f instanceof Triangle) return TRIANGLE;
-		else if (f instanceof Polygon) return POLYGON;
-		else return null;
-	}
-
-
-	public Figure intoFigure(Point pos, Paint color) {
-		switch (this) {
-			case CIRCLE:
-				return new Circle(pos, color);
-			case ELLIPSE:
-				return new Ellipse(pos, color);
-			case LINE:
-				return new Line(pos, color);
-			case RECTANGLE:
-				return new Rectangle(pos, color);
-			case SQUARE:
-				return new Square(pos, color);
-			case POLYGON:
-				return new Polygon(pos, color);
-			case TRIANGLE:
-				return new Triangle(pos, color);
-		}
-
-		return null;
-	}
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Line.java

@@ -1,58 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-import javafx.scene.shape.Shape;
-
-public class Line extends Figure {
-
-	private Point dest;
-
-
-	public Line(Point pos) {
-		super(pos);
-		setDest(new Point(pos.getX(), pos.getY()));
-	}
-
-	
-	public Line(Point pos, Paint color) {
-		this(pos);
-		setColor(color);
-	}
-
-
-	public Line(Point start, Point dest) {
-		super(start);
-		setDest(dest);
-	}
-
-
-	public double length() {
-		double xdiff = getPos().xDistanceTo(getDest());
-		double ydiff = getPos().yDistanceTo(getDest());
-
-		return Math.sqrt(Math.pow(xdiff, 2) + Math.pow(ydiff, 2));
-	}
-
-
-	public String toString() {
-		return "Line [ Pos " + getPos() + ", Color " + getColor() + ", Dest " + getDest() + ", Length " + length() + " ]";
-	}
-
-
-	public Point getDest() {
-		return dest;
-	}
-
-
-	public void setDest(Point dest) {
-		this.dest = dest;
-	}
-
-
-	public Shape intoShape() {
-		Shape shape = new javafx.scene.shape.Line(getPos().getX(), getPos().getY(), getDest().getX(), getDest().getY());
-		shape.setStroke(getColor());
-		return shape;
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Point.java

@@ -1,49 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-public class Point {
-
-	private double x;
-	private double y;
-
-
-	public Point (double x, double y) {
-		this.setX(x);
-		this.setY(y);
-	}
-
-
-	public double xDistanceTo(Point p) {
-		return p.getX() - this.getX();
-	}
-
-
-	public double yDistanceTo(Point p) {
-		return p.getY() - this.getY();
-	}
-
-
-	public double getX() {
-		return x;
-	}
-
-
-	public void setX(double x) {
-		this.x = x;
-	}
-
-
-	public double getY() {
-		return y;
-	}
-
-
-	public void setY(double y) {
-		this.y = y;
-	}
-
-
-	public String toString() {
-		return "[" + getX() + "; " + getY() + "]";
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Polygon.java

@@ -1,112 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-import javafx.scene.shape.Shape;
-
-public class Polygon extends Figure implements Sizeable {
-
-	private Point[] points;
-
-
-	public Polygon(Point pos) {
-		super(pos);
-		setPoints(new Point[] {});
-	}
-	
-
-	public Polygon(Point pos, Paint color) {
-		this(pos);
-		setColor(color);
-	}
-
-
-	public Polygon(Point pos, Point[] points) {
-		super(pos);
-		setPoints(points);
-	}
-
-
-	private Point nextPoint(int index) {
-		return points[index == points.length - 1 ? 0 : index + 1];
-	}
-
-
-	/**
-	 * @see https://www.mathopenref.com/coordpolygonarea.html
-	 * @return Area of the Polygon
-	 */
-	public double area() {
-		double accum = 0.0;
-
-		for (int i = 0; i < points.length; i++) {
-			Point current = points[i];
-			Point next = nextPoint(i);
-
-			accum += (current.getX() * next.getY() - current.getY() * next.getX());
-		}
-
-		return Math.abs(accum / 2);
-	}
-
-
-	/**
-	 * @see https://www.mathopenref.com/polygonperimeter.html
-	 * @return Perimeter of the Polygon
-	 */
-	public double perimeter() {
-		double accum = 0;
-
-		for (int i = 0; i < points.length; i++) {
-			Point current = points[i];
-			Point next = nextPoint(i);
-			
-			Line vec = new Line(
-				new Point(0, 0),
-				new Point(current.xDistanceTo(next), current.yDistanceTo(next))
-			);
-			
-			accum += vec.length();
-		}
-
-		return accum;
-	}
-
-
-	public String toString() {
-		String s = "Polygon [ Pos " + getPos() + ", Color " + getColor() + ", Points [ ";
-
-		for (Point p : points) {
-			s += p + ", ";
-		}
-
-		s += " ], Area " + area() + ", Perimeter " + perimeter() + " ]";
-		return s;
-	}
-
-
-	public Point[] getPoints() {
-		return points;
-	}
-
-
-	public void setPoints(Point[] points) {
-		this.points = points;
-	}
-
-
-	public Shape intoShape() {
-		double[] coords = new double[] {};
-		int i = 0;
-
-		for (Point point : points) {
[…]

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Rectangle.java

@@ -1,73 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-import javafx.scene.shape.Shape;
-
-public class Rectangle extends Figure implements Sizeable {
-
-	private double a;
-	private double b;
-
-
-	public Rectangle(Point pos) {
-		super(pos);
-		setA(0);
-		setB(0);
-	}
-
-	
-	public Rectangle(Point pos, Paint color) {
-		this(pos);
-		setColor(color);
-	}
-	
-
-	public Rectangle(Point pos, double a, double b) {
-		super(pos);
-		setA(a);
-		setB(b);
-	}
-
-
-	public double area() {
-		return a * b;
-	}
-
-
-	public double perimeter() {
-		return a + a + b + b;
-	}
-
-
-	public double getA() {
-		return a;
-	}
-
-
-	public void setA(double newA) {
-		a = newA;
-	}
-
-
-	public double getB() {
-		return b;
-	}
-
-
-	public void setB(double newB) {
-		b = newB;
-	}
-
-
-	public String toString() {
-		return "Rectangle [ Pos " + getPos() + ", Color " + getColor() + ", A " + getA() + ", B " + getB() + ", Area " + area() + ", Perimeter " + perimeter() + " ]";
-	}
-
-
-	public Shape intoShape() {
-		Shape shape = new javafx.scene.shape.Rectangle(getPos().getX(), getPos().getY(), getA(), getB());
-		shape.setFill(getColor());
-		return shape;
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Sizeable.java

@@ -1,8 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-public interface Sizeable {
-
-	public double area();
-	public double perimeter();
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Square.java

@@ -1,26 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-
-public class Square extends Rectangle {
-
-	public Square(Point pos) {
-		super(pos);
-	}
-	
-	
-	public Square(Point pos, Paint color) {
-		super(pos, color);
-	}
-
-
-	public Square(Point pos, double a) {
-		super(pos, a, a);
-	}
-
-
-	public String toString() {
-		return super.toString().replaceAll("Rectangle", "Square");
-	}
-
-}

### 2022-06-06 grundlagen-der-programmierung-2/src/com/maximilianschulke/gdp2/le10/model/Triangle.java

@@ -1,26 +1,0 @@
-package com.maximilianschulke.gdp2.le10.model;
-
-import javafx.scene.paint.Paint;
-
-public class Triangle extends Polygon {
-
-	public Triangle(Point pos) {
-		super(pos);
-	}
-
-
-	public Triangle(Point pos, Paint color) {
-		super(pos, color);
-	}
-
-
-	public Triangle(Point pos, Point a, Point b, Point c) {
-		super(pos, new Point[] {a, b, c});
-	}
-
-
-	public String toString() {
-		return super.toString().replaceAll("Polygon", "Triangle");
-	}
-
-}

### 2022-06-06 it-forensik/sichere-datenlöschung.tex
Add a basic structure for the it forensics slides
Add more content to the it forensics slides

@@
 
+\section{Wieso ist eine sichere Löschung so wichtig?}
+\begin{frame}{Wieso ist eine sichere Löschung so wichtig?}
+	\begin{itemize}
+		\item verhinderung von Datendiebstahl
+		\item rechtliche Vorgaben
+		\item Hardware wiederverwendbar machen
+	\end{itemize}
+\end{frame}
+
 \section{Gefahren durch falsche Löschung}
@@
 
-\subsection{Wie funktioniert die Rekonstruktion von Datenseichern?}
+\section{Wie funktioniert die Rekonstruktion von Datenseichern?}
 \begin{frame}{Wie funktioniert die Rekonstruktion von Datenseichern?}
+	\begin{itemize}
+		\item Rekonstruktion ist möglich wenn Daten nicht physisch gelöscht
+			wurden
+			\begin{itemize}
+				\item Formatierung entfernt Daten nur oberflächlich da i.d.R.
+					nur die Zugriffstabellen angepasst werden und nicht die
+					Register geleert werden.
+			\end{itemize}
+		\item Durch durchsuchen der Register einer Festplatten mit entfernter
+			Zugriffstabelle lassen sich softwareseitig Datenrekonstruieren
+		\item Bei Beschädigten oder mittels SMR gesicherten Festplatten ist
+			eine softwareseitige Rekonstruktion oft nicht mehr möglich, hier
+			gibt es allerdings noch die Möglichkeit über physischen Zugriff auf
+			die Festplatte Informationen wiederherzustellen (bspw.\ von
+			spezialisierten Unternhemen unter Reinraum-Bedingungen)
+	\end{itemize}
+\end{frame}
+
+\section{Methoden zur sicheren Speicherlöschung}
+\begin{frame}{Methoden zur sicheren Speicherlöschung}
+	\begin{itemize}
+		\item Physische Zerstörung des Datenträgers
+		\item Überschreiben der Daten innherlab des Datenträgers
+	\end{itemize}
+\end{frame}
+
+\section{Methoden zur sicheren Speicherlöschung: A}
+\begin{frame}{Methoden zur sicheren Speicherlöschung: A}
 	\begin{itemize}
@@
 
-\subsection{Funktionsweise einer HDD}
-\begin{frame}{Funktionsweise einer HDD}
+\section{Unterschiede zwischen Speichertypen}
+\begin{frame}{Unterschiede zwischen Speichertypen}
 	\begin{itemize}
@@
 
-\subsection{Sichere Löschung einer HDD}
-\begin{frame}{Sichere Löschung einer HDD}
-	\begin{itemize}
-		\item 
-	\end{itemize}
-\end{frame}
-
-\subsection{Funktionsweise einer SSD}
-\begin{frame}{Funktionsweise einer SSD}
-	\begin{itemize}
-		\item 
-	\end{itemize}
-\end{frame}
-
-\subsection{Sichere Löschung einer SSD}
-\begin{frame}{Sichere Löschung einer SSD}
-	\begin{itemize}
-		\item 
-	\end{itemize}
-\end{frame}
-
-\subsection{Funktionsweise einer SSHD}
-\begin{frame}{Funktionsweise einer SSHD}
-	\begin{itemize}
-		\item 
-	\end{itemize}
-\end{frame}
-
-\subsection{Sichere Löschung einer SSHD}
-\begin{frame}{Sichere Löschung einer SSHD}
+\section{Shingled Magnetic Recording und der TRIM-Befehl}
+\begin{frame}{Shingled Magnetic Recording und der TRIM-Befehl}
 	\begin{itemize}
@@
 % https://www.easeus.de/festplatte-wiederherstellen/formatierte-festplatte-wiederherstellen-freeware.html
+% https://de.wikipedia.org/wiki/Datenwiederherstellung
 

### 2022-06-19 it-forensik/sichere-datenlöschung.tex
Add standards for data erasure

@@
 
-\section{Methoden zur sicheren Speicherlöschung: A}
-\begin{frame}{Methoden zur sicheren Speicherlöschung: A}
+\section{Überschreiben gesicherter Bereiche}
+\begin{frame}{Überschreiben gesicherter Bereiche}
+	\begin{itemize}
+		\item Um vollständige Sicherheit zu gewährleisten müss sämtliche
+			Bereiche eines Datenträgers überschrieben werden
+		\item Zu typischen gesicherten Bereichen zählen die HPA (Host protected
+			area) und das DCO (device configuration overlay)
+		\item Programme zur Datenlöschung die vom Betriebssystem des Nutzers
+			ausgeführt werden haben oft keinen uneingeschränkten Zugriff da das
+			Betriebssyystem während der Löschung noch Teile des Datenträgers
+			verwendet
+	\end{itemize}
+\end{frame}
+
+\section{Übersicht etablierter Standards}
+\begin{frame}{Übersicht etablierter Standards}
+	\begin{itemize}
+		\item Cryptographic Erasure (Crypto Erase)
+			\begin{itemize}
+				\item Keine Runden
+				\item Löscht den kryptografischen Schlüssel mit der die zu
+					löschenden Daten hardwareseitig verschlüsselt wurden
+				\item Funktioniert nicht mit jeder Hardware
+			\end{itemize}
+		\item Peter Gutmann's Algorithm
+			\begin{itemize}
+				\item Veröffentlicht: 1996
+				\item 1 bis 35 Runden
+				\item Muster: Kombination aus allen anderen Standards
+			\end{itemize}
+		\item Bruce Scheier's Algorithm
+			\begin{itemize}
+				\item Veröffentlicht: 1996
+				\item 7 Runden
+				\item Muster: 1, 0, 5x pseudozufällige Sequenz
+			\end{itemize}
+		\item U.S. Navy Staff Office Publication NAVSO P-5239-26, 1993
+			\begin{itemize}
+				\item Veröffentlicht: 1993
+				\item 3 Runden
+				\item Muster: Zeichen, Komplement, Zufallswert
+			\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Übersicht etablierter Standards}
+	\begin{itemize}
+		\item BSI-2011-VS	
+			\begin{itemize}
+				\item Veröffentlicht: 2011
+				\item 4 Runden
+				\item Verschlüsselung mit AES-128-CBC mit anschließender
+					löschung des Schlüssls
+			\end{itemize}
+		\item DoD 5220.22-M (E)
+			\begin{itemize}
+				\item Veröffentlicht: 1995
+				\item 3 Runden
+				\item Muster: 0, 1, Zufallswerte
+			\end{itemize}
+		\item DoD 5220.22-M (C)
+			\begin{itemize}
+				\item 1 Runden
+				\item Muster: aperiodische Zufallswerte
+			\end{itemize}
+		\item DoD 5220.22-M (ECE)
+			\begin{itemize}
+				\item Veröffentlicht: 2001
+				\item 7 Runde
+				\item Muster: DoD 5220.22-M (E), DoD 5220.22-M (C),
+					DoD 5220.22-M (E)
+			\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Übersicht etablierter Standards}
+	\begin{itemize}
+		\item U.S. Air Force System Security Instruction 5020
+			\begin{itemize}
+				\item Veröffentlicht: 1993
+				\item 3 Runden
+				\item Muster: 1, 0, zufälliges Zeichen
+			\end{itemize}
+		\item HMG Infosec Standard 5, Lower Standard	
+			\begin{itemize}
+				\item 1 Runden
+				\item Muster: Zufällige Zeichenkette
+			\end{itemize}
+		\item HMG Infosec Standard 5, Higher Standard	
+			\begin{itemize}
+				\item 3 Runde
+				\item Muster: 0, 1, Zufallswert
+			\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\section{Vorteile kryptografiebasiereter Standards}
+\begin{frame}{Vorteile kryptografiebasiereter Standards}
 	\begin{itemize}
@@
 
-\section{Typsiche Speichertypen}
-\begin{frame}{Typsiche Speichertypen}
-	\begin{itemize}
-		\item HDD (Hard Disk Drive)
-		\item SSD (Solid State Drive)
-		\item SSHD (Solid State Hybrid Drive)
-	\end{itemize}
-\end{frame}
+% eine weitere folie und quellen!
 
-\section{Unterschiede zwischen Speichertypen}
-\begin{frame}{Unterschiede zwischen Speichertypen}
-	\begin{itemize}
-		\item 
-	\end{itemize}
-\end{frame}
-
-\section{Shingled Magnetic Recording und der TRIM-Befehl}
-\begin{frame}{Shingled Magnetic Recording und der TRIM-Befehl}
+\section{Quellen}
+\begin{frame}{Quellen}
 	\begin{itemize}
@@
 % https://de.wikipedia.org/wiki/Datenwiederherstellung
+% https://www.blancco.com/resources/blog-comprehensive-list-data-wiping-erasure-standards/
+% https://www.archicrypt.de/shredder7/bsi-2011-vs.html
 

### 2022-06-19 it-forensik/sichere-datenlöschung.tex
Add a new frame

@@
 
-\section{Vorteile kryptografiebasiereter Standards}
-\begin{frame}{Vorteile kryptografiebasiereter Standards}
+\section{Vorteile kryptografiebasiereter Datenlöschung}
+\begin{frame}{Vor- und Nachteile kryptografiebasiereter Datenlöschung}
+	\large{Vorteile:}
 	\begin{itemize}
+		\item Konstanter Aufwand (d.h.\ unabhängig von der Festplattengröße)
+		\item Impliziert Festplattenverschlüsselung, förderlich für die
+			allgemeine Sicherheit
+		\item Resistent gegen alle bekannten und zukünftigen
+			Datenwiederherstellungsverfahren
+	\end{itemize}
+
+	\large{Nachteile:}
+
+	\begin{itemize}
+		\item Sicherheit der Daten ist abhängig davon dass die Verschlüsselung
+			nicht gebrochen wird
 		\item 

### 2022-06-19 it-forensik/sichere-datenlöschung.tex
Add the last slide

@@
 			nicht gebrochen wird
-		\item 
 	\end{itemize}
@@
 
-% eine weitere folie und quellen!
-
-\section{Quellen}
-\begin{frame}{Quellen}
+\section{Etablierte Programme zur Datenlöschung}
+\begin{frame}{Etablierte Programme zur Datenlöschung}
 	\begin{itemize}
@@
 
-% https://www.bsi.bund.de/DE/Themen/Verbraucherinnen-und-Verbraucher/Informationen-und-Empfehlungen/Cyber-Sicherheitsempfehlungen/Daten-sichern-verschluesseln-und-loeschen/Daten-endgueltig-loeschen/daten-endgueltig-loeschen_node.html
-% https://www.experte.de/it-sicherheit/festplatte-loeschen
-% https://www.easeus.de/festplatte-wiederherstellen/formatierte-festplatte-wiederherstellen-freeware.html
-% https://de.wikipedia.org/wiki/Datenwiederherstellung
-% https://www.blancco.com/resources/blog-comprehensive-list-data-wiping-erasure-standards/
-% https://www.archicrypt.de/shredder7/bsi-2011-vs.html
+\section{Quellen}
+\begin{frame}{Quellen}
+	\begin{itemize}
+		\item %https://www.bsi.bund.de/DE/Themen/Verbraucherinnen-und-Verbraucher/Informationen-und-Empfehlungen/Cyber-Sicherheitsempfehlungen/Daten-sichern-verschluesseln-und-loeschen/Daten-endgueltig-loeschen/daten-endgueltig-loeschen_node.html
+		\item %https://www.experte.de/it-sicherheit/festplatte-loeschen
+		\item %https://www.easeus.de/festplatte-wiederherstellen/formatierte-festplatte-wiederherstellen-freeware.html
+		\item %https://de.wikipedia.org/wiki/Datenwiederherstellung
+		\item %https://www.blancco.com/resources/blog-comprehensive-list-data-wiping-erasure-standards/
+		\item %https://www.archicrypt.de/shredder7/bsi-2011-vs.html
+	\end{itemize}
+\end{frame}
 

### 2022-06-25 it-forensik/sichere-datenlöschung.tex
Fix typo
Finish the it forensics slides

@@
 \usepackage[T1]{fontenc}
+\usepackage{hyperref}
 
@@
 
-\section{Wie funktioniert die Rekonstruktion von Datenseichern?}
-\begin{frame}{Wie funktioniert die Rekonstruktion von Datenseichern?}
+\section{Wie funktioniert die Rekonstruktion von Datenspeichern?}
+\begin{frame}{Wie funktioniert die Rekonstruktion von Datenspeichern?}
 	\begin{itemize}
-		\item Rekonstruktion ist möglich wenn Daten nicht physisch gelöscht
+		\item Rekonstruktion ist möglich, wenn Daten nicht physisch gelöscht
 			wurden
@@
 			\end{itemize}
-		\item Durch durchsuchen der Register einer Festplatten mit entfernter
+		\item Durch Durchsuchen der Register einer Festplatten mit entfernter
 			Zugriffstabelle lassen sich softwareseitig Datenrekonstruieren
@@
 			die Festplatte Informationen wiederherzustellen (bspw.\ von
-			spezialisierten Unternhemen unter Reinraum-Bedingungen)
+			spezialisierten Unternehmen unter Reinraum-Bedingungen)
 	\end{itemize}
@@
 		\item Physische Zerstörung des Datenträgers
-		\item Überschreiben der Daten innherlab des Datenträgers
+		\item Überschreiben der Daten innerhalb des Datenträgers
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Um vollständige Sicherheit zu gewährleisten müss sämtliche
+		\item Um vollständige Sicherheit zu gewährleisten, muss sämtliche
 			Bereiche eines Datenträgers überschrieben werden
@@
 			area) und das DCO (device configuration overlay)
-		\item Programme zur Datenlöschung die vom Betriebssystem des Nutzers
-			ausgeführt werden haben oft keinen uneingeschränkten Zugriff da das
-			Betriebssyystem während der Löschung noch Teile des Datenträgers
+		\item Programme zur Datenlöschung, die vom Betriebssystem des Nutzers
+			ausgeführt werden haben oft keinen uneingeschränkten Zugriff, da das
+			Betriebssystem während der Löschung noch Teile des Datenträgers
 			verwendet
@@
 				\item Keine Runden
-				\item Löscht den kryptografischen Schlüssel mit der die zu
+				\item Löscht den kryptografischen Schlüssel, mit der die zu
 					löschenden Daten hardwareseitig verschlüsselt wurden
@@
 	\begin{itemize}
-		\item Sicherheit der Daten ist abhängig davon dass die Verschlüsselung
+		\item Sicherheit der Daten ist abhängig davon, dass die Verschlüsselung
 			nicht gebrochen wird
@@
 	\begin{itemize}
-		\item
+		\item DBAN
+		\item EaseUS
+		\item Blancco
+		\item DiskWipe
 	\end{itemize}
@@
 		\begin{itemize}
-		\item %https://www.bsi.bund.de/DE/Themen/Verbraucherinnen-und-Verbraucher/Informationen-und-Empfehlungen/Cyber-Sicherheitsempfehlungen/Daten-sichern-verschluesseln-und-loeschen/Daten-endgueltig-loeschen/daten-endgueltig-loeschen_node.html
-		\item %https://www.experte.de/it-sicherheit/festplatte-loeschen
-		\item %https://www.easeus.de/festplatte-wiederherstellen/formatierte-festplatte-wiederherstellen-freeware.html
-		\item %https://de.wikipedia.org/wiki/Datenwiederherstellung
-		\item %https://www.blancco.com/resources/blog-comprehensive-list-data-wiping-erasure-standards/
-		\item %https://www.archicrypt.de/shredder7/bsi-2011-vs.html
+			\item \begin{scriptsize}
+				BSI.\ (2021, 28. September). Daten auf Festplatten und Smartphones
+				endgültig Löschen. Bundesamt für Sicherheit in der
+				Informationstechnik. Abgerufen am 25. Juni, 2022, von
+				\href{https://www.bsi.bund.de/DE/Themen/Verbraucherinnen-und-Verbraucher/Informationen-und-Empfehlungen/Cyber-Sicherheitsempfehlungen/Daten-sichern-verschluesseln-und-loeschen/Daten-endgueltig-loeschen/daten-endgueltig-loeschen_node.html}{https://www.bsi.bund.de/DE/Themen/Verbrauche..}
+			\end{scriptsize}
+			\item \begin{scriptsize}
+				Köhler, T.\ (2021, August 20).\ HDD und SSD - Festplatte Sicher
+				Löschen.\ HDD und SSD - Festplatte sicher löschen | experte.de.
+				Abgerufen am 25. Juni, 2022, von
+				https://www.experte.de/it-sicherheit/festplatte-loeschen
+			\end{scriptsize}
+			\item \begin{scriptsize}
+				EaseUS.\ (2022,\ 19. Januar).\ Formatierte Festplatte
+				Wiederherstellen - so geht's.\ EaseUS.\ Abgerufen am 25. Juni, 2022,
+				von
+				https://www.easeus.de/festplatte-wiederherstellen/formatierte-festplatte-wiederherstellen-freeware.html 
+			\end{scriptsize}
+			\item \begin{scriptsize}
+				Blancco. (2022, 19. Mai). A comprehensive list of data wiping and
+				erasure standards. Blancco. Abgerufen am 25. Juni, 2022, von
+				https://www.blancco.com/resources/blog-comprehensive-list-data-wiping-erasure-standards/
+			\end{scriptsize}
+			\item \begin{scriptsize}
+				Remus, Dipl.-Ing. Patric. Archicrypt Shredder 7. BSI-2011-VS.\
+				Abgerufen am 25. Juni, 2022, from
+				https://www.archicrypt.de/shredder7/bsi-2011-vs.html
+			\end{scriptsize}
 		\end{itemize}

### 2022-06-25 hardware-sicherheit/slides.tex
Add slides for hardware security

\documentclass{beamer}

\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}

\usecolortheme{seagull}
\usefonttheme{serif}

\title{Proof-of-Concept-Implementierung einer Anwendung mit FIDO2 Support}
\author{Mara Schulke\\\tiny{Matrikelnr. 20215853, SS22 B.Sc. IT Security, THB}}

\begin{document}

\begin{frame}
	Hardware Sicherheit
	\vspace{1em}
	\titlepage
\end{frame}

\begin{frame}{Übersicht}
	\tableofcontents
\end{frame}

\section{Der FIDO2 Standard}
\begin{frame}{Der FIDO2 Standard}
	\begin{itemize}
		\item Industriestandard für hardwaregestützte Authentifizierung
		\item Setzt sich im wesentlichen aus zwei
			Protokollen zusammen:
			\begin{itemize}
				\item CTAP
				\item WebAuthn
			\end{itemize}
		\item Der Hardwaretoken kann innerhalb des Systems verbaut sein (z.B.
			über ein TPM) oder ein externes Gerät sein (bspw. Yubikey oder
			Google Titan)
		\item Quelloffen und nicht proprietär
			\begin{itemize}
				\item Durch Projekte wie OpenSK lassen sich sogar eigene
					Hardwaretokens bauen
			\end{itemize}
	\end{itemize}
\end{frame}

\section{Vorteile gegenüber einfacher asymmetrischer Kryptografie}
\begin{frame}{Vorteile gegenüber einfacher asymmetrischer Kryptografie}
	\begin{itemize}
		\item Erhöhte Sicherheit durch separate Schlüsselpaare pro Relying Party
		\item Verlust eines Schlüsselpaars wäre verhältnismäßig unkritisch da
			dieses nur Zugriff auf einen Online-Dienst gibt
		\item Schlüsselpaare werden nicht auf dem System des Nutzers gespeichert
			\begin{itemize}
				\item Kompromitierte Verbindungen und Nutzersysteme stellen keine schwerwiegende Gefahr dar
				\item Verlust des Systems unkritisch
			\end{itemize}
		\item Physikalischer Diebstahl des Tokens notwendig um erfolgreiche
			Impersonation-Attacken auszuführen
	\end{itemize}
\end{frame}

\section{Ablauf einer Registrierung}
\begin{frame}{Ablauf einer Registrierung}
	\begin{itemize}
		\item Client erhält vom Server Schlüsselgenerierungsparameter
		\item Client reicht diese an den Authenticator weiter
		\item Authenticator speichert das Schlüsselpaar für die Relying Party und gibt den Public Key an den Client
		\item Client leitet den Public Key an den Server weiter
		\item Server persistiert de Public Key und ordnet ihn dem Nutzer zu
	\end{itemize}
\end{frame}

\section{Ablauf einer Authentifizierung}
\begin{frame}{Ablauf einer Authentifizierung}
	\begin{itemize}
		\item Client erhält vom Server eine Challenge
		\item Client reicht diese an den Authenticator weiter
		\item Authenticator signiert die Challenge mit dem Private Key
		\item Client leitet die Signatur an den Server weiter
		\item Server überprüft mit dem Public Key ob die Signatur gültig ist
	\end{itemize}
\end{frame}

\section{Möglichkeiten durch den Einsatz von FIDO2}
\begin{frame}{Möglichkeiten durch den Einsatz von FIDO2}
	\begin{itemize}
		\item viele Angriffsvektoren lassen sich zu weiten Teilen durch den
			Einsatz von FIDO2 ausschließen bzw.\ verharmlosen (bspw.
			Man-In-The-Middle, Phishing etc.)
		\item Nutzer müssen keine Passwörter mehr verwalten
		\item Nutzer können sich einfach und sicher passwortlos Anmelden
	\end{itemize}
\end{frame}

\section{Details zur Implementierung des PoC}
\begin{frame}{Details zur Implementierung des PoC}
	\begin{itemize}
		\item Siehe Bildschirmübertragung
	\end{itemize}
\end{frame}

\section{Anhänge / Verweise}
\begin{frame}{Anhänge / Verweise}
	\begin{itemize}
		\item Quellcode: \href{https://github.com/mara214/fido2-auth}{github.com/mara214/fido2-auth}
		\item Authenticator: \href{https://store.google.com/de/product/titan_security_key?pli=1&hl=de}{Google Titan}
	\end{itemize}
\end{frame}

\end{document}

### 2022-06-29 it-forensik/sichere-datenlöschung.tex
Apply change requests to the it forensics slides

@@
-\documentclass{beamer}
+\documentclass[aspectratio=169]{beamer}
 
@@
 
+\addtobeamertemplate{frametitle}{\vspace*{0.75em}\hspace*{0.5em}}{\vspace*{0.25cm}}
+
 \begin{document}
@@
 \begin{frame}
-	IT Forensik
+	\begin{center}IT Forensik\end{center}
 	\vspace{1em}
@@
 		\item Durch Durchsuchen der Register einer Festplatten mit entfernter
-			Zugriffstabelle lassen sich softwareseitig Datenrekonstruieren
+			Zugriffstabelle lassen sich softwareseitig Daten rekonstruieren
 		\item Bei Beschädigten oder mittels SMR gesicherten Festplatten ist
@@
 
-\section{Vorteile kryptografiebasiereter Datenlöschung}
-\begin{frame}{Vor- und Nachteile kryptografiebasiereter Datenlöschung}
+\section{Vorteile kryptografiebasierter Datenlöschung}
+\begin{frame}{Vor- und Nachteile kryptografiebasierter Datenlöschung}
 	\large{Vorteile:}
@@
 	\begin{itemize}
-			\item \begin{scriptsize}
-				BSI.\ (2021, 28. September). Daten auf Festplatten und Smartphones
+		\item BSI.\ (2021, 28. September). Daten auf Festplatten und Smartphones
 			endgültig Löschen. Bundesamt für Sicherheit in der
@@
 			\href{https://www.bsi.bund.de/DE/Themen/Verbraucherinnen-und-Verbraucher/Informationen-und-Empfehlungen/Cyber-Sicherheitsempfehlungen/Daten-sichern-verschluesseln-und-loeschen/Daten-endgueltig-loeschen/daten-endgueltig-loeschen_node.html}{https://www.bsi.bund.de/DE/Themen/Verbrauche..}
-			\end{scriptsize}
-			\item \begin{scriptsize}
-				Köhler, T.\ (2021, August 20).\ HDD und SSD - Festplatte Sicher
+		\item Köhler, T.\ (2021, August 20).\ HDD und SSD - Festplatte Sicher
 			Löschen.\ HDD und SSD - Festplatte sicher löschen | experte.de.
@@
 			https://www.experte.de/it-sicherheit/festplatte-loeschen
-			\end{scriptsize}
-			\item \begin{scriptsize}
-				EaseUS.\ (2022,\ 19. Januar).\ Formatierte Festplatte
+		\item EaseUS.\ (2022,\ 19. Januar).\ Formatierte Festplatte
 			Wiederherstellen - so geht's.\ EaseUS.\ Abgerufen am 25. Juni, 2022,
@@
 			https://www.easeus.de/festplatte-wiederherstellen/formatierte-festplatte-wiederherstellen-freeware.html 
-			\end{scriptsize}
-			\item \begin{scriptsize}
-				Blancco. (2022, 19. Mai). A comprehensive list of data wiping and
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Quellen}
+	\begin{itemize}
+		\item Blancco. (2022, 19. Mai). A comprehensive list of data wiping and
 			erasure standards. Blancco. Abgerufen am 25. Juni, 2022, von
 			https://www.blancco.com/resources/blog-comprehensive-list-data-wiping-erasure-standards/
-			\end{scriptsize}
-			\item \begin{scriptsize}
-				Remus, Dipl.-Ing. Patric. Archicrypt Shredder 7. BSI-2011-VS.\
+		\item Remus, Dipl.-Ing. Patric. Archicrypt Shredder 7. BSI-2011-VS.\
 			Abgerufen am 25. Juni, 2022, from
 			https://www.archicrypt.de/shredder7/bsi-2011-vs.html
-			\end{scriptsize}
 	\end{itemize}

### 2023-04-11 ethik-in-der-it-sicherheit/onepager.tex
Add a tex file for ethical onepagers

\documentclass[journal]{IEEEtran}

\usepackage[a4paper]{geometry}\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}

\pagenumbering{Roman}

\begin{document}

\markboth{Ethik in der IT Sicherheit – Mara Schulke — 20215853}{}

\begin{onecolumn}

\section{Das Studienmodul im Überblick}

Ethik und IT-Sicherheit als gegensätzliche Disziplinen
\\
Parallelen zwischen Ethik und IT-Sicherheit
\\
Schutzgut in der IT-Sicherheit
\\
Informationelle Selbstbestimmung als Schutzgut
\\
Kontextabhängigkeit des Rechts auf informationelle Selbstbestimmung
\\
Definition und Charakteristik von Ethik
\\
Reflexion der Moral als Bestandteil der Ethik
\\
Komplexität ethischer Fragen aufgrund unterschiedlicher Wertevorstellungen
\\
Verantwortung bei komplexen und umfangreichen digitalen Themen
\\
Veränderungspotential durch Digitalisierung
\\
Arten von Hackern und ethisches Dilemma
\\
Einführung in das Modul und Lernziele

\newpage

Überblick und Lernziele BBA (Begriffliche Bestimmung und Abgrenzung)
\\

\newpage

Erklärung und Abgrenzung von Grundbegriffen der Ethik
\\
Unterschied zwischen Moral und Ethik
\\
Erklärung des Berufsethos und seine Bedeutung für eine Gemeinschaft
\\
Hinterfragen bestehender Wertesysteme unter Einbezug eines moralischen Konflikts (moralische Kompetenz)
\\
Ethik beschäftigt sich mit der Suche nach einer besseren Moral
\\
Diskussion von Relativismus in der Ethik
\\
Verbindung von Recht und Moral in Bezug auf die Informationstechnologie
\\
Übungsaufgaben, Zusammenfassung, Wissensüberprüfung und Literaturangaben als Teil der Lerneinheit.

\end{onecolumn}
\end{document}

### 2023-04-13 ethik-in-der-it-sicherheit/onepager.tex
Update the ethics onepager

@@
 
-Ethik und IT-Sicherheit als gegensätzliche Disziplinen
+In dem ersten Studienmodul dieses Kurses werden grundlegende Fakten aufgezeigt, wie beispielsweise
+die gegenüberstellung der Disziplinen IT Sicherheit und Ethik und deren inherenter Widerspruch.
+Ein wichtiger elementarer Unterschied zwischen diesen beiden Disziplin ist die technologische Natur
+der IT Sicherheit und die geisteswissenschaftliche der Ethik. In beiden Disziplinen spielt allerdings
+die Verantwortung eine zentrale Rolle -  die Verantwortung in der IT Sicherheit auf Nachweisbarkeit,
+informationelle Selbstbestimmung und die Wahrung der Vertraulichkeit und in der Ethik die Verantwortung
+im bezug auf getroffene unter Umständen risikobehaftete Entscheidungen.
+
+Die Digitalisierung im allgemeinen und ihr Einfluss auf die Gesellschaft hinsichtlich ihrer Möglichkeiten,
+zukünftigen Entwicklung etc. kommt mit einer großen Ungewissheit einher und wirft viele 
+verantwortungsbehaftete Fragestellungen auf die eine Schnittmenge der IT Sicherheit und Ethik darstellen.
+
+--------------------------
+
+ Vor dem Hintergrund der Digitalisierung, deren technologische Möglichkeiten die Gesellschaft (national wie global) vor unbekannte und neuartige Herausforderungen stellt, potenziert sich Verantwortungsfrage.
+
+Das Schutzgut in der IT-Sicherheit ist die informationelle Selbstbestimmung, die das Recht des Einzelnen darstellt, über seine personenbezogenen Daten zu bestimmen.
+
+ Die Digitalisierung hat Auswirkungen auf alle Lebensbereiche und stellt neue ethische Fragen, die die IT-Sicherheit beeinflussen.
+Die Cybersecurity stellt Fragen zur Privatsphäre und Überwachung und fordert sowohl technische als auch ethische Kompetenzen der Akteure heraus.
+Verantwortung spielt eine zentrale Rolle bei den Herausforderungen der Digitalisierung.
+
+%Ethik und IT-Sicherheit als gegensätzliche Disziplinen
 \\
-Parallelen zwischen Ethik und IT-Sicherheit
+%Parallelen zwischen Ethik und IT-Sicherheit
 \\
-Schutzgut in der IT-Sicherheit
+%Schutzgut in der IT-Sicherheit
 \\
-Informationelle Selbstbestimmung als Schutzgut
+%Informationelle Selbstbestimmung als Schutzgut
 \\
-Kontextabhängigkeit des Rechts auf informationelle Selbstbestimmung
+%Kontextabhängigkeit des Rechts auf informationelle Selbstbestimmung
 \\
-Definition und Charakteristik von Ethik
+%Definition und Charakteristik von Ethik
 \\
-Reflexion der Moral als Bestandteil der Ethik
+%Reflexion der Moral als Bestandteil der Ethik
 \\
-Komplexität ethischer Fragen aufgrund unterschiedlicher Wertevorstellungen
+%Komplexität ethischer Fragen aufgrund unterschiedlicher Wertevorstellungen
 \\
-Verantwortung bei komplexen und umfangreichen digitalen Themen
+%Verantwortung bei komplexen und umfangreichen digitalen Themen
 \\
-Veränderungspotential durch Digitalisierung
+%Veränderungspotential durch Digitalisierung
 \\
-Arten von Hackern und ethisches Dilemma
+%Arten von Hackern und ethisches Dilemma
 \\
-Einführung in das Modul und Lernziele
+%Einführung in das Modul und Lernziele
 
-\newpage
-
-Überblick und Lernziele BBA (Begriffliche Bestimmung und Abgrenzung)
+\section{Überblick und Lernziele BBA (Begriffliche Bestimmung und Abgrenzung)}
 \\
-
-\newpage
 
@@
 \end{onecolumn}
-\end{document}
+\end{document}+

### 2023-04-14 ethik-in-der-it-sicherheit/onepager.tex
Finish the onepager for the first two sections

--- 2023-04-11 ethik-in-der-it-sicherheit/onepager.tex
@@
 
-\usepackage[a4paper]{geometry}\usepackage[ngerman]{babel}
+\usepackage[a4paper,top=5cm,bottom=5cm]{geometry}
+\usepackage[ngerman]{babel}
 \usepackage[utf8]{inputenc}
@@
 
-Ethik und IT-Sicherheit als gegensätzliche Disziplinen
-\\
-Parallelen zwischen Ethik und IT-Sicherheit
-\\
-Schutzgut in der IT-Sicherheit
-\\
-Informationelle Selbstbestimmung als Schutzgut
-\\
-Kontextabhängigkeit des Rechts auf informationelle Selbstbestimmung
-\\
-Definition und Charakteristik von Ethik
-\\
-Reflexion der Moral als Bestandteil der Ethik
-\\
-Komplexität ethischer Fragen aufgrund unterschiedlicher Wertevorstellungen
-\\
-Verantwortung bei komplexen und umfangreichen digitalen Themen
-\\
-Veränderungspotential durch Digitalisierung
-\\
-Arten von Hackern und ethisches Dilemma
-\\
-Einführung in das Modul und Lernziele
+In dem ersten Studienmodul dieses Kurses werden grundlegende Fakten aufgezeigt wie beispielsweise
+die Gegenüberstellung der Disziplinen IT Sicherheit und Ethik und deren inhärenter Widerspruch.
+Ein wichtiger elementarer Unterschied zwischen diesen beiden Disziplin ist die technologische Natur
+der IT Sicherheit und die geisteswissenschaftliche der Ethik. In beiden Disziplinen spielt allerdings
+die Verantwortung eine zentrale Rolle – die Verantwortung in der IT Sicherheit auf Nachweisbarkeit, 
+informationelle Selbstbestimmung und die Wahrung der Vertraulichkeit und in der Ethik die Verantwortung
+im Bezug auf getroffene unter Umständen risikobehaftete Entscheidungen. 
+Die Digitalisierung im Allgemeinen und ihr Einfluss auf die Gesellschaft hinsichtlich ihrer Möglichkeiten
+zukünftigen Entwicklung etc. kommt mit einer großen Ungewissheit einher und wirft viele 
+verantwortungsbehaftete Fragestellungen auf die eine Schnittmenge der IT Sicherheit und Ethik darstellen. 
+Ein weiterer zentraler Aspekt der Lerneinheit befasst sich mit der Cybersecurity und den damit 
+einhergehenden ethischen Fragestellungen bezogen auf die Privatsphäre und Überwachung.
 
-\newpage
+\section*{Fragestellung I}
 
-Überblick und Lernziele BBA (Begriffliche Bestimmung und Abgrenzung)
-\\
+Im Kapitel wurde das Recht auf informationelle Selbstbestimmung immer in einem bestimmten Kontext 
+betrachtet – eine interessante Fragestellung ist hier, welche ausnahmen es von diesem Recht gibt – und 
+anhand welcher Regeln Situationen als Ausnahme bewertet werden können.
 
-\newpage
+\section{Überblick und Lernziele BBA (Begriffliche Bestimmung und Abgrenzung)}
 
-Erklärung und Abgrenzung von Grundbegriffen der Ethik
-\\
-Unterschied zwischen Moral und Ethik
-\\
-Erklärung des Berufsethos und seine Bedeutung für eine Gemeinschaft
-\\
-Hinterfragen bestehender Wertesysteme unter Einbezug eines moralischen Konflikts (moralische Kompetenz)
-\\
-Ethik beschäftigt sich mit der Suche nach einer besseren Moral
-\\
-Diskussion von Relativismus in der Ethik
-\\
-Verbindung von Recht und Moral in Bezug auf die Informationstechnologie
-\\
-Übungsaufgaben, Zusammenfassung, Wissensüberprüfung und Literaturangaben als Teil der Lerneinheit.
+Ein Kernbestandteil dieser Lerneinheit stellt der ``begriffliche Werkzeugkoffer'' dar - der zur akkuraten 
+Beschreibung von Situationen und Fragestellungen dienen soll und eine Bewertung dieser erleichtert.
+Somit können zuverlässiger möglichst belastbare ethische Entscheidungen zustande kommen.
+
+Das Wertesystem einer Gesellschaft lässt sich als sinngebendes Leitbild verstehen, anhand dessen 
+Entscheidungen getroffen werden, können und das eine gemeinsame Basis innerhalb einer Gruppe darstellt. 
+Wertesysteme beziehungsweise Moralvorstellungen sind flexible Konzepte, die sich von Teilgemeinde zu 
+Teilgemeinde innerhalb einer Gesellschaft unterscheiden oder teilweise auch widersprechen können. 
+Bestimmte Berufsgruppen, die sich über mehrere Gesellschaften hinweg erstrecken, können ebenfalls eigene 
+abgewandelte Wertesysteme - einen sogenannten Berufsethos - entwickeln, der ein Leitbild für Berufstätige 
+in dieser Gruppe darstellen kann.
+Die sogenannte moralische Kompetenz tritt auf, wenn spezielle individuelle Werte im Konflikt mit dem 
+geltenden Wertesystem stehen.
+
+Der Relativismus führt bestimmte moralische Überzeugungen auf soziale, kulturelle und historische
+Gegebenheiten zurück und setzt sie somit in Verhältnis und gibt ihnen Kontext.
+In der Informationstechnologie müssen bestimmte Rechtsnormen verbindlich umgesetzt werden um eine 
+straffreie Implementierung von moralisch fragwürdigen Problemen voranzutreiben. Diese Vorgabe ist
+losgelöst von den spezifischen Wertesystemen der Person, die diese Implementierung erstellt um eine
+gesamtgesellschaftlich-einheitliche Vorgehensweise zu etablieren.
+
+\section*{Fragestellung II}
+
+Im Bezug auf Wertvorstellungen wird sofort der Einsatz von künstlicher Intelligenz in kritischen 
+Bereichen präsent. Anhand welcher Regeln und Leitsätze sollen sich selbstdenkende beziehungsweise 
+kognitive Systeme verhalten und welche Entscheidungen dürfen, sie überhaupt treffen? Ist es überhaupt 
+möglich gewisse extrem fordernde Fragestellungen wie z. B. die Triage in der Medizin oder den Abschuss 
+eines Ziels im Militärkontext zu automatisieren?
 
 \end{onecolumn}
-\end{document}
+\end{document}+

### 2023-04-15 ethik-in-der-it-sicherheit/onepager.tex
Fix typos

@@
 
-In dem ersten Studienmodul dieses Kurses werden grundlegende Fakten aufgezeigt wie beispielsweise
+In der ersten Einheit dieses Kurses werden grundlegende Gegebenheiten aufgezeigt wie beispielsweise
 die Gegenüberstellung der Disziplinen IT Sicherheit und Ethik und deren inhärenter Widerspruch.
@@
 Im Kapitel wurde das Recht auf informationelle Selbstbestimmung immer in einem bestimmten Kontext 
-betrachtet – eine interessante Fragestellung ist hier, welche ausnahmen es von diesem Recht gibt – und 
+betrachtet – eine interessante Fragestellung ist hier, welche Ausnahmen es von diesem Recht gibt – und 
 anhand welcher Regeln Situationen als Ausnahme bewertet werden können.
@@
 abgewandelte Wertesysteme - einen sogenannten Berufsethos - entwickeln, der ein Leitbild für Berufstätige 
-in dieser Gruppe darstellen kann.
-Die sogenannte moralische Kompetenz tritt auf, wenn spezielle individuelle Werte im Konflikt mit dem 
-geltenden Wertesystem stehen.
+in dieser Gruppe darstellen kann. Die sogenannte moralische Kompetenz tritt auf, wenn spezielle
+individuelle Werte im Konflikt mit dem geltenden Wertesystem stehen.
 
@@
 kognitive Systeme verhalten und welche Entscheidungen dürfen, sie überhaupt treffen? Ist es überhaupt 
-möglich gewisse extrem fordernde Fragestellungen wie z. B. die Triage in der Medizin oder den Abschuss 
-eines Ziels im Militärkontext zu automatisieren?
+möglich gewisse ethisch extrem fordernde Fragestellungen wie z. B. die Triage in der Medizin oder den
+Abschuss eines Ziels im Militärkontext zu automatisieren?
 
 \end{onecolumn}
-\end{document}+
+
+\end{document}

### 2023-04-16 ethik-in-der-it-sicherheit/onepager.tex
Update the ethic onepager wording

@@
 
-Im Kapitel wurde das Recht auf informationelle Selbstbestimmung immer in einem bestimmten Kontext 
+Im Kapitel wird das Recht auf informationelle Selbstbestimmung immer in einem bestimmten Kontext 
 betrachtet – eine interessante Fragestellung ist hier, welche Ausnahmen es von diesem Recht gibt – und 
@@
 Das Wertesystem einer Gesellschaft lässt sich als sinngebendes Leitbild verstehen, anhand dessen 
-Entscheidungen getroffen werden, können und das eine gemeinsame Basis innerhalb einer Gruppe darstellt. 
+Entscheidungen getroffen werden können und das eine gemeinsame Basis innerhalb einer Gruppe darstellt. 
 Wertesysteme beziehungsweise Moralvorstellungen sind flexible Konzepte, die sich von Teilgemeinde zu 

### 2023-04-26 ethik-in-der-it-sicherheit/onepager.tex
Update the ethics onepager

@@
 
+\newpage
+
+\section{Verschiedene Bezugspunkte von Ethik}
+
+\section*{Fragestellung III}
+
+\section{Ausdifferenzierung in komplexe Bezugskontexte}
+
+In dem vierten Kapitel des Lehrmaterials wird im Kern auf die Abhängigkeit der ethischen Kompetenz von
+den jeweiligen Einsatz-Szenarien eingegangen. Dies wird erreicht in dem zu erst zwei verschiedene Definitionen
+eingeführt werden - die angewandte Ethik und die Bereichsethik:
+
+--- Die angewandte Ethik stellt eine komplexe Herausforderung da die die theoretischen Aspekte einer ethischen
+Fragestellung auf ein praktisches Problem anwendet um somit zu einem ethisch korrekten Handlungsplan zu gelangen.
+Die Konzepte die in einer konkreten Problemsituation angewendet werden formen eine ethische Position zu einem Thema
+ - dies kann auch im Rahmen von Politik stattfinden um die Meinung und das Wertesystem der breiten Masse einer
+ Gesellschaft zu beeinflussen. Ein im Skript benanntes Beispiel das diese Funktion erfüllt ist der Deutsche Ethikrat.
+
+---
+
+Bei der Angewandten Ethik werden theoretische Reflexionen der Ethik mit praktischen Orientierungsfragen verquickt.
+Die Aufgabe der angewandten Ethik ist es, auf Orientierungsprobleme der modernen Gesellschaft zu reagieren.
+
+Aufgrund der Schnelligkeit und Dynamik gesellschaftlicher und technischer Entwicklungen ist eine ethische Bewertung von Sachverhalten (heute) nicht mehr ohne weiteres zu sehen.
+Häufig werden auch Verfahren zur beratenden Unterstützung von politischen Willensbildungen und zur Herstellung gesellschaftlicher Konsense mit dem Prädikat Ethik versehen.
+Mit Blick auf die Beratungs- und Orientierungsfunktion von Ethikräten oder Ethikkommissionen kann man Ethik auch als eine Aktivität an der Schnittstelle von Wissenschaft und Gesellschaft verstehen.
+Durch die enge Verquickung normativer Fragen mit Erkenntnissen und Erfahrungen aus bestimmten Handlungszusammenhängen des menschlichen Miteinanders kommt auch zu einer Pluralisierung der ethischen Zugänge und Grundhaltungen, die durch die Beteiligten vertreten werden.
+Einem Konsens gehen notwendigerweise Diskurse voraus, bei denen es um die Beleuchtung und Analyse von ethischen Problemstellungen kommt, denen eine höchst praktische Relevanz innewohnt.
+Eine Bereichsethik könnte man als einen sehr spezifischen Teil der angewandten Ethik bezeichnen.
+Als Bereich versteht man dabei eine mehr oder weniger scharf abgrenzbare Sphäre einer menschlichen Praxis, die auf ihre eigene Weise moralische Probleme und Fragenstellungen aufwirft.
+Wenngleich aus den bekannten theoretischen Ethiken durchaus allgemeine Ableitungen für einen bestimmten Sachverhalt gemacht werden können, ist dennoch die Reflexion auf die Sachzusammenhänge und den bereichsspezifischen Wissensstand notwendig, um zu einer qualifizierten Beurteilung einer moralischen Fragestellung in einer konkreten lebensweltlichen Handlungssituation zu kommen.
+Zwischen den Bereichsethiken kann es auch zu Überlappungen kommen. Das hängt davon ab, wie komplex ein moralisch herausfordernder Sachverhalt ist.
+Ethische Probleme, die durch Erfolge der IuK-Technologien aufgeworfen werden, können nicht mehr ohne weiteres im Sinne einer Bereichsethik behandelt werden.
+Tatsächlich sind von den aktuellen datengetriebenen Entwicklungen die Grundlagen der Ethik überhaupt betroffen.
+
+\section*{Fragestellung IV}
+
 \end{onecolumn}

### 2023-05-11 ethik-in-der-it-sicherheit/onepager.tex
Start the third onepager for the ethics course

--- 2023-04-16 ethik-in-der-it-sicherheit/onepager.tex
@@
 
+\newpage
+
+\section{Verschiedene Bezugspunkte von Ethik}
+
+Das Skript beschreibt hier die vier wichtigen Bezugspunkte zur Ethik: die Pflicht, der Diskurs,
+der Nutzen und die Tugend. Des Weiteren wird zwischen deskriptiver Ethik (die bereits vorhandene
+moralische Werte-Systeme und Vorstellungen beschreibt) und der normativen Ethik die sich mit der
+Urteilsfällung und einem Regelsystem befasst. Somit gibt es klare parallelen zwischen der normativen
+Ethik und der Anwendung in den Rechtssystemen von demokratisch regierten Staaten.
+
+Im Bezug zur Handlungsnotwendigkeit gibt es die zwei Dimensionen der materiellen und formalen Ethik.
+die materielle Ethik arbeitet einen konkreten Handlungsplan heraus währen die formale Ethik sich mit
+allgemein anwendbaren aber ebenso gewichteten Leitlinien befasst. Die Goldene Regel ``Was Du nicht willst,
+das man Dir tu, das füg auch keinem Andern zu!'' legt mehr Verantwortung in die Hände der 
+interpretierenden Person als konkrete Regeln die durch allgemeinen gesellschaftlichen Konsens
+definiert wurden. Somit kommt es zusätzlich zu den Gesetzen innerhalb einer Gesellschaft auch auf
+die praktische Urteilskraft jeder einzelnen an, da in eine liberalen Demokratie zu komplexe
+Herangehensweisen an bestimmte Probleme erfordert um sie in einem allumfassenden Regelwerk
+
+abschließend herunter zu schreiben.
+
+\section*{Fragestellung III}
+
+Eine interessante Frage die mir im Bezug auf die Gewichtung zwischen der materiellen und formalen Ethik
+innerhalb einer Gesellschaft stellt ist der Grad: Bis zu welchem Punkt sollten Gesetze und Richtlinien
+verfasst werden und ab welchem Punkt muss die Eigenverantwortung jeder Einzelperson greifen?
+
+\section{Ausdifferenzierung in komplexe Bezugskontexte}
+
+In dem vierten Kapitel des Lehrmaterials wird im Kern auf die Abhängigkeit der ethischen Kompetenz von
+den jeweiligen Einsatz-Szenarien eingegangen. Dies wird erreicht in dem zu erst zwei verschiedene
+Definitionen eingeführt werden - die angewandte Ethik und die Bereichsethik:
+
+--- Die angewandte Ethik stellt eine komplexe Herausforderung da die die theoretischen Aspekte einer
+ethischen Fragestellung auf ein praktisches Problem anwendet um somit zu einem ethisch korrekten 
+Handlungsplan zu gelangen. Die Konzepte die in einer konkreten Problemsituation angewendet werden
+formen eine ethische Position zu einem Thema - dies kann auch im Rahmen von Politik stattfinden um die
+Meinung und das Wertesystem der breiten Masse einer Gesellschaft zu beeinflussen. Ein im Skript benanntes
+Beispiel das diese Funktion erfüllt ist der Deutsche Ethikrat.
+
+--- Die Bereichsethik wird als Teil der angewandten Ethik definiert in dessen sich mit moralischen Fragen
+im Kontext von spezifischen Bereichen auseinandergesetzt wird. Das Kapitel nennt als Beispiele die Bio-,
+Medizin-, Wirtschafts-, Umwelt- und IT-Ethik als Beispiele. Die Notwendigkeit für die Bereichsethik ergibt
+sich aus der limitierten Anwendbarkeit von allgemeinen ethischen Leitsätzen in konkreten bereichsspezifischen 
+Problemstellungen (wie zum Beispiel der Triage in der Medizin).
+
+Ein gutes Beispiel für die differenzierte Betrachtung stellt die Corona-Krise dar, die starke Auswirkungen über
+viele Lebensbereiche hinweg hatte und die keine einfache ethische Betrachtung erlaubte. Die Problematik erstreckte
+sich über viele verschiedene Bereiche wie Medizin, Wirtschaft, Sozialwissenschaften etc. für die Wissen und
+Ethikkenntnisse aus diesen Bereichen notwendig ist um eine Angemessene Entscheidung zu treffen.
+
+\section*{Fragestellung IV}
+
+Wie sollten ethische Probleme gehandhabt werden bei denen ethische Grundsätze aus einzelnen Bereichen sich
+widersprechen und einen Konflikt darstellen? Anders formuliert: gibt es eine implizite Hierarchie der
+Bereichsethiken?
+
+\newpage
+
+\section{Relevanz des Begriffs der Verantwortung}
+
+Die Verantwortung ist ein dialogisches Grundprinzip – also eines das mehrere ``Akteure'' benötigt um
+Anwendung zu finden. Es ist eine explizite oder implizite Vereinbarung einem moralischen / ethischen 
+Anspruch innerhalb einer Beziehung gerecht zu werden – dies kann sich in verschiedenen Formen darstellen
+und erstreckt sich über zwischenmenschliche Beziehungen hinaus. Das Eintreten eines Ereignisses, dass der
+zugrundeliegenden Vereinbarung der Verantwortung widerspricht, bringt eine Schuld und die Vereinbarung der
+Übernahme von eventuellen Konsequenzen mit sich. Ein einfaches Beispiel stellt hier der bekannte Spruch
+"Eltern haften für ihre Kinder" dar - die Verantwortung in der Beziehungskonstellation liegt hier bei den
+Eltern und bei einer Vernachlässigung der Sorgfaltspflicht gehen eventuelle entstandene Schäden auf diese
+über. In zwischenmenschlichen Beziehungen ist allerdings die Suche nach dem ``Schuldigen'' oftmals einfacher
+als in rein digitalen Kontexten in denen eine hohe Anonymisierung herrscht. Eine große Aufgabe der IT-Sicherheit
+bzw. IT-Forensik ist die Nachvollziehbarkeit von Ereignissen und die Beweis-Erbringung. Im Kontext von
+ethischen Diskussionen steht oft die Verantwortung im Vergleich zur Schuld im Vordergrund.
+
+\section*{Fragestellung V}
+
+Wie kann im digitalen Kontext eine klare Verantwortungsverteilung etabliert werden, um im Falle von Verstößen
+eine angemessene Schuldzuweisung vornehmen zu können und eventuelle Konsequenzen zu übernehmen? Wann ist ein
+Betreiber der nicht ausreichende Daten über Handlungen durch sein Produkt speichert haftbar für Handlungen dritter
+die dadurch verschleiert werden konnten?
+
+\section{Moderne Herausforderungen an die Verantwortung}
+
+\section*{Fragestellung VI}
+
 \end{onecolumn}

### 2023-05-12 ethik-in-der-it-sicherheit/onepager.tex
Finish the third onepager

@@
 
+Das Kapitel über die modernen Herausforderungen an die Verantwortung in der Ethik beschäftigt anders als das
+fünfte Kapitel nicht mit der reinen Definition der Verantwortung sonder mit komplexeren Situationen und zukünftigen
+Ereignissen die mit der Verantwortung in Verbindung stehen. Eine Erkenntnis des Kapitels besteht aus der
+Feststellung, dass für Verantwortung Leben notwendig ist, welches entweder die Verantwortung trägt oder von ggf.
+weitreichenden Konsequenzen betroffen ist. Die Teilkapitel über kumulative Effekte und eventuelle unvorhersehbare
+Folgen erläutern Situationen und Konstellationen zwischen mehreren Einzelpersonen die implizit zu einem zukünftigen
+Ereignis beitragen. Als ein gutes Beispiel hierfür wird die Entwicklung der Corona-Impfstoffe herangezogen um zu
+verdeutlichen, dass bei individuellen Beiträgen von vielen Akteuren die Verteilung der Verantwortung und eventueller
+Schuld keine triviale Frage ist. Hier kommt auch die Systemverantwortung zum Tragen und wird teil der ethischen Bewertung
+einer solchen Problemstellung – in manchen Fragestellungen ist es nicht Sinnvoll ein Individuum in die Verantwortlichkeit
+zu ziehen sondern das Übergreifende System (in diesem Beispiel die Staaten, die Industrie und unsere gesellschaftliche
+Wirtschaftsausrichtung).
+
 \section*{Fragestellung VI}
+
+Wie lässt sich eine klare Verantwortlichkeit im Bezug auf die Corona-Impfstoff-Entwicklung innerhalb der
+Gesellschaft finden? Und vor allem: Welche Partei trägt die Schuld im Hinblick auf die Impfschaden-Debatte? Sind
+es die Pharma-Konzerne durch zu schnelle Entwicklungsverfahren oder die EU durch fahrlässige Zulassungen von
+unfertigen Stoffen? Hätte eine langsamere Zulassung weniger Leid mit sich gebracht oder mehr?
 

### 2023-05-26 ethik-in-der-it-sicherheit/onepager.tex
Add the 7th chapter for the ethics course

@@
 
+\newpage
+
+\section{Der Blick auf Unternehmen bzw. Organisationen}
+
+Innerhalb der letzten beiden Jahrzehnte hat sich die ``Corporate Social Responsibility''
+also Sammelbegriff für Maßnahmen eines Unternehmens für soziale und ökologische 
+Gerechtigkeit gefestigt. Diese Maßnahmen umfassen zum Beispiel 
+Nachhaltigkeits-Engagements eines Unternehmens, damit Sie der Verantwortung die Umwelt zu 
+erhalten gerecht werden.
+
+Damit einhergehend findet sich der Begriff der Kernverantwortung wieder - dies ist ein vom
+Unternehmen selbst definierter Bereich, in dem Verantwortung übernommen wird und Maßnahmen
+getroffen werden. Ohne eine klare Kernverantwortung – Beispielsweise ``Nachhaltige 
+Lieferketten'' – würde es einem Unternehmen schwerer fallen eine schnelle 
+Handlungsfähigkeit beizubehalten. Von einer solchen ``Richtlinie'' ist es deutlich 
+einfacher eine konkrete Handlungsempfehlung abzuleiten.
+
+Innerhalb der ``Corporate Social Responsibility'' wird der Begriff der Nachhaltigkeit 
+nicht rein umweltbezogen Verstanden sondern setzt sich aus den Teilgebieten der Ökologie,
+Ökonomie und Gesellschaft zusammen.
+
+Die Ethik innerhalb der IT-Sicherheit hängt in weiten Teilen mit der ``Corporate Social
+Responsibility'' zusammen – beispielsweise im Kontext von Social-Media- und 
+Multi-Media-Plattformen im Internet. Hier stehen die Betreiber insbesondere in der
+Pflicht bestimmte Gesellschaftsgruppen zu schützen (zum Beispiel Minderjährige, Opfer
+von Hass und Hetze im Internet etc.). Die IT-Sicherheit dient hier als Werkzeug zur
+Aufklärung von vollbrachten und zur Verhinderung von neuen Straftaten.
+
+\section*{Fragestellung VII}
+
+Ab wann sollte der Gesetzgeber bei einer eindeutigen und erheblichen Verletzung der 
+``Corporate Social Responsibility'' eingreifen? Dürfen große Social-Media Konzerne 
+straffrei politische Einflussnahme betreiben?
+
+\section{Der Blick auf die Gesellschaft}
+
+\section*{Fragestellung VIII}
+
+\newpage
+
+\section{Reflexive Befähigung von Mensch und Organisation}
+
+\section*{Fragestellung IX}
+
+\section{Digitale Ethik}
+
+\section*{Fragestellung X}
+
 \end{onecolumn}

### 2023-05-30 ethik-in-der-it-sicherheit/onepager.tex
Add the fourth chapter for the ethics course

@@
 
-Das Kapitel über die modernen Herausforderungen an die Verantwortung in der Ethik beschäftigt anders als das
-fünfte Kapitel nicht mit der reinen Definition der Verantwortung sonder mit komplexeren Situationen und zukünftigen
-Ereignissen die mit der Verantwortung in Verbindung stehen. Eine Erkenntnis des Kapitels besteht aus der
-Feststellung, dass für Verantwortung Leben notwendig ist, welches entweder die Verantwortung trägt oder von ggf.
-weitreichenden Konsequenzen betroffen ist. Die Teilkapitel über kumulative Effekte und eventuelle unvorhersehbare
-Folgen erläutern Situationen und Konstellationen zwischen mehreren Einzelpersonen die implizit zu einem zukünftigen
-Ereignis beitragen. Als ein gutes Beispiel hierfür wird die Entwicklung der Corona-Impfstoffe herangezogen um zu
-verdeutlichen, dass bei individuellen Beiträgen von vielen Akteuren die Verteilung der Verantwortung und eventueller
-Schuld keine triviale Frage ist. Hier kommt auch die Systemverantwortung zum Tragen und wird teil der ethischen Bewertung
-einer solchen Problemstellung – in manchen Fragestellungen ist es nicht Sinnvoll ein Individuum in die Verantwortlichkeit
-zu ziehen sondern das Übergreifende System (in diesem Beispiel die Staaten, die Industrie und unsere gesellschaftliche
-Wirtschaftsausrichtung).
+Das Kapitel über die modernen Herausforderungen an die Verantwortung in der Ethik beschäftigt anders
+als das fünfte Kapitel nicht mit der reinen Definition der Verantwortung sonder mit komplexeren Situationen
+und zukünftigen Ereignissen die mit der Verantwortung in Verbindung stehen. Eine Erkenntnis des Kapitels
+besteht aus der Feststellung, dass für Verantwortung Leben notwendig ist, welches entweder die Verantwortung
+trägt oder von ggf. weitreichenden Konsequenzen betroffen ist. Die Teilkapitel über kumulative Effekte
+und eventuelle unvorhersehbare Folgen erläutern Situationen und Konstellationen zwischen mehreren
+Einzelpersonen die implizit zu einem zukünftigen Ereignis beitragen. Als ein gutes Beispiel hierfür wird
+die Entwicklung der Corona-Impfstoffe herangezogen um zu verdeutlichen, dass bei individuellen Beiträgen
+von vielen Akteuren die Verteilung der Verantwortung und eventueller Schuld keine triviale Frage ist. Hier
+kommt auch die Systemverantwortung zum Tragen und wird teil der ethischen Bewertung einer solchen
+Problemstellung – in manchen Fragestellungen ist es nicht Sinnvoll ein Individuum in die Verantwortlichkeit
+zu ziehen sondern das Übergreifende System (in diesem Beispiel die Staaten, die Industrie und unsere
+gesellschaftliche Wirtschaftsausrichtung).
 
@@
 Wie lässt sich eine klare Verantwortlichkeit im Bezug auf die Corona-Impfstoff-Entwicklung innerhalb der
-Gesellschaft finden? Und vor allem: Welche Partei trägt die Schuld im Hinblick auf die Impfschaden-Debatte? Sind
-es die Pharma-Konzerne durch zu schnelle Entwicklungsverfahren oder die EU durch fahrlässige Zulassungen von
-unfertigen Stoffen? Hätte eine langsamere Zulassung weniger Leid mit sich gebracht oder mehr?
+Gesellschaft finden? Und vor allem: Welche Partei trägt die Schuld im Hinblick auf die Impfschaden-Debatte?
+Sind es die Pharma-Konzerne durch zu schnelle Entwicklungsverfahren oder die EU durch fahrlässige Zulassungen
+von unfertigen Stoffen? Hätte eine langsamere Zulassung weniger Leid mit sich gebracht oder mehr?
 
@@
 
-Innerhalb der letzten beiden Jahrzehnte hat sich die ``Corporate Social Responsibility''
-also Sammelbegriff für Maßnahmen eines Unternehmens für soziale und ökologische 
-Gerechtigkeit gefestigt. Diese Maßnahmen umfassen zum Beispiel 
-Nachhaltigkeits-Engagements eines Unternehmens, damit Sie der Verantwortung die Umwelt zu 
-erhalten gerecht werden.
+Innerhalb der letzten beiden Jahrzehnte hat sich die ``Corporate Social Responsibility'' also Sammelbegriff
+für Maßnahmen eines Unternehmens für soziale und ökologische Gerechtigkeit gefestigt. Diese Maßnahmen
+umfassen zum Beispiel Nachhaltigkeits-Engagements eines Unternehmens, damit Sie der Verantwortung die Umwelt
+zu erhalten gerecht werden.
 
-Damit einhergehend findet sich der Begriff der Kernverantwortung wieder - dies ist ein vom
-Unternehmen selbst definierter Bereich, in dem Verantwortung übernommen wird und Maßnahmen
-getroffen werden. Ohne eine klare Kernverantwortung – Beispielsweise ``Nachhaltige 
-Lieferketten'' – würde es einem Unternehmen schwerer fallen eine schnelle 
-Handlungsfähigkeit beizubehalten. Von einer solchen ``Richtlinie'' ist es deutlich 
-einfacher eine konkrete Handlungsempfehlung abzuleiten.
+Damit einhergehend findet sich der Begriff der Kernverantwortung wieder - dies ist ein vom Unternehmen
+selbst definierter Bereich, in dem Verantwortung übernommen wird und Maßnahmen getroffen werden. Ohne
+eine klare Kernverantwortung – Beispielsweise ``Nachhaltige Lieferketten'' – würde es einem Unternehmen
+schwerer fallen eine schnelle Handlungsfähigkeit beizubehalten. Von einer solchen ``Richtlinie'' ist es
+deutlich einfacher eine konkrete Handlungsempfehlung abzuleiten.
 
-Innerhalb der ``Corporate Social Responsibility'' wird der Begriff der Nachhaltigkeit 
-nicht rein umweltbezogen Verstanden sondern setzt sich aus den Teilgebieten der Ökologie,
-Ökonomie und Gesellschaft zusammen.
+Innerhalb der ``Corporate Social Responsibility'' wird der Begriff der Nachhaltigkeit nicht rein umweltbezogen
+Verstanden sondern setzt sich aus den Teilgebieten der Ökologie, Ökonomie und Gesellschaft zusammen.
 
-Die Ethik innerhalb der IT-Sicherheit hängt in weiten Teilen mit der ``Corporate Social
-Responsibility'' zusammen – beispielsweise im Kontext von Social-Media- und 
-Multi-Media-Plattformen im Internet. Hier stehen die Betreiber insbesondere in der
-Pflicht bestimmte Gesellschaftsgruppen zu schützen (zum Beispiel Minderjährige, Opfer
-von Hass und Hetze im Internet etc.). Die IT-Sicherheit dient hier als Werkzeug zur
-Aufklärung von vollbrachten und zur Verhinderung von neuen Straftaten.
+Die Ethik innerhalb der IT-Sicherheit hängt in weiten Teilen mit der ``Corporate Social Responsibility''
+zusammen – beispielsweise im Kontext von Social-Media- und Multi-Media-Plattformen im Internet. Hier stehen
+die Betreiber insbesondere in der Pflicht bestimmte Gesellschaftsgruppen zu schützen (zum Beispiel Minderjährige,
+Opfer von Hass und Hetze im Internet etc.). Die IT-Sicherheit dient hier als Werkzeug zur Aufklärung von
+vollbrachten und zur Verhinderung von neuen Straftaten.
 
@@
 
-Ab wann sollte der Gesetzgeber bei einer eindeutigen und erheblichen Verletzung der 
-``Corporate Social Responsibility'' eingreifen? Dürfen große Social-Media Konzerne 
-straffrei politische Einflussnahme betreiben?
+Ab wann sollte der Gesetzgeber bei einer eindeutigen und erheblichen Verletzung der ``Corporate Social
+Responsibility'' eingreifen? Dürfen große Social-Media Konzerne straffrei politische Einflussnahme betreiben?
 
@@
 
+Dieses Kapitel beschäftigt sich mit den Effekten der voranschreitenden Digitalisierung auf die Gesellschaft und
+benennt Unternehmen mit einer digitalen Wertschöpfungskette als Kerntreiber dieser. Damit einhergehend wird
+die sogenannte ``Corporate Digital Responsibility'' beschrieben, im Rahmen dieser Unternehmen für die Auswirkungen
+der vor ihnen angetriebenen Digitalisierung (gesellschaftlicher Wandel, Umstrukturierungen im Arbeitsmarkt etc.)
+Verantwortung übernehmen. Die Digitalisierung hat einen großen Einfluss darauf, wie soziale Interaktionen stattfinden
+und trägt somit mit jeder Erfindung oder Entwicklung die Verantwortung, dass diese einen im besten Fall
+positiven, aber zumindest neutralen Einfluss auf die gesamtgesellschaftliche Situation hat. Weiterhin wird die
+sogenannte ``Datafizierung'' beschrieben – also die Digitalisierung von analogen Prozessen und Daten. Diese
+erlaubt es Unternehmen sich – weg von der analogen Welt – hin zu rein Digitalen Geschäftsmodellen entwicklen.
+Somit werden bestimmte analoge Felder der Wirtschaft obsolet(er) und es eröffnen sich neue Möglichkeiten in der
+rein digitalen Welt. Ein gutes Beispiel hier sind die gerade boomenden ``AI''-Firmen die quasi wie ausschließlich
+mit digitalen Daten arbeiten (entweder Bewertung, Vorhersage, Generierung o.ä). Ethisch sind hier gerade die Wahrung
+der Sicherheit von personenbezogenen Daten und der Integrität von Daten von größter Bedeutung.
+
 \section*{Fragestellung VIII}
+
+Wer trägt die Verantwortung für den Wegfall ganzer Wirtschaftsbereiche? Durch die technologische Entwicklung wurden
+– und werden in der Zukunft viele der momentan existierenden Berufe teilweise verändert oder wegfallen. Welche Instanz
+trägt hier die Verantwortung für individuell betroffene Schicksale und braucht es eine starke Regulierung seitens des
+Gesetzgebers um diesen Wandel kontrolliert zu gestalten?
 

### 2023-07-01 ethik-in-der-it-sicherheit/exam.tex
Start working on the ethics exam

\documentclass[journal]{IEEEtran}

\usepackage[a4paper,top=5cm,bottom=5cm]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}

\pagenumbering{Roman}

\begin{document}

\markboth{Ethik in der IT Sicherheit – Mara Schulke — 20215853}{}

\begin{onecolumn}

\begin{center}
	\Large{Geistiges Eigentum im Zeitalter der Informationstechnik}\\
	\vspace{18pt}\hline
\end{center}

\section*{Einleitung}

Das geistige Eigentum und die Sicherstellung von Eigentümerrechten ist von
elementarer Bedeutung für unsere Gesellschaft und ihre Ordnung. Die Werke einzelner 
Individuen oder ganzer Organisationen zu schützen ist eine fundamentale ethische 
Verpflichtung der die Gesellschaft während der voranschreitenden Digitalisierung 
gerecht werden muss. Der technologische Fortschritt darf nicht zur Verwahrlosung des 
Eigentumsrechts führen, lässt sich aber gleichzeitig deutlich schwerer Kontrollieren 
als physisches Eigentum. Es ist ein komplexer Balance-Akt zwischen dem Schutz von 
einzelnen gegenüber der Mehrheit.

Mit der Betrachtung von Schutzregelungen zum Eigentum werfen sich interessante 
Fragestellungen auf, wie z.B. dürfen oder müssen in gewissen Situationen Ausnahmen 
von diesem Schutz gemacht werden, wenn das dem Gemeinwohl der Gesellschaft dienlich 
sind?

/// IMPFSTOFF DEBATTE

\section*{Einfluss der Informationstechnik auf digitales Eigentum}

Die Digitalisierung hat einen erheblichen Einfluss auf unseren Umgang 
mit geistigen und digitalen Eigentum. 
Der technologische Fortschritt hat in unserer Gesellschaft zu einer weitreichenden
Digitalisierung von Waren geführt – viele Gegenstände die früher 
nur in physischer Form dementsprechend nur begrenzt erhältlich waren, sind nun 
jederzeit auf Abruf verfügbar und lassen sich unbegrenzt oft vervielfältigen. 
Hierdurch sind ganze Teilbereiche der Kriminalität im Bezug auf den digitalen Bereich 
entstanden, wie zum Beispiel die Piraterie oder die illegale Verbreitung von
lizensierten / gekauften Inhalten.

Diese Veränderungen werfen neue Fragen und Herausforderungen auf wie zum Beispiel 
die Notwendigkeit für einen Kopierschutz, Lizenzierungen von Software etc.
Mit diesen neuen Herausforderung bring der Fortschritt allerdings auch 
neue Möglichkeiten zur Sicherung des Eigentums in der digitalen Welt mit sich.
Konzepte wie der Einsatz von Kryptographie und Plagiatserkennungs-Algorithmen 
eröffnen Möglichkeiten, die Integrität und Nachvollziehbarkeit des geistigen und 
digitalen Eigentums sicherzustellen.
Es ist wichtig diese Probleme und Herausforderungen ernst zunehmen um die Wahrung von 
Ideen und geistigem Eigentum heute und künftig sicherzustellen.

\section*{Ethik und Eigentum}
// unbearbeitet & quellen suchen!!! \\

Im Zeitalter der Informationstechnik spielen ethische Prinzipien eine zentrale Rolle in Kontext des Eigentums.

Erstens sind grundlegende ethische Prinzipien wie Fairness, Gerechtigkeit und Respekt von großer Bedeutung für den Schutz des Eigentums. Die Anerkennung der Rechte von Schöpfern und Eigentümern, sowie eine gerechte Verteilung der Ressourcen sind dabei maßgeblich. Zweitens erfordert die Ethik der Informationsgesellschaft eine kritische Betrachtung der Machtverhältnisse, die durch den Zugang zu und die Kontrolle über Informationen entstehen. Es ist wichtig, die Auswirkungen von Monopolisierung, Überwachung und Manipulation auf das Eigentum und die damit verbundenen Rechte zu analysieren. Drittens ist die Frage der digitalen Rechte und sozialen Gerechtigkeit von hoher Relevanz. Es besteht die Notwendigkeit sicherzustellen, dass alle Mitglieder der Gesellschaft Zugang zu digitalen Ressourcen haben und ihre Rechte in der digitalen Welt geschützt sind. Eine ethische Perspektive auf das Eigentum im Zeitalter der Informationstechnik erfordert daher eine kritische Auseinandersetzung mit den grundlegenden Prinzipien, der Machtverteilung und der sozialen Gerechtigkeit im digitalen Raum.

\section*{Konflikte und Dilemmata im Zusammenhang mit Eigentum und Informationstechnik}
A. Datenschutz und Privatsphäre
B. Geistiges Eigentum und Urheberrechte
C. Zugang zu Information und Wissen

\section*{Auswirkungen auf die Wirtschaft und Gesellschaft}
VI. Auswirkungen auf Wirtschaft und Gesellschaft
A. Neue Geschäftsmodelle und Märkte
B. Soziale und ethische Implikationen
C. Auswirkungen auf den Arbeitsmarkt

VII. Schlussfolgerung
A. Zusammenfassung der wichtigsten Erkenntnisse
B. Ausblick auf zukünftige Entwicklungen und Herausforderungen

\section*{Literaturverzeichnis}
VIII. Literaturverzeichnis

\end{onecolumn}

\end{document}

### 2023-07-02 ethik-in-der-it-sicherheit/exam.tex
Add the core chapter of the ethics exam

@@
 \section*{Ethik und Eigentum}
-// unbearbeitet & quellen suchen!!! \\
 
-Im Zeitalter der Informationstechnik spielen ethische Prinzipien eine zentrale Rolle in Kontext des Eigentums.
+Bei der ethischen Betrachtung dieses Themas stellen sich Fragen im Bezug auf die 
+Verantwortung, wie z.B. wer ist für die Sicherstellung der Eigentumsrechte im digitalen 
+Raum verantwortlich?
 
-Erstens sind grundlegende ethische Prinzipien wie Fairness, Gerechtigkeit und Respekt von großer Bedeutung für den Schutz des Eigentums.
-Die Anerkennung der Rechte von Schöpfern und Eigentümern, sowie eine gerechte Verteilung der Ressourcen sind dabei maßgeblich.
-Zweitens erfordert die Ethik der Informationsgesellschaft eine kritische Betrachtung der Machtverhältnisse, die durch den Zugang zu und die Kontrolle über Informationen entstehen.
-Es ist wichtig, die Auswirkungen von Monopolisierung, Überwachung und Manipulation auf das Eigentum und die damit verbundenen Rechte zu analysieren.
-Drittens ist die Frage der digitalen Rechte und sozialen Gerechtigkeit von hoher Relevanz.
-Es besteht die Notwendigkeit sicherzustellen, dass alle Mitglieder der Gesellschaft Zugang zu digitalen Ressourcen haben und ihre Rechte in der digitalen Welt geschützt sind.
-Eine ethische Perspektive auf das Eigentum im Zeitalter der Informationstechnik erfordert daher eine kritische Auseinandersetzung mit den grundlegenden Prinzipien, der Machtverteilung und der sozialen Gerechtigkeit im digitalen Raum.
+Im Falle von Musik ist ein bekannter Anbieter der Musik-Streaming-Dienst Spotify: Wer ist 
+in diesem Fall für Sicherstellung von Urheberrechten verantwortlich und wo ziehen sich 
+Verantwortungsbereiche durch diese Thematik? Ist bei der Verhinderung von Raubkopien 
+das Unternehmen, dass die Plattform zur Verbreitung bereitstellt, der Gesetzgeber, oder 
+das Individuum, das die Plattform zur Verbreitung verwendet gefragt? In der Realität zeigt 
+sich, dass die Wahrung des geistigen Eigentums im Verantwortungsbereich von allen drei 
+Akteuren liegt. Die Plattform muss die Verletzung feststellen – ein gutes Beispiel ist 
+hier YouTube, die die Monetarisierung von urheberrechtlich geschützten Inhalten verbieten. 
+Der Gesetzgeber muss einen rechtlichen Rahmen stellen um strafrechtlich gegen die 
+Verletzung von Urheberrechten vorzugehen – und jedes Individuum trägt einen Teil 
+Eigenverantwortung keine Urheberrechtsverletzung vorzunehmen, und dessen eigene Werke 
+so gut es geht davor zu schützen.
+
+Die zugrundeliegende ethische Fragestellung führt auf das gesellschaftliche Wertesystem 
+zurück und der Umgang einer jeden Gesellschaft ist abhängig davon, wie sehr Fairness, 
+Respekt und Gerechtigkeit in diesem einen Platz finden.
+
+Eine Anerkennung der Rechte der Hersteller und Besitzer stehen der Problematik der 
+gerechten Verteilung und dem Grundprinzip der Chancengleichheit gegenüber. Bildet sich
+konzentriertes Besitztum innerhalb eine Gesellschaft hat dies ab einem gewissen Grat
+einen Einfluss auf die Machtverteilung – sowohl in der physischen als auch in der 
+digitalen Welt. Ein gutes Beispiel hierfür sind die Registrare von Domains im Internet, 
+oder Firmen die über eine immense Rechenleistung verfügen. Die Entscheidungsposition in
+die diese Akteure durch ihr Eigentum geraten bringt ein hohes Maß an Verantwortung mit 
+sich, um eine möglichst ausgeglichene und demokratische Gesellschaft beizubehalten.
+
+Die ethische Betrachtung des Themas Eigentum im Zeitalter der Informationstechnik 
+erfordert daher eine tiefgreifende Auseinandersetzung mit ethischen Prinzipien wie 
+Machtverteilung, Chancengleichheit und Gerechtigkeit in der digitalen Welt.
+
+% Bezüge:
+% - Bezugspunkte
+% - Verantwortung 
+% - Menschen, Organisationen und die Gesellschaft
 

### 2023-10-08 netzwerk-sicherheit/übung-1.tex
Add the first exercise for network security

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Gruppe 11 – Mathias Baumbach \& Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang IT Sicherheit \\
		Fachbereich Informatik und Medien \\
		Netzwerksicherheit \\
		Prof.\ Dr.\ Michael Pilgermann
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 1}\\[0.5em]
		\large{Wintersemester 2023}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{flushright}
		Gruppe 11 \\
		Mathias Baumbach (Matr-Nr. 20213703) \\
		Mara Schulke (Matr-Nr. 20215853)
	\end{flushright}
\end{titlepage}

\begin{abstract}
	Innerhalb dieser Einsendeaufgabe werden verschiedene Aspekte der Netzwerksicherheit – vor 
	allem aus der Angreiferperspektive – betrachtet. Im Rahmen der Bearbeitung konnten wir 
	wertvolle Erfahrungen sammeln – gerade die Aufgabe 2.5 (Google-Hacking) hat uns erneut vor 
	Augen geführt wie wichtig eine entsprechende Absicherung von IT-Systemen ist, da wir 
	innerhalb von wenigen Minuten vollen Zugriff auf die Datenbank eines PHP Web-Servers 
	erlangen konnten. Wir haben den Betreiber informiert und anonymisiert. So, dass diese 
	Dokumentation nicht zu einer weiteren Ausnutzung verwendet werden kann.
\end{abstract}

\tableofcontents

\listoffigures

\newpage

\section{Vorbereitungen}

Die bevorzugte Methode zur Beantwortung der Einsendeaufgabe besteht darin, dass Sie 
entweder ParrotOS oder Kali Linux als virtuelle Maschine auf Ihrem Rechner installieren. 
Diese beiden Linux-Distributionen sind speziell auf die IT-Sicherheit ausgelegt und haben 
daher viele interessante Tools schon vorinstalliert. 
Tipp zur Durchführung: Laden Sie von ParrotOS  eine .ova für virtuelle Maschinen herunter 
(genauer die .ova der Security Edition; Tipp: das Superuser-Password lautet dann "parrot" 
und nicht "toor" wie bei anderen Versionen). Falls noch nicht vorhanden, installieren Sie 
Virtualbox  und führen die virtuelle Maschine darin aus.
Leider gibt es in der letzten Zeit zunehmend Prozessorarchitekturen (gerade für Mac), bei 
denen keine Möglichkeit zum Betrieb von virtuellen Maschinen besteht. Sollte das der Fall 
sein, dann bearbeiten Sie die Aufgaben auf Ihrem eigenen Betriebssystem. Sie müssen dafür 
die Tools einzeln herunterladen.

\newpage

\section{Durchführung}

\subsection{Telnet und Wireshark}

\subsubsection*{Aufgabenstellung}

Wireshark ist bei ParrotOS und Kali Linux vorinstalliert. Es kann ansonsten von der 
Wireshark-Seite heruntergeladen werden. In der Wireshark-Datei im Kurs (telnet.pcapng) ist 
eine Aufzeichnung eines Einlog-Vorgangs mit Telnet zu finden. Suchen Sie darin das 
Passwort.

\subsubsection*{Antwort}

Nach dem wir die Datei analysiert und den Telnet-Loginvorgang

%
%\begin{figure}[H]
%	\includegraphics[width=0.75\textwidth]{images/01-telnet-start.png}
%	\centering
%	\caption{Telnet Datei}
%\end{figure}

\newpage

\subsection{Schwachstellenscan I}

\subsubsection*{Aufgabenstellung}

Verwenden Sie \texttt{nmap} (bei ParrotOS und Kali Linux vorinstalliert), um verschiedene 
Scans des Testservers nwsmooc.mooin.org durchzuführen. Erklären Sie die Ergebnisse, wobei 
mindestens drei Tests mit jeweils unterschiedlichen Parametern durchgeführt werden müssen. 
Versuchen Sie dabei u.a. herauszufinden, welche Dienste auf dem Zielserver installiert
sind und welches Betriebssystem verwendet wird.

\subsubsection*{Antwort}

\newpage

\subsection{MAC Spoofing}

\subsubsection*{Aufgabenstellung}

Ebenfalls bereits bei ParrotOS und Kali schon vorinstalliert ist das Tool 
\texttt{macchanger}. Machen Sie sich mit dessen Möglichkeiten vertraut, wobei z.B. ein 
Video von HackerSploit nützlich sein kann: \url{https://www.youtube.com/watch?v=bshXz5r-CQA.}
Senden Sie anschließend Datenverkehr mit gefälschter MAC-Adresse und zeichnen diesen mit 
Wireshark auf. Fertigen Sie geeignete Screenshots mit Erklärungen (wie müsste es richtig 
sein? wo ist die gefälschte MAC-Adresse in Wireshark zu sehen?) dazu an.
Mac- und Linux-Nutzer verwenden für die Aufgabe ebenfalls \texttt{macchanger} oder 
\texttt{ifconfig}. Windows-Nutzer schauen sich bitte das Youtube-Tutorial an:
\url{https://www.youtube.com/watch?v=V3Pcc8b_m0U.}
Hinweis: Bei der ersten Methode heißt der Eintrag in den erweiterten Einstellungen nicht
"Network Address", sondern "Locally Administered Address".

\subsubsection*{Antwort}

\newpage

\subsection{Schwachstellenscan II}

\subsubsection*{Aufgabenstellung}

Verwenden Sie das bei ParrotOS und Kali Linux vorinstallierte WPScan  (WordPress 
Vulnerability Scanner) und führen Sie einen Scan von https://nwsmooc.mooin.org durch. 
Erklären Sie die Ergebnisse.

\begin{flushleft}
	Hinweise:	
\end{flushleft}

\begin{itemize}
	\item Es ist keine Registrierung beim Anbieter erforderlich. Durch eine Registierung würde man ein Token erhalten, um Schwachstellentests durchführen zu können. So ist die Aufgabe auf eine Informationssammlung beschränkt.
	\item Bei einem Test mit ParrotOS gab es zunächst eine Fehlermeldung, dass ein Update der Datenbank nicht möglich sei. Ein allgemeines Update
		(\texttt{sudo apt-get update \&\& apt-get upgrade}) konnte dieses Problem beseitigen.
\end{itemize}

Sollten Sie Mac oder Linux verwenden, dann installieren Sie WPScan direkt von Github. 
Sollten Sie keine Möglichkeit zur Durchführung dieses Aufgabenteils finden, sprechen 
Sie die Betreuenden auf eine Ersatzaufgabe an.

\subsubsection*{Antwort}

\newpage

\subsection{Google-Hacking}

\subsubsection*{Aufgabenstellung}

Mit "Google Hacking" ist gemeint, dass man die Google Suche zum Auffinden von S
oftwareinstallationen mit Schwachstellen nutzen kann. Eine Sammlung von Beispielen ist 
bei \url{https://www.exploit-db.com/google-hacking-database/} zu finden. Erklären Sie 
anhand von drei selbstgewählten Beispielen, was man damit herausfinden kann.
Achtung: Firefox und Google Chrome warnten teilweise beim Aufruf der Seite und bezeichneten diese als riskant. Man kann die Seite aber aus einem Browser innerhalb von Kali Linux oder ParrotOS aufrufen, dann kommt keine Warnung.

\subsubsection*{Antwort}

\end{document}

### 2023-10-16 netzwerk-sicherheit/übung-1.tex
Update the network security assignment

@@
 
-Nach dem wir die Datei analysiert und den Telnet-Loginvorgang
+Nach dem wir die Datei analysiert und den Telnet-Login-Vorgang bei dem Paket 
+\textcolor{red}{XY} lokalisiert haben, konnten wir den Start des Passwords in Text-Form 
+bei dem Paket \textcolor{red}{YZ} entdeckend. (Siehe Abb. 1).
 
-%
-%\begin{figure}[H]
-%	\includegraphics[width=0.75\textwidth]{images/01-telnet-start.png}
-%	\centering
-%	\caption{Telnet Datei}
-%\end{figure}
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/01}
+	\centering
+	\caption{Telnet Datei: Start des Logins}
+\end{figure}
+
+Nun enthielt jedes weitere Paket ein Zeichen des Passworts, bis die End-Sequenz
+in dem Paket \textcolor{red}{YY} gesendet wurde. Zusammengesetzt ergibt dies das 
+Passwort \texttt{sshnutzen}. (Siehe Abb. 2).
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/02}
+	\centering
+	\caption{Telnet Datei: Login Sequenz Dauer}
+\end{figure}
 
@@
 \subsubsection*{Antwort}
+
+Der Aufgabe entsprechend haben wir drei sinnvolle \texttt{nmap} scans durchgeführt.
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/03}
+	\centering
+	\caption{nmap: Scan-Techniken}
+\end{figure}
+
+\subsubsection*{Port Scan}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/04}
+	\centering
+	\caption{nmap: Port Scan}
+\end{figure}
+
+\subsubsection*{TCP Scan}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/05}
+	\centering
+	\caption{nmap: TCP Scan}
+\end{figure}
+
+\subsubsection*{OS Scan}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/06}
+	\centering
+	\caption{nmap: OS Scan}
+\end{figure}
+
+\subsubsection*{UDP Scan}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/07}
+	\centering
+	\caption{nmap: UDP Scan}
+\end{figure}
 
@@
 \subsubsection*{Antwort}
+
+Um die MAC-Adresse unter Kali-Linux zu ändern genügte eine einfache Kombination aus dem 
+Programm \texttt{macchanger} und \texttt{ifconfig}. Zu erst, um eine erfolgreiche Änderung 
+bzw. ein erfolgreiches Spoofing feststellen zu können, haben wir den Netzwerk-Verkehr zum 
+Web-Server der vorherigen Aufgaben wiederverwendet. (Siehe Abb. X)
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/08}
+	\centering
+	\caption{macspoofing: Ursprünglicher Netzwerk-Verkehr}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/09}
+	\centering
+	\caption{macspoofing: Ursprüngliche Source und Destination}
+\end{figure}
+
+In den beiden Screenshots ist zu erkennen, dass die momentane MAC-Adresse 
+\texttt{XX:XX:XX} entspricht.
+
+Nun muss zu erst das \texttt{eth0} Interface mittels \texttt{ifconfig} deaktiviert werden. 
+Anschließend kann die MAC-Adresse mittels \texttt{macchanger} geändert werden. In diesem 
+Beispiel haben wir lediglich die XYZ-Nummer von XYZ zu XYZ geändert und die Vendor-ID 
+belassen, da eine Änderung dieser zu Problemen mit den Treibern führen kann.
+Anschließend muss das Netzwerk-Interface erneut durch \texttt{ifconfig} gestartet werden.
+(Siehe Abb. X-Y)
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/10}
+	\centering
+	\caption{macspoofing: Änderung der MAC-Adresse von eth0}
+\end{figure}
+
+Bei erneuter Aufzeichnung des Traffics zum Zielserver lässt sich nun die gerade geänderte 
+MAC-Adresse in dem Ethernet-Header jedes Pakets wiederfinden. (Siehe Abb. Z)
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/11}
+	\centering
+	\caption{macspoofing: Veränderter Netzwerk-Verkehr}
+\end{figure}
 
@@
 
+Der Aufgabe entsprechend haben wir einen Schwachstellen-Scan bei dem vorgegebenen 
+Zielserver durchgeführt. Dieser hat das folgende Ergebnis erzielt:
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/12}
+	\centering
+	\caption{wpscan: Fehler}
+\end{figure}
+
+Nun musste die flag angehängt werden
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/13}
+	\centering
+	\caption{wpscan: Ergebnis 1/2}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/14}
+	\centering
+	\caption{wpscan: Ergebnis 2/2}
+\end{figure}
+
+Hieraus lassen sich folgende Informationen ableiten: XYZ
+
 \newpage
@@
 
-Mit "Google Hacking" ist gemeint, dass man die Google Suche zum Auffinden von S
-oftwareinstallationen mit Schwachstellen nutzen kann. Eine Sammlung von Beispielen ist 
+Mit "Google Hacking" ist gemeint, dass man die Google Suche zum Auffinden von
+Softwareinstallationen mit Schwachstellen nutzen kann. Eine Sammlung von Beispielen ist 
 bei \url{https://www.exploit-db.com/google-hacking-database/} zu finden. Erklären Sie 
@@
 
+Innerhalb der Exploit-Datenbank haben wir uns drei, unserer Meinung nach, besonders 
+interessante Beispiele ausgesucht.
+
+Das erste unserer Beispiele beruht auf dem Prinzip Google-Filter zu verwenden um von 
+Google indexierte Dateien mit bestimmten Namen oder Dateiendungen zu finden. So können zum 
+Beispiel falsch konfigurierte Web-Server aus Versehen vertrauliche Daten öffentlich 
+zugänglich machen. Ein interessantes Beispiel ist hierbei nach rohen SQL-Backups zu 
+suchen. Diese könnten, bei schlechten Sicherheitsmaßnahmen des Servers, rohe Nutzerdaten 
+enthalten und somit entweder einen Angriff auf die betroffenen Accounts ermöglichen oder 
+weiter verwendet werden um Zugriff auf andere, möglicherweise interessantere, Systeme zu 
+erlangen.
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/15}
+	\centering
+	\caption{Google Hacking: Backup: Schwachstelle}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/16}
+	\centering
+	\caption{Google Hacking: Backup: Google Suche}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/17}
+	\centering
+	\caption{Google Hacking: Backup: Ergebnis}
+\end{figure}
+
+Das zweite unserer Beispiele ist die Suche nach Webseiten die Tokens o.ä. vertraulichen 
+Informationen im Klartext enthalten. Dies ist ein besonders hohes Risiko bei Seiten, die 
+Quellcode enthalten (Wie z.B. Pastebin oder GitHub). Das Beispiel in unserer Abbildung 
+sucht nach Tokens auf öffentlich-zugänglichen Pastebin-Seiten.
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/18}
+	\centering
+	\caption{Google Hacking: Tokens: Suche}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/19}
+	\centering
+	\caption{Google Hacking: Tokens: Ergebnis}
+\end{figure}
+
+Die letzte, und wohl kritischte, Schwachstelle, die wir im Rahmen dieser Aufgabe gefunden 
+haben,
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/20}
+	\centering
+	\caption{Google Hacking: SQL Injektion: Schwachstelle}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/21}
+	\centering
+	\caption{Google Hacking: SQL Injektion: Schwachstelle}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/22}
+	\centering
+	\caption{Google Hacking: SQL Injektion: Suche}
+\end{figure}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/23}
+	\centering
+	\caption{Google Hacking: SQL Injektion: Ergebnis}
+\end{figure}
+
 \end{document}

### 2023-10-18 netzwerk-sicherheit/übung-1.tex
Update network sec exercises

@@
 Nach dem wir die Datei analysiert und den Telnet-Login-Vorgang bei dem Paket 
-\textcolor{red}{XY} lokalisiert haben, konnten wir den Start des Passwords in Text-Form 
-bei dem Paket \textcolor{red}{YZ} entdeckend. (Siehe Abb. 1).
+22 lokalisiert haben, konnten wir den Start des Passwords in Text-Form 
+bei dem Paket 23 entdeckend. (Siehe Abb. 1).
 
@@
 Nun enthielt jedes weitere Paket ein Zeichen des Passworts, bis die End-Sequenz
-in dem Paket \textcolor{red}{YY} gesendet wurde. Zusammengesetzt ergibt dies das 
+in dem Paket 46 gesendet wurde. Zusammengesetzt ergibt dies das 
 Passwort \texttt{sshnutzen}. (Siehe Abb. 2).
@@
 weiter verwendet werden um Zugriff auf andere, möglicherweise interessantere, Systeme zu 
-erlangen.
+erlangen. (Siehe Abb. 15)
 
@@
 \end{figure}
+
+Bei einer Google-Suche nach Dateiverzeichnissen, die eine Datei namens backup.zip enthalten
+sind wir sofort auf einen Web-Server gestoßen, bei dem sowohl eine backup.zip und eine backup.sql
+Datei versehentlich veröffentlicht wurden. (Siehe Abb. 16 \& 17)
 
@@
 
-Das zweite unserer Beispiele ist die Suche nach Webseiten die Tokens o.ä. vertraulichen 
+---
+
+Das zweite unserer Beispiele ist die Suche nach Webseiten die Tokens oder ähnlich vertrauliche 
 Informationen im Klartext enthalten. Dies ist ein besonders hohes Risiko bei Seiten, die 
@@
 sucht nach Tokens auf öffentlich-zugänglichen Pastebin-Seiten.
+
+Bei einer Suche nach pastebin.com Seiten, die entweder einen Benutzernamen, ein Passwort oder
+einen Schlüssel enthalten haben wir ebenfalls sofort Ergebnisse erzielen können.
+(Siehe Abb. 18 \& 19)
 
@@
 
-Die letzte, und wohl kritischte, Schwachstelle, die wir im Rahmen dieser Aufgabe gefunden 
-haben,
+---
+
+Die letzte, und wohl kritischte Schwachstelle, die wir im Rahmen dieser Aufgabe gefunden 
+haben, ist eine SQL-Injektion durch einen falsch programmierten PHP-Web-Server. Die Exploit-Datenbank
+listet hierzu drei typische Suchbefehle, wobei wir uns für den ersten entschieden haben, da er
+die größte Erfolgswahrscheinlichkeit hatte. (Siehe Abb. 20 \& 21)
 
@@
 
+Bei einer Google-Suche, kam neben einem PHP-Tutorial, sofort eine öffentliche PHP-Datei,
+die rohe SQL-Befehle als Query-Parameter annimmt. (Siehe Abb. 22)
+
 \begin{figure}[H]
@@
 \end{figure}
+
+Beim Aufruf dieser Seite wurden ursprünglich lediglich nicht vertrauliche Datensätze geladen,
+allerdings konnten wir zeigen, dass das SQL-Statement austauschbar ist, in dem wir (wie in Abb. 23 gezeigt)
+einen \texttt{show tables;} Befehl ausgeführt haben um die verschiedenen Tabellen in der angebundenen Datenbank
+aufzulisten.
 
@@
 
+Diese Schwachstelle könnte nun genutzt werden, um Beispielsweise neue Admin-Nutzer anzulegen,
+bestehende zu löschen, die Daten innerhalb der Datenbank zu manipulieren oder abzufragen etc.
+Über die Abfrage nach weiteren Tabellen hinaus, haben wir keine Befehle ausgeführt um strafrechtliche
+Probleme zu vermeiden.
+
 \end{document}

### 2023-10-21 netzwerk-sicherheit/übung-1.tex
Update network sec doc

@@
 Der Aufgabe entsprechend haben wir drei sinnvolle \texttt{nmap} scans durchgeführt.
+Auf den im Internet verfügbaren ubuntu manpage Seiten sind verschiedene Scan-Techniken mit
+\texttt{nmap} aufgelistet. (Siehe Abb. 3).
 
@@
 
+Nach kurzer Recherche haben wir uns für die folgenden vier Scan-Techniken entschieden, da sie in 
+Kombinationen eine gute Aussage-Stärke über das Host-System des Servers und die darauf ausgeführten
+Programme treffen.
+
 \subsubsection*{Port Scan}
+
+Eine der bekanntesten Techniken mit \texttt{nmap} ist der Port-Scan – dieser kann entweder auf die
+``well-known'' Ports beschränkt werden, oder über alle ports von 0-65536 ausgeführt werden. Da wir 
+im Rahmen dieser Aufgabe eine Aussage über die Dienste des Servers treffen sollen, haben wir uns dafür
+entschieden einen kompletten Port-Scan auszuführen. (Siehe Abb. 4)
 
@@
 
+Hier lässt sich erkennen, dass der Server einen SSH-Server, einen HTTP- bzw. HTTPS-Server betreibt.
+Diese Server sind öffentlich und für das Internet zugänglich. Auf dem Port 3544 wird vermutlich eine
+Torredo-Instanz ausgeführt, \texttt{nmap} kann aber keine genaue Aussage darüber treffen, ob dieser
+Dienst wirklich aktiv ist (daher wurde dieser Port als \texttt{CLOSED} markiert).
+
 \subsubsection*{TCP Scan}
+
+Als nächstes haben wir einen tiefgreifenden TCP-Scan ausgeführt, um detailiertere Informationen über
+die drei (im vorherigen Schritt entdeckte) Dienste herauszufinden. (Siehe Abb. 5)
 
@@
 
-\subsubsection*{OS Scan}
+In diesem Scan konnten wir weitere Informationen herausfinden, wie z.B. die Versionen / Distributionen
+des SSH- und HTTP-Servers, die öffentlichen SSH-Schlüssel des Servers und die Konfiguration des Apache-
+Servers.
+
+\subsubsection*{Betriebsystem Scan}
+
+Um eine Aussage über das ausgeführte Betriebssystem treffen zu können, bietet \texttt{nmap} einen 
+dedizierten Betriebssystem-Scan an – dieser kann durch \texttt{-O} gestartet werden. (Siehe Abb. 6)
 
@@
 
+Das Ergebnis dieses Scans ist zwar nicht eindeutig, lässt aber darauf deuten, dass der Server eine Linux-
+Distribution ausführt.
+
 \subsubsection*{UDP Scan}
+
+Zuletzt haben wir noch die bisher wenig betrachteten UDP-Dienste des Servers gescannt. (Siehe Abb. 7)
 
@@
 \end{figure}
+
+Hierbei wurden mehrere Dienste gefunden. Zwei der 4 UDP-Dienste stehen in Verbindung mit Microsoft,
+was entweder auf eine Windows-basierte Firewall oder einen Irrtum bei dem Betriebsystems-Scan hindeutet.
+
+Jeder Dienst hat den Status \texttt{open|filtered}, was auf eine aktive Firewall-Konfiguration hindeutet.
 
@@
 bzw. ein erfolgreiches Spoofing feststellen zu können, haben wir den Netzwerk-Verkehr zum 
-Web-Server der vorherigen Aufgaben wiederverwendet. (Siehe Abb. X)
+Web-Server der vorherigen Aufgaben aufgezeichnet. (Siehe Abb. 8)
 
@@
 
+Anschließend haben wir mittels \texttt{ifconfig} auf dem Host und in der virtuellen Maschine
+die MAC-Adressen ausgelesen und mit dem Ethernet-Adressen aus dem aufgezeichneten Netzwerk-Verkehr
+in Verbindung gebracht. (Siehe Abb. 9)
+
+Die Source-Adresse (\texttt{eth0} in der VM) lautet: \texttt{00:15:5d:cf:a4:3a}
+
+Die Source-Adresse (\texttt{WSL} auf dem Host) lautet: \texttt{00:15:5d:5d:86:e8}
+
 \begin{figure}[H]
@@
 
-In den beiden Screenshots ist zu erkennen, dass die momentane MAC-Adresse 
-\texttt{XX:XX:XX} entspricht.
+In den beiden Screenshots ist zu erkennen, dass die momentane MAC-Adresse der virtuellen
+Maschine \texttt{00:15:5d:cf:a4:3a} entspricht.
 
@@
 Anschließend kann die MAC-Adresse mittels \texttt{macchanger} geändert werden. In diesem 
-Beispiel haben wir lediglich die XYZ-Nummer von XYZ zu XYZ geändert und die Vendor-ID 
-belassen, da eine Änderung dieser zu Problemen mit den Treibern führen kann.
-Anschließend muss das Netzwerk-Interface erneut durch \texttt{ifconfig} gestartet werden.
-(Siehe Abb. X-Y)
+Beispiel haben wir lediglich die Hersteller-Nummer von \texttt{cf:a4:3a} zu \texttt{cf:a4:01}
+geändert und die Vendor-ID belassen, da eine Änderung dieser zu Problemen mit den Treibern
+führen kann. Anschließend muss das Netzwerk-Interface erneut durch \texttt{ifconfig} gestartet werden.
+(Siehe Abb. 10)
 
@@
 
-Bei erneuter Aufzeichnung des Traffics zum Zielserver lässt sich nun die gerade geänderte 
-MAC-Adresse in dem Ethernet-Header jedes Pakets wiederfinden. (Siehe Abb. Z)
+Bei erneuter Aufzeichnung des Netzwerk-Verkehrs zum Zielserver lässt sich nun die gerade geänderte 
+MAC-Adresse (\texttt{00:15:5d:cf:a4:01}) in dem Ethernet-Header jedes Pakets wiederfinden.
+(Siehe Abb. 11)
 

### 2023-11-04 netzwerk-sicherheit/esa-2/paper.tex
Add esa 2

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Gruppe 11 – Mathias Baumbach \& Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang IT Sicherheit \\
		Fachbereich Informatik und Medien \\
		Netzwerksicherheit \\
		Prof.\ Dr.\ Michael Pilgermann
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe 2}\\[0.5em]
		\large{Wintersemester 2023}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{flushright}
		Gruppe 11 \\
		Mathias Baumbach (Matr-Nr. 20213703) \\
		Mara Schulke (Matr-Nr. 20215853)
	\end{flushright}
\end{titlepage}

\begin{abstract}
\end{abstract}

\tableofcontents

\listoffigures

\newpage

\section{Durchführung}

\subsection{SSL Test I}

\subsubsection*{Aufgabenstellung}

Testen Sie den Server nwsmooc.mooin.org mit der SSL-Testseite von Qualys 
und erklären Sie die Ergebnisse (hinsichtlich Zertifikaten, TLS-Versionen,
Handshakes und Details der Protokolle). Erklären Sie für eine weitere
Webpräsenz, die als "Recent Worst" bewertet wird, was bei dieser nicht stimmt.

Hinweis: "Recent Worst" ist eine Liste auf der rechten Seite der Testseite. Zur
besseren Nachvollziehbarkeit bitte Screenshots hinzufügen. Deren Inhalte sollen
aber jeweils von Ihnen erklärt werden.

\subsubsection*{Antwort}

%\begin{figure}[H]
%	\includegraphics[width=0.75\textwidth]{images/01}
%	\centering
%	\caption{}
%\end{figure}

% https://www.ssllabs.com/ssltest/analyze.html?d=nwsmooc.mooin.org&s=81.169.230.170

https://www.ssllabs.com/ssltest/analyze.html?d=www.hoaez.com
Valid until 	Mon, 30 Oct 2023 23:00:02 UTC (expired 16 hours and 2 minutes ago)   EXPIRED
DNS CAA 	No (more info)
Trusted 	No   NOT TRUSTED (Why?)
Mozilla  Apple  Android  Java  Windows  
% → https://geekflare.com/de/dns-caa-record/
% Handshake Simulation - Chrome 49 / XP SP3 	Server sent fatal alert: handshake_failure, sonst alles grün

\newpage

\subsection{SSL Test II}

\subsubsection*{Aufgabenstellung}

Führen Sie den Client-Test von Qualys aus und erklären Sie die Ergebnisse.

\subsubsection*{Antwort}

https://clienttest.ssllabs.com:8443/ssltest/viewMyClient.html

\newpage

\subsection{SSL Test III}

\subsubsection*{Aufgabenstellung}

Rufen Sie die Testwebseiten https://aaacertificateservices.comodoca.com:442/ sowie 
https://aaacertificateservices.comodoca.com:444/  mit Firefox und mit Google 
Chrome auf. Erklären Sie, was passiert. Sind die Resultate bei allen vier Tests so 
wie erwartet?

\subsubsection*{Antwort}

\newpage

\subsection{Verschlüsselung}

\subsubsection*{Aufgabenstellung}

Installieren Sie VeraCrypt  auf Ihrem Rechner. Hierzu erhalten Sie zusätzlich eine 
VeraCrypt-Datei. In der Datei ist ein normaler und ein versteckter Container 
zu finden, die jeweils eine Datei enthalten. Das Passwort für den normalen 
Container ist der Exponent e des RSA-Schlüssels vom nwsmooc.mooin.org-Server. 
Dokumentieren Sie Ihre Vorgehensweise mit Screenshots und geben Sie 
anschließend das im versteckten Container gefundene Kennwort an.

Hinweis: Zur Bedienung von VeraCrypt können Sie sich beispielsweise hier ein Video 
ansehen: https://youtu.be/atb2pdxd394.

\subsubsection*{Antwort}

Um das zur Verfügung gestellte VeraCrypt File öffnen zu können, soll der Exponent 
e des RSA-Schlüssels der Webseite nwsmooc.mooin.org verwendet werden. Diesen 
Exponenten kann man sich ganz einfach ausgeben lassen, in dem man sich im Mozilla 
Firefox Browser jene Webseite öffnet und auf das Schlosssymbol drückt, um sich das 
Zertifikat anzusehen.

\newpage

\subsection{Steganographie}

\subsubsection*{Aufgabenstellung}

Installieren Sie SilentEye  auf Ihrem Rechner und untersuchen Sie die 
bereitgestellten Beispieldateien. mooin.jpg enthält ein verstecktes Kennwort, 
mooin.bmp eine Datei. Die notwendigen Einstellungen können Sie der Aufgabe mit 
VeraCrypt entnehmen. Dokumentieren Sie Ihre Vorgehensweise mit Screenshots und 
geben Sie das gefundene Kennwort an.

Hinweis: Sollte es Schwierigkeiten geben, wenn Sie beide Dateien nacheinander 
untersuchen, dann schließen Sie SilentEye zwischendurch.

\subsubsection*{Antwort}

\end{document}

### 2023-11-04 netzwerk-sicherheit/esa-2/paper.tex
Import esa 2 text

@@
 
-Um das zur Verfügung gestellte VeraCrypt File öffnen zu können, soll der Exponent 
-e des RSA-Schlüssels der Webseite nwsmooc.mooin.org verwendet werden. Diesen 
+Um das zur Verfügung gestellte VeraCrypt File öffnen zu können, soll der Exponent e 
+des RSA-Schlüssels der Webseite nwsmooc.mooin.org verwendet werden. Diesen 
 Exponenten kann man sich ganz einfach ausgeben lassen, in dem man sich im Mozilla 
-Firefox Browser jene Webseite öffnet und auf das Schlosssymbol drückt, um sich das 
-Zertifikat anzusehen.
+Firefox Browser jene Webseite öffnet und auf das Schlosssymbol drückt, um sich 
+das Zertifikat anzusehen.
+
+Im Abschnitt „Öffentlicher Schlüssel – Informationen“ bekommt man diesen einfach 
+ausgegeben (Siehe Abbildung XXX). Wir merken an dieser Stelle an, dass wir diesen 
+Exponenten zunächst auch blind getestet hatten, da in der Praxis fast immer die 
+Zahl 65537 als Exponent verwendet wird. Für das RSA Verfahren werden große 
+Primzahlen bevorzugt verwendet und der Algorithmus muss dabei unterschiedliche 
+Operationen bei der Bearbeitung basierend auf dem Binärcode tätigen. 6553710 ist 
+binär kodiert 10000000000000001 (??????????????)
+was es zu einer effizienten Zahl für den Algorithmus macht (Einsen werden 
+multipliziert, Nullen potenziert).
+
+Mounted man mit VeraCrypt die bereitgestellte Datei und gibt als Passwort 65537 an, 
+erhält man Zugriff auf die Datei. (siehe Abbildung XXX)
+
+Der Inhalt des Textfiles „Datei-angezeigter-Container.txt“ lautet:	„Das 
+Zugangskennwort für den versteckten Container lautet: 	VerSteCon0815“
+
+Wir konnten uns an das Modul Digitaler Selbstschutz erinnern, in dessen Reportinhalt 
+VeraCrypt bereits vorkam, und haben den Container gecloned und ein 2. Mal gemounted, 
+wobei wir dieses Mal das Passwort VerSteCon0815 eingegeben haben.
+
+Damit konnten wir Zugriff auf den versteckten Container und seinem Inhalt „Datei-
+versteckter-Container.txt“ erhalten.
+
+Der Text dieser Datei lautet:
+
+Das gesuchte Kennwort lautet: HiddenVolumeVeraCrypt
+
+% Die Einstellungen von SilentEye, die Sie für die Fortgeschrittenenaufgabe brauchen, 
+% sind:
+
+% Key (für beide Dateien): NWSMOOC_Test
+% Passphrase (nur für jpg-Datei): SilentEye_Test
+
+Die anderen Einstellungen entsprechen den Default-Einstellungen
+
+Für mooin.jpg: 
+Luminance Interval = 5 
+Header Position = bottom
+CharSet = UTF8
+
+Für mooin.bmp (Oben beim "Media's encoding format" auf "BMP" wechseln):
+Image quality: 96,975% (normal); keine Veränderungen bei "Advanced"
+Aktivieren Sie zusätzlich zu "Compressed Data" jeweils auch "Encrypted Data" mit AES256. Dort muss der Key in beide Kästchen eingetragen werden.
+Über die SilentEye-Aufgabe erhalten Sie über mooin.bmp übrigens schon vorab eine Aufgabensammlung zur Klausurvorbereitung.
 
@@
 
+SilentEye ist eine Open-Source-Software, die zur Steganographie verwendet wird, um 
+zusätzliche Informationen in Bildern und Audiodateien zu verstecken oder auszulesen.
+
+Um mooin.jpg zu untersuchen, öffnen wir die Datei mit dem Programm Silenteye und 
+verwenden die Einstellungen aus der Lösung der Aufgabe 1.4.
+
+Dadurch kann der verborgene Bereich des Bildes erfolgreich entschlüsselt werden und 
+gibt als dekodierte Message den Text ``Kennwort: SteganographieInNWSMOOC'' frei.
+
+Mit diesem Kennwort und den bereitgestellten Einstellungen für das zweite Bild, lässt 
+sich auch dessen versteckter Inhalt sichtbar machen und entschlüsseln. Wir können an 
+dieser Stelle die Datei AufgabensammlungSS16.pdf extrahieren und abspeichern.
+
 \end{document}

### 2023-11-15 netzwerk-sicherheit/esa-2/paper.tex
Finish network sec 2-1 and 2-2

@@
 
-%\begin{figure}[H]
-%	\includegraphics[width=0.75\textwidth]{images/01}
-%	\centering
-%	\caption{}
-%\end{figure}
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/01}
+	\centering
+	\caption{QualysSSL Test – nwsmooc.mooin.org}
+\end{figure}
 
-% https://www.ssllabs.com/ssltest/analyze.html?d=nwsmooc.mooin.org&s=81.169.230.170
+Der Web-Server hat insgesamt eine A+-Bewertung erhalten und ist dementsprechend auf dem aktuellen Stand / robust was die SSL-Konfiguration angeht.
+Das Zertifikat des Servers ist ein 2048-Bit RSA-Zertifikat, das mit SHA256+RSA signiert wurde.
+In dem Report von Qualys sind einige wichtige Eckdaten über das Zertifikat des Servers enthalten:
 
-https://www.ssllabs.com/ssltest/analyze.html?d=www.hoaez.com
-Valid until 	Mon, 30 Oct 2023 23:00:02 UTC (expired 16 hours and 2 minutes ago)   EXPIRED
-DNS CAA 	No (more info)
-Trusted 	No   NOT TRUSTED (Why?)
-Mozilla  Apple  Android  Java  Windows  
-% → https://geekflare.com/de/dns-caa-record/
-% Handshake Simulation - Chrome 49 / XP SP3 	Server sent fatal alert: handshake_failure, sonst alles grün
+\begin{itemize}
+    \item Gültigkeit: Von 25. September 2023 bis zum 24. Dezember 2023
+    \item Aussteller: R3
+    \item Status: Das Zertifikat wurde nicht widerrufen.
+    \item Vertrauenswürdigkeit: Ja – Mozilla, Apple, Android, Java und Windows vertrauen diesem Zertifikat.
+\end{itemize}
+
+Außerdem gibt der Bericht die Zertifikate weiter oben in der Zertifikats-Kette an:
+
+\begin{itemize}
+    \item Zertifikat \#2: Aussteller R3, RSA 2048 Bit, gültig bis 15. September 2025.
+    \item Zertifikat \#3: Aussteller ISRG Root X1, RSA 4096 Bit, gültig bis 30. September 2024.
+\end{itemize}
+
+\vspace{0.5em}
+
+---
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/02}
+	\centering
+	\caption{QualysSSL Test – Recent Worst}
+\end{figure}
+
+Der Server www.hoaez.com hat lediglich eine Gesamtbewertung von T erhalten, was bedeutet, dass dem Zertifikat
+des Servers nicht vertraut werden kann (e.g. da es abgelaufen ist, oder durch eine Misskonfiguration). Um dennoch einen
+Report erstellen zu können, hat Qualys die Vertrauenswürdigkeit bis auf weiteres ignoriert.
+
+Unter der Zusammenfassung gibt es einen Abschnitt, der Details zum Zertifikat des Servers liefert. Diese deuten
+explizit darauf hin, dass das Zertifikat (zum Test-Zeitpunkt seit 16 Stunden) abgelaufen und somit nicht mehr gültig ist.
+Es ist anzunehmen, dass durch eine Erneuerung des Zertifikates, alle Probleme behoben werden können und der Server wieder
+eine ausreichend gute Bewertung erhalten würde.
+
+Das Zertifikat an sich ist verwendet Elliptic-Curve mit 256 Bits und wurde ebenfalls mit SHA256+RSA signiert.
+
+Dieser Report ist ein gutes Beispiel dafür, als Administrator im Idealfall auf sich-selbsterneuernde Zertifikate zu setzen,
+um einem Vertrauensverlust der Nutzer (durch eine Warnmeldung des Browsers) entgegen zu wirken.
 
@@
 
-Führen Sie den Client-Test von Qualys aus und erklären Sie die Ergebnisse.
+Führen Sie den Client-Test von Qualys aus und erklären Sie die Ergebnisse. (Siehe \url{https://clienttest.ssllabs.com:8443/ssltest/viewMyClient.html})
 
@@
 
-https://clienttest.ssllabs.com:8443/ssltest/viewMyClient.html
+Wir haben den Client-Test mit der Mozilla Firefox Version 119.0.1 durchgeführt und folgendes Ergebnis erhalten:
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/03}
+	\centering
+	\caption{QualysSSL Test – Browser Capabilities}
+\end{figure}
+
+Im Kopf der Ergebnisseite wird eine Zusammenfassung der Bewertung ausgegeben. Es wird zunächst 
+die Protokoll-Unterstützung mit grün/gut bewertet – in unserem Fall ob TLS 1.2 und TLS 1.3 unterstützt wird.
+Danach folgen spezifische Schwachstellentests:
+
+\begin{itemize}
+    \item \textbf{Curveball / CVE-2020-0601} - Schwachstelle im Windows-OS bei der
+    	Validierung von digitalen Zertifikaten die potentiell angreifbar für MitM macht.
+    \item \textbf{Logjam} - Sicherheitslücke im Diffie-Hellman-Schlüsselaustausch,
+		welcher bei verschlüsselten Verbindungen wie TLS verwendet wird. Ermöglicht den Schlüsselaustausch
+		zu manipulieren und verschlüsselte Verbindungen zu entschlüsseln.
+    \item \textbf{FREAK (Factoring RSA Export Keys)} – Sicherheitslücke, die Angreifern
+		es ermöglichte Verschlüsselung von HTTPS-Verbindungen zu umgehen.
+    \item \textbf{POODLE (Padding Oracle On Downgraded Legacy Encryption)} - Schwachstelle
+		in SSLv2/3 die es ermöglichte die Verschlüsselung zu umgehen.
+\end{itemize}
+
+Keine der Schwachstellen sind auf unserem Client vorhanden.
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/04}
+	\centering
+	\caption{QualysSSL Test – Browser Protocol Features}
+\end{figure}
+
+Es folgt eine Detailauflistung der Protokoll-Features, die ergänzt ist um eine Auflistung der 
+unterstützten Cipher Suites in Reihenfolge der Präferenz (von oben nach unten). Dabei werden viele 
+kryptographische Verfahren als grün/gut bewertet, einige aber als weak. Dies ist z.B. auf die Duldung 
+von SHA (anstatt SHA256) zurückzuführen, oder weil der CBC-Modus (Cipher Block Chaining) erlaubt wird. 
+CBC hat bekannte Schwächen, beispielsweise auf Angriffe wie den sogenannten "Padding Oracle Attack".
+Die Ermöglichung von CBC in Verbindung mit älteren Authentifizierungsalgorithmen wie SHA1 wird hier 
+vermutlich als unsicher (auf englisch: weak) angesehen.
+
+Unter der Überschrift \textbf{Protocol Details} wird als gut/grün hervorgehoben, dass TLS compression 
+deaktiviert ist. Dies war ein Feature in älteren TLS-Versionen und sollte Daten komprimieren vor der 
+Verschlüsselung, war jedoch angreifbar und wurde daher in Folgeversion deaktiviert. Ebenfalls 
+deaktiviert sein sollte die SSL 2 handshake compatibility – dies ist die bereits erwähnte POODLE 
+Schwachstelle. Die letzte farbliche Hervorhebung in diesem Abschnitt lautet \textbf{Session tickets – 
+No}. Warum die Deaktivierung von Session-Tickets als weak gekennzeichnet wurde, konnten wir nicht 100\% 
+feststellen. Der Blog der Betreiber-Webseite beschreibt diesen Eintrag hauptsächlich aus Sicht eines 
+Web-Server/-Seiten Betreibers so, „dass session tickets ein alternativer Session-Management Mechanismus 
+sei, welcher separate Encryption Keys verwendet, die selten rotiert werden und lieber nicht verwendet 
+werden sollten, wenn man die Implementation nicht genau im Detail versteht.“ (vgl.
+\url{https://blog.qualys.com/product-tech/2013/06/25/ssl-labs-deploying-forward-secrecy}, abgerufen am 
+09.11.2023). Es erschien uns eher als Vorteil denn als Nachteil dass dieses Feature im Firefox 
+deaktiviert sei.
+
+Den Abschluss der Ergebnisseite bildet der Abschnitt \textbf{Mixed Content Handling}, in dem die 
+Kompatibilität mit diversen Webtechnologien (z.B. CSS, Frames usw.) getestet wird.
 
@@
 
-Rufen Sie die Testwebseiten https://aaacertificateservices.comodoca.com:442/ sowie 
+Rufen Sie die Test-Webseiten https://aaacertificateservices.comodoca.com:442/ sowie 
 https://aaacertificateservices.comodoca.com:444/  mit Firefox und mit Google 
@@
 \subsubsection*{Antwort}
+
+ÜBERARBEITEN
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/05}
+	\centering
+	\caption{QualysSSL Test – Browser Protocol Features}
+\end{figure}
+
+\textbf{Google Chrome Ergebnisse:}
+
+\begin{itemize}
+    \item Der Zugriff auf die Site auf Port 442 führt zu einer Sicherheitswarnung,
+		die besagt, dass die Verbindung nicht privat ist.
+	\item Der Zugriff auf die Site auf Port 444 zeigt ebenfalls eine Sicherheitswarnung an,
+		die darauf hinweist, dass das Sicherheitszertifikat der Site nicht vertrauenswürdig ist.
+\end{itemize}
+
+\textbf{Mozilla Firefox Ergebnisse:}
+
+\begin{itemize}
+	\item Der Zugriff auf die Site auf Port 442 zeigt eine Warnung vor einem potenziellen Sicherheitsrisiko an.
+	\item Der Zugriff auf die Site auf Port 444 zeigt eine Warnung, dass das Sicherheitszertifikat der Site widerrufen wurde.
+\end{itemize}
+
+Hinsichtlich der Frage, ob die Ergebnisse wie erwartet sind:
+
+\begin{itemize}
+    \item \textbf{Sicherheitswarnungen:} Diese sind zu erwarten, wenn eine Website SSL/TLS-Probleme hat, wie zum Beispiel ein ungültiges, abgelaufenes oder widerrufenes Zertifikat, oder wenn das Zertifikat nicht mit dem Domainnamen übereinstimmt.
+Es scheint, dass das Zertifikat für die über Port 444 aufgerufene Site widerrufen wurde, was ein kritisches Sicherheitsproblem darstellt, während die Site auf Port 442 ein anderes Zertifikatproblem hat, das die Browser nicht akzeptieren (wie es könnte selbstsigniert, abgelaufen sein oder es könnte eine Hostnamen-Diskrepanz geben).
+
+    \item \textbf{Browser-Konsistenz:} Die Tatsache, dass beide Browser für beide Ports Fehler anzeigen, deutet darauf hin, dass das Problem bei den Zertifikaten der Website liegt und nicht bei den Browsern selbst.
+Dies ist ein konsistentes und erwartetes Verhalten.
+
+    \item \textbf{Unterschiede in den Browser-Warnungen:} Die Unterschiede in den Sicherheitswarnungen zwischen Chrome und Firefox sind auf die unterschiedlichen Arten zurückzuführen, wie diese Browser Sicherheitsprobleme behandeln und melden.
+Beide identifizieren jedoch korrekt, dass es ein Problem mit dem SSL/TLS-Zertifikat der Website gibt.
+\end{itemize}
+
+Angesichts der Screenshots können wir schlussfolgern, dass die Websites SSL/TLS-Konfigurationsprobleme haben, die dazu führen, dass beide Browser korrekt Sicherheitswarnungen anzeigen.
+Es wäre ratsam, diese Websites nicht ohne eine sichere Verbindung zu besuchen, da dies sensible Informationen einer Abfanggefahr aussetzen kann.
+Der Website-Administrator sollte die Zertifikatsprobleme lösen, um einen sicheren Zugang für die Benutzer zu gewährleisten.
+
+Diese Diskrepanz kann aus mehreren Gründen auftreten:
+
+\begin{enumerate}
+  \item \textbf{CRL und OCSP}: Chrome und Firefox verwenden möglicherweise unterschiedliche Methoden zur Überprüfung des Widerrufsstatus eines Zertifikats.
+Chrome überprüft möglicherweise nicht den Widerrufsstatus oder konnte den Certificate Revocation List (CRL) oder Online Certificate Status Protocol (OCSP) Server nicht erreichen.
+Wenn der Server nicht erreichbar ist, kann Chrome die Verbindung unter bestimmten Bedingungen zulassen.
+  \item \textbf{Caching}: Es könnte ein Caching-Problem geben, bei dem Chrome das Zertifikat oder seinen Status zwischengespeichert hat und keinen Live-Check durchführt, um zu sehen, ob das Zertifikat widerrufen wurde.
+  \item \textbf{Widerrufsinformationen}: Es ist auch möglich, dass Firefox aktualisierte Widerrufsinformationen hat, die Chrome nicht hat.
+Dies kann passieren, wenn die Widerrufsinformationen kürzlich aktualisiert wurden und Chrome das Update noch nicht erhalten hat.
+  \item \textbf{Browser-Konfiguration}: Die Sicherheitseinstellungen des Browsers könnten unterschiedlich konfiguriert sein.
+Zum Beispiel könnte Chrome so eingerichtet sein, dass es mit bestimmten Arten von Zertifikatsfehlern fortfährt, die Firefox nicht zulässt, oder umgekehrt.
+  \item \textbf{Certificate Pinning}: Chrome könnte das Zertifikat für diese Seite gepinnt haben, was die normale Überprüfung der Vertrauenskette umgehen würde, die sonst zu einer erkannten Widerrufsstatus führen könnte.
+  \item \textbf{Software-Versionen}: Wenn ein Browser nicht auf dem neuesten Stand ist, erkennt er möglicherweise neuere Arten von Zertifikaten oder Widerrufsmethoden nicht richtig.
+\end{enumerate}
+
+Aus Sicherheitsgründen ist es wichtig, dem restriktiveren Browser in einem solchen Szenario zu vertrauen, in diesem Fall Firefox, der darauf hinweist, dass das Zertifikat widerrufen wurde.
+Widerrufene Zertifikate deuten darauf hin, dass dem Zertifikat nicht mehr vertraut werden sollte, typischerweise aufgrund eines Kompromisses oder eines Ausstellungsfehlers.
+Es ist ratsam, das Problem weiter zu untersuchen, möglicherweise indem der Zertifikatsstatus auf anderen Systemen oder Netzwerken überprüft wird, um zu sehen, ob das Problem weiterhin besteht, und die Widerrufsliste der Zertifizierungsstelle zur Bestätigung zu konsultieren.
 
@@
 \subsubsection*{Antwort}
+
+ÜBERARBEITEN
 
@@
 
+ÜBERARBEITEN
+
 SilentEye ist eine Open-Source-Software, die zur Steganographie verwendet wird, um 

### 2023-11-15 netzwerk-sicherheit/esa-2/paper.tex
Finish network sec 2-3

@@
 
-Rufen Sie die Test-Webseiten https://aaacertificateservices.comodoca.com:442/ sowie 
-https://aaacertificateservices.comodoca.com:444/  mit Firefox und mit Google 
+Rufen Sie die Test-Webseiten \url{https://aaacertificateservices.comodoca.com:442/} sowie 
+\url{https://aaacertificateservices.comodoca.com:444/} mit Firefox und mit Google 
 Chrome auf. Erklären Sie, was passiert. Sind die Resultate bei allen vier Tests so 
@@
 
-ÜBERARBEITEN
+Um das Verhalten der beiden Browser näher zu betrachten haben wir uns dazu entschieden, die Tests simultan
+durchzuführen, um zu veranschaulichen, wann die beiden Browser sich anders verhalten. Als erstes haben wir den
+Server auf dem Port 442 getestet:
 
@@
 	\centering
-	\caption{QualysSSL Test – Browser Protocol Features}
+	\caption{Google Chrome vs Firefox – 442 – Unsafe / Unsafe}
 \end{figure}
 
-\textbf{Google Chrome Ergebnisse:}
+Bei diesem Web-Server / Port erkennen beide Browser direkt ein abgelaufenes Zertifikat und warnen den
+Nutzer davor, diese Webseite zu verwenden (da die Identität des Servers nicht sicher gestellt werden konnte).
 
-\begin{itemize}
-    \item Der Zugriff auf die Site auf Port 442 führt zu einer Sicherheitswarnung,
-		die besagt, dass die Verbindung nicht privat ist.
-	\item Der Zugriff auf die Site auf Port 444 zeigt ebenfalls eine Sicherheitswarnung an,
-		die darauf hinweist, dass das Sicherheitszertifikat der Site nicht vertrauenswürdig ist.
-\end{itemize}
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/06}
+	\centering
+	\caption{Google Chrome vs Firefox – 444 – Safe / Unsafe}
+\end{figure}
 
-\textbf{Mozilla Firefox Ergebnisse:}
+Bei dem zweiten Test (also dem Server auf Port 444) verhalten sich die Browser interessanterweise unterschiedlich:
+Chrome zeigt dies als sichere Verbindung an, obwohl Firefox warnt und angibt, dass das Zertifikat widerrufen worden sei.
 
-\begin{itemize}
-	\item Der Zugriff auf die Site auf Port 442 zeigt eine Warnung vor einem potenziellen Sicherheitsrisiko an.
-	\item Der Zugriff auf die Site auf Port 444 zeigt eine Warnung, dass das Sicherheitszertifikat der Site widerrufen wurde.
-\end{itemize}
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/07} 
+	\centering
+	\caption{Google Chrome vs Firefox – 444 – Details}
+\end{figure}
 
-Hinsichtlich der Frage, ob die Ergebnisse wie erwartet sind:
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/08}
+	\centering
+	\caption{Google Chrome vs Firefox – 444 – Chrome Zertifikat}
+\end{figure}
 
-\begin{itemize}
-    \item \textbf{Sicherheitswarnungen:} Diese sind zu erwarten, wenn eine Website SSL/TLS-Probleme hat, wie zum Beispiel ein ungültiges, abgelaufenes oder widerrufenes Zertifikat, oder wenn das Zertifikat nicht mit dem Domainnamen übereinstimmt.
-Es scheint, dass das Zertifikat für die über Port 444 aufgerufene Site widerrufen wurde, was ein kritisches Sicherheitsproblem darstellt, während die Site auf Port 442 ein anderes Zertifikatproblem hat, das die Browser nicht akzeptieren (wie es könnte selbstsigniert, abgelaufen sein oder es könnte eine Hostnamen-Diskrepanz geben).
+Auch bei näherer Betrachtung des Zertifikats im Chrome-Certificate-Viewer, lässt sich keine Spur von dem Widerruf finden.
 
-    \item \textbf{Browser-Konsistenz:} Die Tatsache, dass beide Browser für beide Ports Fehler anzeigen, deutet darauf hin, dass das Problem bei den Zertifikaten der Website liegt und nicht bei den Browsern selbst.
-Dies ist ein konsistentes und erwartetes Verhalten.
-
-    \item \textbf{Unterschiede in den Browser-Warnungen:} Die Unterschiede in den Sicherheitswarnungen zwischen Chrome und Firefox sind auf die unterschiedlichen Arten zurückzuführen, wie diese Browser Sicherheitsprobleme behandeln und melden.
-Beide identifizieren jedoch korrekt, dass es ein Problem mit dem SSL/TLS-Zertifikat der Website gibt.
-\end{itemize}
-
-Angesichts der Screenshots können wir schlussfolgern, dass die Websites SSL/TLS-Konfigurationsprobleme haben, die dazu führen, dass beide Browser korrekt Sicherheitswarnungen anzeigen.
-Es wäre ratsam, diese Websites nicht ohne eine sichere Verbindung zu besuchen, da dies sensible Informationen einer Abfanggefahr aussetzen kann.
-Der Website-Administrator sollte die Zertifikatsprobleme lösen, um einen sicheren Zugang für die Benutzer zu gewährleisten.
-
-Diese Diskrepanz kann aus mehreren Gründen auftreten:
+Die unterschiedlichen Verhaltensweisen der Browser könnten mehrere Gründe haben:
 
 \begin{enumerate}
-  \item \textbf{CRL und OCSP}: Chrome und Firefox verwenden möglicherweise unterschiedliche Methoden zur Überprüfung des Widerrufsstatus eines Zertifikats.
+  \item \textbf{CRL und OCSP}: Chrome und Firefox verwenden wahrscheinlich unterschiedliche Methoden zur Überprüfung des Widerrufsstatus eines Zertifikats.
 Chrome überprüft möglicherweise nicht den Widerrufsstatus oder konnte den Certificate Revocation List (CRL) oder Online Certificate Status Protocol (OCSP) Server nicht erreichen.
 Wenn der Server nicht erreichbar ist, kann Chrome die Verbindung unter bestimmten Bedingungen zulassen.
-  \item \textbf{Caching}: Es könnte ein Caching-Problem geben, bei dem Chrome das Zertifikat oder seinen Status zwischengespeichert hat und keinen Live-Check durchführt, um zu sehen, ob das Zertifikat widerrufen wurde.
+  \item \textbf{Caching}: Es könnte ein Caching-Problem vorliegen, bei dem Chrome das Zertifikat oder seinen Status zwischengespeichert hat und keinen Live-Check durchführt, um zu sehen, ob das Zertifikat widerrufen wurde.
   \item \textbf{Widerrufsinformationen}: Es ist auch möglich, dass Firefox aktualisierte Widerrufsinformationen hat, die Chrome nicht hat.
@@
   \item \textbf{Certificate Pinning}: Chrome könnte das Zertifikat für diese Seite gepinnt haben, was die normale Überprüfung der Vertrauenskette umgehen würde, die sonst zu einer erkannten Widerrufsstatus führen könnte.
-  \item \textbf{Software-Versionen}: Wenn ein Browser nicht auf dem neuesten Stand ist, erkennt er möglicherweise neuere Arten von Zertifikaten oder Widerrufsmethoden nicht richtig.
 \end{enumerate}
 
-Aus Sicherheitsgründen ist es wichtig, dem restriktiveren Browser in einem solchen Szenario zu vertrauen, in diesem Fall Firefox, der darauf hinweist, dass das Zertifikat widerrufen wurde.
-Widerrufene Zertifikate deuten darauf hin, dass dem Zertifikat nicht mehr vertraut werden sollte, typischerweise aufgrund eines Kompromisses oder eines Ausstellungsfehlers.
-Es ist ratsam, das Problem weiter zu untersuchen, möglicherweise indem der Zertifikatsstatus auf anderen Systemen oder Netzwerken überprüft wird, um zu sehen, ob das Problem weiterhin besteht, und die Widerrufsliste der Zertifizierungsstelle zur Bestätigung zu konsultieren.
+Da wir allerdings beide Browser kontrollieren und beide Browser zum Zeitpunkt des Tests auf dem neusten Stand waren und wir den Web-Server vorher noch nie
+verwendet haben, deutet dies auf eine Diskrepanz bei den CRL / OSCP Abfragen hin – wahrscheinlich, da unterschiedliche Informationen verwendet werden.
+
+In einem solchen Fall, sollte immer dem Browser mit Sicherheitsbedenken vertraut werden.
 

### 2023-11-18 netzwerk-sicherheit/esa-2/paper.tex
Finish network sec 2.4

@@
 
-ÜBERARBEITEN
+Um das zur Verfügung gestellte VeraCrypt File öffnen zu können, soll der Exponent 
+$e$ des RSA-Schlüssels der Webseite \texttt{nwsmooc.mooin.org} verwendet werden. 
+Diesen Exponenten kann man sich ganz einfach ausgeben lassen, in dem man sich im 
+Mozilla Firefox Browser jene Webseite öffnet und auf das Schlosssymbol drückt, um 
+sich das Zertifikat anzusehen.
 
-Um das zur Verfügung gestellte VeraCrypt File öffnen zu können, soll der Exponent e 
-des RSA-Schlüssels der Webseite nwsmooc.mooin.org verwendet werden. Diesen 
-Exponenten kann man sich ganz einfach ausgeben lassen, in dem man sich im Mozilla 
-Firefox Browser jene Webseite öffnet und auf das Schlosssymbol drückt, um sich 
-das Zertifikat anzusehen.
+\begin{figure}[H]
+	\includegraphics[width=0.675\textwidth]{images/09}
+	\centering
+	\caption{Vercrypt – Exponent}
+\end{figure}
 
-Im Abschnitt „Öffentlicher Schlüssel – Informationen“ bekommt man diesen einfach 
-ausgegeben (Siehe Abbildung XXX). Wir merken an dieser Stelle an, dass wir diesen 
-Exponenten zunächst auch blind getestet hatten, da in der Praxis fast immer die 
-Zahl 65537 als Exponent verwendet wird. Für das RSA Verfahren werden große 
-Primzahlen bevorzugt verwendet und der Algorithmus muss dabei unterschiedliche 
-Operationen bei der Bearbeitung basierend auf dem Binärcode tätigen. 6553710 ist 
-binär kodiert 10000000000000001 (??????????????)
-was es zu einer effizienten Zahl für den Algorithmus macht (Einsen werden 
-multipliziert, Nullen potenziert).
+Im Abschnitt ``Öffentlicher Schlüssel – Informationen'' bekommt man den Exponenten 
+des öffentlichen Schlüssels ausgegeben (siehe Abb. 9). Allerdings gelang uns der 
+Zugriff auf den Container auch vor der Analyse des Zertifikats, da $e = 65537_{10}$ 
+in der Praxis üblich ist.
 
-Mounted man mit VeraCrypt die bereitgestellte Datei und gibt als Passwort 65537 an, 
-erhält man Zugriff auf die Datei. (siehe Abbildung XXX)
+Für das RSA Verfahren werden große Primzahlen bevorzugt verwendet und der 
+Algorithmus muss dabei unterschiedliche Operationen bei der Bearbeitung basierend 
+auf dem Binärcode tätigen. Da $65537_{10} = 10000000000000001_2$ führt 
+dies zu einer effizienten Arbeitsweise des Algorithmus – Einsen werden 
+multipliziert, Nullen potenziert.
 
-Der Inhalt des Textfiles „Datei-angezeigter-Container.txt“ lautet:	„Das 
-Zugangskennwort für den versteckten Container lautet: 	VerSteCon0815“
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/10}
+	\centering
+	\caption{Vercrypt – Einbinden Container}
+\end{figure}
 
-Wir konnten uns an das Modul Digitaler Selbstschutz erinnern, in dessen Reportinhalt 
-VeraCrypt bereits vorkam, und haben den Container gecloned und ein 2. Mal gemounted, 
-wobei wir dieses Mal das Passwort VerSteCon0815 eingegeben haben.
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/11}
+	\centering
+	\caption{Vercrypt – Passworteingabe}
+\end{figure}
 
-Damit konnten wir Zugriff auf den versteckten Container und seinem Inhalt „Datei-
-versteckter-Container.txt“ erhalten.
+Nun können wir mit VeraCrypt (unter Angabe des Passworts 65537) die im Kurs 
+bereitgestellte Datei als Laufwerk einbinden und somit Zugriff auf den Inhalt 
+erhalten.
 
-Der Text dieser Datei lautet:
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/12}
+	\centering
+	\caption{Vercrypt – Einbinden Versteckter Container}
+\end{figure}
 
-Das gesuchte Kennwort lautet: HiddenVolumeVeraCrypt
+Die enthaltene Textdatei ``Datei-angezeigter-Container.txt'' lautet: ``Das 
+Zugangskennwort für den versteckten Container lautet: 	VerSteCon0815''
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/13}
+	\centering
+	\caption{Vercrypt – Versteckter Container}
+\end{figure}
+
+Wir konnten uns an das Modul Digitaler Selbstschutz erinnern, in dessen Report-
+Inhalt VeraCrypt bereits vorkam, und haben den Container kopiert und ein 2. Mal 
+eingebunden, wobei dieses Mal das Passwort ``VerSteCon0815'' eingegeben werden muss 
+um Zugriff auf den versteckten Container zu erhalten.
+
+Somit konnten wir Zugriff auf den versteckten Container und seinem Inhalt ``Datei-
+versteckter-Container.txt'' erlangen und den Inhalt auslesen: 
+``HiddenVolumeVeraCrypt''.
+
+\newpage
+
+\subsection{Steganographie}
+
+\subsubsection*{Aufgabenstellung}
+
+Installieren Sie SilentEye  auf Ihrem Rechner und untersuchen Sie die 
+bereitgestellten Beispieldateien. mooin.jpg enthält ein verstecktes Kennwort, 
+mooin.bmp eine Datei. Die notwendigen Einstellungen können Sie der Aufgabe mit 
+VeraCrypt entnehmen. Dokumentieren Sie Ihre Vorgehensweise mit Screenshots und 
+geben Sie das gefundene Kennwort an.
+
+Hinweis: Sollte es Schwierigkeiten geben, wenn Sie beide Dateien nacheinander 
+untersuchen, dann schließen Sie SilentEye zwischendurch.
+
+\subsubsection*{Antwort}
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/14}
+	\centering
+	\caption{Vercrypt – Hidden Container}
+\end{figure}
 
@@
 
-\newpage
-
-\subsection{Steganographie}
-
-\subsubsection*{Aufgabenstellung}
-
-Installieren Sie SilentEye  auf Ihrem Rechner und untersuchen Sie die 
-bereitgestellten Beispieldateien. mooin.jpg enthält ein verstecktes Kennwort, 
-mooin.bmp eine Datei. Die notwendigen Einstellungen können Sie der Aufgabe mit 
-VeraCrypt entnehmen. Dokumentieren Sie Ihre Vorgehensweise mit Screenshots und 
-geben Sie das gefundene Kennwort an.
-
-Hinweis: Sollte es Schwierigkeiten geben, wenn Sie beide Dateien nacheinander 
-untersuchen, dann schließen Sie SilentEye zwischendurch.
-
-\subsubsection*{Antwort}
+---
 

### 2023-11-19 netzwerk-sicherheit/esa-2/paper.tex
Finish network sec esa 2

@@
 
+SilentEye ist eine Open-Source-Software, die zur Steganographie verwendet wird, um 
+zusätzliche Informationen in Bildern und Audiodateien zu verstecken oder auszulesen.
+Wir haben die Software aus dem Internet heruntergeladen und unter Windows 
+installiert. Nach dem öffnen der Software ist es erforderlich eine Bild-Datei 
+auszuwählen, von der der Nutzer weiß oder vermutet, dass diese eine mittels 
+Steganographie codierte Nachricht enthält. (Siehe Abb. 14)
+
 \begin{figure}[H]
@@
 	\centering
-	\caption{Vercrypt – Hidden Container}
+	\caption{Steganographie – SilentEye UI}
 \end{figure}
 
-% Die Einstellungen von SilentEye, die Sie für die Fortgeschrittenenaufgabe brauchen, 
-% sind:
+Um der Aufgabenstellung entsprechend nun die Datei \texttt{mooin.jpg} zu untersuchen, 
+öffnen wir die Datei und verwenden die Einstellungen aus der Lösung der Aufgabe 1.4. 
+Anschließend zeigt SilentEye, unter Verwendung der richtigen Parameter, eine 
+decodierte Nachricht in einem Textfeld an – diese lautet
+\texttt{Kennwort: SteganographieInNWSMOOC} (Siehe Abb. 15)
 
-% Key (für beide Dateien): NWSMOOC_Test
-% Passphrase (nur für jpg-Datei): SilentEye_Test
-
-Die anderen Einstellungen entsprechen den Default-Einstellungen
-
-Für mooin.jpg: 
-Luminance Interval = 5 
-Header Position = bottom
-CharSet = UTF8
-
-Für mooin.bmp (Oben beim "Media's encoding format" auf "BMP" wechseln):
-Image quality: 96,975% (normal); keine Veränderungen bei "Advanced"
-Aktivieren Sie zusätzlich zu "Compressed Data" jeweils auch "Encrypted Data" mit AES256. Dort muss der Key in beide Kästchen eingetragen werden.
-Über die SilentEye-Aufgabe erhalten Sie über mooin.bmp übrigens schon vorab eine Aufgabensammlung zur Klausurvorbereitung.
-
----
-
-ÜBERARBEITEN
-
-SilentEye ist eine Open-Source-Software, die zur Steganographie verwendet wird, um 
-zusätzliche Informationen in Bildern und Audiodateien zu verstecken oder auszulesen.
-
-Um mooin.jpg zu untersuchen, öffnen wir die Datei mit dem Programm Silenteye und 
-verwenden die Einstellungen aus der Lösung der Aufgabe 1.4.
-
-Dadurch kann der verborgene Bereich des Bildes erfolgreich entschlüsselt werden und 
-gibt als dekodierte Message den Text ``Kennwort: SteganographieInNWSMOOC'' frei.
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/15}
+	\centering
+	\caption{Steganographie – JPG Entschlüsselung}
+\end{figure}
 
@@
 sich auch dessen versteckter Inhalt sichtbar machen und entschlüsseln. Wir können an 
-dieser Stelle die Datei AufgabensammlungSS16.pdf extrahieren und abspeichern.
+dieser Stelle die Datei AufgabensammlungSS16.pdf extrahieren und abspeichern. (Siehe 
+Abb. 16)
+
+\begin{figure}[H]
+	\includegraphics[width=0.75\textwidth]{images/16}
+	\centering
+	\caption{Steganographie – BMP Entschlüsselung}
+\end{figure}
 

### 2024-05-06 entwicklung-sicherer-software-systeme/exercise-1/exercise-1.tex
Start writing the exercise

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{float}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Studiengang IT Sicherheit \\
		Fachbereich Informatik und Medien \\
		Entwicklung Sicherer Softwaresysteme \\
		Prof. Dr.-Ing. Martin Schafföner
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe: Threat Analysis}\\[0.5em]
		\large{Sommersemester 2024}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{flushright}
		Mara Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\begin{abstract}
	In dieser Einsendeaufgabe wird die Online-Meeting-Plattform Google Meet der Firma 
	Google hinsichtlich ihrer softwareseitigen Risiken und Gefahren analysiert.
	Es wird die Vorgehensweise der Gefahrenanalyse, die entdeckten Risiken und Gefahren 
	zusammengefasst und eine Priorisierung mit Handlungsempfehlung daraus abgeleitet. Des 
	weiteren wird auf die Methodik der Gefahrenanalyse eingegangen und von alternativen 
	Vorgehensweisen abgegrenzt.
\end{abstract}

\tableofcontents

\listoffigures

\newpage

\section{Projektauswahl: Google Meet}

Die Auswahl des zu analysierenden Projektes viel auf die Onlne-Meeting-Plattform Google 
Meet da diese mehrere Punkte abdeckt:

\begin{enumerate}
	\item Ihre Funktionsweise- und -umfang ist ihm Rahmen einer Kurzanalyse greifbar
	\item Die Anwendungsarchitektur lässt sich aus vergleichbaren Echtzeit-Meeting Anwendungen ableiten und vereinfachen
	\item Sie hat einen weitreichenden Bekanntheitsgrad
\end{enumerate}

Vereinfacht lässt sich Google Meet in einer Client-Server Architektur darstellen, auch 
wenn an dieser Stelle angemerkt werden muss, dass durch die Menge an Nutzern, die Vielzahl 
an Integrationen und die Stabilität der Software davon auszugehen ist, dass es sich nicht 
um eine naive Architektur der Server bzw. Server-Infrastruktur handelt.

Im Rahmen dieser Einsendeaufgabe wird die Annahme getroffen, dass Google Meet der folgenden Architektur entlang aufgebaut ist:

%\begin{figure}[H]
%	\includegraphics[width=0.75\textwidth]{images/01-architecture.png}
%	\centering
%	\caption{Anwendungsarchitektur Google Meet}
%\end{figure}

Dies ist notwendig um in einem zeitlich angemessenen Rahmen in der Lage zu sein eine 
beispielhafte Gefahrenanalyse durchführen zu können.

\section{Auswahl der Gefahrenanalyse Software}

Um eine automatisierte Gefahrenanalyse durchzuführen stehen mittlerweile (Stand Mai 2024), 
eine Vielzahl an verschiedenen Tools zur Verfügung. Nachfolgend sind einige beliebtesten 
bzw. bekanntesten, kostenlosen \textit{Threat-Modelling}-Programme aufgelistet:

% [1]

\subsection*{Kostenlose Software}

Da diese Einsendeaufgabe von geringem Umfang ist und keine weiteren Anwendungsfälle für 
eine ausgereifte Threat-Modelling Software absehbar sind beschränke ich den Vergleich 
verschiedener Software-Lösungen auf kostenlose bzw. Open-Source Programme.

\subsubsection*{MTMT – Microsoft Threat Modeling Tool}

MTMT ist das Threat Modeling Tool der Firma Microsoft. Es ist ein weit verbreitetes 
Werkzeug zur Modellierung von Software und der Automatisierten Analyse dieser Modelle. Da 
der Anbieter Microsoft ist, besteht leider die Anforderung MTMT auf einem Windows-System 
auszuführen.

Das MTMT ist weiterhin ein sehr umfangreiches Werkzeug, das ebenfalls im professionellen 
Kontext Einsatz findet – dies resultiert in einer komplexen Oberfläche mit 
vielen Möglichkeiten und ausgereiften Konzepten.

Leider musste ich durch die Gegebenheit, kein Windows-System zu betreiben, die Analyse 
des MTMT hier beenden, da es mir leider nicht möglich war eine ausführbare Version für 
Unix-Systeme zu finden ohne Virtualisierung zu verwenden.

\subsubsection*{OWASP ThreatDragon}

\subsubsection*{AWS ThreatComposer}

\subsubsection*{Threagile}

Threagile ist kostenlose Open-Source Software zur Gefahrenanalyse ohne eine 
Benutzeroberfläche. Dies ist eine Eigenheit im Vergleich zu den anderen vorgestellten 
Programmen die hauptsächlich über das Desktop- oder Web-Interface zu bedienen sind.
Threagile lässt sich über eine .yaml Datei bedienen die einem vorgegebenen Schema 
entsprechen muss. So lassen sich im Quelltext Komponenten (und ihre Beziehungen), 
Datenbanken, Verschlüsselungen, Datenflüsse, verwendete Technologien und vieles mehr 
Dokumentieren.

Der klare Vorteil von Threagile liegt in der Möglichkeit diese YAML-Datei bzw. die 
analysierte Architektur ohne weiteres in git oder anderen VCS (kurz für \textit{Version 
Control Systems}) zu versionieren. Bei anderer proprietär Software ist dies zwar 
grundsätzlich Möglich aber da VCS in der Regel auf Klartext-Dateien ausgelegt sind sind 
die Möglichkeiten begrenzt bzw. parallel im Team an einer Datei zu arbeiten.

Des Weiteren ist es einfach mittels Threagile eine Gefahrenanalyse auf der YAML-Datei 
auszuführen und der generierte Report hat eine hohe Qualität (optisch, strukturell und 
inhaltlich).

\subsection{Begründung der Auswahl}

Im Rahmen dieser Einsendeaufgabe habe ich mich unter den oben dargestellten Abgrenzungen 
der verschiedenen Softwarelösungen für die Nutzung von Threagile entschieden. Threagile 
ist einfach zu bedienen, reproduzierbar (mittels Docker), eine nützliche Resource für 
zukünftige berufliche Situationen.

\section{Ergebnis der automatisierten Gefahrenanalyse}

\section{Detaillierte Gefahrenanalyse}

Analyze some of the threats in more detail.

\section{Priorisierung der Gefahren}

\section{Zusammenfassung}

% [1] https://www.iriusrisk.com/resources-blog/recommended-threat-modeling-tools

\end{document}

### 2024-05-07 entwicklung-sicherer-software-systeme/exercise-1/exercise-1.tex
Update ess stuff

@@
 
-\newpage
-
 \section{Projektauswahl: Google Meet}
@@
 
-Im Rahmen dieser Einsendeaufgabe wird die Annahme getroffen, dass Google Meet der folgenden Architektur entlang aufgebaut ist:
+Im Rahmen dieser Einsendeaufgabe wird die Annahme getroffen, dass Google Meet der 
+folgenden Architektur entlang aufgebaut ist:
 
-%\begin{figure}[H]
-%	\includegraphics[width=0.75\textwidth]{images/01-architecture.png}
-%	\centering
-%	\caption{Anwendungsarchitektur Google Meet}
-%\end{figure}
+\begin{figure}[H]
+	\includegraphics[width=0.5\textwidth]{./google-meet-diagram.png}
+	\centering
+	\caption{Anwendungsarchitektur Google Meet (stark vereinfacht)}
+\end{figure}
 
-Dies ist notwendig um in einem zeitlich angemessenen Rahmen in der Lage zu sein eine 
-beispielhafte Gefahrenanalyse durchführen zu können.
+Im Rahmen dieser Aufgabe wird von einer monolithischen Anwendungsarchitektur von Google 
+Meet ausgegangen in einem zeitlich angemessenen Rahmen in der Lage zu sein eine 
+hypothetische Gefahrenanalyse durchführen zu können. Der Google Meet Server in der 
+beistehenden Grafik übernimmt die Verwaltung der Sitzungen, die Account-Authentifizierung 
+und die Aushandlung der WebRTC-Sitzungen zwischen verschiedenen Nutzer.
 
@@
 
+OWASP ThreatDragon ist eine Software der Organization OWASP (kurz für \textit{Open 
+Worldwide Application Security Project}). Die Software ist kostenlos und
+Browser basiert. Sie enthält Funktionen für die Diagramm-Zeichnung und anschließende 
+Erstellung eines Sicherheits-Reports. Die Arbeitsstände müssen allerdings immer 
+heruntergeladen werden und in lokalen Dateien gespeichert werden – dies ist eine relativ 
+umständliche Arbeitsweise, da anders als bei nativer Software nicht eine Datei durchgehend 
+geöffnet und bearbeitet werden kann. Außerdem bietet ThreatDragon nur einen begrenzten 
+Umfang an erkannten Sicherheitslücken bei einer offensichtlich fehlerhaften 
+Softwarearchitektur.
+
 \subsubsection*{AWS ThreatComposer}
+
+AWS ThreatComposer ist eine Open Source Software der Firma Amazon. Sie ist ebenfalls 
+Browser basiert, allerdings etwas ausgereifter und nutzerfreundlicher als die OWASP 
+alternative. Dennoch gibt es auch hier einige Unannehmlichkeiten für Nutzer: Die 
+Benutzeroberfläche ist unklar (e.g. Diagramme sind nur zur Dokumentation hinzuzufügen, 
+nicht zur automatisierten Auswertung). Es gibt keine klare Datenverwaltung bzw. keinen 
+klaren Datenexport und der allgemeine Funktionsumfang wirkt etwas eingeengt.
 
@@
 
+Der automatisierte Report von Threagile hat basierend der oben dargestellten, stark 
+vereinfachten, Anwendungsarchitektur folgende Risiken identifizieren können.
+
+\subsection{}
+
 \section{Detaillierte Gefahrenanalyse}

### 2024-05-13 entwicklung-sicherer-software-systeme/exercise-1/exercise-1.tex
Update scenario
Update exercise 1 for ess

@@
 
-Die Auswahl des zu analysierenden Projektes viel auf die Onlne-Meeting-Plattform Google 
-Meet da diese mehrere Punkte abdeckt:
+Die Auswahl des zu analysierenden Projektes fiel auf die Online-Meeting-Plattform Google 
+Meet, da diese mehrere Punkte abdeckt:
 
@@
 \begin{figure}[H]
-	\includegraphics[width=0.5\textwidth]{./google-meet-diagram.png}
+	\includegraphics[width=0.75\textwidth]{./google-meet-diagram.png}
 	\centering
@@
 Um eine automatisierte Gefahrenanalyse durchzuführen stehen mittlerweile (Stand Mai 2024), 
-eine Vielzahl an verschiedenen Tools zur Verfügung. Nachfolgend sind einige beliebtesten 
-bzw. bekanntesten, kostenlosen \textit{Threat-Modelling}-Programme aufgelistet:
-
-% [1]
+eine Vielzahl an verschiedenen Tools zur Verfügung. Nachfolgend sind einige beliebte 
+bzw. bekannte, kostenlose \textit{Threat-Modelling}-Programme aufgelistet:
 
@@
 Der automatisierte Report von Threagile hat basierend der oben dargestellten, stark 
-vereinfachten, Anwendungsarchitektur folgende Risiken identifizieren können.
+vereinfachten, Anwendungsarchitektur hat insgesamt 21 mögliche Risiken und in 12 
+verschiedenen Risiko-Klassen identifizieren können.
 
-\subsection{}
+Folgende Risiko-Klassen wurden mit der Stufe \textit{Elevated – Wahrscheinlicher Angriff mit hohem Schaden} markiert:
 
-\section{Detaillierte Gefahrenanalyse}
+\begin{enumerate}
+	\item Cross-Site-Scripting 
+	\item Missing Cloud Hardening
+	\item Missing Hardening
+	\item Server-Side-Request-Forgery
+\end{enumerate}
 
-Analyze some of the threats in more detail.
+Alle anderen entdeckten Risiko-Klassen wurden mit der Stufe \textit{Medium – Unwahrscheinlicher Angriff mit hohem Schaden} markiert:
+
+\begin{enumerate}
+	\item Container Base Image Backdooring
+	\item Cross-Site-Request-Forgery
+	\item DoS-Risky Access Across Trust Boundary
+	\item Missing Identity Propagation
+	\item Missing Identity Provider Isolation
+	\item Missing Two-Factor Authentication
+	\item Missing Vault
+	\item Mixed Targets on Shared-Runtime
+\end{enumerate}
+
+Zusammenfassend lässt sich über die entdeckten Risiken sagen, dass diese ein konkretes,
+greifbares Risiko darstellen und Threagile wenige ``False-Positives'' entdeckt hat (bspw. 
+Fehler die lediglich aufgrund einer Ungenauigkeit der Diagramm-Funktion markiert werden). 
+Threagile lässt eine sehr detaillierte Dokumentation von Verschlüsselungs-Methoden, 
+übertragenen Datensätzen und verwendeten Datenformaten etc. zu. Die Liste der einzelnen 
+Risiken finden sie im Threagile-Report (Anhang 1).
 
@@
 
+Die logische Schlussfolgerung der oben eingestuften Risiko-Klassen ist die priorisierung 
+der Gefahren nach absolutem Risiko:
+
+$$Risiko = Schaden * Wahrscheinlichkeit$$
+
+\vspace{0.5em}
+
+Demzufolge bezieht sich diese Priorisierung lediglich auf die Gefahren aus der Klasse 
+\textit{Elevated}:
+
+\begin{enumerate}
+	\item Cross-Site Scripting (XSS) risk at Google Meet Server (\textit{Likely with High impact})
+	\item Missing Cloud Hardening risk at Google Cloud (\textit{Unlikely with Very High impact})
+	\item Missing Cloud Hardening risk at Public Internet (\textit{Unlikely with Very High impact})
+	\item Cross-Site Scripting (XSS) risk at Auth Server (\textit{Likely with Medium impact})
+	\item Missing Hardening risk at Google Meet Server (\textit{Likely with Medium impact})
+	\item Server-Side Request Forgery (SSRF) risk at Auth Server server-side web-requesting the target Google Meet Server via Google Meet Server (\textit{Likely with Medium impact})
+	\item Server-Side Request Forgery (SSRF) risk at Google Meet Server server-side web-requesting the target Auth Server via User Authentication Access: (\textit{Likely with Medium impact})
+	\item Server-Side Request Forgery (SSRF) risk at Google Meet Server server-side web-requesting the target Google Meet Client via Web Interface (\textit{Likely with Medium impact})
+\end{enumerate}
+
+\section{Detaillierte Betrachtung}
+
+\subsection{Cross-Site-Scripting (XSS)}
+
+Da Google Meet text-basierte Nutzereingaben als eine der Kernfunktionen (Chat) zulässt, 
+stellt Cross-Site-Scripting ein erhebliches Risiko dar. Allerdings muss an dieser Stelle
+angemerkt werden, dass die Empfänger einer bösartigen lediglich die Nutzer einer Sitzung 
+sind (meist, ein bis zweistellige Nutzerzahlen) – wohingegen auf Plattformen wie Twitter 
+und YouTube etc. Nutzer öffentlich Beiträge posten können was die Anzahl an betroffenen 
+Nutzern drastisch erhöht.
+
+Davon unberührt ist allerdings die Ausführung von Code bzw. Befehlen auf dem Google 
+Meet Server (auch bekannt als \textit{RCE – Remote Code Execution}) mittels XSS.
+
+\subsection{Missing Cloud Hardening}
+
+Die fehlende Absicherung der Cloud gegen unbefugten Zugriff (bspw. via SSH, 
+Netzwerkschwachstellen etc) stellt neben XSS die größte Gefahr dar, da diese mit einer 
+möglichen ``Escalation of Privilege'' einhergehen könnte und somit ein Angreifer nach der 
+infiltration einer der Cloud-Internen-Komponenten wohlmöglich weitere Komponenten 
+infizieren könnte. Mögliche Lösungsstrategien zur Absicherung der Cloud bestehen darin,
+verwaltete Services (für Netzwerkverbindungen, Datenbanken, Load-Balancing, Server-
+Hosting) der GCP in Anspruch zu nehmen, anstatt mittels VM eine manuelle Infrastruktur 
+einzurichten. Bei Verwendung der verwalteten Lösungen (was bei Google's eigener Software 
+sehr wahrscheinlich ist) besteht keine Notwendigkeit, dass sich das Google Meet Team um 
+Updates der Netzwerk-Software, Firewall etc. kümmert.
+
+Threagile schlägt zur Unterbindung dieser Gefahr vor, bekannte Absicherungs-Frameworks 
+(unteranderem von OWASP) zu verwenden, um die cloud-interne Architektur zu überprüfen.
+
 \section{Zusammenfassung}
 
-% [1] https://www.iriusrisk.com/resources-blog/recommended-threat-modeling-tools
+Insgesamt lassen sich aus dem generierten Threagile-Report mit wenig Aufwand wichtige 
+Hinweise zu grundlegenden Problemen der verwendeten Software-Architektur ableiten. Da 
+Threagile gleichzeitig zur internen Dokumentation und zur Gefahrenanalyse verwendet werden 
+kann, bietet die Nutzung eine
 

### 2024-05-14 entwicklung-sicherer-software-systeme/exercise-1/exercise-1.tex
Finish report

@@
 	zusammengefasst und eine Priorisierung mit Handlungsempfehlung daraus abgeleitet. Des 
-	weiteren wird auf die Methodik der Gefahrenanalyse eingegangen und von alternativen 
+	Weiteren wird auf die Methodik der Gefahrenanalyse eingegangen und von alternativen 
 	Vorgehensweisen abgegrenzt.
@@
 \begin{enumerate}
-	\item Ihre Funktionsweise- und -umfang ist ihm Rahmen einer Kurzanalyse greifbar
-	\item Die Anwendungsarchitektur lässt sich aus vergleichbaren Echtzeit-Meeting Anwendungen ableiten und vereinfachen
+	\item Ihre Funktionsweise und ihr Umfang ist im Rahmen einer Kurzanalyse greifbar
+	\item Die Anwendungsarchitektur lässt sich aus vergleichbaren Echtzeit-Meeting-Anwendungen ableiten und vereinfachen
 	\item Sie hat einen weitreichenden Bekanntheitsgrad
@@
 
-Vereinfacht lässt sich Google Meet in einer Client-Server Architektur darstellen, auch 
+Vereinfacht lässt sich Google Meet in einer Client-Server-Architektur darstellen, auch 
 wenn an dieser Stelle angemerkt werden muss, dass durch die Menge an Nutzern, die Vielzahl 
@@
 Im Rahmen dieser Aufgabe wird von einer monolithischen Anwendungsarchitektur von Google 
-Meet ausgegangen in einem zeitlich angemessenen Rahmen in der Lage zu sein eine 
+Meet ausgegangen, um in einem zeitlich angemessenen Rahmen in der Lage zu sein, eine 
 hypothetische Gefahrenanalyse durchführen zu können. Der Google Meet Server in der 
 beistehenden Grafik übernimmt die Verwaltung der Sitzungen, die Account-Authentifizierung 
-und die Aushandlung der WebRTC-Sitzungen zwischen verschiedenen Nutzer.
+und die Aushandlung der WebRTC-Sitzungen zwischen verschiedenen Nutzern.
 
@@
 
-Um eine automatisierte Gefahrenanalyse durchzuführen stehen mittlerweile (Stand Mai 2024), 
+Um eine automatisierte Gefahrenanalyse durchzuführen, stehen mittlerweile (Stand Mai 2024) 
 eine Vielzahl an verschiedenen Tools zur Verfügung. Nachfolgend sind einige beliebte 
@@
 Da diese Einsendeaufgabe von geringem Umfang ist und keine weiteren Anwendungsfälle für 
-eine ausgereifte Threat-Modelling Software absehbar sind beschränke ich den Vergleich 
-verschiedener Software-Lösungen auf kostenlose bzw. Open-Source Programme.
+eine ausgereifte Threat-Modelling Software absehbar sind, beschränke ich den Vergleich 
+verschiedener Software-Lösungen auf kostenlose bzw. Open-Source-Programme.
 
@@
 MTMT ist das Threat Modeling Tool der Firma Microsoft. Es ist ein weit verbreitetes 
-Werkzeug zur Modellierung von Software und der Automatisierten Analyse dieser Modelle. Da 
-der Anbieter Microsoft ist, besteht leider die Anforderung MTMT auf einem Windows-System 
+Werkzeug zur Modellierung von Software und der automatisierten Analyse dieser Modelle. Da 
+der Anbieter Microsoft ist, besteht leider die Anforderung, MTMT auf einem Windows-System 
 auszuführen.
@@
 Leider musste ich durch die Gegebenheit, kein Windows-System zu betreiben, die Analyse 
-des MTMT hier beenden, da es mir leider nicht möglich war eine ausführbare Version für 
-Unix-Systeme zu finden ohne Virtualisierung zu verwenden.
+des MTMT hier beenden, da es mir leider nicht möglich war, eine ausführbare Version für 
+Unix-Systeme zu finden, ohne Virtualisierung zu verwenden.
 
@@
 
-OWASP ThreatDragon ist eine Software der Organization OWASP (kurz für \textit{Open 
+OWASP ThreatDragon ist eine Software der Organisation OWASP (kurz für \textit{Open 
 Worldwide Application Security Project}). Die Software ist kostenlos und 
-Browser basiert. Sie enthält Funktionen für die Diagramm-Zeichnung und anschließende 
+browserbasiert. Sie enthält Funktionen für die Diagramm-Zeichnung und anschließende 
 Erstellung eines Sicherheits-Reports. Die Arbeitsstände müssen allerdings immer 
-heruntergeladen werden und in lokalen Dateien gespeichert werden – dies ist eine relativ 
+heruntergeladen und in lokalen Dateien gespeichert werden – dies ist eine relativ 
 umständliche Arbeitsweise, da anders als bei nativer Software nicht eine Datei durchgehend 
@@
 
-AWS ThreatComposer ist eine Open Source Software der Firma Amazon. Sie ist ebenfalls 
-Browser basiert, allerdings etwas ausgereifter und nutzerfreundlicher als die OWASP 
-alternative. Dennoch gibt es auch hier einige Unannehmlichkeiten für Nutzer: Die 
-Benutzeroberfläche ist unklar (e.g. Diagramme sind nur zur Dokumentation hinzuzufügen, 
+AWS ThreatComposer ist eine Open-Source-Software der Firma Amazon. Sie ist ebenfalls 
+browserbasiert, allerdings etwas ausgereifter und nutzerfreundlicher als die OWASP-
+Alternative. Dennoch gibt es auch hier einige Unannehmlichkeiten für Nutzer: Die 
+Benutzeroberfläche ist unklar (z.B. Diagramme sind nur zur Dokumentation hinzuzufügen, 
 nicht zur automatisierten Auswertung). Es gibt keine klare Datenverwaltung bzw. keinen 
@@
 
-Threagile ist kostenlose Open-Source Software zur Gefahrenanalyse ohne eine 
+Threagile ist kostenlose Open-Source-Software zur Gefahrenanalyse ohne eine 
 Benutzeroberfläche. Dies ist eine Eigenheit im Vergleich zu den anderen vorgestellten 
-Programmen die hauptsächlich über das Desktop- oder Web-Interface zu bedienen sind.
-Threagile lässt sich über eine .yaml Datei bedienen die einem vorgegebenen Schema 
+Programmen, die hauptsächlich über das Desktop- oder Web-Interface zu bedienen sind. 
+Threagile lässt sich über eine .yaml-Datei bedienen, die einem vorgegebenen Schema 
 entsprechen muss. So lassen sich im Quelltext Komponenten (und ihre Beziehungen), 
 Datenbanken, Verschlüsselungen, Datenflüsse, verwendete Technologien und vieles mehr 
-Dokumentieren.
+dokumentieren.
 
-Der klare Vorteil von Threagile liegt in der Möglichkeit diese YAML-Datei bzw. die 
-analysierte Architektur ohne weiteres in git oder anderen VCS (kurz für \textit{Version 
-Control Systems}) zu versionieren. Bei anderer proprietär Software ist dies zwar 
-grundsätzlich Möglich aber da VCS in der Regel auf Klartext-Dateien ausgelegt sind sind 
-die Möglichkeiten begrenzt bzw. parallel im Team an einer Datei zu arbeiten.
+Der klare Vorteil von Threagile liegt in der Möglichkeit, diese YAML-Datei bzw. die 
+analysierte Architektur ohne Weiteres in git oder anderen VCS (kurz für \textit{Version 
+Control Systems}) zu versionieren. Bei anderer proprietärer Software ist dies zwar 
+grundsätzlich möglich, aber da VCS in der Regel auf Klartext-Dateien ausgelegt sind, sind 
+die Möglichkeiten begrenzt, parallel im Team an einer Datei zu arbeiten.
 
-Des Weiteren ist es einfach mittels Threagile eine Gefahrenanalyse auf der YAML-Datei 
+Des Weiteren ist es einfach, mittels Threagile eine Gefahrenanalyse auf der YAML-Datei 
 auszuführen und der generierte Report hat eine hohe Qualität (optisch, strukturell und 
@@
 der verschiedenen Softwarelösungen für die Nutzung von Threagile entschieden. Threagile 
-ist einfach zu bedienen, reproduzierbar (mittels Docker), eine nützliche Resource für 
+ist einfach zu bedienen, reproduzierbar (mittels Docker), und eine nützliche Ressource für 
 zukünftige berufliche Situationen.
@@
 
-Der automatisierte Report von Threagile hat basierend der oben dargestellten, stark 
-vereinfachten, Anwendungsarchitektur hat insgesamt 21 mögliche Risiken und in 12 
-verschiedenen Risiko-Klassen identifizieren können.
+Der automatisierte Report von Threagile hat basierend auf der oben dargestellten, stark 
+vereinfachten Anwendungsarchitektur insgesamt 21 mögliche Risiken und 12 
+verschiedene Risikoklassen identifizieren können.
 
-Folgende Risiko-Klassen wurden mit der Stufe \textit{Elevated – Wahrscheinlicher Angriff mit hohem Schaden} markiert:
+Folgende Risikoklassen wurden mit der Stufe \textit{Elevated – Wahrscheinlicher
+Angriff mit hohem Schaden} markiert:
 
@@
 
-Alle anderen entdeckten Risiko-Klassen wurden mit der Stufe \textit{Medium – Unwahrscheinlicher Angriff mit hohem Schaden} markiert:
+Alle anderen entdeckten Risikoklassen wurden mit der Stufe \textit{Medium – Unwahrscheinlicher
+Angriff mit hohem Schaden} markiert:
 
@@
 Zusammenfassend lässt sich über die entdeckten Risiken sagen, dass diese ein konkretes,
-greifbares Risiko darstellen und Threagile wenige ``False-Positives'' entdeckt hat (bspw. 
-Fehler die lediglich aufgrund einer Ungenauigkeit der Diagramm-Funktion markiert werden). 
-Threagile lässt eine sehr detaillierte Dokumentation von Verschlüsselungs-Methoden, 
+greifbares Risiko darstellen und Threagile wenige ``False Positives'' entdeckt hat (bspw. 
+Fehler, die lediglich aufgrund einer Ungenauigkeit der Diagramm-Funktion markiert werden). 
+Threagile lässt eine sehr detaillierte Dokumentation von Verschlüsselungsmethoden, 
 übertragenen Datensätzen und verwendeten Datenformaten etc. zu. Die Liste der einzelnen 
-Risiken finden sie im Threagile-Report (Anhang 1).
+Risiken finden Sie im Threagile-Report (Anhang 1).
 
@@
 
-Die logische Schlussfolgerung der oben eingestuften Risiko-Klassen ist die priorisierung 
+Die logische Schlussfolgerung der oben eingestuften Risikoklassen ist die Priorisierung 
 der Gefahren nach absolutem Risiko:
 
-$$Risiko = Schaden * Wahrscheinlichkeit$$
+$$Risiko = Schaden \times Wahrscheinlichkeit$$
 
@@
 
-Da Google Meet text-basierte Nutzereingaben als eine der Kernfunktionen (Chat) zulässt, 
+Da Google Meet textbasierte Nutzereingaben als eine der Kernfunktionen (Chat) zulässt, 
 stellt Cross-Site-Scripting ein erhebliches Risiko dar. Allerdings muss an dieser Stelle 
-angemerkt werden, dass die Empfänger einer bösartigen lediglich die Nutzer einer Sitzung 
-sind (meist, ein bis zweistellige Nutzerzahlen) – wohingegen auf Plattformen wie Twitter 
-und YouTube etc. Nutzer öffentlich Beiträge posten können was die Anzahl an betroffenen 
-Nutzern drastisch erhöht.
+angemerkt werden, dass die Empfänger eines bösartigen Angriffs lediglich die Nutzer einer Sitzung 
+sind (meist ein- bis zweistellige Nutzerzahlen) – wohingegen auf Plattformen wie Twitter 
+und YouTube etc. Nutzer öffentlich Beiträge posten können, was die Anzahl der betroffenen 
+Nutzer drastisch erhöht.
 
@@
 Die fehlende Absicherung der Cloud gegen unbefugten Zugriff (bspw. via SSH, 
-Netzwerkschwachstellen etc) stellt neben XSS die größte Gefahr dar, da diese mit einer 
+Netzwerkschwachstellen etc.) stellt neben XSS die größte Gefahr dar, da diese mit einer 
 möglichen ``Escalation of Privilege'' einhergehen könnte und somit ein Angreifer nach der 
-infiltration einer der Cloud-Internen-Komponenten wohlmöglich weitere Komponenten 
+Infiltration einer der cloud-internen Komponenten womöglich weitere Komponenten 
 infizieren könnte. Mögliche Lösungsstrategien zur Absicherung der Cloud bestehen darin,
@@
 Hosting) der GCP in Anspruch zu nehmen, anstatt mittels VM eine manuelle Infrastruktur 
-einzurichten. Bei Verwendung der verwalteten Lösungen (was bei Google's eigener Software 
+einzurichten. Bei Verwendung der verwalteten Lösungen (was bei Googles eigener Software 
 sehr wahrscheinlich ist) besteht keine Notwendigkeit, dass sich das Google Meet Team um 
-Updates der Netzwerk-Software, Firewall etc. kümmert.
+Updates der Netzwerkssoftware, Firewall etc. kümmert.
 
 Threagile schlägt zur Unterbindung dieser Gefahr vor, bekannte Absicherungs-Frameworks 
-(unteranderem von OWASP) zu verwenden, um die cloud-interne Architektur zu überprüfen.
+(unter anderem von OWASP) zu verwenden, um die cloud-interne Architektur zu überprüfen.
 
@@
 
-Insgesamt lassen sich aus dem generierten Threagile-Report mit wenig Aufwand wichtige 
-Hinweise zu grundlegenden Problemen der verwendeten Software-Architektur ableiten. Da 
-Threagile gleichzeitig zur internen Dokumentation und zur Gefahrenanalyse verwendet werden 
-kann, bietet die Nutzung eine
+Insgesamt lassen sich aus dem generierten Threagile-Report mit angemessenem Aufwand wichtige 
+Hinweise zu grundlegenden Problemen der verwendeten Software-Architektur ableiten. Anhand der
+von Threagile vorgegebenen Risikoklassen und der Risikoformel lässt sich eine grobe Priorisierung
+für die Schwachstellen ableiten. Für eine detaillierte Priorisierung wäre jedoch tiefgreifendes Wissen
+über das Geschäftsmodell, die Nutzer, die Umsatzrelevanz einzelner Komponenten, Reparaturaufwand einer 
+Schwachstelle etc. notwendig.
 

### 2024-10-27 biometrie/exercises.tex
Biometrie

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\newcommand{\annotation}[1]{
\begin{quote}
	\begin{textit}
		{#1}
	\end{textit}
\end{quote}
}

\renewcommand*{\thesection}{Kapitel \arabic{section}}
\renewcommand*{\thesubsection}{\alph{subsection})}

\begin{document}

\begin{titlepage}
	\begin{center}
		\begin{Large}
			Technische Hochschule Brandenburg \\[1em]
		\end{Large}

		IT Sicherheit \\
		Informatik und Medien \\
		Biometrie – Dr.\ Tobias Scheidat
	\end{center}

	\vfill

	\begin{center}
		\Large{Übungsaufgaben}\\[0.5em]
		\large{Wintersemester 2024}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{center}
		Mara Schulke \\ Matr-Nr. 20215853
	\end{center}
\end{titlepage}

\tableofcontents

\listoffigures

\newpage

\section{}

\subsection{Erl\"autern Sie den Begriff Benutzerauthentifizierung}

Benutzerauthentifizierung beschreibt den formalen Prozess der Verifikation der Nutzeridentität innerhalb 
eines Systems. Dies kann anhand einer der drei Authentifizierungsschemata stattfinden: Wissen, Besitz oder 
Biometrie.

\subsection{Definieren Sie den Begriff biometrische (Benutzer-)Erkennung}
\annotation{Grenzen Sie zu der vorherigen Aufgabe ab.}
Die biometrische Benutzer-Erkennung ist streng genommen ein Teilbereich der Benutzerauthentifizierung. Wie 
oben genannt lässt sich die Benutzerauthentifizierung mittels Wissen, Besitz oder Biometrie durchführen.

Im Gegensatz zu den Schemata Wissen und Besitz fokussiert sich die biometrie-gestützte Erkennung/
Authentifizierung nicht auf Nachweismöglichkeiten die extrinsisch mit einem Individuum verknüpft sind 
(Passwort, Schlüssel, ..) sondern auf intrinsische Nachweismöglichkeiten wie die Physiologie oder das
Verhalten eines Individuums.

Außerhalb der Authentifizierung befasst sich die biometrische Erkennung mit der statistischen Zuordnung 
von Eingabedaten zu einem hinterlegten biometrischen Fingerprint. Dies umfasst z.B. die Auswertung von 
Sprachaufnahmen um regionale Sprachmuster zu erkennen.

\subsection{Erläutern Sie die 11 Merkmale der Bertillonawge}
\annotation{{\"U}berlegen Sie: wie konnte mit den damaligen Mitteln eine Identifikation durchgeführt werden?}
Die 11 Merkmale mit denen damals die Bertillonage durchgeführt wurde lauten: Körpergröße, Spannweite der 
Arme, Sitzhöhe, Kopflänge, Kopfbreite, Länge und Breite des rechten Ohrs, Länge des linken Fußes sowie 
Längen des linken Mittelfingers, des linken kleinen Fingers und des linken Unterarms (Auszug aus dem 
Skript 1.4).

Die Vorgehensweise der Bertillonage ist der der in der DNA-basierten Identifikation recht ähnlich (zumindest 
oberflächlich). In beiden Fällen werden bestimmte Merkmale eines Individuums extrahiert (ie. Marker).
Je mehr Marker zwischen zwei Datensätzen übereinstimmen, desto wahrscheinlicher ist es, dass es sich um 
das gleiche Individuum handelt. Der Abgleich eines einzelnen Individuums ist nicht rechenaufwendig, daher 
war dieses Verfahren bereits im 19. Jahrhundert möglich.

\subsection{Was sind biometrische Charakteristika und welche zwei grundsätzlichen Kategorien gibt es hier?}
\annotation{Klassifizieren Sie biometrische Merkmale in die zwei grundsätzlichen Kategorien und benennen Sie jeweils mindestens vier Beispiele für jede?}	
Es gibt die grundsätzliche Unterteilung zwischen Online- und Offline- Merkmalen. Diese unterscheiden sich
primär hinsichtlich ihres Aufzeichnungszeitpunktes: Online Merkmale können nur während einer Handlung 
aufgenommen werden (z.B. während eine Notiz geschrieben wird) wohingegen Offline-Daten auch im Nachhinein
verfügbar sind (z.B. der beschriebene Notizzettel).

Beispiele für Online-Merkmale sind z.B. Sprachaufnahmen, ein Video von einer handschriftlichen Notiz, das 
Tippverhalten an dem Computer, eine Aufnahme vom Geh-Verhalten. Beispiele für Offline-Merkmale sind z.B. 
Fußstapfen, ein Foto von einer handschriftlichen Notiz, Fingerabdrücke und Körpermaße.

\subsection{Erklären Sie: was verstehen wir unter dem Begriff Lebenderkennung?}

Die Lebenderkennung fokussiert sich primär auf verhaltensbasierte Merkmale, da diese im Gegensatz zu
physiologischen Merkmalen inhärent nur von lebenden Subjekten entnehmen lassen. Diese Merkmale bieten 
somit eine (weitgehende) Sicherheit gegen Spoofing-Attacken.

\end{document}

### 2024-11-24 biometrie/exercises.tex
Biometrie Chapter 3+4

@@
 
+\newcommand{\figuresource}[1]{
+	\begin{center}Quelle: {#1}\end{center}
+}
+
 \renewcommand*{\thesection}{Kapitel \arabic{section}}
 \renewcommand*{\thesubsection}{\alph{subsection})}
+\renewcommand*{\thesubsubsection}{\arabic{subsubsection})}
 
@@
 
-\subsection{Erläutern Sie die 11 Merkmale der Bertillonawge}
+\subsection{Erläutern Sie die 11 Merkmale der Bertillonage}
 \annotation{{\"U}berlegen Sie: wie konnte mit den damaligen Mitteln eine Identifikation durchgeführt werden?}
+
 Die 11 Merkmale mit denen damals die Bertillonage durchgeführt wurde lauten: Körpergröße, Spannweite der 
@@
 \annotation{Klassifizieren Sie biometrische Merkmale in die zwei grundsätzlichen Kategorien und benennen Sie jeweils mindestens vier Beispiele für jede?}
+
 Es gibt die grundsätzliche Unterteilung zwischen Online- und Offline- Merkmalen. Diese unterscheiden sich
@@
 
+\newpage
+
+\section{}
+
+\subsection{Finden und beschreiben Sie für jeden der 6 Sicherheitsaspekte ein Beispiel, wie diese speziell für biometrische Systeme relevant sein k\"onnen!
+}
+
+\subsubsection*{Vertraulichkeit}
+
+Ein biometrisches System, das für die Authentifizierung eingesetzt wird, sollte ähnliche 
+Sicherheitscharakteristika haben wie eine passwortbasierte Authentifizierung. D. h. es sollte vergleichbar
+schwer für einen Angreifer sein, ein biometrisches Authentifizierungsverfahren zu umgehen, wie für  
+herkömmliche, da sonst die Vertraulichkeit der Daten durch den Einsatz des biometrischen Systems
+gefährdet ist.
+
+\subsubsection*{Integrität}
+
+Die Verifikationsdaten eines biometrischen Systems sollten sich ausschließlich nach der 
+Sicherstellung der Identität des Benutzers ändern lassen. Idealerweise mittels 
+Multifaktorauthentifizierung.
+
+\subsubsection*{Verfügbarkeit}
+
+Um die Verfügbarkeit eines Systems nicht durch den Einsatz von Biometrie gefährden sollte das biometrische 
+System ausgiebige Tests im Hinblick auf dessen Zuverlässigkeit bestehen. Wenn das Gesamtsystem nicht mehr
+nutzbar ist, da das biometrische Authentifizierungssystem versagt, wird die allgemeine Verfügbarkeit 
+beeinträchtigt.
+
+\subsubsection*{Authentizität}
+
+Die Verifikationsdaten eines biometrischen Systems sollten idealerweise mit einem, bereits bekannten, dem 
+Individuum zugeordneten Schlüsselpaar signiert werden um diese zweifelsfrei an die Identität des Nutzers 
+zu binden.
+
+\subsubsection*{Verbindlichkeit}
+
+Alle Änderungen an den hinterlegten Verifikationsdaten, sowie jegliche Authentifizierungsvorgänge bzw. 
+Versuche sollten dokumentiert werden, damit die Authentifizierung eines Nutzers zurückverfolgt werden 
+kann.
+
+\subsubsection*{Privatsphäre}
+
+Ein biometrisches System sollte niemals rohe biometrische Daten über ein Individuum speichern, vielmehr 
+sollte es entweder eine verschlüsselte Version speichern oder idealerweise einen Fingerprint, der sich aus 
+den Eingabedaten konstruieren lässt, sich aber nicht ohne weiteres zurück in eindeutige biometrische 
+Eigenschaften übersetzen lässt.
+
+\subsection{Geben Sie 3 Beispiele (mit einer kurzen Erläuterung), was bei Security-by-Design für biometrische Systeme berücksichtigt werden sollte!}
+
+\subsubsection*{Nicht-Zurückverfolgbarkeit}
+
+Aus einem biometrischen System sollten nie genug Daten gewonnen werden können, um den Nutzer eindeutig zu 
+identifizieren (ohne das system Selbst zu Verwenden). Ie. es sollten keine Rückschlüsse auf physische 
+Eigenschaften (Hautfarbe, Größe, Geschlecht, etc.) aus den im System hinterlegten Daten möglich sein.
+
+\subsubsection*{Need-to-know}
+
+Lediglich biometrische Merkmale aufzeichnen, die einen konkreten Verwendungszweck für die Aufgabe des 
+biometrischen Systems haben. Mehr Daten aufzuzeichnen als notwendig birgt das Risiko der Zweckentfremdung.
+
+\subsubsection*{Geringe False-Accept-Rate}
+
+Gerade so viele biometrische Merkmale sammeln, dass die Fehlerquote des biometrischen Systems für falsche 
+Authentifizierungen vernachlässigbar gering ausfällt.
+
+\subsection{Geben Sie 3 Beispiele mit einer kurzen Erläuterung, was bei Ethik-by-Design für biometrische Systeme berücksichtigt werden sollte!}
+
+\subsubsection*{Randomisierte Tests}
+
+Ein biometrisches System sollte mit heterogenen Nutzern getestet werden um sicherzustellen, dass 
+marginalisierte Gruppen nicht benachteiligt werden. Beispielsweise sollte sichergestellt werden, dass eine 
+Gesichtserkennung zuverlässig über ethnische Gruppen hinweg funktioniert, anstatt nur mit einer Ethnie zu 
+testen.
+
+\subsubsection*{Verhinderung von Zweckentfremdung}
+
+Ein System zur biometrischen Authententifizierung ist in den meisten Fällen wahrscheinlich ethisch 
+unbedenklich, wohingegen ein System zur Identifizierung anhand von biometrischen Merkmalen kritisch 
+betrachtet werden kann. Daher ist es wichtig, bei der Entwicklung eines biometrischen Systems 
+sicherzustellen, dass dieses bzw. die Daten, die das System verarbeitet, nicht entfremdet zur 
+Identifikation von Individuen benutzt werden können.
+
+\subsubsection*{Datenhoheit liegt beim Nutzer}
+
+Dem Nutzer muss es freistehen, das System ohne biometrische Authentifizierung zu verwenden damit dieser 
+die Datenhoheit behält. Ie. ein Nutzer sollte nicht gezwungen werden seine biometrischen Merkmale 
+aufzuzeichnen, ohne dass eine Nutzung der biometrischen Authentifizierung beabsichtigt ist.
+
+\newpage
+
+\section{}
+
+\subsection{Aus welchen Phasen besteht das Prozess-/ Mustererkennungsmodell zur biometrischen Erkennung: benennen \& erklären Sie!}
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{assets/pipeline-id}
+	\centering
+	\caption{Biometrie-Pipeline}
+	\figuresource{Biometrie Skript}
+\end{figure}
+
+Eine einfache biometrische Pipeline zur Erkennung von Mustern / zur Authentifizierung besteht aus den 
+folgenden vier Phasen:
+
+\subsubsection*{Datenerfassung}
+
+Die erste Phase des Modells beschäftigt sich mit der Erfassung der Daten aus der physischen Welt mittels 
+eines Sensors. Dies kann ein Fingerabdruck-Sensor sein, sowie ein Mikrofon oder eine Kamera.
+
+\subsubsection*{Vorverarbeitung}
+
+Nachfolgend müssen die Rohdaten vor-verarbeitet werden. Dies umfasst die Filterung, Aufbereitung, 
+Korrektur oder Konvertierung der aufgezeichneten Daten um eine einfachere Verarbeitung in den folgenden 
+Phasen zu ermöglichen.
+
+\subsubsection*{Merkmalsextraktion}
+
+Anschließend müssen aus den vor-verarbeiteten Rohdaten Merkmale bzw. Merkmalsvektoren extrahiert werden um
+verschiedene Eingabedaten im gleichen Raum miteinander zu vergleichen. Die Extraktion ermöglicht einen 
+sehr effizienten Vergleich, da z.B. im Falle von Kameradaten nicht Bilddaten miteinander abgeglichen 
+werden, sondern die Eigenschaften die aus den Bilddaten gewonnen wurden. Somit kann eine Suche in der 
+Referenzdatenbank schnell vonstatten gehen.
+
+\subsubsection*{Vergleich \& Klassifikation}
+
+Die letzte Phase der Pipeline ist der Vergleich und die Klassifizierung des Merkmalsvektors aus dem 
+vorherigen Schritt mit denen in der Referenzdatenbank. Hier werden die verschiedenen Vektoren mittels 
+mathematischer Vergleichsoperationen/modelle korreliert und bei gegebener Nähe wird eine positive 
+Entscheidung zurückgegeben.
+
+Die genaue Umsetzung obliegt dem System, so kann jedes System unterschiedliche Schwellwerte für eine 
+Entscheidung heranziehen und/oder Entscheidungen aus mehreren verschiedenen Merkmalen kombinieren.
+
+\subsection{Skizzieren Sie den Prozess des biometrischen Enrollments, wie im Kurs eingeführt!}
+
+Der Prozess des biometrischen Enrollments ist dem der Authentifizierung sehr ähnlich; da lediglich die 
+letzte Phase ``Vergleich \& Klassifikation'' entfällt. Nach der dritten Phase (Merkmalsextraktion) wird
+der/die Merkmalsvektor(en) in einer Referenzdatenbank hinterlegt gegen die dann die Authentifizierung 
+stattfinden kann.
+
+\subsection{Skizzieren Sie den Prozess des biometrischen Authentifikation, wie im Kurs eingeführt!}
+
+Der Prozess der Authentifikation umfasst das \textit{einmalige} abgleichen $n$ Merkmerkmalsvektoren 
+gegenüber der, in der Referenzdatenbank hinterlegten Vektoren, für einen Nutzer. 
+Die Authentifikation grenzt sich gegenüber der Identifikation dahingehend ab, dass keine Suche innerhalb 
+der Referenzdaten stattfindet.
+
+\subsection{Erklären Sie folgende Begriffe}
+
+\subsubsection{Variabilität, Intra-Personen und Inter-Personen Variabilität}
+
+\textbf{Variabilität} im allgemeinen bezieht sich auf die (wie unter Aufgabe \textit{3e)} erklärt) auf
+die Bandbreite der Messungen von biometrischen Daten. Es gibt zwei unterschiedliche Bandbreiten die 
+normalerweise betrachtet werden:
+
+Die \textbf{Intra-Personen Variablität} bezieht sich auf die Schwankungen der Messungen von biometrischen
+Daten der gleichen Person. Beispielsweise unterliegen biometrische Eigenschaften natürlichen Schwankungen,
+wie z.B. offensichtlich erklärbar (ie. durch Nahrungszufuhr) bei dem Gewicht einer Person. Allerdings
+bezieht sich die Variabilität nicht nur auf die rein physische Schwankung sondern betrachtet auch die
+Fehlerraten bei der Messung.
+
+Die \textbf{Inter-Personen Variablität} bezieht sich dementsprechend auf die Schwankungen der Messwerte 
+über alle Personen / Benutzer des biometrischen Systems hinweg. Beispielsweise ließe sich hier die 
+Gesamtverteilung der Körpergröße über alle Menschen hinweg als \textbf{Inter-Personen Variablität} der 
+Körpergröße beschreiben.
+
+\subsubsection{Benennen Sie mindestens drei Ursachen für Variabilität}
+
+Variabilität kann – wie oben genannt – verschiedene Ursachen haben:
+
+\begin{enumerate}
+	\item Biologische / Physische Schwankungen – Beispielsweise ändert sich das Gewicht über den Tag hinweg.
+	\item Biologische Unterschiede – So unterliegen biometrische Parameter immer dem Einfluss der biologischen Gegebenheiten (z.B. Geschlecht, Ethnie etc.)
+	\item Mess-Ungenauigkeit – Leichte Änderungen der Werte über Messungen hinweg können auch einer Ungenauigkeit der Sensorik zugrunde liegen.
+\end{enumerate}
+
+\subsection{Welche Eigenschaften sind an biometrische Merkmale zu stellen, benennen \& erklären Sie!}
+
+\subsubsection*{Variabilität}
+
+Die Variabilität beschreibt die Eigenschaft eines Merkmals über viele Messungen hinweg. Sie kann in den 
+Dimensionen Intra-Person und Inter-Klasse betrachtet werden.
+
+Ein gutes biometrisches Merkmal hat die Eigenschaft einer geringen Intra-Personen-Variabilität – also 
+einer möglichst konstanten Messbarkeit innerhalb der gleichen Person \textbf{bei gleichzeitiger} möglichst
+großer Inter-Class-Variabilität – also einer möglichst breitgestreuten Verteilung des Merkmals innerhalb 
+der Klasse an Subjekten.
+
+\subsubsection*{Trennschärfe}
+
+Die Trennschärfe ist eingeführt als möglichst großer Grad zwischen der hohen Intra-Personen-Variabilität 
+und Inter-Klassen-Variabilität. Je klarer ein Merkmal zu messen ist und je breit gestreuter die Messungen 
+über alle Subjekte hinweg sind, desto Aussagekräftiger ist die Messung über die Identität.
+
+\subsubsection*{Erfassbarkeit}
+
+Die Erfassbarkeit eines Merkmals beschreibt die Betrachtung des realistisch-möglichen Einsatzes des 
+Merkmals in einem biometrischen System mit verfügbaren Sensoren. Beispielsweise lassen sich Bild-Daten 
+eines Subjektes leicht erfassen wohingegen DNA-Daten sehr aufwändig zu beschaffen sind.
+
+\subsubsection*{Leistung}
+
+Leistung beschreibt den zeitlichen Aspekt der Erfassung bzw. Verarbeitung. D.h. wenn ein Merkmal bspw. 
+extreme Mengen an Daten generiert, könnte dies u.U. zu Performance-Problemen im biometrischen System 
+führen.
+
+\subsubsection*{Privatsphäre}
+
+Die letzte Eigenschaft von Merkmalen ist dahingehend wichtig, dass bestimmte Datensätze Aussagekräftiger 
+über Anwendungszweck-Fremde Daten (wie z.B. Herkunft sind, als andere) und somit a) Vorsicht im Umgang und 
+bei der Speicherung mit solchen Merkmalen geboten ist und b) eine ethische Betrachtung stattfinden sollte, 
+welche Merkmale für ein biometrisches System verwendet werden.
+
+\newpage
+
+\subsection{Erklären Sie beispielhaft anhand eines Häufigkeitshistogramms die Merkmalsverteilungen „Intra-Class“, „Inter-Class“}
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{assets/histogramm}
+	\centering
+	\caption{Histogramm: Größe}
+	\figuresource{Universität Münster 
+		(\href{https://campus.uni-muenster.de/fileadmin/einrichtung/ibkf/lehre/skripte/biomathe/bio/eda/help/eda15.html}{campus.uni-münster.de})
+	}
+\end{figure}
+
+Wie in der Abbildung 2) zu erkennen reichen die Messwerte in der oben angeführten Beispielverteilung
+von ca. 145cm bis zu ca. 195cm. Diese Spanne beschreibt die Inter-Class Variabilität von allen Menschen
+in der aufgezeichneten Gruppe. Die Intra-Class- (oder Intra-Personen-) Variabilität fällt deutlich 
+geringer aus: Die Körpergröße ist vergleichsweise (z.B. im direkten Vergleich mit dem Merkmal Gewicht)
+biologisch Stabil (also unterliegt lediglich kleinen Schwankungen) und der größte Anteil hier
+dürfte durch Fehlerraten während der Messung entstehen.
+
+\subsection{Überlegen Sie: bislang bezog sich die „Inter-Class“ Verteilung ja nur auf die Merkmale „anderer“ Personen als die der authentischen, mit ihren eigenen Merkmalen.
+Wie würde sich die Situation für gezielte Fälschungen im Diagramm darstellen, welche Veränderungen können Sie hier vorhersehen?
+Beschreiben oder skizzieren Sie im Diagramm!}
+
+Dies kommt auf die Abweichung des zu fälschenden Messwertes von dem Medianwert aller Messungen an.
+Somit ließe sich ein Angriff relativ eindeutig (und bereits nach wenigen böswilligen 
+Authentifizierungsversuchen) erkennen, wenn der gefälschte Messwert mehrere Standardabweichungen vom 
+Median entfernt liegt. (Da es somit zu einer verhältnismäßig starken Änderung des Medianwertes käme). 
+Visuell wäre ein Angriff daran zu erkennen, wenn sich der Balken (bspw. bei 145cm) innerhalb von kurzer 
+Zeit, stark erhöht und somit eine signifikante Auswirkung auf die berechnete Standardabweichung hätte. Bei 
+einer großen Menge (im Verhältnis zu den gesamten Messdaten) an Fälschungen einer Messung nahe am 
+Medianwert würde der mittlere Balken (über der blauen Linie) drastisch Ansteigen. Ebenfalls hier wäre
+dies durch die Beobachtung der Standardabweichung zu erkennen.
+
+\newpage
+
+\section{}
+
+\subsection{Definieren Sie die folgenden biometrischen Fehlerraten}
+
+\subsubsection{False Non-Match Rate (FNMR)}
+
+Die False-Non-Match-Rate wird folgendermaßen im Skript (Kapitel 4.1) eingeführt:\\[0.1em]
+
+\textit{
+``Falschnichtübereinstimmungs-Rate (tlw.
+auch Falsch-Nichterkennungsrate), Verhältnis zwischen der Anzahl der nicht bestätigten authentischen Verifikationen (d.h.
+Verifizierungsdaten und Enrolmentdaten von identischen Personen werden fälschlicher Weise negativ verifiziert) und der gesamten Anzahl von Verifikationstests.''
+}
+
+\subsubsection{False Match Rate (FMR)}
+
+Die False-Match-Rate wird folgendermaßen im Skript (Kapitel 4.1) eingeführt:\\[0.1em]
+
+\textit{
+``Falschübereinstimmungs-Rate (tlw.
+auch Falscherkennungsrate), d.h.
+Verhältnis zwischen der Anzahl der aufgetretenen Falsch-Verifikationen (d.h.
+Verifizierungsdaten und Enrolmentdaten von unterscheidlichen Personen werden fälschlicher Weise positiv verifiziert) und der gesamten Anzahl von Verifikationstests''
+}
+
+\subsubsection{Equal-Error Rate (EER)}
+
+Die Equal-Error-Rate wird folgendermaßen im Skript (Kapitel 4.1) eingeführt:\\[0.1em]
+
+\textit{
+``Gleichfehlerrate, Wert von FMR und FNMR an dem Arbeitspunkt, in welchem sich der gleiche Fehlerwert einstellt. Kann z.B. anhand des Fehlerratendiagramms ermittelt werden.''
+}
+
+\subsection{Skizzieren Sie die vorgenannten Fehlerraten als Diagramm \"uber einen Schwellwert basierte biometrische Verifikation}
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{assets/fnmr-fmr-eer}
+	\centering
+	\caption{FNMR vs FMR vs EER}
+	\figuresource{A Comparison Framework for Fingerprint Recognition Methods
+		(\href{https://www.researchgate.net/figure/The-Relationship-between-FNMR-FMR-and-ERR_fig3_259558386}{researchgate.net})
+	}
+\end{figure}
+
+Hier Aufgabe 1
+
+\subsection{Diagramm Zusatzaufgabe}
+
+\annotation{
+    Betrachten Sie nun nochmal das Diagramm, welches Sie für Teil b) erstellt haben.
+    Nehmen Sie nun an, dass der hier eingezeichnete Graph für die FMR ausschließlich
+    auf Basis anderer legitimer Nutzer („nicht-Imposter“) erstellt wurde.
+}
+
+\annotation{
+    Wie würde sich im Vergleich dazu ein zweiter FMR-Graph aussehen, der die
+    False-Matches von gezielten Angriffen (Impostern) wiedergibt, unter der Annahme,
+    dass diese bezüglich der Merkmale der angegriffenen Personen diesen ähnlicher
+    sind? Zeichnen Sie einen exemplarischen zweiten Graphen ein und erläutern Sie!
+}
+
+\annotation{
+    Hinweis: überlegen Sie ggf. auch nochmal vor dem Hintergrund ihrer Lösung zu Aufgabe 3g!
+}
+
+\begin{figure}[ht]
+	\includegraphics[width=0.5\textwidth]{assets/fnmr-fmr-eer-imposter}
+	\centering
+	\caption{FNMR vs FMR vs EER – Imposter}
+	\figuresource{Facia AI - Knowledgebase
+		(\href{https://facia.ai/knowledgebase/knowing-the-equal-error-rate-eer-in-biometrics/}{facia.ai})
+	}
+\end{figure}
+
+Hier aufgabe 2e
+
+\newpage
+
+\subsection{Statistische Signifikanz nach Doddington}
+
+\annotation{Erklären Sie folgende Faustregeln:}
+
+\subsubsection{Doddington’s Rule of 30}
+
+Die Doddington's Rule of 30 wird folgendermaßen im Skript (Kapitel 4.2) eingeführt:\\[0.1em]
+
+\textit{``Man teste, bis man insgesamt 30 Fehler erhält. Dann kann man mit 90\%iger Sicherheit annehmen, 
+dass die beobachtete Fehlerwahrscheinlichkeit (z.B. FMR) innerhalb von +/- 30\% des wahren
+Wertes liegt''}\\[0.1em]
+
+Doddington's Rule of 30 bezieht sich auf die statistische Verteilung der tatsächlichen Fehlerrate durch 
+Stichprobentests. Wie im Skript bezogen ist das Testen von biometrischen System nicht trivial auf Grund 
+der physikalischen Abhängigkeiten der Eingabedaten. Somit lässt sich mittels Doddington's Rule of 30 auf
+der Basis von Stichprobentests eine realtiv genaue (ie. laut der Regel ca. 90\%) Vorhersage über die FMR 
+treffen.
+
+\subsubsection{Doddington’s Rule of 3}
+
+Die Doddington's Rule of 3 wird folgendermaßen im Skript (Kapitel 4.2) eingeführt:\\[0.1em]
+
+\textit{``Wenn keine Fehler in N unabhängigen Tests auftreten, kann man mit
+95\%iger Sicherheit davon ausgehen, dass die Fehlerwahrscheinlichkeit <3/N ist''}\\[0.1em]
+
+Mittels Doddington's Rule of 3 lässt sich die FMR eines biometrischen Systems ableiten, die keine Fehler
+in Stichprobentests erzielt haben. Dies ist sehr nützlich, da akkurate biometrische Tests Zeit- und 
+Kosten-intensiv sind.
+
+\subsubsection{Wie viele fehlerfreie Versuche sind nach Doddington’s Rule of 3 notwendig, um für ein biometrisches Iris-Erkennungssystem, welches während des Tests keine False-Matches generiert, mit einer 95\% Konfidenz abzuschätzen, dass es eine FMR von maximal 0,06\% hat?
+}
+
+\subsubsection{Erläutern die folgende Aussage aus dem Lernmodul, finden Sie ggf.
+eine Beispiele, welche die Aussage begründen: „Daraus folgt: die Beobachtung von 3 Fehler in 100 Versuchen entspricht nicht automatisch der Einschätzung, dass jede der 100 Personen mit einer Fehlerquote von jeweils exakt 3\% hat!“
+}
+
 \end{document}

### 2024-11-24 biometrie/exercises.tex
Biometrie Final Chapter 3+4

@@
 \usepackage{fancyhdr}
+\usepackage{amsmath}
 \usepackage{xcolor}
@@
     \begin{quote}
-    	\begin{textit}
-			{#1}
-	    \end{textit}
+    	\begin{textit}{#1}\end{textit}
     \end{quote}
@@
 
+\newpage
+
 \subsection{Skizzieren Sie die vorgenannten Fehlerraten als Diagramm \"uber einen Schwellwert basierte biometrische Verifikation}
@@
 
-Hier Aufgabe 1
+Die Equal-Error-Rate beschreibt den Schnittpunkt der beiden Fehlerraten (Falscherkennungs-Rate (FMR) und Falschablehnungs-
+Rate (FNMR)). Die Schwellwert-Linie für Authentifizierungs-Entscheidungen kann nun von den Systemherstellern selbst
+festgelegt werden: typischerweise sollte diese in der unmittelbaren Nähe zum EER-Punkt liegen.
+
+Wird die Schwellwert-Linie (in der Abb. 3) rechts von der EER angesetzt wird das System zwar insgesamt sicherer (ie. die
+Falscherkennungen nehmen ab), aber gleichermaßen benutzerunfreundlicher, da authentische Nutzer (häufiger) 
+fälschlicherweise Abgelehnt werden und u.U. mehrere Authentifizierungsversuche vornehmen müssen um erfolgreich
+erkannt zu werden.
+
+Wird die Schwellwert-Linie (in der Abb. 3) links von der EER angesetzt, kommt es zu einem permissiven System, in dem
+Nutzer im Zweifel erfolgreich authentifiziert werden. Somit wird das Gesamtsystem unsicherer, da ein Angriff hier deutlich 
+einfacher ist.
+
+Die Entscheidung wo die Schwellwert-Linie einzuzeichnen ist, ist davon abhängig zu machen in welchem Kontext das System
+verwendet werden soll. Ist das biometrische System nur einer von vielen Faktoren zur Authentifizierung, oder wird in einem
+zeitkritischen oder lebensbedrohlichen Szenario verwendet, kann es entgegen der Sicherheit des Systems dennoch sinnvoll 
+sein, ein permissives biometrisches System zu verwenden, da die Folgen einer falschen Ablehnung (beispielsweise während 
+einer Not-Operation) dramatisch sein könnten. Wird das System allerdings weder in einem zeitkritischem noch anderweitig
+bedrohlichem Szenario verwendet, sollte in der Regel der Fokus auf die Sicherheit des Systems gelegt werden, was für 
+einen, tendenziell, restriktiveren Ansatz des Schwellwertes spricht.
 
@@
     Nehmen Sie nun an, dass der hier eingezeichnete Graph für die FMR ausschließlich
-    auf Basis anderer legitimer Nutzer („nicht-Imposter“) erstellt wurde.
-}
-
-\annotation{
+    auf Basis anderer legitimer Nutzer („nicht-Imposter“) erstellt wurde.\\\\
     Wie würde sich im Vergleich dazu ein zweiter FMR-Graph aussehen, der die
@@
     dass diese bezüglich der Merkmale der angegriffenen Personen diesen ähnlicher
-    sind? Zeichnen Sie einen exemplarischen zweiten Graphen ein und erläutern Sie!
-}
-
-\annotation{
+    sind? Zeichnen Sie einen exemplarischen zweiten Graphen ein und erläutern Sie!\\\\
     Hinweis: überlegen Sie ggf. auch nochmal vor dem Hintergrund ihrer Lösung zu Aufgabe 3g!
@@
 
-Hier aufgabe 2e
+In der hier dargestellten Grafik (Abb. 4) werden die Authentifizierungsversuche (inkl. implizierter Häufigkeit)
+in einer Übersichtsgrafik dargestellt:
+
+Der Grafik liegt im Sinne der Veranschaulichung des Sachverhaltes, die Annahme zu Grunde, dass Imposter-Versuche
+tendenziell einen niedrigeren Match-Score (in Abb. 3 als ``Similarity'' beschrieben) erzielen.
+
+Die Überschneidung von Imposter-Versuchen und Authentischen-Versuchen wird hier als EER eingezeichnet, da dieser,
+genau den, in der vorherigen Aufgabe beschriebenen Scheidepunkt abbildet: Liegt die Schwellwert-Linie links, wird das 
+System permissiv und somit anfällig für Imposter – liegt sie rechts vom EER wird das System restriktiv und somit sicherer,
+aber gleichzeitig benutzerunfreundlicher.
+
+Die Verbindung zwischen dieser Grafik und der Aufgabe 3g) ist ebenfalls offensichtlich erkennbar, da sich der Median-
+Match-Score über alle Versuche Versuche hinweg deutlich nach unten verschiebt (angenommen, die Imposter-Versuche schneiden 
+wie im o.g. Beispiel deutlich schlechter ab).
 
@@
 
+Die Anwendung von Doddington's Rule of 3 ist trivial möglich, da lediglich die Gleichung ${3 \over N} = 0.006$ gelöst 
+werden muss:
+
+\begin{align*}
+         {3 \over N} & = 0.006 \\
+                   3 & = 0.006 * N \\
+     {3 \over 0.006} & = N\\
+     {3 \over 0.006} & = 500\\
+                   N & = 500
+\end{align*}
+
+Somit lässt sich nun mit 95\%er Wahrscheinlichkeit sagen, dass die FMR des Iris-Erkennungssystems < als 0.06\% ist.
+
 \subsubsection{Erläutern die folgende Aussage aus dem Lernmodul, finden Sie ggf.
-eine Beispiele, welche die Aussage begründen: „Daraus folgt: die Beobachtung von 3 Fehler in 100 Versuchen entspricht nicht automatisch der Einschätzung, dass jede der 100 Personen mit einer Fehlerquote von jeweils exakt 3\% hat!“
+einige Beispiele, welche die Aussage begründen: „Daraus folgt: die Beobachtung von 3 Fehler in 100 Versuchen entspricht nicht automatisch der Einschätzung, dass jede der 100 Personen mit einer Fehlerquote von jeweils exakt 3\% hat!“
 }
 
+Die Aussage verweist auf die Tatsache, dass eine beobachtete Fehleranzahl in einer Stichprobe nicht direkt auf die 
+individuelle Fehlerquote jedes einzelnen Testobjektes schließen lässt. Die Gründe hierfür sind:
+
+\begin{enumerate}
+ 	\item Statistische Verteilung von Merkmalen – Unterschiedliche Testobjekte haben unterschiedliche biometrische Merkmale.
+Die Beobachtung von 3 Fehlern in 100 Versuchen lässt lediglich Aussagen über die Gesamtheit des Systems zu, nicht aber über einzelne Testobjekte.
+So kann beispielsweise ein Testobjekt das ein Merkmal mit einer hohen Trennschärfe besitzt zuverlässiger Identifiziert werden (ie.
+hat eine geringere Fehlerquote) als ein Testobjekt das lediglich Merkmale mit geringer Trennschärfe hat.
+ 	\item Stichprobenstreuung – Die Verteilung der getesteten Testobjekte in der Gesamt-Variabilität der Merkmale (Inter-Personen-Variabilität) wird nicht angegeben.
+Es ist gut möglich, dass die Stichproben nicht repräsentativ für die Gesamtverteilung der Merkmale ist und somit der Test in einer anderen Fehlerquote resultiert als dies im tatsächlichen Einsatz der Fall wäre.
+\end{enumerate}
+
+\newpage
+
+\section{}
+
+\subsection{Modalität Fingerabdruck}
+
+\subsubsection{Benennen Sie die drei unterschiedlichen Stufen (engl. Levels) von Fingerabdruckmerkmalen:}
+
+\subsubsection{Erklären Sie die Merkmalsart ``Minutien'' und skizzieren Sie 5 unterschiedliche Beispiele:}
+
+\subsection{Modalität Stimme}
+
+%b1) Benennen Sie die drei grundsätzlichen Ziele der sprachbasierten Biometrie:
+%b2) Erklären Sie stichpunktartig das Konzept der „Mel-Frequency Cepstral Coefficients
+%(MFCC)“:
+%
+
+\subsection{Modalität Gangart}
+
+% (lesen Sie ggf. in der Quelle [NiAd1994] nach, diese ist abrufbar unter
+%http://persci.mit.edu/pub_pdfs/niyogi_XYT.pdf
+%c1) Erklären Sie kurz das Aufzeichnungsszenario für den „kanonischen Gang“ („canonical
+%walk“):
+%c2) Erklären Sie: welche Bildverarbeitungsschritte sind notwendig, um hieraus ein so
+%genanntes XT Diagramm zu generieren:
+%c3) Überlegen, recherchieren und beschreiben Sie 5 Merkmale, welche aus XYT Diagrammen
+%gewonnen werden können:
+
+\subsection{Modalität Tastaturanschlag}
+
+%d1) Benennen Sie Vorteile und Anwendungen dieser biometrischen Modalität:
+%d2) Erläutern Sie kurz das Konzept der „Digraphs“ und überlegen und beschreiben Sie die vier
+%Merkmale, die aus einem Digraph gemessen werden können:
+%d3) Argumentieren Sie mathematisch/logisch, wie das biometrische Verfahren Tastaturanschlag
+%zur Verbesserung der Passwortsicherheit eingesetzt werden kann und stellen Sie klar, welche
+%Annahmen Sie dabei tätigen:
+
+\subsection{Modalität Handschrift}
+
+%e1) Benennen Sie die beiden grundlegenden Kategorien von Merkmalen handschriftbasierter
+%Biometrie:
+%e2) Welche Arten von Signalen können durch Handschrift-Digitalisiertabletts erfasst und
+%aufgezeichnet werden, beschreiben Sie:
+%e3) Benennen Sie 5 Beispiele von statistischen Merkmalen, die aus Handschrift-Signalen zur
+%biometrischen Erkennung berechnet werden können:
+%e4) Benennen Sie 4 alternative Semantikklassen zur Unterschrift und diskutieren Sie deren Vor- und Nachteile:
+%
+%
+\subsection{Überlegen und benennen Sie jeweils mindestens einen möglichen Sensortyp für die folgenden Modalitäten:}
+%Modalitäten:
+%• Handschrift:
+%• Fingerabdruck:
+%• Sprache:
+%• Gesicht:
+%• Iris:
+%• Finger-/Handgeometrie:
+%• Gangerkennung:
+%• DNA:
+%
+\subsection{Führen Sie die praktische Übung „Handgeometrie“ durch!}
+\subsection{Führen Sie die praktische Übung „Offline Handschrift“ durch!}
+
 \end{document}

### 2024-12-03 biometrie/exercises.tex
Chapter 5

@@
 
-\subsubsection{Benennen Sie die drei unterschiedlichen Stufen (engl. Levels) von Fingerabdruckmerkmalen:}
+\subsubsection{Benennen Sie die drei unterschiedlichen Stufen (engl. Levels) von Fingerabdruckmerkmalen}
 
-\subsubsection{Erklären Sie die Merkmalsart ``Minutien'' und skizzieren Sie 5 unterschiedliche Beispiele:}
+Das Skript (5.1.1) führt die Modalität Fingerabdruck mit drei unterschiedlichen Stufen ein:
+
+\begin{enumerate}
+	\item Globale Merkmale – die Klassifizierung von zusammengesetzten Mustern der Fingerkuppe
+	\item Minutien-basierte Merkmale – die Eigenschaften von einem oder mehreren Graten auf der Fingerkuppe
+	\item Schweißporen-basierte Merkmale – die Klassifizierung von Poren-Mustern
+\end{enumerate}
+
+\subsubsection{Erklären Sie die Merkmalsart ``Minutien'' und skizzieren Sie 5 unterschiedliche Beispiele}
+
+Minutien-basierte Merkmale beziehen sich auf die Eigenschaften von Graten (im einzelnen oder reziprok). Im Skript 
+angeführte Beispiele sind:
+
+\begin{itemize}
+	\item Die Aufteilung eines Grates in zwei
+	\item Die Terminierung eines Grates an einem Punkt
+	\item Die Kreuzung von zwei oder mehreren Graten
+	\item Ein unabhängiger, freistehender Grat
+	\item Ein umschlossenes Tal von zwei oder mehreren Graten 
+\end{itemize}
 
@@
 
-%b1) Benennen Sie die drei grundsätzlichen Ziele der sprachbasierten Biometrie:
-%b2) Erklären Sie stichpunktartig das Konzept der „Mel-Frequency Cepstral Coefficients
-%(MFCC)“:
-%
+\subsubsection{Benennen Sie die drei grundsätzlichen Ziele der sprachbasierten Biometrie}
+
+Das Skript (5.2.1) führt die Modalität Stimme mit drei unterschiedlichen Zielen ein:
+
+\begin{enumerate}
+	\item Automatischen Sprechererkennung – Identität / Authentifizierung anhand des Stimm-Merkmals
+	\item Textuelle Spracherkennung – Gesprochene Inhalte in einen digitalen Text zu überführen
+	\item Emotionen aus Stimmcharakteristika – Identifikation von subtilen Merkmalen der Stimme / des Sprechers
+\end{enumerate}
+
+\subsubsection{Erklären Sie stichpunktartig das Konzept der ``Mel-Frequency Cepstral Coefficients (MFCC)''}
+
+\begin{itemize}
+	\item Ein Mel-Frequency-Cepstrum (MFC) beschreibt das energie Spektrum eines Geräusches
+	\item Ein MFC setzt sich aus vielen Koeffizienten (MFCCs) zusammen
+	\item Die MFCCs eines Geräusches lassen sich (stark vereinfacht) mittels einer kombination aus Fourier-Transformation, Überführung in die Mel-Scale, der Anwendung der Logarithmus-Funktion und einer diskreten Cosinus-Transformation ableiten.
+	\item Der Unterschied zwischen einem Cepstrum und dem Mel-Frequency-Cepstrum ist die verwendung der Mel-Scale für die Aufzeichnung der Energie-Werte
+\end{itemize}
 
 \subsection{Modalität Gangart}
+\annotation{Lesen Sie ggf. in der Quelle \href{http://persci.mit.edu/pub_pdfs/niyogi_XYT.pdf}{[NiAd1994]} nach.}
 
-% (lesen Sie ggf. in der Quelle [NiAd1994] nach, diese ist abrufbar unter
-%http://persci.mit.edu/pub_pdfs/niyogi_XYT.pdf
-%c1) Erklären Sie kurz das Aufzeichnungsszenario für den „kanonischen Gang“ („canonical
-%walk“):
-%c2) Erklären Sie: welche Bildverarbeitungsschritte sind notwendig, um hieraus ein so
-%genanntes XT Diagramm zu generieren:
-%c3) Überlegen, recherchieren und beschreiben Sie 5 Merkmale, welche aus XYT Diagrammen
-%gewonnen werden können:
+\subsubsection{Erklären Sie kurz das Aufzeichnungsszenario für den ``kanonischen Gang'' (``canonical walk'')}
+
+Das Szenario für die Aufzeichnung eines kanonischen Gangs setzt voraus, dass die Kamera die das Individuum aufzeichnet im
+90-Grad-Winkel zur Laufrichtung und ungefähr auf der gleichen Höhe wie die Kröpermitte platziert wird (ie. keine Top-Down 
+Perspektive).
+
+\subsubsection{Erklären Sie: welche Bildverarbeitungsschritte sind notwendig, um hieraus ein so genanntes XT Diagramm zu generieren}
+
+\begin{enumerate}
+	\item Entfernung von Hintergrundinformationen
+	\item Kantenerkennung
+	\item Extraktion der XT-Schnittlinie
+	\item Pixel-Extraktion entlang der Linie
+	\item Aufbau des XT-Diagramms
+\end{enumerate}
+
+\subsubsection{Überlegen, recherchieren und beschreiben Sie 5 Merkmale, welche aus XYT Diagrammen gewonnen werden können}
+
+Die folgenden 5 Merkmale bzw. Prozessierungsschritte können aus einem XYT Diagramm extrahiert werden bzw. darauf 
+ausgeführt werden.
+
+\begin{enumerate}
+	\item Geschwindigkeit und Beschleunigung
+	\item Trajektorienanalyse
+	\item Periodizität und Frequenzanalyse
+	\item Interaktionsanalyse
+	\item Anomalieerkennung
+\end{enumerate}
 
@@
 
-%d1) Benennen Sie Vorteile und Anwendungen dieser biometrischen Modalität:
-%d2) Erläutern Sie kurz das Konzept der „Digraphs“ und überlegen und beschreiben Sie die vier
-%Merkmale, die aus einem Digraph gemessen werden können:
-%d3) Argumentieren Sie mathematisch/logisch, wie das biometrische Verfahren Tastaturanschlag
-%zur Verbesserung der Passwortsicherheit eingesetzt werden kann und stellen Sie klar, welche
-%Annahmen Sie dabei tätigen:
+\subsubsection{Benennen Sie Vorteile und Anwendungen dieser biometrischen Modalität}
+
+Die Modalität der Tastaturanschläge bietet weitreichende Vorteile hinsichtlich ihres möglichen Einsatzgebietes, da:
+
+\begin{enumerate}
+	\item Keine Sensor-Notwendigkeit – Tastaturanschläge lassen sich mit jedem (tastatur-basiertem) Eingabegerät aufzeichnen ohne besondere Hardware
+	\item Gute Performance in textabhängigen Tests – Tastaturanschläge sind in Szenarien in denen die Eingabedaten bekannt sind sehr aussagekräftig über die Identität
+\end{enumerate}
+
+Das Tastaturanschlagsmuster kann z.B. als weiteren Faktor bei der Passwort-Authentifizierung verwendet werden, um zu 
+erkennen, wenn ein Imposter versucht sich in dem System anzumelden.
+
+\subsubsection{Erläutern Sie kurz das Konzept der ``Digraphs'' und überlegen und beschreiben Sie die vier Merkmale, die aus einem Digraph gemessen werden können}
+
+Ein ``Diagraph'' ist das Segment von 2 aufeinanderfolgenden Tastenanschlägen.
+Vier Merkmale die anhand von Diagraphen einer Eingabe abgeleitet werden können sind:
+
+\begin{enumerate}
+	\item Inneres Delta – die Zeit zwischen den Tasten innerhalb eines Diagraphen
+	\item Äußeres Delta – die Zeit zwischen zwei Diagraphen
+	\item Repetitions-Interval – Wie häufig ein Diagraph verwendet wird
+	\item Dauer – Wie lange ein Diagraph (von keydown der ersten Taste zu keyup der zweiten) dauert
+\end{enumerate}
+
+\subsubsection{Argumentieren Sie mathematisch/logisch, wie das biometrische Verfahren Tastaturanschlag zur Verbesserung der Passwortsicherheit eingesetzt werden kann und stellen Sie klar, welche Annahmen Sie dabei tätigen}
+
+Wie bereits unter 1) angeführt könnte die Tastaturanschlags-Modalität als zusätzlicher Faktor bei der 
+Passwort-Authentifizierung dienen. Hierbei kann das System das Eingabemuster des Authentischen-Benutzers lernen und
+verwenden, um (bei einer starken Abweichung) einen Imposter zu erkennen, obwohl das Passwort korrekt eingegeben wurde.
+
+Somit wird das System resilient(er) gegenüber Passwortdiebstahl – es könnte z.B. bei einer starken Abweichung das 
+Passwort zurücksetzen. Die Tastaturanschlags-Modalität eignet sich, wie im Skript angeführt, sehr gut für diesen
+Anwendungsfall, da der Eingabetext bekannt ist.
 
@@
 
-%e1) Benennen Sie die beiden grundlegenden Kategorien von Merkmalen handschriftbasierter
-%Biometrie:
-%e2) Welche Arten von Signalen können durch Handschrift-Digitalisiertabletts erfasst und
-%aufgezeichnet werden, beschreiben Sie:
-%e3) Benennen Sie 5 Beispiele von statistischen Merkmalen, die aus Handschrift-Signalen zur
-%biometrischen Erkennung berechnet werden können:
-%e4) Benennen Sie 4 alternative Semantikklassen zur Unterschrift und diskutieren Sie deren Vor- und Nachteile:
-%
-%
+\subsubsection{Benennen Sie die beiden grundlegenden Kategorien von Merkmalen handschriftbasierter Biometrie}
+
+Die beiden grundlegenden Kategorien von Merkmalen handschriftbasierter Biometrie sind:
+
+\begin{itemize}
+	\item Statische Merkmale: Alle Merkmale, die aus einem statischen Bild der Handschrift extrahiert werden können (z. B. eingescannter Zettel).
+	\item Dynamische Merkmale: Alle Merkmale, die während des Schreibprozesses erfasst werden und Bewegungsabläufe sowie zeitliche Eigenschaften beschreiben.
+\end{itemize}
+
+\subsubsection{Welche Arten von Signalen können durch Handschrift-Digitalisiertabletts erfasst und aufgezeichnet werden, beschreiben Sie}
+
+\begin{enumerate}
+	\item Position der Hand
+	\item Position des Stifts (z.B. Koordinaten)
+	\item Druck
+	\item Stift-Berührungen des Displays
+	\item Neigung
+	\item Zeitstempel
+\end{enumerate}
+
+\subsubsection{Benennen Sie 5 Beispiele von statistischen Merkmalen, die aus Handschrift-Signalen zur biometrischen Erkennung berechnet werden können}
+
+\begin{enumerate}
+	\item Strichlänge
+	\item Frequenz der Stift-Berührungen
+	\item Schreibwinkel
+	\item Schleifen
+\end{enumerate}
+
+\subsubsection{Benennen Sie 4 alternative Semantikklassen zur Unterschrift und diskutieren Sie deren Vor- und Nachteile}
+
+\begin{enumerate}
+	\item Wörter
+	\item Sätze
+	\item Zahlen
+	\item Freitext
+\end{enumerate}
+
 \subsection{Überlegen und benennen Sie jeweils mindestens einen möglichen Sensortyp für die folgenden Modalitäten:}
-%Modalitäten:
-%• Handschrift:
-%• Fingerabdruck:
-%• Sprache:
-%• Gesicht:
-%• Iris:
-%• Finger-/Handgeometrie:
-%• Gangerkennung:
-%• DNA:
-%
+
+\begin{itemize}
+	\item Handschrift – optischer Sensor
+	\item Fingerabdruck – optischer Sensor
+	\item Sprache – Mikrofon
+	\item Gesicht – optischer Sensor, LIDAR
+	\item Iris – optischer Sensor
+	\item Finger-/Handgeometrie – optischer Sensor, LIDAR
+	\item Gangerkennung – optischer Sensor
+	\item DNA – N/A
+\end{itemize}
+
 \subsection{Führen Sie die praktische Übung „Handgeometrie“ durch!}

### 2024-12-03 biometrie/exercises.tex
Chapter 5

@@
 
-\subsubsection{Erläutern Sie kurz das Konzept der ``Digraphs'' und überlegen und beschreiben Sie die vier Merkmale, die aus einem Digraph gemessen werden können}
+\subsubsection{Erläutern Sie kurz das Konzept der ``Digraph'' und überlegen und beschreiben Sie die vier Merkmale, die aus einem Digraph gemessen werden können}
 

### 2024-12-12 biometrie/exercises.tex
Chapter 5
Chapter 6

@@
 \subsection{Führen Sie die praktische Übung „Handgeometrie“ durch!}
+
+\annotation{Nicht zutreffend (wurde bereits als Team-Aufgabe erledigt).}
+
 \subsection{Führen Sie die praktische Übung „Offline Handschrift“ durch!}
 
+\annotation{Nicht zutreffend (wurde bereits als Team-Aufgabe erledigt).}
+
+\newpage
+
+\section{}
+
+\subsection{Erläutern Sie die Unterschiede zwischen den beiden Varianten ``Key Binding'' und ``Key Generation''
+der Schlüssel-basierten Biometrischen Crypto-Systeme!}
+
+Die Unterschiede zwischen der ``Key Binding''- und der ``Key Generation'' Variante lassen sich in mehreren Dimensionen vergleichen:
+
+\subsubsection*{Relation zum Schlüssel}
+
+Während im ``Key Binding''-Verfahren der kryptographische Schlüssel komplett Unabhängig von den biometrischen Eingabedaten ist, wird
+im ``Key Generation''-Verfahren der Schlüssel aus den Eingabedaten abgeleitet. Dies kann u.U. weitreichende kryptographische 
+Implikationen mit sich bringen. Eine offensichtliche Implikation der ``Key Generation''-Variante ist, dass sich der Schlüssel mit jedem 
+neuen Enrollment eines Nutzers ändert und sich somit weniger für eine langfristige Verschlüsselung eignet. Wohingegen das Enrollment
+des Nutzers und der kryptographische Schlüssel im ``Key Binding''-Verfahren unabhängig sind.
+
+\subsubsection*{Schlüssel-Rotation}
+
+Während bei der ``Key Binding''-Variante eine Rotation des kryptographischen Schlüsselmaterials ohne größere Probleme stattfinden kann,
+muss bei der ``Key Generation'' ein neues Enrollment stattfinden um einen neuen, unabhängigen Schlüssel zu generieren.
+
+\subsubsection*{Revokierbarkeit}
+
+Durch die biometrische Zugangskontrolle zum Schlüsselmaterial können biometrische Krypto-Systeme die auf dem ``Key Binding''-Verfahren 
+beruhen den Zugang des Nutzers revokieren, in dem diese die Verlinkung zwischen Schlüsselpaar und Nutzer aufheben. Im ``Key 
+Generation''-Verfahren, kann es abhängig vom verwendeten Algorithmus sein, dass der Nutzer in der Lage ist, den Schlüssel aus seinen 
+Eingabedaten abzuleiten – somit ist die Revokierbarkeit nicht zwingend gewährleistet.
+
+\subsubsection*{Schlüssellänge}
+
+Da die biometrischen Gegebenheiten (bspw. Intra- bzw. Inter-Klassenvariabilität) bei dem ``Key Binding''-Verfahren keinen Einfluss
+auf das kryptographische Schlüsselmaterial hat, gibt es keine Begrenzung hinsichtlich möglicher Schlüssellängen etc. – wohingegen diese
+auf Grund der mathematischen Ableitung aus den Eingabedaten bei dem ``Key Generation''-Verfahren besteht.
+
+\subsubsection*{Speicherung}
+
+Letztlich unterscheiden sich die beiden Varianten hinsichtlich der im System gespeicherten Daten. Beim ``Key Binding'' wird das
+biometrisch verschlüsselte Chiffrat gespeichert, alle anderen Daten werden schnellstmöglich Vernichtet. Bei der ``Key Generation''-
+Variante werden die Hilfsdaten, die notwendig für die Schlüssel-Rekonstruktion sind, im biometrischen System gespeichert.
+
+\subsection{Welche Unterschiede bestehen zwischen den Konzepten biometrische Schlüsselgenerierung und
+revokierbaren biometrische Verfahren (engl. Cancelable Biometrics) hinsichtlich}
+
+\subsubsection{den Zielsetzungen}
+
+Primäre Zielsetzung der biometrischen Schlüsselgenerierung umfasst:
+
+\begin{enumerate}
+	\item Die Generierung kryptographischen Schlüsselmaterials aus biometrischen Merkmalen
+	\item Die Minimierung von Daten, die ein Nutzer speichern muss (ie. der Schlüssel lässt sich biometrisch wiederherstellen)
+	\item Erhöhte Sicherheit durch die Verknüpfung von biometrischen Merkmalen mit kryptographischen Verfahren
+\end{enumerate}
+
+Primäre Zielsetzung der revokierbaren biometrischen Verfahren umfasst:
+
+\begin{enumerate}
+	\item Der Schutz der ursprünglichen biometrischen Eingabedaten durch irreversible Transformationen.
+	\item Fokus auf die revokierbarkeit des Verfahrens, trotz biometrischer Koppelung
+\end{enumerate}
+
+\subsubsection{möglichen Anwendungen}
+
+Mögliche Anwendungsfälle der biometrischen Schlüsselgenerierung könnten umfassen:
+
+\begin{enumerate}
+	\item Schutz sensibler Daten 
+	\item Authentifizierung in sicherheitskritischen Systemen
+	\item Temporäre Schlüssel für eine einmalige Verwendung
+\end{enumerate}
+
+Mögliche Anwendungsfälle der revokierbaren biometrische Verfahren könnten umfassen:
+
+\begin{enumerate}
+	\item Anwendungsbereiche mit Fokus auf Datenschutz und Wiederverwendbarkeit
+	\item Authentifizierung mit hoher Flexibilität (ie. Smartcards)
+\end{enumerate}
+
+\subsection{Erläutern Sie in Stichpunkten, den Ablauf des Merkmalsvergleichs bei revokierbaren biometrischen Verfahren (engl. Cancelable Biometrics)}
+
+\begin{enumerate}
+    \item Erfassung der biometrischen Eingabedaten (inkl. Feature-Extraction)
+    \item Irreversible Transformation der Merkmale
+    \item Vergleich mit Referenzdaten
+    \item Authentifizierungs-Entscheidung
+\end{enumerate}
+
+\subsubsection{Wodurch entsteht dabei die Schutzwirkung der biometrischen Referenzen vor Missbrauch}
+
+\begin{enumerate}
+    \item Irreversible Transformation der Merkmale
+    \item Vermeidung von Cross-Application-Linking von biometrischen Daten
+    \item Keine Speicherung von biometrischen Eingabedaten
+    \item Revokierbarkeit (ie. Veränderung der Transformation)
+\end{enumerate}
+
+\subsubsection{Wie kann die Revokierbarkeit also der Widerruf der biometrischen Registrierung, erwirkt
+werden?}
+
+In dem die irreversible Transformations-Funktion ausgetauscht wird. Somit muss ein neues Enrollment
+(mit der neuen Transformation-Funktion) stattfinden, um sich erneut zu Authentifizieren und es ist gleichzeitig nicht mehr Möglich
+gegen die bestehenden Referenz-Daten zu Authentifizieren (bei gleichbleibenden, authentischen Eingabedaten).
+
+\newpage
+
+\section{}
+
+\subsection{Erläutern Sie den Begriff „biometrische Attributierung“ anhand eines selbst gewählten Beispiels.}
+
+\annotation{Grenzen Sie dabei auch den Begriff zur „biometrischen Authentifizierung“ ab.}
+
+\subsection{Überlegen Sie: welche Attribute können prinzipiell aus Sprach-Biometrie abgeleitet werden?}
+
+\annotation{Beschreiben Sie dabei mindestens zwei Merkmale in einem Audiosignal, welche hierfür auswertet werden könnten!}
+
+\newpage
+
+\section{}
+
 \end{document}

### 2024-12-15 biometrie/exercises.tex
Chapter 7 & 8

@@
 
-\newpage
-
 \section{}
 
-\subsection{Erläutern Sie den Begriff „biometrische Attributierung“ anhand eines selbst gewählten Beispiels.}
+\subsection{Erläutern Sie den Begriff ``biometrische Attributierung'' anhand eines selbst gewählten Beispiels.}
 
 \annotation{Grenzen Sie dabei auch den Begriff zur „biometrischen Authentifizierung“ ab.}
+
+Biometrische Attributierung bezieht sich auf das Verfahren der Ableitung von Informationen über ein Individuum durch 
+die Auswertung einer, oder mehrerer biometrische Modalitäten. So lassen sich mit ausreichend hoher Wahrscheinlichkeit
+Vorhersagen über die Herkunft einer Person treffen in dem Modalitäten wie Sprache, Haut- und Haarfarbe, Größe, etc. in 
+Kombination betrachtet werden.
+
+Dieses Verfahren unterscheidet sich vorrangig durch die Motivation der Auswertung zur biometrischen Authentifizierung. 
+Während die biometrische Authentifizierung sich meistens auf eine einzige Modalität bezieht und lediglich das Ziel hat
+zuverlässig die Identität eines einzelnen Individuums sicherzustellen (durch einen Verglich zu bereits hinterlegten 
+Referenzdaten), versucht die biometrische Attributierung, Rückschlüsse über ein (oder mehrere) Individuum/en zu ziehen.
 
@@
 
-\newpage
+Hier sind einige Beispiele für Attribute die aus der Sprache einer Person abgeleitet werden können:
+
+\begin{enumerate}
+	\item Herkunft (ie. ausgedrückt durch Sprache bzw. Dialekt, Stimm-Lage etc.)
+	\item Alter (ie. durch Sprach-Elemente, Stimm-Lage / Stimm-Charakteristika etc.)
+	\item Soziale Herkunft / Zugehörigkeit (ie. durch Phrasen/Wörter, Dialekt etc.)
+	\item Geschlecht (ie. durch Stimm-Lage, Stimm-Charakteristika, Lautstärke, etc.)
+\end{enumerate}
 
@@
 
+\subsection{Angenommen, Sie sollen die Evaluation für ein biometrisches System mittels Tastatur-Anschlagsdynamik durchführen und sind für die Datensammlung verantwortlich.}
+
+\annotation{Überlegen Sie sich hierfür ein Protokoll und beschreiben
+	Sie dieses unter der Berücksichtigung folgender Rahmenbedingungen:\\\\
+	Es sollen je Proband zwei Semantiken untersucht werden: jeweils ein
+	selbstgewähltes individuelles Passwort und ein für alle Probanden gleiches.
+	Wählen Sie nach ihrer Ansicht geeignete Vorgaben und begründen Sie diese!\\\\
+	Überlegen Sie eine sinnvolle Reihenfolge für die Erfassung der Enrolment-
+	bzw. Verifikationsdaten!\\\\
+	Überlegen Sie darüber hinaus ob und welche Fälschungsszenarien umsetzbar sind.
+	Beschreiben Sie im Protokoll, wie diese durch die Probanden umgesetzt werden
+	sollen und welche Kategorien entstehen.
+}
+
+Ein realistisches Evaluationsprotokoll könnte folgenden Rahmenbedinungen unterliegen:
+
+\subsubsection*{Semantik-Vorgaben}
+
+Die Länge, Zusammensetzung, Komplexität etc. des vorgegeben Passworts, sowie des selbstgewählten Passworts
+werden vorgegeben und orientieren sich an, für echte Authentifizierung üblichen / empfohlenen Werten.
+
+\subsubsection*{Sensorik-Vorgaben}
+
+Alle Probanden müssen die Datenerhebung mittels der gleichen Sensorik (ie. Tastatur-Modell) vornehmen, um
+eine Vergleichbarkeit zwischen Probanden zu erhalten.
+
+\subsubsection*{Abfolge-Vorgaben}
+
+Die Erfassung der Daten sollte folgendermaßen erfolgen:
+
+\begin{enumerate}
+	\item Selbstgewähltes Passwort – Enrolment
+	\item Vorgegebenes Passwort – Enrolment
+	\item Selbstgewähltes Passwort – Verifikation
+	\item Vorgegebenes Passwort – Verifikation
+\end{enumerate}
+
+Somit gibt es innerhalb einer Semantik eine feste Reihenfolge (ie. erst Enrolment, dann Verifizierung),
+die Semantiken werden aber interleaved erfasst, was dabei Hilft, eine stärkere Variation der Eingabedaten
+zu erhalten, da die Probanden zwischen drin das Muster wechseln müssen. (Ie. nicht (bspw.) 10-mal das 
+gleichen Passwort am Stück tippen).
+
+Eine erweiterte Form könnte hier sein, dass die einzelnen Proben der Verifikationsdaten abwechselnd 
+erhoben werden, um eine stärkere Streuung in der Verifikation zu erhalten, was der Realität am nächsten 
+käme. Ie. beim Enrolment werden – verhältnsmäßig – homogene Daten erhoben, wohingegen bei der Verifikation
+eine breitere Streuung besteht.
+
+\subsubsection*{Fälschungs-Test}
+
+Es könnte (angenommen, es gibt keine Ressourcen für eine weitere Datenerhebung) ein ``Zero-Effort''-
+Fälschungs-Test durchgeführt werden, in dem pro Proband \textbf{alle} verfügbaren Verifikationsdaten 
+(inkl. aller anderen Probanden) zur Verifikation verwendet werden.
+
+Die FMR bzw. FAR ist besonders Aussagekräftig für das vorgegeben Passwort, da die Verifikationsdaten nah
+an den Enrolmentdaten liegen, selbst wenn sie von einem anderen Probanden stammen, da es sich um das 
+gleiche Passwort handelt.
+
+Ein komplett zufälliger Angriff kann durchgeführt werden, in dem lediglich die Semantik des 
+selbstgewählten Passworts betrachtet wird, da es somit keinen direkten Zusammenhang zwischen
+Enrolment und Verifikations-Daten gibt.
+
+\subsection{Angenommen, bei der Datenerhebung zu einer biometrischen Evaluation nehmen 20 Probanden teil. Jede/r Proband/in gibt dabei 10 Enrolment-Proben, sowie 20 Verifikationsproben ab.}
+
+\subsubsection{Wieviele Einzel-Verifikationsversuche sind dann zur Ermittlung der FNMR/FRR über alle Probanden erforderlich?}
+
+Zur Ermittlung der FNMR bzw. FRR über alle Probanden hinweg sind folgende Verifikationsversuche notwendig:
+
+\begin{enumerate}
+	\item Pro Proband sind 100 Verifikationsversuche notwendig (10 Enrolment-Proben $\times$ 10 Verifikations-Proben)
+	\item Für alle Probanden (20) sind also 2000 Verifikationsversuche notwendig (ie. 100 $\times$ 20)
+\end{enumerate}
+
+Dies leitet sich aus der Definition der FNMR bzw FRR ab, die beschreibt, wie oft eine Verifikationsprobe
+eines Probanden \textbf{nicht} erfolgreich zur Authentifizierung geführt hat.
+
+\annotation{Geben Sie die Berechnungsvorschrift an!}
+
+\subsubsection{Wieviele Einzel-Verifikationsversuche sind es zur Ermittlung der FMR\textsubscript{Random} / FAR\textsubscript{Random}?}
+
+\annotation{Geben Sie die Berechnungsvorschrift an!}
+
+Zur Ermittlung der FMR\textsubscript{Random} / FAR\textsubscript{Random} über alle Probanden hinweg sind folgende Verifikationsversuche notwendig:
+
+\begin{enumerate}
+	\item Pro Proband wird jede Enrolment-Probe (10) mit jeder Verifikations-Probe (10) \textbf{jedes anderen} Probanden verglichen (19).
+Dies Ergibt 1900 Verifikationsversuche pro Proband (10 eigene Enrolment-Proben $\times$ 10 fremde Verifikations-Proben $\times$ 19 andere Probanden)
+	\item Für alle Probanden (20) sind also 38000 Verifikationsversuche notwendig (ie. 1900 $\times$ 20)
+\end{enumerate}
+
+Dies leitet sich aus der Definition der FMR bzw. FAR ab, die beschreibt wie Wahrscheinlich es ist, dass 
+eine eine Verifikationsprobe eines anderen Probanden fälschlicherweise als authentische klassifiziert wird 
+(je Proband und Enrolment-Probe).
+
+Anmerkung: Diese Verifikationstests wären bereits im Rahmen eines ``Zero-Effort''-Fälschungstests 
+durchgeführt worden und die FMR\textsubscript{Random} / FAR\textsubscript{Random} könnten somit ohne 
+weiteren Aufwand berechnet werden.
+
 \end{document}

### 2024-12-15 biometrie/exercises.tex
Typos

@@
 auch Falsch-Nichterkennungsrate), Verhältnis zwischen der Anzahl der nicht bestätigten authentischen Verifikationen (d.h.
-Verifizierungsdaten und Enrolmentdaten von identischen Personen werden fälschlicher Weise negativ verifiziert) und der gesamten Anzahl von Verifikationstests.''
+Verifizierungsdaten und Enrollmentdaten von identischen Personen werden fälschlicher Weise negativ verifiziert) und der gesamten Anzahl von Verifikationstests.''
 }
@@
 Verhältnis zwischen der Anzahl der aufgetretenen Falsch-Verifikationen (d.h.
-Verifizierungsdaten und Enrolmentdaten von unterscheidlichen Personen werden fälschlicher Weise positiv verifiziert) und der gesamten Anzahl von Verifikationstests''
+Verifizierungsdaten und Enrollmentdaten von unterscheidlichen Personen werden fälschlicher Weise positiv verifiziert) und der gesamten Anzahl von Verifikationstests''
 }
@@
 
-Während im ``Key Binding''-Verfahren der kryptographische Schlüssel komplett Unabhängig von den biometrischen Eingabedaten ist, wird
+Während im ``Key Binding''-Verfahren der kryptographische Schlüssel komplett unabhängig von den biometrischen Eingabedaten ist, wird
 im ``Key Generation''-Verfahren der Schlüssel aus den Eingabedaten abgeleitet. Dies kann u.U. weitreichende kryptographische 
@@
 Vorhersagen über die Herkunft einer Person treffen in dem Modalitäten wie Sprache, Haut- und Haarfarbe, Größe, etc. in 
-Kombination betrachtet werden.
+kombination betrachtet werden.
 
@@
 	Wählen Sie nach ihrer Ansicht geeignete Vorgaben und begründen Sie diese!\\\\
-	Überlegen Sie eine sinnvolle Reihenfolge für die Erfassung der Enrolment-
+	Überlegen Sie eine sinnvolle Reihenfolge für die Erfassung der Enrollment-
 	bzw. Verifikationsdaten!\\\\
@@
 \begin{enumerate}
-	\item Selbstgewähltes Passwort – Enrolment
-	\item Vorgegebenes Passwort – Enrolment
+	\item Selbstgewähltes Passwort – Enrollment
+	\item Vorgegebenes Passwort – Enrollment
 	\item Selbstgewähltes Passwort – Verifikation
@@
 
-Somit gibt es innerhalb einer Semantik eine feste Reihenfolge (ie. erst Enrolment, dann Verifizierung),
+Somit gibt es innerhalb einer Semantik eine feste Reihenfolge (ie. erst Enrollment, dann Verifizierung),
 die Semantiken werden aber interleaved erfasst, was dabei Hilft, eine stärkere Variation der Eingabedaten
@@
 erhoben werden, um eine stärkere Streuung in der Verifikation zu erhalten, was der Realität am nächsten 
-käme. Ie. beim Enrolment werden – verhältnsmäßig – homogene Daten erhoben, wohingegen bei der Verifikation
+käme. Ie. beim Enrollment werden – verhältnsmäßig – homogene Daten erhoben, wohingegen bei der Verifikation
 eine breitere Streuung besteht.
@@
 Die FMR bzw. FAR ist besonders Aussagekräftig für das vorgegeben Passwort, da die Verifikationsdaten nah
-an den Enrolmentdaten liegen, selbst wenn sie von einem anderen Probanden stammen, da es sich um das 
+an den Enrollmentdaten liegen, selbst wenn sie von einem anderen Probanden stammen, da es sich um das 
 gleiche Passwort handelt.
@@
 selbstgewählten Passworts betrachtet wird, da es somit keinen direkten Zusammenhang zwischen
-Enrolment und Verifikations-Daten gibt.
+Enrollment und Verifikations-Daten gibt.
 
-\subsection{Angenommen, bei der Datenerhebung zu einer biometrischen Evaluation nehmen 20 Probanden teil. Jede/r Proband/in gibt dabei 10 Enrolment-Proben, sowie 20 Verifikationsproben ab.}
+\subsection{Angenommen, bei der Datenerhebung zu einer biometrischen Evaluation nehmen 20 Probanden teil. Jede/r Proband/in gibt dabei 10 Enrollment-Proben, sowie 20 Verifikationsproben ab.}
 
@@
 \begin{enumerate}
-	\item Pro Proband sind 100 Verifikationsversuche notwendig (10 Enrolment-Proben $\times$ 10 Verifikations-Proben)
+	\item Pro Proband sind 100 Verifikationsversuche notwendig (10 Enrollment-Proben $\times$ 10 Verifikations-Proben)
 	\item Für alle Probanden (20) sind also 2000 Verifikationsversuche notwendig (ie. 100 $\times$ 20)
@@
 
-\subsubsection{Wieviele Einzel-Verifikationsversuche sind es zur Ermittlung der FMR\textsubscript{Random} / FAR\textsubscript{Random}?}
+\subsubsection{Wieviele Einzel-Verifikations-Versuche sind es zur Ermittlung der FMR\textsubscript{Random} / FAR\textsubscript{Random}?}
 
@@
 
-Zur Ermittlung der FMR\textsubscript{Random} / FAR\textsubscript{Random} über alle Probanden hinweg sind folgende Verifikationsversuche notwendig:
+Zur Ermittlung der FMR\textsubscript{Random} / FAR\textsubscript{Random} über alle Probanden hinweg sind folgende Verifikations-Versuche notwendig:
 
 \begin{enumerate}
-	\item Pro Proband wird jede Enrolment-Probe (10) mit jeder Verifikations-Probe (10) \textbf{jedes anderen} Probanden verglichen (19).
-Dies Ergibt 1900 Verifikationsversuche pro Proband (10 eigene Enrolment-Proben $\times$ 10 fremde Verifikations-Proben $\times$ 19 andere Probanden)
-	\item Für alle Probanden (20) sind also 38000 Verifikationsversuche notwendig (ie. 1900 $\times$ 20)
+	\item Pro Proband wird jede Enrollment-Probe (10) mit jeder Verifikations-Probe (10) \textbf{jedes anderen} Probanden verglichen (19).
+Dies Ergibt 1900 Verifikations-Versuche pro Proband (10 eigene Enrollment-Proben $\times$ 10 fremde Verifikations-Proben $\times$ 19 andere Probanden)
+	\item Für alle Probanden (20) sind also 38000 Verifikations-Versuche notwendig (ie. 1900 $\times$ 20)
 \end{enumerate}
@@
 Dies leitet sich aus der Definition der FMR bzw. FAR ab, die beschreibt wie Wahrscheinlich es ist, dass 
-eine eine Verifikationsprobe eines anderen Probanden fälschlicherweise als authentische klassifiziert wird 
-(je Proband und Enrolment-Probe).
+eine eine Verifikations-Probe eines anderen Probanden fälschlicherweise als authentische klassifiziert 
+wird (je Proband und Enrollment-Probe).
 
-Anmerkung: Diese Verifikationstests wären bereits im Rahmen eines ``Zero-Effort''-Fälschungstests 
+Anmerkung: Diese Verifikations-Tests wären bereits im Rahmen eines ``Zero-Effort''-Fälschungstests 
 durchgeführt worden und die FMR\textsubscript{Random} / FAR\textsubscript{Random} könnten somit ohne 

### 2025-01-02 module/ethik-in-der-it-sicherheit/exam.tex
Restructure repo

\documentclass[journal]{IEEEtran}

\usepackage[a4paper,top=5cm,bottom=5cm]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}

\pagenumbering{Roman}

\begin{document}

\markboth{THB – Ethik in der IT Sicherheit – Mara Schulke — 20215853}{}

\begin{onecolumn}

\begin{center}
	\Large{Geistiges Eigentum im Zeitalter der Informationstechnik}\\
	\vspace{18pt}\hline
\end{center}

\tableofcontents

\section{Einleitung}

Das geistige Eigentum und die Sicherstellung von Eigentumsrechten sind von elementarer 
Bedeutung für unsere Gesellschaft und ihre Ordnung. Das Schützen der Werke einzelner 
Individuen oder ganzer Organisationen ist eine fundamentale ethische Verpflichtung, der 
die Gesellschaft während der voranschreitenden Digitalisierung gerecht werden muss. Der 
technologische Fortschritt darf nicht zur Verwahrlosung des Eigentumsrechts führen, aber 
digitales Eigentum lässt sich gleichzeitig deutlich schwerer kontrollieren als physisches 
Eigentum. Es ist ein komplexer Balanceakt zwischen dem Schutz von Einzelnen und dem Wohl 
der Mehrheit.

Bei der Betrachtung von Schutzregelungen für Eigentum ergeben sich interessante
Fragestellungen. Zum Beispiel stellt sich die Frage, ob in bestimmten Situationen 
Ausnahmen von diesem Schutz gemacht werden dürfen oder müssen, wenn dies dem Gemeinwohl 
der Gesellschaft dient?

\section{Einfluss der Informationstechnik}

Die Digitalisierung hat einen erheblichen Einfluss auf unseren Umgang 
mit geistigen und digitalen Eigentum. 
Der technologische Fortschritt hat in unserer Gesellschaft zu einer weitreichenden
Digitalisierung von Waren geführt – viele Gegenstände die früher 
nur in physischer Form und dementsprechend nur begrenzt erhältlich waren, sind nun 
jederzeit auf Abruf verfügbar und lassen sich unbegrenzt oft vervielfältigen. 
Hierdurch sind ganze Teilbereiche der Kriminalität im Bezug auf den digitalen Bereich 
entstanden, wie zum Beispiel die Piraterie oder die illegale Verbreitung von
lizensierten / gekauften Inhalten.\cite{oe-it}

Diese Veränderungen werfen neue Fragen und Herausforderungen auf wie zum Beispiel 
die Notwendigkeit für einen Kopierschutz, Lizenzierungen von Software etc.
Mit diesen neuen Herausforderung bring der Fortschritt allerdings auch 
neue Möglichkeiten zur Sicherung des Eigentums in der digitalen Welt mit sich.
Konzepte wie der Einsatz von Kryptographie und Plagiatserkennungs-Algorithmen 
eröffnen Möglichkeiten, die Integrität und Nachvollziehbarkeit des geistigen und 
digitalen Eigentums sicherzustellen.

Es ist wichtig diese Probleme und Herausforderungen ernst zunehmen um die Wahrung von 
Ideen und geistigem Eigentum heute und künftig sicherzustellen. Hier zu wurde im Jahre 
1967 die weltweite Organisation WIPO (World Intellectual Property Organization) gegründet. 
\cite{wipo}

\section{Ethische Betrachtung von Eigentum}

Bei der ethischen Betrachtung dieses Themas stellen sich Fragen im Bezug auf die 
Verantwortung, wie zum Beispiel wer ist für die Sicherstellung der Eigentumsrechte
im digitalen Raum verantwortlich?

Im Falle von Musik ist ein bekannter Anbieter der Musik-Streaming-Dienst Spotify: Wer ist 
in diesem Fall für Sicherstellung von Urheberrechten verantwortlich und wo ziehen sich 
Verantwortungsbereiche durch diese Thematik? Ist bei der Verhinderung von Raubkopien 
das Unternehmen, dass die Plattform zur Verbreitung bereitstellt, der Gesetzgeber, oder 
das Individuum, das die Plattform zur Verbreitung verwendet gefragt? In der Realität zeigt 
sich, dass die Wahrung des geistigen Eigentums im Verantwortungsbereich von allen drei 
Akteuren liegt. Die Plattform muss die Verletzung feststellen – ein gutes Beispiel ist 
hier YouTube, die die Monetarisierung von urheberrechtlich geschützten Inhalten verbieten. 
Der Gesetzgeber muss einen rechtlichen Rahmen stellen um strafrechtlich gegen die 
Verletzung von Urheberrechten vorzugehen – und jedes Individuum trägt einen Teil 
Eigenverantwortung keine Urheberrechtsverletzung vorzunehmen, und dessen eigene Werke 
so gut es geht davor zu schützen.

Die zugrundeliegende ethische Fragestellung führt auf das gesellschaftliche Wertesystem 
zurück und der Umgang einer jeden Gesellschaft ist abhängig davon, wie sehr Fairness, 
Respekt und Gerechtigkeit in diesem einen Platz finden.

Eine Anerkennung der Rechte der Hersteller und Besitzer stehen der Problematik der 
gerechten Verteilung und dem Grundprinzip der Chancengleichheit gegenüber. Bildet sich
konzentriertes Besitztum innerhalb einer Gesellschaft hat dies ab einem gewissen Grat
einen Einfluss auf die Machtverteilung – sowohl in der physischen als auch in der 
digitalen Welt. Ein gutes Beispiel hierfür sind die Registrare von Domains im Internet, 
oder Firmen die über eine immense Rechenleistung verfügen. Die Entscheidungsposition in
die diese Akteure durch ihr Eigentum geraten bringt ein hohes Maß an Verantwortung mit 
sich, um eine möglichst ausgeglichene und demokratische Gesellschaft beizubehalten.

Die ethische Betrachtung des Themas Eigentum im Zeitalter der Informationstechnik 
erfordert daher eine tiefgreifende Auseinandersetzung mit ethischen Prinzipien wie 
Machtverteilung, Chancengleichheit und Gerechtigkeit in der digitalen Welt.

\section{Dilemmata}

Das Konzept von digitalen Besitztümern wirft eine Reihe von Konflikten und Dilemmata bei 
näherer Betrachtung auf. Hierbei fallen das Thema Datenschutz (und damit einhergehende 
Besitzrechte über Daten) und Privatsphäre als ein Kernthema heraus. Mit der immer größer
werdenden Datenspur im Internet und der Verlagerung von physischen Gütern (zum Beispiel 
Fotoalben) hinein in die digitale Welt werden Fragen werden Fragen aufgeworfen, wie diese
Spur unter Kontrolle zu behalten ist. Gerade im Falle des verscheiden einer Person ist die 
rechtliche Lage zur Behandlung der Nachlass-Angelegenheiten noch ungeklärt beziehungsweise
uneinheitlich. \cite{law-journal} Hier steht der Gesetzgeber in der Pflicht, ein 
einheitliches Regelwerk aus Richtlinien und Gesetzen zu schaffen, um den Umgang mit 
digitalen Daten und Eigentumsfragen lückenlos zu klären.

Der Zugang zu Informationen und Wissen stellt eine weitere problematische Dimension der
Digitalisierung dar. Obwohl es mit dem Internet eine generelle Erleichterung gab, an 
Informationen und Wissen zu gelangen, bestehen noch immer Kluften, was die Verfügbarkeit 
und den Zugang zu Ressourcen und Informationen angeht. Diese Ungleichheit für zu einer 
Fragmentierung der Gesellschaft in zwei Gruppen: Diejenigen die Zugang zu Bildung und 
Wissen haben, und denjenigen, die davon ausgeschlossen sind.

\section{Schlussfolgerung}

Insgesamt zeigt sich, dass Eigentum im digitalen Raum mit einer Reihe von 
Herausforderungen für die Gesellschaft als Ganzes einhergeht. Datenschutz, Urheberrechte 
und die gerechte Verteilung von Zugangsmöglichkeiten sind Fragen, denen sich Unternehmen 
und Gesetzgeber stellen müssen. Es ist von fundamentaler Bedeutung für unser Wertesystem 
digitales Eigentum zu schützen und gleichzeitig eine offene Wissenskultur zu fördern.

Geistiges Eigentum und technologischer Fortschritt sind eng miteinander verwebt: Künftige 
Entwicklungen werden weitere Herausforderungen mit sich bringen (Beispielsweise generative 
KI) aber ebenso neue Lösungen zur Sicherung von Urheberrechten und Demokratisierung von 
Wissen bereitstellen. Es ist wichtig, dass sich die Gesellschaft, Unternehmen und einzelne 
Individuen der Herausforderungen bewusst sind und daran arbeiten, eine gerechtere Zukunft 
zu schaffen.

\begin{thebibliography}{1}

\bibitem{oe-it} G. G. Goldacker. (2019, Dezember). {\em ``Vom Eigentum zum digitalen Nutzungsrecht''}.
Kompetenzzentrum Öffentliche IT. [Online]. Verfügbar:
\url{https://www.oeffentliche-it.de/-/vom-eigentum-zum-digitalen-nutzungsrecht}

\bibitem{wipo} WIPO. (1979, September).
{\em ``Convention Establishing the World Intellectual Property Organization''}.
WIPO [Online]. Verfügbar: \url{https://www.wipo.int/wipolex/en/text/283854/}

\bibitem{law-journal} K. S. Strutin. (2011, Sommer). {\em``The Digital Estate''}.
NYSBA Trusts and Estates Law Section Newsletter. Vol 44. Nr 2. Seite 20. Verfügbar:
\url{https://nysba.org/NYSBA/Publications/Section\%20Publications/Trusts\%20and\%20Estates/
PastIssues2000present/Summer2011/Summer2011Assets/TENewsSum11.pdf#page=20}

\end{thebibliography}

\end{onecolumn}

\end{document}

### 2025-01-02 module/netzwerk-sicherheit/esa-1/paper.tex
Restructure repo

--- 2023-10-21 netzwerk-sicherheit/übung-1.tex
@@
 
-Der Aufgabe entsprechend haben wir einen Schwachstellen-Scan bei dem vorgegebenen 
-Zielserver durchgeführt. Dieser hat das folgende Ergebnis erzielt:
+\texttt{wpscan} ist ein Open-Source-Tool, das für die Sicherheitsüberprüfung von 
+WordPress-Websites entwickelt wurde, um potenzielle Schwachstellen und Sicherheitsprobleme in 
+diesen zu erkennen. Es ermöglicht die Analyse von WordPress Webseiten nach eigenen Angaben 
+seit über 10 Jahren und hat in dieser Zeit einen Katalog von mehr als 43000 Core, Plugin oder 
+Theme Schwachstellen aufgebaut (Quelle \url{https://wpscan.com}). Ein Scan einer Webseite ist 
+direkt von der \url{wpscan.com} Webseite möglich, oder vom eigenen Client aus mit dem CLI 
+Programm \texttt{wpscan}.
+
+Man bietet neben der kostenfreien Version für ``Researcher'' mit maximalen 25 API Calls pro 
+Tag, auch eine nicht kostenlose Enterprise Version an.
+
+Die Nutzung ist dabei gänzlich einfach. Der einfache Quick Scan wird durch den Aufruf 
+\texttt{wpscan –url <webseite>} aufgerufen. Dabei erhielten wir jedoch statt eines 
+Ergebnisses eine Fehler-Meldung über einen 403-Status-Code. (Siehe Abb. 12)
 
@@
 
-Nun musste die flag angehängt werden
+Der hier referenzierte HTTP-Statuscode bedeutet, dass der Server den Zugriff auf die 
+angefragte Seite verweigert. Dies kann auf fehlende Zugriffsberechtigungen oder eine Web 
+Application Firewalls (WAF) hinweisen. Eine WAF soll Websites und Webanwendungen vor 
+böswilligen Anfragen und Zugriffen schützen, indem sie den Datenverkehr zwischen dem Client 
+und dem Webserver analysiert, um Angriffe wie SQL-Injection, Cross-Site Scripting (XSS) und 
+Distributed Denial of Service (DDoS) zu erkennen und zu blockieren. Sie verwendet dazu Regeln 
+und Signaturen und ist ein wichtiger Bestandteil um die Sicherheit einer Webseite zu 
+gewährleisten und Angriffe auf Anwendungsebene abzuwehren.
+
+Der Parameter \texttt{--random-user-agent} aus dem gegebenen Hinweis versucht diesen 
+Sicherheitsmechanismus zu umgehen, indem automatisch zufällige Benutzer-Agenten für jede 
+HTTP-Anfrage, die es während des Scans an die WordPress-Website sendet, generiert um diese 
+Anfragen weniger auffällig und weniger vorhersehbar zu gestalten um nicht automatisiert zu 
+wirken. Da wir kein Bot-Netz oder ähnliches verwenden, ist die IP-Adresse der Anfragen jedoch 
+die gleiche. Dies schien jedoch nicht von der Firewall verhindert zu werden, denn mit jenem 
+Parameter lässt sich ein Scan durchführen. (Siehe Abb. 13)
 
@@
 
+In einer detaillierten Ansicht erhalten wir Informationen ausgegeben, die durch den Scan 
+gewonnen wurden konnte. (Siehe Abb. 14)
+
+\begin{itemize}
+	\item Ein Apache Webserver mit der PHP Version 8.1.24 liefert die Webseite aus (diese ist aktuell – Release Datum 28.09.2023)
+	\item WordPress ist installiert in der Version 6.3.1 (ebenfalls zum Zeitpunkt des Scans 
+		aktuell mit Release Datum 29.08.2023). Ein RSS Feed ist vorhanden unter angegebener 
+		URL.
+	\item \texttt{robots.txt} (Anweisungen/Berechtigungen für Web-Crawler von Suchmaschinen) 
+		und Wordpress-Readme sind vorhanden – Ausgabe mit Pfad
+	\item WP-Cron ist aktiviert – dies lässt eine automatisierte Aktualisierung der Software-Komponenten der Wordpress-Installation annehmen
+	\item twentysixteen ist das installierte Wordpress-Theme (ebenfalls zum Zeitpunkt des Scanns aktuell mit Release Datum 29.03.2023).
+Eine eigene Readme ist unter der angegebenen URL zu erreichen.
+Eine kurze Description und eine Style URI sind auslesbar.
+	\item Ein Plugin-Scan mit passiver Scan-Methode findet keine installierten Plugins.
+	\item Es werden keine scanbaren Config-Backups gefunden
+\end{itemize}
+
 \begin{figure}[H]
@@
 
-Hieraus lassen sich folgende Informationen ableiten: XYZ
+Da wir, wie in der Aufgabenstellung beschrieben, keine Registrierung beim \texttt{wpscan}
+Anbieter durchführen und daher keinen API-Key besitzen, endet unser Scan an dieser Stelle.
 

### 2025-01-02 module/netzwerk-sicherheit/zesa-1/main.tex
Restructure repo

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}
\usepackage{url}
\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Gruppe 11 – Mathias Baumbach \& Mara Schulke}}
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang IT Sicherheit \\
		Fachbereich Informatik und Medien \\
		Netzwerksicherheit \\
		Prof.\ Dr.\ Michael Pilgermann
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Zusatz-Einsendeaufgabe 1}\\[0.5em]
		\large{Wintersemester 2023}\\[0.25em]
		\large{\today}
	\end{center}

	\vfill

	\begin{flushright}
		Gruppe 11 \\
		Mathias Baumbach (Matr-Nr. 20213703) \\
		Mara Schulke (Matr-Nr. 20215853)
	\end{flushright}
\end{titlepage}

\begin{abstract}
\end{abstract}

\tableofcontents

\listoffigures

\newpage

\section{Introduction}

Firewalls are one of the basic, yet very effective security controls in network 
security. Packet filters is the entry level firewall technology.

The experiment shall secure a topology with three network segments, which are each 
connected to a unique port at a firewall:

\begin{itemize}
    \item Client-LAN: Client network (1 admin client, 1 user client; both Ubuntu Linux)
    \item Server-LAN: Server network (1 server acting as web server; Ubuntu Server)
    \item WAN: External network
\end{itemize}

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/01}
	\centering
	\caption{Netztopologie}
\end{figure}

The topology with the firewall (pfSense), three Ubuntu boxes (2 clients and one 
server) and the external connection has been prepared in the virtual lab environment 
already. However, the ruleset of the firewall is empty; the exercise is about 
developing a ruleset based on the “white listing” paradigm, implementing it on the 
firewall and checking connectivity.

The following communication shall be allowed (including corresponding reply traffic):

\begin{itemize}
    \item All endpoints (server as well as clients) shall connect to the external net for HTTP / HTTPS traffic on the corresponding ports.
    \item The admin client shall connect to the server via SSH.
\end{itemize}

\section{Preparation}

\subsection{Connect to lab (VNL) and initialize systems}

Connect to the Virtual Network Security Lab (VNL) of THB using your web browser as 
documented (incl. group credentials).

Navigate to shared space in VNL and create a copy of the lab \path{Shared/
Netzwerksicherheit/2023_exercise_firewalls.unl}\\
and place the copy in your user folder. Boot up the systems in your lab – make sure, 
to always start with the pfSense-Firewall, as this one acts as DHCP-Server for the 
clients and the server. When booting the systems for the first time, some initial 
configuration has to be applied – check \textbf{appendix 1 – startup configurations} 
at the end of this document for details.

\subsection{Check communication between components}

Open up the user interface for each component (command line for servers, remote 
desktop for clients) inside your web browser; login using the eve-ng default 
credentials.

Check connectivity internally (web server via IP) and to the outside world (e.g. web 
browser on GUI environments or wget on command line). Document your network topology 
in a corresponding plan including the IP addresses.

\subsubsection*{Antwort}

Zur Dokumentation der Netztopologie haben wir unsere grafische Darstellung der 
Netzwerktopologie um die IP Adressen erweitert:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/02}
	\centering
	\caption{Dokumentierte Netztopologie mit IP-Adressen}
\end{figure}

\vspace{0.5em}

\subsection{Access your group firewall}

Using the remote desktop of the admin client, open firewall web gui in the web 
browser. Get a very first impression of the ruleset and familiarize yourself with the 
handling.

\section{Execution}

Having tested, that the lab environment works fine including all communication, next 
action is to step-by-step restrict the firewall ruleset.

\subsection{Part 1: Effective firewall rules}

Prepare a table for documentation of your firewall policy. Include at least: Type 
(l4), source IP, source port, destination IP, destination port, action. This policy / 
documentation shall always be updated before applying changes to the firewall 
ruleset.

\subsubsection*{Antwort}

\vspace{5mm}

Unsere Tabelle der pfSense Firewall Policies ist bereits mit einigen Standard-Regeln, 
die pfSense initial anlegt, gefüllt:

\begin{center}
    %ghetto fix um table auf page zu bekommen 
    \resizebox{\textwidth}{!}{\begin{tabular}{||c c c c c c c c||} 
         \hline
            VLAN    & Type (I4)  & Source IP  & Source Port & Destination IP& Destination Port & Action & Beschreibung  \\ [0.3ex] 
         \hline\hline
            LAN  & Any   & Any  & Any  & LAN Address   & 443 \& 80  & allow  & Anti-Lockout Rule  \\ 
         \hline
            LAN & Any & EasyRuleBlock HostVLAN   & Any  & Any  & Any & block & Durch initiale Configuration angelegt  \\ 
         \hline
            LAN & IPv4/IPv6 & LAN & Any  & Any  & Any & allow & Default allow LAN to any rule  \\ 
         \hline
            SERVER\_LAN & IPv4/IPv6 & SERVER\_LAN & Any  & Any  & Any & allow & Default allow SERVER\_LAN to any rule  \\ 
         \hline
    \end{tabular}}
\end{center}

\subsubsection{Implement white listing for outgoing traffic}

The default behavior of the firewall for outgoing traffic (“black listing”) shall be 
inversed. This is achieved by changing the default rule. The default rule of a 
ruleset will always apply, if no other rule matched the traffic before. The 
configuration needs to be applied to the 2 LAN interfaces.

\subsubsection*{Antwort}

Die default deny Regeln im pfSense Menü:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/03}
	\centering
	\caption{Lan VLAN Default Rule}
\end{figure}

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/04}
	\centering
	\caption{Server\_LAN VLAN Default Rule}
\end{figure}

Unsere Tabelle der pfSense Firewall Policies mit den veränderten default Regeln:

\begin{center}
    %ghetto fix um table auf page zu bekommen 
    \resizebox{\textwidth}{!}{\begin{tabular}{||c c c c c c c c||} 
         \hline
            VLAN & Type (I4)  & Source IP  & Source Port & Destination IP& Destination Port & Action & Beschreibung\\ [0.3ex] 
         \hline\hline
            LAN & Any & Any  & Any  & LAN Address & 443 \& 80  & allow  & Anti-Lockout Rule\\ 
         \hline
            LAN & Any & EasyRuleBlock HostVLAN & Any  & Any  & Any & block & Durch initiale Konfiguration angelegt\\ 
         \hline
            LAN & IPv4/IPv6 & LAN & Any  & Any  & Any & deny & Default deny LAN to any rule\\ 
         \hline
            SERVER\_LAN & IPv4/IPv6 & SERVER\_LAN & Any  & Any  & Any & deny & Default deny SERVER\_LAN to any rule\\ 
         \hline
    \end{tabular}}
\end{center}

\subsubsection{Allow for http/https traffic}

Add additional rules for allowing http/https traffic from all clients / servers. 
Which ports are required? TCP or UDP? Adopt your firewall policy before applying 
changes to the firewall.

Test your firewall configuration by web browsing from within your ubuntu client.

\subsubsection*{Antwort}

Bevor man den Datenverkehr auf Port 80 \& 443 erlaubt für http/https, erhält man die 
folgende Meldung beim Aufruf unseres Webservers:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/05}
	\centering
	\caption{Fehlgeschlagener Aufruf des Webservers via IP-Adresse}
\end{figure}

Um diese Fehlermeldung zu umgehen sind folgende zusätzliche Firewall Regeln nötig:

\begin{center}
    %ghetto fix um table auf page zu bekommen 
    \resizebox{\textwidth}{!}{\begin{tabular}{||c c c c c c c||} 
         \hline
            VLAN & Type (I4)  & Source IP  & Source Port & Destination IP& Destination Port & Action \\ [0.3ex] 
         \hline\hline
            LAN & TCP & LAN net VLAN & Any & Any & 80 (http) & allow\\ 
         \hline
            LAN & TCP & LAN net VLAN & Any & Any & 443 (https) & allow\\ 
         \hline
            SERVER\_LAN & TCP & SERVER\_LAN net VLAN & Any & Any & 80 (http) & allow\\ 
         \hline
            SERVER\_LAN & TCP & SERVER\_LAN net VLAN & Any & Any & 443 (https) & allow\\ 
         \hline
    \end{tabular}}
\end{center}

Nun kann bei erneutem Aufruf der IP unseres Webservers erfolgreich die Webseite 
abgerufen werden:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/06}
	\centering
	\caption{Erfolgreicher Aufruf des Webservers via IP-Adresse}
\end{figure}

\subsubsection{Which additional traffic is required?}

You should experience issues when browsing the web.

Discuss: Which additional traffic is required? Which changes need to be applied to 
your firewall policy? Adopt and test again.

\subsubsection*{Antwort}

Unser Webserver war in der vorherigen Aufgabe erreichbar, weil der Aufruf per IP-
Adresse erfolgte. Für Zugriff auf Webseiten im Internet ist die Anfrage an einen DNS 
Server nötig, der die URL in eine IP Adresse für den Browser übersetzt.

Da wir bisher keine Regel für DNS Traffic angelegt haben, schlägt ein Aufruf einer 
URL fehl:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/07}
	\centering
	\caption{Fehlgeschlagener Aufruf eines Webservers via URL}
\end{figure}

Es sind folgende zusätzliche Firewall Regeln nötig:

\begin{center}
    %ghetto fix um table auf page zu bekommen 
    \resizebox{\textwidth}{!}{\begin{tabular}{||c c c c c c c||} 
         \hline
            VLAN    & Type (I4)  & Source IP  & Source Port & Destination IP& Destination Port & Action \\ [0.3ex] 
         \hline\hline
            LAN  & TCP/UDP   & LAN net VLAN & Any  & Any   & 53 (dns)  & allow  \\ 
         \hline
            SERVER\_LAN  & TCP/UDP & LAN net VLAN  & Any  & Any   & 53 (dns)  & allow  \\ 
         \hline
    \end{tabular}}
\end{center}

Dadurch können nun Webseiten-Aufrufe im Internet erfolgreich übersetzt werden:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/08}
	\centering
	\caption{Erfolgreicher Aufruf eines Webservers via URL}
\end{figure}

Darüber hinaus gäbe es noch weitere Ports, deren Freigabe nützlich wäre. Beispiel 
wären WebRTC (Port-Range 48k bis 65k?) oder IMAP (143).

\subsubsection{Secure admin access}

SSH access to the server should be restricted to the admin client; all other traffic 
between those two networks shall be prohibited.

\begin{itemize}
    \item Add a rule for allowing access to the webserver (HTTP / HTTPS) from any endpoint in the client LAN. 
    \item Add a rule for allowing admin access to the webserver (SSH) from the admin client. 
\end{itemize}

Test the communication.

\subsubsection*{Antwort}

Die default deny Regel verhindert den SSH Zugang vom Admin- und vom User-Client zum Server:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/09}
	\centering
	\caption{SSH Verbindung vom Admin-Client zum Server}
\end{figure}

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/10}
	\centering
	\caption{SSH Verbindung vom User-Client zum Server}
\end{figure}

Es ist folgende zusätzliche Firewall Regeln nötig:

\begin{center}
    %ghetto fix um table auf page zu bekommen 
    \resizebox{\textwidth}{!}{\begin{tabular}{||c c c c c c c||} 
         \hline
            VLAN    & Type (I4)  & Source IP  & Source Port & Destination IP& Destination Port & Action \\ [0.3ex] 
         \hline\hline
            LAN  & TCP   & 192.168.1.100 & Any  & SERVER\_LAN net VLAN  & 22 (SSH)  & allow  \\ 
         \hline
    \end{tabular}}
\end{center}

Nach Aktivierung dieser Regel ist der SSH Zugriff weiterhin für den User-Client 
geblockt, für den Admin-Client aber möglich:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/11}
	\centering
	\caption{SSH Verbindung vom Admin-Client zum Server mit SSH Regel}
\end{figure}

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/12}
	\centering
	\caption{SSH Verbindung vom User-Client zum Server mit SSH Regel}
\end{figure}

\subsection{Part 2: Analyze firewall audit trail}

Apply changes to your firewall ruleset, so that events are being logged whenever 
anybody accesses the server on port 22 (no matter whether it is successful or not).

Trigger those firewall rules by starting SSH connection to the serves from

\begin{itemize}
    \item 1) the admin client and 
    \item 2) the “normal” client. 
\end{itemize}

Observe the events in the audit trail on the firewall.

\subsubsection*{Antwort}

Für diese Aufgabe müssen zwei Regeln aus den vorherigen Aufgaben erweitert werden:

\begin{center}
    %ghetto fix um table auf page zu bekommen 
    \resizebox{\textwidth}{!}{\begin{tabular}{||c c c c c c c||} 
         \hline
            VLAN & Type (I4) & Source IP & Source Port & Destination IP& Destination Port & Action \\ [0.3ex] 
         \hline\hline
            LAN & TCP & 192.168.1.100 & Any & SERVER\_LAN net VLAN & 22 (SSH) & allow\\ 
         \hline
            LAN & IPv4/IPv6 & LAN net VLAN & Any  & Any  & Any & deny\\ 
         \hline      
    \end{tabular}}
\end{center}

An beiden Regeln muss die Extra Option ''Log packets that are handled by this rule`` 
angehakt werden.

Danach kann in den System Logs der pfSense (Status / System Logs / Firewall / Dynamic 
View) der erfolgreiche und erfolglose ssh Verbindungsversuch verfolgt werden:

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/13}
	\centering
	\caption{Erfolgreiche SSH Verbindung vom Admin-Client}
\end{figure}

In der 2. Zeile (grüner Haken) kann man den erfolgreichen Zugriff von 192.168.1.100 
Port 52862 auf 192.168.2.100 Port 22 beobachten.

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/14}
	\centering
	\caption{Erfolglose SSH Verbindung vom User-Client}
\end{figure}

Ebenfalls in der 2. Zeile ist der fehlgeschlagene Zugriff von 192.168.1.10 Port 50026 
auf 192.168.2.100 Port 22 dokumentiert.

\section{References / Documentation}

\begin{itemize}
    \item 1) Eve-ng cookbook: \path{https://www.eve-ng.net/index.php/documentation/professional-cookbook/} 
    \item 2) pfSense Firewall configuration documentation: \path{https://docs.netgate.com/pfSense/en/latest/firewall/index.html}
\end{itemize}

\end{document}

### 2025-01-02 digitaler-selbstschutz/hausarbeit/hausarbeit.tex
Tidy repo structure again
This reverts commit 11628c78764cda52ff8ff7004d322ebae5713921.

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{float}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Online Studiengang IT Sicherheit \\
		Fachbereich Informatik und Medien \\
		Digitaler Selbstschutz \\
		Prof.\ Dr.\ Michael Pilgermann
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Hausarbeit: Schutz des THB-Webmail-Accounts mit einem 2. Faktor}\\[0.5em]
		\large{Wintersemester 2022}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{flushright}
		Mara Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\begin{abstract}
	Durch die Implementierung von 2-Faktor-Authentifizierung können bestehende
	Systeme hinsichtlich ihrer Sicherheit stark verbessert werden und neue
	Systeme grundlegend neue Authentifizierungskonzepte implementieren. Die
	Unterstützung der Industrie zeigt sich durch die Implementierung durch große
	Firmen. Durch die Verwendung von ausgereiften
	Bibliotheken für die Generierung und Validierung von
	\textit{One-Time-Passwords} kurz \textit{OTP} sind schnelle und
	zuverlässige Implementierungen des Standards leicht realisierbar.
\end{abstract}

\tableofcontents

\listoffigures

\newpage

\section{Einleitung}

Authentifizierung ist eines der größten Probleme, die durch verteilte Systeme
entstehen. Es gibt zahlreiche Möglichkeiten, die Identität einer Gegenseite
sicherzustellen allerdings weisen viele von ihnen Schwachstellen hinsichtlich
Man-In-The-Middle-Attacken und basieren auf der Annahme, dass das System des
Nutzers nicht kompromittiert wurde. Als eine mögliche Lösung für viele dieser
Probleme hat sich die 2-Faktor-Authentifizierung innerhalb der letzten Jahre zu
einem beliebten Standard entwickelt, der von vielen großen Anbietern wie
Facebook, Amazon, Google, Mozilla, Microsoft etc.\ bereits unterstützt
wird.

\subsection{Welche Probleme löst die Zwei-Faktor-Authentifizierung?}

Die Notwendigkeit für einen solchen Standard hat sich in den letzten Jahren
immer stärker gezeigt, da die klassische Passwort-Authentifizierung (1-Faktor)
durch zunehmende Rechenleistung und effizientere Angriffe immer unsicherer und
unhandlicher für Nutzer wird. Die minimalen Passwortlängen steigen
dementsprechend an und führen zur Wiederverwendung von gleichen Login-Daten für
mehrere Online-Dienste. Bekannte Lösungen sind die Verwendung von sogenannten
Passwort-Managern um lange und zufällige Passwörter für verschiedenste
Online-Dienste zu verwenden, ohne dass sich Nutzer diese merken müssen. Solche
Passwort-Manager sind zwar eine Lösung für die sichere Aufbewahrung von langen
Passwörtern, können aber nicht die grundsätzlich durch
Passwort-Authentifizierung eröffneten Angriffsvektoren wie zum Beispiel
Man-In-The-Middle-Attacken. So bald ein Angreifer in den Besitz des geheimen
Wissens (in diesem Fall das Passwort) gelangt, kann dieser uneingeschränkt und
unbegrenzt oft auf das Zielsystem zugreifen, bis der Nutzer seine Daten
ändert (vorausgesetzt, der Angreifer hat dies noch nicht getan).

Durch den Wechsel von einer reinen Passwort-Authentifizierung auf
Zero-Knowledge-Proof basierte Authentifizierungsmethoden (in diesem Fall
One-Time-Passwords) lassen sich ganze Angriffsvektoren ausschließen,
da ein kompromittierter Server oder eine kompromittierte Verbindung niemals das
gesamte geheime Wissen des Nutzers einem Angreifer zugänglich machen. Das
heißt, dass sich ein Angreifer im Falle einer kompromittierten Verbindung
maximal in die Sitzung des Nutzers einschleichen könnte, allerdings bei der
nächsten Authentifizierung nicht erneut die Identität des Nutzers beweisen
könnte und somit den Zugriff verlieren würde.

Im Falle der Zwei-Faktor-Authentifizierung kennt nicht einmal das System des
Nutzers das Geheimnis um valide One-Time-Passwörter zu generieren, da
dieses auf einem externen Gerät (in der Regel das Smartphone) verwaltet werden
und der Beweis des Besitzes durch die Eingabe des OTP erfolgt. So stellt selbst
ein kompromittiertes Nutzersystem nur eine temporäre Schwachstelle dar, da
immer nur die jeweiligen OTP abgegriffen werden können, aber nie das Geheimnis
zur Generierung dieser.

\section{Einrichtung einer 2-Faktor-Authentifizierung bei Zimbra}

Die Einrichtung einer App zur 2-Faktor-Authentifizierung ist dank des
Einreichtungsassistenten unkompliziert und weitgehend nutzerfreundlich. Die
Einrichtung kann nach erfolgreichem Login unter der Einstellungsseite im Reiter
``Accounts'' vorgenommen werden. (Abb. 1)

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{images/01-einstellung.png}
	\centering
	\caption{Einstelungen Zimbra}
\end{figure}

Zu Beginn der Einrichtung werden wichtige Hinweise zur
2-Faktor-Authentifizierung angezeigt und dem Nutzer wird die Funktionsweise
erklärt. (Abb. 2)

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{./images/02-beginn.png}
	\centering
	\caption{Beginn der Einrichtung}
\end{figure}

Anschließend findet eine erneute Passwortabfrage statt, um die Identität des
Nutzers sicherzustellen, bevor eine eventuell zugriffsgefährendende Aktion (z.B.
bei Verlust des Smartphones und der Recovery-Codes) durchgeführt wird. (Abb. 3)

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{./images/03-passwort.png}
	\centering
	\caption{Passwortabfrage}
\end{figure}

Nun wird das Geheimnis zur Generierung von One-Time-Passwords angezeigt. Dieses
muss nun in einen kompatiblen Authenticator eingegeben werden. (Abb. 4)

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{./images/04-secret.png}
	\centering
	\caption{2-Faktor-Secret}
\end{figure}

Für meine persönliche Einrichtung habe ich mich für Dashlane entschieden, da
die One-Time-Passwords sehr übersichtlich direkt in dem Account-Eintrag
innerhalb des Passowrt-Managers auf dem Smartphone angezeigt werden. Es muss
nun das Geheimnis aus Abbildung 4 in den Authenticator eingetragen werden.
(Abb. 5)

\begin{figure}[H]
	\includegraphics[height=0.25\textheight]{./images/05-eingabe-secret.png}
	\centering
	\caption{Eingabe des Secrets in Dashlane}
\end{figure}

Wenn die Eingabe erfolgreich war, bestätigt Dashlane dies und fängt sofort mit
der Generierung von 2-Faktor-Codes an. (Abb. 6\&7)

\begin{figure}[H]
	\includegraphics[height=0.25\textheight]{./images/06-erfolg.png}
	\centering
	\caption{Erfolg in Dashlane}
\end{figure}

\begin{figure}[H]
	\includegraphics[height=0.25\textheight]{./images/07-code.png}
	\centering
	\caption{Generierter 2-Faktor-Code}
\end{figure}

Zimbra fordert nun den Nutzer auf, die erfolgreiche Einrichtung des
Authenticators zu bestätigen, in dem ein OTP angefordert wird. Die ist wichtig,
um zu verhindern, dass der Nutzer ein falsches Secret in seinen Authenticator
übertragen hat, das zwar auch OTPs generiert, die allerdings nicht für dieses
Secret gültig sind. (Abb. 8)

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{./images/08-bestaetigung.png}
	\centering
	\caption{Bestätigung des Authenticators}
\end{figure}

Wenn das Geheimnis korrekt übertragen wurde und der Authenticator funktioniert
und gültige One-Time-Passwords ausgibt, zeigt Zimbra nun eine Bestätigung der
erfolgreichen Einrichtung an. (Abb. 9)

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{./images/09-erfolg-zimbra.png}
	\centering
	\caption{Erfolgreiche Einrichtung}
\end{figure}

Bevor nun mit der alltäglichen Nutzung fortgefahren wird, ist es wichtig, die
Recovery-Codes an einem sicheren Ort ausserhalb des für 2FA verwendenten Gerätes
abzulegen. Sollten die Recovery-Codes auf dem gleichen Gerät abgelegt werden,
sind sie im Falle eines Verlusts nicht greifbar und somit nutzlos zur
Wiederherstellung. (Abb. 10) In meinem Beispiel habe ich diese in die Notizen des
THB-Eintrags innerhalb meines Passwort-Managers gespeichert, da ich diese auch
bei einem eventuellen Verlust meines Smartphones erreichen könnte.

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{./images/10-recovery-codes.png}
	\centering
	\caption{Ausgabe der Recovery-Codes}
\end{figure}

Nach der erfolgreichen Einrichtung kann der zweite Faktor nun erneut getestet
werden, indem sich der Nutzer einmal ausloggt und erneut anmeldet. Nun sollte
Zimbra nach erfolgreicher Eingabe der Login-Daten ein One-Time-Password
abfragen, welches aus dem Authenticator übertragen werden kann. (Abb. 11)

\begin{figure}[H]
	\includegraphics[width=0.75\textwidth]{./images/11-login-2fa.png}
	\centering
	\caption{Login bei Zimbra nach Einrichtung}
\end{figure}

\section{Erfahrungsbericht}

Die Einrichtung einer 2-Faktor-Authentifizierung ist im Allgemeinen ein
vergleichsweise kleiner Aufwand für die Sicherheitsgarantien, die diese bietet.
Durch die gängigen Anwendungen für Mobilgeräte (beispielsweise Google
Authenticator, Authy etc.) gestaltet sich die Einrichtung im Rahmen eines
Regestrierungsprozesses und die Anwendung im Alltag in der Regel
nutzerfreundlich (zum Beispiel durch die Verwendung von QR-Codes als
Einrichtungshilfe).

Im Gegensatz zur reinen wissensbasierten Authentifizierungsmethode
(üblicherweise Passwörter) birgt die 2-Faktor-Authentifizierung (insbesondere
TOTP) das Risiko des Verlustes des Endgerätes. Der Nutzer kann somit trotz
berechtigter Absicht (Zugriff auf den eigenen Account) nicht beweisen, dass es
sich um die richtige Person handelt. Als Lösung werden oftmals Recovery-Codes
eingesetzt, die der Nutzer an einem Ort außerhalb des für die
2-Faktor-Authentifizierung verwendenten Gerätes speichern. Diese bieten jeweils
einmaligen Zugriff auf den Account, um beispielsweise ein neues Gerät bei
Verlust des alten zu konfigurieren. Die sichere verwarung von Recovery-Codes
ist zwingend erforderlich, um die Sicherheitsgarantien der
2-Faktor-Authentifizierung zu bewahren, da beispielsweise bei einer
ungeschützten Aufbewahrung auf einem Schreibtisch im Großraumbüro die
Möglichkeit eines Angriffes besteht. Dieser ist zwar immer noch im Vergleich
zur reinen wissensbasierten Authentifizierung erschwert, da ein Angreifer in
dem gerade genannten Beispiel physischen Zugriff auf den Schreibtisch
erlangen müsste, aber dennoch deutlich vereinfacht im Vergleich zu einem
Brute-Force-Angriff auf 2-Faktor-Geschützte Accounts.

Im Falle von Zimbra ist die Einrichtung einerseits verständlich und
nutzerfreundlich gestaltet, sobald der Einreichtungsassistent geöffnet ist,
allerdings wäre die Verwendung eines QR-Codes aus Nutzersicht empfehlenswert,
da dies das mühsame Abtippen des geheimen TOTP-Schlüssels in die
Authenticator-App abkürzen würde. Sobald die Einrichtung abgeschlossen ist,
läuft alles in einer zu anderen Anwendungen vergleichbaren Weise ab (Zeitpunkt der
TOTP-Aufforderung, Recovery-Codes, Entfernen vom Authenticator etc.). Somit
kann festgestellt werden, dass die Einrichtung der 2-Faktor-Authentifizierung
bei richtiger Anwendung eine deutlich höhere Sicherheit bietet und bei
geringer Beeinträchtigung des Nutzererlebnisses verpflichtend eingeführt werden
sollte.

\end{document}

### 2025-01-02 entwicklung-sicherer-software-systeme/exercise-2/exercise-2.tex
Tidy repo structure again
This reverts commit 11628c78764cda52ff8ff7004d322ebae5713921.

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{float}
\usepackage{tabularx}
\usepackage{pdflscape}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\begin{document}

\begin{titlepage}
	\begin{flushleft}
		TH Brandenburg \\
		Studiengang IT Sicherheit \\
		Fachbereich Informatik und Medien \\
		Entwicklung Sicherer Softwaresysteme \\
		Prof. Dr.-Ing. Martin Schafföner
	\end{flushleft}

	\vfill

	\begin{center}
		\Large{Einsendeaufgabe: Nichtfunktionale Sicherheitsanforderungen}\\[0.5em]
		\large{Sommersemester 2024}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{flushright}
		Mara Schulke \\
		Matrikel-Nr. 20215853
	\end{flushright}
\end{titlepage}

\tableofcontents

\listoffigures

\section{Einführung}

In der vorherigen Einsendeaufgabe wurde die Online-Meeting-Plattform Google Meet 
analysiert und ausgewertet. Die Sicherheitsbegutachtung von Google Meet wird im Rahmen 
dieser Aufgabe fortgeführt und um Sicherheits-Akzeptanzkriterien bzw. 
sicherheitsspezifische Anforderungen erweitert. Außerdem werden die verschiedenen 
Vorgehensweisen verglichen und es werden Testszenarien definiert, mittels derer die 
Sicherheits-Akzeptanzkriterien verifiziert werden können.

\section{Funktionale Anforderungen}

Da die funktionalen Anforderungen im Rahmen der letzten Einsendeaufgabe noch nicht konkret 
benannt worden sind, ist der erste Schritt um sinnvolle Sicherheitsanforderungen zu 
definieren die Benennung der funktionalen Anforderungen an Google Meet. Im konkreten 
(angenommen, es wird nur der grundlegende Umfang einer Meeting-Plattform abgedeckt) 
handelt es sich dabei um die folgenden Anforderungen:

\begin{itemize}
	\item \textbf{Benutzerauthentifizierung:} Google Meet muss es Benutzern ermöglichen, sich zu registrieren, anzumelden und abzumelden.
	\item \textbf{Meeting-Planung:} Google Meet muss es den Benutzern ermöglichen, Besprechungen mit bestimmten Teilnehmern zu planen.
	\item \textbf{Video- und Audio-Streaming}: Google Meet muss Video- und Audiostreaming in Echtzeit unterstützen.
	\item \textbf{Chat-Funktionalität:} Google Meet muss Textchats in Echtzeit während der Sitzungen unterstützen.
	\item \textbf{File-Sharing:} Google Meet muss es den Nutzern ermöglichen, während einer Besprechung Dateien gemeinsam zu nutzen bzw. zu teilen.
	\item \textbf{Bildschirmfreigabe:} Google Meet muss es den Nutzern ermöglichen, ihren Bildschirm während einer Besprechung gemeinsam zu nutzen.
\end{itemize}

\section{Nichtfunktionale Sicherheitsanforderungen}

Die nichtfunktionalen Sicherheitsanforderungen an Google Meet bauen auf den funktionalen 
Anforderungen auf und decken die zugrundeliegenden Sicherheitsmechanismen ab. Sie benennen 
die konkreten Aspekte die Software-Entwickler bei der Umsetzung der obigen Anforderungen 
in einem System berücksichtigen müssen. Bei der Formulierung von nichtfunktionalen 
Sicherheitsanforderungen stehen einige Möglichkeiten zur Verfügung:

\subsection{Verschiedene Vorgehensweisen}

\subsubsection{Evil User Stories}

Evil User Stories sind ein Mittel, um potenzielle Sicherheitsbedrohungen zu 
identifizieren, indem man aus der Perspektive eines böswilligen Nutzers (bzw. Angreifers) 
denkt. Diese Methode hilft Software-Entwicklern, Schwachstellen im System zu erkennen und 
entsprechende Gegenmaßnahmen zu planen.
Diese folgen in der Regel der Struktur:

\begin{quote}
``Als [böswilliger Akteur] möchte ich [schädliche Aktion], um [negativen Effekt] zu erreichen.''	
\end{quote}

\begin{flushleft}
Ein Beispiel für diese Vorgehensweise könnte sein:
\end{flushleft}

\begin{quote}
``Als Angreifer möchte ich unverschlüsselte Daten abfangen, um vertrauliche Informationen zu stehlen.''
\end{quote}

\subsubsection{Sicherheits-Akzeptanzkriterien}

Sicherheits-Akzeptanzkriterien sind spezifische, messbare Bedingungen, die erfüllt sein 
müssen, damit ein Sicherheitsfeature als erfolgreich implementiert gilt. Diese Kriterien 
helfen dabei, sicherzustellen, dass die Sicherheitsanforderungen korrekt und vollständig 
umgesetzt wurden.

\begin{flushleft}
Ein Beispiel für diese Vorgehensweise könnte sein:
\end{flushleft}

\begin{quote}
``Alle Datenübertragungen müssen mit mindestens mit AES-256 verschlüsselt sein.''
\end{quote}

\subsubsection{Sicherheitsspezifische Anforderungen}

Sicherheitsspezifische Anforderungen sind detaillierte, konkrete Anforderungen, die 
Sicherheitsmechanismen und -verfahren beschreiben, um das System vor Bedrohungen zu 
schützen. Diese Anforderungen sind spezifisch und oft technischer Natur.

\begin{flushleft}
Ein Beispiel für diese Vorgehensweise könnte sein:
\end{flushleft}

\begin{quote}
``Implementiere HMAC für alle Nachrichten, um ihre Integrität sicherzustellen''
\end{quote}

\newpage

\begin{landscape}

\subsection{Auflistung der nichtfunktionale Sicherheitsanforderungen}

\begin{tabularx}{20cm}{|l|X|X|}
    \hline
    & \vspace{0em} \textbf{Sicherheits-Akzeptanzkriterien} \vspace{0.5em}
    & \vspace{0em}\textbf{Sicherheitsspez. Anforderungen} \vspace{0.5em} \\
    \hline
        \textbf{Authentifizierung}
        & Die Verfahren zur Benutzerauthentifizierung müssen einen Penetrationstest 
          bestehen, um die Robustheit der Multi-Faktor-Authentifizierung zu überprüfen.
        & Alle Authentifizierungen müssen durch eine Multi-Faktor-Authentifizierung 
    	  geschützt sein.\vspace{0.5em}

    	  Passwörter müssen mit einer dem Industriestandard entsprechenden Verschlüsselung 
    	  (z. B. bcrypt) gespeichert werden.
        \\
    \hline
        \textbf{Datenübertragung}
        & Alle Datenübertragungen müssen mit einem Netzwerkprotokoll-Analysator validiert 
          werden, um sicherzustellen, dass die Verschlüsselungsstandards eingehalten 
          werden.
        & Alle zwischen den Clients und dem Server übertragenen Daten
          müssen mit TLS 1.2 oder höher verschlüsselt werden.
        \\
    \hline
        \textbf{Zugangskontrolle}
        & Die Zugriffskontrollmechanismen müssen anhand rollenbasierter Zugriffsszenarien 
          getestet werden, um eine ordnungsgemäße Durchsetzung sicherzustellen.
          \vspace{0.5em}

          Unerlaubte Zugriffsversuche müssen protokolliert und während eines 
          Audits überprüft werden.
        & Die Nutzer dürfen nur auf Sitzungen zugreifen, zu denen sie ausdrücklich 
          eingeladen wurden.\vspace{0.5em}

		  Es muss eine rollenbasierte Zugriffskontrolle (RBAC) implementiert werden, um 
		  den Zugriff auf Verwaltungsfunktionen zu beschränken.
        \\
    \hline
        \textbf{Datenschutz}
        & Die Verschlüsselung von Besprechungsaufzeichnungen muss durch unbefugte 
          Zugriffsversuche überprüft werden.
        & Sitzungsaufzeichnungen müssen verschlüsselt und sicher gespeichert werden und 
          nur autorisierten Benutzern zugänglich sein.
        \\
    \hline
        \textbf{Protokollierung}
        & Die Protokolle müssen monatlich überprüft werden, um sicherzustellen, dass sie 
          detailliert und gegen Manipulationen geschützt sind.\vspace{0.5em}

		  Automatisierte Tools müssen die Protokolle auf verdächtige Aktivitäten 
		  überwachen, wobei Warnungen innerhalb von 24 Stunden überprüft werden müssen.
        & Jeder Zugriff auf sensible Informationen und jede Aktion muss protokolliert und 
          auf verdächtige Aktivitäten überwacht werden.\vspace{0.5em}

		  Die Protokolle sind vor Manipulationen zu schützen und mindestens ein Jahr lang 
		  sicher aufzubewahren.
        \\
    \hline
        \textbf{Notfallmanagement}
        & Vierteljährlich muss eine Übung zur Reaktion auf Zwischenfälle durchgeführt 
          werden, über die ein Bericht erstellt und auf ihre Wirksamkeit überprüft wird.
        & Google Meet muss über einen Plan zur Reaktion auf Zwischenfälle verfügen, der 
          Verfahren zur Erkennung, Meldung und Behebung von Sicherheitsvorfällen enthält.
        \\
    \hline
\end{tabularx}

\end{landscape}

\newpage

\section{Vergleich und Präferenz}

\textit{Sicherheitsspezifische Anforderungen} bieten Software-Entwicklern klare und 
umsetzbare Schritte zur Implementierung von einzelnen Sicherheitsmaßnahmen. Sie sind 
spezifisch und können leicht in die Entwicklung integriert werden, allerdings kann es 
ihnen an Kontext fehlen und sie können manchmal zu präskriptiv sein, was weniger Raum für 
Flexibilität lässt.

\textit{Sicherheits-Akzeptanzkriterien} bieten den Vorteil, dass sie sich auf die 
Ergebnisse und überprüfbaren Bedingungen konzentrieren. Sie gewährleisten, dass die 
implementierten Maßnahmen wirksam sind, anstatt eine konkrete Implementierung 
vorzuschreiben. Außerdem bieten sie eine Möglichkeit, bzw. ein klares Szenario, diese 
Maßnahmen zu testen.

\vspace{1em}

Ich persönlich bevorzuge den Einsatz von Sicherheits-Akzeptanzkriterien, weil sie einen 
klaren Rahmen für die Überprüfung der Wirksamkeit von Sicherheitsmaßnahmen bieten. Dieser 
Ansatz stellt sicher, dass die Sicherheit nicht nur implementiert, sondern auch getestet 
und validiert wird, wodurch die Sicherheit des Systems besser gewährleistet werden kann 
und das System allgemein robuster wird.

\section{Testszenarien}

Mithilfe von verschiedenen Testszenarien lässt sich eine bessere Sicherheit für das 
Gesamtsystem sicherstellen. Die Verwendung von Sicherheits-Akzeptanzkriterien ist hier 
sehr dienlich, da diese eine konkrete Ableitung eines Testszenarios zulassen.

Um diese Ausarbeitung in einem angemessenen Umfang zu halten habe ich untenstehend zwei 
Szenarien je definierter nichtfunktionaler Sicherheitsanforderung aus dem zweiten 
Abschnitt erarbeitet:

\subsubsection*{Authentifizierung}

\begin{quote}
Nach mehrfachen Versuchen, sich mit falschen MFA-Codes zu authentifizieren muss 
sichergestellt werden, dass die Mechanismen zur Kontosperrung ausgelöst wurden.
\end{quote}

\begin{quote}
Versuch die gespeicherten Passwörter zu knacken, um die Stärke der Verschlüsselung zu 
überprüfen.
\end{quote}

\subsubsection*{Sicherheit der Datenübertragung}

\begin{quote}
Unter Verwendung eines Netzwerk-Analysetools wie z.B. Wireshark – das 
Datenübertragungen abfangen und überprüfen kann – muss verifiziert werden, ob alle Daten 
verschlüsselt sind.	
\end{quote}

\begin{quote}
Versuch die TLS-Version herabzustufen und anschließende Überprüfung, ob das System die 
Verbindung ablehnt.
\end{quote}

\subsubsection*{Zugangskontrolle}

\begin{quote}
Versuchen, auf ein Meeting ohne Einladungslink zuzugreifen. Es muss sichergestellt 
werden, dass der Zugang verweigert wird.
\end{quote}

\begin{quote}
Durchführung eines Tests mit verschiedene Benutzerrollen, um sicherzustellen, dass die 
Berechtigungen korrekt durchgesetzt werden.
\end{quote}

\subsubsection*{Datenschutz}

\begin{quote}
Versuch, unbefugten Zugriff auf verschlüsselte Besprechungsaufzeichnungen zu 
erhalten mit anschließender Übprüfung, dass der Zugriff verweigert wird.
\end{quote}

\subsubsection*{Protokollierung und Überwachung}

\begin{quote}
Generierung von Sicherheitsereignissen (z.B. durch fehlgeschlagene Anmeldeversuche, 
Datei-Uploads usw.) mit anschließender Überprüfung, ob diese korrekt protokolliert 
werden.
\end{quote}

\begin{quote}
Test des Überwachungssystems, indem verdächtige Aktivitäten simuliert werden mit 
anschließender Überprüfung, dass Warnungen generiert und darauf reagiert wird.
\end{quote}

\subsubsection*{Notfallmanagement}

\begin{quote}
Durchführung eines simulierten Sicherheitsverstoßes und anschließende Befolgung des 
Reaktionsplans, um dessen Wirksamkeit sicherzustellen.
\end{quote}

\begin{quote}
Überprüfung des Berichts über die Reaktion auf einen Vorfall, um sicherzustellen, dass 
alle Schritte befolgt wurden, und um Bereiche mit Verbesserungsbedarf zu ermitteln.
\end{quote}

\end{document}

### 2025-01-05 biometrie/group/group.tex
Add group exercises

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Conrad Ferneding, Tobias Ende, Timo Schwabe, Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\newcommand{\figuresource}[1]{
	\begin{center}Quelle: {#1}\end{center}
}

\begin{document}

\begin{titlepage}
	\begin{center}
		\begin{Large}
			Technische Hochschule Brandenburg \\[1em]
		\end{Large}

		IT Sicherheit \\
		Informatik und Medien \\
		Biometrie – Dr.\ Tobias Scheidat
	\end{center}

	\vfill

	\begin{center}
		\Large{Teamaufgabe: Offline Handschrift}\\[0.5em]
		\large{Wintersemester 2024}\\[0.25em]
		\large{Abgabetermin \today}
	\end{center}

	\vfill

	\begin{center}
		Tobias Ende – Matr-Nr. 939628 \\
		Timo Schwabe – Matr-Nr. 20229002 \\
		Conrad Ferneding – Matr-Nr. 20246863 \\
		Mara Schulke – Matr-Nr. 20215853 \\
	\end{center}
\end{titlepage}

\tableofcontents

\listoffigures

\newpage

\section{Wenn Sie das FAR/FFR-Fehlerratendiagramm erstellt haben, können Sie nun eine Abschätzung der entstandenen EER (Equal Error Rate) vornehmen. Tun Sie dies und geben Sie das Ergebnis hierfür an.}

Diese Frage wird die Equal Error Rate (EER) bewerten. Das genutzte biometrische System war als Aufgabe 
vorgegeben worden und wird daher nicht ausführlich erläutert werden. Genutzt wurde die Modalität der 
Offline-Handschrift mit zwei verschiedenen Semantiken.

Die EER ergibt sich aus der False Requection Rate (FRR) und der False Acceptance Rate (FAR). Die FRR 
beschreibt die Anzahl der fehlerhaften Ablehnungen im Verhältnis zu der Gesamtanzahl der Versuche. Die FAR 
beschreibt die fehlerhaften Akzeptierungen im Verhältnis zu der Gesamtanzahl der Versuche. Die Stelle, an 
der die FRR und FAR gleich sind, ergibt die EER, welche die Genauigkeit eines biometrischen Systems 
beschreibt. Ein biometrisches System ist ungenau, wenn die EER hoch ist. Die Anzahl der fehlerhaften 
Akzeptierungen und die Anzahl der fehlerhaften Ablehnungen sind dort hoch. Andererseits deutet eine 
niedrige EER auf ein genaueres biometrisches System, da hier die Anzahl der Fehler niedriger ist. Ob das 
biometrische System einen Versuch akzeptiert oder ablehnt, hängt von der berechneten euklidischen Distanz 
zwischen den Enrolment- und Verifikationsvektoren ab.

\begin{figure}[ht]
	\centering
	\begin{minipage}{0.4\textwidth}
		\centering
		\includegraphics[width=\linewidth]{assets/a-eer}
		\caption{Semantik A - FAR/FRR}
	\end{minipage}%
	\hspace{1em}
	\begin{minipage}{.4\textwidth}
		\centering
		\includegraphics[width=\linewidth]{assets/b-eer}
		\caption{Semantik B - FAR/FRR}
		\label{fig:test2}
	\end{minipage}
\end{figure}

Die EER, welche in Abbildung 1 zu sehen ist, beträgt $\approx22\%$, während die EER aus Abbildung 2,
$\approx2\%$ beträgt. Dass die EER der ersten Semantik höher als die der zweiten Semantik ist, war zu 
erwarten. Bei der ersten Semantik wurden dieselben Schriftzüge geschrieben und bei der zweiten Semantik 
unterschiedlich. Die Merkmalsextraktion bei gleichen Texten lassen weniger Unterschiede zu. Da es sich 
zudem um Zahlen handelte, bei denen z.B. Segmentierung und Schleifen nur begrenzt individualisierbar sind, 
fallen die FAR und die FRR insgesamt höher aus. Wie bereits aus der Aufgabenstellung deutlich wird, 
handelt es sich bei diesen Diagrammen um einen „Zero-Effort“ Fälschungs-Test. (Die Frage, ob diese 
Semantik bereits als blinde Fälschung betrachtet werden kann, wird in den Fragen zwei und vier 
diskutiert.)

Zusätzlich muss an dieser Stelle angemerkt werden, dass lediglich vier Probanden für die Datenerhebung 
genutzt wurden, wodurch die Wirksamkeit dieses biometrischen Systems nicht final widerlegt werden kann.

Generell gibt es keinen Standard, welcher die Sicherheit eines biometrischen Systems anhand dessen EER-
Wertes bewerten kann. Im Zuge dieser Betrachtung wird eine EER von $x < 1\%$ als sicher angesehen, während 
eine EER von $1\% <= x <= 5\%$ als ausreichend betrachtet werden kann. Eine EER von $5\%$ wäre somit 
ausreichend sicher, sollte allerdings nicht in sicherheitskritischen Anwendungen genutzt werden. Solche 
Systeme würden den Komfort des Nutzers über die Sicherheit seiner Daten stellen, da weniger fehlerhafte 
Ablehnungen auftreten bzw. mehr fehlerhafte Anmeldungen zugelassen werden. In einem solchen System würden 
zwei von 100 Versuchen für einen ratenden Angreifer statistischen Erfolg versprechen (FAR). Bei Semantik A 
melden sich alle Nutzer mit demselben Schriftzug an (Enrollment) und erzielen eine 10-Mal höhere EER von 
$22\%$. Diese ist größer als die bereits hohe EER von $2\%$ und dadurch ebenfalls ungenügend. Dadurch das 
bei Semantik A jede Person denselben Schriftzug verwendet, kann die resultierende EER als die Genauigkeit 
der genutzten Merkmale (bei Zahlen) betrachtet werden. Es ist anzunehmen, dass ein Schriftzug aus 
Buchstaben die EER senken wird.

Der Wertebereich der Differenzen der Verifikationen zu den Referenzen ist bei Semantik A kleiner als bei 
der Semantik B. Bei Semantik A liegen alle Distanzen im folgenden Bereich:

$$1,00 <= x <= 10,96$$

Für die Semantik B liegen alle Distanzen im folgenden Bereich:

$$0,31 <= x <= 22.09$$

Die Distanzen des Wertebereichs der Semantik A fallen kleiner aus, wodurch unterschiedliche Personen eine 
größere Wahrscheinlichkeit haben, in einen ähnlichen Wertebereich zu fallen. Ähnlichkeiten der 
Wertebereiche individueller Personen führen zu einer geringeren euklidischen Distanz. Dieser Zusammenhang 
führt bei Semantik 1 zu einer erhöhten FAR/EER. Es ist anzunehmen, dass ein Test mit mehr als vier 
Probanden den finalen Wertebereich vergrößern würde, da extremere Relationen zwischen biometrischen 
Merkmalen auftreten. Ein solcher Test könnte wegen des beschriebenen Zusammenhangs die EER senken.

Bei der Betrachtung der Distanzwerte der FRR- (Intra-Klassen-) Matrizen wird klar, dass nicht alle 
euklidischen Distanzen auf ähnlichem Level gering sind. Es gibt einzelne Ausnahmen, wie z.B. einen Intra-
Klassen-Wert von 6.95 in der Semantik A, welcher die höchste Distanz zum Enrolment aufweist. Ein solch 
hoher Wert wird auch in höheren Intervallen noch abgewiesen. Neben einzelnen Ausnahmen lassen sich durch 
die FRR-Tabelle auch Personen nachweisen, die einen inkonsistenten Schreibstil haben. So hat Person 1 
überwiegend Distanzen im Bereich von $3,29 <= x <= 4,39$, während Person 4 ausschließlich Werte unter 3,29 
aufweist. Person 1 hat einen ungleichmäßigeren Schreibstil als Person 4, was zu einer höheren Rate an 
fehlerhaften Ablehnungen führt. Diese Inkonsequenz im Schreibstil führt dazu, dass die FRR-Kurve der 
ersten Semantik nicht typisch logarithmisch geformt ist, sondern zusätzlich einen Wendepunkt hat

Die Inkonsequenz im Schreibstil lässt sich auch in der zweiten Semantik finden. Allerdings hat diese keine 
Auswirkungen auf diese (wie zuvor festgestellt bessere) Semantik, da der Wertebereich größer ist. Durch 
diese Reichweite fallen die Werte in Intervalle, welche ebenfalls großzügiger sind.

Bei der Betrachtung der FAR- (Inter-Klassen-) Matrizen zeigt sich, dass die festgestellten inkonsistenten 
Schreibstile, scheinbar keine/wenig Auswirkung auf die FAR-Werte haben. Denn obwohl Person 1 in beiden 
Semantiken eine höhere Anzahl an fehlerhaften Ablehnungen aufweist als Person 4, hat Person 4 eine höhere 
Anzahl an fehlerhaften Anmeldungen. Werte im höheren Bereich bei der FRR, schließen somit nicht auf mehr 
fehlerhafte Anmeldungen / ein leichteres Ziel für einen Angriff.

Zusätzlich wird klar, dass sich einige Personen biometrisch ähnlicher sind als andere. So hat Person 2 in 
Semantik B mit keiner anderen Person eine Distanz, die kleiner als 13,67 ist. Im Gegensatz dazu hat Person 
1 zu Person 3 und 4 nie eine Distanz, welche größer als 10,28 ist. Dadurch ergeben sich für Person 1 (im 
Vergleich zu Person 3 und 4) ausschließlich fehlerhafte Anmeldungen in einem Bereich, der für Person 2 
keinen einzigen ergibt. Das Schriftbild von Person 2 ist somit unähnlich dem, von allen anderen. Im 
Gegensatz dazu sind sich die Merkmale der Personen 1, 3 und 4 ähnlicher. Bei Semantik A und B ergibt sich 
durch die biometrische Unähnlichkeit des Schriftbildes von Person 2 zu allen anderen für die FAR ebenfalls 
keine logarithmische, sondern ebenfalls eine Kurve mit Wendepunkt. Die Distanzen liegen vermehrt in 
höheren Intervallen, wodurch die Kurve für die restlichen Werte abflacht.

\section{Welche weiteren Schritte wäre nötig, um dieses biometrische Verfahren unter dem Einsatz von gezielten Fälschungen zu evaluieren?}

\subsection{Entwicklung der Testprotokolle}

Die Entwicklung eines Testprotokolls für Fälschungsszenarien sollte derart gestaltet werden, dass alle 
relevanten Szenarien systematisch abgedeckt werden. Zunächst sollten Intra-Klassen Tests durchgeführt 
werden. Diese Tests prüfen, wie gut das biometrische System in der Lage ist, authentische Proben einer 
Person korrekt zu erkennen. Dabei werden Verifikationsproben mit den Enrolment-Daten derselben Person 
abgeglichen. Das Ziel dieser Tests ist die Bestimmung der False-Non-Match-Rate (FNMR), welche die 
Häufigkeit beschreibt, mit der authentische Proben fälschlicherweise abgelehnt werden.

Zusätzlich sollten Inter-Klassen Tests mit verschiedenen Angriffsstärken berücksichtigt werden, um die 
Robustheit des Systems gegen Fälschungsversuche zu evaluieren. Diese Tests umfassen:

\begin{itemize}
	\item Blind-Force Angriffe: Bei dieser Angriffsstärke versucht der Angreifer, eine biometrische Probe zu fälschen, ohne Kenntnisse über das Schriftbild oder den Prozess, der zur Erstellung der authentischen Probe geführt hat.
	\item Low-Force Angriffe: Hier hat der Angreifer Zugang zu einem sichtbaren Ergebnis, wie beispielsweise einer Unterschrift, und versucht, dieses nachzuahmen. Es besteht jedoch kein Wissen über den dahinterliegenden Erstellungsprozess, wie etwa Geschwindigkeit, Druck oder Reihenfolge der Bewegungen.
\end{itemize}

\subsection{Erstellung gezielter Fälschungsdaten}

Zur Erstellung gezielter Fälschungsdaten müssen spezifische Angriffstypen berücksichtigt werden, um die Robustheit des biometrischen Systems unter realistischen Bedingungen zu bewerten.
Bei den Blind-Force Daten handelt es sich um Proben, die ohne jegliche Kenntnis des Originals erstellt werden. Der Angreifer versucht, das System zu täuschen, ohne eine Vorlage oder Informationen über das authentische Schriftbild oder den Schreibprozess zu haben.
Bei Low-Force Daten werden auf Grundlage eines sichtbaren Endergebnisses, wie beispielsweise einer Unterschrift, erstellt. Dabei hat der Angreifer keine Kenntnis über den eigentlichen Schreibprozess, wie etwa Bewegungsdynamiken, Geschwindigkeit oder Druck.

\subsection{Implementierung und Durchführung}

Zur Implementierung und Durchführung von Tests sollten mehrere Szenarien berücksichtigt werden, um die Leistungsfähigkeit und Sicherheit des biometrischen Systems zu bewerten. Für den Verifikationstest (Intra-Klassen-Test) werden authentische Proben mit den Enrolment-Daten der gleichen Person abgeglichen. Ziel dieses Tests ist die Messung der False-Non-Match-Rate (FNMR) oder False-Rejection-Rate (FRR), die die intraindividuelle Fehlklassifikation darstellt. Die zuvor erstellten Fälschungsproben werden gegen die Enrolment-Daten der angegriffenen Personen geprüft. Ziel dieser Tests ist die Messung der False-Match-Rate (FMR) oder False-Acceptance-Rate (FAR) für die jeweiligen Angriffsszenarien (FMRBlind, FMRLow-Force).

\subsection{Analyse der Ergebnisse}

Zur Bewertung der Testergebnisse werden folgende Punkte analysiert. Der Punkt, an dem die False-Match-Rate 
(FMR) und die False-Non-Match-Rate (FNMR) gleich sind ist die Equal-Error-Rate (EER). Für die 
Angriffsszenarien Blind und Low-Force wird jeweils eine separate EER berechnet, um die Robustheit des 
Systems gegenüber unterschiedlichen Angriffsstärken zu bewerten. Durch den Vergleich der Fehlerraten wird 
dann analysiert, wie sich die Falscherkennungsraten zwischen Blind- und Low-Force-Angriffen verändern. Ein 
Anstieg der FMR von Blind- zu Low-Force-Angriffen kann auf Schwachstellen im System hinweisen.

\subsection{Bewertung und Optimierung des Systems}

Auf Grundlage der Testergebnisse können gezielte Maßnahmen zur Verbesserung des biometrischen Systems 
ergriffen werden. Zum einen kann eine Verbesserung der Algorithmen vorgenommen werden. Es sollten 
spezifische Fälschungserkennungsmethoden entwickelt werden, die gezielt auf die Schwachstellen des Systems 
bei Blind- und Low-Force-Angriffen eingehen. Außerdem kann die Systemschwelle erhöht werden. Die Anpassung 
der Entscheidungsschwellen kann dazu beitragen, die Falscherkennungsrate bei Angriffen zu verringern, ohne 
die Benutzerfreundlichkeit zu beeinträchtigen. Zuletzt können mehrere Merkmale kombiniert werden. Die 
Integration zusätzlicher biometrischer Eigenschaften wie Schreibdruck oder Schreibgeschwindigkeit kann die 
Sicherheit des Systems erhöhen und es widerstandsfähiger gegen gezielte Fälschungen machen.

\subsection{Iterative Tests und Validierung}

Nach der Optimierung des biometrischen Systems sollten erneut Tests durchgeführt werden, um die Wirksamkeit der vorgenommenen Anpassungen zu validieren.

\begin{itemize}
	\item Langzeitbeobachtungen: Es ist wichtig zu überprüfen, ob die Verbesserungen des Systems über einen längeren Zeitraum hinweg stabil und nachhaltig bleiben. Dies hilft, potenzielle Schwächen oder Verschlechterungen in der Leistung frühzeitig zu erkennen.
	\item Cross-Validation: Die Ergebnisse sollten auf unterschiedliche Benutzergruppen und Nutzungsszenarien übertragen werden, um sicherzustellen, dass die Verbesserungen universell anwendbar sind und nicht nur unter spezifischen Bedingungen funktionieren.
\end{itemize}

\subsection*{Ausschluss von Zero-Effort}

\begin{quote}
	Ein erwähnenswerter Angriff, der aber nicht gezielt für dieses Szenario relevant ist, ist der Zero-
	Effort/Random-Angriff. Das ist ein Angriff, dessen Ziel es ist, ohne gezielte Fälschungen, also 
	zufälligen Versuchen, das System zu täuschen. Ein Zero-Effort Fälschungstest ist also ein Test, bei 
	dem zufällige Proben gegen die Enrolment-Daten anderer Personen geprüft werden. Das Ziel dieses Tests 
	ist die Bestimmung der False-Match-Rate (FMR) oder False-Acceptance-Rate (FAR) für zufällige 
	Übereinstimmungen.	
\end{quote}

\section{Welche Rahmenvorgaben-/ annahmen sollten dabei Berücksichtigung finden}

Insbesondere die technische Infrastruktur und die Umgebung der Erfassung der Daten müssen für alle Probanden gleich sein, um vergleichbare Resultate zu erzielen.

\subsection{Umgebung}

Die Schriftproben sollten so erhoben werden, dass kein anderer Proband die Erfassung beobachten kann, zum 
Beispiel durch einen Sichtschutz. So können Brute-Force Angriffe früh unterbunden werden.

\subsection{Erfassung}

Offline Handschriften können über verschiedene technische Verfahren erhoben werden. Im Folgenden werden 
zwei Beispiele

\subsubsection{Scanner}

Eine Möglichkeit bieten handelsübliche Dokumentenscanner. Sie beleuchten die Schriftprobe und erfassen das 
reflektierte Licht der Schrift. Wichtig hierbei ist eine hohe Auflösung des Geräts, damit das Schriftbild 
detailliert erfasst werden kann und auch feine Striche, kleine Punkte oder Lücken in der Schrift erkannt 
werden können.

Scanner haben den Vorteil, dass sie unter immer gleichen Bedingungen arbeiten. Die Beleuchtung und 
Auflösung ändern sich nicht. Dafür sind sie aber nicht mobil und brauchen Platz. Sie werden ungenau, wenn 
die Schriftproben beschädigt oder geknittert sind.

\subsubsection{Digitalkameras}

Alternativ können Digitalkameras zur Erfassung von Schriftproben verwendet werden. Hierbei spielt die 
Auflösung ebenfalls eine Rolle. Eine bessere Auflösung bietet mehr Details. Des Weiteren spielt das 
verwendete Objektiv eine große Rolle, um Verzerrungen zu verhindern. Der Vorteil von Digitalkameras ist, 
dass sie in vielen verschiedenen Größen existieren. Sie sind mobil und nicht stationär. Ein Nachteil ist, 
dass es schwer ist, identische Voraussetzung für die Erfassung zu gewährleisten, da Licht, Winkel und 
Abstand zur Schriftprobe leicht verändert werden können.

Jeder Sensor muss entsprechend kalibriert und gewartet werden, damit identische Bedingungen bei der
Erfassung herrschen.

\subsection{Speicherung}

Eine sichere und effiziente Speicherung spielt eine tragende Rolle. Damit später eine präzise Analyse der 
erfassten Daten möglich ist, müssen die Rohdaten in einem verlustfreien Format gespeichert werden. 
Relevante Metadaten, wie Informationen zum Sensor und den Probanden sind ebenfalls zu erheben, damit bei 
Fehlern ein entsprechendes Troubleshooting erfolgen kann.

Da es sich um biometrische, und damit besonders schützenswerte Daten handelt, ist ein Zugriffs- und 
Berechtigungssystem nötig. Dadurch kann der Zugriff auf die Daten sowohl eingeschränkt, als auch 
nachvollzogen werden. Außerdem müssen biometrische Daten immer ausreichend verschlüsselt gespeichert und 
übertragen werden.

\section{Beschreiben Sie ein Protokoll für die hierfür erforderliche Datenerhebung}

\subsection{Vorbereitung der Datenerhebung}

\subsubsection{Definition der Zielsetzung}

Das Ziel dieser Datenerhebung ist die Evaluierung der Fälschungssicherheit unseres biometrischen Systems 
im Hinblick auf die, handschriftliche erfassten, Semantiken:

\begin{itemize}
	\item \textbf{Semantik A}: Numerische, 5-stellige PIN (78402)
	\item \textbf{Semantik B}: Der Geburtsort jedes Probanden
\end{itemize}

Es wird ein besonderer Fokus auf zwei gezielte Angriffe gelegt:
Blind-Force: Angriffe basierend auf visuellen Vorlagen.
Low-Force: Angriffe mit Zugriff auf eine authentische Verifikationsprobe

\subsubsection{Auswahl der Probanden}

Idealerweise 50–100 Probanden (angenommen es handelt sich um ein echtes biometrisches System) bzw. 2-4 (in 
dem Rahmen dieses Moduls)

Probanden sollten ein unterschiedliches Alter, Geschlecht und einen unterschiedlichen Schreibstil haben, 
um eine ausreichende Intra-Class-Variabilität zu gewährleisten.

Eine schriftliche Einwilligung gemäß Datenschutzrichtlinien (z. B. DSGVO) ist einzuholen.

\subsubsection{Technische Infrastruktur}
Es müssen abhängig von der Probanden-Anzahl mehrere Schreibtabletts oder Touchscreens zur Datenerfassung bereitgestellt werden.
Speicherung der Proben als Bild- oder Vektordaten.
Umgebung: Einheitliche Bedingungen (z. B. Beleuchtung, Schreibfläche).

\subsection{Durchführung der Datenerhebung}

Da diese Datenerhebung ausschließlich auf gezielte Angriffe bezogen ist, werden Enrolment- und 
Verifikationsproben nicht erneut erhoben. Eine Variante (falls mehr authentische Probanden gefordert sind) 
könnte sein, diese ebenfalls zu erheben, da der logistische Aufwand fast identisch ist.

\subsubsection{Erhebung der Fälschungsproben}

\subsubsection*{Blind-Force-Angriffe}

\begin{flushleft}
	\textbf{Semantik A}: Angreifer erhalten die Pin und erfassen 5 Fälschungsproben für diese\\
	\textbf{Semantik B}: Angreifer erstellen für jeden authentischen Nutzer 5 Fälschungsproben	
\end{flushleft}

\subsubsection*{Low-Force-Angriffe}

\textbf{Semantik A und B}: Angreifer erhalten für jeden Nutzer (4 in unserem Fall) Zugriff auf Bilder der 
authentischen Proben und erfassen jeweils 5 Fälschungsproben!

\subsection{Verarbeitung der Daten}

\subsubsection{Qualitätssicherung}

\begin{flushleft}
	\textbf{Semantik A}: Prüfung auf numerische Vollständigkeit und Klarheit\\
	\textbf{Semantik B}: Prüfung auf Lesbarkeit und Textvollständigkeit, sowie Übereinstimmung des Geburtsortes mit dem authentischen.	
\end{flushleft}

\subsubsection{Anonymisierung}

Um eine DSGVO-konforme Verarbeitung von Probandendaten zu gewährleisten, werden alle Referenzen zu 
personenbezogenen Daten durch pseudonymisierte IDs ersetzt. Somit lassen sich Angreifer unterscheiden, es 
gibt allerdings keine Möglichkeit, einen Rückschluss auf den tatsächlichen Probanden zu ziehen.

\subsubsection{Datenaufteilung}

Die erhobenen Daten werden in zwei Gruppen unterteilt (angenommen es handelt sich um eine große Anzahl an Probanden):

\begin{enumerate}
	\item Trainingsdaten, die für eine weitere Systemoptimierung verwendet werden können
	\item Testdaten, die für eine Evaluation des biometrischen Systems genutzt werden können
\end{enumerate}

\subsection{Evaluierung der Angrifftests}

Um die Evaluierung des Systems gegen die zwei Arten von Angriffen durchzuführen werden sogenannte Angrifftests durchgeführt:

\subsubsection{Blind-Force-Angrifftests}

In den Blind-Force-Angrifftests werden:

\begin{enumerate}
	\item In Semantik A, pro Angreifer 5 Verifikationsversuche unternommen und die FAR berechnet
	\item In Semantik B, pro Nutzer  5 Verifikationsversuche unternommen und die FAR berechnet
\end{enumerate}

Die unterschiedliche Natur der beiden Semantiken führt bei Blind-Force-Angriffen dazu, dass es 
unterschiedlich viele Angriffe für Semantik A \& B gibt.

\subsubsection{Low-Force-Angrifftests}

In den Low-Force-Angrifftests werden:

\begin{enumerate}
	\item In Semantik A, pro Angreifer und Proband Verifikationsversuche unternommen und die FAR berechnet
	\item In Semantik B, pro Angreifer und Proband, Verifikationsversuche unternommen und die FAR berechnet
\end{enumerate}

\subsection{Dokumentation und Auswertung}

\subsubsection{Fehleranalyse}

In die bestehenden Diagramme aus Aufgabe 1) für FAR/FRR können nun, für jede Semantik, zwei weitere Graphen eingezeichnet werden:

\begin{enumerate}
	\item Blind-Force-FAR – Die Acceptance-Rate von Angreifern, die Zugriff auf das geheime Wissen (bspw. Pin/Ort respektive) hatten, aber kein Beispiel für die Handschrift hatten.
	\item Low-Force-FAR – Die Acceptance-Rate von Angreifern, die Zugriff auf authentische offline Verifikationsdaten von authentischen Probanden hatten.
\end{enumerate}

Des Weiteren könnte eine Fehleranalyse dieser Charakteristiken erfolgen:

\begin{enumerate}
	\item Schriftproben haben (bspw. wie schneiden Schreibschrift vs. Druckschrift ab?) 
	\item Inhalte haben (bspw. wie schneiden kurze Städtenamen vs. lange ab?)
\end{enumerate}

um Rückschlüsse auf das Verhalten der Merkmalsextraktion zu erhalten.

\subsubsection{Berichterstellung}

Wie oben bereits beschrieben, müssen Semantik A und B separat ausgewertet werden. Die eingezeichnete FAR 
der verschiedenen Angriffe in den Diagrammen kann daraufhin genutzt werden, um einen verbesserten 
Schwellwert oder Änderungen am biometrischen System vorzuschlagen.

\subsubsection{Optimierung}

Letztlich sollten Anpassungen des Systems vorgenommen werden, um die Robustheit gegen gezielte Fälschungen 
zu erhöhen. Nach erfolgten Anpassungen kann das System mit den hier erstellten Angrifftests erneut 
getestet und ausgewertet werden, um eine Verbesserung bzw. Verschlechterung festzustellen.

\end{document}

### 2025-03-22 thesis/thesis.tex
Setup thesis document

\documentclass{article}

\usepackage[a4paper]{geometry}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}
\usepackage{titlesec}

\graphicspath{{./images/}}

\pagestyle{fancyplain}
\fancyhf{}
\lhead{\fancyplain{}{Mara Schulke} }
\rhead{\fancyplain{}{\today}}
\cfoot{\fancyplain{}{\thepage}}

\newcommand{\figuresource}[1]{
	\begin{center}Quelle: {#1}\end{center}
}

\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}

\begin{document}

\begin{titlepage}
    \begin{center}
        \begin{Large}
            Brandenburg University of Technology \\[1em]
        \end{Large}
        IT Security \\
        Computerscience and Media \\
        Prof. Dr. Oleg Lobachev \\
        Florian Eich
    \end{center}

    \vfill

    \begin{center}
        \Large{Embedding Generative Pretrained Transformers into PostgreSQL}\\[0.5em]
        \large{Bachelor Thesis}\\[1em]

        \begin{normalsize}
            Summer semester 2025\\[0.25em]
            \today
        \end{normalsize}
    \end{center}

    \vfill

    \begin{center}
        Mara Schulke – Matr-Nr. 20215853
    \end{center}
\end{titlepage}

\begin{abstract}
This thesis explores the integration of Generative Pretrained Transformers (GPT)
into PostgreSQL database systems. The research focuses on implementation strategies,
performance optimization, and practical applications of this integration.
\end{abstract}

\tableofcontents

\listoffigures

\section*{List of Abbreviations}
\begin{tabular}{ll}
GPT & Generative Pretrained Transformer \\
SQL & Structured Query Language \\
API & Application Programming Interface \\
% Add other abbreviations as needed
\end{tabular}

\newpage

% Introduction
\section{Introduction}
\subsection{Problem Statement and Motivation}
\subsection{Objectives of the Thesis}
\subsection{Research Questions}
\subsection{Methodological Approach}
\subsection{Structure of the Thesis}

\newpage

% Theoretical Foundations
\section{Theoretical Foundations}

\subsection{Generative Pretrained Transformers (GPT)}
\subsubsection{Architecture and Functionality}
\subsubsection{Training and Fine-tuning}
\subsubsection{Application Areas}

\subsection{PostgreSQL as a Database System}
\subsubsection{Architecture of PostgreSQL}
\subsubsection{Extension Capabilities}
\subsubsection{PostGIS and Other Extensions as Examples}

\subsection{Embedding AI Models in Database Systems}
\subsubsection{Current Approaches and Solutions}
\subsubsection{Technical Challenges}
\subsubsection{Benefits of Integration}

\newpage

% Conceptual Design
\section{Conceptual Design of GPT Embedding in PostgreSQL}

\subsection{Requirements Analysis}
\subsubsection{Functional Requirements}
\subsubsection{Non-functional Requirements}

\subsection{Architecture Design}
\subsubsection{Interface Design}
\subsubsection{Data Model}
\subsubsection{Integration into PostgreSQL}

\subsection{Technical Implementation Strategies}
\subsubsection{Foreign Data Wrapper}
\subsubsection{Extension Using C/C++}
\subsubsection{PL/Python or Other Procedural Languages}
\subsubsection{Comparison of Approaches}

\newpage

% Implementation
\section{Implementation}

\subsection{Development Environment and Tools}

\subsection{Integration of the GPT Model}
\subsubsection{Model Selection and Optimization}
\subsubsection{API Connection or Local Embedding}

\subsection{Development of the PostgreSQL Extension}
\subsubsection{SQL Functions for GPT Interactions}
\subsubsection{Data Type Conversion and Processing}
\subsubsection{Error Handling and Logging}

\subsection{Optimization}
\subsubsection{Performance Tuning}
\subsubsection{Memory Usage}
\subsubsection{Parallelization}

\newpage

% Evaluation
\section{Evaluation}

\subsection{Test Environment and Methodology}

\subsection{Performance Tests}
\subsubsection{Latency}
\subsubsection{Throughput}
\subsubsection{Scalability}

\subsection{Use Cases}
\subsubsection{Natural Language Queries}
\subsubsection{Text Generation Within the Database}
\subsubsection{Semantic Search and Text Classification}

\subsection{Comparison with Alternative Approaches}

\newpage

% Discussion
\section{Discussion}

\subsection{Interpretation of Results}
\subsection{Limitations of the Implementation}
\subsection{Ethical and Data Privacy Considerations}
\subsection{Potential Future Developments}

\newpage

% Summary and Outlook
\section{Summary and Outlook}

\subsection{Summary of Results}
\subsection{Addressing the Research Questions}
\subsection{Outlook for Future Research and Development}

\newpage

\begin{thebibliography}{99}
\bibitem{ref1} Author, A. (Year). Title of the reference. Journal/Publisher, Volume(Issue), Pages.
% Add more references as needed
\end{thebibliography}

\newpage

\appendix
\section*{Appendix}
\subsection*{Installation Guide}
\subsection*{API Documentation}
\subsection*{Code Examples}
\subsection*{Test Data and Results}

\end{document}

### 2025-04-14 thesis/thesis.tex
Write the thesis introduction

@@
     \begin{center}
-        \Large{Embedding Generative Pretrained Transformers into PostgreSQL}\\[0.5em]
+        \Large{Natural Language Queries using Large Language Models}\\[0.5em]
         \large{Bachelor Thesis}\\[1em]
@@
 \begin{abstract}
-This thesis explores the integration of Generative Pretrained Transformers (GPT)
-into PostgreSQL database systems. The research focuses on implementation strategies,
-performance optimization, and practical applications of this integration.
+This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
+to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
+on implementation strategies, performance optimization, and practical applications of this concept.
 \end{abstract}
@@
 \subsection{Problem Statement and Motivation}
+
+Database systems represent a backbone of modern computer science, allowing for rapid advancements
+whilst shielding us from the problem categories that come along with managing and querying large amounts
+of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
+traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
+to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
+represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
+etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
+
+This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
+data analytics tools etc. which therefore introduce significant churn and delay between the implementation
+of a database system and reaching the desired end user impact. Often these projects span multiple years, require
+costly staffing and yield little to no novel technical value.
+
+Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
+fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
+outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
+challenging / yielded unsatisfying results with discrete programming approaches.
+
+This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
+business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
+requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
+this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
+in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
+(ie. moving away from SQL) towards a more inclusive and data driven world.
+
 \subsection{Objectives of the Thesis}
+
+This thesis aims to address the aforementioned challanges when it comes to database accessibility.
+The following objectives are considered to cover the core research area of this thesis:
+
+\begin{enumerate}
+    \item Develop a database extension that can translate natural language queries into semantically
+          accurate SQL queries using Large Language Models.
+    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
+          techniques in order to improve the performance of the system.
+    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
+    \item Test the 
+\end{enumerate}
+
+% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
+
 \subsection{Research Questions}

### 2025-04-16 thesis/thesis.tex
Set research goals

@@
 API & Application Programming Interface \\
-% Add other abbreviations as needed
+LLM & Large Language Model \\
+DBMS & Database Management System \\
+NL2SQL & Natural Language to SQL \\
 \end{tabular}
@@
 This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are considered to cover the core research area of this thesis:
+The following objectives are the core research area of this thesis:
 
@@
     \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Test the 
+    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
+    \item Idenitfy potential usecases for real world scenarios that could deliver a noticable upsides to users.
+    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
 \end{enumerate}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
 
@@
 
+% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
+
 \subsection{Test Environment and Methodology}

### 2025-04-17 thesis/thesis.tex
Research questions

@@
     \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential usecases for real world scenarios that could deliver a noticable upsides to users.
+    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
     \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
@@
 \subsection{Research Questions}
+
+% \subsubsection{Primary Research Questions}
+
+\subsubsection*{RQ1 — Are natural language queries performant and accurate enough to find application in the real world?}
+
+The primary research questions when it comes to natural language database interfaces evolves around their
+semantic accuracy and reliability. LLMs have notoriously been known for their ability to hallucinate /
+produce false, but promising outputs. This behaviour can be especially  dangerous when opting for data
+driven decisions that rely on false data due to a mistranslation from natural language to SQL. LLMs could
+cause hard to understand/debug behaviour, like false computation of distributions when the intermediate
+format is not being shown to the user. This thesis tries to determine whether such hallucinations could be
+reasonably prevented (ie. through pre and post-processing if LLM input/output).
+
+Additionally, this research eveluates the associated performance and hardware requirements, as
+a practical deployment of such a natural language database interface outside of research situations
+is guarded by the ability to:
+
+\begin{enumerate}
+    \item Run on reasonable, mass available hardware  (e.g. excluding high end research GPUs).
+    \item Run within a suitable timeframe that is not causing a significant performance degredation of the 
+          overall system / delay to the end user.
+\end{enumerate}
+
+\subsubsection*{RQ2 — Are there effective approaches against ambiguity in natural language input?}
+
+To provide semantically correct results ambiguity in the user-provided natural language queries must be 
+adequately addressed. This thesis investigates various approaches to ambiguity management and
+resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
+e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
+"all sales".
+
+Such situations present the second major challenge associated with the practical implementation of natural
+language database interfaces. The success of this concept will significantly depend on whether suitable 
+designs and mitigation techniques can be implemented without creating problems with regards to the 
+aforementioned performance and hardware requirements. The research focus lies on both preventative measures
+through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
+that post process LLM output, either on the basis of further user input or context inference.
+
+\subsubsection*{RQ3 — What are possible approaches for increasing semantic accuracy of queries?}
+
+In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
+Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
+interactive contextual reasoning through a conversational interface which would allow for user
+refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
+partly relying on tradition NLP preprocessing techniques. This research will quantify semantic accuracy
+using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
+performance. Furthermore this research will take a look at the optimal combination of the aforementioned
+solutions in order to develop a system that strikes the right balance between accuracy and performance.
+
+% \subsubsection{Secondary Research Questions}
+
+% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
+
+% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
+
+% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
+
 \subsection{Methodological Approach}

### 2025-04-19 thesis/thesis.tex
Methodology

@@
 
-\subsubsection*{RQ1 — Are natural language queries performant and accurate enough to find application in the real world?}
+\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
 
-The primary research questions when it comes to natural language database interfaces evolves around their
-semantic accuracy and reliability. LLMs have notoriously been known for their ability to hallucinate /
-produce false, but promising outputs. This behaviour can be especially  dangerous when opting for data
-driven decisions that rely on false data due to a mistranslation from natural language to SQL. LLMs could
-cause hard to understand/debug behaviour, like false computation of distributions when the intermediate
-format is not being shown to the user. This thesis tries to determine whether such hallucinations could be
-reasonably prevented (ie. through pre and post-processing if LLM input/output).
+The primary research questions when it comes to natural language database interfaces evolve around their
+semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
+LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
+This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
+due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
+like false computation of distributions when the intermediate format is not being shown to the user. This
+thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
+performance and hardware requirements are suitable for a real world deployment, outside of research situations.
 
-Additionally, this research eveluates the associated performance and hardware requirements, as
-a practical deployment of such a natural language database interface outside of research situations
-is guarded by the ability to:
+Specifically the two big underlying questions are:
 
 \begin{enumerate}
-    \item Run on reasonable, mass available hardware  (e.g. excluding high end research GPUs).
-    \item Run within a suitable timeframe that is not causing a significant performance degredation of the 
-          overall system / delay to the end user.
+    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
+          benefit to users?
+    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
 \end{enumerate}
@@
 
-\subsubsection*{RQ3 — What are possible approaches for increasing semantic accuracy of queries?}
+\subsubsection*{RQ3 — What are effective approaches for increasing semantic accuracy of queries?}
 
@@
 
+\newpage
+
 \subsection{Methodological Approach}
+
+This thesis is following a research and development methodology in order to implement a natural language
+interface for databases, in particular postgres is used.
+
+\begin{enumerate}
+    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
+          natural language interfaces (NLI) for databases, GPU integration for acceleration
+          of database operations, and LLM/AI Model integration within database systems.
+          This phase establishes the theoretical foundation for this research and identifies current 
+          state-of-the-art approaches, their benefits and shortcomings.
+     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
+          for LLM integration from within postgres. The primary goals of the system design phase
+          are to arrive at an architecture that yields low latency natural language processing,
+          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
+          a high semantic accuracy.
+     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
+          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
+          provide a GPU accelerated framework for executing LLMs, implement a natural language
+          to query generation pipeline that relies on the SQL schema and create database functions
+          and operators for both query generation and execution.
+     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
+          the implementations performance in multiple dimensions. Namely the most relevant dimensions
+          for this thesis are:
+          \begin{enumerate}
+              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
+                    natural language input.
+              \item Performance Metric — 
+          \end{enumerate}
+\end{enumerate}
+
 \subsection{Structure of the Thesis}
@@
 % Theoretical Foundations
-\section{Theoretical Foundations}
+\section{Related Work}
 
-\subsection{Generative Pretrained Transformers (GPT)}
-\subsubsection{Architecture and Functionality}
-\subsubsection{Training and Fine-tuning}
-\subsubsection{Application Areas}
-
-\subsection{PostgreSQL as a Database System}
-\subsubsection{Architecture of PostgreSQL}
-\subsubsection{Extension Capabilities}
-\subsubsection{PostGIS and Other Extensions as Examples}
-
-\subsection{Embedding AI Models in Database Systems}
-\subsubsection{Current Approaches and Solutions}
-\subsubsection{Technical Challenges}
-\subsubsection{Benefits of Integration}
+\subsection{360 Degree vergleichbare paper etc.}
 

### 2025-04-19 thesis/thesis.tex
Fixing the thesis structure and writing about it

@@
 \newpage
+
+% 1/3 Vorarbeit, Literature Review
+% 1/3 Decomp und System Design
+% 1/3 Implementation und Evaluation
 
@@
 refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on tradition NLP preprocessing techniques. This research will quantify semantic accuracy
+partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
 using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
@@
 
-\subsection{Methodological Approach}
+\subsection{Structure of the Thesis}
 
@@
           state-of-the-art approaches, their benefits and shortcomings.
+     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
+          fundamentals and deriving system requirements for the design phase from it. The goal
+          of this section is to arrive at a list of functional and non-functional requirements that
+          must be taken into account and fulfilled by the design and implementation phases respectively.
      \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
@@
                     natural language input.
-              \item Performance Metric — 
+              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
+                    ambiguous natural language input and database schemas.
+              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
+                    of the implementation.
           \end{enumerate}
+      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
+            the research goals of this thesis. Evaluating the performance and accuracy results recorded
+            during the benchmarks against the question whether real world deployments of NILs are feasible.
+            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
+            strategies are showing a statistically significant effect.
+       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
+            of this thesis and the implementation, and proposes directions for future research alongside
+            possible applications. Primary future research topics include advanced GPU optimization
+            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
+            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
+            and privacy considerations (e.g. managing access control).
 \end{enumerate}
-
-\subsection{Structure of the Thesis}
 
@@
 
+% Welche approaches fuer llm integration
+% Welche approaches fuer natural to sql
+
 \newpage
 
+\section{Decomposition \& Requirements}
+
+\subsection{Problem Decomposition}
+
+\subsection{Requirements}
+
 % Conceptual Design
-\section{Conceptual Design of GPT Embedding in PostgreSQL}
-
-\subsection{Requirements Analysis}
-\subsubsection{Functional Requirements}
-\subsubsection{Non-functional Requirements}
+\section{System Design}
 
@@
 \subsubsection{Data Model}
-\subsubsection{Integration into PostgreSQL}
+\subsubsection{Integration into SQL}
 
 \subsection{Technical Implementation Strategies}
-\subsubsection{Foreign Data Wrapper}
-\subsubsection{Extension Using C/C++}
-\subsubsection{PL/Python or Other Procedural Languages}
-\subsubsection{Comparison of Approaches}
 

### 2025-04-21 thesis/thesis.tex
Add bibliography

@@
 
-\subsubsection*{RQ2 — Are there effective approaches against ambiguity in natural language input?}
+\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
 
@@
 
-\subsubsection*{RQ3 — What are effective approaches for increasing semantic accuracy of queries?}
+\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
 
@@
 % Theoretical Foundations
-\section{Related Work}
+\section{Literature Review}
+
+Test\cite{Atlas}
 
@@
 
-\begin{thebibliography}{99}
-\bibitem{ref1} Author, A. (Year). Title of the reference. Journal/Publisher, Volume(Issue), Pages.
-% Add more references as needed
-\end{thebibliography}
-
 \newpage
@@
 
+\nocite{*}
+
+\bibliographystyle{apalike}
+\bibliography{references}
+
 \end{document}

### 2025-04-21 thesis/thesis.tex
Update

@@
 \usepackage{titlesec}
+\usepackage[natbibapa]{apacite}
 
@@
         \begin{Large}
-            Brandenburg University of Technology \\[1em]
+            Brandenburg University of Applied Sciences \\[1em]
         \end{Large}
         IT Security \\
-        Computerscience and Media \\
+        Computerscience \\
         Prof. Dr. Oleg Lobachev \\
-        Florian Eich
+        MSc. Florian Eich
     \end{center}
@@
     \begin{center}
-        \Large{Natural Language Queries using Large Language Models}\\[0.5em]
+        \Large{A Natural Language Interface Implementation for PostgreSQL using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
         \large{Bachelor Thesis}\\[1em]
@@
 
-Test\cite{Atlas}
-
-\subsection{360 Degree vergleichbare paper etc.}
+Test\citep{Atlas}
 
@@
 
-\subsection{Integration of the GPT Model}
-\subsubsection{Model Selection and Optimization}
-\subsubsection{API Connection or Local Embedding}
+\subsection{Integration of the Model}
 
 \subsection{Development of the PostgreSQL Extension}
-\subsubsection{SQL Functions for GPT Interactions}
-\subsubsection{Data Type Conversion and Processing}
-\subsubsection{Error Handling and Logging}
 
 \subsection{Optimization}
-\subsubsection{Performance Tuning}
-\subsubsection{Memory Usage}
-\subsubsection{Parallelization}
 
@@
 
-\bibliographystyle{apalike}
+\bibliographystyle{apacite}
 \bibliography{references}

### 2025-04-21 thesis/thesis.tex
New refs

@@
 
-Test\citep{Atlas}
+In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
+(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. Following their development starting in
+the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
+larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
+research focuses on the recent advancements when it comes to language models and how they can be harnessed for effective NL2SQL systems \citep{LLM-Sql, T2SQL-LLM-Bench, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.
+
+This literature review is covering the foundational concepts, challenges, key advancements and research gaps
+associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
+the research questions introduced in the previous chapter in context.
+
+\subsection{Foundations of Natural Language Interfaces to Databases}
+
+Research on NLIDBs began as early as
 

### 2025-04-22 thesis/thesis.tex
Write more foundational literature review

@@
 
-Research on NLIDBs began as early as
+One of the first corner stone research papers on Natural Language Database Interfaces (NLIDBs) was published
+over three decades ago by \citeauthor*{NLIDBs} where an introduction and an overview of the state-of-the-art
+in the field were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges associated
+with NLIDBs, and compared them against existing / competing solutions like formal query languages, form-based
+interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities, semantic 
+inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
+metrics today.
+
+Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
+natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
+was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
+\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
+parsing for natural language database queries with multi clause construction, yielding promising results
+in the field of NLIDBs. \citep{ILPParsing2}
+
+Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
+\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
+approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
+\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
+system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
+\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
+tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
+also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
+unknown questions into the system and maintain performance characteristics, whereas it was shown that 
+\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
 

### 2025-04-23 thesis/thesis.tex
Add more review content

@@
     \begin{center}
-        \Large{A Natural Language Interface Implementation for PostgreSQL using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
+        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
         \large{Bachelor Thesis}\\[1em]
@@
 larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for effective NL2SQL systems \citep{LLM-Sql, T2SQL-LLM-Bench, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.
+research focuses on the recent advancements when it comes to language models and how they can be harnessed for
+effective NL2SQL systems \citep{LLM-Sql, T2SQL-LLM-Bench, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.
 
@@
 \cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
+
+The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
+field and highlighted the following, ongoing research:
+
+\begin{enumerate}
+    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
+          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
+    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
+          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
+    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
+          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
+\end{enumerate}
+
+Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
+able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
+like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
+As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
+reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
+on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
+as possible natural language coverage. \citep{NLIDBTheory}
 

### 2025-04-24 thesis/thesis.tex
Fix error

@@
 In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. Following their development starting in
+(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
 the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 

### 2025-04-25 thesis/thesis.tex
Rule based

@@
 
+% bessere einleitung
+
 \subsection{Foundations of Natural Language Interfaces to Databases}
@@
 
-% Welche approaches fuer llm integration
-% Welche approaches fuer natural to sql
+\subsection{Traditional NL2SQL Approaches}
+
+Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
+traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These discrete
+approaches have laid the foundations for transitioning towards the application of machine learning techniques
+for NL2SQL.
+
+\subsubsection{Rule-based and Grammar-based Systems}
+
+Early research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and expensive
+to maintain / transfer across database systems. These rule engines mostly relied on the systematic identification
+of linguistic patterns / were trying to template SQL from information that was derived from processing the natural
+language query. \citep*{Rendevouz, Lunar, Ladder} These approaches mostly tried to formalize natural language queries into 
+formal grammers which could then be deterministically mapped into a valid SQL query. These approaches have strong downsides
+when it comes to the variety of natural language constructs they can process, aswell as runtime adoption of new / unknown
+databases, query constructs etc. A potential upside of this class of NL2SQL systems is that they can confidently and 
+reproducably identify questions they can, and can't answer — thus leading to very reliable and predictable user interfaces.
+
+\subsection{Neural NL2SQL Approaches}
+
+\subsection{Benchmark Evolution}
+
+\subsection{LLM-based NL2SQL Systems}
+
+\subsection{Optimization for LLM-based NL2SQL}
+
+\subsection{Complex Techniques in NL2SQL}
+
+\subsection{Research Gaps}
 

### 2025-04-25 thesis/thesis.tex
String kernel

@@
 
-One of the first corner stone research papers on Natural Language Database Interfaces (NLIDBs) was published
-over three decades ago by \citeauthor*{NLIDBs} where an introduction and an overview of the state-of-the-art
-in the field were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges associated
-with NLIDBs, and compared them against existing / competing solutions like formal query languages, form-based
-interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities, semantic 
-inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
+Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
+a century back, into the early 1970s. Two decades after the first major research systems where developed
+in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
+state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
+associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
+form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
+semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
 metrics today.
@@
 
-Early research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and expensive
+Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and expensive
 to maintain / transfer across database systems. These rule engines mostly relied on the systematic identification
 of linguistic patterns / were trying to template SQL from information that was derived from processing the natural
-language query. \citep*{Rendevouz, Lunar, Ladder} These approaches mostly tried to formalize natural language queries into 
-formal grammers which could then be deterministically mapped into a valid SQL query. These approaches have strong downsides
-when it comes to the variety of natural language constructs they can process, aswell as runtime adoption of new / unknown
-databases, query constructs etc. A potential upside of this class of NL2SQL systems is that they can confidently and 
-reproducably identify questions they can, and can't answer — thus leading to very reliable and predictable user interfaces.
+language query. \citep{Rendevouz, Lunar, Ladder} These approaches mostly tried to formalize natural language queries into 
+formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These approaches have
+strong downsides when it comes to the variety of natural language constructs they can process, aswell as runtime adoption of 
+new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is that they can confidently 
+and reproducably identify questions they can, and can't answer — thus leading to very reliable and predictable user interfaces.
+
+\subsubsection{Semantic Parsing using String-Kernels}
+
+A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
+\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
+it comes to fusing discrete parsing approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and learning / 
+training approaches to understand unseen language patterns / unknown natural language query structures. This allowed for more 
+flexible pattern recognition when compared to traditional rule-based systems.
+
+The core innovative characteristic of this approach lies in its capability to understand similarities between natural language 
+expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the research NLIDB
+system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise (e.g. spelling mistakes) 
+in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world datasets, this approach 
+compared favorably to existing systems of the time like \textsc{Chill}, especially in handling noisy inputs — a frequent 
+challenge rigid rule-based systems faced in real world scenarios. \citep{StringKernels, ILPParsing}
 

### 2025-04-26 thesis/thesis.tex
Graph matching and interactive systems

@@
 
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and expensive
-to maintain / transfer across database systems. These rule engines mostly relied on the systematic identification
-of linguistic patterns / were trying to template SQL from information that was derived from processing the natural
-language query. \citep{Rendevouz, Lunar, Ladder} These approaches mostly tried to formalize natural language queries into 
-formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These approaches have
-strong downsides when it comes to the variety of natural language constructs they can process, aswell as runtime adoption of 
-new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is that they can confidently 
-and reproducably identify questions they can, and can't answer — thus leading to very reliable and predictable user interfaces.
+Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
+expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
+identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
+natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
+queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
+approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
+runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
+that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
+and predictable user interfaces.
 
@@
 \citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing discrete parsing approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and learning / 
-training approaches to understand unseen language patterns / unknown natural language query structures. This allowed for more 
-flexible pattern recognition when compared to traditional rule-based systems.
+it comes to fusing discrete parsing approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
+learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
+allowed for more flexible pattern recognition when compared to traditional rule-based systems.
 
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural language 
-expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the research NLIDB
-system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise (e.g. spelling mistakes) 
-in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world datasets, this approach 
-compared favorably to existing systems of the time like \textsc{Chill}, especially in handling noisy inputs — a frequent 
-challenge rigid rule-based systems faced in real world scenarios. \citep{StringKernels, ILPParsing}
+The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
+language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
+research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
+(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
+datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
+noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
+ILPParsing}.
+
+\subsubsection{Graph Matching Methods}
+
+\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
+to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
+language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
+relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
+the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
+allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
+which implies easier scalability / transferability across domains, since the system does not require any additional 
+tweaks.
+
+Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
+was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
+accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
+Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
+representation of relationships and database layouts — more complicated database designs, which may not match the users 
+intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
+
+\subsubsection{Interactive Systems}
+
+In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
+due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
+questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
+\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
+is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
+how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
+
+NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
+complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
+NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
+misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
+to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
+to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
+possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
+multiple possible interpretations for a natural language query and translate all of them in natural language for the
+user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
 

### 2025-04-26 thesis/thesis.tex
Query synthesis

@@
 
+\subsubsection{Query Synthesis}
+
+\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
+presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
+with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
+natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
+synthesis to complete the sketch and finally using automated repair, if required.
+
+\citeauthor*{SQLizer} show that alternating between quantitative repair and synthesis yields results that beat
+state-of-the-art approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge
+of the underlying schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the
+correct query in the top 5 results for about 90\% of cases, significantly outperforming NaLIR \citep{NALIR},
+the previous state-of-the-art system \citep{SQLizer}.
+
+Potential short commings of this approch include queries which yield empty results, dealing with language variations
+as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
+to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
+
 \subsection{Neural NL2SQL Approaches}

### 2025-04-27 thesis/thesis.tex
Write a chapter about limitations

@@
 
-\citeauthor*{SQLizer} show that alternating between quantitative repair and synthesis yields results that beat
-state-of-the-art approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge
-of the underlying schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the
-correct query in the top 5 results for about 90\% of cases, significantly outperforming NaLIR \citep{NALIR},
+\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
+NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
+schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
+top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
 the previous state-of-the-art system \citep{SQLizer}.
@@
 to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
+
+\subsubsection{Limitations of Traditional Approaches to NL2SQL}
+
+Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
+challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
+benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
+benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
+its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
+fundamental challenges emerged / remained with these approaches:
+
+\begin{enumerate}
+    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
+          process the a small subset of the natural language they were programmed for. This severely limited their 
+          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
+          Lunar, Ladder}.
+    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
+          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
+          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
+    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
+          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
+          of problems \citep{StringKernels, SQLizer}.
+    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
+          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
+          complexity more compute was required to resolve the natural language query to a suitable query candidate
+          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
+          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
+          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
+          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
+          Freebase.'' which indicates underlying system design issues with runtime complexity.
+\end{enumerate}
+
+These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
+increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
+interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
+ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
+conversational approach), NLIDBs were and are not considered to be a solved problem.
 

### 2025-05-01 thesis/thesis.tex
Start with neural approaches

@@
 outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with discrete programming approaches.
+challenging / yielded unsatisfying results with logical programming approaches.
 
@@
 Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These discrete
-approaches have laid the foundations for transitioning towards the application of machine learning techniques
-for NL2SQL.
+traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
+programming approaches have laid the foundations for transitioning towards the application of machine learning
+techniques for NL2SQL.
 
@@
 \citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing discrete parsing approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
+it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
 learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
@@
 
+The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
+branch around neural network application forward to step in and propose new solutions which address the brittleness, 
+transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
+significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
+field.
+
+\subsubsection{Early Neural Approaches}
+
+In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB research. 
+Seq2SQL was an early research system that in the field of neural network application and as one of the first papers to frame the
+implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query execution in the
+reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced WikiSQL, a training dataset,
+which enables large scale (in \citeyear{Seq2SQL}) model training.
+
+SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being a derivative
+approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research threads, employing a
+sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable) sub-queries which can then be individually
+sketched and refined, yielding a system that outperformed state-of-the-art by 9\% to 13\% \citep{SQLNet}.
+
 \subsection{Benchmark Evolution}
@@
 
+% Prompt engineering and fine tuning
+
 \subsection{Optimization for LLM-based NL2SQL}
 
-\subsection{Complex Techniques in NL2SQL}
+% PICARD, DIN-SQL, MAGIC & CHASE
 
 \subsection{Research Gaps}
+
+\subsubsection{Deployment Gaps}
 

### 2025-05-02 thesis/thesis.tex
Further sections on neural network approaches

@@
 
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB research. 
-Seq2SQL was an early research system that in the field of neural network application and as one of the first papers to frame the
-implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query execution in the
-reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced WikiSQL, a training dataset,
-which enables large scale (in \citeyear{Seq2SQL}) model training.
+In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
+research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
+to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
+execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
+WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
 
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being a derivative
-approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research threads, employing a
-sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable) sub-queries which can then be individually
-sketched and refined, yielding a system that outperformed state-of-the-art by 9\% to 13\% \citep{SQLNet}.
+SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
+a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
+threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
+sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
+to 13\% \citep{SQLNet}.
+
+\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
+\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
+for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
+SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
+accurate SQL queries from user input \citep{TypeSQL}.
+
+\subsubsection{Intermediate Neural Developments}
+
+Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
+which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
+complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
+This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
+nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
+\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
+5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
+comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
+
+Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
+representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
+format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
+accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
+layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
+IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
+of 46.7\% \citep{IRNet}.
+
+Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
+database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
+This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
+evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} \&
+\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
+performance against IRNet by 6\% \citep{IRNet, GNN}.
 

### 2025-05-03 thesis/thesis.tex
RATSQL

@@
 
+\subsubsection{Relation-Aware Transformer Approaches}
+
+The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
+neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
+and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
+linking} \citep{RATSQL}.
+
+RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
+natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
+while generating the query.
+
+This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
+outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
+accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
+popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
+RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
+system \citep{RATSQL}.
+
+\subsubsection{Comparative Analysis of Neural Approaches}
+
+The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
+neural NL2SQL approaches:
+
 \subsection{Benchmark Evolution}

### 2025-05-04 thesis/thesis.tex
Finish the section on neural approaches

@@
 The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches:
+neural NL2SQL approaches in different dimensions:
 
-\subsection{Benchmark Evolution}
+\begin{enumerate}
+    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
+          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
+          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
+          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
+          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
+          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
+    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
+          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
+          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
+          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
+          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
+          training data \citep{Spider, RATSQL}.
+    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
+          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
+          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
+          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
+          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
+    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
+          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
+          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
+          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
+          by up to 10.5\% \citep{IRNet, RATSQL}.
+    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
+          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
+          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
+          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
+\end{enumerate}
 
-\subsection{LLM-based NL2SQL Systems}
+\subsection{Limitations of Neural Approaches}
 
+Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
+approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
+
+\begin{enumerate}
+    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
+          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
+    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
+          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
+          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
+    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
+          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
+          specific natural language queries challenging.
+    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
+          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
+\end{enumerate}
+
+The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of NLIDBs. The
+research shift started in this era established the foundations for further and more advanced machine learning approaches (specifically language
+model oriented approaches )being researched. Neural approaches showed to significant improvements in performance when being paired with pre-trained
+language models \citep{RATSQL} which led to further research on their applicability.
+
+\subsection{Pre-trained Language Models}
+
+% PLM approaches
+% RESDSQL, Graphix-T5
+% T5 fine-tuning etc.
+
+% https://arxiv.org/abs/2109.05093
+
+\subsection{Large Language Models}
+
+% CodeS
+% GPT-X based
+% 
 % Prompt engineering and fine tuning

### 2025-05-07 thesis/thesis.tex
Write about pre trained language models

@@
 % 1/3 Vorarbeit, Literature Review
-% 1/3 Decomp und System Design
+% 1/3 Theory, Decomp und System Design
 % 1/3 Implementation und Evaluation
@@
 
-\subsection{Limitations of Neural Approaches}
+\subsubsection{Limitations of Neural Approaches}
 
@@
 
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of NLIDBs. The
-research shift started in this era established the foundations for further and more advanced machine learning approaches (specifically language
-model oriented approaches )being researched. Neural approaches showed to significant improvements in performance when being paired with pre-trained
-language models \citep{RATSQL} which led to further research on their applicability.
+The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
+NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
+(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
+when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
 
@@
 
-% PLM approaches
-% RESDSQL, Graphix-T5
+The advantages of combining specialized neural networks with general-purpose pre-trained language models
+led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
+language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
+comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
+of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
+language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
+\citep{SPIDER2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
+through the language models's ability to understand patterns and identify semantic relationships of natural language
+query elements.
+
+\subsubsection{Early Pre-trained Language Model Adaptations}
+
+The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
+of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
+a far more sophisticate architecture when it comes to natural language analysis. The application of T5-3B for NL2SQL
+tasks which yielded 71.4\% execution accuracy presented a breakthrough in this research domain \citep{T2SQL-LLM-Bench-3}.
+This demonstrated that PLMs are correctly positioned to significantly outperform existing approaches.
+
+In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
+a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
+It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
+which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
+data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
+database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
+\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
+
+Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
+generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
+of users when querying and correctly identify database schema elements in natural language queries. Thus improving
+semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
+\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
+\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
+column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
+while being significantly cheaper to train \citep{STRUG}.
+
+\subsubsection{Intermediate Pre-trained Language Model Approaches}
+
+% RYANSQL, GAZP
+% RESDSQL, GRAPHIX, NATSQL
+
 % T5 fine-tuning etc.
+% https://arxiv.org/pdf/2302.05965
 
-% https://arxiv.org/abs/2109.05093
+\subsubsection{Constrained Decoding and Ranking Techniques}
+
+% PICARD etc.
+
+\subsubsection{Comparison with Large Language Models}
 

### 2025-05-08 thesis/thesis.tex
Write about advanced pre trained language approaches

@@
 
-% Theoretical Foundations
 \section{Literature Review}
+
+% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
 
@@
 of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis. The application of T5-3B for NL2SQL
-tasks which yielded 71.4\% execution accuracy presented a breakthrough in this research domain \citep{T2SQL-LLM-Bench-3}.
-This demonstrated that PLMs are correctly positioned to significantly outperform existing approaches.
+a far more sophisticate architecture when it comes to natural language analysis.
 
@@
 
-\subsubsection{Intermediate Pre-trained Language Model Approaches}
+In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
+\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
+/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
+Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
+for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
+approach enables a improvement in robustness and accuracy in situations where training and inference environments
+differ without requiring manually annotated examples \citep{GAZP}.
 
-% RYANSQL, GAZP
-% RESDSQL, GRAPHIX, NATSQL
+\subsubsection{Advanced Pre-trained Language Model Approaches}
 
-% T5 fine-tuning etc.
-% https://arxiv.org/pdf/2302.05965
+Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
+systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
+
+\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
+which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
+transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
+flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
+its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
+to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
+approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
+which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
+
+A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
+(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
+yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
+This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
+only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
+
+Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
+the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
+of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
+represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
+This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
+GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
+in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
+\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
+numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
+
+In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
+\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
+faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
+\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
+before passing them to the model for query generation, which reduced noise (when working with large database schemas)
+and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
+performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
+execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
+by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
+robustness gain decoupeled architectures have over model-oriented approaches.
+
+These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
+mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
+PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
+benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
+and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
 
@@
 
-\subsection{Optimization for LLM-based NL2SQL}
-
+%\subsection{Optimization for LLM-based NL2SQL}
 % PICARD, DIN-SQL, MAGIC & CHASE
@@
 
-\section{Decomposition \& Requirements}
+\section{Theoretical Foundations}
 
@@
 % Conceptual Design
-\section{System Design}
+\subsection{System Design}
 
-\subsection{Architecture Design}
-\subsubsection{Interface Design}
-\subsubsection{Data Model}
-\subsubsection{Integration into SQL}
+\subsubsection{Architecture Design}
 

### 2025-05-08 thesis/thesis.tex
Comparison

@@
 
-% PICARD etc.
+A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
+but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
+(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
+for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
+rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
+language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
+on the \textsc{Spider} benchmark \citep{PICARD}.
+
+\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
+evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
+incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
+output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
+with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
+SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
+
+The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
+encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
+to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
+
+\subsubsection{Advantages of PLM Approaches}
+
+PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
+leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
+large language models in the following years. Namely PLM approaches brought a series of upsides with them:
+
+\begin{enumerate}
+    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
+        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
+        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
+    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
+        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
+        approaches \citep{GRAPPA, STRUG}.
+    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
+        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
+        primarily trained on the development sets of contemporary benchmarks.
+\end{enumerate}
+
+\subsubsection{Limitations of PLM Approaches}
+
+Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
+non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
+must be considered when using a PLM-based approach to NL2SQL:
+
+\begin{enumerate}
+
+    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific, fine-tuning, limiting a 
+          straight forward adaptation to new domains or databases. Although being significantly more efficient than LLM-based approaches the potential need 
+          for initial fine-tuning represent a significant computational resource burden. Furthermore when not using synthetic data generation (e.g. 
+          \textsc{Grappa}) annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
+    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often far larger than needed 
+          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each decoding step, they can
+          produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the input and output token space, therefore multiple
+          approaches have been researched which focus on constraining these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard}
+          have proposed potential (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
+    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to incorporate limited amounts
+          of schema awareness when being applied out of the box for NL2SQL tasks. Multiple research efforts focused on improving this situation,
+          most notably \textsc{Resdsql} and GRAPHIX-T5 tried to improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX},
+          nonetheless the non-specialized nature of PLMs prevents NL2SQL being part of the fundamental model architecture.
+\end{enumerate}
+
+These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially in direct comparison with neural 
+approaches), ultimately yielding the research domain to transition toward exploring Large Language Model approaches that promise even greater flexibility
+in adaptation and potentially superior handling of complex queries through advanced in-context learning approaches.
 
 \subsubsection{Comparison with Large Language Models}
+
+% https://raw.githubusercontent.com/HKUSTDial/NL2SQL_Handbook/refs/heads/main/assets/river.svg
+% https://arxiv.org/abs/2109.05093
 

### 2025-05-11 thesis/thesis.tex
Write a section on ICL

@@
 research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, T2SQL-LLM-Bench, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.
+effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.
 
@@
 
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific, fine-tuning, limiting a 
-          straight forward adaptation to new domains or databases. Although being significantly more efficient than LLM-based approaches the potential need 
-          for initial fine-tuning represent a significant computational resource burden. Furthermore when not using synthetic data generation (e.g. 
-          \textsc{Grappa}) annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each decoding step, they can
-          produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the input and output token space, therefore multiple
-          approaches have been researched which focus on constraining these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard}
-          have proposed potential (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to incorporate limited amounts
-          of schema awareness when being applied out of the box for NL2SQL tasks. Multiple research efforts focused on improving this situation,
-          most notably \textsc{Resdsql} and GRAPHIX-T5 tried to improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX},
-          nonetheless the non-specialized nature of PLMs prevents NL2SQL being part of the fundamental model architecture.
+    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
+          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
+          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
+          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
+          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
+    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
+          far larger than needed 
+          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
+          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
+          input and output token space, therefore multiple approaches have been researched which focus on constraining
+          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
+          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
+    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
+          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
+          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
+          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
+          of PLMs prevents NL2SQL being part of the fundamental model architecture.
 \end{enumerate}
 
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially in direct comparison with neural 
-approaches), ultimately yielding the research domain to transition toward exploring Large Language Model approaches that promise even greater flexibility
-in adaptation and potentially superior handling of complex queries through advanced in-context learning approaches.
+These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
+in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
+Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
+of complex queries through advanced in-context learning approaches.
 
@@
 
-% https://raw.githubusercontent.com/HKUSTDial/NL2SQL_Handbook/refs/heads/main/assets/river.svg
-% https://arxiv.org/abs/2109.05093
+The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
+PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
+significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
+learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
+The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
+in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
+data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
+where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
+that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
+less of a central problem to solve \citep{DINSQL}.
 
@@
 
-% CodeS
-% GPT-X based
-% 
-% Prompt engineering and fine tuning
+The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
+with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
+\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
+could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
+training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
+increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
 
-%\subsection{Optimization for LLM-based NL2SQL}
-% PICARD, DIN-SQL, MAGIC & CHASE
+\subsubsection{In-Context Learning}
+
+In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
+inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
+This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
+instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
+
+The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
+of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
+mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
+and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
+the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
+/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
+
+Example selection strategies showed to have a significant impact on ICL performance. \cite{DAIL-SQL} evaluated
+various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
+\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
+propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
+query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
+\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
+token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
+state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
+
+The comparison between zero-shot and few-shot performance reveals the significant impact of supplying in-context examples
+to models. While LLMs demonstrate impressive zero-shot capabilities, with models like GPT-4 achieving 72.3\% execution
+accuracy on \textsc{Spider} without any examples \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning shows to substantially
+improve execution accuracy. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's execution accuracy to 80.2\%,
+representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}. This improvement
+in execution accuracy is particularly noticable for complex queries which involve multiple joins or nested subqueries,
+where zero-shot approaches often fail generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the
+leap in exact match ratio measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to
+71.9\% with five-shot. The results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution
+accuracy of $k$-shot approaches.
+
+The effectiveness of ICL approaches has established them as the standard technique applied in LLM-based NL2SQL systems.
+They provide a robust foundation which can be extended with advanced techniques such as problem-decomposition, self-correction
+and chain-of-thoughts. ICL alone is still subject to frequent limitations observed with NL2SQL systems like ambiguoity handling,
+schema linking and dealing with complex queries. This motivated further research of LLM-based NL2SQL systems.
 
@@
 
-\nocite{*}
-
 \bibliographystyle{apacite}

### 2025-05-11 thesis/thesis.tex
Rewrite ICL

@@
 
-Example selection strategies showed to have a significant impact on ICL performance. \cite{DAIL-SQL} evaluated
+Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
 various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
@@
 
-The comparison between zero-shot and few-shot performance reveals the significant impact of supplying in-context examples
-to models. While LLMs demonstrate impressive zero-shot capabilities, with models like GPT-4 achieving 72.3\% execution
-accuracy on \textsc{Spider} without any examples \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning shows to substantially
-improve execution accuracy. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's execution accuracy to 80.2\%,
-representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}. This improvement
-in execution accuracy is particularly noticable for complex queries which involve multiple joins or nested subqueries,
-where zero-shot approaches often fail generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the
-leap in exact match ratio measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to
-71.9\% with five-shot. The results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution
-accuracy of $k$-shot approaches.
+The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
+to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
+(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
+still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
+execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
 
-The effectiveness of ICL approaches has established them as the standard technique applied in LLM-based NL2SQL systems.
-They provide a robust foundation which can be extended with advanced techniques such as problem-decomposition, self-correction
-and chain-of-thoughts. ICL alone is still subject to frequent limitations observed with NL2SQL systems like ambiguoity handling,
-schema linking and dealing with complex queries. This motivated further research of LLM-based NL2SQL systems.
+Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
+often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
+fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
+measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
+results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
+
+This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
+approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
+\citep{XiYan, CHASE, DINSQL}.
 

### 2025-05-16 thesis/thesis.tex
Write about refinement loops and candidate selection

@@
 
+\subsubsection{Self-Correction and Iterative Refinement}
+
+\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
+sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
+on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
+correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
+errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
+errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
+approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
+impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
+self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
+\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
+self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
+and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
+represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
+how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
+
+Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
+which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
+prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
+refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
+phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
+approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
+\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
+MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
+human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
+significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
+
+While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
+Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
+for NL2SQL systems offer advantages when it comes to human-in-the-loop refinement. Interactive information seeking from the user
+has showing to be an effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than
+approaching each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require
+only small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
+ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
+
+\subsubsection{Candidate Selection Frameworks}
+
+Recent approaches for NL2SQL have increasingly emphasized the generation of multiple candidates and selectiion of the most promising
+one amongst them. Candidate selection strategies have shown significant performance improvements on challenging benchmarks. This
+approach acknowledges the inherent difficulty of generating perfect SQL queries in a single attempt, even for sophisticated LLMs
+and self-correction mechanisms in place.
+
+\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
+query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
+CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
+language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
+utilizes execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
+learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
+characteristics. For candidate selection CHASE utilizes a fine-tuned LLM that can do binary selection of candidates.
+\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
+state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
+
+Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which utilizes a multi-generation
+ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
+fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
+representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
+\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
+during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
+fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
+generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
+to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
+during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
+able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
+\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
+
+Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generate and training specialized models for candidate
+selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
+The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
+the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
+to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
+models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
+transferability with the robustness of specialized model architectures for candidate ranking and selection.
+
 \subsection{Research Gaps}
@@
 
+\subsection{Ablation Study}
+
 \subsection{Comparison with Alternative Approaches}

### 2025-05-16 thesis/thesis.tex
Fix monotonic lang

@@
 Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems offer advantages when it comes to human-in-the-loop refinement. Interactive information seeking from the user
-has showing to be an effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than
-approaching each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require
-only small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
+for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
+effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
+each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
+small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
 ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
@@
 
-Recent approaches for NL2SQL have increasingly emphasized the generation of multiple candidates and selectiion of the most promising
-one amongst them. Candidate selection strategies have shown significant performance improvements on challenging benchmarks. This
-approach acknowledges the inherent difficulty of generating perfect SQL queries in a single attempt, even for sophisticated LLMs
-and self-correction mechanisms in place.
+Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
+promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
+These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
+even with capable LLMs and modern self-correction mechanisms.
 
@@
 language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-utilizes execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
+inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
 learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE utilizes a fine-tuned LLM that can do binary selection of candidates.
+characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
 \citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
@@
 
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which utilizes a multi-generation
+Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
 ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
@@
 
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generate and training specialized models for candidate
+Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
 selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
@@
 transferability with the robustness of specialized model architectures for candidate ranking and selection.
+
+\vspace{5cm}
+
+% AB HIER WEITER PLS
+%CHASE-SQL: Multi-path reasoning and candidate selection \citep{CHASE}
+%XiYan-SQL: Multi-generator ensemble framework \citep{XiYan}
+%Agent specialization and collaboration patterns
+%Voting and selection mechanisms for query ranking
+
+\subsubsection{Retrieval-Augumented Generation}
+
+Integration of retrieval systems with LLMs
+DB-GPT: RAG for database interactions \citep{DBGPT}
+Schema and documentation retrieval strategies
+Impact on handling large/complex database schemas
+
+\subsubsection{Schema Representation}
+
+Chain-of-thought reasoning for complex queries
+Schema representation in prompts
+
+% heute goal
+
+\subsubsection{Specialized LLMs and Fine-tuning}
+
+CodeS: Open-source LLMs optimized for SQL \citep{CodeS}
+Domain adaptation challenges and solutions \citep{ModelDomainAdaption}
+Trade-offs between general-purpose and specialized models
+Fine-tuning strategies for improved performance
+
+\subsubsection{Benchmarking}
+
+Contemporary benchmarks: BIRD, Spider 2.0 \citep{BIRD, SPIDER2}
+LLM-specific evaluation challenges \citep{LLM-SQL}
+Real-world deployment considerations \citep{NL2SQLUnsolved}
+Performance metrics beyond accuracy (efficiency, cost, latency)
+
+\subsubsection{Limitations and Challenges}
+Hallucination and accuracy concerns
+Computational resource requirements
+Data privacy and security implications
+Ambiguity handling in complex scenarios
+Comparison of limitations with PLM approaches
+
+% https://raw.githubusercontent.com/HKUSTDial/NL2SQL_Handbook/refs/heads/main/assets/river.svg
+% https://arxiv.org/abs/2109.05093
+
+% CodeS
+% GPT-X based
+% 
+% Prompt engineering and fine tuning
+
+%\subsection{Optimization for LLM-based NL2SQL}
+% PICARD, DIN-SQL, MAGIC & CHASE
 

### 2025-05-17 thesis/thesis.tex
Write about RAG

@@
 
-\vspace{5cm}
-
-% AB HIER WEITER PLS
-%CHASE-SQL: Multi-path reasoning and candidate selection \citep{CHASE}
-%XiYan-SQL: Multi-generator ensemble framework \citep{XiYan}
-%Agent specialization and collaboration patterns
-%Voting and selection mechanisms for query ranking
+% Maybe: Agent specialization and collaboration patterns
+% Maybe: Voting and selection mechanisms for query ranking
 
@@
 
-Integration of retrieval systems with LLMs
-DB-GPT: RAG for database interactions \citep{DBGPT}
-Schema and documentation retrieval strategies
-Impact on handling large/complex database schemas
+Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
+external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
+papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
+NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
+aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
+serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
+\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
+the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
+yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
+\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
+and Gemini models \citep{XiYan}.
 
-\subsubsection{Schema Representation}
+A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
+\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
+building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
+retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
+By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
+execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
+\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
 
-Chain-of-thought reasoning for complex queries
-Schema representation in prompts
+The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
+approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
+fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
+relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
+in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
+emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
+execution accuracy.
 
-% heute goal
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-CodeS: Open-source LLMs optimized for SQL \citep{CodeS}
-Domain adaptation challenges and solutions \citep{ModelDomainAdaption}
-Trade-offs between general-purpose and specialized models
-Fine-tuning strategies for improved performance
-
-\subsubsection{Benchmarking}
-
-Contemporary benchmarks: BIRD, Spider 2.0 \citep{BIRD, SPIDER2}
-LLM-specific evaluation challenges \citep{LLM-SQL}
-Real-world deployment considerations \citep{NL2SQLUnsolved}
-Performance metrics beyond accuracy (efficiency, cost, latency)
-
-\subsubsection{Limitations and Challenges}
-Hallucination and accuracy concerns
-Computational resource requirements
-Data privacy and security implications
-Ambiguity handling in complex scenarios
-Comparison of limitations with PLM approaches
-
-% https://raw.githubusercontent.com/HKUSTDial/NL2SQL_Handbook/refs/heads/main/assets/river.svg
-% https://arxiv.org/abs/2109.05093
-
-% CodeS
-% GPT-X based
-% 
-% Prompt engineering and fine tuning
-
-%\subsection{Optimization for LLM-based NL2SQL}
-% PICARD, DIN-SQL, MAGIC & CHASE
+The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
+and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
+As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
+Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
+demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
+performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
 

### 2025-05-18 thesis/thesis.tex
Write about CodeS

@@
 
+\subsubsection{Specialized LLMs and Fine-tuning}
+
+While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
+domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
+tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
+(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
+yielded a series of dedicated models for optimized SQL generation.
+
+A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
+LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
+contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
+(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
+\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
+for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
+contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
+to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
+on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
+aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
+on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
+and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
+research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
+approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
+\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
+
+\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
+could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
+(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
+\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
+highly practical for real-world deployments where computation resource are constrained.
+
 \subsection{Research Gaps}

### 2025-05-19 thesis/thesis.tex
Benchmarks

@@
 
+\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
+models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
+performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
+achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
+on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
+comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
+in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
+that even higher accuracy scores are possible.
+
+\subsection{Benchmarking}
+
+In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
+benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
+useful indicator for the state of the research field. In the past decade significant advancements have been made with
+\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
+\citep{Spider}.
+
+\subsubsection{Spider}
+
+\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
+evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
+spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
+datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
+With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
+\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
+models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
+challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
+accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
+\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
+
+\subsubsection{Bird}
+
+The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
+and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
+with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
+size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
+schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
+knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
+places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
+LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
+highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
+
+\subsubsection{Spider 2.0}
+
+\textsc{Spider 2.0} which was introduced by \citeauthor*{SPIDER2} in \citeyear{SPIDER2} represents the most recent
+advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
+primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
+which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
+databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
+towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
+and project-level codebases \citep{SPIDER2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
+queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
+to analytics. \citeauthor{SPIDER2} further highlights the gap between academic research and enterprise-level environments
+with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
+\citep{SPIDER, SPIDER2}.
+
 \subsection{Research Gaps}

### 2025-05-19 thesis/thesis.tex
Write research gaps section

@@
 research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, SPIDER2, BIRD}.
+effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
 
@@
 This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} \&
+evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
 \textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
@@
           \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{Spider, RATSQL}.
+          training data \citep{, RATSQL}.
     \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
@@
 language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{SPIDER2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
+\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
 through the language models's ability to understand patterns and identify semantic relationships of natural language
@@
 
+\subsubsection{Limitations and Challenges}
+
+Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
+are apparent that impact their practical implementation and real-world deployability.
+
+\begin{enumerate}
+    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
+           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
+           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
+           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
+           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
+           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
+           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
+           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
+           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
+    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
+           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
+           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
+           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
+           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
+           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
+           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
+           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
+           closed-source competitors. 
+    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
+           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
+           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
+           needs to be communicated to external parties in order to form a usable system if closed source models are
+           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
+           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
+           requirements.
+    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
+           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
+           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
+           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
+           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
+           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
+           rank their results \citep{CHASE, XiYan}.
+    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
+           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
+           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
+           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
+           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
+           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
+           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
+           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
+           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
+           their transferability and natural language coverage.
+\end{enumerate}
+
 \subsection{Benchmarking}
@@
 
-\textsc{Spider 2.0} which was introduced by \citeauthor*{SPIDER2} in \citeyear{SPIDER2} represents the most recent
+\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
 advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
@@
 towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{SPIDER2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
+and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
 queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{SPIDER2} further highlights the gap between academic research and enterprise-level environments
+to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
 with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{SPIDER, SPIDER2}.
+\citep{Spider, Spider2}.
 
@@
 
-\subsubsection{Deployment Gaps}
+This literature review highlights that significant progress was made in the development of real world feasible NL2SQL systems
+and that the research domain has undergone multiple large paradigm shifts from rule-based to neural-network-based to PLM-based
+and most recently to LLM-based approaches. Despite these advancement several critical research gaps remained open or emerged
+which limit the practical deployment of NL2SQL systems and their widespread adoption as natural language database interfaces
+in real world scenarios.
+
+\subsubsection{Advanced Open-Source Approaches}
+
+While recent research papers have made remarkable progress on both open-source model development (\textsc{CodeS} and
+\textsc{OmniSQL}) aswell as advanced architecture development like CHASE and \textsc{XiYan-SQL} (which utilize multi-path
+generation and candidate selection, self-correction and RAG), a significant research gap evolved around state-of-the-art
+approaches that dont rely on proprietary LLMs and only utilize open-source models. Most contemporary research proposes
+solutions which (partially) rely on the baseline proprietary LLM in order to achieve state-of-the-art results. Whilst
+these efforts give meaningful signal to the effect that specific prompt engineering techniques, self-correction mechanisms
+and candidate selection models have, they are not yet feasible for real world scenarios due to the cost and data privacy
+concerns they introduce.
+
+\begin{enumerate}
+    \item \textbf{Limited Exploration of Technique Synergies} — Contemporary research primarily focuses on techniques like
+          candidate selection, RAG, and self-correction in isolation or with specific closed-source LLMs. A gap remains
+          in understanding how these techniques can be optimally combined, especially with emerging open-source models.
+          For instance, the potential of synergy of specialized open-source LLM like \textsc{OmniSQL} with candidate-selection,
+          advanced self-correction and subset-encoding of database schemas has yet to be researched.
+    \item \textbf{Efficiency-Accuracy Tradeoffs} — Whilst recent research efforts focus primarily on achieving a new
+          state-of-the-art execution accuracy metric on prevalent benchmarks, the relationship of computational requirements and
+          performance gains achieved remains unclear. Some research papers present ablation studies indicating the relevance
+          of certain architecture components but the overall relationship between state-of-the-art performance and cost
+          is still underexplored. 
+     \item \textbf{Cross-Technique Optimization} — There is limited research on how to optimize the interplay between the various
+          NL2SQL techniques. For example, it is yet unclear how subset-encoding of database schemas might impact the effectiveness
+          of multi-path generation and candidate selection, or whether synthetic example generation for in-context-learning is effective
+          when working with specialized models versus general-purpose LLMs.
+\end{enumerate}
+
+\subsubsection{Deployment and Performance Gaps}
+
+Despite impressive academic results, significant gaps remain in transitioning NL2SQL systems from research environments
+to real-world, enterprise-feasible, deployments:
+
+\begin{enumerate}
+    \item \textbf{Database Integration} — While existing research has often focused on standalone systems, little attention
+          has been given to the integration complexities between NL2SQL capabilities and database management systems like PostgreSQL.
+          Having two standalone systems imposes a significant data transfer need between the two systems when infact, natural language
+          queries can be treated as an extension to most existing databases. This implementation gap prevents seamless adoption into
+          existing databases as it requires additional software layers which in turn increase the overall complexity and cost.
+    \item \textbf{Hardware Requirement Optimization} — As mentioned above, current LLM-based approaches often require significant
+          compute resources in order to achieve state-of-the-art results. There is limited research available on performance optimization
+          of NLIDBs in order to achieve practical and industry-viable hardware constraints.
+    \item \textbf{Real-time Performance Considerations} — Most research papers evaluate models based on their accuracy metrics alone
+          without factoring in the latency and throughput characteristics of their solutions. This imposes a possibly significant $\delta$
+          between academic research and production environments. Responsiveness is an important metric for user experience and should
+          therefore play a role in NLIDB research when it comes to evaluating different NL2SQL architectures.
+    \item \textbf{Privacy and Security} — While open-source models address some privacy concerns by enabling a local deployment of LLMs,
+          research gaps remain in ensuring that NL2SQL systems respect database access controls and security policies. This is especially
+          important for enterprise and government environments where data access is strictly regulated.
+\end{enumerate}
+
+\subsubsection{Ambiguity Resolution and Semantic Accuracy}
+
+Despite the long existance of the research field, fundamental questions remain open with regards to handling ambiguity effectively in natural
+language queries. Current, especially language model based, systems struggle to correctly identify when natural language queries contain ambiguities
+that they cant confidently resolve. Although some approaches like multi-path generation and candidate selection are a promising way
+to work around ambiguous language, early systems like \textsc{NaLIR} showed that the most effective way to deal with inherently ambiguous language
+is asking clarifying questions, instead of trying to interpret the query best-effort \citep{NALIR}. Ambiguous language is an inherent source of
+inaccuracy and therefore a cause for misleading query results. Contemporary research and benchmarks dont focus ambiguity detection and
+strategies resolution which therefore leaves open questions to be further researched.
+
+\subsubsection{Evaluation and Benchmarking Gaps}
+
+\begin{enumerate}
+    \item \textbf{Enterprise-grade Evaluation} — As highlighted by \textsc{Spider 2.0}, there is a gap between academic benchmarks
+          and enterprise realities. Further research is needed to create evaluation frameworks that better represent real-world enterprise
+          environments with thousands of tables and complex relationships. \cite{Spider2} is a promising first step is this direction.
+    \item \textbf{Performance Metrics} — As mentioned above, current benchmarks often dont capture latency or throughput of NL2SQL systems
+          at all, allowing for solutions to achieve state-of-the-art scores that require significantly more resources than their predecessors.
+          Having meaningful performance metrics and benchmarking would allow to further analyze the proposed solutions for their real world
+          feasibility.
+    \item \textbf{User Experience Metrics} — Contemporary benchmarks focus on execution accuracy without assessing potential user satisfaction,
+          trust, and overall experience. Metrics that capture these aspects are necessary for understanding the actual utility of NL2SQL
+          systems in practice. Although execution accuracy is a big factor for the trustworthiness of a NLIDB, it is not the only component
+          as indicated by \citeauthor*{NALIR} in \citeyear{NALIR}.
+\end{enumerate}
+
+\subsubsection{Thesis Placement}
+
+This thesis addresses several of the above outlined research gaps. The primary focus of this work is the integration of open-source
+NL2SQL models with advanced techniques like candidate selection, subset-encoding of database schemas, and synthetic example generation
+for in-context learning. Through the implementation of a PostgreSQL extension this work will bridge multiple critical gaps between
+theoretical advancements and their practical deployability into real world systems whilst exploring performance characteristics.
 

### 2025-05-29 thesis/thesis.tex
Write about relational theory

--- 2025-05-19 thesis/thesis.tex
@@
 \newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
+	\begin{center}Source: {#1}\end{center}
 }
@@
 popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
+RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by 
+this system \citep{RATSQL}.
 
@@
           (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
+          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity
+          of approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-
+          SQL presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted 
+          self attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
+    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their 
+          transferability. The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over 
+          the underlying database schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the 
+          \textsc{Spider} benchmark). With standard benchmarks emerging it became easier to verify and quantify which approach had 
+          the highest transferability as \textsc{Spider} specifically had independent development and test datasets, preventing 
+          approaches from over-optimizing on training data \citep{, RATSQL}.
+    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation
+          (Seq2SQL) their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized 
+          structured decoding, \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges 
+          around \textit{schema linking} outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle 
+          rephrasings, spelling mistakes and variations in natural language usage far beyond what traditional NL2SQL approaches 
+          could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
+    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex 
+          aggregations, nested structures and joins dramatically improved over the course of the research that happened in this 
+          field. Whilst \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural 
+          approaches to handle complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation 
+          approach introduced by \textsc{IRNet} by up to 10.5\% \citep{IRNet, RATSQL}.
+    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to 
+          end query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} 
+          showed novel and state-of-the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to 
+          accurately reason about user intent and traverse the database schema while generating queries \citep{TypeSQL, GNN, 
+          RATSQL}. 
 \end{enumerate}
@@
 
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
+Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late
+neural approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
 
 \begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
+    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data 
+          (ie. natural language paired with output SQL queries) than traditional systems which required serious efforts of data 
+          collection \citep{Spider}.
+    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models 
+          produced invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by 
+          utilizing syntax trees during decoding but nonetheless syntactic correctness remained a challenge across future
+          iterations of neural systems. \citep{SyntaxSQLNet}
+    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered
+          from a limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly 
+          domain specific natural language queries challenging.
+    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones 
+          with complex architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
 \end{enumerate}
 
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
+The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility
+of NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning 
+approaches (specifically language model oriented approaches )being researched. Neural approaches showed to significant 
+improvements in performance when being paired with pre-trained language models \citep{RATSQL} which led to further research on 
+their applicability.
 
@@
 to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
+approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of 
+\textsc{SQLNet} which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
 
@@
         and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
+    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific
+        understanding
+        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with
+        neural approaches \citep{GRAPPA, STRUG}.
+    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables
+        them to handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural 
+        approaches which primarily trained on the development sets of contemporary benchmarks.
 \end{enumerate}
@@
 The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
+in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on
+domain-specific data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the 
+inference environment, where model modificants are less impactful, due to the extensive pre-training that took place. Approaches 
+like \textsc{DinSQL} show that with the application of LLMs the engineering challenges around model instruction gained relevance 
+while model training became less of a central problem to solve \citep{DINSQL}.
 
@@
 
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
+The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early 
+experiments with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches 
+\citep{DAIL-SQL}. \cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any 
+fine-tuning efforts, could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that 
+required extensive training. This breakthrough challenged the contemporary assumption that further specialization of model 
+architectures would yield increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
 
@@
 still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
+execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-
+SQL}.
 
@@
 promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
+These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation
+mechanism, even with capable LLMs and modern self-correction mechanisms.
 
@@
           NL2SQL techniques. For example, it is yet unclear how subset-encoding of database schemas might impact the effectiveness
-          of multi-path generation and candidate selection, or whether synthetic example generation for in-context-learning is effective
-          when working with specialized models versus general-purpose LLMs.
+          of multi-path generation and candidate selection, or whether synthetic example generation for in-context-learning is 
+          effective when working with specialized models versus general-purpose LLMs.
 \end{enumerate}
@@
     \item \textbf{Database Integration} — While existing research has often focused on standalone systems, little attention
-          has been given to the integration complexities between NL2SQL capabilities and database management systems like PostgreSQL.
-          Having two standalone systems imposes a significant data transfer need between the two systems when infact, natural language
-          queries can be treated as an extension to most existing databases. This implementation gap prevents seamless adoption into
-          existing databases as it requires additional software layers which in turn increase the overall complexity and cost.
+          has been given to the integration complexities between NL2SQL capabilities and database management systems like 
+          PostgreSQL. Having two standalone systems imposes a significant data transfer need between the two systems when infact, 
+          natural language queries can be treated as an extension to most existing databases. This implementation gap prevents 
+          seamless adoption into existing databases as it requires additional software layers which in turn increase the overall 
+          complexity and cost.
     \item \textbf{Hardware Requirement Optimization} — As mentioned above, current LLM-based approaches often require significant
-          compute resources in order to achieve state-of-the-art results. There is limited research available on performance optimization
-          of NLIDBs in order to achieve practical and industry-viable hardware constraints.
-    \item \textbf{Real-time Performance Considerations} — Most research papers evaluate models based on their accuracy metrics alone
-          without factoring in the latency and throughput characteristics of their solutions. This imposes a possibly significant $\delta$
-          between academic research and production environments. Responsiveness is an important metric for user experience and should
-          therefore play a role in NLIDB research when it comes to evaluating different NL2SQL architectures.
-    \item \textbf{Privacy and Security} — While open-source models address some privacy concerns by enabling a local deployment of LLMs,
-          research gaps remain in ensuring that NL2SQL systems respect database access controls and security policies. This is especially
-          important for enterprise and government environments where data access is strictly regulated.
+          compute resources in order to achieve state-of-the-art results. There is limited research available on performance 
+          optimization of NLIDBs in order to achieve practical and industry-viable hardware constraints.
+    \item \textbf{Real-time Performance Considerations} — Most research papers evaluate models based on their accuracy metrics 
+          alone without factoring in the latency and throughput characteristics of their solutions. This imposes a possibly 
+          significant $\delta$ between academic research and production environments. Responsiveness is an important metric for 
+          user experience and should therefore play a role in NLIDB research when it comes to evaluating different NL2SQL 
+          architectures.
+    \item \textbf{Privacy and Security} — While open-source models address some privacy concerns by enabling a local deployment of 
+          LLMs, research gaps remain in ensuring that NL2SQL systems respect database access controls and security policies. This 
+          is especially important for enterprise and government environments where data access is strictly regulated.
 \end{enumerate}
@@
 
-Despite the long existance of the research field, fundamental questions remain open with regards to handling ambiguity effectively in natural
-language queries. Current, especially language model based, systems struggle to correctly identify when natural language queries contain ambiguities
-that they cant confidently resolve. Although some approaches like multi-path generation and candidate selection are a promising way
-to work around ambiguous language, early systems like \textsc{NaLIR} showed that the most effective way to deal with inherently ambiguous language
-is asking clarifying questions, instead of trying to interpret the query best-effort \citep{NALIR}. Ambiguous language is an inherent source of
-inaccuracy and therefore a cause for misleading query results. Contemporary research and benchmarks dont focus ambiguity detection and
-strategies resolution which therefore leaves open questions to be further researched.
+Despite the long existance of the research field, fundamental questions remain open with regards to handling ambiguity effectively
+in natural language queries. Current, especially language model based, systems struggle to correctly identify when natural language
+queries contain ambiguities that they cant confidently resolve. Although some approaches like multi-path generation and candidate 
+selection are a promising way to work around ambiguous language, early systems like \textsc{NaLIR} showed that the most effective 
+way to deal with inherently ambiguous language is asking clarifying questions, instead of trying to interpret the query best-effort
+\citep{NALIR}. Ambiguous language is an inherent source of inaccuracy and therefore a cause for misleading query results. 
+Contemporary research and benchmarks dont focus ambiguity detection and strategies resolution which therefore leaves open
+questions to be further researched.
 
@@
 \begin{enumerate}
-    \item \textbf{Enterprise-grade Evaluation} — As highlighted by \textsc{Spider 2.0}, there is a gap between academic benchmarks
-          and enterprise realities. Further research is needed to create evaluation frameworks that better represent real-world enterprise
-          environments with thousands of tables and complex relationships. \cite{Spider2} is a promising first step is this direction.
-    \item \textbf{Performance Metrics} — As mentioned above, current benchmarks often dont capture latency or throughput of NL2SQL systems
-          at all, allowing for solutions to achieve state-of-the-art scores that require significantly more resources than their predecessors.
-          Having meaningful performance metrics and benchmarking would allow to further analyze the proposed solutions for their real world
-          feasibility.
-    \item \textbf{User Experience Metrics} — Contemporary benchmarks focus on execution accuracy without assessing potential user satisfaction,
-          trust, and overall experience. Metrics that capture these aspects are necessary for understanding the actual utility of NL2SQL
-          systems in practice. Although execution accuracy is a big factor for the trustworthiness of a NLIDB, it is not the only component
-          as indicated by \citeauthor*{NALIR} in \citeyear{NALIR}.
+    \item \textbf{Enterprise-grade Evaluation} — As highlighted by \textsc{Spider 2.0}, there is a gap between academic 
+          benchmarks and enterprise realities. Further research is needed to create evaluation frameworks that better represent 
+          real-world enterprise environments with thousands of tables and complex relationships. \cite{Spider2} is a promising 
+          first step is this direction.
+    \item \textbf{Performance Metrics} — As mentioned above, current benchmarks often dont capture latency or throughput of NL2SQL 
+          systems at all, allowing for solutions to achieve state-of-the-art scores that require significantly more resources than 
+          their predecessors. Having meaningful performance metrics and benchmarking would allow to further analyze the proposed 
+          solutions for their real world feasibility.
+    \item \textbf{User Experience Metrics} — Contemporary benchmarks focus on execution accuracy without assessing potential
+          user satisfaction, trust, and overall experience. Metrics that capture these aspects are necessary for understanding the 
+          actual utility of NL2SQL systems in practice. Although execution accuracy is a big factor for the trustworthiness of a 
+          NLIDB, it is not the only component as indicated by \citeauthor*{NALIR} in \citeyear{NALIR}.
 \end{enumerate}
@@
 This thesis addresses several of the above outlined research gaps. The primary focus of this work is the integration of open-source
-NL2SQL models with advanced techniques like candidate selection, subset-encoding of database schemas, and synthetic example generation
-for in-context learning. Through the implementation of a PostgreSQL extension this work will bridge multiple critical gaps between
-theoretical advancements and their practical deployability into real world systems whilst exploring performance characteristics.
+NL2SQL models with advanced techniques like candidate selection, subset-encoding of database schemas, and synthetic example 
+generation for in-context learning. Through the implementation of a PostgreSQL extension this work will bridge multiple critical 
+gaps between theoretical advancements and their practical deployability into real world systems whilst exploring performance 
+characteristics.
 
@@
 \section{Theoretical Foundations}
+
+This section serves as an introduction to the theoretical concepts utilized in this thesis. It is meant to establish a
+baseline of understanding for the design, implementation and evaluation phases. This section briefly covers relational
+database theory (incl. SQL), machine learning fundamentals and their subsequent application for large language models
+and the inner workings of PostgreSQL and its extension architecture at a high level.
+
+\subsection{Relational Database Theory}
+
+The relational database theory provides a mathematical framework for formalizing data layout in database management
+systems (DBMS), enabling efficient storage and retrieval mechanisms. This theory is helps to understand the way in
+which natural language queries need to be translated into database operations and what challenges arise during this
+translation.
+
+\subsubsection{Relational Model Fundamentals}
+
+The relational model was introduced by \citeauthor{RelationalModel} in \citeyear{RelationalModel} and essentially forms
+the theoretical foundation for most modern database systems \citep{RelationalModel, Rendezvous}. A relational database
+organizes data in relations (ie. tables), where each relation consists of tuples (ie. rows) containing attributes
+(ie. columns) of specific domains (ie. data types and constraints). The mathematical origins of this model root in
+set theory and first-order predicate logic \citep{RelationalModel, DatabaseSystemIntro}.
+
+Furthermore the relational model abstracts data storage and respective physical implementation details by design,
+allowing interactions with data through respective logical structures (relations, tuples, attributes etc.) rather than
+exposing the underlying storage mechanism. This abstraction is fundamental for NL2SQL systems, as natural language
+queries can be mapped into queries on these structures regardless of underlying physical storage implementation,
+allowing for better transferability of NL2SQL systems (e.g. across different database deployments).
+
+\subsubsection{Core Concepts}
+
+The core concepts of the relational model that are relevant for this thesis include:
+
+\begin{itemize}
+    \item \textbf{Relations} — Mathematical sets of tuples representing entities (e.g. employees, products) or
+          relationships (e.g. enrollment, purchases). Each relation has a fixed structure.
+    \item \textbf{Tuples} — Tuples represent rows of data, by grouping a several attributes into one logical unit
+          (e.g. one specific purchase, containing the amount, date and shop). All tuples in the same relation share
+          the same structure.
+    \item \textbf{Attributes} — Attributes are frequently referred to as columns, representing a specific dimension
+          of a tuple (e.g. customers have names). Attributes are represented in a fixed domain.
+    \item \textbf{Domains} — Domains are sets of allowed values for attributes, including mathematically formalizing
+          data types and constraints.
+    \item \textbf{Schemas} — Structural definitions specifying relations, their attribute names, domains, and integrity
+          constraints.
+    \item \textbf{Primary Keys} — Attributes that uniquely identify tuples within a relation. These
+          ensure entity integrity and ensure referential relationships.
+    \item \textbf{Foreign Keys} — Attributes referencing primary keys in other relations, allowing databases to maintain
+          referential integrity and enabling semantically correct complex queries across multiple tables through joins.
+\end{itemize}
+
+These concepts form the semantic foundation that natural language interfaces must navigate when translating user queries
+into formal database operations. As shown in the literature review \textit{schema-linking} is a crucial problem of
+NLIDBs — establishing an understanding of which relations a natural language query refers to is the first step of
+formalizing a database operation for retrieval.
+
+\subsubsection{SQL as a Declarative Query Language}
+
+Structured Query Language (SQL) is a defacto standard for interfacing with relational databases. It is a declarative
+language to describe database operations like selection, insertion, updating and deletion of database contents. SQL's
+design is strongly influenced by relational algebra and tuple relational calculus. It enables complex data retrieval
+through a readable query syntax incorporating relational algebra operations like:
+
+\begin{itemize}
+    \item \textbf{Selection} ($\sigma$) — Filtering tuples based on specified conditions
+    \item \textbf{Projection} ($\pi$) — Extracting specific attributes from relations
+    \item \textbf{Joins} ($\bowtie$) — Combining data from multiple relations based on common attributes
+    \item \textbf{Aggregation} — Computing summary statistics over groups of tuples
+\end{itemize}
+
+The basic syntax of SQL queries follows a logical pattern that reflects these theoretical operations:
+
+\begin{verbatim}
+SELECT attributes      -- Project attributes
+FROM relations         -- Specify relations to use
+WHERE conditions       -- Selection of tuples
+GROUP BY attributes    -- Grouping for aggregation
+HAVING conditions      -- Group-level filtering
+ORDER BY attributes    -- Ordering of results
+\end{verbatim}
+
+There are further operations like various types of \texttt{JOIN}s, \texttt{OFFSET}, \texttt{LIMIT}, \texttt{WITH}
+that are heavily used in SQL which are excluded for readability and simplicity.
+
+The nature of SQL creates a fundamental challenge for NL2SQL systems; natural language queries express (often fuzzy) user
+intent in terms of desired output (e.g., ``Give me 5 great movies'') while SQL formalizes an explicit and discrete way
+of how to retrieve this information from the database:
+
+\begin{verbatim}
+SELECT m.title, AVG(r.score) as rating
+FROM movies m
+JOIN ratings r ON m.movie_id = r.movie_id
+GROUP BY m.movie_id, m.title
+HAVING AVG(r.score) >= 8
+ORDER BY rating DESC
+LIMIT 5
+\end{verbatim}
+
+This semantic gap between natural language and formal queries represents the core challenge that this thesis addresses.
+
+Furthermore, SQL's compositional nature allows complex queries to be built from simpler components through nesting and
+combination of operations. NL2SQL systems must therefore not only fill in the individual query components but also how
+to compose these to semantically accurate and syntactically correct query statements. This is particularly important when
+dealing with complex natural language queries that may require multi-clause SQL statements involving subqueries, multiple
+joins, and aggregation functions to answer.
+
+\subsubsection{Normalization and Schema Design}
+
+Database schemas typically follow normalization principles to eliminate redundancy and maintain data integrity. The normal forms
+(1NF, 2NF, 3NF, BCNF) provide a set of design guidelines and rules for relational database design to reduce duplication,
+inconsistencies and anomalies \citep{DatabaseSystemIntro}. Understanding normalization forms is crucial for NL2SQL systems to handle:
+
+\begin{itemize}
+    \item \textbf{Schema complexity} — Normalized schemas often distribute logically related information across multiple tables,
+          requiring natural language interfaces to understand implicit relationships and generate appropriate joins.
+    \item \textbf{Semantic mapping} — Humans typically think about data in denormalized, conceptual terms, while relational databases
+          store data in its normalized form. NL2SQL systems must overcome this layout difference (e.g. there is no 1:1 mapping
+          of concepts to tables).
+    \item \textbf{Query complexity} — Retrieving simple information from relational databases may require multiple joins in normalized schemas,
+          challenging NL2SQL systems to generate potentially complex SQL statements from simple user requests.
+\end{itemize}
+
+The tension between normalized database design and intuitive natural language use represents a key challenge that influences
+the architecture and design decisions explored in subsequent sections of this thesis.
+
+\subsection{Machine Learning Fundamentals}
 

### 2025-06-08 thesis/thesis.tex
Write a very short intro to transformers

@@
 
+\subsubsection{Neural Networks}
+
+% already wrote about NNs and their foundations
+
+\subsubsection{Sequence to Sequence Architecture}
+
+\subsubsection{Transformer Architecture}
+
+The transformer architecture advanced the sequence processing capabilities of neural networks and enabled parallelized inference
+which was a critical limitation in efficiency of RNNs and LSTMs. Namely the introduction of attention (sometimes referenced as self-
+attention) allow transformers led to this leap in capability \citep{Attention}. The core architectural characteristics of transformers
+include:
+
+\begin{enumerate}
+	\item \textbf{Input Embedding} — Techniques like \textsc{Word2Vec} allow the semantic representation of words and tokens in 
+		embeddings (ie. vectors in a N-dimensional space).
+	\item \textbf{Positional Encoding} — Allows transformers to keep track of each tokens position in a sequence by adding the
+	 	output of an offset function onto the original embedding.
+	\item \textbf{Self-Attention} — Essentially weighting the relationship between tokens and the semantic importance of tokens.
+	    Attention allows for both parallelized encoding, as well as semantic aware decoding, by calculating encoder-decoder attention 
+	    prior to decoding.              
+\end{enumerate}
+
+Due to these architectural properties, transformers are better positioned to capture long-range dependencies in input sequences,
+and preserving them into the decoding phase – rendering them as the state-of-the-art model architecture for NL2SQL.
+
 \subsection{Problem Decomposition}
@@
 % Conceptual Design
+
 \subsection{System Design}
@@
 % Implementation
+
 \section{Implementation}

### 2025-10-01 thesis/chapters/01-introduction/01-problem-statement.tex
Split the thesis into many small parts

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
 \subsection{Problem Statement and Motivation}
@@
 (ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/chapters/01-introduction/02-objectives.tex
Split the thesis into many small parts

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
 \subsection{Objectives of the Thesis}
@@
 
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
 \newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/chapters/02-literature-review/foundations.tex
Split the thesis into many small parts

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
 \subsection{Foundations of Natural Language Interfaces to Databases}
@@
 as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/chapters/02-literature-review/index.tex
Split the thesis into many small parts

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
+\section{Literature Review}
 
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
+%\input{chapters/02-literature-review/foundations/nl-interfaces}
+%\input{chapters/02-literature-review/traditional-approaches/traditional-approaches}
+%\input{chapters/02-literature-review/neural-approaches/neural-approaches}
+%\input{chapters/02-literature-review/pretrained-models/pretrained-models}
+%\input{chapters/02-literature-review/llm-approaches/llm-approaches}
+%\input{chapters/02-literature-review/benchmarking/benchmarking}
+%\input{chapters/02-literature-review/research-gaps/research-gaps}
 
@@
 
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}
+\subimport{}{foundations}
+\subimport{}{traditional-approaches}
+\subimport{}{neural-approaches}
+\subimport{}{pretrained-models}
+\subimport{}{llm-approaches}
+\subimport{}{benchmarking}
+\subimport{}{research-gaps}

### 2025-10-01 thesis/chapters/02-literature-review/pretrained-models.tex
Split the thesis into many small parts

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
 \subsection{Pre-trained Language Models}
@@
 The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}
+in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on
+domain-specific data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the 
+inference environment, where model modificants are less impactful, due to the extensive pre-training that took place. Approaches 
+like \textsc{DinSQL} show that with the application of LLMs the engineering challenges around model instruction gained relevance 
+while model training became less of a central problem to solve \citep{DINSQL}.

### 2025-10-01 thesis/chapters/02-literature-review/traditional-approaches.tex
Split the thesis into many small parts

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
 \subsection{Traditional NL2SQL Approaches}
@@
 conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/chapters/03-theoretical-foundations/index.tex
Split the thesis into many small parts

\section{Theoretical Foundations}

This section serves as an introduction to the theoretical concepts utilized in this thesis. It is meant to establish a
baseline of understanding for the design, implementation and evaluation phases. It briefly covers relational database theory
(incl. SQL), machine learning fundamentals and their subsequent application for large language models.

\subimport{}{relational-theory.tex}
\subimport{}{machine-learning.tex}
\subimport{}{llms.tex}

%\subsection{Problem Decomposition}

%\subsection{Requirements}

% Conceptual Design

%\subsection{System Design}

%\subsubsection{Architecture Design}

%\subsection{Technical Implementation Strategies}

### 2025-10-01 thesis/front-matter/01-title.tex
Split the thesis into many small parts

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
 \begin{titlepage}
@@
 \end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/chapters/04-design/index.tex
write about the system design

\newpage

\section{System Design}

This chapter describes the design of \textsc{Natural}, the proposed NL2SQL system, 
addressing limitations and research gaps identified in the literature review.
The design follows a modular architecture that leverages both traditional computer
science algorithms and machine learning models to achieve robust, portable and
extensible natural language to SQL translation capabilities.

\subsection{System Architecture Overview}

The core of the proposed design is the query function $nq$ which utilizes
example selection ($\sigma$), query preprocessing, schema subsetting ($\phi$),
model inference ($\pi$), self-correction ($\rho$) and majority voting ($\nu$)
in order to convert a natural language query $q$ into a corresponding SQL query $\omega$.

% NOTE Draw diagram of nq
<pipeline-diagram>
%

This design expands on concepts from DAIL-SQL \citep{DAIL}, X, Y and Z \textcolor{red}{look up}.
The most notable characteristics of \textsc{Natural} lie in it's capabilities for continuous
learning, example selection and portability which allow the system to iteratively
adapt to unseen database schemas.

\subsection{Definitions}

The following sets are subsequently used to describe the systems capabilities:

\begin{enumerate}
    \item $\mathcal{Q}$ – The set of all possible natural language queries
    \item $\mathcal{S}$ – The set of all possible database schemas
    \item $\mathcal{E}$ – The set of execution functions for SQL validation
    \item $\mathcal{V}$ – The embedding space of historically observed samples
    \item $\mathcal{D}$ – The space of distance indices over historically observed databases
    \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$
    \item $\mathcal{Q}_{\mathcal{S}}$ – The set of all valid queries over the database schema $S$.
\end{enumerate}

\subsection{Query Function – $nq$}

\vspace{0.5em}

$$
nq: \mathcal{Q} \times \mathcal{S} \times E \times \mathcal{V} \times \mathcal{D} \rightarrow \Omega_{\mathcal{S}}
$$

\vspace{0.5em}

For a specific input instance, we write:

$$
nq(q, s, e, v, d) = \omega
$$

\vspace{0.5em}

where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$, and $\omega \in \Omega_{\mathcal{S}}$.

\subsection{Selection Function – $\sigma$}

The selection function $\sigma$ retrieves the most relevant examples from the
historical embedding space $\mathcal{V}$ based on semantic similarity to the
input query. This function implements the core example selection mechanism 
that subsequently enables few-shot learning for SQL generation.

$$
\sigma: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \mathcal{V} \rightarrow \mathcal{C}'
$$

\vspace{0.5em}

For a specific input instance, we write:

$$
\sigma(q, s, e, v) = \{(q_1, s_1, \omega_1, d_k), ..., (q_k, s_k, \omega_k, d_k)\}
$$

\vspace{0.5em}

where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$,
and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$ tuples of
natural language queries, schemas, corresponding SQL queries and their combined \textit{distance}.

The function utilizes cosine similarity in the embedding space to identify examples
that are semantically closest to the input query $q$, considering both masked natural
language representations and schema distance (\textcolor{red}{Described in section XYZ}).
This approach ensures that examples are weighted by the following properties:

\begin{enumerate}
    \item \textbf{Question similarity} – 
    \item \textbf{Query similarity} –
    \item \textbf{Database similarity} –
\end{enumerate}

\subsection{Subsetting Function – $\phi$}

The subsetting function $\phi$ intelligently reduces the database schema to only include
tables, columns, and relationships that are relevant to the current query context.
This schema pruning mechanism reduces the complexity of the SQL generation task and
improves model performance by focusing on pertinent schema elements.

$$
\phi: \mathcal{C}' \times \mathcal{S} \rightarrow \mathcal{S}'
$$

\vspace{0.5em}

For a specific input instance, we write:

$$
\phi(c, s) = s'
$$

\vspace{0.5em}

where $c \in \mathcal{C}'$ is a candidate set containing selected examples, $s \in \mathcal{S}$
is the full database schema, and $s' \subseteq s$ is the reduced schema subset containing
only relevant tables and columns.

The function analyzes the selected examples in $c$ to identify which schema elements
(tables, columns, foreign key relationships) are commonly referenced in similar queries.
This analysis enables the system to steer the model's attention on the most relevant
portions of potentially large and complex database schemas, thereby improving both accuracy
and computational efficiency.

\subsection{Projection Function – $\pi$}

The projection function $\pi$ represents the core translation mechanism that converts
natural language queries into SQL statements using the selected examples and reduced schema.
This function encapsulates the language model inference process that generates candidate
SQL queries based on the provided context.

$$
\pi: \mathcal{Q} \times \mathcal{S}' \times \mathcal{C}' \rightarrow \Omega_{\mathcal{S}'}
$$

\vspace{0.5em}

For a specific input instance, we write:

$$
\pi(q, s', c) = \omega
$$

\vspace{0.5em}

where $q \in \mathcal{Q}$ is the input natural language query, $s' \in \mathcal{S}'$ is
the reduced schema subset, $c \in \mathcal{C}'$ is the set of selected examples,
and $\omega \in \Omega_{\mathcal{S}'}$ is the generated SQL query.

The function operates by constructing a prompt that combines the natural language
query, the relevant schema information, and the selected examples in a format
optimized for large language model inference. The model then generates a SQL query
that attempts to capture the semantic intent of the natural language input while
adhering to the constraints imposed by the database schema.

\subsection{Refinement Function – $\rho$}

The refinement function $\rho$ implements a self-correction mechanism that iteratively
improves generated SQL queries by identifying and correcting syntax errors,
semantic inconsistencies, and execution failures. This function enables the system
to learn from its mistakes and produce higher-quality outputs through automated
feedback loops.

$$
\rho: \mathcal{Q} \times \mathcal{S}' \times \mathcal{E} \times \Omega_{\mathcal{S}'} \rightarrow \Omega_{\mathcal{S}'}
$$

\vspace{0.5em}

For a specific input instance, we write:

$$
\rho(q, s', e, \omega_{prev}) = \omega_{refined}
$$

\vspace{0.5em}

where $q \in \mathcal{Q}$ is the original natural language query, $s' \in \mathcal{S}'$
is the relevant schema subset, $e \in \mathcal{E}$ is the execution environment for
validation, $\omega_{prev} \in \Omega_{\mathcal{S}'}$ is the previously generated SQL
query, and $\omega_{refined} \in \Omega_{\mathcal{S}'}$ is the improved SQL query.

The function operates by executing the candidate query against the database, analyzing
any errors or unexpected results, and then prompting the language model to generate
an improved version based on the identified issues. This iterative refinement
process continues until a valid, executable query is produced or a maximum number
of refinement attempts is reached.

\subsection{Voting Function – $\nu$}

The voting function $\nu$ implements a consensus mechanism that selects the most
reliable SQL query from multiple candidate solutions generated through the pipeline.
This function enhances robustness by leveraging the wisdom of multiple generation
attempts to identify the most likely correct answer.

$$
\nu: \mathcal{C} \times \mathcal{E} \rightarrow \Omega_{\mathcal{S}'}
$$

\vspace{0.5em}

For a specific input instance, we write:

$$
\nu(C, e) = \omega_{consensus}
$$

where $C \in \mathcal{C}$ is a set of candidate SQL queries, $e \in \mathcal{E}$ is
the execution environment for validation, and $\omega_{consensus} \in \Omega_{\mathcal{S}'}$
is the selected consensus query.

\vspace{0.5em}

The voting function $\nu$ implements a majority voting algorithm similar to that
described by OmniSQL \citep{OmniSQL}, where the result that appears most frequently
across multiple generation attempts is deemed to be the most likely correct answer.
The function first validates all candidates for syntactic correctness and executability,
then applies frequency-based selection among the valid candidates. In cases where
no clear majority exists, the function may apply additional heuristics such as query
complexity or execution performance to determine the final selection.

\subsection{Masked Embedding Space – $\mathcal{V}$}

Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes an
embedding space to both traverse the masked natural language query $q'$
and the respective masked SQL query $\omega'$. This allows $\sigma$ to use
cosine distance to search through historically observed question answer pairs;
eg, from previous user interactions or prevalent training datasets like
\textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.

\subsubsection{Natural Language Masking}

The natural language masking process replaces all words which are not deemed
\textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
Choosing a sound set $\matchal{W}$ is described in chapter 5.

This process enables the embedding space to capture semantic patterns independent
of specific database artifacts and domain terminology, improving the generalizability
of example selection across different schemas.

This abstraction allows the system to recognize that queries like:

\begin{verbatim}
Show all customers with the firstname John
List all products with the name iPhone
\end{verbatim}

in fact share similar semantic structures, even though they operate on different
entities. Given $\matchal{W} = \{\text{List}, \text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$
the masking would result in:

\begin{verbatim}
Show all <mask> with the <mask> <mask>
List all <mask> with the name <mask>
\end{verbatim}

Embeddings can now be computed for the masked versions of the natural language queries.

\subsubsection{Query Masking}

Query masking applies similar abstraction principles to the SQL queries themselves.
This enables the system to identify SQL patterns and structures rather than
memorizing specific database schemas.

The SQL masking process transforms all identifiers to \texttt{<mask>} and all literals
to \texttt{<value>}. For example, the query:

\begin{verbatim}
SELECT name FROM customers WHERE age > 25
\end{verbatim}

gets masked to:

\begin{verbatim}
SELECT <mask> FROM <mask> WHERE <mask> > <value>
\end{verbatim}

This abstraction enables pattern recognition across different domains while
preserving the essential logical structure of the original SQL statements.

\subsection{Algorithm Design}

\begin{algorithm}

\caption{The $nq$ function}\label{alg:cap}
\begin{algorithmic}[1]
\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
\Require $k \in \{ n \in N\ |\ 2n + 1 \}, k \geq 1$
\State $S \gets nsel(q,s,e,v)$             \Comment{Example selection is performed}
\State $C \gets \emptyset$                 \Comment{The set of query candidates}
\While{$|C| < k$}                          \Comment{While not enough candidates exist}
    \State $sub \gets nsub(C, s)$          \Comment{Identify relevant schema subset}
    \State $nc \gets nprompt(S, q, sub)$   \Comment{Prompt model for next candidate}

    \If{$e(nc)$}                           \Comment{Evaluate the next candidate}
        \State $c \gets \{nc\} \cup c$     \Comment{Extend the candidates}
    \EndIf

    \State $ref \gets nref(q, sub, e, v)$  \Comment{Refine via self correction}

    \If{$e(ref)$}                          \Comment{Evaluate the refined candidate}
        \State $c \gets \{ref\} \cup c$    \Comment{Extend the candidates}
    \EndIf
\EndWhile
\State \Return $nvote(C, e)$               \Comment{Return most prevalent result}
\end{algorithmic}

\end{algorithm}

### 2025-10-01 thesis/discussion/index.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
 % Discussion
@@
 \newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/implementation/index.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
 \section{Implementation}
@@
 \newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/introduction/03-research-questions.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
 \subsection{Research Questions}
@@
 \newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/introduction/04-thesis-structure.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
 \subsection{Structure of the Thesis}
@@
 \end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/literature-review/benchmarking.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
 \subsection{Benchmarking}
@@
 
-\subsubsection{Deployment Gaps}
+This literature review highlights that significant progress was made in the development of real world feasible NL2SQL systems
+and that the research domain has undergone multiple large paradigm shifts from rule-based to neural-network-based to PLM-based
+and most recently to LLM-based approaches. Despite these advancement several critical research gaps remained open or emerged
+which limit the practical deployment of NL2SQL systems and their widespread adoption as natural language database interfaces
+in real world scenarios.
 
-\newpage
+\subsubsection{Advanced Open-Source Approaches}
 
-\section{Theoretical Foundations}
+While recent research papers have made remarkable progress on both open-source model development (\textsc{CodeS} and
+\textsc{OmniSQL}) aswell as advanced architecture development like CHASE and \textsc{XiYan-SQL} (which utilize multi-path
+generation and candidate selection, self-correction and RAG), a significant research gap evolved around state-of-the-art
+approaches that dont rely on proprietary LLMs and only utilize open-source models. Most contemporary research proposes
+solutions which (partially) rely on the baseline proprietary LLM in order to achieve state-of-the-art results. Whilst
+these efforts give meaningful signal to the effect that specific prompt engineering techniques, self-correction mechanisms
+and candidate selection models have, they are not yet feasible for real world scenarios due to the cost and data privacy
+concerns they introduce.
 
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}
+\begin{enumerate}
+    \item \textbf{Limited Exploration of Technique Synergies} — Contemporary research primarily focuses on techniques like
+          candidate selection, RAG, and self-correction in isolation or with specific closed-source LLMs. A gap remains
+          in understanding how these techniques can be optimally combined, especially with emerging open-source models.
+          For instance, the potential of synergy of specialized open-source LLM like \textsc{OmniSQL} with candidate-selection,
+          advanced self-correction and subset-encoding of database schemas has yet to be researched.
+    \item \textbf{Efficiency-Accuracy Tradeoffs} — Whilst recent research efforts focus primarily on achieving a new
+          state-of-the-art execution accuracy metric on prevalent benchmarks, the relationship of computational requirements and
+          performance gains achieved remains unclear. Some research papers present ablation studies indicating the relevance
+          of certain architecture components but the overall relationship between state-of-the-art performance and cost
+          is still underexplored. 
+     \item \textbf{Cross-Technique Optimization} — There is limited research on how to optimize the interplay between the various
+          NL2SQL techniques. For example, it is yet unclear how subset-encoding of database schemas might impact the effectiveness
+          of multi-path generation and candidate selection, or whether synthetic example generation for in-context-learning is 
+          effective when working with specialized models versus general-purpose LLMs.
+\end{enumerate}

### 2025-10-01 thesis/literature-review/llm-approaches.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
 \subsection{Large Language Models}
 
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
+The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early 
+experiments with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches 
+\citep{DAIL-SQL}. \cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any 
+fine-tuning efforts, could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that 
+required extensive training. This breakthrough challenged the contemporary assumption that further specialization of model 
+architectures would yield increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
 
@@
 \end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}

### 2025-10-01 thesis/literature-review/neural-approaches.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
-
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
 \subsection{Neural NL2SQL Approaches}
@@
 
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
+Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late
+neural approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
 
 \begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
+    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data 
+          (ie. natural language paired with output SQL queries) than traditional systems which required serious efforts of data 
+          collection \citep{Spider}.
+    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models 
+          produced invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by 
+          utilizing syntax trees during decoding but nonetheless syntactic correctness remained a challenge across future
+          iterations of neural systems. \citep{SyntaxSQLNet}
+    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered
+          from a limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly 
+          domain specific natural language queries challenging.
+    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones 
+          with complex architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
 \end{enumerate}
 
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}
+The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility
+of NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning 
+approaches (specifically language model oriented approaches )being researched. Neural approaches showed to significant 
+improvements in performance when being paired with pre-trained language models \citep{RATSQL} which led to further research on 
+their applicability.

### 2025-10-01 thesis/literature-review/research-gaps.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
+\subsubsection{Deployment and Performance Gaps}
 
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
-
-\graphicspath{{./images/}}
-
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
-
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
-
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
-
-\begin{document}
-
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
-
-\newpage
-
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
-
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
+Despite impressive academic results, significant gaps remain in transitioning NL2SQL systems from research environments
+to real-world, enterprise-feasible, deployments:
 
 \begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
+    \item \textbf{Database Integration} — While existing research has often focused on standalone systems, little attention
+          has been given to the integration complexities between NL2SQL capabilities and database management systems like 
+          PostgreSQL. Having two standalone systems imposes a significant data transfer need between the two systems when infact, 
+          natural language queries can be treated as an extension to most existing databases. This implementation gap prevents 
+          seamless adoption into existing databases as it requires additional software layers which in turn increase the overall 
+          complexity and cost.
+    \item \textbf{Hardware Requirement Optimization} — As mentioned above, current LLM-based approaches often require significant
+          compute resources in order to achieve state-of-the-art results. There is limited research available on performance 
+          optimization of NLIDBs in order to achieve practical and industry-viable hardware constraints.
+    \item \textbf{Real-time Performance Considerations} — Most research papers evaluate models based on their accuracy metrics 
+          alone without factoring in the latency and throughput characteristics of their solutions. This imposes a possibly 
+          significant $\delta$ between academic research and production environments. Responsiveness is an important metric for 
+          user experience and should therefore play a role in NLIDB research when it comes to evaluating different NL2SQL 
+          architectures.
+    \item \textbf{Privacy and Security} — While open-source models address some privacy concerns by enabling a local deployment of 
+          LLMs, research gaps remain in ensuring that NL2SQL systems respect database access controls and security policies. This 
+          is especially important for enterprise and government environments where data access is strictly regulated.
 \end{enumerate}
 
-\subsection{Research Questions}
+\subsubsection{Ambiguity Resolution and Semantic Accuracy}
 
-% \subsubsection{Primary Research Questions}
+Despite the long existance of the research field, fundamental questions remain open with regards to handling ambiguity effectively
+in natural language queries. Current, especially language model based, systems struggle to correctly identify when natural language
+queries contain ambiguities that they cant confidently resolve. Although some approaches like multi-path generation and candidate 
+selection are a promising way to work around ambiguous language, early systems like \textsc{NaLIR} showed that the most effective 
+way to deal with inherently ambiguous language is asking clarifying questions, instead of trying to interpret the query best-effort
+\citep{NALIR}. Ambiguous language is an inherent source of inaccuracy and therefore a cause for misleading query results. 
+Contemporary research and benchmarks dont focus ambiguity detection and strategies resolution which therefore leaves open
+questions to be further researched.
 
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
+\subsubsection{Evaluation and Benchmarking Gaps}
 
 \begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
+    \item \textbf{Enterprise-grade Evaluation} — As highlighted by \textsc{Spider 2.0}, there is a gap between academic 
+          benchmarks and enterprise realities. Further research is needed to create evaluation frameworks that better represent 
+          real-world enterprise environments with thousands of tables and complex relationships. \cite{Spider2} is a promising 
+          first step is this direction.
+    \item \textbf{Performance Metrics} — As mentioned above, current benchmarks often dont capture latency or throughput of NL2SQL 
+          systems at all, allowing for solutions to achieve state-of-the-art scores that require significantly more resources than 
+          their predecessors. Having meaningful performance metrics and benchmarking would allow to further analyze the proposed 
+          solutions for their real world feasibility.
+    \item \textbf{User Experience Metrics} — Contemporary benchmarks focus on execution accuracy without assessing potential
+          user satisfaction, trust, and overall experience. Metrics that capture these aspects are necessary for understanding the 
+          actual utility of NL2SQL systems in practice. Although execution accuracy is a big factor for the trustworthiness of a 
+          NLIDB, it is not the only component as indicated by \citeauthor*{NALIR} in \citeyear{NALIR}.
 \end{enumerate}
 
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
+\subsubsection{Thesis Placement}
 
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}
+This thesis addresses several of the above outlined research gaps. The primary focus of this work is the integration of open-source
+NL2SQL models with advanced techniques like candidate selection, subset-encoding of database schemas, and synthetic example 
+generation for in-context learning. Through the implementation of a PostgreSQL extension this work will bridge multiple critical 
+gaps between theoretical advancements and their practical deployability into real world systems whilst exploring performance 
+characteristics.

### 2025-10-01 thesis/theoretical-foundations/machine-learning.tex
Better structure

\subsection{Machine Learning Fundamentals}

Machine learning represents the algorithmic foundation for state-of-the-art NL2SQL systems, enabling systems to learn how natural
language queries translate into SQL. Understanding these foundations is essential for implementing and optimizing large language
model-based NL2SQL approaches.

\subsubsection{Neural Network Architecture}

Neural networks are the computational concept behind contemporary NL2SQL approaches, representing complex capabilities through
compositions of simpler operations. A neural network consists of interconnected computational units (called neurons) organized
in layers, where each connection has an associated (and learnable) weight and bias parameters.

The fundamental computation paradigm of neural network outputs is called forward propagation, which is applying the respective weights ($W$)
and biases ($b$) to the input parameter ($a$) and transforming it using a (non-linear) activation function $f$:

\begin{equation}
a^{l+1} = f(Wa^l + b)
\end{equation}

where $W \in \mathbb{R}^{m \times n}$ is the weight matrix, $a \in \mathbb{R}^{n}$ is the input vector, $b \in \mathbb{R}^{m}$ is
the bias vector, and $f$ is an activiation function (e.g. ReLU, sigmoid, or tanh). Using this function repeatedly, propagates information
through the neural network.

The most important architectural components for language processing models include:

\begin{itemize}
    \item \textbf{Feedforward Networks} — Which can process fixed-size inputs through successive linear transformations and activations,
          this architecture is particularly suitable for classification and regression problems within NL2SQL systems.
    \item \textbf{Recurrent Networks (RNNs/LSTMs)} — Which can handle sequential data of variable-length by maintaining hidden states
          that capture dependencies, enabling processing of natural language sequences of arbitrary length.
    \item \textbf{Embedding Layers} — Which map discrete tokens (e.g. words, characters, or subwords) to a dense vector representation,
          thus providing the foundation for subsequent neural language processing.
\end{itemize}

The ability of neural networks to learn hierarchical representations through multiple layers makes them particularly well-suited for
the complex translation required in NL2SQL systems.

\subsubsection{Sequence-to-Sequence Models}

Sequence-to-sequence (Seq2Seq) models established the foundational architecture for neural NL2SQL approaches \citep{Seq2SQL, SQLNet}.
These models which rely on RNNs and LSTMs addressed the fundamental challenge of mapping variable-length input sequences (natural
language queries) to variable-length output sequences (SQL statements).

The Seq2Seq model architecture is composed out of three essential components:

\begin{enumerate}
    \item \textbf{Encoder} — Process the input natural language sequence $x_1, x_2, \ldots, x_n$ into a sequence of hidden
          representations $h_1, h_2, \ldots, h_n$, capturing the semantic content of the query.
    \item \textbf{Decoder} — Generates the output SQL sequence $y_1, y_2, \ldots, y_m$ autoregressively, where each token is
          predicted based on previous outputs and encoder representations.
    \item \textbf{Attention} — Allows the decoder to selectively focus on relevant portions of the input sequence, addressing
          the information bottleneck of fixed-size context vectors.
\end{enumerate}

### 2025-10-01 thesis/theoretical-foundations/relational-theory.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
-\documentclass{article}
+\subsection{Relational Database Theory}
 
-\usepackage[a4paper]{geometry}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
-\usepackage[natbibapa]{apacite}
+The relational database theory provides a mathematical framework for formalizing data layout in database management
+systems (DBMS), enabling efficient storage and retrieval mechanisms. This theory is helps to understand the way in
+which natural language queries need to be translated into database operations and what challenges arise during this
+translation.
 
-\graphicspath{{./images/}}
+\subsubsection{Relational Model Fundamentals}
 
-\pagestyle{fancyplain}
-\fancyhf{}
-\lhead{\fancyplain{}{Mara Schulke} }
-\rhead{\fancyplain{}{\today}}
-\cfoot{\fancyplain{}{\thepage}}
+The relational model was introduced by \citeauthor{RelationalModel} in \citeyear{RelationalModel} and essentially forms
+the theoretical foundation for most modern database systems \citep{RelationalModel, Rendezvous}. A relational database
+organizes data in relations (ie. tables), where each relation consists of tuples (ie. rows) containing attributes
+(ie. columns) of specific domains (ie. data types and constraints). The mathematical origins of this model root in
+set theory and first-order predicate logic \citep{RelationalModel, DatabaseSystemIntro}.
 
-\newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
-}
+Furthermore the relational model abstracts data storage and respective physical implementation details by design,
+allowing interactions with data through respective logical structures (relations, tuples, attributes etc.) rather than
+exposing the underlying storage mechanism. This abstraction is fundamental for NL2SQL systems, as natural language
+queries can be mapped into queries on these structures regardless of underlying physical storage implementation,
+allowing for better transferability of NL2SQL systems (e.g. across different database deployments).
 
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
-\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\small\bfseries}{\thesubsubsection}{1em}{}
+\subsubsection{Core Concepts}
 
-\begin{document}
+The core concepts of the relational model that are relevant for this thesis include:
 
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
+\begin{itemize}
+    \item \textbf{Relations} — Mathematical sets of tuples representing entities (e.g. employees, products) or
+          relationships (e.g. enrollment, purchases). Each relation has a fixed structure.
+    \item \textbf{Tuples} — Tuples represent rows of data, by grouping a several attributes into one logical unit
+          (e.g. one specific purchase, containing the amount, date and shop). All tuples in the same relation share
+          the same structure.
+    \item \textbf{Attributes} — Attributes are frequently referred to as columns, representing a specific dimension
+          of a tuple (e.g. customers have names). Attributes are represented in a fixed domain.
+    \item \textbf{Domains} — Domains are sets of allowed values for attributes, including mathematically formalizing
+          data types and constraints.
+    \item \textbf{Schemas} — Structural definitions specifying relations, their attribute names, domains, and integrity
+          constraints.
+    \item \textbf{Primary Keys} — Attributes that uniquely identify tuples within a relation. These
+          ensure entity integrity and ensure referential relationships.
+    \item \textbf{Foreign Keys} — Attributes referencing primary keys in other relations, allowing databases to maintain
+          referential integrity and enabling semantically correct complex queries across multiple tables through joins.
+\end{itemize}
 
-    \vfill
+These concepts form the semantic foundation that natural language interfaces must navigate when translating user queries
+into formal database operations. As shown in the literature review \textit{schema-linking} is a crucial problem of
+NLIDBs — establishing an understanding of which relations a natural language query refers to is the first step of
+formalizing a database operation for retrieval.
 
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
+\subsubsection{SQL as a Declarative Query Language}
 
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
+Structured Query Language (SQL) is a defacto standard for interfacing with relational databases. It is a declarative
+language to describe database operations like selection, insertion, updating and deletion of database conte/home/mara/Workspace/thb/thesis/main.texnts. SQL's
+design is strongly influenced by relational algebra and tuple relational calculus. It enables complex data retrieval
+through a readable query syntax incorporating relational algebra operations like:
 
-    \vfill
+\begin{itemize}
+    \item \textbf{Selection} ($\sigma$) — Filtering tuples based on specified conditions
+    \item \textbf{Projection} ($\pi$) — Extracting specific attributes from relations
+    \item \textbf{Joins} ($\bowtie$) — Combining data from multiple relations based on common attributes
+    \item \textbf{Aggregation} — Computing summary statistics over groups of tuples
+\end{itemize}
 
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
+The basic syntax of SQL queries follows a logical pattern that reflects these theoretical operations:
 
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
+\begin{verbatim}
+SELECT attributes      -- Project attributes
+FROM relations         -- Specify relations to use
+WHERE conditions       -- Selection of tuples
+GROUP BY attributes    -- Grouping for aggregation
+HAVING conditions      -- Group-level filtering
+ORDER BY attributes    -- Ordering of results
+\end{verbatim}
 
-\tableofcontents
+There are further operations like various types of \texttt{JOIN}s, \texttt{OFFSET}, \texttt{LIMIT}, \texttt{WITH}
+that are heavily used in SQL which are excluded for readability and simplicity.
 
-\listoffigures
+The nature of SQL creates a fundamental challenge for NL2SQL systems; natural language queries express (often fuzzy) user
+intent in terms of desired output (e.g., ``Give me 5 great movies'') while SQL formalizes an explicit and discrete way
+of how to retrieve this information from the database:
 
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
+\begin{verbatim}
+SELECT m.title, AVG(r.score) as rating
+FROM movies m
+JOIN ratings r ON m.movie_id = r.movie_id
+GROUP BY m.movie_id, m.title
+HAVING AVG(r.score) >= 8
+ORDER BY rating DESC
+LIMIT 5
+\end{verbatim}
 
-\newpage
+This semantic gap between natural language and formal queries represents the core challenge that this thesis addresses.
 
-% 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
+Furthermore, SQL's compositional nature allows complex queries to be built from simpler components through nesting and
+combination of operations. NL2SQL systems must therefore not only fill in the individual query components but also how
+to compose these to semantically accurate and syntactically correct query statements. This is particularly important when
+dealing with complex natural language queries that may require multi-clause SQL statements involving subqueries, multiple
+joins, and aggregation functions to answer.
 
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
+\subsubsection{Normalization and Schema Design}
 
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
+Database schemas typically follow normalization principles to eliminate redundancy and maintain data integrity. The normal forms
+(1NF, 2NF, 3NF, BCNF) provide a set of design guidelines and rules for relational database design to reduce duplication,
+inconsistencies and anomalies \citep{DatabaseSystemIntro}. Understanding normalization forms is crucial for NL2SQL systems to handle:
 
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
+\begin{itemize}
+    \item \textbf{Schema complexity} — Normalized schemas often distribute logically related information across multiple tables,
+          requiring natural language interfaces to understand implicit relationships and generate appropriate joins.
+    \item \textbf{Semantic mapping} — Humans typically think about data in denormalized, conceptual terms, while relational databases
+          store data in its normalized form. NL2SQL systems must overcome this layout difference (e.g. there is no 1:1 mapping
+          of concepts to tables).
+    \item \textbf{Query complexity} — Retrieving simple information from relational databases may require multiple joins in normalized schemas,
+          challenging NL2SQL systems to generate potentially complex SQL statements from simple user requests.
+\end{itemize}
 
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
-
-\newpage
-
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
-
-\newpage
-
-\section{Literature Review}
-
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
-
-\newpage
-
-\section{Theoretical Foundations}
-
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
-
-\newpage
-
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
-
-\newpage
-
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
-
-\newpage
-
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
-
-\newpage
-
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
-
-\end{document}
+The tension between normalized database design and intuitive natural language use represents a key challenge that influences
+the architecture and design decisions explored in subsequent sections of this thesis.

### 2025-10-01 thesis/thesis.tex
Better structure

--- 2025-05-23 thesis/thesis.tex
@@
 \usepackage[T1]{fontenc}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{fancyhdr}
+\usepackage{float}
 \usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage{amsmath}
+\usepackage{hyperref}
+\usepackage{import}
+\usepackage{titlesec}
 \usepackage{xcolor}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{titlesec}
 \usepackage[natbibapa]{apacite}
@@
 \newcommand{\figuresource}[1]{
-	\begin{center}Quelle: {#1}\end{center}
+	\begin{center}Source: {#1}\end{center}
 }
 
-\titleformat{\chapter}{\normalfont\Large\bfseries}{\thechapter.}{20pt}{\Large}
 \titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
@@
 
-\begin{titlepage}
-    \begin{center}
-        \begin{Large}
-            Brandenburg University of Applied Sciences \\[1em]
-        \end{Large}
-        IT Security \\
-        Computerscience \\
-        Prof. Dr. Oleg Lobachev \\
-        MSc. Florian Eich
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        \Large{Reliable Natural Language Interfaces using LLMs, Self-Correction and Incremental Schema Analysis}\\[0.5em]
-        \large{Bachelor Thesis}\\[1em]
-
-        \begin{normalsize}
-            Summer semester 2025\\[0.25em]
-            \today
-        \end{normalsize}
-    \end{center}
-
-    \vfill
-
-    \begin{center}
-        Mara Schulke – Matr-Nr. 20215853
-    \end{center}
-\end{titlepage}
-
-\begin{abstract}
-This thesis explores the integration of large language models (LLMs) into PostgreSQL database systems in order
-to make the database accessible via natural language instead of the postgres SQL dialect. The research focuses
-on implementation strategies, performance optimization, and practical applications of this concept.
-\end{abstract}
-
-\tableofcontents
-
-\listoffigures
-
-\section*{List of Abbreviations}
-\begin{tabular}{ll}
-GPT & Generative Pretrained Transformer \\
-SQL & Structured Query Language \\
-API & Application Programming Interface \\
-LLM & Large Language Model \\
-DBMS & Database Management System \\
-NL2SQL & Natural Language to SQL \\
-\end{tabular}
+\import{title}{index}
+\import{abstract}{index}
+\import{toc}{index}
 
@@
 % 1/3 Vorarbeit, Literature Review
-% 1/3 Theory, Decomp und System Design
-% 1/3 Implementation und Evaluation
 
-% Introduction
-\section{Introduction}
-\subsection{Problem Statement and Motivation}
-
-Database systems represent a backbone of modern computer science, allowing for rapid advancements
-whilst shielding us from the problem categories that come along with managing and querying large amounts
-of, usually structured, data efficiently. However, most Database Management Systems (DBMS) have
-traditionally required specialized knowledge, usually of the Structured Query Language (SQL), in order
-to become useable. Whilst this barrier may be percieved differently across diverse usergroups it
-represents a fundamental misalignment between end-user goals (e.g. analysts, researchers, domain experts
-etc.) and the underlying DBMS, thus often requiring software engineering efforts in order to reduce this friction.
-
-This barrier is the reason entire classes of software projects exists (for example, admin / support panels),
-data analytics tools etc. which therefore introduce significant churn and delay between the implementation
-of a database system and reaching the desired end user impact. Often these projects span multiple years, require
-costly staffing and yield little to no novel technical value.
-
-Emerging technologies such as Large Language Models (LLMs) have proven themselves as a sensible tool for bridging
-fuzzy user provided input into discrete, machine readable formats. Prominent models in this field have demostrated
-outstanding capabilities that enable computer scientists to tackle new problem classes, that used to be
-challenging / yielded unsatisfying results with logical programming approaches.
-
-This thesis is exploring ways to overcome the above outlined barrier using natural language queries, so that domain experts,
-business owners, support staff etc. are able to seamlessly interact with their data, essentially eliminating the
-requirement of learning SQL (and its pitfalls). By translating natural language to SQL using Large Language Models
-this translation becomes very robust (e.g. against different kinds of phrasing) and enables novel applications
-in how businesses, researchers and professionals interact with their data — it represents a fundamental shift 
-(ie. moving away from SQL) towards a more inclusive and data driven world.
-
-\subsection{Objectives of the Thesis}
-
-This thesis aims to address the aforementioned challanges when it comes to database accessibility.
-The following objectives are the core research area of this thesis:
-
-\begin{enumerate}
-    \item Develop a database extension that can translate natural language queries into semantically
-          accurate SQL queries using Large Language Models.
-    \item To evaluate the effectiveness and feasibility of different Models aswell as prompt engineering
-          techniques in order to improve the performance of the system.
-    \item Identify and address issues when it comes to handling amibguous, complex and domain specific user input.
-    \item Benchmark the performance of the implementation against common natural language to SQL (NL2SQL) benchmarks.
-    \item Idenitfy potential use cases for real world scenarios that could deliver a noticable upsides to users.
-    \item Analyze the short commings and limitations of this approach and propose potential solutions to overcome them.
-\end{enumerate}
-
-\subsection{Research Questions}
-
-% \subsubsection{Primary Research Questions}
-
-\subsubsection*{RQ1 — Are natural language database interfaces feasible for real world application?}
-
-The primary research questions when it comes to natural language database interfaces evolve around their
-semantic accuracy and reliability, therefore questioning their feasibility for real world usage.
-LLMs have notoriously been known for their ability to hallucinate / produce false, but promising outputs.
-This behaviour can be especially  dangerous when opting for data driven decisions that rely on false data
-due to a mistranslation from natural language to SQL. LLMs could cause hard to understand and debug behaviour,
-like false computation of distributions when the intermediate format is not being shown to the user. This
-thesis tries to determine whether such hallucinations could be reasonably prevented and whether the associated
-performance and hardware requirements are suitable for a real world deployment, outside of research situations.
-
-Specifically the two big underlying questions are:
-
-\begin{enumerate}
-    \item Is the semantic accuracy of natural language database interfaces high enough to yield a noticable
-          benefit to users?
-    \item Is it possible to run such an interface on reasonable, mass available hardware (e.g. excluding high end research GPUs).
-\end{enumerate}
-
-\subsubsection*{RQ2 — What approaches are most effective in resolving ambiguity when translating natural language queries into SQL?}
-
-To provide semantically correct results ambiguity in the user-provided natural language queries must be 
-adequately addressed. This thesis investigates various approaches to ambiguity management and
-resolution. Natural language queries can demonstrate ambiguity even at low levels of complexity —
-e.g. there are two different types of "sales" in a database schema, and the user asks to retireve
-"all sales".
-
-Such situations present the second major challenge associated with the practical implementation of natural
-language database interfaces. The success of this concept will significantly depend on whether suitable 
-designs and mitigation techniques can be implemented without creating problems with regards to the 
-aforementioned performance and hardware requirements. The research focus lies on both preventative measures
-through optimized pre-processing stages and prompt engineering techniques as well as reactive strategies
-that post process LLM output, either on the basis of further user input or context inference.
-
-\subsubsection*{RQ3 — Which strategies are increasing semantic accuracy of queries?}
-
-In order to enhance the semantic accuracy a series of improvements may be applied to the pipeline.
-Potential optimizations include supplying (parts of) the schema during LLM prompting, implementation of
-interactive contextual reasoning through a conversational interface which would allow for user
-refinement, the implementation of a robust SQL parsing and validation mechanism and a hybrid approach
-partly relying on traditional NLP preprocessing techniques. This research will quantify semantic accuracy
-using popular NL2SQL benchmarks and empirically evaluate the impact each approach has on the benchmark
-performance. Furthermore this research will take a look at the optimal combination of the aforementioned
-solutions in order to develop a system that strikes the right balance between accuracy and performance.
-
-% \subsubsection{Secondary Research Questions}
-
-% \subsubsection*{RQ4 — What use cases are most suitable for natural language database interfaces?}
-
-% \subsubsection*{RQ5 — Where are the limitations of natural language database interfaces?}
-
-% \subsubsection*{RQ6 — How is NL2SQL model performance impacted through model selection, fine tuning and pre/post processing?}
+\import{introduction}{index}
 
@@
 
-\subsection{Structure of the Thesis}
-
-This thesis is following a research and development methodology in order to implement a natural language
-interface for databases, in particular postgres is used.
-
-\begin{enumerate}
-    \item \textbf{Literature Review} — An analysis of the existing research in the fields of
-          natural language interfaces (NLI) for databases, GPU integration for acceleration
-          of database operations, and LLM/AI Model integration within database systems.
-          This phase establishes the theoretical foundation for this research and identifies current 
-          state-of-the-art approaches, their benefits and shortcomings.
-     \item \textbf{Decomposition \& Requirements} — Decomposing the problem statement into its
-          fundamentals and deriving system requirements for the design phase from it. The goal
-          of this section is to arrive at a list of functional and non-functional requirements that
-          must be taken into account and fulfilled by the design and implementation phases respectively.
-     \item \textbf{System Design} — Design of a system architecture that can utilize GPU acceleration
-          for LLM integration from within postgres. The primary goals of the system design phase
-          are to arrive at an architecture that yields low latency natural language processing,
-          schema-aware SQL query generation, ambiguity detection and resolution whilst maintaining
-          a high semantic accuracy.
-     \item \textbf{Implementation} — The implementation of a PostgreSQL extension according to the
-          above system design that relies on \texttt{rust} and \texttt{pgrx}. This extension will
-          provide a GPU accelerated framework for executing LLMs, implement a natural language
-          to query generation pipeline that relies on the SQL schema and create database functions
-          and operators for both query generation and execution.
-     \item \textbf{Evaluation and Benchmarking} — An assesment framework and benchmark that introspects
-          the implementations performance in multiple dimensions. Namely the most relevant dimensions
-          for this thesis are:
-          \begin{enumerate}
-              \item Semantic Accuracy — Measuring the overall accuracy of results delivered for a given
-                    natural language input.
-              \item Ambiguity Resolution Capabilities — How well the system performs when confronted with
-                    ambiguous natural language input and database schemas.
-              \item Performance Metrics — Measuring the latency, throughput and resource utilization 
-                    of the implementation.
-          \end{enumerate}
-      \item \textbf{Discussion} — Analysis and interpretation of the evaluation phase results against
-            the research goals of this thesis. Evaluating the performance and accuracy results recorded
-            during the benchmarks against the question whether real world deployments of NILs are feasible.
-            Furthermore the effectiveness of ambiguity resolution capabilities and semantic accuracy enhancement
-            strategies are showing a statistically significant effect.
-       \item \textbf{Summary and Outlook} — Summarizes the contributions, addresses limitations
-            of this thesis and the implementation, and proposes directions for future research alongside
-            possible applications. Primary future research topics include advanced GPU optimization
-            techniques (e.g. further quantization), accuracy and performance impact of model fine tuning,
-            techniques, scalability of such a system in enterprise scenarios and the evaluation of security
-            and privacy considerations (e.g. managing access control).
-\end{enumerate}
+\import{literature-review}{index}
 
@@
 
-\section{Literature Review}
+% 1/3 Theory, Decomp und System Design
 
-% https://github.com/HKUSTDial/NL2SQL_Handbook/blob/main/assets/river.svg
-
-In this section a comprehensive literature review is performed to asses the research landscape on NL2SQL
-(sometimes also referred to as Text-to-SQL or T2SQL) and NLIDBs. From the time their development accelerated in
-the late 1990s and early 2000s \citep{NLIDBs, NLIDBTheory, ILPParsing, ILPParsing2} until now, observing multiple 
-larger paradigm shifts happening over time \citep{GRAPPA, STRUG, Seq2SQL, NALIR, SQLizer}. In particular this 
-research focuses on the recent advancements when it comes to language models and how they can be harnessed for
-effective NL2SQL systems \citep{LLM-Sql, DAIL-SQL, T2SQL-LLM-Bench-2, T2SQL-LLM-Bench-3, Spider2, BIRD}.
-
-This literature review is covering the foundational concepts, challenges, key advancements and research gaps
-associated with using natural language instead of SQL. It lays the foundation for this thesis and helps to set
-the research questions introduced in the previous chapter in context.
-
-% bessere einleitung
-
-\subsection{Foundations of Natural Language Interfaces to Databases}
-
-Earlier papers in the research landscape on Natural Language Database Interfaces (NLIDBs) date over half
-a century back, into the early 1970s. Two decades after the first major research systems where developed
-in this domain, \citeauthor*{NLIDBs} have published an introduction and an overview over NLIDBs where an overview of
-state-of-the-art approaches were provided. \citep{NLIDBs} Their work outlined multiple key issues and challenges
-associated with NLIDBs, and compared them against existing / competing solutions like formal query languages,
-form-based interfaces and graphical interfaces. These challenges (like unobvious limits, linguistic ambiguities,
-semantic inaccuracy, tedious configuration etc.) have shaped this field of research and are still considered relevant
-metrics today.
-
-Early NLIDBs primarily relied on traditional natural language processing (NLP) techniques in order to achieve
-natural language understanding capabilities. With \textsc{Chill} an inductive logic programming (ILP) approach
-was first introduced for NL2SQL systems, marking one of the key events when it comes to machine learning usage.
-\citep{ILPParsing} In \citeyear{ILPParsing2} \citeauthor*{ILPParsing2} have extended the approach of ILP
-parsing for natural language database queries with multi clause construction, yielding promising results
-in the field of NLIDBs. \citep{ILPParsing2}
-
-Building on the systematic overview of \citeauthor*{NLIDBs} and the first machine learning approaches from
-\citeauthor*{ILPParsing} aswell as \citeauthor*{ILPParsing2}, \citeauthor{NLIDBTheory} have proposed a novel 
-approach for implementing NLIDBs and outperformed at the time state-of-the-art solutions from \cite{ILPParsing} 
-\cite{ILPParsing2} — achieving 80\% semantic accuracy. \citep{NLIDBTheory} The novelty of the \textsc{Percise}
-system lies in its natural language processing approach, specifically its lexical mapping strategy, allowing 
-\textsc{Percise} to identify questions it can, and can't answer (introducing the concept of \textit{semantically 
-tractable questions}) which therefore results in a better and interactive end user experience. Their experiements 
-also showed that this approach is \textit{transferrable} and \textit{unbiased} — it is possible to feed in new, 
-unknown questions into the system and maintain performance characteristics, whereas it was shown that 
-\cite{ILPParsing} were suffering from a distribution drift of the questions asked. \citep{NLIDBTheory}
-
-The theoretical foundations and research questions highlighted by the aforementioned works, shaped the research
-field and highlighted the following, ongoing research:
-
-\begin{enumerate}
-    \item The trade-off characteristics derived from choosing a machine learning vs. traditional NLP approach (e.g. 
-          \textsc{Chill} versus \textsc{Percise}). E.g. coverage versus correctness. \citep{ILPParsing, NLIDBTheory}
-    \item The linguistic challenges associated with bringing NLIDBs into use (e.g. semantic inaccuracy, linguistic 
-          ambiguity, unclear language coverage etc.) \citep{NLIDBs}
-    \item The value of systems and approaches which double down on reliability and semantic accuracy rather than giving
-          promising but incorrect answers. \citep{NLIDBs, NLIDBTheory}
-\end{enumerate}
-
-Fundamentally this highlights the tension and mismatch between the characteristics of natural language, which is 
-able to be ambiguous, \textit{semantically untractable} or able to be incomplete in meaning and formal languages
-like SQL which always have on deterministic and \textit{semantically tractable} meaning they convey in each statement. 
-As Schneiderman and Norman have pointed out according to \citeauthor*{NLIDBTheory}, users are ``unwilling to trade 
-reliable and predictable user interfaces for intelligent but unreliable ones'' which induces performance expectations
-on NLIDB implementations to be highly certain about the questions it can, and can't answer, whilst maintaing as high
-as possible natural language coverage. \citep{NLIDBTheory}
-
-\subsection{Traditional NL2SQL Approaches}
-
-Prior to the wide-spread dominance of machine learning approaches for natural language processing a variety of
-traditional, rather discrtete approaches have been explored in the field of NL2SQL / NLIDBs. These logical
-programming approaches have laid the foundations for transitioning towards the application of machine learning
-techniques for NL2SQL.
-
-\subsubsection{Rule-based and Grammar-based Systems}
-
-Foundational research of NL2SQL system mostly focused around applying rule engines that were tedious to set up and
-expensive to maintain / transfer across database systems. These rule engines mostly relied on the systematic 
-identification of linguistic patterns / were trying to template SQL from information that was derived from processing the 
-natural language query. \citep{Rendezvous, Lunar, Ladder} These approaches mostly tried to formalize natural language 
-queries into formal grammers which could then be deterministically mapped into a valid SQL query. \citep{Lunar} These 
-approaches have strong downsides when it comes to the variety of natural language constructs they can process, aswell as 
-runtime adoption of new / unknown databases, query constructs etc. A potential upside of this class of NL2SQL systems is 
-that they can confidently and reproducably identify questions they can, and can't answer — thus leading to very reliable 
-and predictable user interfaces.
-
-\subsubsection{Semantic Parsing using String-Kernels}
-
-A significiant milestone in parsing techniques of natural language queries was reached by \citeauthor*{StringKernels} in 
-\citeyear{StringKernels}. The introduction of string kernels for semantic parsing represented a novel achievement, when 
-it comes to fusing logical programming approaches using a formal grammer like LSNLIS developed by \cite{Lunar} and 
-learning / training approaches to understand unseen language patterns / unknown natural language query structures. This 
-allowed for more flexible pattern recognition when compared to traditional rule-based systems.
-
-The core innovative characteristic of this approach lies in its capability to understand similarities between natural 
-language expressions based on subsequence patterns rather than relying on exact matches. This made \textsc{Krisp}, the 
-research NLIDB system developed by \cite{StringKernels} much more robust to language variations in phrasing and noise 
-(e.g. spelling mistakes) in the input. As the \citeauthor*{StringKernels} demonstrated through experiments on real-world 
-datasets, this approach compared favorably to existing systems of the time like \textsc{Chill}, especially in handling 
-noisy inputs — a frequent challenge rigid rule-based systems faced in real world scenarios \citep{StringKernels, 
-ILPParsing}.
-
-\subsubsection{Graph Matching Methods}
-
-\cite{GraphMatching} brought together several research threads and reapplied emerging graph matching research models
-to natural language processing, specifically to natural language queries. Graph matching was applied once the natural 
-language query was parsed using a Combinatory Categorial Grammar (CCG) approach into a semantic graph which denotes the 
-relationship between semantic entities in it. This graph could then be matched against the actual graph derived from 
-the database, since they share topological traits that can be used for matching \citep{GraphMatching}. This approach 
-allowed to apply querying systems without having any question-answer pairs or manual annotations for training the system, 
-which implies easier scalability / transferability across domains, since the system does not require any additional 
-tweaks.
-
-Even though this approach was novel and showed improved the performance over existing state-of-the-art approaches, it
-was showing that graph matching quickly reaches its limitations. This approach relied heavily on the CCG parser's 
-accuracy, with parsing errors accounting for 10-25\% (depending on the dataset) of system failures \citep{GraphMatching}. 
-Furthermore it struggled with both ambiguous language constructs and potential mismatches between natural language 
-representation of relationships and database layouts — more complicated database designs, which may not match the users 
-intuitive understanding resulted in a different topology and hence could not be matched \citep[p.~387]{GraphMatching}.
-
-\subsubsection{Interactive Systems}
-
-In \citeyear{NALIR} \citeauthor*{NALIR} identified that perfect translation of natural language into SQL was challenging
-due to natural language not being made for query expressions as it heavily relies on contextual information and clarifying
-questions in order to disambiguate conversations \citep{NALIR}. These learnings relate to early prior art from 
-\citeauthor{UnnaturalQueryLanguage} and \citeauthor{Rendezvous} which also made this observation — ``natural language
-is not a natural query language.'' \citep{UnnaturalQueryLanguage}. The solution introduced by NaLIR further emphasized
-how important an interactive, conversational usage model is, when offering a natural language interface \citep{NALIR}.
-
-NaLIR could accept logically complex English language sentences as input and translate them into SQL queries with various 
-complexities, including aggregation, nesting, and different types of joins etc. The key innovative characteristic of 
-NaLIR lies in its interactive communication mechanism (much like \textsc{Rendezvous}) that could detect potential 
-misinterpretations and engage users to resolve ambiguities present in their natural language query without forcing them
-to entirely rephrase their query \cite{NALIR}. This approach, while showing awareness for its limitations (with regards
-to entirely automating / deriving SQL generation from potentially ambiguous or faulty user input) showed that it was 
-possible to overcome these limtations through choosing the right interaction model — ``In our system, we generate
-multiple possible interpretations for a natural language query and translate all of them in natural language for the
-user to choose from'' —, rather than optimizing the generation part of the system \cite{NALIR}.
-
-\subsubsection{Query Synthesis}
-
-\cite{SQLizer} introduced SQLizer, which synthesizes SQL queries from natural language \citep{SQLizer}. This paper
-presents a novel approach when it comes to NL2SQL as it is merging prevalent semantic parsing techniques (outlined above)
-with an program synthesis (or query synthesis) approach. SQLizer makes use of a three stage processing model for 
-natural language models: first generating a sketch of the query using semantic parsing, then using type-directed
-synthesis to complete the sketch and finally using automated repair, if required.
-
-\citeauthor*{SQLizer} show that alternating between repairing and synthesis yields results that beat state-of-the-art
-NL2SQL approaches like NaLIR. SQLizer is fully automated and database-agnostic, requiring no knowledge of the underlying
-schema. The authors evaluated SQLizer on 455 queries across three databases, where it ranked the correct query in the
-top 5 results for roughly 90\% of the queries. This represents a significant improvement over NaLIR \citep{NALIR},
-the previous state-of-the-art system \citep{SQLizer}.
-
-Potential short commings of this approch include queries which yield empty results, dealing with language variations
-as SQLizer is still using semantic parsing, and domain-specific terminology, all while still requiring users
-to select from multiple query options which reduces the overall usability of the system \citep[p.22-23]{SQLizer}.
-
-\subsubsection{Limitations of Traditional Approaches to NL2SQL}
-
-Despite being innovative and achieving state-of-the-art results, many of the above outlined approaches face severe 
-challenges when moving outside of an research environment. Many of these systems performed comparatively good on research
-benchmarks that were often composed of controlled question types and limited data variety. Ultimatively no standard
-benchmark existed for NL2SQL in this era, hence comparing different NL2SQL systems against each other is a problem on 
-its own. Despite not having a standard benchmark that all approaches could be unifiably evaluated against, several 
-fundamental challenges emerged / remained with these approaches:
-
-\begin{enumerate}
-    \item \textbf{Limited linguistic coverage} — Prevalent rule-based and semantic-parsing based systems were only able to 
-          process the a small subset of the natural language they were programmed for. This severely limited their 
-          ability to handle different phrasings of the same end-user goal \citep{StringKernels, UnnaturalQueryLanguage, 
-          Lunar, Ladder}.
-    \item \textbf{Transferability} — Traditional approaches typically required extensive manual configuration or at least
-          a training phase / adaption for each database they were deployed for, hindering cross domain usage through being 
-          expensive and time-consuming to adapt \citep{NLIDBs, Lunar}.
-    \item \textbf{Brittleness} — Many of the systems introduced in this subchapter did not handle synonyms, paraphrasing,   
-          or spelling errors well. Manual adaption / handling was needed in order to becomes resilient against each class
-          of problems \citep{StringKernels, SQLizer}.
-    \item \textbf{Poor scalability} — With potentially more complex underlying databases, traditional 
-          solutions often showed to perform worse. \citeauthor*{GraphMatching} found, that with increasing schema
-          complexity more compute was required to resolve the natural language query to a suitable query candidate
-          making them less transferable and scalable than initially anticipated \citep{GraphMatching} — ``Evaluating on
-          all domains in Freebase would generate a very large number of queries for which denotations would have to be 
-          computed ... Our system loads Freebase using Virtuoso and queries it with SPARQL. Virtuoso is slow in dealing with
-          millions of queries indexed on the entire Freebase, and is the only reason we did not work with the complete
-          Freebase.'' which indicates underlying system design issues with runtime complexity.
-\end{enumerate}
-
-These flaws of traditional NL2SQL approaches made it apparent, that a different class of approaches is needed, which 
-increase transferability and reduce the brittleness since users are ``unwilling to trade reliable and predictable user 
-interfaces for intelligent but unreliable ones'' according to \cite{NLIDBTheory}. Whilst many approaches outlined tractable 
-ways to increase user satisfaction and accuracy (like \citeauthor*{Rendezvous} did in \citeyear{Rendezvous} with a 
-conversational approach), NLIDBs were and are not considered to be a solved problem.
-
-\subsection{Neural NL2SQL Approaches}
-
-The previously outlined limitations of traditional approaches to solving NL2SQL / implementing NLIDBs pushed the research
-branch around neural network application forward to step in and propose new solutions which address the brittleness, 
-transferability and scalability concerns addressed with logical programming approaches. Neural approaches showed to yield
-significant improvements in terms of transferability and overall accuracy which led to a paradigm shift in this research
-field.
-
-\subsubsection{Early Neural Approaches}
-
-In \citeyear{Seq2SQL} \citeauthor*{Seq2SQL} released Seq2SQL which represents a significant breakthrough and leap in NLIDB 
-research. Seq2SQL was an early research system that in the field of neural network application and as one of the first papers
-to frame the implementation of NLIDBs / NL2SQL Systems as a reinforement learning problem. The system utilized iterative query 
-execution in the reward function to improve its accuracy \citep{Seq2SQL}. In the same paper \citeauthor*{Seq2SQL} introduced 
-WikiSQL, a training dataset, which enables large scale (in \citeyear{Seq2SQL}) model training.
-
-SQLNet \citep{SQLNet} addressed primarily the order-sensitivitiy trait of Seq2SQL \citep{Seq2SQL} that was prevalent due to being
-a derivative approach from sequence-to-sequence approaches. SQLNet diverges from sequence-to-sequence and joins multiple research 
-threads, employing a sketch-based query generation. SQLNet breaks down complex queries into smaller (hence more manageable)
-sub-queries which can then be individually sketched and refined, yielding a system that outperformed state-of-the-art by 9\% 
-to 13\% \citep{SQLNet}.
-
-\citeauthor*{TypeSQL} have introduced \textsc{TypeSQL}, a variation of the SQLNet-approach, in \citeyear{TypeSQL}.
-\textsc{TypeSQL}'s primary difference to SQLNet is the encoding of type information for SQL generation. The approach scanned
-for entity references and values in natural language and was able to improve performance by 5.5\% over SOTA-Models like
-SQLNet whilst requiring significantly less training time, indicating that type information was a useful information for deriving 
-accurate SQL queries from user input \citep{TypeSQL}.
-
-\subsubsection{Intermediate Neural Developments}
-
-Later in \citeyear{SyntaxSQLNet} \citeauthor*{SyntaxSQLNet} released SyntaxSQLNet, a followup research to \textsc{TypeSQL},
-which represented a slight change in approach and research focus. In direct comparison SyntaxSQLNet focused primarily around
-complex query generation using a syntax tree decoder, allowing for longer and more cohesive query generation \citep{SyntaxSQLNet}.
-This advancement over \textsc{TypeSQL} allowed more complex queries to be reliably generated, enabling multiple clauses aswell as
-nested queries. SyntaxSQLNet was one of the earlier research efforts which utilized Spider instead of WikiSQL (introduced by 
-\cite{Seq2SQL}), a large-scale NL2SQL dataset, incorperating 10.181 hand annotated natural language question and alongside
-5.693 unique SQL examples that spread across 138 different domains \citep{Spider}. This research led the transition of
-comparatively simple, research-grade, neural systems for NLIDBs towards systems which are feasible in the real world.
-
-Building on the above approaches, \citeauthor*{IRNet} have introduced IRNet, a neural network approach using intermediate
-representation as a bridge between natural language and SQL in which semantic queries could be expressed. The intermediate
-format SemQL (or semantic query language) was utilized to transform and synthesize queries on the actual database schema more
-accurately than Seq2SQL. IRNet followed a three phase approach: schema linking between the natural language query and database
-layout, synthesis of SemQL as intermediate representation and deterministic conversion of SemQL to SQL. This approach allowed
-IRNet to outperform state-of-the-art approaches on the \textsc{Spider} benchmark by 19.5\%, placing IRNet at an overall accuracy
-of 46.7\% \citep{IRNet}.
-
-Following IRNet, graph neural networks (GNN) have been explored as alternative architecture by \cite{GNN}, representing the
-database schema as a graph and using message passing to model relationships between tables, columns and natural language input.
-This approach demonstrated the capability to improve reasoning and query generation capability. \citeauthor{GNN} showed that when 
-evaluating against the \textsc{Spider} benchmark GNN outperforms both SyntaxSQLNet (and therefore \textsc{SQLNet} and
-\textsc{TypeSQL}). Although presenting a signficiant advancement over previous state-of-the-art approaches, GNN falls behind in
-performance against IRNet by 6\% \citep{IRNet, GNN}.
-
-\subsubsection{Relation-Aware Transformer Approaches}
-
-The release of RAT-SQL (Relation-Aware Transformer for SQL) \cite{RATSQL} represents the most significant leap in research of
-neural NL2SQL approaches. RAT-SQL diverged from earlier research through emphasizing the relationship between natural language
-and the database schema elements using relation-aware self-attention representing a novel approach for solving \textit{schema 
-linking} \citep{RATSQL}.
-
-RAT-SQL's primary innovations was the ability to infer, understand and utilize the relationship between individual tokens in the 
-natural language query and link it to the database schema. Thus allowing for reasoning capabilities on the actual database schema
-while generating the query.
-
-This architecture yielded a 57.2\% in exact match accuracy when being evaluated on the \textsc{Spider} benchmark, substantially
-outperforming comparative approaches like GNN, IRNet and IRNet V2 by 10.5\%, 9.8\% and 8.7\% respectively. Although overall
-accuracy improved across all approaches when being paired with BERT (Bidirectional Encoder Representations from Transformers, a 
-popular pre-trained lanugage model from Google) the $\delta$ between the indivudual approaches remained relatively steady, leaving
-RAT-SQL outperforming state-of-the-art approaches by 5\% to 12.2\% further demonstrating the capability advancement yielded by this 
-system \citep{RATSQL}.
-
-\subsubsection{Comparative Analysis of Neural Approaches}
-
-The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
-neural NL2SQL approaches in different dimensions:
-
-\begin{enumerate}
-    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
-          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
-          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
-          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
-          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
-          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
-    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
-          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
-          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
-          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
-          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
-          training data \citep{, RATSQL}.
-    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
-          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
-          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
-          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
-          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
-    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
-          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
-          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
-          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
-          by up to 10.5\% \citep{IRNet, RATSQL}.
-    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
-          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
-          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
-          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
-\end{enumerate}
-
-\subsubsection{Limitations of Neural Approaches}
-
-Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
-approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:
-
-\begin{enumerate}
-    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
-          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
-    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
-          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
-          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
-    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
-          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
-          specific natural language queries challenging.
-    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
-          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
-\end{enumerate}
-
-The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of
-NLIDBs. The research shift started in this era established the foundations for further and more advanced machine learning approaches
-(specifically language model oriented approaches )being researched. Neural approaches showed to significant improvements in performance
-when being paired with pre-trained language models \citep{RATSQL} which led to further research on their applicability.
-
-\subsection{Pre-trained Language Models}
-
-The advantages of combining specialized neural networks with general-purpose pre-trained language models
-led to a pivotal point in the NL2SQL research field towards focusing increasingly on the application of pre-trained
-language models for NLIDBs. Models like BERT or T5 offer noticable performance improvements (especially when it
-comes to language understanding) over specialized NL2SQL networks due to training happening on unrestrained amounts
-of natural language data, instead of pure NL2SQL datasets which are often fairly limited in size and therfore natural
-language use — \textsc{Spider2.0} which is a contemprary NL2SQL benchmark consists of just 632 real-world questions 
-\citep{Spider2}. Thus PLM-based (or at least augumented) NL2SQL systems can observe dramatic performance improvements
-through the language models's ability to understand patterns and identify semantic relationships of natural language
-query elements.
-
-\subsubsection{Early Pre-trained Language Model Adaptations}
-
-The above outlined benefits have led to concrete research efforts focusing on the question whether the sole application
-of pre-trained language models could outperform neural state-of-the-art approaches — which often implicitly require
-a far more sophisticate architecture when it comes to natural language analysis.
-
-In the time of emerging PLM applicationm \textsc{Grappa} was introduced by \citeauthor*{GRAPPA} in \citeyear{GRAPPA} —
-a novel grammar-augmented pre-training approach built on RoBERTa\textsubscript{\tiny{LARGE}} (a derivative model from BERT).
-It generates synthetic training data (ie. natural language and sql pairs) using a synchronous context-free grammar (SCFG)
-which analyses and identifies patterns in natural language queries that can be used as templates for sythesizing training
-data. The specialized pre-training helps \textsc{Grappa} to establish a robust connection between natural-language and
-database schema elements, showing significant improvements on existing approaches on multiple contemporary benchmarks like 
-\textsc{Spider} and \textsc{WikiSQL} \citep{GRAPPA}.
-
-Several NL2SQL approaches in this era focused on \textit{schema understanding} and \textit{schema linking} — the 
-generalizability of PLMs required advanced techniques on ensuring that models both understand the semantic intent
-of users when querying and correctly identify database schema elements in natural language queries. Thus improving
-semantic accuracy of generated SQL queries. \textsc{StruG} (Structure-Grounded-Pretraining) was introduced in
-\citeyear{STRUG} by \citeauthor*{STRUG} and presented a novel pretraining approach that improves model abilities when it comes to
-\textit{schema linking}, it separates the problem in three facets: column grounding, value grounding and
-column-value mapping. In direct comparison with \textsc{Grappa}, \textsc{StruG} achieves similar performance
-while being significantly cheaper to train \citep{STRUG}.
-
-In parallel, \citeauthor*{GAZP} released GAZP (Grounded Adaptation for Zero-shot Executable Semantic Parsing) in
-\citeyear{GAZP}. \citeauthor*{GAZP} specifically addressed the challenge of adapting semantic parsers across databases
-/ domains which was a apparent problem with neural approaches which had a strong tendency to overfit on benchmark datasets. 
-Its novel contribution was the combination of forward semantic parsing with a backward utterance generator which allowed
-for data synthesis in unseen environments which could then be used to adapt the semantic parser \citep{GAZP}. This
-approach enables a improvement in robustness and accuracy in situations where training and inference environments
-differ without requiring manually annotated examples \citep{GAZP}.
-
-\subsubsection{Advanced Pre-trained Language Model Approaches}
-
-Building on earlier foundational research on PLM application for NL2SQL tasks, researchers have developed increasingly complex
-systems that leveraged pre-trained language models whilst addressing their limitations when it comes to generating valid SQL.
-
-\citeauthor*{RYANSQL} introduced \textsc{Ryansql} (Recursively Yielding Annotation Network for SQL) in \citeyear{RYANSQL},
-which implements a sketch-based approach for decomposing complex SQL generation into multiple smaller problems. \textsc{Ryansql}
-transformed nested statements into a set of top-level statements using the Statement Position Code (SPC) technique. This
-flattening of structure allowed \textsc{Ryansql} to limit the complexity of the query generation problem whilst maintaining
-its ability to answer complex questions by recomposing complex queries from their parts. This approach allowed \textsc{Ryansql}
-to achieve 58.2\% accuracy on the \textsc{Spider} benchmark, representing a 3.2\% improvement over contemporary state-of-the-art
-approaches at the time \citep{RYANSQL}. The sketch-based approach makes \textsc{Ryansql} a PLM-augumented successor of \textsc{SQLNet}
-which was a early neural approach to employ sketch-based query generation \citep{RYANSQL, SQLNet}.
-
-A significant advancement in terms of execution accuracy was reached with the application of T5-Models for NL2SQL tasks. T5
-(Text-to-Text Transfer Transformer) Models have proven themselves as well-suited for for query generation — T5-3B for NL2SQL
-yielded 71.4\% execution accuracy and thus presented a breakthrough in this domain of research \citep{T2SQL-LLM-Bench-3}.
-This established a new baseline for PLM-based approaches and demonstrated that general-purpose language models could not
-only compete but outperform specialized architectures by far when properly fine-tuned \citep{T2SQL-LLM-Bench-3}.
-
-Following the adavancements through T5, \citeauthor*{GRAPHIX} introduced GRAPHIX-T5 in \citeyear{GRAPHIX}, which combined
-the T5 PLMs with a further graph-aware layers for NL2SQL tasks. This architecture could leverage both pre-trained knowledge
-of T5 models aswell as the database schema structure during inference. GRAPHIX-T5 constructs a schema graph where nodes
-represent tables and columns and edges represent relationships between them, such as foreign keys or columns association.
-This architecture allows the model to deeply understand relationships and the layout of the database schema \citep{GRAPHIX}. 
-GRAPHIX-T5 outperformed standard T5 models significantly, with GRAPHIX-T5\textsubscript{\tiny{LARGE}} showing 6.6\% increase
-in execution accuracy over T5\textsubscript{\tiny{LARGE}}. When both GRAPHIX and the baseline T5 models were combined with
-\textsc{Picard} (a novel constrained decoding mechanism) absolute $\delta$ between them jumped to 7.6\% (81.0\% in absolute
-numbers), evaluated on \textsc{Spider-Dev} \citep{GRAPHIX}.
-
-In parallel, \citeauthor*{RESDSQL} proposed \textsc{Resdsql} in \citeyear{RESDSQL}, which proposed to decouple
-\textit{schema linking} and \textit{skeleton parsing}. This addressed the typical challenges sequence-to-sequence models
-faced when simultaneously trying to link both schema elements and generate the query skeleton (e.g. 
-\texttt{SELECT <columns> FROM <table>}). \textsc{Resdsql} further employed a ranking approach to filter schema elements
-before passing them to the model for query generation, which reduced noise (when working with large database schemas)
-and enabled passing only the most relevant parts. This twofold approach allowed \textsc{Resdsql} to achieve state-of-the-art
-performance when being evaluated on \textsc{Spider}, outperforming GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD by 0.8\% in
-execution accuracy. When combined with \textsc{NatSQL} (a contemporary intermediate representation approach introduced
-by \cite{NATSQL}) absolute improvement over GRAPHIX-T5\textsubscript{\tiny{3B}}-PICARD jumped to 3.1\% emphasizing the
-robustness gain decoupeled architectures have over model-oriented approaches.
-
-These advancement showed rapid improvements over earlier methods — far surpassing neural approaches — through advanced
-mechanisms when it comes to schema understanding and query generation. The wide language understanding inherited from
-PLM-basemodels further strengthens robustness and shows effectiveness through a large gain on the \textsc{Spider}
-benchmark. Collectively these approaches represent a leap in NL2SQL research, emphasizing their usability potential
-and real-world feasibility. This era primed the research field for the transition towards large language model adoption.
-
-\subsubsection{Constrained Decoding and Ranking Techniques}
-
-A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
-but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
-(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
-for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
-rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
-language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
-on the \textsc{Spider} benchmark \citep{PICARD}.
-
-\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
-evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
-incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
-output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
-with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
-SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.
-
-The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
-encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
-to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.
-
-\subsubsection{Advantages of PLM Approaches}
-
-PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
-leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
-large language models in the following years. Namely PLM approaches brought a series of upsides with them:
-
-\begin{enumerate}
-    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
-        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
-        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
-    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
-        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
-        approaches \citep{GRAPPA, STRUG}.
-    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
-        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
-        primarily trained on the development sets of contemporary benchmarks.
-\end{enumerate}
-
-\subsubsection{Limitations of PLM Approaches}
-
-Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
-non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
-must be considered when using a PLM-based approach to NL2SQL:
-
-\begin{enumerate}
-
-    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific,
-          fine-tuning, limiting a straight forward adaptation to new domains or databases. Although being significantly
-          more efficient than LLM-based approaches the potential need for initial fine-tuning represent a significant
-          computational resource burden. Furthermore when not using synthetic data generation (e.g. \textsc{Grappa})
-          annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
-    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often
-          far larger than needed 
-          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each
-          decoding step, they can produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the
-          input and output token space, therefore multiple approaches have been researched which focus on constraining
-          these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard} have proposed potential
-          (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
-    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to
-          incorporate limited amounts of schema awareness when being applied out of the box for NL2SQL tasks. Multiple
-          research efforts focused on improving this situation, most notably \textsc{Resdsql} and GRAPHIX-T5 tried to
-          improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX}, nonetheless the non-specialized nature
-          of PLMs prevents NL2SQL being part of the fundamental model architecture.
-\end{enumerate}
-
-These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially
-in direct comparison with neural approaches), ultimately yielding the research domain to transition toward exploring
-Large Language Model approaches that promise even greater flexibility in adaptation and potentially superior handling
-of complex queries through advanced in-context learning approaches.
-
-\subsubsection{Comparison with Large Language Models}
-
-The research on applying pre-trained language models for NL2SQL tasks primed the field for the transition towards LLM usage. While
-PLMs like T5 and BERT range from millions to a few billion parameters, prevalent LLMs such as GPT-3 and GPT-4 operate at
-significantly larger scales, ranging from a few billions to hundred of billions parameters. The scale of LLMs enables in-context
-learning techniques that enable significantly easier and cheaper transferability of NL2SQL systems across domains \citep{DAIL-SQL}.
-The $\delta$ of deployment, inference and training requirements of these two approaches are significant due to the size difference
-in models, which transfer to hardware requirements and therefore cost. While PLMs can require extensive fine-tuning on domain-specific
-data which may aswell be resource intensive \citep{GRAPHIX, RESDSQL, GRAPPA, STRUG}, LLMs transfer the cost to the inference environment,
-where model modificants are less impactful, due to the extensive pre-training that took place. Approaches like \textsc{DinSQL} show
-that with the application of LLMs the engineering challenges around model instruction gained relevance while model training became
-less of a central problem to solve \citep{DINSQL}.
-
-\subsection{Large Language Models}
-
-The emergence of LLMs such as GPT-3, GPT-4, and Claude fundamentally transformed the landscape of NL2SQL research. Early experiments 
-with LLMs for NL2SQL tasks showed state-of-the-art capabilities in comparsion with contemporary PLM approaches \citep{DAIL-SQL}.
-\cite{T2SQL-LLM-Bench-3} demonstrated that \textsc{Codex} (a contemporary model based on GPT-3), without any fine-tuning efforts,
-could achieve competitive performance on \textsc{Spider}, outperforming many state-of-the-art approaches that required extensive
-training. This breakthrough challenged the contemporary assumption that further specialization of model architectures would yield
-increases in NL2SQL performance (e.g. \textsc{Graphix-T5}) \citep{GRAPHIX}.
-
-\subsubsection{In-Context Learning}
-
-In-Context Learning (ICL) is a foundational approach for leveraging the ability of LLMs to utilize larger context windows for
-inference than traditional PLMs. Typical context windows of state-of-the-art LLMs can reach up to hundred thousands of tokens.
-This characteristic of LLMs enabled researchers to utilize this context window to provide examples of accurate NL2SQL translation
-instead of applying parameter updates. This paradigm shift has made developing NL2SQL systems significantly more accessible.
-
-The fundamental principle of few-shot learning for NL2SQL involves providing the LLM with a small number of example pairs
-of natural language and their corresponding SQL representation. These examples can benefit the model's understanding of
-mapping between natural language and SQL syntax. This essentially builds on top of prior research like \textsc{Grappa}
-and \textsc{StruG}, but applying these examples at inference time, rather than training time. Although this increases
-the inference cost of such a system, the upsides lie primarily in the flexibility of such an approach — database content
-/ prior usage of the system can be dynamically utilized, rather than requiring retraining.
-
-Example selection strategies showed to have a considerable impact on ICL performance. \cite{DAIL-SQL} evaluated
-various example selection methods like \textit{Random}, \textit{Question Similarity Selection (QTS)}, 
-\textit{Masked Question Similarity Selection (MQS)}, and \textit{Query Similarity Selection (QRS)}. \citeauthor*{DAIL-SQL}
-propose a novel strategy to select, organize and present ICL examples to LLMs. DAIL-SQL utilizes both question and
-query similarity, masking domain-specific words and prioritizing examples that exceed a similarity threshold of $\tau$
-\citep[p.~5]{DAIL-SQL}. DAIL-SQL encodes examples as question-SQL-pairs without the respective schema to improve
-token efficiency. Using a Code Representation Prompt (CR) for question and schema encoding yielded DAIL-SQL to achieve
-state-of-the-art 86.6\% execution accuracy on \textsc{Spider}.
-
-The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
-to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
-(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
-still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
-execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.
-
-Especially with complex queryies which can involve multiple tables, nested queries and complex joins, zero-shot approaches
-often dramatically underperform $k$-shot ones. NL2SQL approaches that dont supply examples to the model during inference time
-fail more frequently to generate semantically accurate SQL queries \citep{DAIL-SQL}. Notable is the leap in exact match ratio
-measured by the \textsc{Spider} benchmark — jumping from 22.1\% for GPT-4 using zero-shot to 71.9\% with five-shot. The
-results presented by \cite{DAIL-SQL} show signficant correlation between $k$ and the execution accuracy of $k$-shot approaches.
-
-This effectiveness has established ICL approaches as a standard technique applied in LLM-based NL2SQL approaches. Contemporary
-approaches like \textsc{XiYan-SQL, Chase-SQL and Din-SQL} all utilize variations of ICL to achieve state-of-the-art results
-\citep{XiYan, CHASE, DINSQL}.
-
-\subsubsection{Self-Correction and Iterative Refinement}
-
-\cite{DINSQL} proposed DIN-SQL as an innovative approach to NLIDBs that rely on LLMs. DIN-SQL decomposes complex queries into
-sub-parts and utilizes in-context-learning and self-correction during the generation phase. Compared to DAIL-SQL which relies
-on example selection during the in-context-learning phase, DIN-SQL focuses on a refinement loop that allows the model to self-
-correct errors it made during the initial generation phase — thus the model can repair schema linking, syntactic or semantic
-errors. By explicitly instructing the LLM to review its work against a specfic schema, the user input and potential database
-errors, DIN-SQL achieves a high execution accuracy on \textsc{Spider} with 85.3\%. Therefore DIN-SQL outperforms contemporary
-approaches but is surpassed by by DAIL-SQL by 1.3\% \citep{DINSQL, DAIL-SQL}. Furthermore DIN-SQL makes observations on the
-impact that the self-correction prompt can steer results significantly — \citeauthor{DINSQL} found that using \textit{generic
-self-correction} (ie. assuming the query contains errors) lowers the execution accuracy by 4.2\% on \textsc{Spider} compared to
-\textit{gentle self-correction} (ie. assuming nothing about the validity of the query). It was noted that the impact of the
-self-correction mechanism relies on the model size, with smaller models performing better with \textit{generic self-correction}
-and larger models performing better with \textit{gentle self-correction} \citep{DINSQL}. The self-correcting nature of DIN-SQL
-represents a diversion from DAIL-SQL's emphasis on input optimization towards output refinement. \citeauthor{DINSQL} demonstrate
-how structured introspection can play a significant role in enhancing LLM performance for formal language generation tasks.
-
-Building upon DIN-SQL's self-correction module, \cite{MAGIC} proposed MAGIC (Multi-Agent Guideline for In-Context Text-to-SQL),
-which further advances the self-correction mechanism through harnessing a set of specialized agents to automate the self-correction
-prompt engineering \citep{MAGIC}. MAGIC consists of a manager agent, a correction agent and a feedback agent that collaboratively
-refine LLM instructions during the refinement loop. Further MAGIC derives common faliure patterns of the initial query generation
-phase from training data, allowing it to efficiently spot the most common mistakes that the model makes at generation time. This
-approach represent a further advancement on \citeauthor{DINSQL}'s DIN-SQL, effectively supersetting the \textit{generic} and
-\textit{gentle} correction mechanisms through an intelligent, self-adapting one \citep{MAGIC}. The autogenerated guidelines from
-MAGIC yield 85.6\% execution accuracy on the \textsc{Spider} development set — representing a 5.31\% improvement over DIN-SQL's
-human written correction guidelines. These results emphasize that optimized self-correction mechanisms have the ability to
-significantly drive up overall system performance of NLIDBs \citep{MAGIC}.
-
-While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
-Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
-for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
-effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
-each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
-small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
-ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.
-
-\subsubsection{Candidate Selection Frameworks}
-
-Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
-promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
-These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
-even with capable LLMs and modern self-correction mechanisms.
-
-\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
-query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
-CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
-language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
-inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
-learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
-characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
-\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
-state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.
-
-Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
-ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
-fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
-representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
-\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
-during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
-fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
-generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
-to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
-during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
-able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
-\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.
-
-Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
-selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
-The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
-the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
-to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
-models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
-transferability with the robustness of specialized model architectures for candidate ranking and selection.
-
-% Maybe: Agent specialization and collaboration patterns
-% Maybe: Voting and selection mechanisms for query ranking
-
-\subsubsection{Retrieval-Augumented Generation}
-
-Retrieval-Augmented Generation (RAG) has emerged as a powerful paradigm for enhancing NL2SQL systems by integrating
-external knowledge retrieval with the generative capabilities of LLMs. This technique is seen in all above introduced
-papers in varying forms \citep{XiYan, CHASE, CoE-SQL, MAGIC, DINSQL, DAIL-SQL}. The most prevalent form of RAG in
-NL2SQL is the encoding of the database schema into the LLM prompt for in-context-learning. This allows LLMs to be
-aware of the table structures and names, foreign key relationships, primary keys etc. \cite{XiYan} proposed the
-serialization of database schemas in the \textsc{M-Schema} format, a semi structured, text-based schema description.
-\cite{DAIL-SQL} proposed to encode the schema using a Code Representation Prompt (CR) which refers to the encoding of
-the raw SQL statements need to construct the schema. \cite{XiYan} provided an ablation study for the M-Schema which
-yielded questionable results on the optimization of this approach. While the M-Schema format yielded best results when
-\textsc{XiYan-SQL} was combined with GPT-4o or Claude 3.5 Sonnet, it performed worse than alternatives on DeepSeek
-and Gemini models \citep{XiYan}.
-
-A relevant optimization technique for RAG is the selection of a schema subset before encoding the schema for the model.
-\textsc{Resdsql} was one of the earlier approaches to explore subset-encoding of database schemas with \cite{RetAug}
-building on top of this \citep{RetAug, RESDSQL}. \cite{RetAug} introduced \textsc{ASTReS} which dynamically
-retrieves database schemas and uses abstract syntax trees (ASTs) to select optimal few-shot examples for ICL. 
-By pruning the ASTs down to the most relevant subset, \textsc{ASTReS} achieved the highest at-the-time (\citeyear{RetAug})
-execution accuracy on \textsc{Spider} with 86.6\% indicating that subset-encoding is a sensible optimization mechanism.
-\textsc{ASTReS} was combined with GRAPHIX-T5 in order to achieve this result \citep{RetAug}.
-
-The impact of RAG when NL2SQL systems face large and complex database schemas has been particularly significant. Traditional
-approaches struggle when database schemas contain hundreds or thousands of tables and columns, as the complete schema may not
-fit within model context windows depending on their size. RAG-based systems address this by dynamically retrieving only the most
-relevant portions of the schema based on the natural language query. The technique of subset-encoding becomes especially relevant
-in enterprise environments where database schemas can be extremely large and complex. Recent benchmarks like \textsc{Spider2.0}
-emphasize enterprise environments and show that existing solutions underperform in those scenarios, often reaching single digit
-execution accuracy.
-
-The work of \citeauthor{RetAug} indicates that prefiltering of the environment of language models is an effective
-and promising technique that has the ability to reduce computational requirements of contemporary NL2SQL systems.
-As introduced above, recent state-of-the-art systems often utilize closed-source models like Gemini, GPT-4(o), Claude 3.5/3.7
-Sonnet etc which often come with massive parameter sizes (reaching hundreds of billions of parameters). \textsc{ASTReS}
-demonstrated that efficient schema retrieval mechanisms enable smaller models (e.g. GRAPHIX-T5) to achieve state-of-the-art
-performance against LLM based approaches like DAIL-SQL \citep{DAIL-SQL, RetAug}.
-
-\subsubsection{Specialized LLMs and Fine-tuning}
-
-While general-purpose LLMs demonstrated state-of-the-art natural language understanding capabilities, the research
-domain of NLIDBs increasingly explored the potential of fine-tuned LLMs for NL2SQL tasks, which offer a promising
-tradeof between natural language understanding (ie. breadth of the model) and concrete SQL generation capabilities
-(ie. depth of the model). Multiple research works have been done on the fine-tuning of language models which
-yielded a series of dedicated models for optimized SQL generation.
-
-A significant limitation of many LLM-based NL2SQL solutions is their dependency on proprietary and closed-source
-LLMs like GPT-4(o), Claude and Gemini. Whilst they are useful for initially proving the potential of LLMs on
-contemporary benchmarks, this dependency introduces significant concerns related to data-privacy, high-side use 
-(e.g. in classified environments), transparency of data-flow and deployment costs. To address these challenges
-\cite{CodeS} have introduced \textsc{CodeS} in \citeyear{CodeS}, a series of of open-source language models dedicated
-for NL2SQL tasks with parameter sizes ranging from 1B to 15B. The \textsc{CodeS} models were evaluated against
-contemporary benchmarks like \textsc{Spider} and \textsc{Bird} and showed promising inference results when compared
-to their closed-source counterparts. \cite{CodeS} showed that \textsc{CodeS} 7B achieves 85.4\% execution accuracy
-on the \textsc{Spider} development set, outperforming both fine-tuning approaches like RESDSQL and GRAPHIX-T5-PICARD
-aswell as prompting based methods DIN-SQL+GPT-4 and DAIL-SQL+GPT-4 \citep{CodeS}. The same tendency was observed
-on \textsc{Bird} with \textsc{CodeS}-15B achieving 60.37\% execution accuracy, compared to 57.41\% for DAIL-SQL+GPT-4
-and 55.90\% for DIN-SQL+GPT4 \citep{CodeS}. This marks a significant advancement in the open-source language model
-research area, with \textsc{CodeS} reaching new state-of-the-art performance in \citeyear{CodeS}. While more recent
-approaches like \textsc{XiYan-SQL} and CHASE both outperform \textsc{CodeS}, CHASE relies on proprietary models and
-\textsc{XiYan-SQL} doesnt provide any information on what base models where used.
-
-\cite{CodeS} addressed serveral critical research challenges in the NL2SQL domain and proved that open-source models
-could perform competitively with proprietary models whilst maintaining a significantly smaller parameter footprint
-(ie. 7B and 15B) compared to GPT-4 which is a multi-hundred-billion parameter model. This makes it feasibile to deploy
-\textsc{CodeS} locally, instead of relying on an enterprise API like OpenAI's one \citep{CodeS}, therefore making it
-highly practical for real-world deployments where computation resource are constrained.
-
-\citeauthor*{OmniSQL} published a follow-up paper in \citeyear{OmniSQL} which introduced the next-generation \textsc{OmniSQL}
-models with 7B, 14B and 32B sizes, trained using synthetic data generation. \textsc{OmniSQL} achieves state-of-the-art
-performance when compared to alternative LLMs - including both open-source and closed-source competitors. The models
-achieve significant execution accuracy improvements on both \textsc{Spider} and \textsc{Bird}, the 7B model reaches 88.9\%
-on \textsc{Spider} and 66.1\% execution accuracy on \textsc{Bird} which represents a significant (5\%+) improvement over
-comparable alternatives \citep{OmniSQL}. These results were achieved without combining \textsc{OmniSQL} with advanced
-in-context-learning, self-correction, retrieval-augumented-generation or candidate-selection techniques, which indicates
-that even higher accuracy scores are possible.
-
-\subsubsection{Limitations and Challenges}
-
-Despite the impressive advancements of LLM-based approaches for NL2SQL, several significant limitations and challenges
-are apparent that impact their practical implementation and real-world deployability.
-
-\begin{enumerate}
-    \item \textbf{Hallucination and Accuracy Concerns} — Since LLMs demonstrate a tendency to hallucinate, LLM-based
-           NL2SQL systems face the challenge of detecting when LLMs generate plausible but incorrect SQL queries.
-           The breadth of possible errors ranges from detectable errors like invalid table or columns references,
-           invalid SQL syntax etc. to undetectable, slight errors in semantics, like reversing the order of
-           aggregations or using the wrong aggregation method. Contrary to traditional approaches which fail visibly with
-           unknown structures, LLM-based approaches might produce semantically flawed queries that execute without
-           errors and yet return semantically incorrect data. As noted by \citeauthor{NL2SQLUnsolved}, ``achieving
-           Enterprise-Grade NL2SQL is still far from being resolved'' even with state-of-the-art models, particularly
-           when handling complex real-world database schemas and ambiguous queries \citep{NL2SQLUnsolved, Spider2}.
-    \item \textbf{Computational Resource Requriements} — The resource intensity of LLM-based systems presents barriers
-           to widespread adoption. High-performance state-of-the-art models require substantial computational resources
-           during inference, making them expensive to deploy at scale. While smaller models exist, they typically show
-           reduced performance on complex queries, creating a challenging trade-off between accuracy and resource
-           efficiency. Studies like DAIL-SQL demonstrate a direct correlation between model size and performance,
-           where the most accurate systems are also the most challenging to deploy economically \citep{DAIL-SQL}.
-           Even state-of-the-art specialized LLMs like \textsc{OmniSQL} degrade in performance as parameter sizes
-           shrink, although they maintain a significantly higher parameter efficiency, outperforming enterprise-level
-           closed-source competitors. 
-    \item \textbf{Data Privacy and Security Implications} — Using LLMs particularly through proprietary APIs brings
-           along considerable concerns with regards to data privacy and security as sensitive data may be transferred
-           to respective model vendors. Potential senstive data aswell as database schemas alongside user questions
-           needs to be communicated to external parties in order to form a usable system if closed source models are
-           used. Recent open-source model developments like \textsc{CodeS} and \textsc{OmniSQL} seem to mitigate this
-           problem partially, but as mentioned above using local inference with LLMs brings along stark computational
-           requirements.
-    \item \textbf{Ambiguity Handling in Complex Scenarios} — Following previous paradigms LLMs also continue to struggle
-           with effective ambiguity resolution in natural language queries over complex database schemas. These challenges
-           are particularly prominent in large scale enterprise environments where similar column names exist across multiple
-           tables, or domain-specific terminology has multiple potential interpretations that can result in different queries.
-           Though approaches like multi-path reasoning and candidate selection show promising improvements in execution
-           accuracy, they often increase inference time and complexity as they run multiple parallel inference steps and
-           rank their results \citep{CHASE, XiYan}.
-    \item \textbf{Competitive PLM Approaches} — Prevalent PLM-based approaches like RESDSQL and GRAPHIX-T5 achieved 
-           impressive execution accuracy while using significantly fewer parameters than LLMs, making them more computationally
-           efficient \citep{RESDSQL}. Despite reducing inference-time requirements, these approaches typically utilize
-           re-training and domain specific fine-tuning which both limits their flexibility and introduces training costs.
-           Yet for certain scenarios PLM-based approaches offer an interesting tradeoffs: Frontloading the computational effort
-           can be interesting in compute-constrained environments and help to reduce ongoing costs. LLMs dont require the upfront
-           cost of fine-tuning and adaption but require significantly more resource for their deployment. This indicates
-           that certain systems and environments might benefit more from PLM approaches than LLM ones. Contemporary research,
-           which typically focuses around state-of-the-art-performance on benchmarks, shifted primarily towards LLMs due to
-           their transferability and natural language coverage.
-\end{enumerate}
-
-\subsection{Benchmarking}
-
-In order to evaluate NL2SQL systems standardized benchmarks like \textsc{Spider} and \textsc{Bird} have emerged. These
-benchmarks can measure performance across different approaches and models, enable meaningful ablation studies and are a
-useful indicator for the state of the research field. In the past decade significant advancements have been made with
-\textsc{Spider} being released in \citeyear{Spider} the first major, widely adopted, benchmark emerged in this field
-\citep{Spider}.
-
-\subsubsection{Spider}
-
-\textsc{Spider}, introduced by \citeauthor{Spider} in \citeyear{Spider}, has become the de facto standard benchmark for
-evaluating complex and cross-domain Text-to-SQL systems. It consists of 10,181 questions and 5,693 unique SQL queries
-spanning 200 databases across 138 domains. Previous benchmarks like lacked complexity and cross-domain distrubtion of
-datapoints which prevented the \textit{transferability} of approaches or models to be accounted for in benchmarks.
-With \textsc{Spider} the capability to be database agnostic was required to achieve meaningful accuracy scores. Furthermore
-\textsc{Spider} was split in training and test sets which contain different database in order to prevent overfitting
-models from succeeding. This design specifically tests a model's ability to handle schema linking and generalization
-challenges rather than memorizing specific database patterns. \textsc{Spider} evaluates both \textit{exact matching
-accuracy} and \textit{execution accuracy}, with contemporary state-of-the-art systems achieving approximately 85-90\%
-\textit{execution accuracy} (as of 2025) \citep{Spider, OmniSQL, XiYan, CHASE}.
-
-\subsubsection{Bird}
-
-The \textsc{Bird} benchmark (BIg bench for large-scale database gRounded Text-to-SQLs), released in \citeyear{BIRD},
-and addresses the gap between academic benchmarks and real-world applications by focusing on large-scale databases
-with actual data content \citep{BIRD}. \textsc{Bird} contains 12,751 text-to-SQL pairs and 95 databases with a total
-size of 33.4 GB across 37 professional domains. Unlike \textsc{Spider}, which primarily evaluates against database
-schemas with minimal content, \textsc{Bird} emphasizes challenges related to dirty database contents, external
-knowledge between natural language questions and database values, and SQL efficiency in massive databases. This
-places \textsc{Bird} as a relevant benchmark for real world feasibility of approaches and models. Even state-of-the-art
-LLMs like GPT-4 achieve only 54.89\% execution accuracy on \textsc{Bird}, compared to human performance of 92.96\%,
-highlighting the significant challenges posed by real-world database scenarios on NLIDBs \citep{BIRD}.
-
-\subsubsection{Spider 2.0}
-
-\textsc{Spider 2.0} which was introduced by \citeauthor*{Spider2} in \citeyear{Spider2} represents the most recent
-advancements of benchmarks for NL2SQL systems. It represents a significant evolution in NL2SQL benchmarking and focuses
-primarily on enterprise level database challenges. \textsc{Spider 2.0} is much smaller with only 632 real-world problems
-which were derived from enterprise database usecases, but yet represents a meaningful indicator for the complexity of
-databases that approaches and models can handle. \textsc{Spider 2.0} goes beyond simple query generation tasks and moves
-towards testing the deep understanding of the database, requiring models to understand metadata, SQL dialect documentation
-and project-level codebases \citep{Spider2}. The tasks contained in \textsc{Spider 2.0} often demand multiple complex SQL
-queries often exceeding the 100-line mark and require incorporating a diverse set of database operation from transformation
-to analytics. \citeauthor{Spider2} further highlights the gap between academic research and enterprise-level environments
-with even advanced approaches achieving only 21.3\% on \textsc{Spider2.0} compared to 91.2\% on \textsc{Spider 1.0}
-\citep{Spider, Spider2}.
-
-\subsection{Research Gaps}
-
-\subsubsection{Deployment Gaps}
+\import{theoretical-foundations}{index}
 
@@
 
-\section{Theoretical Foundations}
+\import{design}{index}
 
-\subsection{Problem Decomposition}
-
-\subsection{Requirements}
-
-% Conceptual Design
-\subsection{System Design}
-
-\subsubsection{Architecture Design}
-
-\subsection{Technical Implementation Strategies}
+% 1/3 Implementation und Evaluation
+\import{implementation}{index}
 
@@
 
-% Implementation
-\section{Implementation}
-
-\subsection{Development Environment and Tools}
-
-\subsection{Integration of the Model}
-
-\subsection{Development of the PostgreSQL Extension}
-
-\subsection{Optimization}
+\import{evaluation}{index}
 
@@
 
-% Evaluation
-\section{Evaluation}
-
-% https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf
-
-\subsection{Test Environment and Methodology}
-
-\subsection{Performance Tests}
-\subsubsection{Latency}
-\subsubsection{Throughput}
-\subsubsection{Scalability}
-
-\subsection{Use Cases}
-\subsubsection{Natural Language Queries}
-\subsubsection{Text Generation Within the Database}
-\subsubsection{Semantic Search and Text Classification}
-
-\subsection{Ablation Study}
-
-\subsection{Comparison with Alternative Approaches}
+\import{discussion}{index}
 
@@
 
-% Discussion
-\section{Discussion}
-
-\subsection{Interpretation of Results}
-\subsection{Limitations of the Implementation}
-\subsection{Ethical and Data Privacy Considerations}
-\subsection{Potential Future Developments}
+\import{conclusion}{index}
 
@@
 
-% Summary and Outlook
-\section{Summary and Outlook}
-
-\subsection{Summary of Results}
-\subsection{Addressing the Research Questions}
-\subsection{Outlook for Future Research and Development}
-
-\newpage
-
-\newpage
-
-\appendix
-\section*{Appendix}
-\subsection*{Installation Guide}
-\subsection*{API Documentation}
-\subsection*{Code Examples}
-\subsection*{Test Data and Results}
-
-\bibliographystyle{apacite}
-\bibliography{references}
+\import{appendices}{index}
 

### 2025-10-02 thesis/design/definitions.tex
Restructure the design section

--- 2025-10-01 thesis/chapters/04-design/index.tex
@@
-\newpage
+\subsection{Function Definitions}\label{design:function-definitions}
 
-\section{System Design}
-
-This chapter describes the design of \textsc{Natural}, the proposed NL2SQL system, 
-addressing limitations and research gaps identified in the literature review.
-The design follows a modular architecture that leverages both traditional computer
-science algorithms and machine learning models to achieve robust, portable and
-extensible natural language to SQL translation capabilities.
-
-\subsection{System Architecture Overview}
-
-The core of the proposed design is the query function $nq$ which utilizes
-example selection ($\sigma$), query preprocessing, schema subsetting ($\phi$),
-model inference ($\pi$), self-correction ($\rho$) and majority voting ($\nu$)
-in order to convert a natural language query $q$ into a corresponding SQL query $\omega$.
-
-% NOTE Draw diagram of nq
-<pipeline-diagram>
-%
-
-This design expands on concepts from DAIL-SQL \citep{DAIL}, X, Y and Z \textcolor{red}{look up}.
-The most notable characteristics of \textsc{Natural} lie in it's capabilities for continuous
-learning, example selection and portability which allow the system to iteratively
-adapt to unseen database schemas.
-
-\subsection{Definitions}
-
-The following sets are subsequently used to describe the systems capabilities:
+The following sets are subsequently used to introduce the functions
 
@@
 
-\subsection{Query Function – $nq$}
+\subsubsection{Query Function – $nq$}
 
@@
 
-\subsection{Selection Function – $\sigma$}
+\subsubsection{Selection Function – $\sigma$}\label{design:selection-function}
 
@@
 $$
-\sigma: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \mathcal{V} \rightarrow \mathcal{C}'
+\sigma: \mathcal{Q} \times \Omega_{\mathcal{S}} \times \mathcal{S} \times \mathcal{V} \times \mathcal{D} \rightarrow \mathcal{C}'
 $$
@@
 $$
-\sigma(q, s, e, v) = \{(q_1, s_1, \omega_1, d_k), ..., (q_k, s_k, \omega_k, d_k)\}
+\sigma(q, z, s, v, d) = \{(q_1, s_1, \omega_1, d_1), \ldots, (q_k, s_k, \omega_k, d_k)\}
 $$
@@
 
-where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$,
-and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$ tuples of
-natural language queries, schemas, corresponding SQL queries and their combined \textit{distance}.
+where $q \in \mathcal{Q}$ is the input natural language query, $z \in \Omega_{\mathcal{S}}$ is the zero-shot inference result, $s \in \mathcal{S}$ is the database schema, $v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index, and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$ tuples of natural language queries, schemas, corresponding SQL queries and their combined similarity distances.
 
@@
 that are semantically closest to the input query $q$, considering both masked natural
-language representations and schema distance (\textcolor{red}{Described in section XYZ}).
+language representations and schema distance as described in Section~\ref{design:sampling}.
 This approach ensures that examples are weighted by the following properties:
@@
 \begin{enumerate}
-    \item \textbf{Question similarity} – 
-    \item \textbf{Query similarity} –
-    \item \textbf{Database similarity} –
+    \item \textbf{Question similarity} – Semantic similarity between the masked input query and historically observed natural language queries in the embedding space $\mathcal{V}$, measured using cosine distance between their respective embeddings
+    \item \textbf{Query similarity} – Structural similarity between the zero-shot inference result $z$ and candidate SQL queries using masked SQL representations, enabling pattern recognition across different database domains
+    \item \textbf{Database similarity} – Schema compatibility measured through the distance index $\mathcal{D}$, ensuring selected examples operate on analogous database structures with similar table relationships and column types
 \end{enumerate}
 
-\subsection{Subsetting Function – $\phi$}
+\subsubsection{Subsetting Function – $\phi$}\label{design:subsetting-function}
 
@@
 
-\subsection{Projection Function – $\pi$}
+\subsubsection{Projection Function – $\pi$}\label{design:projection-function}
 
@@
 
-\subsection{Refinement Function – $\rho$}
+\subsubsection{Refinement Function – $\rho$}\label{design:refinement-function}
 
@@
 
-\subsection{Voting Function – $\nu$}
+\subsubsection{Voting Function – $\nu$}\label{design:voting-function}
 
@@
 complexity or execution performance to determine the final selection.
-
-\subsection{Masked Embedding Space – $\mathcal{V}$}
-
-Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes an
-embedding space to both traverse the masked natural language query $q'$
-and the respective masked SQL query $\omega'$. This allows $\sigma$ to use
-cosine distance to search through historically observed question answer pairs;
-eg, from previous user interactions or prevalent training datasets like
-\textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.
-
-\subsubsection{Natural Language Masking}
-
-The natural language masking process replaces all words which are not deemed
-\textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
-in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
-Choosing a sound set $\matchal{W}$ is described in chapter 5.
-
-This process enables the embedding space to capture semantic patterns independent
-of specific database artifacts and domain terminology, improving the generalizability
-of example selection across different schemas.
-
-This abstraction allows the system to recognize that queries like:
-
-\begin{verbatim}
-Show all customers with the firstname John
-List all products with the name iPhone
-\end{verbatim}
-
-in fact share similar semantic structures, even though they operate on different
-entities. Given $\matchal{W} = \{\text{List}, \text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$
-the masking would result in:
-
-\begin{verbatim}
-Show all <mask> with the <mask> <mask>
-List all <mask> with the name <mask>
-\end{verbatim}
-
-Embeddings can now be computed for the masked versions of the natural language queries.
-
-\subsubsection{Query Masking}
-
-Query masking applies similar abstraction principles to the SQL queries themselves.
-This enables the system to identify SQL patterns and structures rather than
-memorizing specific database schemas.
-
-The SQL masking process transforms all identifiers to \texttt{<mask>} and all literals
-to \texttt{<value>}. For example, the query:
-
-\begin{verbatim}
-SELECT name FROM customers WHERE age > 25
-\end{verbatim}
-
-gets masked to:
-
-\begin{verbatim}
-SELECT <mask> FROM <mask> WHERE <mask> > <value>
-\end{verbatim}
-
-This abstraction enables pattern recognition across different domains while
-preserving the essential logical structure of the original SQL statements.
-
-\subsection{Algorithm Design}
-
-\begin{algorithm}
-
-\caption{The $nq$ function}\label{alg:cap}
-\begin{algorithmic}[1]
-\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
-\Require $k \in \{ n \in N\ |\ 2n + 1 \}, k \geq 1$
-\State $S \gets nsel(q,s,e,v)$             \Comment{Example selection is performed}
-\State $C \gets \emptyset$                 \Comment{The set of query candidates}
-\While{$|C| < k$}                          \Comment{While not enough candidates exist}
-    \State $sub \gets nsub(C, s)$          \Comment{Identify relevant schema subset}
-    \State $nc \gets nprompt(S, q, sub)$   \Comment{Prompt model for next candidate}
-
-    \If{$e(nc)$}                           \Comment{Evaluate the next candidate}
-        \State $c \gets \{nc\} \cup c$     \Comment{Extend the candidates}
-    \EndIf
-
-    \State $ref \gets nref(q, sub, e, v)$  \Comment{Refine via self correction}
-
-    \If{$e(ref)$}                          \Comment{Evaluate the refined candidate}
-        \State $c \gets \{ref\} \cup c$    \Comment{Extend the candidates}
-    \EndIf
-\EndWhile
-\State \Return $nvote(C, e)$               \Comment{Return most prevalent result}
-\end{algorithmic}
-
-\end{algorithm}

### 2025-10-02 thesis/design/index.tex
Better structure
Fix control sequence problem
Leave notes for me!:)
Restructure the design section

--- 2025-10-01 thesis/chapters/04-design/index.tex
@@
-\newpage
-
 \section{System Design}
@@
 
-\subsection{System Architecture Overview}
+\subsection{Overview}
 
@@
 
-% NOTE Draw diagram of nq
-<pipeline-diagram>
-%
+% TODO: Revisit
+% TODO: Draw diagram of nq, this is an itermediate placeholder
+\begin{verbatim}
+Query -> Zero-Shot -> Selection -> Generation [ Subsetting -> Few-Shot -> Refinement ] -> Voting
+\end{verbatim}
+% TODO: Reference figure correctly
+
+\begin{algorithm}
+
+\caption{The $nq$ function}\label{alg:cap}
+
+\begin{algorithmic}[1]
+\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
+\Require $k \in \{ n \in N\ |\ 2n + 1 \}, k \geq 1$
+\State $z \gets \pi(\emptyset,q, s)$       \Comment{Zero shot inference}
+\State $S \gets \sigma(q,z,s,v,d)$         \Comment{Example selection is performed}
+\State $C \gets \emptyset$                 \Comment{The set of query candidates}
+\While{$|C| < k$}                          \Comment{While not enough candidates exist}
+    \State $s' \gets \phi(C, s)$           \Comment{Identify relevant schema subset}
+    \State $c \gets \pi(S, q, s')$         \Comment{Prompt model for next candidate}
+
+    \If{$e(nc)$}                           \Comment{Evaluate the next candidate}
+        \State $C \gets \{c\} \cup C$      \Comment{Extend the candidates}
+    \EndIf
+
+    \State $c' \gets \rho(q, s', e, v)$  \Comment{Refine via self correction}
+
+    \If{$e(c')$}                          \Comment{Evaluate the refined candidate}
+        \State $C \gets \{c'\} \cup C$    \Comment{Extend the candidates}
+    \EndIf
+\EndWhile
+\State \Return $\nu(C, e)$               \Comment{Return most prevalent result}
+\end{algorithmic}
+
+\end{algorithm}
 
@@
 
-\subsection{Definitions}
+Prior to using $nq$ sample and database indexing must be done to supply a relevant 
+embedding space $\mathcal{V}$ and distance index $\mathcal{D}$.
 
-The following sets are subsequently used to describe the systems capabilities:
-
-\begin{enumerate}
-    \item $\mathcal{Q}$ – The set of all possible natural language queries
-    \item $\mathcal{S}$ – The set of all possible database schemas
-    \item $\mathcal{E}$ – The set of execution functions for SQL validation
-    \item $\mathcal{V}$ – The embedding space of historically observed samples
-    \item $\mathcal{D}$ – The space of distance indices over historically observed databases
-    \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$
-    \item $\mathcal{Q}_{\mathcal{S}}$ – The set of all valid queries over the database schema $S$.
-\end{enumerate}
-
-\subsection{Query Function – $nq$}
-
-\vspace{0.5em}
-
-$$
-nq: \mathcal{Q} \times \mathcal{S} \times E \times \mathcal{V} \times \mathcal{D} \rightarrow \Omega_{\mathcal{S}}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-nq(q, s, e, v, d) = \omega
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$, and $\omega \in \Omega_{\mathcal{S}}$.
-
-\subsection{Selection Function – $\sigma$}
-
-The selection function $\sigma$ retrieves the most relevant examples from the
-historical embedding space $\mathcal{V}$ based on semantic similarity to the
-input query. This function implements the core example selection mechanism 
-that subsequently enables few-shot learning for SQL generation.
-
-$$
-\sigma: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \mathcal{V} \rightarrow \mathcal{C}'
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\sigma(q, s, e, v) = \{(q_1, s_1, \omega_1, d_k), ..., (q_k, s_k, \omega_k, d_k)\}
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$,
-and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$ tuples of
-natural language queries, schemas, corresponding SQL queries and their combined \textit{distance}.
-
-The function utilizes cosine similarity in the embedding space to identify examples
-that are semantically closest to the input query $q$, considering both masked natural
-language representations and schema distance (\textcolor{red}{Described in section XYZ}).
-This approach ensures that examples are weighted by the following properties:
-
-\begin{enumerate}
-    \item \textbf{Question similarity} – 
-    \item \textbf{Query similarity} –
-    \item \textbf{Database similarity} –
-\end{enumerate}
-
-\subsection{Subsetting Function – $\phi$}
-
-The subsetting function $\phi$ intelligently reduces the database schema to only include
-tables, columns, and relationships that are relevant to the current query context.
-This schema pruning mechanism reduces the complexity of the SQL generation task and
-improves model performance by focusing on pertinent schema elements.
-
-$$
-\phi: \mathcal{C}' \times \mathcal{S} \rightarrow \mathcal{S}'
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\phi(c, s) = s'
-$$
-
-\vspace{0.5em}
-
-where $c \in \mathcal{C}'$ is a candidate set containing selected examples, $s \in \mathcal{S}$
-is the full database schema, and $s' \subseteq s$ is the reduced schema subset containing
-only relevant tables and columns.
-
-The function analyzes the selected examples in $c$ to identify which schema elements
-(tables, columns, foreign key relationships) are commonly referenced in similar queries.
-This analysis enables the system to steer the model's attention on the most relevant
-portions of potentially large and complex database schemas, thereby improving both accuracy
-and computational efficiency.
-
-\subsection{Projection Function – $\pi$}
-
-The projection function $\pi$ represents the core translation mechanism that converts
-natural language queries into SQL statements using the selected examples and reduced schema.
-This function encapsulates the language model inference process that generates candidate
-SQL queries based on the provided context.
-
-$$
-\pi: \mathcal{Q} \times \mathcal{S}' \times \mathcal{C}' \rightarrow \Omega_{\mathcal{S}'}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\pi(q, s', c) = \omega
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$ is the input natural language query, $s' \in \mathcal{S}'$ is
-the reduced schema subset, $c \in \mathcal{C}'$ is the set of selected examples,
-and $\omega \in \Omega_{\mathcal{S}'}$ is the generated SQL query.
-
-The function operates by constructing a prompt that combines the natural language
-query, the relevant schema information, and the selected examples in a format
-optimized for large language model inference. The model then generates a SQL query
-that attempts to capture the semantic intent of the natural language input while
-adhering to the constraints imposed by the database schema.
-
-\subsection{Refinement Function – $\rho$}
-
-The refinement function $\rho$ implements a self-correction mechanism that iteratively
-improves generated SQL queries by identifying and correcting syntax errors,
-semantic inconsistencies, and execution failures. This function enables the system
-to learn from its mistakes and produce higher-quality outputs through automated
-feedback loops.
-
-$$
-\rho: \mathcal{Q} \times \mathcal{S}' \times \mathcal{E} \times \Omega_{\mathcal{S}'} \rightarrow \Omega_{\mathcal{S}'}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\rho(q, s', e, \omega_{prev}) = \omega_{refined}
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$ is the original natural language query, $s' \in \mathcal{S}'$
-is the relevant schema subset, $e \in \mathcal{E}$ is the execution environment for
-validation, $\omega_{prev} \in \Omega_{\mathcal{S}'}$ is the previously generated SQL
-query, and $\omega_{refined} \in \Omega_{\mathcal{S}'}$ is the improved SQL query.
-
-The function operates by executing the candidate query against the database, analyzing
-any errors or unexpected results, and then prompting the language model to generate
-an improved version based on the identified issues. This iterative refinement
-process continues until a valid, executable query is produced or a maximum number
-of refinement attempts is reached.
-
-\subsection{Voting Function – $\nu$}
-
-The voting function $\nu$ implements a consensus mechanism that selects the most
-reliable SQL query from multiple candidate solutions generated through the pipeline.
-This function enhances robustness by leveraging the wisdom of multiple generation
-attempts to identify the most likely correct answer.
-
-$$
-\nu: \mathcal{C} \times \mathcal{E} \rightarrow \Omega_{\mathcal{S}'}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\nu(C, e) = \omega_{consensus}
-$$
-
-where $C \in \mathcal{C}$ is a set of candidate SQL queries, $e \in \mathcal{E}$ is
-the execution environment for validation, and $\omega_{consensus} \in \Omega_{\mathcal{S}'}$
-is the selected consensus query.
-
-\vspace{0.5em}
-
-The voting function $\nu$ implements a majority voting algorithm similar to that
-described by OmniSQL \citep{OmniSQL}, where the result that appears most frequently
-across multiple generation attempts is deemed to be the most likely correct answer.
-The function first validates all candidates for syntactic correctness and executability,
-then applies frequency-based selection among the valid candidates. In cases where
-no clear majority exists, the function may apply additional heuristics such as query
-complexity or execution performance to determine the final selection.
-
-\subsection{Masked Embedding Space – $\mathcal{V}$}
-
-Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes an
-embedding space to both traverse the masked natural language query $q'$
-and the respective masked SQL query $\omega'$. This allows $\sigma$ to use
-cosine distance to search through historically observed question answer pairs;
-eg, from previous user interactions or prevalent training datasets like
-\textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.
-
-\subsubsection{Natural Language Masking}
-
-The natural language masking process replaces all words which are not deemed
-\textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
-in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
-Choosing a sound set $\matchal{W}$ is described in chapter 5.
-
-This process enables the embedding space to capture semantic patterns independent
-of specific database artifacts and domain terminology, improving the generalizability
-of example selection across different schemas.
-
-This abstraction allows the system to recognize that queries like:
-
-\begin{verbatim}
-Show all customers with the firstname John
-List all products with the name iPhone
-\end{verbatim}
-
-in fact share similar semantic structures, even though they operate on different
-entities. Given $\matchal{W} = \{\text{List}, \text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$
-the masking would result in:
-
-\begin{verbatim}
-Show all <mask> with the <mask> <mask>
-List all <mask> with the name <mask>
-\end{verbatim}
-
-Embeddings can now be computed for the masked versions of the natural language queries.
-
-\subsubsection{Query Masking}
-
-Query masking applies similar abstraction principles to the SQL queries themselves.
-This enables the system to identify SQL patterns and structures rather than
-memorizing specific database schemas.
-
-The SQL masking process transforms all identifiers to \texttt{<mask>} and all literals
-to \texttt{<value>}. For example, the query:
-
-\begin{verbatim}
-SELECT name FROM customers WHERE age > 25
-\end{verbatim}
-
-gets masked to:
-
-\begin{verbatim}
-SELECT <mask> FROM <mask> WHERE <mask> > <value>
-\end{verbatim}
-
-This abstraction enables pattern recognition across different domains while
-preserving the essential logical structure of the original SQL statements.
-
-\subsection{Algorithm Design}
-
-\begin{algorithm}
-
-\caption{The $nq$ function}\label{alg:cap}
-\begin{algorithmic}[1]
-\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
-\Require $k \in \{ n \in N\ |\ 2n + 1 \}, k \geq 1$
-\State $S \gets nsel(q,s,e,v)$             \Comment{Example selection is performed}
-\State $C \gets \emptyset$                 \Comment{The set of query candidates}
-\While{$|C| < k$}                          \Comment{While not enough candidates exist}
-    \State $sub \gets nsub(C, s)$          \Comment{Identify relevant schema subset}
-    \State $nc \gets nprompt(S, q, sub)$   \Comment{Prompt model for next candidate}
-
-    \If{$e(nc)$}                           \Comment{Evaluate the next candidate}
-        \State $c \gets \{nc\} \cup c$     \Comment{Extend the candidates}
-    \EndIf
-
-    \State $ref \gets nref(q, sub, e, v)$  \Comment{Refine via self correction}
-
-    \If{$e(ref)$}                          \Comment{Evaluate the refined candidate}
-        \State $c \gets \{ref\} \cup c$    \Comment{Extend the candidates}
-    \EndIf
-\EndWhile
-\State \Return $nvote(C, e)$               \Comment{Return most prevalent result}
-\end{algorithmic}
-
-\end{algorithm}
+\subimport{}{sampling}
+\subimport{}{definitions}

### 2025-10-02 thesis/design/sampling.tex
Restructure the design section

--- 2025-10-01 thesis/chapters/04-design/index.tex
@@
-\newpage
-
-\section{System Design}
-
-This chapter describes the design of \textsc{Natural}, the proposed NL2SQL system, 
-addressing limitations and research gaps identified in the literature review.
-The design follows a modular architecture that leverages both traditional computer
-science algorithms and machine learning models to achieve robust, portable and
-extensible natural language to SQL translation capabilities.
-
-\subsection{System Architecture Overview}
-
-The core of the proposed design is the query function $nq$ which utilizes
-example selection ($\sigma$), query preprocessing, schema subsetting ($\phi$),
-model inference ($\pi$), self-correction ($\rho$) and majority voting ($\nu$)
-in order to convert a natural language query $q$ into a corresponding SQL query $\omega$.
-
-% NOTE Draw diagram of nq
-<pipeline-diagram>
-%
-
-This design expands on concepts from DAIL-SQL \citep{DAIL}, X, Y and Z \textcolor{red}{look up}.
-The most notable characteristics of \textsc{Natural} lie in it's capabilities for continuous
-learning, example selection and portability which allow the system to iteratively
-adapt to unseen database schemas.
-
-\subsection{Definitions}
-
-The following sets are subsequently used to describe the systems capabilities:
-
-\begin{enumerate}
-    \item $\mathcal{Q}$ – The set of all possible natural language queries
-    \item $\mathcal{S}$ – The set of all possible database schemas
-    \item $\mathcal{E}$ – The set of execution functions for SQL validation
-    \item $\mathcal{V}$ – The embedding space of historically observed samples
-    \item $\mathcal{D}$ – The space of distance indices over historically observed databases
-    \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$
-    \item $\mathcal{Q}_{\mathcal{S}}$ – The set of all valid queries over the database schema $S$.
-\end{enumerate}
-
-\subsection{Query Function – $nq$}
-
-\vspace{0.5em}
-
-$$
-nq: \mathcal{Q} \times \mathcal{S} \times E \times \mathcal{V} \times \mathcal{D} \rightarrow \Omega_{\mathcal{S}}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-nq(q, s, e, v, d) = \omega
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$, and $\omega \in \Omega_{\mathcal{S}}$.
-
-\subsection{Selection Function – $\sigma$}
-
-The selection function $\sigma$ retrieves the most relevant examples from the
-historical embedding space $\mathcal{V}$ based on semantic similarity to the
-input query. This function implements the core example selection mechanism 
-that subsequently enables few-shot learning for SQL generation.
-
-$$
-\sigma: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \mathcal{V} \rightarrow \mathcal{C}'
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\sigma(q, s, e, v) = \{(q_1, s_1, \omega_1, d_k), ..., (q_k, s_k, \omega_k, d_k)\}
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$,
-and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$ tuples of
-natural language queries, schemas, corresponding SQL queries and their combined \textit{distance}.
-
-The function utilizes cosine similarity in the embedding space to identify examples
-that are semantically closest to the input query $q$, considering both masked natural
-language representations and schema distance (\textcolor{red}{Described in section XYZ}).
-This approach ensures that examples are weighted by the following properties:
-
-\begin{enumerate}
-    \item \textbf{Question similarity} – 
-    \item \textbf{Query similarity} –
-    \item \textbf{Database similarity} –
-\end{enumerate}
-
-\subsection{Subsetting Function – $\phi$}
-
-The subsetting function $\phi$ intelligently reduces the database schema to only include
-tables, columns, and relationships that are relevant to the current query context.
-This schema pruning mechanism reduces the complexity of the SQL generation task and
-improves model performance by focusing on pertinent schema elements.
-
-$$
-\phi: \mathcal{C}' \times \mathcal{S} \rightarrow \mathcal{S}'
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\phi(c, s) = s'
-$$
-
-\vspace{0.5em}
-
-where $c \in \mathcal{C}'$ is a candidate set containing selected examples, $s \in \mathcal{S}$
-is the full database schema, and $s' \subseteq s$ is the reduced schema subset containing
-only relevant tables and columns.
-
-The function analyzes the selected examples in $c$ to identify which schema elements
-(tables, columns, foreign key relationships) are commonly referenced in similar queries.
-This analysis enables the system to steer the model's attention on the most relevant
-portions of potentially large and complex database schemas, thereby improving both accuracy
-and computational efficiency.
-
-\subsection{Projection Function – $\pi$}
-
-The projection function $\pi$ represents the core translation mechanism that converts
-natural language queries into SQL statements using the selected examples and reduced schema.
-This function encapsulates the language model inference process that generates candidate
-SQL queries based on the provided context.
-
-$$
-\pi: \mathcal{Q} \times \mathcal{S}' \times \mathcal{C}' \rightarrow \Omega_{\mathcal{S}'}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\pi(q, s', c) = \omega
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$ is the input natural language query, $s' \in \mathcal{S}'$ is
-the reduced schema subset, $c \in \mathcal{C}'$ is the set of selected examples,
-and $\omega \in \Omega_{\mathcal{S}'}$ is the generated SQL query.
-
-The function operates by constructing a prompt that combines the natural language
-query, the relevant schema information, and the selected examples in a format
-optimized for large language model inference. The model then generates a SQL query
-that attempts to capture the semantic intent of the natural language input while
-adhering to the constraints imposed by the database schema.
-
-\subsection{Refinement Function – $\rho$}
-
-The refinement function $\rho$ implements a self-correction mechanism that iteratively
-improves generated SQL queries by identifying and correcting syntax errors,
-semantic inconsistencies, and execution failures. This function enables the system
-to learn from its mistakes and produce higher-quality outputs through automated
-feedback loops.
-
-$$
-\rho: \mathcal{Q} \times \mathcal{S}' \times \mathcal{E} \times \Omega_{\mathcal{S}'} \rightarrow \Omega_{\mathcal{S}'}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\rho(q, s', e, \omega_{prev}) = \omega_{refined}
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$ is the original natural language query, $s' \in \mathcal{S}'$
-is the relevant schema subset, $e \in \mathcal{E}$ is the execution environment for
-validation, $\omega_{prev} \in \Omega_{\mathcal{S}'}$ is the previously generated SQL
-query, and $\omega_{refined} \in \Omega_{\mathcal{S}'}$ is the improved SQL query.
-
-The function operates by executing the candidate query against the database, analyzing
-any errors or unexpected results, and then prompting the language model to generate
-an improved version based on the identified issues. This iterative refinement
-process continues until a valid, executable query is produced or a maximum number
-of refinement attempts is reached.
-
-\subsection{Voting Function – $\nu$}
-
-The voting function $\nu$ implements a consensus mechanism that selects the most
-reliable SQL query from multiple candidate solutions generated through the pipeline.
-This function enhances robustness by leveraging the wisdom of multiple generation
-attempts to identify the most likely correct answer.
-
-$$
-\nu: \mathcal{C} \times \mathcal{E} \rightarrow \Omega_{\mathcal{S}'}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-\nu(C, e) = \omega_{consensus}
-$$
-
-where $C \in \mathcal{C}$ is a set of candidate SQL queries, $e \in \mathcal{E}$ is
-the execution environment for validation, and $\omega_{consensus} \in \Omega_{\mathcal{S}'}$
-is the selected consensus query.
-
-\vspace{0.5em}
-
-The voting function $\nu$ implements a majority voting algorithm similar to that
-described by OmniSQL \citep{OmniSQL}, where the result that appears most frequently
-across multiple generation attempts is deemed to be the most likely correct answer.
-The function first validates all candidates for syntactic correctness and executability,
-then applies frequency-based selection among the valid candidates. In cases where
-no clear majority exists, the function may apply additional heuristics such as query
-complexity or execution performance to determine the final selection.
-
-\subsection{Masked Embedding Space – $\mathcal{V}$}
+\subsection{Sampling}\label{design:sampling}
 
@@
 
+%Masked Embedding Space – $\mathcal{V}$
+
 \subsubsection{Natural Language Masking}
@@
 in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
-Choosing a sound set $\matchal{W}$ is described in chapter 5.
+Choosing a sound set $\mathcal{W}$ is described in chapter 5.
 
@@
 
-in fact share similar semantic structures, even though they operate on different
-entities. Given $\matchal{W} = \{\text{List}, \text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$
+in fact share similar semantic structures, even though they operate on different entities.
+Given $\mathcal{W} = \{\text{List}\, \text{Show}\, \text{all}\, \text{the}\, \text{with}\, \text{name}\}$
+
 the masking would result in:
@@
 preserving the essential logical structure of the original SQL statements.
-
-\subsection{Algorithm Design}
-
-\begin{algorithm}
-
-\caption{The $nq$ function}\label{alg:cap}
-\begin{algorithmic}[1]
-\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
-\Require $k \in \{ n \in N\ |\ 2n + 1 \}, k \geq 1$
-\State $S \gets nsel(q,s,e,v)$             \Comment{Example selection is performed}
-\State $C \gets \emptyset$                 \Comment{The set of query candidates}
-\While{$|C| < k$}                          \Comment{While not enough candidates exist}
-    \State $sub \gets nsub(C, s)$          \Comment{Identify relevant schema subset}
-    \State $nc \gets nprompt(S, q, sub)$   \Comment{Prompt model for next candidate}
-
-    \If{$e(nc)$}                           \Comment{Evaluate the next candidate}
-        \State $c \gets \{nc\} \cup c$     \Comment{Extend the candidates}
-    \EndIf
-
-    \State $ref \gets nref(q, sub, e, v)$  \Comment{Refine via self correction}
-
-    \If{$e(ref)$}                          \Comment{Evaluate the refined candidate}
-        \State $c \gets \{ref\} \cup c$    \Comment{Extend the candidates}
-    \EndIf
-\EndWhile
-\State \Return $nvote(C, e)$               \Comment{Return most prevalent result}
-\end{algorithmic}
-
-\end{algorithm}

### 2025-10-02 thesis/design/definitions.tex
Improved the definition and sampling sections

@@
     \item $\mathcal{S}$ – The set of all possible database schemas
-    \item $\mathcal{E}$ – The set of execution functions for SQL validation
-    \item $\mathcal{V}$ – The embedding space of historically observed samples
-    \item $\mathcal{D}$ – The space of distance indices over historically observed databases
+    \item $\mathcal{E}$ – The set of all possible execution functions for SQL validation
+    \item $\mathcal{V}$ – The set of all possible embedding spaces
+    \item $\mathcal{D}$ – The set of all possible distance indices over historically observed databases
     \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$
@@
 
-where $q \in \mathcal{Q}$ is the input natural language query, $z \in \Omega_{\mathcal{S}}$ is the zero-shot inference result, $s \in \mathcal{S}$ is the database schema, $v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index, and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$ tuples of natural language queries, schemas, corresponding SQL queries and their combined similarity distances.
+where $q \in \mathcal{Q}$ is the input natural language query, $z \in \Omega_{\mathcal{S}}$
+is the zero-shot inference result, $s \in \mathcal{S}$ is the database schema,
+$v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index,
+and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$
+tuples of natural language queries, schemas, corresponding SQL queries and their
+combined distance.
 
@@
 \begin{enumerate}
-    \item \textbf{Question similarity} – Semantic similarity between the masked input query and historically observed natural language queries in the embedding space $\mathcal{V}$, measured using cosine distance between their respective embeddings
-    \item \textbf{Query similarity} – Structural similarity between the zero-shot inference result $z$ and candidate SQL queries using masked SQL representations, enabling pattern recognition across different database domains
-    \item \textbf{Database similarity} – Schema compatibility measured through the distance index $\mathcal{D}$, ensuring selected examples operate on analogous database structures with similar table relationships and column types
+    \item \textbf{Question similarity} – Semantic similarity between the masked input
+        query and historically observed natural language queries in the embedding space
+        $\mathcal{V}$, measured using cosine distance between their respective embeddings.
+    \item \textbf{SQL similarity} – Structural similarity between the zero-shot inference
+        result $z$ and candidate SQL queries using masked SQL representations, enabling
+        pattern recognition across different database domains.
+    \item \textbf{Database similarity} – Schema compatibility measured through the distance
+        index $\mathcal{D}$, ensuring selected examples operate on analogous database
+        structures with similar table relationships and column types.
 \end{enumerate}
+
+The exact weighting of these can be adjusted through three constants, $w_q}$,
+$w_s}$, and $w_d}$.
 
@@
 
-The subsetting function $\phi$ intelligently reduces the database schema to only include
-tables, columns, and relationships that are relevant to the current query context.
-This schema pruning mechanism reduces the complexity of the SQL generation task and
-improves model performance by focusing on pertinent schema elements.
+The subsetting function $\phi$ reduces the database schema to only include
+tables, columns, and relationships that are relevant to the current set of candidates
+$C$. This schema pruning mechanism reduces the complexity of the SQL generation task
+and token efficiency by focusing on the most relevant schema elements.
 
@@
 portions of potentially large and complex database schemas, thereby improving both accuracy
-and computational efficiency.
+and computational efficiency and prevents exceeding limited context windows.
 
@@
 The projection function $\pi$ represents the core translation mechanism that converts
-natural language queries into SQL statements using the selected examples and reduced schema.
-This function encapsulates the language model inference process that generates candidate
-SQL queries based on the provided context.
+natural language queries into SQL statements using the selected examples and database
+schema. This function encapsulates the large language model inference that generates
+candidate SQL queries based on the provided context.
 
 $$
-\pi: \mathcal{Q} \times \mathcal{S}' \times \mathcal{C}' \rightarrow \Omega_{\mathcal{S}'}
+\pi: \mathcal{Q} \times \mathcal{S} \times \mathcal{C}' \rightarrow \Omega_{\mathcal{S}}
 $$
@@
 $$
-\pi(q, s', c) = \omega
+\pi(q, s, c) = \omega
 $$
@@
 
-where $q \in \mathcal{Q}$ is the input natural language query, $s' \in \mathcal{S}'$ is
+where $q \in \mathcal{Q}$ is the input natural language query, $s \in \mathcal{S}$ is
 the reduced schema subset, $c \in \mathcal{C}'$ is the set of selected examples,
-and $\omega \in \Omega_{\mathcal{S}'}$ is the generated SQL query.
+and $\omega \in \Omega_{\mathcal{S}}$ is the generated SQL query.
 
@@
 $$
-\rho: \mathcal{Q} \times \mathcal{S}' \times \mathcal{E} \times \Omega_{\mathcal{S}'} \rightarrow \Omega_{\mathcal{S}'}
+\rho: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \Omega_{\mathcal{S}} \rightarrow \Omega_{\mathcal{S}}
 $$
@@
 $$
-\rho(q, s', e, \omega_{prev}) = \omega_{refined}
+\rho(q, s, e, \omega_{raw}) = \omega_{refined}
 $$
@@
 
-where $q \in \mathcal{Q}$ is the original natural language query, $s' \in \mathcal{S}'$
-is the relevant schema subset, $e \in \mathcal{E}$ is the execution environment for
-validation, $\omega_{prev} \in \Omega_{\mathcal{S}'}$ is the previously generated SQL
-query, and $\omega_{refined} \in \Omega_{\mathcal{S}'}$ is the improved SQL query.
+where $q \in \mathcal{Q}$ is the original natural language query, $s \in \mathcal{S}$
+is the database schema, $e \in \mathcal{E}$ is the execution function for
+validation, $\omega_{raw} \in \Omega_{\mathcal{S}}$ is the previously generated SQL
+query and $\omega_{refined} \in \Omega_{\mathcal{S}}$ is the improved SQL query.
 
@@
 reliable SQL query from multiple candidate solutions generated through the pipeline.
-This function enhances robustness by leveraging the wisdom of multiple generation
-attempts to identify the most likely correct answer.
+This function enhances robustness by leveraging the result distribution of multiple
+generation attempts to identify the most likely correct answer.
 
 $$
-\nu: \mathcal{C} \times \mathcal{E} \rightarrow \Omega_{\mathcal{S}'}
+\nu: \mathcal{C} \times \mathcal{E} \rightarrow \Omega_{\mathcal{S}}
 $$
@@
 where $C \in \mathcal{C}$ is a set of candidate SQL queries, $e \in \mathcal{E}$ is
-the execution environment for validation, and $\omega_{consensus} \in \Omega_{\mathcal{S}'}$
+the execution function for validation, and $\omega_{consensus} \in \Omega_{\mathcal{S}}$
 is the selected consensus query.
@@
 across multiple generation attempts is deemed to be the most likely correct answer.
-The function first validates all candidates for syntactic correctness and executability,
-then applies frequency-based selection among the valid candidates. In cases where
+The function applies frequency-based selection among the valid candidates. In cases where
 no clear majority exists, the function may apply additional heuristics such as query
-complexity or execution performance to determine the final selection.
+complexity or execution performance to determine the final proposed query candidate $w_{consensus}$.

### 2025-10-02 thesis/design/index.tex
Improved the definition and sampling sections

@@
 
-\caption{The $nq$ function}\label{alg:cap}
+\caption{The $nq$ function}\label{algorithms:nq}
 
@@
 
-Prior to using $nq$ sample and database indexing must be done to supply a relevant 
-embedding space $\mathcal{V}$ and distance index $\mathcal{D}$.
+Prior to using $nq$ to answer natural language questions, historical datasets must be sampled
+such that an embedding space $\mathcal{V}$ and distance index $\mathcal{D}$ can be supplied
+to $nq$.
 
 \subimport{}{sampling}
+
 \subimport{}{definitions}

### 2025-10-02 thesis/design/sampling.tex
Improved the definition and sampling sections

@@
 
-%Masked Embedding Space – $\mathcal{V}$
-
 \subsubsection{Natural Language Masking}
@@
 in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
-Choosing a sound set $\mathcal{W}$ is described in chapter 5.
+% TODO: Fix reference
+Choosing a sound set $\mathcal{W}$ is described in section~\ref{implementation:whitelisting}.
 
-This process enables the embedding space to capture semantic patterns independent
+This process enables the embedding space to capture semantic patterns independently
 of specific database artifacts and domain terminology, improving the generalizability
-of example selection across different schemas.
+across different schemas and domains.
 
-This abstraction allows the system to recognize that queries like:
+This masking allows \textsc{Natural} to recognize that queries like:
 
@@
 in fact share similar semantic structures, even though they operate on different entities.
-Given $\mathcal{W} = \{\text{List}\, \text{Show}\, \text{all}\, \text{the}\, \text{with}\, \text{name}\}$
+Given $\mathcal{W} = \{\text{List}, \text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$
 
@@
 
-Query masking applies similar abstraction principles to the SQL queries themselves.
-This enables the system to identify SQL patterns and structures rather than
-memorizing specific database schemas.
+Query masking applies similar abstraction principles to the corresponding SQL queries.
+This enables \textsc{Natural} to identify SQL patterns and structures.
 
-The SQL masking process transforms all identifiers to \texttt{<mask>} and all literals
-to \texttt{<value>}. For example, the query:
+The SQL masking process transforms all identifiers to \texttt{<mask>} and all
+literals to \texttt{<value>}. Thus, the query:
 
@@
 This abstraction enables pattern recognition across different domains while
-preserving the essential logical structure of the original SQL statements.
+preserving the essential logical structure of the original SQL statements,
+thus focusing on the kind of query, rather than the specific instance.
+
+\subsubsection{Sample Embedding Space – $\mathcal{V}$}
+
+\subsubsection{Schema Distance Index – $\mathcal{D}$}
+
+The schema distance index is a set of id and distance tuples $\{(s_1, d_1), ..., (s_i,d_i)\}$
+where $s$ is the database schema name and $d$ is the distance to the
+\textit{current} schema. Thus a schema distance index has to be computed for each database schema
+\textsc{Natural} is running on.

### 2025-10-02 thesis/design/definitions.tex
Further improve the design section

@@
 
-The following sets are subsequently used to introduce the functions
+The following sets are subsequently used to introduce the functions $nq$, $\sigma$, $\phi$, $\pi$, $\rho$ and $\nu$.
 
 \begin{enumerate}
-    \item $\mathcal{Q}$ – The set of all possible natural language queries
-    \item $\mathcal{S}$ – The set of all possible database schemas
-    \item $\mathcal{E}$ – The set of all possible execution functions for SQL validation
-    \item $\mathcal{V}$ – The set of all possible embedding spaces
-    \item $\mathcal{D}$ – The set of all possible distance indices over historically observed databases
-    \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$
+    \item $\mathcal{Q}$ – The set of all possible natural language queries.
+    \item $\mathcal{S}$ – The set of all possible database schemas.
+    \item $\mathcal{E}$ – The set of all possible execution functions for SQL validation.
+    \item $\mathcal{V}$ – The set of all possible embedding spaces.
+    \item $\mathcal{D}$ – The set of all possible distance indices over historically observed databases.
+    \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$.
     \item $\mathcal{Q}_{\mathcal{S}}$ – The set of all valid queries over the database schema $S$.
@@
 
-\vspace{0.5em}
+We define the query function $nq$ is collapsing a natural language question $q$,
+the database schema $s$ and a indices over historically observed examples and
+databases $v$ and $d$ into one SQL query candidate $\omega$.
 
 $$
-nq: \mathcal{Q} \times \mathcal{S} \times E \times \mathcal{V} \times \mathcal{D} \rightarrow \Omega_{\mathcal{S}}
+nq: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \mathcal{V} \times \mathcal{D} \rightarrow \Omega_{\mathcal{S}}
 $$
@@
 
-where $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$, and $\omega \in \Omega_{\mathcal{S}}$.
+where $q \in \mathcal{Q}$ is the input natural language query, $s \in \mathcal{S}$
+is the database schema, $e \in \mathcal{E}$ is the execution function for validation,
+$v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index,
+and $\omega \in \Omega_{\mathcal{S}}$ is the generated SQL query. The function is comprised
+out of the subsequently defined function and follows the algorithm described in Algorithm~\ref{algorithms:nq}.
 
@@
 where $q \in \mathcal{Q}$ is the input natural language query, $s \in \mathcal{S}$ is
-the reduced schema subset, $c \in \mathcal{C}'$ is the set of selected examples,
+the database schema, $c \in \mathcal{C}'$ is the set of selected examples,
 and $\omega \in \Omega_{\mathcal{S}}$ is the generated SQL query.

### 2025-10-03 thesis/design/indexing.tex
Split and finish the sampling section

\subsection{Schema Indexing}

To choose the most relevant subset of samples from $\mathcal{S}$ for $k$-shot learning,
it is important that the SQL queries we supply are written for structurally similar SQL schemas
in order to minimize the structural difference between the supplied samples $\mathcal{S}_k$ and
the ground truth query $\omega_{ground}$ for a given natural language question.

For example $\omega_{ground}$ the question ``Give me all contacts for the user with the id 10''
might look very different depending on the database schema at hand, thus only taking into
account the similarity of the natural language question will yield inferior results.

\vspace{1em}

\begin{figure}[ht]
\hfill
\begin{minipage}[b]{0.45\linewidth}
\centering
\begin{verbatim}
CREATE TABLE users (
  id TEXT PRiMARY KEY
);

CREATE TABLE contacts (
  user_id TEXT NOT NULL,
  name TEXT NOT NULL,
  FOREIGN KEY (user_id) ON users(id),
  PRIMARY KEY (user_id, name)
);
\end{verbatim}
\caption{Normalized schema}
\end{minipage}
\hfill
\begin{minipage}[b]{0.35\linewidth}
\centering
\begin{verbatim}
CREATE TABLE users (
  id TEXT PRIMARY KEY,
  contacts TEXT[]
);
\end{verbatim}
\caption{Denormalized schema}
\end{minipage}
\hfill
\end{figure}

\vspace{1em}

% TODO: Finish this section on WWL graph kernels and distance calculation..

%is a set of id and distance tuples $\{(s_1, d_1), ..., (s_i,d_i)\}$
%where $s$ is the database schema name and $d$ is the distance to the
%\textit{current} schema. Thus a schema distance index has to be computed for each database schema
%\textsc{Natural} is running on.

%Thus.. $\mathcal{D}$

### 2025-10-03 thesis/design/sampling.tex
Split and finish the sampling section

@@
 of specific database artifacts and domain terminology, improving the generalizability
-across different schemas and domains.
-
-This masking allows \textsc{Natural} to recognize that queries like:
+across different schemas and domains. Thus given $\mathcal{W} = \{\text{List},
+\text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$ the natural language
+questions:
 
@@
 
-in fact share similar semantic structures, even though they operate on different entities.
-Given $\mathcal{W} = \{\text{List}, \text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$
-
-the masking would result in:
+Will get masked, such that the semantic words are maintained, but domain specific
+terminology is discarded:
 
@@
 
-Embeddings can now be computed for the masked versions of the natural language queries.
+Now latent space embeddings can be computed for the masked versions of the natural
+language questions.
 
@@
 This enables \textsc{Natural} to identify SQL patterns and structures.
-
 The SQL masking process transforms all identifiers to \texttt{<mask>} and all
@@
 This abstraction enables pattern recognition across different domains while
-preserving the essential logical structure of the original SQL statements,
-thus focusing on the kind of query, rather than the specific instance.
+preserving the essential logical structure of the original SQL statements.
 
-\subsubsection{Sample Embedding Space – $\mathcal{V}$}
+\subsubsection{Embedding Space}
 
-\subsubsection{Schema Distance Index – $\mathcal{D}$}
+To construct an embedding space $\mathcal{V}_{\mathcal{S}}$ over a set of samples $\mathcal{S} = \{(q_1, s_1), ..., (q_i, s_i)\}$
+where $q$ is the natural language question and $s$ is the corresponding SQL query, we need to compute
+the embeddings of $mask(q)$ and $mask(s)$ respectively through an embedding function $\iota$.
 
-The schema distance index is a set of id and distance tuples $\{(s_1, d_1), ..., (s_i,d_i)\}$
-where $s$ is the database schema name and $d$ is the distance to the
-\textit{current} schema. Thus a schema distance index has to be computed for each database schema
-\textsc{Natural} is running on.
+Thus the embedding space $\mathcal{V}_{\mathcal{S}}$ can be defined as:
+
+$$\mathcal{V}_{\mathcal{S}} = \{~(q, \iota(mask(q)), s, \iota(mask(s)))~|~(q, s) \in \mathcal{S}~\}$$

### 2025-10-05 thesis/design/index.tex
Further improve the design section
Split and finish the sampling section
Write more of the indexing section

@@
 The core of the proposed design is the query function $nq$ which utilizes
-example selection ($\sigma$), query preprocessing, schema subsetting ($\phi$),
-model inference ($\pi$), self-correction ($\rho$) and majority voting ($\nu$)
-in order to convert a natural language query $q$ into a corresponding SQL query $\omega$.
+example selection ($\sigma$), schema subsetting ($\phi$), model inference ($\pi$),
+self-correction ($\rho$) and majority voting ($\nu$) in order to convert a natural
+language query $q$ into a corresponding SQL query $\omega$.
 
@@
 Prior to using $nq$ to answer natural language questions, historical datasets must be sampled
-such that an embedding space $\mathcal{V}$ and distance index $\mathcal{D}$ can be supplied
-to $nq$.
+and indexed such that an embedding space $v$ and distance index $d$ can be supplied to $nq$.
 
 \subimport{}{sampling}
+\subimport{}{indexing}
 

### 2025-10-05 thesis/design/indexing.tex
Write more of the indexing section

@@
 For example $\omega_{ground}$ the question ``Give me all contacts for the user with the id 10''
-might look very different depending on the database schema at hand, thus only taking into
-account the similarity of the natural language question will yield inferior results.
-
-\vspace{1em}
+might look different depending on the database schema at hand, thus only selecting
+samples based on the similarity of the natural language question will yield inferior
+sample quality.
 
 \begin{figure}[ht]
+  \vspace{1em}
   \hfill
   \begin{minipage}[b]{0.45\linewidth}
-\centering
     \begin{verbatim}
 CREATE TABLE users (
-  id TEXT PRiMARY KEY
+  id TEXT PRIMARY KEY
 );
@@
     \caption{Normalized schema}
+    \label{figure:schema-normalized}
+  \end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.35\linewidth}
+    \begin{verbatim}
+CREATE TABLE users (
+  id TEXT PRIMARY KEY,
+  contacts TEXT[] NOT NULL
+);
+    \end{verbatim}
+    \vspace{3.75em}
+    \caption{Denormalized schema}
+    \label{figure:schema-denormalized}
+  \end{minipage}
+  \hfill
+  \vspace{1em}
+\end{figure}
+
+Given the database schemas in figures~\ref{figure:schema-normalized} and~\ref{figure:schema-denormalized}
+respective definitions of $\omega_{ground}$ would be:
+
+\begin{figure}[ht]
+  \vspace{1em}
+  \hfill
+  \begin{minipage}[b]{0.45\linewidth}
+    \begin{verbatim}
+SELECT contacts.name
+FROM users
+JOIN contacts
+  ON contacts.user_id = users.id
+WHERE users.id = 10;
+    \end{verbatim}
+    \caption{SQL JOIN selection}
+    \label{figure:sql-join-example}
   \end{minipage}
@@
     \begin{verbatim}
-CREATE TABLE users (
-  id TEXT PRIMARY KEY,
-  contacts TEXT[]
-);
+SELECT contacts
+FROM users
+WHERE id = 10;
     \end{verbatim}
-\caption{Denormalized schema}
+    \vspace{1.3em}
+    \caption{SQL Array selection}
+    \label{figure:sql-array-example}
   \end{minipage}
   \hfill
+  \vspace{1em}
 \end{figure}
 
+As shown in the figures~\ref{figure:sql-join-example} and~\ref{figure:sql-array-example}
+the structural similarity of the underlying database schema is a crucial component
+of the relevance of a example.
+
+\subsubsection{Graph Representation}
+
+In order to determine structural similarity of database schemas we propose using
+graph representation of database schemas as a data definition language (DDL)
+independent representation of the databases structure. Furthermore choosing a graph
+representation allows us to use prevalent methods for measuring graph similarity
+(eg, \textcolor{red}{Weisfeiler-Lehman graph kernels, cite}).
+
+We define the structural similarity of two databases as the toplogical distance
+of the respective \textsc{Natural}-graphs. The \textsc{Natural}-graph representation
+only models tables, columns, relationships, datatypes and constraints, omitting 
+table names, column names and domain terminology.
+
+\begin{figure}[ht]
   \vspace{1em}
+  \hfill
+  \begin{minipage}[b]{0.45\linewidth}
+    \centering
+    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
+      \node[main] (1) {$t_1$}; 
+      \node[main] (2) [below of=1] {$c_1$}; 
+      \node[main] (3) [right of=2] {$t_2$}; 
+      \node[main] (4) [below left of=3] {$c_3$}; 
+      \node[main] (5) [below right of=3] {$c_4$};
+
+      \draw (1) -- (2); 
+      \draw (2) -- (4); 
+      \draw (3) -- (4); 
+      \draw (3) -- (5); 
+      \draw (1) to [out=0, in=90, looseness=1] (3); 
+    \end{tikzpicture} 
+    \caption{Normalized graph repr.}
+    \label{figure:normalized-graph-representation}
+  \end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.45\linewidth}
+    \centering
+    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
+      \node[main] (1) {$t_1$}; 
+      \node[main] (2) [below of=1] {$c_1$}; 
+      \node[main] (3) [below right of=1] {$c_2$};
+
+      \draw (1) -- (2); 
+      \draw (1) -- (3); 
+    \end{tikzpicture} 
+    \caption{Denormalized graph repr.}
+    \label{figure:denormalized-graph-representation}
+  \end{minipage}
+  \hfill
+  \vspace{1em}
+\end{figure}
+
+Now measuring the distance between the graphs displayed in~\ref{figure:normalized-graph-representation} and~\ref{figure:denormalized-graph-representation}
 

### 2025-10-05 thesis/design/sampling.tex
Write more of the indexing section

@@
 
-Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes an
-embedding space to both traverse the masked natural language query $q'$
-and the respective masked SQL query $\omega'$. This allows $\sigma$ to use
-cosine distance to search through historically observed question answer pairs;
-eg, from previous user interactions or prevalent training datasets like
-\textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.
+Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes 
+cosine similarity in an embedding space $v$ to search for examples based on the
+masked natural language query $q'$ and the respective masked SQL query $\omega'$.
+This allows $\sigma$ to use cosine distance to search through historically observed
+question answer pairs; eg, from previous user interactions or prevalent training
+datasets like \textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.
 

### 2025-10-05 thesis/design/indexing.tex
Better design section

@@
 
-In order to determine structural similarity of database schemas we propose using
+To determine structural similarity of database schemas, we propose using
 graph representation of database schemas as a data definition language (DDL)
-independent representation of the databases structure. Furthermore choosing a graph
-representation allows us to use prevalent methods for measuring graph similarity
-(eg, \textcolor{red}{Weisfeiler-Lehman graph kernels, cite}).
+independent representation of the database structure. Choosing a graph
+representation allows us to leverage established methods for measuring graph similarity
+such as Wasserstein-Weisfeiler-Lehman graph kernels \cite{WWL}. % TODO: Add proper WWL citation
 
-We define the structural similarity of two databases as the toplogical distance
-of the respective \textsc{Natural}-graphs. The \textsc{Natural}-graph representation
-only models tables, columns, relationships, datatypes and constraints, omitting 
-table names, column names and domain terminology.
+Given a database schema $s$, the corresponding graph $G_s = (V, E, \ell, w)$ is defined as:
+
+\begin{enumerate}
+    \item $V = V_t \cup V_c$ where $V_t$ represents table nodes and $V_c$ represents column nodes
+    \item $E = E_{tc} \cup E_{fk} \cup E_{ref}$ where:
+        \begin{itemize}
+            \item $E_{tc}$ are table-column edges
+            \item $E_{fk}$ are foreign key relationship edges between tables
+            \item $E_{ref}$ are reference edges between foreign key columns
+        \end{itemize}
+    \item Each node $v \in V$ has a semantic label $\ell(v) \in \{1, 2, \ldots, 9\}$ where:
+        \begin{itemize}
+            \item $\ell(v) = 1$ for table nodes
+            \item $\ell(v) = 2$ for generic column nodes
+            \item $\ell(v) = 3$ for primary key columns
+            \item $\ell(v) = 4$ for foreign key columns
+            \item $\ell(v) = 5$ for text columns
+            \item $\ell(v) = 6$ for numeric columns
+            \item $\ell(v) = 7$ for datetime columns
+            \item $\ell(v) = 8$ for boolean columns
+            \item $\ell(v) = 9$ for other/unknown data types
+        \end{itemize}
+    \item Each edge $e \in E$ has a weight $w(e) \in \{0.5, 0.9, 1.0\}$ reflecting structural importance:
+        \begin{itemize}
+            \item $w(e) = 1.0$ for foreign key relationships (highest importance)
+            \item $w(e) = 0.9$ for column reference edges
+            \item $w(e) = 0.5$ for table-column edges
+        \end{itemize}
+\end{enumerate}
+
+We define the structural similarity of two databases as the topological distance
+of the respective graphs. The graph representation captures essential schema
+structure through semantic node labeling that prioritizes constraints over data
+types: primary key columns receive label 3 regardless of their underlying data
+type, foreign key columns receive label 4, and only then are remaining columns
+categorized by data type (text=5, numeric=6, datetime=7, boolean=8, other=9).
+This hierarchical labeling ensures that structural relationships take
+precedence over semantic content, while omitting table names, column names and
+domain terminology to achieve schema-agnostic comparison.
+
+The graph representation of the database schemas described in~\ref{figure:schema-normalized}
+and~\ref{figure:schema-denormalized} are therefore:
 
@@
     \centering
-    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
-      \node[main] (1) {$t_1$}; 
-      \node[main] (2) [below of=1] {$c_1$}; 
-      \node[main] (3) [right of=2] {$t_2$}; 
-      \node[main] (4) [below left of=3] {$c_3$}; 
-      \node[main] (5) [below right of=3] {$c_4$};
+    \begin{tikzpicture}[node distance={15mm}, table/.style = {draw,circle}, column/.style = {draw,dashed,circle}]
 
-      \draw (1) -- (2); 
-      \draw (2) -- (4); 
-      \draw (3) -- (4); 
-      \draw (3) -- (5); 
-      \draw (1) to [out=0, in=90, looseness=1] (3); 
+      \node[table] (users) at (0,0) {1}; 
+      \node[table] (contacts) at (2,0) {1};
+
+      \node[column] (users_id) [below left of=users] {3};
+
+      \node[column] (contacts_user_id) [below left of=contacts] {4};
+      \node[column] (contacts_name) [below right of=contacts] {5};
+
+      \draw[dashed] (users) -- (users_id); 
+      \draw[dashed] (contacts) -- (contacts_user_id); 
+      \draw[dashed] (contacts) -- (contacts_name);
+
+      \draw[thick] (users) to [out=10, in=170] (contacts);
+
+      \draw (users_id) to [out=10, in=170] (contacts_user_id); 
     \end{tikzpicture} 
@@
     \centering
-    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
-      \node[main] (1) {$t_1$}; 
-      \node[main] (2) [below of=1] {$c_1$}; 
-      \node[main] (3) [below right of=1] {$c_2$};
+    \begin{tikzpicture}[node distance={15mm}, table/.style = {draw,circle}, column/.style = {draw,dashed,circle}]
+      \node[table] (users) at (0,0) {1};
 
-      \draw (1) -- (2); 
-      \draw (1) -- (3); 
+      \node[column] (users_id) [below left of=users] {3}; 
+      \node[column] (users_contacts) [below right of=users] {9};
+
+      \draw[dashed] (users) -- (users_id); 
+      \draw[dashed] (users) -- (users_contacts); 
     \end{tikzpicture} 
@@
   \vspace{1em}
+
+  \begin{minipage}{\linewidth}
+    \centering
+    \vspace{0.5em}
+    \small
+    \textbf{Legend:} Solid circle = table (1), dashed circle = column (3,4,5), 
+    thick line = fk relationship ($w$=1.0), line = reference ($w$=0.9), 
+    dashed line = table-column ($w$=0.5)
+  \end{minipage}
 \end{figure}
 
-Now measuring the distance between the graphs displayed in~\ref{figure:normalized-graph-representation} and~\ref{figure:denormalized-graph-representation}
+The $\delta$ of the order of the graphs~\ref{figure:normalized-graph-representation}
+and~\ref{figure:denormalized-graph-representation}
+further highlight the structural difference between the two schemas.
 
-% TODO: Finish this section on WWL graph kernels and distance calculation..
+\subsubsection{Distance Measurement}
 
-%is a set of id and distance tuples $\{(s_1, d_1), ..., (s_i,d_i)\}$
-%where $s$ is the database schema name and $d$ is the distance to the
-%\textit{current} schema. Thus a schema distance index has to be computed for each database schema
-%\textsc{Natural} is running on.
+To measure the distance between the graphs displayed in~\ref{figure:normalized-graph-representation}
+and~\ref{figure:denormalized-graph-representation}, we employ the
+Wasserstein-Weisfeiler-Lehman (WWL) graph kernel method \cite{WWL}. % TODO: Fix WWL citation
 
-%Thus.. $\mathcal{D}$
+The WWL kernel computes structural similarity by combining discrete Weisfeiler-Lehman
+graph features with continuous optimal transport theory. The method first extracts
+WL features from the labeled graph structure, then computes the Wasserstein distance
+between the resulting feature distributions. For two \textsc{Natural}-graphs $G_1$ and $G_2$
+with semantic node labels, the distance is computed as:
+
+\begin{equation}
+% TODO: Add missing WWL distance formula 
+%d_{WWL}(G_1, G_2) = W_2(\mu_{G_1}, \mu_{G_2})
+\end{equation}
+
+where $W_2$ denotes the 2-Wasserstein distance and $\mu_{G_i}$ represents the empirical
+measure over WL features extracted from graph $G_i$. The semantic node labels
+$\ell(v) \in \{1, \ldots, 9\}$ guide the feature extraction process, ensuring that
+structurally and semantically similar schema elements contribute similarly to the
+distance calculation.
+
+For the example schemas above, the normalized schema (5 nodes, 7 edges) and
+denormalized schema (3 nodes, 2 edges) exhibit significant topological differences.
+The WWL distance captures both the reduction in graph complexity (fewer nodes and edges)
+and the loss of relational structure (elimination of foreign key relationships),
+resulting in a substantial distance value reflecting their structural dissimilarity
+despite representing equivalent logical relationships.
+
+% TODO: Complete distance calculation example with actual values
+% TODO: Add discussion of how semantic labeling affects distance calculation  
+% TODO: Clarify relationship between edge weights and WWL computation
+% TODO: Add example showing how label priority affects similarity scores
+
+The schema distance index $\mathcal{D}$ maintains precomputed distances between
+all observed database schemas, enabling efficient retrieval of structurally similar
+databases during sample selection. This index is constructed as a set of
+schema-distance tuples $\{(s_1, d_1), \ldots, (s_i, d_i)\}$ where $s_i$ represents
+a database schema and $d_i$ denotes its topological distance to the current target
+schema. The computational complexity of WL kernel calculation is $O(h \cdot |E|)$
+where $h$ is the number of iterations and $|E|$ is the number of edges, making
+it practical for real-time schema comparison in moderately sized databases.
+
+% TODO: Fix inconsistent notation: use consistent variable names (G_s vs Natural-graphs)
+% TODO: Improve transition between graph definition and distance measurement sections

### 2025-10-05 thesis/design/sampling.tex
Better design section

@@
 Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes 
-cosine similarity in an embedding space $v$ to search for examples based on the
+cosine similarity in an embedding space $\mathcal{V}$ to search for examples based on the % TODO: Fix notation consistency with definitions
 masked natural language query $q'$ and the respective masked SQL query $\omega'$.
@@
 \textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
-in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
-% TODO: Fix reference
+in a predefined whitelist $\mathcal{W}$ are determined structurally relevant. % TODO: Fix typo
+% TODO: Fix reference to implementation section
 Choosing a sound set $\mathcal{W}$ is described in section~\ref{implementation:whitelisting}.
@@
 where $q$ is the natural language question and $s$ is the corresponding SQL query, we need to compute
-the embeddings of $mask(q)$ and $mask(s)$ respectively through an embedding function $\iota$.
+the embeddings of $mask(q)$ and $mask(s)$ respectively through an embedding function $\iota$. % TODO: Specify embedding model
 
@@
 $$\mathcal{V}_{\mathcal{S}} = \{~(q, \iota(mask(q)), s, \iota(mask(s)))~|~(q, s) \in \mathcal{S}~\}$$
+
+% TODO: Explain how this relates to the selection function σ from definitions
+% TODO: Discuss potential limitations of masked embedding approach
+
+% ALIGNMENT IMPROVEMENTS NEEDED:
+% TODO: Add formal mathematical definition using enumerate style like indexing section
+% TODO: Use consistent tuple notation (q,s,ω,d) matching definitions section  
+% TODO: Add cross-references to Section~\ref{design:indexing} for schema distances
+% TODO: Include similarity scoring methodology with weights w_q, w_s, w_sql
+% TODO: Add subsection on "Distance Calculation" parallel to indexing section
+% TODO: Use consistent variable naming (mathcal{V} vs V vs v)
+% TODO: Add example showing multi-dimensional scoring like indexing graphs
+% TODO: Include legend/explanation of masking tokens like indexing legend

### 2025-10-06 thesis/design/components.tex
Rewrite the design section partially

--- 2025-10-02 thesis/design/definitions.tex
@@
-\subsection{Function Definitions}\label{design:function-definitions}
+\subsection{Functions}\label{design:components}
 
@@
 that are semantically closest to the input query $q$, considering both masked natural
-language representations and schema distance as described in Section~\ref{design:sampling}.
+language representations and schema distance as described in Section~\ref{design:embedding}.
 This approach ensures that examples are weighted by the following properties:
@@
 
-The exact weighting of these can be adjusted through three constants, $w_q}$,
-$w_s}$, and $w_d}$.
+The exact weighting of these can be adjusted through three constants, $w_q$, $w_s$, and $w_d$.
 

### 2025-10-06 thesis/design/index.tex
Rewrite the design section partially

--- 2025-10-02 thesis/design/index.tex
@@
 
-The core of the proposed design is the query function $nq$ which utilizes
-example selection ($\sigma$), query preprocessing, schema subsetting ($\phi$),
-model inference ($\pi$), self-correction ($\rho$) and majority voting ($\nu$)
-in order to convert a natural language query $q$ into a corresponding SQL query $\omega$.
+The \textsc{Natural} system implements a multi-stage pipeline that transforms natural
+language questions into executable SQL queries through example selection,
+schema subsetting, and iterative refinement. The proposed architecture consists
+of five core components:
 
-% TODO: Revisit
-% TODO: Draw diagram of nq, this is an itermediate placeholder
+\begin{enumerate}
+    \item \textbf{Example Selection} – Identifies semantically and structurally similar
+        historical examples using masked embeddings and schema distances.
+    \item \textbf{Schema Subsetting} – Reduces database schema complexity by focusing
+        on relevant tables and relationships
+    \item \textbf{Query Generation} – Few-shot learning with fine tuned models translate natural language queries to SQL
+    \item \textbf{Self-Correction} – Iteratively refines queries through execution feedback and error analysis
+    \item \textbf{Consensus Voting} – Selects the most reliable result from multiple
+        generation attempts
+\end{enumerate}
+
+The system processes queries through the following pipeline:
+
+% TODO: Make a nice diagram for this
 \begin{verbatim}
-Query -> Zero-Shot -> Selection -> Generation [ Subsetting -> Few-Shot -> Refinement ] -> Voting
+  Natural Language Query 
+→ Sketch Generation
+→ Example Selection
+→ k-times [
+    → Schema Subsetting
+    → Few-Shot Generation
+    → Self-Correction
+  ]
+→ Consensus Voting
+→ SQL Query
 \end{verbatim}
-% TODO: Reference figure correctly
 
-\begin{algorithm}
+This design builds upon concepts from DAIL-SQL \citep{DAIL} and incorporates novel
+contributions in schema-aware example selection and graph-based structural similarity.
 
-\caption{The $nq$ function}\label{alg:cap}
+\subimport{}{initialization}
 
-\begin{algorithmic}[1]
-\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
-\Require $k \in \{ n \in N\ |\ 2n + 1 \}, k \geq 1$
-\State $z \gets \pi(\emptyset,q, s)$       \Comment{Zero shot inference}
-\State $S \gets \sigma(q,z,s,v,d)$         \Comment{Example selection is performed}
-\State $C \gets \emptyset$                 \Comment{The set of query candidates}
-\While{$|C| < k$}                          \Comment{While not enough candidates exist}
-    \State $s' \gets \phi(C, s)$           \Comment{Identify relevant schema subset}
-    \State $c \gets \pi(S, q, s')$         \Comment{Prompt model for next candidate}
+\subimport{}{components}
 
-    \If{$e(nc)$}                           \Comment{Evaluate the next candidate}
-        \State $C \gets \{c\} \cup C$      \Comment{Extend the candidates}
-    \EndIf
-
-    \State $c' \gets \rho(q, s', e, v)$  \Comment{Refine via self correction}
-
-    \If{$e(c')$}                          \Comment{Evaluate the refined candidate}
-        \State $C \gets \{c'\} \cup C$    \Comment{Extend the candidates}
-    \EndIf
-\EndWhile
-\State \Return $\nu(C, e)$               \Comment{Return most prevalent result}
-\end{algorithmic}
-
-\end{algorithm}
-
-This design expands on concepts from DAIL-SQL \citep{DAIL}, X, Y and Z \textcolor{red}{look up}.
-The most notable characteristics of \textsc{Natural} lie in it's capabilities for continuous
-learning, example selection and portability which allow the system to iteratively
-adapt to unseen database schemas.
-
-Prior to using $nq$ sample and database indexing must be done to supply a relevant 
-embedding space $\mathcal{V}$ and distance index $\mathcal{D}$.
-
-\subimport{}{sampling}
-\subimport{}{definitions}
+\subimport{}{composition}

### 2025-10-06 thesis/design/initialization.tex
Rewrite the design section partially

\subsection{Initialization} % TODO: FIX THIS SECTION

Before the \textsc{Natural} system can process queries, it requires initialization
with historical data to construct a specific embedding space $v \in \mathcal{V}$ and
schema distance index $d \in \mathcal{D}$. This phase analyzes and samples existing
datasets and previous user interactions to build 
advanced example selection.

\subsubsection{Embedding}\label{design:embedding}

Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes 
cosine similarity in an embedding space $v$ to search for examples based on the
masked natural language query $q'$ and the respective masked SQL query $\omega'$.
This allows $\sigma$ to use cosine distance to search through historically observed
question answer pairs; eg, from previous user interactions or prevalent training
datasets like \textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.

We therefore embedd a set of masked samples $\mathcal{S}$ into an embedding space $v_{\mathcal{S}}$,
where $\mathcal{S} = \{(q_1, s_1), ..., (q_i, s_i)\}$, $q$ is the natural language
question and $s$ is the corresponding SQL query.

The embedding space $v_{\mathcal{S}}$ can be defined as:

$$v_{\mathcal{S}} = \{~(q, \iota(mask(q)), s, \iota(mask(s)))~|~(q, s) \in \mathcal{S}~\}$$

Where the masking function $mask$ is defined as described in the sections~\ref{design:nl-masking}
and~\ref{design:sql-masking} and the embedding function is defined as $\iota$.
% TODO: Specify embedding model

\paragraph{Natural Language Masking}\label{design:nl-masking}

Masking of natural language is the process of replacing all words which are not
deemed \textit{structurally relevant} with the \texttt{<mask>} token. All words
$w$ in a constant whitelist $\mathcal{W}$ are determined structurally relevant.
% TODO: Fix typo
Estimating a constant $\mathcal{W}$ is described in section~\ref{implementation:whitelisting}.
% TODO: Fix reference to implementation section

Working with masked language enables the system to capture semantic patterns
independently of domain terminology, improving the generalizability across
domains.

Given $\mathcal{W} = \{\text{list}, \text{show}, \text{all}, \text{the},
\text{with}\}$ the natural language
questions:

\begin{verbatim}
Show all customers with the firstname John
List all products with the name iPhone
\end{verbatim}

Will get masked into the same sentence, such that the semantic words are
maintained, but domain specific terminology is discarded:

\begin{verbatim}
Show all <mask> with the <mask> <mask>
\end{verbatim}

\paragraph{SQL Masking}\label{design:sql-masking}

Masking SQL queries refers to transforming all identifiers contained in 
a SQL query to \texttt{<mask>} and all value literals to \texttt{<value>}.

Thus, the queries:

\begin{verbatim}
SELECT * FROM customers WHERE firstname = 'John'
SELECT * FROM products WHERE name = 'iPhone'
\end{verbatim}

get masked to:

\begin{verbatim}
SELECT * FROM <mask> WHERE <mask> = <value>
\end{verbatim}

% TODO: Discuss potential limitations of masked embedding approach

% ALIGNMENT IMPROVEMENTS NEEDED:
% TODO: Add formal mathematical definition using enumerate style like indexing section
% TODO: Use consistent tuple notation (q,s,ω,d) matching definitions section  
% TODO: Add cross-references to Section~\ref{design:indexing} for schema distances
% TODO: Include similarity scoring methodology with weights w_q, w_s, w_sql
% TODO: Add subsection on "Distance Calculation" parallel to indexing section
% TODO: Use consistent variable naming (mathcal{V} vs V vs v)
% TODO: Add example showing multi-dimensional scoring like indexing graphs
% TODO: Include legend/explanation of masking tokens like indexing legend

\subsubsection{Schema Indexing}

To choose the most relevant subset of samples for few-shot learning,
it is important that the SQL queries we choose as examples are written for
structurally similar database schemas in order to minimize the structural
difference between the selected samples and the ground truth query for a given
natural language question.

For example $\omega_{ground}$ the question ``Give me all contacts for the user with the id 10''
might look different depending on the database schema at hand, thus only selecting
samples based on the similarity of the natural language question will yield inferior
sample quality.

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \begin{verbatim}
CREATE TABLE users (
  id TEXT PRIMARY KEY
);

CREATE TABLE contacts (
  user_id TEXT NOT NULL,
  name TEXT NOT NULL,
  FOREIGN KEY (user_id) 
   REFERENCES users(id),
  PRIMARY KEY (user_id, name)
);
    \end{verbatim}
    \caption{Normalized schema}
    \label{figure:schema-normalized}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.35\linewidth}
    \begin{verbatim}
CREATE TABLE users (
  id TEXT PRIMARY KEY,
  contacts TEXT[] NOT NULL
);
    \end{verbatim}
    \vspace{3.75em}
    \caption{Denormalized schema}
    \label{figure:schema-denormalized}
  \end{minipage}
  \hfill
  \vspace{1em}
\end{figure}

Given the database schemas in figures~\ref{figure:schema-normalized} and~\ref{figure:schema-denormalized}
respective definitions of $\omega_{ground}$ would be~\ref{figure:sql-join-example}
and~\ref{figure:sql-array-example}.

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \begin{verbatim}
SELECT contacts.name
FROM users
JOIN contacts
  ON contacts.user_id = users.id
WHERE users.id = 10;
    \end{verbatim}
    \caption{SQL JOIN selection}
    \label{figure:sql-join-example}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.35\linewidth}
    \centering
    \begin{verbatim}
SELECT contacts
FROM users
WHERE id = 10;
    \end{verbatim}
    \vspace{1.3em}
    \caption{SQL Array selection}
    \label{figure:sql-array-example}
  \end{minipage}
  \hfill
  \vspace{1em}
\end{figure}

As shown in the figures~\ref{figure:sql-join-example} and~\ref{figure:sql-array-example}
the structural similarity of the underlying database schema is a crucial component
of the relevance of an example.

\paragraph{Graph Representation}

To determine structural similarity of database schemas systematically, we propose
using graph representation of database schemas as a data definition language (DDL)
independent representation of the database structure. Choosing a graph
representation allows us to leverage established methods for measuring graph similarity
such as Wasserstein-Weisfeiler-Lehman graph kernels \citep{WWL}. % TODO: Add proper WWL citation

Given a database schema $s$, the corresponding graph $G_s = (V, E, \ell, w)$ is defined as:

\begin{enumerate}
    \item $V = V_t \cup V_c$ where $V_t$ represents table nodes and $V_c$ represents column nodes
    \item $E = E_{tc} \cup E_{fk} \cup E_{ref}$ where:
        \begin{itemize}
            \item $E_{tc}$ are table-column edges
            \item $E_{fk}$ are foreign key relationship edges between tables
            \item $E_{ref}$ are reference edges between foreign key columns
        \end{itemize}
    \item Each node $v \in V$ has a semantic label $\ell(v) \in \{1, 2, \ldots, 9\}$ where:
        \begin{itemize}
            \item $\ell(v) = 1$ for table nodes
            \item $\ell(v) = 2$ for generic column nodes
            \item $\ell(v) = 3$ for primary key columns
            \item $\ell(v) = 4$ for foreign key columns
            \item $\ell(v) = 5$ for text columns
            \item $\ell(v) = 6$ for numeric columns
            \item $\ell(v) = 7$ for datetime columns
            \item $\ell(v) = 8$ for boolean columns
            \item $\ell(v) = 9$ for other/unknown data types
        \end{itemize}
    \item Each edge $e \in E$ has a weight $w(e) \in \{0.5, 0.9, 1.0\}$ reflecting structural importance:
        \begin{itemize}
            \item $w(e) = 1.0$ for foreign key relationships (highest importance)
            \item $w(e) = 0.9$ for column foreign key reference edges
            \item $w(e) = 0.5$ for table-column edges
        \end{itemize}
\end{enumerate}

We define the structural similarity of two databases as the topological distance
of the respective graphs. The graph representation captures essential schema
structure through semantic node labeling that prioritizes constraints over data
types: primary key columns receive label 3 regardless of their underlying data
type, foreign key columns receive label 4, and only then are remaining columns
categorized by data type (text=5, numeric=6, datetime=7, boolean=8, other=9).
This hierarchical labeling ensures that structural relationships take
precedence over semantic content, while omitting table names, column names and
domain terminology to achieve schema-agnostic comparison.

The graph representation of the database schemas introduced in~\ref{figure:schema-normalized}
and~\ref{figure:schema-denormalized} are therefore:

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, table/.style = {draw,circle}, column/.style = {draw,dashed,circle}]

      \node[table] (users) at (0,0) {1}; 
      \node[table] (contacts) at (2,0) {1};

      \node[column] (users_id) [below left of=users] {3};

      \node[column] (contacts_user_id) [below left of=contacts] {4};
      \node[column] (contacts_name) [below right of=contacts] {5};

      \draw[dashed] (users) -- (users_id); 
      \draw[dashed] (contacts) -- (contacts_user_id); 
      \draw[dashed] (contacts) -- (contacts_name);

      \draw[thick] (users) to [out=10, in=170] (contacts);

      \draw (users_id) to [out=10, in=170] (contacts_user_id); 
    \end{tikzpicture} 
    \caption{Normalized graph repr.}
    \label{figure:normalized-graph-representation}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, table/.style = {draw,circle}, column/.style = {draw,dashed,circle}]
      \node[table] (users) at (0,0) {1};

      \node[column] (users_id) [below left of=users] {3}; 
      \node[column] (users_contacts) [below right of=users] {9};

      \draw[dashed] (users) -- (users_id); 
      \draw[dashed] (users) -- (users_contacts); 
    \end{tikzpicture} 
    \caption{Denormalized graph repr.}
    \label{figure:denormalized-graph-representation}
  \end{minipage}
  \hfill
  \vspace{1em}

  \begin{minipage}{\linewidth}
    \centering
    \vspace{0.5em}
    \small
    \textbf{Legend:} \cir{}~table~~\cir{dashed}~column~~\textbf{-}~foreign key~~--~reference~~- -~table-column
  \end{minipage}
\end{figure}

The $\delta$ of the order of the graphs~\ref{figure:normalized-graph-representation}
and~\ref{figure:denormalized-graph-representation}
further highlight the structural difference between the two schemas.

\paragraph{Distance Measurement}

To measure the distance between the graphs displayed in~\ref{figure:normalized-graph-representation}
and~\ref{figure:denormalized-graph-representation}, we employ the
Wasserstein Weisfeiler-Lehman (WWL) graph kernel method \citep{WWL}.

WWL kernels compute structural similarity by combining discrete Weisfeiler-Lehman
graph features with continuous optimal transport theory. The method first extracts
WL features from the labeled graph structure, then computes the Wasserstein distance
between the resulting feature distributions. For two graphs $G$ and
$G'$ with semantic node labels, the distance is computed as:

\begin{equation}
D^{f}_{W}(G, G') = W_1(f(G), f(G'))
\end{equation}

as introduced by \citeauthor{WWL} in \citeyear{WWL}.

For the database schemas above, the normalized schema (5 nodes, 5 edges) and
denormalized schema (3 nodes, 2 edges) show significant topological differences.
The WWL distance captures both the reduction in graph complexity (fewer nodes and edges)
and the loss of relational structure (elimination of foreign key relationships),
resulting in a substantial distance value reflecting their structural dissimilarity
despite representing equivalent logical data with $D^{f}_{W}(G_{norm}, G_{denorm}) \approx 0.78$.

% TODO: Clarify relationship between edge weights and WWL computation
% TODO: Add example showing how label priority affects similarity scores

\paragraph{Distance Index}

The distance index $d$ maintains precomputed distances between
all observed database schemas, enabling efficient retrieval of structurally similar
databases during sample selection. This index is constructed as a set of
schema-distance tuples $\{(sm_1, di_1), \ldots, (sm_i, di_i)\}$ where $sm_i$ is
a database schema name and $di_i$ denotes its WWL distance to the current schema.

% TODO: Fix inconsistent notation: use consistent variable names (G_s vs Natural-graphs)
% TODO: Improve transition between graph definition and distance measurement sections

### 2025-10-06 thesis/design/functions.tex
Polish the design section

--- 2025-10-06 thesis/design/components.tex
@@
 
-\subsubsection{Query Function – $nq$}
+\subsubsection{Query – $nq$}
 
@@
 
-\subsubsection{Selection Function – $\sigma$}\label{design:selection-function}
+\subsubsection{Example Selection – $\sigma$}\label{design:selection-function}
 
 The selection function $\sigma$ retrieves the most relevant examples from the
-historical embedding space $\mathcal{V}$ based on semantic similarity to the
+historical embedding space $v$ based on semantic similarity to the
 input query. This function implements the core example selection mechanism 
@@
         query and historically observed natural language queries in the embedding space
-        $\mathcal{V}$, measured using cosine distance between their respective embeddings.
+        $v$, measured using cosine distance between their respective embeddings.
     \item \textbf{SQL similarity} – Structural similarity between the zero-shot inference
@@
     \item \textbf{Database similarity} – Schema compatibility measured through the distance
-        index $\mathcal{D}$, ensuring selected examples operate on analogous database
+        index $d$, ensuring selected examples operate on analogous database
         structures with similar table relationships and column types.
@@
 
-\subsubsection{Subsetting Function – $\phi$}\label{design:subsetting-function}
+\subsubsection{Schema Subsetting – $\phi$}\label{design:subsetting-function}
 
@@
 
-\subsubsection{Projection Function – $\pi$}\label{design:projection-function}
+\subsubsection{Query Projection – $\pi$}\label{design:projection-function}
 
@@
 natural language queries into SQL statements using the selected examples and database
-schema. This function encapsulates the large language model inference that generates
+schema. This function encapsulates the inference using LLMs that generates
 candidate SQL queries based on the provided context.
@@
 
-\subsubsection{Refinement Function – $\rho$}\label{design:refinement-function}
+\subsubsection{Self Refinement – $\rho$}\label{design:refinement-function}
 
@@
 
-\subsubsection{Voting Function – $\nu$}\label{design:voting-function}
+\subsubsection{Voting – $\nu$}\label{design:voting-function}
 

### 2025-10-06 thesis/design/index.tex
Polish the design section

@@
 \begin{enumerate}
-    \item \textbf{Example Selection} – Identifies semantically and structurally similar
-        historical examples using masked embeddings and schema distances.
-    \item \textbf{Schema Subsetting} – Reduces database schema complexity by focusing
-        on relevant tables and relationships
-    \item \textbf{Query Generation} – Few-shot learning with fine tuned models translate natural language queries to SQL
-    \item \textbf{Self-Correction} – Iteratively refines queries through execution feedback and error analysis
-    \item \textbf{Consensus Voting} – Selects the most reliable result from multiple
-        generation attempts
+    \item \textbf{Example Selection} $\sigma$ – Identifies semantically and structurally similar examples using cosine similarity and schema distance.
+    \item \textbf{Schema Subsetting} $\phi$ – Reducing schema complexity by subsetting the schema to the relevant subset of tables and relationships.
+    \item \textbf{Query Projection} $\pi$ – Few-shot learning with a finetuned model to project natural language queries to SQL statements.
+    \item \textbf{Self Refinement} $\rho$ – Self refinement of generated SQL statements through execution feedback and error analysis.
+    \item \textbf{Voting} $\nu$ – Self consensus voting mechanism to choose the most likely result from multiple generation attempts.
 \end{enumerate}
@@
 % TODO: Make a nice diagram for this
+
 \begin{verbatim}
@@
 
-This design builds upon concepts from DAIL-SQL \citep{DAIL} and incorporates novel
-contributions in schema-aware example selection and graph-based structural similarity.
+This design largely builds upon in context learning concepts from DAIL-SQL \citep{DAIL-SQL}
+and OmniSQL \citep{OmniSQL} but incorporates novel contributions in schema-aware example
+selection by harnessing a graph-based structural similarity metric.
 
 \subimport{}{initialization}
-
-\subimport{}{components}
-
-\subimport{}{composition}
+\subimport{}{functions}

### 2025-10-06 thesis/design/composition.tex
Improve design section layout

\subsection{Composition – $nq$}

The $nq$ function composes the system by first establishing a
zero-shot baseline, then using the precomputed embedding space $v$ and distance
index $d$ from the initialization phase to guide few-shot generation while
self refining results through execution feedback and finally yield the most likely
candidate $\omega$ through majority voting.

$$
nq: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \mathcal{V} \times \mathcal{D} \rightarrow \Omega_{\mathcal{S}}
$$

\vspace{0.5em}

For a specific input instance, we write:

$$
nq(q, s, e, v, d) = \omega
$$

\vspace{0.5em}

where $q \in \mathcal{Q}$ is the input natural language query, $s \in \mathcal{S}$
is the database schema, $e \in \mathcal{E}$ is the execution function for validation,
$v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index,
and $\omega \in \Omega_{\mathcal{S}}$ is the generated SQL query.

The $nq$ function follows the algorithm~\ref{algorithms:nq} definition.

\begin{algorithm}
\caption{$nq$}\label{algorithms:nq}
\begin{algorithmic}[1]
\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
\Require $k \in \mathbb{N}$, $k \geq 1$             \Comment{Number of candidates to generate}
\State $z \gets \pi(\emptyset, q, s)$               \Comment{Generate zero-shot baseline}
\State $\mathcal{C}' \gets \sigma(q, z, s, v, d)$   \Comment{Select relevant examples}
\State $C \gets \emptyset$                          \Comment{Initialize candidate set}
\While{$|C| < k$}                                   \Comment{Generate k candidates}
    \State $s' \gets \phi(\mathcal{C}', s)$         \Comment{Subset schema to relevant elements}
    \State $\omega \gets \pi(\mathcal{C}', q, s')$  \Comment{Generate candidate query}
    \If{$e(\omega, s')$}                            \Comment{Validate syntactic correctness}
        \State $C \gets C \cup \{\omega\}$          \Comment{Add valid candidate}
    \EndIf

    \State $\omega' \gets \rho(q, s', e, \omega)$   \Comment{Attempt self-correction}
    \If{$e(\omega', s')$}                           \Comment{Validate corrected query}
        \State $C \gets C \cup \{\omega'\}$         \Comment{Add corrected candidate}
    \EndIf
\EndWhile
\State \Return $\nu(C, e)$                          \Comment{Select consensus result}
\end{algorithmic}
\end{algorithm}

### 2025-10-06 thesis/design/functions.tex
Improve design section layout

@@
 \end{enumerate}
-
-\subsubsection{Query – $nq$}
-
-We define the query function $nq$ is collapsing a natural language question $q$,
-the database schema $s$ and a indices over historically observed examples and
-databases $v$ and $d$ into one SQL query candidate $\omega$.
-
-$$
-nq: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \mathcal{V} \times \mathcal{D} \rightarrow \Omega_{\mathcal{S}}
-$$
-
-\vspace{0.5em}
-
-For a specific input instance, we write:
-
-$$
-nq(q, s, e, v, d) = \omega
-$$
-
-\vspace{0.5em}
-
-where $q \in \mathcal{Q}$ is the input natural language query, $s \in \mathcal{S}$
-is the database schema, $e \in \mathcal{E}$ is the execution function for validation,
-$v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index,
-and $\omega \in \Omega_{\mathcal{S}}$ is the generated SQL query. The function is comprised
-out of the subsequently defined function and follows the algorithm described in Algorithm~\ref{algorithms:nq}.
 

### 2025-10-06 thesis/design/index.tex
Improve design section layout

@@
 
-This chapter describes the design of \textsc{Natural}, the proposed NL2SQL system, 
+This chapter describes the design of \textsc{Natural}, our proposed NL2SQL system, 
 addressing limitations and research gaps identified in the literature review.
-The design follows a modular architecture that leverages both traditional computer
-science algorithms and machine learning models to achieve robust, portable and
-extensible natural language to SQL translation capabilities.
 
-\subsection{Overview}
-
-The \textsc{Natural} system implements a multi-stage pipeline that transforms natural
-language questions into executable SQL queries through example selection,
-schema subsetting, and iterative refinement. The proposed architecture consists
-of five core components:
+\textsc{Natural} is architected as a pipeline that transforms natural
+language questions into SQL queries using example selection ($\sigma$),
+schema subsetting ($\phi$), and self refinement ($\rho$) and voting ($\nu$).
+The proposed system consists of five components:
 
@@
 
-The system processes queries through the following pipeline:
+The system processes queries using the following algorithm:
 
@@
 
-This design largely builds upon in context learning concepts from DAIL-SQL \citep{DAIL-SQL}
+This design largely builds upon few-shot learning concepts from DAIL-SQL \citep{DAIL-SQL}
 and OmniSQL \citep{OmniSQL} but incorporates novel contributions in schema-aware example
@@
 \subimport{}{functions}
+\subimport{}{composition}

### 2025-10-06 thesis/design/initialization.tex
Improve design section layout

@@
 
-Given a database schema $s$, the corresponding graph $G_s = (V, E, \ell, w)$ is defined as:
+Given a database schema $s$ we define the corresponding graph $G_s = (V, E, \ell, w)$ as:
 
@@
         \end{itemize}
-    \item Each edge $e \in E$ has a weight $w(e) \in \{0.5, 0.9, 1.0\}$ reflecting structural importance:
+    \item Each edge $e \in E$ has a weight $w(e) \in \{0.5, 0.9, 1.0\}$ reflecting its structural importance:
         \begin{itemize}
@@
 This hierarchical labeling ensures that structural relationships take
-precedence over semantic content, while omitting table names, column names and
+precedence over exact content, while omitting table names, column names and
 domain terminology to achieve schema-agnostic comparison.
@@
 
-For the database schemas above, the normalized schema (5 nodes, 5 edges) and
-denormalized schema (3 nodes, 2 edges) show significant topological differences.
+For the database schemas above, the graph representation of the normalized
+schema $G_{norm}$ (see figure~\ref{figure:normalized-graph-representation})
+with 5 nodes and 5 edges and denormalized schema $G_{denorm}$ (see figure~\ref{figure:denormalized-graph-representation})
+with 3 nodes and 2 edges show significant topological differences.
 The WWL distance captures both the reduction in graph complexity (fewer nodes and edges)

### 2025-10-06 thesis/design/functions.tex
Add placeholder algorithms

@@
 
-The following sets are subsequently used to introduce the functions $nq$, $\sigma$, $\phi$, $\pi$, $\rho$ and $\nu$.
+The following sets are subsequently used to introduce the functions $\sigma$, $\phi$, $\pi$, $\rho$ and $\nu$.
 
@@
 
+\begin{algorithm}
+\caption{$\sigma$ - Example Selection}\label{algorithms:sigma}
+\begin{algorithmic}[1]
+%\Require $q \in \mathcal{Q}$, $z \in \Omega_{\mathcal{S}}$, $s \in \mathcal{S}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
+%\Require $k \in \mathbb{N}$, $k \geq 1$             \Comment{Number of examples to select}
+%\State $q' \gets mask(q)$                           \Comment{Mask input query}
+%\State $z' \gets mask(z)$                           \Comment{Mask zero-shot result}
+%\State $candidates \gets \emptyset$                 \Comment{Initialize candidate set}
+%\For{each $(q_i, s_i, \omega_i) \in v$}            \Comment{Iterate through embedding space}
+    %\State $sim_q \gets cosine(\iota(q'), \iota(mask(q_i)))$  \Comment{Query similarity}
+    %\State $sim_s \gets cosine(\iota(z'), \iota(mask(\omega_i)))$  \Comment{SQL similarity}
+    %\State $sim_d \gets d(s, s_i)$                  \Comment{Schema distance}
+    %\State $score \gets w_q \cdot sim_q + w_s \cdot sim_s + w_d \cdot sim_d$
+    %\State $candidates \gets candidates \cup \{(q_i, s_i, \omega_i, score)\}$
+%\EndFor
+%\State \Return $top\_k(candidates, k)$              \Comment{Return k highest scoring examples}
+\end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Schema Subsetting – $\phi$}\label{design:subsetting-function}
@@
 
+\begin{algorithm}
+\caption{$\phi$ - Schema Subsetting}\label{algorithms:phi}
+\begin{algorithmic}[1]
+%\Require $c \in \mathcal{C}'$, $s \in \mathcal{S}$
+%\State $tables \gets \emptyset$, $columns \gets \emptyset$, $relations \gets \emptyset$
+%\For{each $(q_i, s_i, \omega_i, d_i) \in c$}       \Comment{Analyze selected examples}
+    %\State $parsed \gets parse\_sql(\omega_i)$      \Comment{Extract schema elements from SQL}
+    %\State $tables \gets tables \cup tables(parsed)$
+    %\State $columns \gets columns \cup columns(parsed)$
+    %\State $relations \gets relations \cup foreign\_keys(parsed)$
+%\EndFor
+%\State $s' \gets construct\_schema(tables, columns, relations)$  \Comment{Build reduced schema}
+%\State \Return $s'$                                 \Comment{Return schema subset}
+\end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Query Projection – $\pi$}\label{design:projection-function}
@@
 adhering to the constraints imposed by the database schema.
+
+\begin{algorithm}
+\caption{$\pi$ - Query Projection}\label{algorithms:pi}
+\begin{algorithmic}[1]
+\end{algorithmic}
+\end{algorithm}
 
@@
 
+\begin{algorithm}
+\caption{$\rho$ - Self Refinement}\label{algorithms:rho}
+\begin{algorithmic}[1]
+%\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $\omega_{raw} \in \Omega_{\mathcal{S}}$
+%\Require $max\_attempts \in \mathbb{N}$             \Comment{Maximum refinement attempts}
+%\State $\omega_{current} \gets \omega_{raw}$
+%\State $attempts \gets 0$
+%\While{$attempts < max\_attempts$}
+    %\If{$e(\omega_{current}, s)$}                   \Comment{Check if query is valid}
+        %\State \Return $\omega_{current}$           \Comment{Return if valid}
+    %\EndIf
+    %\State $error \gets get\_error(e, \omega_{current}, s)$  \Comment{Extract error details}
+    %\State $prompt \gets construct\_refinement\_prompt(q, s, \omega_{current}, error)$
+    %\State $\omega_{current} \gets \pi(prompt, s, \emptyset)$  \Comment{Generate refined query}
+    %\State $attempts \gets attempts + 1$
+%\EndWhile
+%\State \Return $\omega_{current}$                   \Comment{Return best attempt}
+\end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Voting – $\nu$}\label{design:voting-function}
@@
 no clear majority exists, the function may apply additional heuristics such as query
-complexity or execution performance to determine the final proposed query candidate $w_{consensus}$.
+complexity or execution performance to determine the final proposed query candidate $\omega_{consensus}$.
+
+\begin{algorithm}
+\caption{$\nu$ - Consensus Voting}\label{algorithms:nu}
+\begin{algorithmic}[1]
+%\Require $C \in \mathcal{C}$, $e \in \mathcal{E}$
+%\State $valid\_queries \gets \{\omega \in C : e(\omega, s)\}$  \Comment{Filter valid queries}
+%\If{$|valid\_queries| = 0$}
+    %\State \Return $null$                           \Comment{No valid candidates}
+%\EndIf
+%\State $frequency\_map \gets \{\}$                  \Comment{Count query frequencies}
+%\For{each $\omega \in valid\_queries$}
+    %\State $result \gets execute(\omega, s)$        \Comment{Execute query to get result}
+    %\State $frequency\_map[result] \gets frequency\_map[result] + 1$
+%\EndFor
+%\State $max\_freq \gets max(frequency\_map.values())$
+%\State $consensus\_result \gets argmax(frequency\_map, max\_freq)$
+%\For{each $\omega \in valid\_queries$}             \Comment{Find query producing consensus result}
+    %\If{$execute(\omega, s) = consensus\_result$}
+        %\State \Return $\omega$                     \Comment{Return consensus query}
+    %\EndIf
+%\EndFor
+\end{algorithmic}
+\end{algorithm}

### 2025-10-06 thesis/design/initialization.tex
Improve the init section

@@
-\subsection{Initialization} % TODO: FIX THIS SECTION
+\subsection{Initialization}
 
@@
 with historical data to construct a specific embedding space $v \in \mathcal{V}$ and
-schema distance index $d \in \mathcal{D}$. This phase analyzes and samples existing
-datasets and previous user interactions to build 
-advanced example selection.
+schema distance index $d \in \mathcal{D}$. This preprocessing phase analyzes existing
+datasets and previous user interactions to enable semantically-aware example selection.
+
+The initialization process consists of two main components: (1) embedding generation
+for semantic similarity computation, and (2) schema indexing for structural similarity
+measurement. These components work together to support the example selection function $\sigma$.
 
@@
 
-Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes 
-cosine similarity in an embedding space $v$ to search for examples based on the
-masked natural language query $q'$ and the respective masked SQL query $\omega'$.
-This allows $\sigma$ to use cosine distance to search through historically observed
-question answer pairs; eg, from previous user interactions or prevalent training
-datasets like \textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.
+Building upon DAIL-SQL's masked example selection approach, \textsc{Natural} constructs
+an embedding space $v$ to enable semantic similarity search. The system uses cosine
+similarity to identify relevant examples based on masked representations of both
+natural language queries and SQL statements, allowing efficient retrieval from
+large collections of question-answer pairs.
 
-We therefore embedd a set of masked samples $\mathcal{S}$ into an embedding space $v_{\mathcal{S}}$,
-where $\mathcal{S} = \{(q_1, s_1), ..., (q_i, s_i)\}$, $q$ is the natural language
-question and $s$ is the corresponding SQL query.
+We therefore embed a set of training samples $\mathcal{T} = \{(q_1, \omega_1), ..., (q_n, \omega_n)\}$
+into an embedding space $v$, where each $q_i$ is a natural language question and $\omega_i$
+is the corresponding SQL query.
 
-The embedding space $v_{\mathcal{S}}$ can be defined as:
+The embedding space $v$ can be formally defined as:
 
-$$v_{\mathcal{S}} = \{~(q, \iota(mask(q)), s, \iota(mask(s)))~|~(q, s) \in \mathcal{S}~\}$$
+$$v = \{~(q, \iota(mask(q)), \omega, \iota(mask(\omega)))~|~(q, \omega) \in \mathcal{T}~\}$$
 
-Where the masking function $mask$ is defined as described in the sections~\ref{design:nl-masking}
-and~\ref{design:sql-masking} and the embedding function is defined as $\iota$.
-% TODO: Specify embedding model
+where $mask$ represents the masking function described in sections~\ref{design:nl-masking}
+and~\ref{design:sql-masking}, and $\iota$ is the embedding function that maps text to vector representations.
 
@@
 $w$ in a constant whitelist $\mathcal{W}$ are determined structurally relevant.
-% TODO: Fix typo
-Estimating a constant $\mathcal{W}$ is described in section~\ref{implementation:whitelisting}.
-% TODO: Fix reference to implementation section
+The constant $\mathcal{W}$ contains common structural SQL keywords and is determined empirically.
 
@@
 \end{verbatim}
-
-% TODO: Discuss potential limitations of masked embedding approach
-
-% ALIGNMENT IMPROVEMENTS NEEDED:
-% TODO: Add formal mathematical definition using enumerate style like indexing section
-% TODO: Use consistent tuple notation (q,s,ω,d) matching definitions section  
-% TODO: Add cross-references to Section~\ref{design:indexing} for schema distances
-% TODO: Include similarity scoring methodology with weights w_q, w_s, w_sql
-% TODO: Add subsection on "Distance Calculation" parallel to indexing section
-% TODO: Use consistent variable naming (mathcal{V} vs V vs v)
-% TODO: Add example showing multi-dimensional scoring like indexing graphs
-% TODO: Include legend/explanation of masking tokens like indexing legend
 
@@
 representation allows us to leverage established methods for measuring graph similarity
-such as Wasserstein-Weisfeiler-Lehman graph kernels \citep{WWL}. % TODO: Add proper WWL citation
-
+such as Wasserstein-Weisfeiler-Lehman graph kernels \citep{WWL}. 
 Given a database schema $s$ we define the corresponding graph $G_s = (V, E, \ell, w)$ as:
@@
 
-% TODO: Clarify relationship between edge weights and WWL computation
-% TODO: Add example showing how label priority affects similarity scores
-
 \paragraph{Distance Index}
@@
 a database schema name and $di_i$ denotes its WWL distance to the current schema.
-
-% TODO: Fix inconsistent notation: use consistent variable names (G_s vs Natural-graphs)
-% TODO: Improve transition between graph definition and distance measurement sections

### 2025-10-06 thesis/theoretical-foundations/graphs.tex
Add theory sections i need to write

\subsection{Graph Theory Fundamentals}

% TODO: Add graph theory basics section (needed for WWL kernel approach)
% - Basic graph definitions: vertices, edges, labeled graphs
% - Graph representations and adjacency structures
% - Graph similarity metrics and kernels
% - Weisfeiler-Lehman graph features
% - Wasserstein distance and optimal transport theory
% - Applications to database schema representation

% TODO: Connect to design section
% - How database schemas map to graph structures
% - WWL kernel computation for schema similarity
% - Distance indexing for efficient retrieval
% - Integration with example selection function Ã

### 2025-10-06 thesis/theoretical-foundations/llms.tex
Add theory sections i need to write

\subsection{Large Language Models}

% TODO: Complete LLM theoretical foundations section (essential for LLM-based approach)
% - Transformer architecture fundamentals
% - Attention mechanisms and self-attention
% - Pre-training vs fine-tuning paradigms
% - In-context learning and few-shot learning
% - Prompt engineering and instruction following
% - Model scaling laws and emergent capabilities
% - Limitations: hallucination, context windows, reasoning

% TODO: Connect to design section
% - How LLMs enable natural language understanding for SQL generation
% - Role in query projection function À
% - Self-refinement capabilities Á
% - Voting and consensus mechanisms ½

### 2025-10-06 thesis/theoretical-foundations/nlp.tex
Add theory sections i need to write

\subsection{Natural Language Processing}

% TODO: Complete NLP theoretical foundations section
% - Text preprocessing and tokenization
% - Word embeddings and semantic representations
% - Language modeling fundamentals
% - Syntactic and semantic parsing
% - Named entity recognition and linking
% - Text similarity metrics (cosine, semantic distance)
% - Masked language modeling approaches

% TODO: Connect to design section
% - Role in natural language query understanding
% - Embedding space construction for example selection
% - Masking techniques for domain generalization
% - Semantic similarity computation in function Ã

### 2025-10-12 thesis/design/functions.tex
Finish the design section for now

--- 2025-10-06 thesis/design/functions.tex
@@
 
+\begin{algorithm}
+\caption{$\sigma$ - Example Selection}\label{algorithms:sigma}
+\begin{algorithmic}[1]
+\Require $q \in \mathcal{Q}$, $z \in \Omega_{\mathcal{S}}$, $s \in \mathcal{S}$, $v \in \mathcal{V}$, $d \in \mathcal{D}$
+\Require $k \in \mathbb{N}$, $k \geq 1$                            \Comment{Number of examples to select}
+\State $q' \gets mask(q)$                                          \Comment{Mask input question}
+\State $z' \gets mask(z)$                                          \Comment{Mask zero-shot result}
+\State $candidates  \gets \emptyset$                               \Comment{Initialize candidate set}
+\For{$(q_i, s_i, \omega_i) \in v$}                            \Comment{For each sample}
+    \State $sim_q \gets cosine(\iota(q'), \iota(mask(q_i)))$       \Comment{Calculate question similarity}
+    \State $sim_s \gets cosine(\iota(z'), \iota(mask(\omega_i)))$  \Comment{Calculate SQL similarity}
+    \State $sim_d \gets d(s, s_i)$                                 \Comment{Calculate schema distance}
+    \State $score \gets w_q \cdot sim_q + w_s \cdot sim_s + w_d \cdot sim_d$
+    \State $candidates \gets candidates \cup \{(q_i, s_i, \omega_i, score)\}$
+\EndFor
+\State \Return $top_{k}(candidates)$                             \Comment{Return k highest scoring examples}
+\end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Schema Subsetting – $\phi$}\label{design:subsetting-function}
@@
 The function analyzes the selected examples in $c$ to identify which schema elements
-(tables, columns, foreign key relationships) are commonly referenced in similar queries.
-This analysis enables the system to steer the model's attention on the most relevant
-portions of potentially large and complex database schemas, thereby improving both accuracy
-and computational efficiency and prevents exceeding limited context windows.
+are commonly referenced in similar queries. This analysis enables the system to
+steer the model's attention on the most relevant portions of potentially large
+and complex database schemas, thereby improving both accuracy and computational
+efficiency and prevents exceeding limited context windows.
+
+\begin{algorithm}
+\caption{$\phi$ - Schema Subsetting}\label{algorithms:phi}
+\begin{algorithmic}[1]
+\Require $c \in \mathcal{C}'$, $s \in \mathcal{S}$
+\State $s' \gets \emptyset$                          \Comment{Empty schema subset}
+\For{$table \in s$}
+    \If{$\exists~r \in references(c, table)$}        \Comment{Any candidates reference the table}
+        \State $s' \gets s' \cup \{table\}$          \Comment{Extend subsetted schema}
+    \EndIf
+\EndFor
+\State \Return $s'$                                  \Comment{Return subsetted schema}
+\end{algorithmic}
+\end{algorithm}
 
@@
 adhering to the constraints imposed by the database schema.
+
+\begin{algorithm}
+\caption{$\pi$ - Query Projection}\label{algorithms:pi}
+\begin{algorithmic}[1]
+\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $c \in \mathcal{C}'$
+\Require $\tau \in \mathbb{R}$, $\tau > 0$                     \Comment{Relevance threshold}
+\State $c_{rel} \gets \{~(q_i, s_i, \omega_i, d_i) \in c : d_i < \tau~\}$  \Comment{Filter by relevance}
+\State $ex \gets \{~fmt(q_i, \omega_i, d_i)~|~(q_i, s_i, \omega_i, d_i) \in c_{rel}~\}$  \Comment{Format examples}
+\State $prompt \gets prompt(q, s, ex)$                         \Comment{Construct prompt}
+\State $out \gets model(prompt)$                               \Comment{Model inference}
+\State $\Omega_{cand} \gets extract_{sql}(out)$                \Comment{Extract SQL candidates}
+\For{$\omega_{raw} \in \Omega_{cand}$}                         \Comment{Validate candidates}
+    \If{$\omega_{raw} \in \Omega_{\mathcal{S}}$}               \Comment{Check syntactic validity}
+        \State \Return $\omega_{raw}$                          \Comment{Return first valid query}
+    \EndIf
+\EndFor
+\end{algorithmic}
+\end{algorithm}
 
@@
 
+\begin{algorithm}
+\caption{$\rho$ - Self Refinement}\label{algorithms:rho}
+\begin{algorithmic}[1]
+\Require $q \in \mathcal{Q}$, $s \in \mathcal{S}$, $e \in \mathcal{E}$, $\omega_{raw} \in \Omega_{\mathcal{S}}$
+\State $exec \gets e(s, \omega_{raw})$                          \Comment{Verify execution output}
+\State $prompt \gets prompt_{refine}(q, s, \omega_{raw}, exec)$ \Comment{Construct refinement prompt}
+\State $out \gets model(prompt)$                                \Comment{Generate refinement output}
+\State $\Omega_{ref} \gets extract_{sql}(out)$                  \Comment{Extract refined candidates}
+\For{$\omega_{ref} \in \Omega_{ref}$}                           \Comment{Validate refined queries}
+    \If{$\omega_{ref} \in \Omega_{\mathcal{S}}$}                \Comment{Check syntactic validity}
+        \State \Return $\omega_{ref}$                           \Comment{Return first valid refinement}
+    \EndIf
+\EndFor
+\end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Voting – $\nu$}\label{design:voting-function}
@@
 no clear majority exists, the function may apply additional heuristics such as query
-complexity or execution performance to determine the final proposed query candidate $w_{consensus}$.
+complexity or execution performance to determine the final proposed query candidate $\omega_{consensus}$.
+
+\begin{algorithm}
+\caption{$\nu$ - Consensus Voting}\label{algorithms:nu}
+\begin{algorithmic}[1]
+\Require $C \in \mathcal{C}$, $e \in \mathcal{E}$
+\State $results \gets \{\}$                               \Comment{Map from result sets to candidate lists}
+\For{$\omega \in C$}                                      \Comment{Execute each candidate}
+    \State $result \gets e(\omega)$                       \Comment{Execute query}
+    \State $results[result] \gets results[result] \cup \{\omega\}$
+\EndFor
+\State $dist \gets \{results[r] : r \in results\}$        \Comment{Get all result groups}
+\State $dist \gets sort(dist, |group| \mapsto |group|)$   \Comment{Sort by group size (largest first)}
+\State \Return $top_1(top_1(dist))$                       \Comment{Return largest group}
+\end{algorithmic}
+\end{algorithm}

### 2025-10-22 thesis/implementation/index.tex
Split and finish the sampling section
Write an outline for the impl section
Finish the design section for now
implementation section!

@@
 
-\subsection{Development Environment and Tools}
+% ??????/
 
-\subsection{Integration of the Model}
+\subsection{Architecture and Infrastructure}
 
-\subsection{Development of the PostgreSQL Extension}
+The implementation of the system design phase is split into inference, sampling
+and evaluation code. The runtime code focuses on the actual algorithm
+implementations outlined in section \ref{section:design}, sampling code focuses on
+indexing samples and computing distance indices and the evaluation code runs
+the \textsc{Natural} pipeline on prevalent benchmarks.
 
-\subsection{Optimization}
+This subsection focuses on outlining the scope of each technological component,
+the design rationale behind them and and discusses the technology stack
+decisions.
 
-\newpage
+\subsubsection{Software Architecture}
+
+\textsc{Natural} consists of five software components:
+
+\begin{enumerate}
+    \item \texttt{natural-models} – For handling the actual model
+        execution on GPUs for inference and embedding.
+    \item \texttt{natural-graphs} – Graph library for representing
+        database schemas in graphs as well as computing graph similarities and
+        distances.
+    \item \texttt{natural-inference} – The core pipeline implementation using
+        other software components during inference time.
+    \item \texttt{natural-sampling} – The sampling setup that focuses on
+        indexing samples and computing graph distance indices.
+    \item \texttt{natural-benchmark} – The benchmarking setup used to run
+        experiments, continuously evaluate the accuracy of the system and
+        compile statistics.
+\end{enumerate}
+
+% Inter-crate communication patterns and shared data structures
+\begin{verbatim}
+    THIS NEEDS A DIAGRAM TO UNDERSTAND THE SOFTWARE ARCH.
+\end{verbatim}
+
+This separation of concerns emerged as the split between inference, sampling
+and evaluation code required shared fundamentals like model execution and graph
+representation which in turn improves testability and maintainability. Furthermore
+separating the inference, sampling and evaluation code yields a smaller
+footprint when embedding the inference code into databases.
+
+Challenges associated with this multi component architecture are mostly tied to
+dependency management and increased build complexity although these can be
+mitiaged through \texttt{cargo}'s workspace support.
+
+\subsubsection{Resource Management Strategy}
+
+Given the constrained 24 gigabytes of VRAM capacity of the RTX 3090 not all
+available model sizes can be used. While 14B model variants theoretically work,
+the need for an embedding model typically exhausts the VRAM unless using
+steep quantization formats. The \textsc{OmniSQL} 7B models where shown to have
+an average performance degredation of only 0.3\% compared to their larger 14B
+counterparts by \citeauthor{OmniSQL} in \citeyear{OmniSQL}. On the spider test
+dataset the 7B model surprisingly outperformed it's 14B counterpart by 0.6\%.
+Due to the unclear performance gains of the 14B model, the significant increase
+in inference time and the limited VRAM available, this thesis focuses on the 7B
+variant of \textsc{OmniSQL}.
+
+As an embedding model is required for doing semantic search of samples in the
+pipeline a small companion model is required to embedd both the sample datasets
+(\textsc{SynSQL}, \textsc{Bird} and \textsc{Spider}) during sampling time and
+the user provided natural language question during inference. The Qwen3 series
+embedding models ranked first place on the MTEB multilingual leaderboard. With
+the Q8\_0 quantized version of the 8B model consumes 8 gigabytes of VRAM which
+is not possible to fit into 24 gigabytes of VRAM when accounting for KV-Cache
+requirements. Therefore, instead of choosing a heavily quantized version of the
+8B variant (ie, Q4\_K\_M or below), the 4B model with Q8\_0 shows similar
+performance characteristics on simple usecase while offering a significantly
+smaller memory footprint at 2 gigabytes.
+\textcolor{red}{https://huggingface.co/Qwen/Qwen3-Embedding-8B cite}.
+
+As models have significant loading times, the system loads models gloablly and 
+at startup and hands around references using atomic reference counting
+(\texttt{std::sync::Arc}) to ensure that models are not loaded twice, can be
+reused between different inference calls and GPU memory is not exhausted which
+would lead to a program crash.
+
+\subsubsection{Technology Stack Decisions}
+
+\paragraph{Language and Ecosystem}
+
+The most apparent and impactful decision is likely the language and ecosystem
+choice made. Viable languages for implementing natural language processing and
+machine learning heavy systems are Python, R, Julia, Rust, C / C++ and Java as
+well as other general purpose languages.
+
+While interpreted languages like Python, R and Julia tend to be significantly
+easier to use for rapid prototyping and approach validation due to their loose
+type systems and great scientific ecosystem, they come with serious drawbacks
+with regards to deployability, speed and robustness compared to compiled and
+strongly typed languages.
+
+Languages like Java, C / C++ and Rust offer greater stability at runtime,
+better interoperability into other programs (eg, database extensions) and better
+resource utilization they represent an interesting tradeoff between performance
+optimization and portability vs. speed of iteration and research ecosystems.
+
+R and Julia are inferior to Python when it comes to adoption, machine learning
+frameworks and natural language processing. Java requires a Java Virtual
+Machine (JVM) at runtime which yields worse portability than C / C++ and Rust
+while offering little advantage over interpreted languages like Python. Thus
+Python, C / C++ and Rust emerge as strong contenders for the implementation of
+\textsc{Natural}. C and C++ have the primary downside that they are prone to
+program crashes and memory safety issues whereas Rust resolves most of these
+downsides while maintaining similar performance, memory management and
+portability characteristics.
+
+Given that performance plays a critical role when developing machine learning
+systems with limited access to hardware, the path to a potential production
+deployments of NL2SQL is easier and the language ecosystem offers bindings
+for the most notable scientific libraries, Rust offers a great value
+proposition for ML systems at the cost of development speed.
+
+\paragraph{Inference Framework}
+
+For local model inference, two primary frameworks emerged: llama.cpp (FFI), 
+Candle (Rust-native). While Rust-native frameworks generally offer better 
+portability, llama.cpp provided a better balance between performance, ecosystem 
+maturity, and implementation simplicity.
+
+llama.cpp provides out-of-the-box optimizations with comprehensive GGUF 
+quantization support, yielding \textcolor{red}{25-30 tokens per second} on
+consumer grade RTX 3090 hardware. It offers high-level abstractions for model
+loading and tokenization and sampling. The mature GGML ecosystem and advanced
+quantization schemes (like \texttt{Q4\_K\_M}, \texttt{Q8\_0}) justified the FFI
+complexity.
+
+Candle, even though it offered comparable raw inference speeds, required
+rather extensive low-level implementation work and proved incompatiblity with
+scenarios where the resulting binary is embedded into a database (e.g. PostgreSQL).
+The combination of manual tensor management, more convoluted quantization support,
+and deployment constraints made llama.cpp a better alternative for
+production-ready systems.
+
+\paragraph{Similarity Search Framework}
+
+For lightweight similarity search SQLite in combination with it's
+\texttt{sqlite-vec} extension is a sensible option. Especially for smaller data
+loads introducing the complexity of \texttt{faiss} and comparable vector search
+solutions like \texttt{qdrant} outweighs their speed benefits.
+
+The \texttt{sqlite-vec} extension advertises to be a ``fast enough'' vector
+search solution, allowing for reduced complexity and compatibility with
+graphical database interfaces that support SQLite to inspect the embedding
+space.
+
+\paragraph{Trade-offs}
+
+Overall the development overhead of choosing Rust over Python for the
+implementation phase was noticable; Rust's machine learning and natural
+language processing frameworks are less advanced, porting research code written
+in python from other papers (like \cite{OmniSQL}) turned out to be non trivial.
+This decision likely doubled the time needed for the implementation, but in
+turn provides a clear path for the algorithms in this thesis to be
+productionized. The outcome of the implementation has significantly better
+performance and portability characteristics than using Python would have
+allowed for.

### 2025-10-25 thesis/implementation/index.tex
Better impl section

@@
 
-% ??????/
+This chapter presents the practical implementation of the \textsc{Natural}
+system, translating the theoretical framework outlined in Chapter
+\ref{section:design} into a working NL2SQL system.
+
+Primary attention is given to the engineering challenges encountered,
+performance bottlenecks identified, and optimization strategies implemented to
+achieve practical deployment viability on consumer hardware.
+The chapter is comprised of a software architecture and infrastructure
+discussion as well as implementation outlines of each pipeline component.
 
@@
 This subsection focuses on outlining the scope of each technological component,
-the design rationale behind them and and discusses the technology stack
+the design rationale behind them and discusses the technology stack
 decisions.
@@
 dependency management and increased build complexity although these can be
-mitiaged through \texttt{cargo}'s workspace support.
+mitigated through \texttt{cargo}'s workspace support.
 
@@
 the need for an embedding model typically exhausts the VRAM unless using
-steep quantization formats. The \textsc{OmniSQL} 7B models where shown to have
-an average performance degredation of only 0.3\% compared to their larger 14B
+steep quantization formats. The \textsc{OmniSQL} 7B models were shown to have
+an average performance degradation of only 0.3\% compared to their larger 14B
 counterparts by \citeauthor{OmniSQL} in \citeyear{OmniSQL}. On the spider test
-dataset the 7B model surprisingly outperformed it's 14B counterpart by 0.6\%.
+dataset the 7B model surprisingly outperformed its 14B counterpart by 0.6\%.
 Due to the unclear performance gains of the 14B model, the significant increase
@@
 As an embedding model is required for doing semantic search of samples in the
-pipeline a small companion model is required to embedd both the sample datasets
+pipeline a small companion model is required to embed both the sample datasets
 (\textsc{SynSQL}, \textsc{Bird} and \textsc{Spider}) during sampling time and
@@
 8B variant (ie, Q4\_K\_M or below), the 4B model with Q8\_0 shows similar
-performance characteristics on simple usecase while offering a significantly
+performance characteristics on simple use cases while offering a significantly
 smaller memory footprint at 2 gigabytes.
@@
 
-As models have significant loading times, the system loads models gloablly and 
+As models have significant loading times, the system loads models globally and 
 at startup and hands around references using atomic reference counting
@@
 \subsubsection{Technology Stack Decisions}
+
+The choice of programming language, inference frameworks, and supporting
+libraries implies the system's performance and deployment characteristics,
+as well as development velocity. This section analyzes the key decisions
+made for \textsc{Natural}, discussing their trade-offs between research
+flexibility and production readiness, performance optimization and development
+speed, and ecosystem maturity versus cutting-edge capabilities.
+
+The critical decisions are the programming language, the model
+inference framework, and the vector similarity search solution. Each
+decision was evaluated against the constraints of limited hardware resources
+(24GB VRAM), the need for database integration capabilities, and the
+requirement for reproducible research outcomes.
 
@@
 better interoperability into other programs (eg, database extensions) and better
-resource utilization they represent an interesting tradeoff between performance
+resource utilization they represent an interesting trade-off between performance
 optimization and portability vs. speed of iteration and research ecosystems.
@@
 Candle, even though it offered comparable raw inference speeds, required
-rather extensive low-level implementation work and proved incompatiblity with
+rather extensive low-level implementation work and proved incompatibility with
 scenarios where the resulting binary is embedded into a database (e.g. PostgreSQL).
@@
 Overall the development overhead of choosing Rust over Python for the
-implementation phase was noticable; Rust's machine learning and natural
+implementation phase was noticeable; Rust's machine learning and natural
 language processing frameworks are less advanced, porting research code written
@@
 allowed for.
+
+\subsection{Pipeline Implementation}
+
+This subsection describes the system design realisation of each core pipeline
+component ($\sigma$, $\phi$, $\pi$, $\rho$, $\nu$) into working code, their
+algorithmic details, performance characteristics and trade-offs made.
+
+\subsubsection{Example Selection Engine ($\sigma$)}
+
+The example selection is implemented using a multi-dimensional similarity
+scoring system that identifies the most relevant samples that were previously
+indexed for subsequent in-context learning. The component addresses the
+fundamental challenge of selecting contextually appropriate examples from large
+training corpus while balancing semantic relevance and structural compatibility.
+
+\paragraph{Similarity Computation}
+
+The selection mechanism combines three distinct similarity measures implemented
+in the selection module in \texttt{natural-sampling::selection}. The 
+\texttt{selection} algorithm computes semantic similarity through cosine
+distance of the masked question embedding. Structural SQL similarity is
+computed via measuring the cosine distance of the masked SQL embedding, and
+schema compatibility using WWL kernel distance from the \texttt{natural-graphs}
+component. The embeddings are computed using the Qwen3-Embedding model (4B)
+using the query masking algorithm in \texttt{natural-sampling::masking}.
+
+\paragraph{Weighting Strategy and Performance Characteristics}
+
+As this algorithm is aware of three distinct ways to measure sample similarity
+(masked question, masked query, structural). The implementation uses
+empirically optimized weights defined as constants: 70\% for semantic question
+similarity and 30\% for SQL structural similarity. Additionally, the final
+scoring combines 70\% sample-level similarity with 30\% schema-level
+compatibility.
+
+The selection algorithm yields a maximum of 32 candidates (\texttt{TOP\_K =
+32}) to limit computational overhead while maintaining selection quality.
+Vector similarity queries are performed using SQLite with the sqlite-vec
+extension, achieving sub-second retrieval times for vector databases exceeding
+2,000,000 samples.
+
+\paragraph{Implementation Architecture}
+
+Using the \texttt{Selector} struct the embedding computation and representation
+is encapsulated from the actual selection algorithm. Thus via
+\texttt{Selector::new} the consuming code can compute the masked embeddings and
+use them subsequently to run example selection.
+
+The \texttt{selection} algorithm implementation follows a two-stage approach,
+split into initial candidate retrieval through vector search and subsequent
+reweighting using the above described weights and a precomputed WWL distance
+index from \texttt{natural-graphs}.

### 2025-10-25 thesis/implementation/index.tex
More impl stuff

@@
 The \texttt{selection} algorithm implementation follows a two-stage approach,
-split into initial candidate retrieval through vector search and subsequent
-reweighting using the above described weights and a precomputed WWL distance
-index from \texttt{natural-graphs}.
+split into initial candidate retrieval through vector search (using the
+\texttt{Vector} struct and SQLite) and subsequent reweighting using the above
+described weights and a precomputed WWL distance index from
+\texttt{natural-graphs}.
+
+\subsubsection{Schema Subsetting System ($\phi$)}
+
+% - Execution-based validation vs heuristic approaches
+% - Dynamic pruning: 500-800ms for 20+ table schemas vs 10-15ms heuristic
+% - In-memory database creation and testing for candidate queries
+% - Schema dependency analysis challenges (nested subqueries, CTEs)
+% - Correctness vs performance trade-off: eliminates pipeline failures
+
+The schema subsetting system is implemented in
+\texttt{natural-inference::pipeline::subsetting} as
+\texttt{SchemaSubsetter} struct. The core algorithm is the \texttt{optimize}
+method (lines 21-47) which performs execution-based validation to determine
+which tables are crucial for the query candidates provided.
+
+\paragraph{Execution-based Validation}
+
+The implementation uses a trial-and-error approach where for each query
+candidate and each table in the schema, it creates an in-memory SQLite database
+using \texttt{Connection::open\_in\_memory()} and populates it's schema with all
+tables except the current one being tested. It then attempts to prepare every
+query candidate against this reduced schema.
+
+If the query preparation fails, this indicates the removed table is crucial
+for the query, and it gets added to the list of crucial tables.
+This process is repeated for all query-table combinations to build a minimal
+schema containing only essential tables for the provided set of query candidates.
+
+\paragraph{Performance and Trade-off Characteristics}
+
+The algorithm creates a new in-memory database for each table-query
+combination, leading to $O(queries \times tables)$ complexity.
+
+\textcolor{red}{For schemas with XXX+ tables and multiple query candidates, this results in XXX-XXXms processing time
+compared to 10-15ms for heuristic approaches.}
+
+However, this execution-based approach provides correctness guarantees by
+eliminating pipeline failures caused by missing table dependencies. The
+method handles complex SQL dependencies including nested subqueries, CTEs,
+and foreign key relationships that heuristic approaches often miss.
+
+\paragraph{Pipeline Integration}
+
+The \texttt{SchemaSubsetter} is used prior to prompting the model using
+in-context learning to ensure that the model context is used optimally.
+The \texttt{optimize} method returns a \texttt{Schema} object
+containing only the tables identified as crucial through execution testing
+which is in turn processed by the subsequent pipeline stages like generation
+and refinement.

### 2025-10-25 thesis/implementation/index.tex
More impl content

--- 2025-10-25 thesis/implementation/index.tex
@@
 
-The example selection is implemented using a multi-dimensional similarity
+The example selection system is implemented using a multi-dimensional similarity
 scoring system that identifies the most relevant samples that were previously
@@
 fundamental challenge of selecting contextually appropriate examples from large
-training corpus while balancing semantic relevance and structural compatibility.
+training corpora whilst balancing semantic relevance and structural compatibility.
 
@@
 As this algorithm is aware of three distinct ways to measure sample similarity
-(masked question, masked query, structural). The implementation uses
+(masked question, masked query, structural). The implementation employs
 empirically optimized weights defined as constants: 70\% for semantic question
@@
 The selection algorithm yields a maximum of 32 candidates (\texttt{TOP\_K =
-32}) to limit computational overhead while maintaining selection quality.
-Vector similarity queries are performed using SQLite with the sqlite-vec
+32}) to limit computational overhead whilst maintaining selection quality.
+Vector similarity queries are performed using SQLite with the \texttt{sqlite-vec}
 extension, achieving sub-second retrieval times for vector databases exceeding
@@
 The \texttt{selection} algorithm implementation follows a two-stage approach,
-split into initial candidate retrieval through vector search and subsequent
-reweighting using the above described weights and a precomputed WWL distance
-index from \texttt{natural-graphs}.
+split into initial candidate retrieval through vector search (using the
+\texttt{Vector} struct and SQLite) and subsequent reweighting using the above
+described weights and a precomputed WWL distance index from
+\texttt{natural-graphs}.
+
+\subsubsection{Schema Subsetting System ($\phi$)}
+
+The schema subsetting system is implemented in
+\texttt{natural-inference::pipeline::subsetting} as
+\texttt{SchemaSubsetter} struct. The core algorithm is the \texttt{optimize}
+method (lines 21-47) which performs query validation to determine
+which tables are crucial for the query candidates provided.
+
+\paragraph{Query Validation}
+
+Query validation employs a trial-and-error approach where for each query
+candidate and table in the schema, an in-memory SQLite database
+is created using \texttt{Connection::open\_in\_memory()}. Subsequently the
+connection is initialized using the schema with all tables in the schema except
+the current one.
+
+Once the connection is ready, the \texttt{SchemaSubsetter} prepares every
+query candidate against this reduced schema. If preparation fails, this
+indicates the table removed is crucial for the query, thus it gets added to the
+list of crucial tables. This process is repeated for all query-table
+combinations to build a minimal schema containing only essential tables for the
+execution of the set of query candidates provided to the subsetting algorithm.
+
+\paragraph{Performance and Trade-off Characteristics}
+
+As a new in-memory database for each table-query combination is created, this
+leads to $O(queries \times tables)$ complexity. \textcolor{red}{For schemas
+with XXX+ tables and multiple query candidates, this results in XXX-XXXms
+processing time compared to 10-15ms for heuristic approaches.}
+
+However, this execution-based approach provides superior correctness guarantees since heuristic
+approaches cannot determine whether candidates will execute successfully in real
+database environments. Using a real SQLite in memory instance both the correctness of
+query candidates and the actually referenced set of tables can be known prior
+to actual query execution.
+
+\paragraph{Pipeline Integration}
+
+The \texttt{SchemaSubsetter} is used prior to prompting the model using ICL to
+ensure that the model context is used efficiently and attention is given to the
+relevant parts of the schema. The \texttt{optimize} method returns a
+\texttt{Schema} object containing only the tables identified as crucial through
+execution testing which is in turn processed by the subsequent pipeline stages
+like generation and refinement.
+
+\subsubsection{Query Projection ($\pi$)}
+
+The ICL module is implementing the query projection algorithm described in
+section~\ref{design:projection-function}. It is implemented in the
+\texttt{ICLGenerator} struct and wraps the \texttt{SqlModel} struct from
+\texttt{natural-models}. Using llama.cpp it runs model inference using a
+prompt optimized for in-context-learning with \textsc{OmniSQL}.
+
+\paragraph{In-Context Learning}
+
+The \texttt{ICLGenerator::generate} method implements few-shot prompting
+with relevance filtering. Relevance filtering refers to removing all selected
+samples with a similarity of less than 0.5 (this value was emperically derived
+from evaluations). Thus only semantically or structurally similar samples are
+actually provided to the model.
+
+The prompt is constructed based on an adapted version of the \textsc{OmniSQL}
+format \citep{OmniSQL}: task overview, sql schema, filtered examples with
+similarity scores, explicit instructions, and the target question.
+The biggest differentiation to the prompt of \citeauthor{OmniSQL} is the
+example section including similarity. For the actual SQL query presentation the
+code representation prompt format is used, inspired by DAIL-SQL
+\citep{DAIL-SQL}. Every example includes the masked question, similarity score,
+and formatted SQL query for clarity.
+
+\begin{verbatim}
+EXAMPLE OF THE PROMPT
+\end{verbatim}
+
+\paragraph{Prompt Engineering Strategy}
+
+As outlined above the \textsc{OmniSQL} prompt was used as base for
+\textsc{Natural}'s prompt together with a . \textsc{Natural} uses more explicit instructions
+compared to DAIL-SQL. Key differentiations include the inclusion of similarity
+scores to give the model the ability to weight the samples itself and
+
+The prompt emphasizes precision and chain-of-thought reasoning. 
+This aims to address common LLM issues with over-generating information or missing
+query constraints, as well as the hallucination and accuracy concerns
+identified in literature review section where LLMs generate plausible but
+incorrect SQL queries.
+
+\paragraph{Model Integration and Performance}
+
+The \texttt{SqlModel} is wrapped around \texttt{llama-cpp-2} bindings and
+loaded globally at startup to avoid a 30-60 second initialization times per
+query. Using the \texttt{prompt} method from \texttt{natural-models}
+tokenization, inference, and decoding with configurable PromptParams for
+context size and generation limits is handled automatically.
+
+\paragraph{Output Processing and Validation}
+
+As \textsc{OmniSQL} was finetuned to output its thoughts and predictions using
+markdown a markdown postprocessing module is needed, as well as a module to
+identify whether a SQL query is syntactically valid.
+
+To extract possible candidates all generated responses are post-processed
+through a markdown parser (\texttt{pulldown-cmark}) which parses the model
+output and looks for the code-block fence characters \texttt{\`\`\` }.
+
+Subsequently all candidates are processed in reversed order and the first full
+valid query is returned as potential candidate. The reversing of processing
+order is needed as the model outputs it's thoughts top-to-bottom as a markdown
+document with the most likely answer usually being output at the end. This
+approach ensures that a returned query candidates are executable and
+ensured to contain valid SQL. Thus \textsc{Natural} can aid the
+difficulty of generating perfect SQL queries by acknowledging
+limitations from large language models and implementing recovery and refinement
+mechanisms in the subsequent pipeline flow.

### 2025-10-25 thesis/implementation/index.tex
Fix typos

@@
 
+\paragraph{Prompt Engineering Strategy}
+
+As outlined above the \textsc{OmniSQL} prompt was used as base for
+\textsc{Natural}'s prompt together with a code representation prompt.
+\textsc{Natural} uses more explicit instructions compared to DAIL-SQL. Key
+differentiations include the inclusion of similarity scores to give the model
+the ability to weight the samples itself.
+
+This prompt steers the model towards precision and chain-of-thought reasoning. 
+It addresses apparent LLM issues like verbosity, over complexity or missing
+query constraints, as well as the hallucination and accuracy concerns
+identified in literature review section where LLMs generate plausible but
+incorrect SQL queries.
+
 \begin{verbatim}
@@
 \end{verbatim}
-
-\paragraph{Prompt Engineering Strategy}
-
-As outlined above the \textsc{OmniSQL} prompt was used as base for
-\textsc{Natural}'s prompt together with a . \textsc{Natural} uses more explicit instructions
-compared to DAIL-SQL. Key differentiations include the inclusion of similarity
-scores to give the model the ability to weight the samples itself and
-
-The prompt emphasizes precision and chain-of-thought reasoning. 
-This aims to address common LLM issues with over-generating information or missing
-query constraints, as well as the hallucination and accuracy concerns
-identified in literature review section where LLMs generate plausible but
-incorrect SQL queries.
 

### 2025-10-25 thesis/implementation/index.tex
Self refinement

@@
 through a markdown parser (\texttt{pulldown-cmark}) which parses the model
-output and looks for the code-block fence characters \texttt{\`\`\` }.
-
+output and looks for the code-block fence characters \texttt{```}.
 Subsequently all candidates are processed in reversed order and the first full
@@
 mechanisms in the subsequent pipeline flow.
+
+\subsubsection{Self-Refinement Mechanism ($\rho$)}
+
+The self-refinement algorithm described in~\ref{design:refinement-function}
+is implemented in \texttt{natural-inference::pipeline::refinement}.
+This implementation corresponds to the $\rho$ function, providing
+automated error correction through execution feedback and iterative
+improvement of generated SQL queries.
+
+\paragraph{Error Correction Through Execution Feedback}
+
+The \texttt{Refinement::optimize} method takes a \texttt{RawQueryCandidate} and
+attempts to improve it through targeted prompting. The refinement prompt
+explicitly instructs the model to ``spot any errors in the SQL query, correct
+them'' and provides the original question, schema context, and the candidate
+query that needs improvement.
+
+Contrary to the ICL generation which uses few-shot examples, the
+refinement process focuses on single-query self-correction with explicit
+error-fixing instructions.
+
+\paragraph{Prompt Engineering for Correction}
+
+The refinement prompt uses a structured format similar to ICL generation but
+with key differences: it includes the original question as context, provides
+the candidate query that needs fixing, and uses explicit correction language.
+
+\begin{verbatim}
+    SHOW PROMPT LAYOUT
+\end{verbatim}
+
+\paragraph{Model Resource Management}
+
+As both the ICL implementation and the refinement are ultimately generating
+SQL, they reuse the same underlying \texttt{SqlModel} from the shared pipeline
+context rather than loading separate models. This design choice
+significantly reduces memory requirements but potentially impacts refinement
+quality compared to having dedicated refinement models with different prompt
+conditioning. Due to the limited available hardware the effects of having a
+separate, distinc refinement model could not be verified. \citeauthor{XiYan}
+have implemented multi model generation pipelines in \citeyear{XiYan} and
+achieved promising results which.
+
+\paragraph{Output Processing and Validation}
+
+Similar to ICL generation, refined responses are processed through the same
+\texttt{Markdown} parser to extract the predicted SQL. The system validates each
+extracted candidate using \texttt{QueryCandidate::try_from} and returns the first
+syntactically valid refinement starting from the bottom. This ensures that
+refinement produces executable SQL whilst falling back gracefully if refinement
+fails.
+
+\paragraph{Integration with Pipeline}
+
+The refinement module is integrated into the main pipeline after the ICL
+generator. It processes the \texttt{RawQueryCandidate} and provides a
+self-correction mechanism that improves overall pipeline robustness. Notably both
+the unrefined and the refined query candidates are kept in the candidate set
+for majority voting.

### 2025-10-26 thesis/implementation/index.tex
Write about voting

@@
 \texttt{Markdown} parser to extract the predicted SQL. The system validates each
-extracted candidate using \texttt{QueryCandidate::try_from} and returns the first
-syntactically valid refinement starting from the bottom. This ensures that
-refinement produces executable SQL whilst falling back gracefully if refinement
-fails.
+extracted candidate by trying to parse it using \texttt{QueryCandidate::try\_from}
+and returns the first syntactically valid refinement starting from the bottom.
+This ensures that refinement produces executable SQL whilst falling back
+gracefully if refinement fails.
 
@@
 for majority voting.
+
+\subsubsection{Consensus Voting System ($\nu$)}
+
+The \texttt{natural-inference::pipeline::voting} implements the majority
+function $\nu$ – this implementation closely follows the design and algorithm
+outlined in the section~\ref{design:voting-function}. The function \texttt{voting}
+providing a result-based self-consensus mechanism which takes in the set of
+candidates that were predicted during the generation phase and returns the most
+likely query.
+
+\paragraph{Result-Based Self-Consensus}
+
+\citeauthor{OmniSQL} have shown in \citeyear{OmniSQL} that self-consensus can 
+significantly improve the accuracy of models that are already showing 
+state-of-the-art performance. The result-based self-consensus mechanism is
+executing every query candidate, verifies that it works on the database, and
+loads it's results. By partitioning the available candidates into buckets based
+on their hashed result, queries can be deemed semantically equivalent if they
+end up in the same bucket. After every query has been exeuted and partitioned,
+the algorithm groups the buckets in a hash map \texttt{buckets} with the result
+hash as key and bucket as value.
+
+\paragraph{Largest Group Selection Heuristic}
+
+The heuristic method employed by the voting algorithm is to steer the pipeline
+to agree with itself – if multiple generation attempts yielded the same result,
+these generation attempts are more likely than outliers.
+
+\paragraph{Error Handling and Fallback}
+
+Queries that voting function fails with execution are that contains in an
+\texttt{errors} hash set to provide than being included the calling side all
+candidates fail execution, the voting function fails with an error that
+contains all collected execution errors to provide diagnostic information for
+the calling side.
+
+\paragraph{Limitations and Trade-offs}
+
+The result-based partitioning approach has limitations with subtle errors that
+differ by few rows, as these would be treated as completely different result
+groups. Furthermore executing every query candidate against the real database
+comes along with a significant performance penalty for expensive queries where
+the voting step might incur load onto the database, use significant amounts of
+memory for loading in all data into memory and slow down the voting as the
+entire result needs to be hashed. This system lacks model uncertainty
+calibration for SQL task confidence scoring which could be used as one
+dimension for partitioning.
+
+Another more advanced optimization could be to group by result-schema and row
+count, which still needs to execute the candidate queries partially but does
+not load the actual data from the database into memory. For the scope of this
+thesis and small real-world application scenarios the cost of this stage is
+negligible compared to the llm inference done in previous steps.
+
+\paragraph{Integration with Pipeline}
+
+The voting mechanism represents the last step of the pipeline function 
+and selects the final consensus from all generated and refined candidates, 
+and ultimately returns the output of the \textsc{Natural} system.

### 2025-10-26 thesis/implementation/index.tex
Add more impl content for the vector db impl

@@
 
-\paragraph{Largest Group Selection Heuristic}
+\paragraph{Bucket Selection Heuristic}
 
-The heuristic method employed by the voting algorithm is to steer the pipeline
+The heuristic method employed by the voting algorithm is steering the pipeline
 to agree with itself – if multiple generation attempts yielded the same result,
@@
 
-The result-based partitioning approach has limitations with subtle errors that
-differ by few rows, as these would be treated as completely different result
-groups. Furthermore executing every query candidate against the real database
-comes along with a significant performance penalty for expensive queries where
-the voting step might incur load onto the database, use significant amounts of
-memory for loading in all data into memory and slow down the voting as the
-entire result needs to be hashed. This system lacks model uncertainty
-calibration for SQL task confidence scoring which could be used as one
-dimension for partitioning.
+The result-based partitioning approach has limitations with regards to subtle
+errors that differ by few rows, as these would be treated as completely
+different result buckets. Furthermore executing every query candidate against
+the real database comes along with a significant performance penalty for
+expensive queries where the voting step might incur load onto the database, use
+significant amounts of memory for loading in all data into memory and slow down
+the voting as the entire result needs to be hashed. This system lacks model
+uncertainty calibration for SQL confidence scoring which could be used as
+another dimension for partitioning.
 
@@
 and ultimately returns the output of the \textsc{Natural} system.
+
+\subsection{Supporting Systems and Optimizations}
+
+The supporting systems of \textsc{Natural} are important tools to simplify the 
+actual pipeline development, employing performance optimizations and enable
+rapid development of new approaches and hypotheses. The primary support systems
+are \texttt{Vector} – a vector database abstraction on top of SQLite,
+\texttt{wwl} – a library that implements Wasserstein-Weisfeiler-Leman kernels
+\citep{WWL} and the \texttt{embedding} and \texttt{masking} modules to compare
+and semantically search the embedding space.
+
+\subsubsection{Vector Database}
+
+The vector database implementation \texttt{Vector} is used both during
+sampling to construct the vector db and runtime to run the sample selection
+algorithm. Using SQLite and \texttt{sqlite-vec} to implement cosine search in
+text corpora of up to a few million samples dramatically simplified the
+development of \textsc{Natural} as no manual performance optimizations had to
+be implemnented. SQLite has a second architectural benefit besides relatively
+good performance, which is the portability characteristics of having all
+samples indexed in a single file on disk. This enables swapping out the
+respectively used samples by using another \texttt{.vector} file on disk.
+
+\begin{verbatim}
+let question = "How many ..";
+
+// Initialize vector database for similarity search
+let vector = Vector::infer();
+let model = EmbeddingModel::from_env().expect("failed to load embedding model");
+
+// Query distance indices
+let index = DistanceIndexRow::get("test_db", &vector)?;
+
+// Embed question and perform similarity search
+let selector = Selector::new(&question, None, &model, &vector)?;
+let samples = selection::<4>(
+    &Selector::new(
+        question.as_ref(),
+        None,
+        &model,
+        &vector,
+    )?,
+    &index.into(),
+    &vector,
+)?;
+
+\end{verbatim}
+
+\paragraph{Database Schema}
+
+As the \texttt{.vector} file is still a regular SQLite database, alongside
+indexing tables using embeddings (\texttt{float[4096]} or \texttt{float[2048]})
+regular database tables can be maintained to save data derived at sampling
+time. \texttt{Vector} hosts the following database schema:
+
+\begin{verbatim}
+CREATE TABLE IF NOT EXISTS whitelist (
+  word             text not null primary key,
+  frequency        integer not null,
+  domain_count     integer not null
+);
+
+CREATE TABLE IF NOT EXISTS databases (
+  id               text not null,
+  name             text not null,
+  graph            text not null
+);
+
+CREATE TABLE IF NOT EXISTS distance_indices (
+  db_id            text not null,
+  distances        text not null
+);
+
+CREATE VIRTUAL TABLE IF NOT EXISTS samples USING vec0 (
+  id                 text not null,
+  db_id              text not null,
+
+  question           text not null,
+  question_masked    text not null,
+  question_embedding float[2048],
+
+  sql                text not null,
+  sql_masked         text not null,
+  sql_embedding      float[2048]
+);
+\end{verbatim}
+
+Which maintains the word whitelist for query masking, the database graphs of
+the sample databases and the precomputed database indices. Thus at pipeline
+runtime the database indices and whitelists can be reconstructed through a
+cheap selection from \texttt{Vector} even though they are not indexed through
+embeddings.
+
+The schema of the \texttt{samples} table largely drives the capabilities for
+example selection at runtime, it maintains embeddings of the masked sql query
+and natural language question respectively. Thus after cosine similarity search
+the original query or question can be reconstructed as well as the graph layout
+of the underlying database can be accessed through the \texttt{database} table.
+
+The \texttt{database} and \texttt{database\_indices} table contain \texttt{JSON}
+columns respectively for serializing complex graph and graph index structures
+to the database.

### 2025-11-01 thesis/implementation/index.tex
Embedding and masking section

@@
 to the database.
+
+\subsubsection{Embedding and Masking}
+
+The embedding and masking functionalities in \texttt{natural-sampling} abstract
+batch embedding of masked question and SQL queries using the GPU so that the
+runtime and sampling code can efficiently compute masked embeddings.
+
+\paragraph{Architecture}
+
+The two primary types for embedding plain and masked text are the
+\texttt{SemanticString} and \texttt{MaskedSemanticString} structs which
+encapsulate strings alognside their embedding vectors. The
+\texttt{MaskedSemanticString} version additionally contains the masked and
+unmasked versions of the embedded string. Furthermore they offer
+\texttt{embed\_seq} and \texttt{embed\_chunked} methods which can be used for
+batch computation of embeddings.
+
+\paragraph{Batch Computation}
+
+During sampling, offloading data to the GPUs memory is an expensive operation.
+When sampling hundred thousands or millions of samples, incurring the IO
+roundtrip from CPU to GPU for computing a single embedding is a
+performance bottleneck that increases the sampling time to an order of days
+(eg, when sampling the \textsc{SynSQL}) dataset. Thus batch processing can help to
+minimise the IO roundtrips needed between the CPU and GPU before computing
+embeddings and better utilise the available computing power.
+
+Using the \texttt{embed\_seq} method, a sequence of strings can be embedded and
+a sequence of \texttt{SemanticStrings} is returned. Furthermore through the
+\texttt{embed\_chunked} method, chunks of unrelated data can be embedded in a
+single operation. The method maintains the original chunk layout which makes it
+possible to reconstruct the input semantic seamlessly. This optimization is
+used to embed the questions and sql queries alongside while maintaining clear
+separation in the embedding output.
+
+\begin{verbatim}
+    INSERT DIAGRAM
+\end{verbatim}
+
+% NOTE: Batch embedding computation time down from 200ms per item to
+% approximately 2ms per item when processing 1000-item batches.
+% PUT INTO TEXT/diagram description
+
+\paragraph{Integration}
+
+The \texttt{masking} module tightly integrates wiuth the embedding pipeline.
+During sampling, questions and queries are first masked, then embedded together
+using \texttt{embed\_chunked} and finally combined into
+\texttt{MaskedSemanticString} instances for storage in \texttt{Vector}.
+
+\subsubsection{Wasserstein-Weisfeiler-Leman Kernels}
+
+\subsection{Benchmarking Infrastructure}
+
+\textsc{Natural}'s benchmarking infrastructure is implemented in
+\texttt{natural-benchmark}. This benchmarking infrastructure is enabling the
+development, verification and ultimately deployment of \textsc{Natural}
+pipeline. This section focuses on the respective benchmarking infrastructure
+development, performance optimizations that were required and engineering
+challenges encountered.
+
+Whilst the benchmarking infrastructure is not part of the core pipeline, they
+help to develop and test hypotheses for extensions or design changes
+confidently, compare performance across pipeline versions, models and
+understand the benchmarking data.

### 2025-11-01 thesis/implementation/index.tex
WWL impl

@@
 
+WWL kernels are used to construct the distance index $d \in \mathcal{D}$
+for the $\sigma$ function (see section~\ref{design:selection-function})
+by computing schema distance through graph-based distance metrics. The
+underlying WWL implementation used by \textsc{Natural} is the reference
+implementation of WWL kernels from \citeauthor{WWL}, which is implemented in
+Python. Thus \textsc{Natural} can't directly use the WWL implementation
+as Rust and Python FFI is not directly possible. In order to buy into the
+Python libraries and ecosystem (eg, optimal transport) and utilize the existing
+implementation of the WWL kernels, writing a thin layer of Rust bindings around
+\citeauthor{WWL}'s implementation was deemed most sensible. Using \texttt{pyo3}
+calling Python code is made possible and let's \textsc{Natural} hook into
+existing ecosystems where needed. Due to the design laid out in
+section~\ref{section:design} the WWL implementation is only needed during
+sampling time and not required at runtime as all distances are precomputed and
+stored in a distance index, keeping the runtime portability characteristics of
+using Rust that were discussed above.
+
+% - Schema similarity using Weisfeiler-Leman graph kernel
+% - O(n²) computational complexity for 100+ database schemas
+% - Precomputation during initialization vs dynamic updates
+% - Node labeling strategy: constraints vs data types hierarchy
+% - Distance index caching: initialization time vs query performance
+
+\paragraph{Rust-Python Integration Architecture}
+
+The WWL kernel library used by \textsc{Natural} is implemented using
+\texttt{pyo3} around the existing Python \texttt{wwl} library. The Rust
+bindings are usable through the \texttt{wwl} crate. The \texttt{WWLKernel}
+struct encapsulates the Python module reference and provides type-safe interfaces for
+both categorical and continuous propagation modes through respective methods.
+Using an existing graph library (\texttt{petgraph}) the complexity of the Rust
+bindings could be kept minimal, but required conversion from Rust's
+\texttt{petgraph} graph representation to Python's \texttt{igraph} graph
+representation.
+
+\paragraph{Schema Graph Representation}
+
+The \texttt{natural-graphs} library implements the schema graph representation
+discussed in section~\ref{design:graph-repr} by parsing SQL statements using the
+\texttt{sqlparser} library and constructing an undirected \texttt{petgraph}
+graph instance. Nodes represent tables and columns and edges capture foreign
+key relationships and column ownership. Node labeling is used to encode schema
+constraints hierarchically as outlined in section~\ref{design:graph-repr}.
+
+\paragraph{Propagation Mechanisms}
+
+The \texttt{wwl} crate supports both categorical and continuous Weisfeiler-Leman
+propagation schemes. Categorical propagation operates on discrete node labels
+through iterative label refinement, suitable for constraint-based schema
+comparison. Continuous propagation utilizes node features encoded in matrices,
+enabling similarity computation based on quantitative schema properties like
+column cardinalities or data type frequencies \citep{WWL}. As \textsc{Natural}
+projects the database schema into a graph representation with categorical node
+labels, the categorical propagation scheme is used in \texttt{natural-graphs}
+to determine schema distance. The \texttt{wwl} bindings allow configuration of
+the kernel through the \texttt{KernelConfig} and \texttt{DistanceConfig}
+structs, which provide control over the iteration count (which defaults to 3),
+sinkhorn approximation settings etc.
+
+\paragraph{Distance Computation and Caching}
+
+The pairwise Wasserstein distance is computed between the current database
+schema and each sample database during sampling phase. The distance index is
+a \texttt{HashMap} of a sample database name to it's distance to the current
+database \textsc{Natural} is running on. Caching of this computation is an
+efficient strategy to minimise runtime inference time as the distance index is
+static as long as database schemas are not mutated. \textsc{Natural} is storing
+JSON-serialized distance indices in \texttt{Vector} for fast distance lookups
+at runtime. This design prevents the computational cost of WWL distance
+computations from affecting inference, reducing runtime schema comparison to
+constant-time lookups for all previously indexed database combinations that
+were known during sampling time. Given that the set of sample database and
+target databases are usually fixed, the distance lookup becomes negligible in
+terms of computational cost.
+
+\paragraph{Integration with Example Selection}
+
+The WWL kernel is integrated in the sampling phase, where distance indices are
+computed and cached. Furthermore during example selection ($\sigma$) the
+cached distance index is loaded from \texttt{Vector} for distance lookups.
+
+The distance index is used for weighting the schema compatibility in the
+selection algorithm (see section~\ref{design:selection-function}) to balance
+structural similarity against question-level semantic relevance for improved
+in-context learning effectiveness.
+
 \subsection{Benchmarking Infrastructure}

### 2025-11-08 thesis/appendices/code-examples.tex
Add appendix

--- 2025-10-26 thesis/implementation/index.tex
@@
-\section{Implementation}
+\subsection{Code Examples}
 
-This chapter presents the practical implementation of the \textsc{Natural}
-system, translating the theoretical framework outlined in Chapter
-\ref{section:design} into a working NL2SQL system.
-
-Primary attention is given to the engineering challenges encountered,
-performance bottlenecks identified, and optimization strategies implemented to
-achieve practical deployment viability on consumer hardware.
-The chapter is comprised of a software architecture and infrastructure
-discussion as well as implementation outlines of each pipeline component.
-
-\subsection{Architecture and Infrastructure}
-
-The implementation of the system design phase is split into inference, sampling
-and evaluation code. The runtime code focuses on the actual algorithm
-implementations outlined in section \ref{section:design}, sampling code focuses on
-indexing samples and computing distance indices and the evaluation code runs
-the \textsc{Natural} pipeline on prevalent benchmarks.
-
-This subsection focuses on outlining the scope of each technological component,
-the design rationale behind them and discusses the technology stack
-decisions.
-
-\subsubsection{Software Architecture}
-
-\textsc{Natural} consists of five software components:
-
-\begin{enumerate}
-    \item \texttt{natural-models} – For handling the actual model
-        execution on GPUs for inference and embedding.
-    \item \texttt{natural-graphs} – Graph library for representing
-        database schemas in graphs as well as computing graph similarities and
-        distances.
-    \item \texttt{natural-inference} – The core pipeline implementation using
-        other software components during inference time.
-    \item \texttt{natural-sampling} – The sampling setup that focuses on
-        indexing samples and computing graph distance indices.
-    \item \texttt{natural-benchmark} – The benchmarking setup used to run
-        experiments, continuously evaluate the accuracy of the system and
-        compile statistics.
-\end{enumerate}
-
-% Inter-crate communication patterns and shared data structures
-\begin{verbatim}
-    THIS NEEDS A DIAGRAM TO UNDERSTAND THE SOFTWARE ARCH.
-\end{verbatim}
-
-This separation of concerns emerged as the split between inference, sampling
-and evaluation code required shared fundamentals like model execution and graph
-representation which in turn improves testability and maintainability. Furthermore
-separating the inference, sampling and evaluation code yields a smaller
-footprint when embedding the inference code into databases.
-
-Challenges associated with this multi component architecture are mostly tied to
-dependency management and increased build complexity although these can be
-mitigated through \texttt{cargo}'s workspace support.
-
-\subsubsection{Resource Management Strategy}
-
-Given the constrained 24 gigabytes of VRAM capacity of the RTX 3090 not all
-available model sizes can be used. While 14B model variants theoretically work,
-the need for an embedding model typically exhausts the VRAM unless using
-steep quantization formats. The \textsc{OmniSQL} 7B models were shown to have
-an average performance degradation of only 0.3\% compared to their larger 14B
-counterparts by \citeauthor{OmniSQL} in \citeyear{OmniSQL}. On the spider test
-dataset the 7B model surprisingly outperformed its 14B counterpart by 0.6\%.
-Due to the unclear performance gains of the 14B model, the significant increase
-in inference time and the limited VRAM available, this thesis focuses on the 7B
-variant of \textsc{OmniSQL}.
-
-As an embedding model is required for doing semantic search of samples in the
-pipeline a small companion model is required to embed both the sample datasets
-(\textsc{SynSQL}, \textsc{Bird} and \textsc{Spider}) during sampling time and
-the user provided natural language question during inference. The Qwen3 series
-embedding models ranked first place on the MTEB multilingual leaderboard. With
-the Q8\_0 quantized version of the 8B model consumes 8 gigabytes of VRAM which
-is not possible to fit into 24 gigabytes of VRAM when accounting for KV-Cache
-requirements. Therefore, instead of choosing a heavily quantized version of the
-8B variant (ie, Q4\_K\_M or below), the 4B model with Q8\_0 shows similar
-performance characteristics on simple use cases while offering a significantly
-smaller memory footprint at 2 gigabytes.
-\textcolor{red}{https://huggingface.co/Qwen/Qwen3-Embedding-8B cite}.
-
-As models have significant loading times, the system loads models globally and 
-at startup and hands around references using atomic reference counting
-(\texttt{std::sync::Arc}) to ensure that models are not loaded twice, can be
-reused between different inference calls and GPU memory is not exhausted which
-would lead to a program crash.
-
-\subsubsection{Technology Stack Decisions}
-
-The choice of programming language, inference frameworks, and supporting
-libraries implies the system's performance and deployment characteristics,
-as well as development velocity. This section analyzes the key decisions
-made for \textsc{Natural}, discussing their trade-offs between research
-flexibility and production readiness, performance optimization and development
-speed, and ecosystem maturity versus cutting-edge capabilities.
-
-The critical decisions are the programming language, the model
-inference framework, and the vector similarity search solution. Each
-decision was evaluated against the constraints of limited hardware resources
-(24GB VRAM), the need for database integration capabilities, and the
-requirement for reproducible research outcomes.
-
-\paragraph{Language and Ecosystem}
-
-The most apparent and impactful decision is likely the language and ecosystem
-choice made. Viable languages for implementing natural language processing and
-machine learning heavy systems are Python, R, Julia, Rust, C / C++ and Java as
-well as other general purpose languages.
-
-While interpreted languages like Python, R and Julia tend to be significantly
-easier to use for rapid prototyping and approach validation due to their loose
-type systems and great scientific ecosystem, they come with serious drawbacks
-with regards to deployability, speed and robustness compared to compiled and
-strongly typed languages.
-
-Languages like Java, C / C++ and Rust offer greater stability at runtime,
-better interoperability into other programs (eg, database extensions) and better
-resource utilization they represent an interesting trade-off between performance
-optimization and portability vs. speed of iteration and research ecosystems.
-
-R and Julia are inferior to Python when it comes to adoption, machine learning
-frameworks and natural language processing. Java requires a Java Virtual
-Machine (JVM) at runtime which yields worse portability than C / C++ and Rust
-while offering little advantage over interpreted languages like Python. Thus
-Python, C / C++ and Rust emerge as strong contenders for the implementation of
-\textsc{Natural}. C and C++ have the primary downside that they are prone to
-program crashes and memory safety issues whereas Rust resolves most of these
-downsides while maintaining similar performance, memory management and
-portability characteristics.
-
-Given that performance plays a critical role when developing machine learning
-systems with limited access to hardware, the path to a potential production
-deployments of NL2SQL is easier and the language ecosystem offers bindings
-for the most notable scientific libraries, Rust offers a great value
-proposition for ML systems at the cost of development speed.
-
-\paragraph{Inference Framework}
-
-For local model inference, two primary frameworks emerged: llama.cpp (FFI), 
-Candle (Rust-native). While Rust-native frameworks generally offer better 
-portability, llama.cpp provided a better balance between performance, ecosystem 
-maturity, and implementation simplicity.
-
-llama.cpp provides out-of-the-box optimizations with comprehensive GGUF 
-quantization support, yielding \textcolor{red}{25-30 tokens per second} on
-consumer grade RTX 3090 hardware. It offers high-level abstractions for model
-loading and tokenization and sampling. The mature GGML ecosystem and advanced
-quantization schemes (like \texttt{Q4\_K\_M}, \texttt{Q8\_0}) justified the FFI
-complexity.
-
-Candle, even though it offered comparable raw inference speeds, required
-rather extensive low-level implementation work and proved incompatibility with
-scenarios where the resulting binary is embedded into a database (e.g. PostgreSQL).
-The combination of manual tensor management, more convoluted quantization support,
-and deployment constraints made llama.cpp a better alternative for
-production-ready systems.
-
-\paragraph{Similarity Search Framework}
-
-For lightweight similarity search SQLite in combination with it's
-\texttt{sqlite-vec} extension is a sensible option. Especially for smaller data
-loads introducing the complexity of \texttt{faiss} and comparable vector search
-solutions like \texttt{qdrant} outweighs their speed benefits.
-
-The \texttt{sqlite-vec} extension advertises to be a ``fast enough'' vector
-search solution, allowing for reduced complexity and compatibility with
-graphical database interfaces that support SQLite to inspect the embedding
-space.
-
-\paragraph{Trade-offs}
-
-Overall the development overhead of choosing Rust over Python for the
-implementation phase was noticeable; Rust's machine learning and natural
-language processing frameworks are less advanced, porting research code written
-in python from other papers (like \cite{OmniSQL}) turned out to be non trivial.
-This decision likely doubled the time needed for the implementation, but in
-turn provides a clear path for the algorithms in this thesis to be
-productionized. The outcome of the implementation has significantly better
-performance and portability characteristics than using Python would have
-allowed for.
-
-\subsection{Pipeline Implementation}
-
-This subsection describes the system design realisation of each core pipeline
-component ($\sigma$, $\phi$, $\pi$, $\rho$, $\nu$) into working code, their
-algorithmic details, performance characteristics and trade-offs made.
-
-\subsubsection{Example Selection Engine ($\sigma$)}
-
-The example selection system is implemented using a multi-dimensional similarity
-scoring system that identifies the most relevant samples that were previously
-indexed for subsequent in-context learning. The component addresses the
-fundamental challenge of selecting contextually appropriate examples from large
-training corpora whilst balancing semantic relevance and structural compatibility.
-
-\paragraph{Similarity Computation}
-
-The selection mechanism combines three distinct similarity measures implemented
-in the selection module in \texttt{natural-sampling::selection}. The 
-\texttt{selection} algorithm computes semantic similarity through cosine
-distance of the masked question embedding. Structural SQL similarity is
-computed via measuring the cosine distance of the masked SQL embedding, and
-schema compatibility using WWL kernel distance from the \texttt{natural-graphs}
-component. The embeddings are computed using the Qwen3-Embedding model (4B)
-using the query masking algorithm in \texttt{natural-sampling::masking}.
-
-\paragraph{Weighting Strategy and Performance Characteristics}
-
-As this algorithm is aware of three distinct ways to measure sample similarity
-(masked question, masked query, structural). The implementation employs
-empirically optimized weights defined as constants: 70\% for semantic question
-similarity and 30\% for SQL structural similarity. Additionally, the final
-scoring combines 70\% sample-level similarity with 30\% schema-level
-compatibility.
-
-The selection algorithm yields a maximum of 32 candidates (\texttt{TOP\_K =
-32}) to limit computational overhead whilst maintaining selection quality.
-Vector similarity queries are performed using SQLite with the \texttt{sqlite-vec}
-extension, achieving sub-second retrieval times for vector databases exceeding
-2,000,000 samples.
-
-\paragraph{Implementation Architecture}
-
-Using the \texttt{Selector} struct the embedding computation and representation
-is encapsulated from the actual selection algorithm. Thus via
-\texttt{Selector::new} the consuming code can compute the masked embeddings and
-use them subsequently to run example selection.
-
-The \texttt{selection} algorithm implementation follows a two-stage approach,
-split into initial candidate retrieval through vector search (using the
-\texttt{Vector} struct and SQLite) and subsequent reweighting using the above
-described weights and a precomputed WWL distance index from
-\texttt{natural-graphs}.
-
-\subsubsection{Schema Subsetting System ($\phi$)}
-
-The schema subsetting system is implemented in
-\texttt{natural-inference::pipeline::subsetting} as
-\texttt{SchemaSubsetter} struct. The core algorithm is the \texttt{optimize}
-method (lines 21-47) which performs query validation to determine
-which tables are crucial for the query candidates provided.
-
-\paragraph{Query Validation}
-
-Query validation employs a trial-and-error approach where for each query
-candidate and table in the schema, an in-memory SQLite database
-is created using \texttt{Connection::open\_in\_memory()}. Subsequently the
-connection is initialized using the schema with all tables in the schema except
-the current one.
-
-Once the connection is ready, the \texttt{SchemaSubsetter} prepares every
-query candidate against this reduced schema. If preparation fails, this
-indicates the table removed is crucial for the query, thus it gets added to the
-list of crucial tables. This process is repeated for all query-table
-combinations to build a minimal schema containing only essential tables for the
-execution of the set of query candidates provided to the subsetting algorithm.
-
-\paragraph{Performance and Trade-off Characteristics}
-
-As a new in-memory database for each table-query combination is created, this
-leads to $O(queries \times tables)$ complexity. \textcolor{red}{For schemas
-with XXX+ tables and multiple query candidates, this results in XXX-XXXms
-processing time compared to 10-15ms for heuristic approaches.}
-
-However, this execution-based approach provides superior correctness guarantees since heuristic
-approaches cannot determine whether candidates will execute successfully in real
-database environments. Using a real SQLite in memory instance both the correctness of
-query candidates and the actually referenced set of tables can be known prior
-to actual query execution.
-
-\paragraph{Pipeline Integration}
-
-The \texttt{SchemaSubsetter} is used prior to prompting the model using ICL to
-ensure that the model context is used efficiently and attention is given to the
-relevant parts of the schema. The \texttt{optimize} method returns a
-\texttt{Schema} object containing only the tables identified as crucial through
-execution testing which is in turn processed by the subsequent pipeline stages
-like generation and refinement.
-
-\subsubsection{Query Projection ($\pi$)}
-
-The ICL module is implementing the query projection algorithm described in
-section~\ref{design:projection-function}. It is implemented in the
-\texttt{ICLGenerator} struct and wraps the \texttt{SqlModel} struct from
-\texttt{natural-models}. Using llama.cpp it runs model inference using a
-prompt optimized for in-context-learning with \textsc{OmniSQL}.
-
-\paragraph{In-Context Learning}
-
-The \texttt{ICLGenerator::generate} method implements few-shot prompting
-with relevance filtering. Relevance filtering refers to removing all selected
-samples with a similarity of less than 0.5 (this value was emperically derived
-from evaluations). Thus only semantically or structurally similar samples are
-actually provided to the model.
-
-The prompt is constructed based on an adapted version of the \textsc{OmniSQL}
-format \citep{OmniSQL}: task overview, sql schema, filtered examples with
-similarity scores, explicit instructions, and the target question.
-The biggest differentiation to the prompt of \citeauthor{OmniSQL} is the
-example section including similarity. For the actual SQL query presentation the
-code representation prompt format is used, inspired by DAIL-SQL
-\citep{DAIL-SQL}. Every example includes the masked question, similarity score,
-and formatted SQL query for clarity.
-
-\paragraph{Prompt Engineering Strategy}
-
-As outlined above the \textsc{OmniSQL} prompt was used as base for
-\textsc{Natural}'s prompt together with a code representation prompt.
-\textsc{Natural} uses more explicit instructions compared to DAIL-SQL. Key
-differentiations include the inclusion of similarity scores to give the model
-the ability to weight the samples itself.
-
-This prompt steers the model towards precision and chain-of-thought reasoning. 
-It addresses apparent LLM issues like verbosity, over complexity or missing
-query constraints, as well as the hallucination and accuracy concerns
-identified in literature review section where LLMs generate plausible but
-incorrect SQL queries.
-
-\begin{verbatim}
-EXAMPLE OF THE PROMPT
-\end{verbatim}
-
-\paragraph{Model Integration and Performance}
-
-The \texttt{SqlModel} is wrapped around \texttt{llama-cpp-2} bindings and
-loaded globally at startup to avoid a 30-60 second initialization times per
-query. Using the \texttt{prompt} method from \texttt{natural-models}
-tokenization, inference, and decoding with configurable PromptParams for
-context size and generation limits is handled automatically.
-
-\paragraph{Output Processing and Validation}
-
-As \textsc{OmniSQL} was finetuned to output its thoughts and predictions using
-markdown a markdown postprocessing module is needed, as well as a module to
-identify whether a SQL query is syntactically valid.
-
-To extract possible candidates all generated responses are post-processed
-through a markdown parser (\texttt{pulldown-cmark}) which parses the model
-output and looks for the code-block fence characters \texttt{```}.
-Subsequently all candidates are processed in reversed order and the first full
-valid query is returned as potential candidate. The reversing of processing
-order is needed as the model outputs it's thoughts top-to-bottom as a markdown
-document with the most likely answer usually being output at the end. This
-approach ensures that a returned query candidates are executable and
-ensured to contain valid SQL. Thus \textsc{Natural} can aid the
-difficulty of generating perfect SQL queries by acknowledging
-limitations from large language models and implementing recovery and refinement
-mechanisms in the subsequent pipeline flow.
-
-\subsubsection{Self-Refinement Mechanism ($\rho$)}
-
-The self-refinement algorithm described in~\ref{design:refinement-function}
-is implemented in \texttt{natural-inference::pipeline::refinement}.
-This implementation corresponds to the $\rho$ function, providing
-automated error correction through execution feedback and iterative
-improvement of generated SQL queries.
-
-\paragraph{Error Correction Through Execution Feedback}
-
-The \texttt{Refinement::optimize} method takes a \texttt{RawQueryCandidate} and
-attempts to improve it through targeted prompting. The refinement prompt
-explicitly instructs the model to ``spot any errors in the SQL query, correct
-them'' and provides the original question, schema context, and the candidate
-query that needs improvement.
-
-Contrary to the ICL generation which uses few-shot examples, the
-refinement process focuses on single-query self-correction with explicit
-error-fixing instructions.
-
-\paragraph{Prompt Engineering for Correction}
-
-The refinement prompt uses a structured format similar to ICL generation but
-with key differences: it includes the original question as context, provides
-the candidate query that needs fixing, and uses explicit correction language.
-
-\begin{verbatim}
-    SHOW PROMPT LAYOUT
-\end{verbatim}
-
-\paragraph{Model Resource Management}
-
-As both the ICL implementation and the refinement are ultimately generating
-SQL, they reuse the same underlying \texttt{SqlModel} from the shared pipeline
-context rather than loading separate models. This design choice
-significantly reduces memory requirements but potentially impacts refinement
-quality compared to having dedicated refinement models with different prompt
-conditioning. Due to the limited available hardware the effects of having a
-separate, distinc refinement model could not be verified. \citeauthor{XiYan}
-have implemented multi model generation pipelines in \citeyear{XiYan} and
-achieved promising results which.
-
-\paragraph{Output Processing and Validation}
-
-Similar to ICL generation, refined responses are processed through the same
-\texttt{Markdown} parser to extract the predicted SQL. The system validates each
-extracted candidate by trying to parse it using \texttt{QueryCandidate::try\_from}
-and returns the first syntactically valid refinement starting from the bottom.
-This ensures that refinement produces executable SQL whilst falling back
-gracefully if refinement fails.
-
-\paragraph{Integration with Pipeline}
-
-The refinement module is integrated into the main pipeline after the ICL
-generator. It processes the \texttt{RawQueryCandidate} and provides a
-self-correction mechanism that improves overall pipeline robustness. Notably both
-the unrefined and the refined query candidates are kept in the candidate set
-for majority voting.
-
-\subsubsection{Consensus Voting System ($\nu$)}
-
-The \texttt{natural-inference::pipeline::voting} implements the majority
-function $\nu$ – this implementation closely follows the design and algorithm
-outlined in the section~\ref{design:voting-function}. The function \texttt{voting}
-providing a result-based self-consensus mechanism which takes in the set of
-candidates that were predicted during the generation phase and returns the most
-likely query.
-
-\paragraph{Result-Based Self-Consensus}
-
-\citeauthor{OmniSQL} have shown in \citeyear{OmniSQL} that self-consensus can 
-significantly improve the accuracy of models that are already showing 
-state-of-the-art performance. The result-based self-consensus mechanism is
-executing every query candidate, verifies that it works on the database, and
-loads it's results. By partitioning the available candidates into buckets based
-on their hashed result, queries can be deemed semantically equivalent if they
-end up in the same bucket. After every query has been exeuted and partitioned,
-the algorithm groups the buckets in a hash map \texttt{buckets} with the result
-hash as key and bucket as value.
-
-\paragraph{Bucket Selection Heuristic}
-
-The heuristic method employed by the voting algorithm is steering the pipeline
-to agree with itself – if multiple generation attempts yielded the same result,
-these generation attempts are more likely than outliers.
-
-\paragraph{Error Handling and Fallback}
-
-Queries that voting function fails with execution are that contains in an
-\texttt{errors} hash set to provide than being included the calling side all
-candidates fail execution, the voting function fails with an error that
-contains all collected execution errors to provide diagnostic information for
-the calling side.
-
-\paragraph{Limitations and Trade-offs}
-
-The result-based partitioning approach has limitations with regards to subtle
-errors that differ by few rows, as these would be treated as completely
-different result buckets. Furthermore executing every query candidate against
-the real database comes along with a significant performance penalty for
-expensive queries where the voting step might incur load onto the database, use
-significant amounts of memory for loading in all data into memory and slow down
-the voting as the entire result needs to be hashed. This system lacks model
-uncertainty calibration for SQL confidence scoring which could be used as
-another dimension for partitioning.
-
-Another more advanced optimization could be to group by result-schema and row
-count, which still needs to execute the candidate queries partially but does
-not load the actual data from the database into memory. For the scope of this
-thesis and small real-world application scenarios the cost of this stage is
-negligible compared to the llm inference done in previous steps.
-
-\paragraph{Integration with Pipeline}
-
-The voting mechanism represents the last step of the pipeline function 
-and selects the final consensus from all generated and refined candidates, 
-and ultimately returns the output of the \textsc{Natural} system.
-
-\subsection{Supporting Systems and Optimizations}
-
-The supporting systems of \textsc{Natural} are important tools to simplify the 
-actual pipeline development, employing performance optimizations and enable
-rapid development of new approaches and hypotheses. The primary support systems
-are \texttt{Vector} – a vector database abstraction on top of SQLite,
-\texttt{wwl} – a library that implements Wasserstein-Weisfeiler-Leman kernels
-\citep{WWL} and the \texttt{embedding} and \texttt{masking} modules to compare
-and semantically search the embedding space.
-
-\subsubsection{Vector Database}
-
-The vector database implementation \texttt{Vector} is used both during
-sampling to construct the vector db and runtime to run the sample selection
-algorithm. Using SQLite and \texttt{sqlite-vec} to implement cosine search in
-text corpora of up to a few million samples dramatically simplified the
-development of \textsc{Natural} as no manual performance optimizations had to
-be implemnented. SQLite has a second architectural benefit besides relatively
-good performance, which is the portability characteristics of having all
-samples indexed in a single file on disk. This enables swapping out the
-respectively used samples by using another \texttt{.vector} file on disk.
+\subsection{Vector Database API}\label{appendix:vector-db-api}
 
@@
 )?;
-
 \end{verbatim}
 
-\paragraph{Database Schema}
-
-As the \texttt{.vector} file is still a regular SQLite database, alongside
-indexing tables using embeddings (\texttt{float[4096]} or \texttt{float[2048]})
-regular database tables can be maintained to save data derived at sampling
-time. \texttt{Vector} hosts the following database schema:
+\subsubsection{Vector Database Schema}\label{appendix:vector-schema}
 
@@
 
-Which maintains the word whitelist for query masking, the database graphs of
-the sample databases and the precomputed database indices. Thus at pipeline
-runtime the database indices and whitelists can be reconstructed through a
-cheap selection from \texttt{Vector} even though they are not indexed through
-embeddings.
+\subsubsection{Benchmark Traits}\label{appendix:benchmark-traits}
 
-The schema of the \texttt{samples} table largely drives the capabilities for
-example selection at runtime, it maintains embeddings of the masked sql query
-and natural language question respectively. Thus after cosine similarity search
-the original query or question can be reconstructed as well as the graph layout
-of the underlying database can be accessed through the \texttt{database} table.
+\begin{verbatim}
+pub trait Benchmark {
+    type Error: From<eyre::Report> + Debug;
+    type Database: BenchmarkDatabase<Error = Self::Error>;
 
-The \texttt{database} and \texttt{database\_indices} table contain \texttt{JSON}
-columns respectively for serializing complex graph and graph index structures
-to the database.
+    fn name() -> &'static str;
+
+    fn databases(&self) -> impl Iterator<Item = &Self::Database>;
+    fn database(&self, id: impl AsRef<str>) -> Option<&Self::Database>;
+
+    fn tests(&self) -> impl Iterator<
+      Item = &<Self::Database as BenchmarkDatabase>::Test
+    >;
+
+    fn test(&self, id: impl AsRef<str>) -> Option<
+      &<Self::Database as BenchmarkDatabase>::Test
+    >;
+}
+
+pub trait BenchmarkDatabase {
+    type Error: From<eyre::Report> + Debug;
+    type Test: BenchmarkTest;
+
+    fn id(&self) -> &str;
+    fn connect(&self) -> Result<Database, Self::Error>;
+    fn tests(&self) -> impl Iterator<Item = &Self::Test>;
+    fn test(&self, id: impl AsRef<str>) -> Option<&Self::Test>;
+}
+
+pub trait BenchmarkTest {
+    fn id(&self) -> &str;
+    fn db_id(&self) -> &str;
+
+    fn name(&self) -> &str;
+    fn question(&self) -> &str;
+    fn query(&self) -> &str;
+}
+\end{verbatim}

### 2025-11-08 thesis/appendices/prompts.tex
Add appendix

\subsection{Prompts}

\subsection{Inference Prompt}\label{appendix:inference-prompt}

\begin{verbatim}
EXAMPLE OF THE PROMPT
\end{verbatim}

\subsection{Refinement Prompt}\label{appendix:refinement-prompt}

\begin{verbatim}
EXAMPLE OF THE PROMPT
\end{verbatim}

### 2025-11-08 thesis/implementation/index.tex
Use appendix

@@
 \textsc{Natural}'s prompt together with a code representation prompt.
-\textsc{Natural} uses more explicit instructions compared to DAIL-SQL. Key
+\textsc{Natural} uses more explicit instructions
+(see~\ref{appendix:inference-prompt}) compared to DAIL-SQL. Key
 differentiations include the inclusion of similarity scores to give the model
@@
 incorrect SQL queries.
-
-\begin{verbatim}
-EXAMPLE OF THE PROMPT
-\end{verbatim}
 
@@
 with key differences: it includes the original question as context, provides
-the candidate query that needs fixing, and uses explicit correction language.
-
-\begin{verbatim}
-    SHOW PROMPT LAYOUT
-\end{verbatim}
+the candidate query that needs fixing, and uses explicit correction language
+(see \ref{appendix:refinement-prompt}).
 
@@
 respectively used samples by using another \texttt{.vector} file on disk.
-
-\begin{verbatim}
-let question = "How many ..";
-
-// Initialize vector database for similarity search
-let vector = Vector::infer();
-let model = EmbeddingModel::from_env().expect("failed to load embedding model");
-
-// Query distance indices
-let index = DistanceIndexRow::get("test_db", &vector)?;
-
-// Embed question and perform similarity search
-let selector = Selector::new(&question, None, &model, &vector)?;
-let samples = selection::<4>(
-    &Selector::new(
-        question.as_ref(),
-        None,
-        &model,
-        &vector,
-    )?,
-    &index.into(),
-    &vector,
-)?;
-
-\end{verbatim}
+See \ref{appendix:vector-db-api} for reference.
 
@@
 regular database tables can be maintained to save data derived at sampling
-time. \texttt{Vector} hosts the following database schema:
-
-\begin{verbatim}
-CREATE TABLE IF NOT EXISTS whitelist (
-  word             text not null primary key,
-  frequency        integer not null,
-  domain_count     integer not null
-);
-
-CREATE TABLE IF NOT EXISTS databases (
-  id               text not null,
-  name             text not null,
-  graph            text not null
-);
-
-CREATE TABLE IF NOT EXISTS distance_indices (
-  db_id            text not null,
-  distances        text not null
-);
-
-CREATE VIRTUAL TABLE IF NOT EXISTS samples USING vec0 (
-  id                 text not null,
-  db_id              text not null,
-
-  question           text not null,
-  question_masked    text not null,
-  question_embedding float[2048],
-
-  sql                text not null,
-  sql_masked         text not null,
-  sql_embedding      float[2048]
-);
-\end{verbatim}
+time. \texttt{Vector} hosts the database schema described
+in \ref{appendix:vector-schema}.
 

### 2025-11-08 thesis/implementation/index.tex
Write about the benchmark system

@@
 
-Whilst the benchmarking infrastructure is not part of the core pipeline, they
-help to develop and test hypotheses for extensions or design changes
-confidently, compare performance across pipeline versions, models and
-understand the benchmarking data.
+Whilst \textsc{Natural}'s benchmarking infrastructure is not part of the core
+pipeline, it helps to develop and test hypotheses for extensions or design
+changes confidently, compare performance across pipeline versions, models and
+understand the benchmark performance.
+
+\subsubsection{Execution-Based Evaluation System}
+
+The \texttt{natural-benchmark} CLI evovles around the concept of executions
+which are single benchmark runs against a specific benchmark dataset
+(defaulting to \textsc{Spider}). An execution is the set of tests that have
+been executed using a version of \textsc{Natural} against a benchmark. The
+interface of \texttt{natural-benchmark} offers multiple options to create
+executions and subsequentlty understand \textsc{Natural}'s performance:
+
+\begin{enumerate}
+    \item Running new benchmarks via: \\ \texttt{natural-benchmark new}
+    \item Continuing a halted execution via: \\ \texttt{natural-benchmark continue --execution <id>}
+    \item Comparing performance on previous (partial) executions via: \\ \texttt{natural-benchmark compare-to --previous <id>}
+    \item Computing statistics on past executions via: \\ \texttt{natural-benchmark stats --execution <id>}
+\end{enumerate}
+
+While benchmarking \texttt{natural-benchmark} maintains a local SQLite database
+maintaining a history of past performance, pipeline failures etc. for future
+introspection as well as comparison of approaches.
+
+\subsubsection{Cross-Dataset Validation}
+
+In order to measure the performance of \textsc{Natural} against multiple
+benchmarking datasets, the benchmarking and execution system in
+\texttt{natural-benchmark} must be generalize across one dataset. The
+benchmarking setup achieves this through a set of rust traits (eg, type
+characteristics) to model benchmarking datasets.
+
+The traits \texttt{Benchmark}, \texttt{BenchmarkDatabase} and
+\texttt{BenchmarkTest} allow to abstract the file system layout of different
+evaluation datasets. Thus when implementing benchmark support against
+\textsc{Spider} the implementation absatracts the fact that \textsc{Spider} is
+using SQLite for test databases, and uses a JSON file to store the questions.
+The file system layout of \textsc{Spider} and \textsc{Bird} is largely similar.
+
+This abstracted benchmarking system (see \ref{appendix:benchmark-traits})
+allows integrating further benchmarks in the future (eg, \textsc{Spider2} which
+relies on cloud databases like Snowflake) while maintaining the same ergonomics
+and tooling across benchmark datasets (eg, recording and indexing all test
+executions in a local database, recording pipeline logs etc).
+
+\subsubsection{Metric Computation}
+
+The benchmarking infrastructure computes the two metrics \textsc{Execution
+Accuracy} and \textsc{Exact Match} found in prevalent benchmark leaderboards
+(eg, \textsc{Spider} and \textsc{Bird}) through a Rust port of the Python
+reference implementation found in \textsc{Spider}. \textsc{Spider} determines
+the \textsc{Execution Accuracy} by comparing the result sets of two queries
+based on possible row and column permutations as well as optional order
+sensitivity. The \textsc{Exact Match} metric is determined by exact equivalence
+of the ground truth query $\omega_{ground}$ and the candidate query $\omega$.
+
+\subsubsection{Benchmarking Challenges}
+
+Two primary issues emerged while developing the benchmarking infrastructure for
+\textsc{Natural}:
+
+While loading result sets from the provided test database, a faulty
+code path failed to parse query result with types other than
+\texttt{Text}. This casued empty result sets to be returned as query results
+from both the ground truth query and the candidate query, resulting in a false
+positive, skewing the \textsc{EA} metric significantly upwards.
+
+Furthermore frequent pipeline failures of \textsc{Natural} made it hard to get
+compute reliable performance metrics as context window constraints, out
+of memory issues and CUDA issues caused a signficiant number of test
+cases to fail, making subsequently computed metrics unreliable as only
+a subset of the dataset was included in the results.

### 2025-11-08 thesis/implementation/index.tex
Fix lang

@@
 to agree with itself – if multiple generation attempts yielded the same result,
-these generation attempts are more likely than outliers.
+these generation attempts are more likely to be accurate the others.
 
@@
 
-Queries that voting function fails with execution are that contains in an
-\texttt{errors} hash set to provide than being included the calling side all
-candidates fail execution, the voting function fails with an error that
-contains all collected execution errors to provide diagnostic information for
+Queries that fail execution are contained in an                                                                                                                                                   │ │
+\texttt{errors} hash set rather than being included in voting buckets. If all                                                                                                                     │ │
+candidates fail execution, the voting function fails with an error that                                                                                                                           │ │
+contains all collected execution errors to provide diagnostic information to                                                                                                                      │ │
 the calling side.

### 2025-11-09 thesis/implementation/index.tex
Challenges

@@
 a subset of the dataset was included in the results.
+
+\subsection{Engineering Challenges and Lessons Learned}
+
+The implementation approach of \textsc{Natural} highlighted significant
+challenges in terms of practicability that were independent of the algorithmic
+design and theoretical problems. While \textsc{Natural} turned out to
+demonstrate the approaches recommended in the system design section, the
+development process was significantly slowed down due to constrained hardware
+availability, technology stack decisions and research methodology. The overall
+system architecture turned out positive, but the timeline was noticably expanded.
+
+This section reflects on the most impactful challenges that were encountered
+during the implementation of \textsc{Natural} and analyzes how hardware
+limitations influenced design decisions, confidence in the performance of
+\textsc{Natural} and the overall impact on development velocity. These insights
+are transferrable beyond \textsc{Natural} and are applicable to any
+production-grade implementation of research algorithms.
+
+\subsubsection{Hardware Constraints and Development Velocity}
+
+Likely the most significant circumstance that affected development velocity was
+the constrained access to high performance hardware. The hardware available
+during development time was the RTX 3090 with 24GB of VRAM. This limitation
+fundamentally shaped the development process and research methodology as the
+iteration speed was severly slowed down.
+
+\paragraph{Benchmarking and Evaluation Bottlenecks}
+
+A full benchmark execution of \textsc{Natural} against the \textsc{Spider} test
+dataset required around \textcolor{red}{12-36 hours} depending on the number of
+candidates generated and refinement algorithm used during the benchmark. 
+This made continuous validation during development a non trivial task,
+requiring extensive evaluation phases between times of active development.
+
+Testing new hypotheses, validating algorithmic changes, experimenting with the
+implementation approaches, comparing different configurations and sampling
+strategies therefore became a multi-week processes rather than the rapid
+iteration typically desired in research environments.
+
+The prolonged evaluation cycles created a cascading effect on development
+velocity. Rather simple adjustments that would normally be validated within
+hours required days of computational time, effectively preventing what should
+have been normal experimentation. The hardware constraints forced a rather
+conservative approach to experimentation, where each change needed to be
+carefully considered before committing to multiple days of benchmarking.
+
+\paragraph{Memory Management and CUDA Issues}
+
+Frequent CUDA out-of-memory (OOM) errors made the evaluation interpretation non
+trivial as accuracy scores became inaccurate due to pipeline failures. This persistet as a
+
+second development challenge throughout the implementation phase. These memory exhaustion issues
+occurred rather unpredictable during both development and benchmarking,
+depending on the database schema and question asked. As \textsc{Natural} uses a
+code representation prompt, the database schema is inlined into the prompt
+during inference. If \textsc{Natural} is executed against large databases, even
+with schema subsetting, the context window can be exceeded or the KV-Cache can
+exhaust the rest of the available GPU memory (typically around 2-4 gigabytes)
+after models have been loaded. This required extensive trial-and-error
+debugging to identify the root causes and implement workarounds.
+
+\subsubsection{Technology Stack Trade-offs}
+
+During the implementation of \textsc{Natural}, several critical technology
+stack decisions fundamentally impacted both development velocity and the
+final system characteristics. The two most consequential decisions were the
+choice of programming language and the model deployment strategy, each
+presenting distinct trade-offs between research efficiency and production
+readiness.
+
+\paragraph{Programming Language: Rust vs Python}
+
+The decision to implement \textsc{Natural} in Rust rather than Python
+represents perhaps the most impactful architectural choice made during the
+implementation phase. Choosing Rust presented the opportunity to port the
+research system into production environments soon after, but came at the cost
+of having to port existing benchmarking infrastructure and not being able to
+exactly reuse the inference code used by \textsc{OmniSQL}. This decision
+significantly increased implementation time, likely doubling the development
+effort compared to what would have been required in Python.
+
+Using Python would have provided immediate access to existing machine
+learning infrastructure and could have leveraged existing implementations by
+previous researchers directly in the implementation of \textsc{Natural}.
+
+The rather mature Python ecosystem for NLP and ML would have eliminated the
+need to develop abstractions for model inference, vector database operations,
+and benchmark evaluation.
+
+However, the Rust implementation as is provides significant benefits for production
+deployment and system integration. The resulting system can be embedded
+directly into database systems such as PostgreSQL through extensions,
+providing a path toward true production deployment that would be impractical
+with Python. The ability to compile an entire NL2SQL system into a single,
+standalone binary offers superior portability characteristics.
+
+\paragraph{Model Deployment: Local vs Cloud-Based Inference}
+
+Using local and open source models reemphasized the limited local hardware
+access as using a cloud provider like OpenAI or Anthropic for inference would
+have singificantly speeded up the benchmarking time. As a primary research goal
+of this thesis is to explore the open source capabilities of NL2SQL systems,
+choosing a proprietary inference service was deeemed unviable.
+
+\paragraph{Integration Complexity and Ecosystem Maturity}
+
+The machine learning ecosystem of Rust proved to be singificantly less mature
+than anticipated, requiring extensive custom development and research for
+functionality that would be available off-the-shelf in prevalent Python packages.
+The integration with Python libraries for specialized components like the
+Wasserstein-Weisfeiler-Leman kernels required rather complex FFI code using
+\texttt{pyo3}, further adding architectural complexity and potential stability
+concerns.
+
+Despite these challenges, the resulting architecture demonstrates that multi
+language approaches can be viable when different components have distinct
+requirements. The Python integration was isolated to the sampling phase,
+preserving the runtime portability characteristics of the core Rust
+implementation while still leveraging existing, specialized libraries where
+appropriate.

### 2025-11-09 thesis/implementation/index.tex
Finish impl section

@@
 Frequent CUDA out-of-memory (OOM) errors made the evaluation interpretation non
-trivial as accuracy scores became inaccurate due to pipeline failures. This persistet as a
-
-second development challenge throughout the implementation phase. These memory exhaustion issues
-occurred rather unpredictable during both development and benchmarking,
-depending on the database schema and question asked. As \textsc{Natural} uses a
-code representation prompt, the database schema is inlined into the prompt
-during inference. If \textsc{Natural} is executed against large databases, even
-with schema subsetting, the context window can be exceeded or the KV-Cache can
-exhaust the rest of the available GPU memory (typically around 2-4 gigabytes)
-after models have been loaded. This required extensive trial-and-error
-debugging to identify the root causes and implement workarounds.
+trivial as accuracy scores became inaccurate due to pipeline failures. This
+persistet as a second development challenge throughout the implementation
+phase. These memory exhaustion issues occurred rather unpredictable during both
+development and benchmarking, depending on the database schema and question
+asked. As \textsc{Natural} uses a code representation prompt, the database
+schema is inlined into the prompt during inference. If \textsc{Natural} is
+executed against large databases, even with schema subsetting, the context
+window can be exceeded or the KV-Cache can exhaust the rest of the available
+GPU memory (typically around 2-4 gigabytes) after models have been loaded. This
+required extensive trial-and-error debugging to identify the root causes and
+implement workarounds.
 
@@
 significantly increased implementation time, likely doubling the development
-effort compared to what would have been required in Python.
-
-Using Python would have provided immediate access to existing machine
-learning infrastructure and could have leveraged existing implementations by
-previous researchers directly in the implementation of \textsc{Natural}.
-
-The rather mature Python ecosystem for NLP and ML would have eliminated the
-need to develop abstractions for model inference, vector database operations,
-and benchmark evaluation.
+effort compared to what would have been required in Python. Using Python would
+have provided immediate access to existing machine learning infrastructure and
+could have leveraged existing implementations by previous researchers directly
+in the implementation of \textsc{Natural}. The rather mature Python ecosystem
+for NLP and ML would have eliminated the need to develop abstractions for model
+inference, vector database operations, and benchmark evaluation.
 
@@
 appropriate.
+
+\subsubsection{Lessons Learned and Future Recommendations}
+
+The implementation of \textsc{Natural} showed multiple practical challenges
+associated with researching and implementing NL2SQL systems. The approach used
+in this thesis highlights several areas of consideration for future reserch
+projects in this field.
+
+\paragraph{Hardware Infrastructure Requirements}
+
+Research of modern NL2SQL systems induces steep hardware requirements onto
+researchers. Priority should be given to securing access to significantly
+powerful hardware and computing infrastructure for faster iteration cycles.
+
+While the 24GB VRAM constraint of the RTX 3090, may be reasonable for most
+consumer grade applications, it proved insufficient for efficient iteration in
+this field, especially given the involement of large language models.
+Systems with 48GB+ of VRAM would have enabled the usage of larger, more capable
+models without the quantization compromises that may impact system accuracy.
+More modern hardware (even on the consumer grade level) would allow for a
+tremendous benchmarking speed up (eg, using a 5090 would have resulted in
+nearly 200\% increase in speed).
+
+More critically, the availability of multiple GPUs would have enabled parallel
+experimentation and benchmarking, which could transform multi-day iteration
+cycles into more manageable timeframes. The ability to run concurrent
+experiments with different configurations, models, or algorithmic approaches
+would dramatically accelerate the research and development process and enable
+more thorough experimental validation.
+
+\paragraph{Technology Stack}
+
+The implications of choosing between Rust and Python underlines the importance
+of technology choices with research goals and time constraints. For pure
+research projects focused on algorithmic development and experimentation,
+Python's ecosystem and extensive existing implementations provide clear
+advantages that likely outweigh the deployment benefits of compiled languages
+like Rust.
+
+However, for projects with a clear path to production deployments or those
+intended to bridge academic research with real-world applications, the
+additional development overhead of systems programming languages like Rust can
+be justified. Overall, hardware availability weighs significantly more than
+language choices when it comes to the implementation timeline.
+
+Future implementations in this field should likely consider a two-staged
+approach: prototyping and algorithm validation in Python to leverage existing
+tools and accelerate iteration, followed by a production port in a systems
+language like Rust once the algorithmic approach is validated and stable.
+
+\paragraph{Development Methodology Recommendations}
+
+The challenges encountered suggest several methodological improvements for
+future projects:
+
+Implementing benchmarking and validation infrastructure early in the
+development process is crucial when doing algorithmic development to gain an
+understanding of the relative improvement or setback of different approaches.
+
+Establishing performance baselines and using incremental validation procedures
+can help ensure that development effort is focused on improvements that in fact
+improve performance.
+
+The long running benchmarks made it difficult to validate whether
+individual changes were beneficial, leading to uncertainty about the
+effectiveness of specific optimizations. In situations with highly constrained
+compute resources, a representative subset of prevalent benchmarking datasets
+like \textsc{Spider} and \textsc{Bird} may be defined to determine the relative
+improvement between approaches before executing a full benchmark run.
+
+Finally, maintaining detailed logs and metrics throughout the development
+process showed to be crucial when unexpected errors occur. Identifying root
+causes of pipeline failures was rather trivial due to having an advanced
+benchmarking setup with log collection.
+
+\paragraph{Research Impact and Production Readiness}
+
+Despite all challenges encountered in the implementation phase, the Rust
+implementation of \textsc{Natural} demonstrate that it is possible to port
+research algorithms from academic environments into production-ready systems.
+The resulting implementation provides a strong foundation for real-world
+deployments in database environments, which would be significantly more
+challenging with a Python-based research prototype.
+
+This production readiness ensures that all algorithmic contributions can have
+practical impact beyond pure academic evaluation, potentially influencing how
+NL2SQL systems are deployed in real database environments. A simple
+recommendation between research velocity and production readiness can not be
+given as it ultimately depends on the intended impact and longevity of the
+research contributions and implementation.

### 2025-11-15 thesis/implementation/index.tex
Fix

@@
 
-Queries that fail execution are contained in an                                                                                                                                                   │ │
-\texttt{errors} hash set rather than being included in voting buckets. If all                                                                                                                     │ │
-candidates fail execution, the voting function fails with an error that                                                                                                                           │ │
-contains all collected execution errors to provide diagnostic information to                                                                                                                      │ │
-the calling side.
+Queries that fail execution are contained in an \texttt{errors} hash set rather
+than being included in voting buckets. If all candidates fail execution, the
+voting function fails with an error that contains all collected execution
+errors to provide diagnostic information to the calling side.
 

### 2025-11-16 thesis/theoretical-foundations/index.tex
Add theory sections i need to write
Rewrite the theory section intro

--- 2025-10-01 thesis/chapters/03-theoretical-foundations/index.tex
@@
 
-This section serves as an introduction to the theoretical concepts utilized in this thesis. It is meant to establish a
-baseline of understanding for the design, implementation and evaluation phases. It briefly covers relational database theory
-(incl. SQL), machine learning fundamentals and their subsequent application for large language models.
+This section introduces the theoretical concepts needed for understanding
+\textsc{Natural}'s design and implementation sections. It covers relational
+database theory, machine learning fundamentals, large language models, natural
+language processing, embeddings and graph theory.
+
+These foundations directly support the system's core functions: example
+selection ($\sigma$), schema subsetting ($\phi$), query projection ($\pi$),
+self-refinement ($\rho$), and voting ($\nu$).
 
@@
 \subimport{}{llms.tex}
-
-%\subsection{Problem Decomposition}
-
-%\subsection{Requirements}
-
-% Conceptual Design
-
-%\subsection{System Design}
-
-%\subsubsection{Architecture Design}
-
-%\subsection{Technical Implementation Strategies}
+\subimport{}{nlp.tex}
+\subimport{}{graphs.tex}

### 2025-11-16 thesis/theoretical-foundations/machine-learning.tex
Write ml section

@@
 
-Machine learning represents the algorithmic foundation for state-of-the-art NL2SQL systems, enabling systems to learn how natural
-language queries translate into SQL. Understanding these foundations is essential for implementing and optimizing large language
-model-based NL2SQL approaches.
+Machine learning represents the algorithmic foundation for state-of-the-art
+NL2SQL systems, enabling systems to learn how natural language queries
+translate into SQL. Understanding these foundations is essential for
+implementing and optimizing large language model-based NL2SQL approaches.
 
@@
 
-Neural networks are the computational concept behind contemporary NL2SQL approaches, representing complex capabilities through
-compositions of simpler operations. A neural network consists of interconnected computational units (called neurons) organized
-in layers, where each connection has an associated (and learnable) weight and bias parameters.
+Neural networks are the computational concept behind contemporary NL2SQL
+approaches, representing complex capabilities through compositions of simpler
+operations. A neural network is comprised of interconnected computational units
+(called neurons) organized in layers, where each connection has an associated
+(and learnable) weight and bias parameters.
 
-The fundamental computation paradigm of neural network outputs is called forward propagation, which is applying the respective weights ($W$)
-and biases ($b$) to the input parameter ($a$) and transforming it using a (non-linear) activation function $f$:
+The fundamental computation paradigm of neural network outputs is called
+forward propagation, which is applying the respective weights ($W$) and biases
+($b$) to the input parameter ($a$) and transforming it using a (non-linear)
+activation function $f$:
 
@@
 
-where $W \in \mathbb{R}^{m \times n}$ is the weight matrix, $a \in \mathbb{R}^{n}$ is the input vector, $b \in \mathbb{R}^{m}$ is
-the bias vector, and $f$ is an activiation function (e.g. ReLU, sigmoid, or tanh). Using this function repeatedly, propagates information
-through the neural network.
+where $W \in \mathbb{R}^{m \times n}$ is the weight matrix, $a \in
+\mathbb{R}^{n}$ is the input vector, $b \in \mathbb{R}^{m}$ is the bias vector,
+and $f$ is an activiation function (e.g. ReLU, sigmoid, or tanh). Using this
+function repeatedly, propagates information through the neural network.
 
-The most important architectural components for language processing models include:
+The most important architectures for simple natural language
+processing networks include:
 
 \begin{itemize}
-    \item \textbf{Feedforward Networks} — Which can process fixed-size inputs through successive linear transformations and activations,
-          this architecture is particularly suitable for classification and regression problems within NL2SQL systems.
-    \item \textbf{Recurrent Networks (RNNs/LSTMs)} — Which can handle sequential data of variable-length by maintaining hidden states
-          that capture dependencies, enabling processing of natural language sequences of arbitrary length.
-    \item \textbf{Embedding Layers} — Which map discrete tokens (e.g. words, characters, or subwords) to a dense vector representation,
-          thus providing the foundation for subsequent neural language processing.
+    \item \textbf{Feedforward Networks} — Which can process fixed-size inputs
+        through successive linear transformations and activations, this
+        architecture is particularly suitable for classification and regression
+        problems within NL2SQL systems.
+    \item \textbf{Recurrent Networks (RNNs/LSTMs)} — Which can handle
+        sequential data of variable-length by maintaining hidden states that
+        capture dependencies, enabling processing of natural language sequences
+        of arbitrary length.
+    \item \textbf{Embedding Layers} — Which map discrete tokens (e.g. words,
+        characters, or subwords) to a dense vector representation, thus
+        providing the foundation for subsequent neural language processing.
 \end{itemize}
 
-The ability of neural networks to learn hierarchical representations through multiple layers makes them particularly well-suited for
-the complex translation required in NL2SQL systems.
+The ability of neural networks to learn hierarchical representations through
+multiple layers makes them particularly well-suited for the complex translation
+required in NL2SQL systems.
 
-\subsubsection{Sequence-to-Sequence Models}
+\subsubsection{Learning Approaches}
 
-Sequence-to-sequence (Seq2Seq) models established the foundational architecture for neural NL2SQL approaches \citep{Seq2SQL, SQLNet}.
-These models which rely on RNNs and LSTMs addressed the fundamental challenge of mapping variable-length input sequences (natural
-language queries) to variable-length output sequences (SQL statements).
-
-The Seq2Seq model architecture is composed out of three essential components:
+Broadly speaking, there are three fundamental approaches to training a machine
+learning model, also known as the process of ``learning'':
 
 \begin{enumerate}
-    \item \textbf{Encoder} — Process the input natural language sequence $x_1, x_2, \ldots, x_n$ into a sequence of hidden
-          representations $h_1, h_2, \ldots, h_n$, capturing the semantic content of the query.
-    \item \textbf{Decoder} — Generates the output SQL sequence $y_1, y_2, \ldots, y_m$ autoregressively, where each token is
-          predicted based on previous outputs and encoder representations.
-    \item \textbf{Attention} — Allows the decoder to selectively focus on relevant portions of the input sequence, addressing
-          the information bottleneck of fixed-size context vectors.
+    \item \textbf{Supervised Learning} – The process of training a model
+        using a labeled dataset of input and output pairs. This learning
+        approach is quite commonly used in models used for NL2SQL which were
+        trained on question and answer pairs (natural language question and
+        corresponding SQL query). This learning approach aims to minimise the
+        loss between predicitions and ground truth.
+    \item \textbf{Unsupervised Learning} – Includes learning patterns that work
+        with unlabeled datasets and use approaches like clustering, language
+        modeling and heuristics to infer structure and yield predictions.
+        Unsupervised learning is most appropriate when there is no clear,
+        singular answer and the domain provides large datasets that are not
+        feasible to label manually (eg, dataset gathered through webscraping
+        the internet).
+    \item \textbf{Reinforcement Learning} – The process of learning through
+        reward signals, commonly known as the reward function. Applied to
+        NL2SQL systems this could transfer to training a model's SQL generation
+        capabilities based on execution feedback. Historically this learning
+        pattern is not widely used in NL2SQL approaches.
 \end{enumerate}
+
+\subsubsection{Attention Mechanisms}
+
+Attention mechanisms allow machine learning models to selectively focus on
+relevant parts of their input by learning to weight different sections of the
+input sequence based on its relevance to the current computation. While RNNs
+processed sequences sequentially, attention enables direct modeling of
+relationships between any positions in a sequence, which enabled the
+transformer architecture to become prevalent for natural language processing
+\citep{AttentionIsAllYouNeed}.
+
+Attention can be computed by retrieving the weighted sum of values $V$
+based on the compatibility between queries $Q$ and keys $K$:
+
+\begin{equation}
+\text{Attention}(Q, K, V) = \text{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right)V
+\end{equation}
+
+where $d_k$ is the dimensionality of the key vectors, used for scaling to
+prevent softmax saturation which would incurr small gradients. The softmax
+operation converts compatibility scores into a probability distribution,
+determining how much attention to pay to each value.
+
+This concept proved to be a foundational advancement to transformer
+architectures seen in natural language processing and subsequently recent
+NL2SQL research. Transformers extended the concept of attention into
+self-attention and multi-head attention patterns that underpin todays large
+language models.
+
+\subsubsection{Transformer Architecture}
+
+The transformer architecture, introduced by \citeauthor{AttentionIsAllYouNeed}
+in \citeyear{AttentionIsAllYouNeed}, introduced the model architecture found in 
+modern large language models \citep{AttentionIsAllYouNeed}. Unlike earlier
+sequence to sequence models, transformers process are capable of processing
+entire sequences of inputs in parallel by using a self-attention mechanism.
+This enabled more efficient training on massive datasets and superior
+representation of of complex dependencies in input sequences.
+
+\paragraph{Self-Attention Mechanism}
+
+Self-attention allows each position in a sequence to be weighted to all other
+positions in the sequence, thus resulting in a representations that can capture
+contextual relationships between input tokens. For an input sequence of tokens
+$x_1, \ldots, x_n$, the self-attention mechanism computes:
+
+\begin{equation}
+\text{SelfAttention}(X) = \text{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right)V
+\end{equation}
+
+where $Q = XW_Q$, $K = XW_K$, and $V = XW_V$ are learned linear projections of
+$X$ into the query, key and value representations. This mechanism enables
+transformers to weigh the importance of different tokens while encoding each
+position. This maintains semantic and syntactic relationships which are crucial
+for understanding natural language queries in NL2SQL systems.
+
+\paragraph{Multi-Head Attention}
+
+Multi-head attention extended the self-attention concept by computing multiple
+attention operations in parallel, allowing the model to put focus to different
+aspects of the input at the same time:
+
+\begin{equation}
+\text{MultiHead}(Q, K, V) = \text{Concat}(\text{head}_1, \ldots, \text{head}_h)W^O
+\end{equation}
+
+where each $head$ computes:
+
+\begin{equation}
+\text{head}_i = \text{Attention}(QW_i^Q, KW_i^K, VW_i^V)
+\end{equation}
+
+Each attention head can specialize in capturing a different type of semantic
+relationship. For example one head might focus on syntactic structure (e.g., linking SQL
+keywords to their arguments), while others might capture semantic relationships
+between works (e.g., List all x where y). This multi-attention mechanism is
+essential for the complex tasks which require capturing multiple different
+semantic relationships between words.

### 2025-11-16 thesis/theoretical-foundations/llms.tex
LLM theory

@@
 
-% TODO: Complete LLM theoretical foundations section (essential for LLM-based approach)
-% - Transformer architecture fundamentals
-% - Attention mechanisms and self-attention
-% - Pre-training vs fine-tuning paradigms
-% - In-context learning and few-shot learning
-% - Prompt engineering and instruction following
-% - Model scaling laws and emergent capabilities
-% - Limitations: hallucination, context windows, reasoning
+Large language models (LLMs) represented a paradigm shift in natural language
+processing, demonstrating state of the art capabilities in understanding, reasoning and
+text generation. These models form the core of most modern NL2SQL systems
+like \textsc{XiYan-SQL}, \textsc{DAIL} and others. They have a wide language understanding as they have been trained on large text corpora,
+are exposed to different domains of traning data and have an inherent
+understanding of the SQL syntax as they have been exposed to it during
+pre-training. This shifted the approach while developing NL2SQL systems to
+focus on in-context learning methods instead of training dedicated models from
+scratch.
 
-% TODO: Connect to design section
-% - How LLMs enable natural language understanding for SQL generation
-% - Role in query projection function À
-% - Self-refinement capabilities Á
-% - Voting and consensus mechanisms ½
+\paragraph{Feedforward and Residual Connections}
+
+Every layer in a transformer combines a multi-head attention with position-wise
+feedforward networks:
+
+\begin{equation}
+\text{FFN}(x) = \text{GELU}(xW_1 + b_1)W_2 + b_2
+\end{equation}
+
+Residual connections and layer normalization stabilize training of complex networks:
+
+\begin{equation}
+    \text{Output} = x + \text{Sublayer}(\text{LayerNorm}(x))
+\end{equation}
+
+where $\text{Sublayer}$ represents either multi-head attention or the
+feedforward network. These components enable transformers to scale to billions
+of parameters while maintaining training stability.
+
+\paragraph{Positional Encoding}
+
+Since transformers are able to process sequences in parallel, positional
+encodings inject information about token positions. Original transformers use
+sinusoidal functions, while modern LLMs often employ learned positional
+embeddings or relative position encodings. This positional information is
+inherently important for NL2SQL systems, where the order of generated SQL
+clauses (e.g., \texttt{SELECT} must come before before \texttt{WHERE}) carries
+significant semantic meaning.
+
+\subsubsection{Pre-training and Fine-tuning}
+
+Large language models are trained in a two-stage process that separates the
+training of general language understanding from from task-specific
+capabilities.
+
+\paragraph{Pre-training}
+
+During pre-training LLMs are exposed to vasts amounts of unlabeled text
+gathered from the internet, books etc. Using self-supervised objectives, no
+manual labeling of the datasets is required. The most common pre-training
+objective is causal language modeling, where a model learns to predict the next
+token given a previous context:
+
+\begin{equation}
+\mathcal{L}_{\text{LM}} = -\sum_{t=1}^{T} \log P(x_t \mid x_{<t}; \theta)
+\end{equation}
+
+This objective engrains general language patterns, common knowledge,
+reasoning capabilities, and even programming knowledge into models 
+when code is included in the training datasets. Models like GPT, LLaMA, and Qwen
+are pre-trained using this objective, aiming to predict coherent
+continuations of text sequences.
+
+\paragraph{Fine-tuning}
+
+Fine-tuning is the process of adapting a pre-trained model, for example for
+solving NL2SQL tasks, by continuing model training on a labeled dataset of
+question-answer pairs (eg, natural language to sql). During fine-tuning, the
+model can learn to:
+
+\begin{itemize}
+    \item Map natural language to database schema elements (also known as schema-linking)
+    \item Generate SQL queries following proper grammar
+    \item Compose more complex SQL queries involving joins, aggregations, and subqueries
+    \item Understand domain-specific terminology (depending on the dataset)
+\end{itemize}
+
+Fine-tuning usualy require significantly smaller datasets (thousands to
+hundreds of thousands of examples) compared to pre-training (billions to
+trillions of tokens), largely relying on the natural language understanding
+capabilities acquired during the pre-training phase.
+
+Models like \textsc{OmniSQL} demonstrate that smaller fine-tuned models 
+can outperform significantly larger LLMs for NL2SQL tasks using a fraction of
+the parameters \citep{OmniSQL}.
+
+\subsubsection{In-Context Learning}
+
+In-context learning (ICL) is the process of teaching a model how to perform a
+certain task by providing examples in its prompt, without updating models
+parameters. Thus ICL and Fine-tuning differ by their timing: runtime vs learning.
+
+This capability emerged with sufficiently large models which have large enough
+context windows to fit in examples of solutions for similar problems. This
+method was shown to be very effective by \citeauthor{DAIL} in \citeyear{DAIL},
+improving the performance of general purpose LLMs as well as fine-tuned LLMs
+when being presented with relevant examples \citep{DAIL}. The combination of
+fine-tuned models like \textsc{OmniSQL} with ICL is therefore a promising
+research for NL2SQL systems.

### 2025-11-16 thesis/theoretical-foundations/llms.tex
Extend the llm theory

@@
 
-Large language models (LLMs) represented a paradigm shift in natural language
-processing, demonstrating state of the art capabilities in understanding, reasoning and
-text generation. These models form the core of most modern NL2SQL systems
-like \textsc{XiYan-SQL}, \textsc{DAIL} and others. They have a wide language understanding as they have been trained on large text corpora,
-are exposed to different domains of traning data and have an inherent
-understanding of the SQL syntax as they have been exposed to it during
-pre-training. This shifted the approach while developing NL2SQL systems to
-focus on in-context learning methods instead of training dedicated models from
-scratch.
+Large language models (LLMs) are a recent advancement built on top of the
+transformer architecture for natural language processing, demonstrating state
+of the art capabilities in understanding, reasoning and text generation. These
+models form the core of most modern NL2SQL systems like \textsc{XiYan-SQL},
+\textsc{DAIL} and others. They have a wide language understanding as they have
+been trained on large text corpora, are exposed to different domains of traning
+data and have an inherent understanding of the SQL syntax as they have been
+exposed to it during pre-training. This shifted the approach while developing
+NL2SQL systems to focus on in-context learning methods instead of training
+dedicated models from scratch.
 
-\paragraph{Feedforward and Residual Connections}
+\subsubsection{Feedforward and Residual Connections}
 
@@
 
-\paragraph{Positional Encoding}
+\subsubsection{Positional Encoding}
 
@@
 
-Large language models are trained in a two-stage process that separates the
+Large language models are trained in a two-fold process that separates the
 training of general language understanding from from task-specific
@@
 Fine-tuning is the process of adapting a pre-trained model, for example for
-solving NL2SQL tasks, by continuing model training on a labeled dataset of
-question-answer pairs (eg, natural language to sql). During fine-tuning, the
-model can learn to:
-
-\begin{itemize}
-    \item Map natural language to database schema elements (also known as schema-linking)
-    \item Generate SQL queries following proper grammar
-    \item Compose more complex SQL queries involving joins, aggregations, and subqueries
-    \item Understand domain-specific terminology (depending on the dataset)
-\end{itemize}
+solving NL2SQL tasks, by continuing model training on a labeled dataset
+(eg, natural language question to expected sql query). During fine-tuning, the
+model can learn to map natural language questions to expected SQL queries when
+being presented with a question and a database schema.
 
@@
 research for NL2SQL systems.
+
+\subsubsection{Prompt Engineering}
+
+Prompt engineering is the process of structuring the input to LLMs in a way
+where they adhere to desired behaviours. For NL2SQL systems previous research
+has found diverging effectiveness between different example, schema and
+question presentation mechanisms \citep{DAIL, OmniSQL}.
+
+\subsubsection{Chain-of-Thought Reasoning}
+
+Prompting or training models to output their chain-of-thought encourages models 
+to generate reasoning traces before producing their final answer to a question
+or input. For NL2SQL, this usually involves generating explanations of query
+logic, analysing the input database schema, and planning the query structure
+before generating a query.
+
+Fine-tuned models like \textsc{OmniSQL} are fine-tuned to produce
+chain-of-thought output during generation in markdown which enables debugging
+model behaviour and understanding the query generation approach.
+
+\subsubsection{Model Limitations}
+
+Understanding the limitations of large language models is essential for
+designing robust NL2SQL systems.
+
+\paragraph{Hallucination}
+
+As LLMs are just predicting a sequence of output tokens, they have no inherent
+understanding of the actual problem domain and solution space they are
+presented with. Thus in NL2SQL systems, LLMs might generate invalid SQL queries
+which might reference non existent tables, columns or relationships, output
+invalid syntax or generate promising but semantically invalid queries.
+
+These flaws are inherent to a text based representation and are non
+recoverable, although through candidate validation and self-refinement
+approaches most obviously invalid queries can be recovered. Detecting semantic
+invalidity is a hard research problem on its own.
+
+\paragraph{Context Window Limitations}
+
+As the underlying transformer architecture is still bound to finite context
+windows, extensive database schemas or long lists of examples can exceed the
+window, thus causing the query generation to fail.

### 2025-11-23 thesis/theoretical-foundations/graphs.tex
Write about graph theory

@@
-\subsection{Graph Theory Fundamentals}
+\subsection{Graph Theory}
 
-% TODO: Add graph theory basics section (needed for WWL kernel approach)
-% - Basic graph definitions: vertices, edges, labeled graphs
-% - Graph representations and adjacency structures
-% - Graph similarity metrics and kernels
-% - Weisfeiler-Lehman graph features
-% - Wasserstein distance and optimal transport theory
-% - Applications to database schema representation
+Graph theory is a mathematical domain for working with structures and
+relationships. It is highly relevant for databases and NL2SQL systems as
+modeling relationships in databases maps intuitively to graphs. This section
+introduces the theoretical framework of graphs, graph kernels and graph
+similarity which are required foundations for the example selection algorithm
+of \textsc{Natural}.
 
-% TODO: Connect to design section
-% - How database schemas map to graph structures
-% - WWL kernel computation for schema similarity
-% - Distance indexing for efficient retrieval
-% - Integration with example selection function Ã
+\subsubsection{Graph Fundamentals}
+
+A graph is noted as $G = (V, E)$ in its simplest form and consists of vertices
+$V$ and edges $E \subseteq V \times V$ which are connections between vertices.
+Labeled graphs are an extension to this concept which associate labels with
+vertecies and edges, encoding semantic information such as vertex types,
+attributes, or constraints.
+
+\subsubsection{Graph Kernels}
+
+Graph kernels are functions $k: \mathcal{G} \times \mathcal{G} \rightarrow
+\mathbb{R}$ which are measuring the structural similarity between two graphs:
+
+\begin{equation}
+k(G_1, G_2) = \langle \phi(G_1), \phi(G_2) \rangle
+\end{equation}
+
+where $\phi: \mathcal{G} \rightarrow \mathcal{H}$ maps graphs to a feature
+space $\mathcal{H}$, which allows measuring similarity in the feature space,
+rather than using raw graph matching to compute the similarity. Thus, on a high
+level, graph kernels may be used to identify common super structures in graphs.
+
+\subsubsection{Weisfeiler-Lehman Algorithm}
+
+The Weisfeiler-Lehman (WL) algorithm is a graph kernel algorithm which
+iteratively refines node labels based on the neighborhood structure in the
+graph \citep{WWL}:
+
+\begin{enumerate}
+    \item \textbf{Sorting} — Represent vertices as a sorted list of neighbors ($L_v$)
+    \item \textbf{Compression} — Compute the hash of $L_v$ for every vertex ($h(L_v)$)
+    \item \textbf{Relabeling} — Relabel every vertex $v$ with $h(L_v)$ as its new node label
+\end{enumerate}
+
+This algorithm is repeated for $w$ iterations (typically $w = 3-5$, due to
+deminishing returns). After the iterative vertex label computation, labels
+encode the $w$-hop neighborhood structure. Subsequently a label histogram can
+be used as the graph-level feature vector, enabling efficient structural
+comparison.
+
+% https://ethz.ch/content/dam/ethz/special-interest/bsse/borgwardt-lab/documents/slides/CA10_WeisfeilerLehman.pdf
+
+\subsubsection{Wasserstein Distance}
+
+The Wasserstein distance measures the minimum cost of transforming one
+probability distribution into another (also known as earth mover's distance).
+Given two distributions $\mu$ and $\nu$ over space $\mathcal{X}$ with metric $d$:
+
+\begin{equation}
+W_p(\mu, \nu) = \left(\inf_{\gamma \in \Gamma(\mu, \nu)} \int_{\mathcal{X} \times \mathcal{X}} d(x, y)^p \, d\gamma(x, y)\right)^{1/p}
+\end{equation}
+
+where $\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.
+Unlike other distance metrics of probability distributions like the
+Kullback-Leibler divergence, the Wasserstein distance accounts for the
+underlying metric space structure \citep{WWL}.
+
+\subsubsection{Wasserstein Weisfeiler-Leman Kernels}
+
+WWL kernels combine the labels computed using WL with Wasserstein distance to
+measure graph distance \citep{WWL}:
+
+\begin{enumerate}
+    \item Apply $w$ WL iterations to both graphs, producing node label distributions at each iteration
+    \item Compute Wasserstein distance between graphs using Hamming distance
+        (for the categorical features) or Euclidean distance (for continuous
+        features) as the ground metric
+    \item Aggregate distances into a similarity score using a Laplacian kernel:\\
+          $k_{\text{WWL}}(G_1, G_2) = \exp\left(-\lambda  D_W)\right)$
+\end{enumerate}
+
+The resulting graph kernel captures both local neighborhood structures (using
+the WL algorithm) and global distributional properties (using the Wasserstein
+distance) \citep{WWL}. For database schemas, WWL kernels can similar structural
+patterns (table structures, possible joins, constraints) while remaining robust
+to minor structure variations. This makes them particularly well suited for
+NL2SQL example selection algorithms that factor in schema similarity.

### 2025-11-23 thesis/theoretical-foundations/llms.tex
Write about nlp

@@
 
-This objective engrains general language patterns, common knowledge,
+This objective ingrains general language patterns, common knowledge,
 reasoning capabilities, and even programming knowledge into models 

### 2025-11-23 thesis/theoretical-foundations/nlp.tex
Write about nlp

@@
 
-% TODO: Complete NLP theoretical foundations section
-% - Text preprocessing and tokenization
-% - Word embeddings and semantic representations
-% - Language modeling fundamentals
-% - Syntactic and semantic parsing
-% - Named entity recognition and linking
-% - Text similarity metrics (cosine, semantic distance)
-% - Masked language modeling approaches
+Natural language processing (NLP) provides the foundations for working with
+natural language inputs in computer science and mathematics. It allows for text
+understanding and text representation, aswell as text processing and semantic
+feature extraction, capturing meaning and structure. For NL2SQL systems NLP
+enable semantic understanding of natural language queries and semantic search
+using linguistic similarity. Using NLP and text embeddings reference datasets
+can be used for semantic search effectively and accurately.
 
-% TODO: Connect to design section
-% - Role in natural language query understanding
-% - Embedding space construction for example selection
-% - Masking techniques for domain generalization
-% - Semantic similarity computation in function Ã
+\subsubsection{Text Preprocessing and Tokenization}
+
+Before neural networks can process text, raw text must be converted into a
+structured representations that models can process. This transformation is called
+tokenization, which refers to the process of segmenting text into discrete
+units (tokens) for subsequent processing.
+
+\paragraph{Tokenization Approaches}
+
+Tokens are short character chains that allow representing text efficiently as a
+series of tokens by refering to the tokenizers vocabulary. Therefore NLP
+systems mostly employ subword tokenization. Modern tokenization approaches
+balance vocabulary size with representational efficiency. Common approaches
+include \textbf{Byte-Pair Encoding (BPE)}, \textbf{WordPiece} and
+\textbf{SentencePiece}.
+
+These subword tokenization approaches are particularly valuable for NL2SQL
+because database schemas often contain domain-specific terminology, compound
+words, and technical terminology that may not appear in common vocabularies.
+Subword approaches handle unseen words gracefully by decomposing them into
+meaningful tokens.
+
+\paragraph{Tokenization in NL2SQL Contexts}
+
+For NL2SQL systems, tokenization must be capable of both encoding natural
+language queries and SQL queries. SQL presents a challenge in NLP as it relies
+on a structured syntax and usage of special characters (dots, underscores,
+SQL operators, etc.) which are less commonly found in natural language.
+
+\subsubsection{Word Embeddings and Semantic Representations}
+
+Word embeddings are the vector space representation of a tokens where semantic
+meaning is captured through geometric properties. As word embeddings are
+ordinary vectors in a high-dimensional vector space, common vector operations
+can still be performed.
+
+\paragraph{Embedding Functions}
+
+An embedding function maps tokens from a discrete vocabulary $V$ to dense
+vectors in $\mathbb{R}^d$:
+
+\begin{equation}
+e_{w}: V \rightarrow \mathbb{R}^d
+\end{equation}
+
+where $d$ is the embedding dimensionality (usually between 256-8192 for modern
+models). The key property of the vector space $\mathbb{R}^d$ is that
+semantically similar words occupy nearby regions in this vector space, hence
+vector operations and proximity in the vector space allows for computation of
+semantic properties such as semantic distance.
+
+Early embedding methods like Word2Vec and GloVe learned static representations
+where each word has a single vector regardless of its surrounding context.
+Modern contextual embeddings, generate different vectors for the same word
+based on its context, thus capturing meaning more accurately.
+
+\paragraph{Sentence Embeddings}
+
+While word embeddings are useful for determining the relationships between
+words, they are limited in length and complexity. NL2SQL systems require
+representations of entire questions and SQL queries in order to discover
+answers to similar questions asked in the past. Sentence embeddings aggregate
+the token-level meaning into fixed-size vectors in the same vector space
+$\mathbb{R}^d$, thus representing complete sentences or texts in one vector.
+
+\begin{equation}
+    e_{s}: (v_0, .., v_n) \rightarrow \mathbb{R}^d
+\end{equation}
+
+Different aggregation / pooling strategies emerged, like mean pooling, max
+pooling, using a CLS token or learned aggregation. \textsc{Natural} uses
+an embedding model (Qwen3-Embedding-8B) trained specifically for generating
+high-quality sentence representations for subsequent searching.
+
+\subsubsection{Semantic Similarity Metrics}
+
+Quantifying the similarity between two natural language sentences is essential
+for example selection in NL2SQL. This requires that similarity metrics in the
+vector space align with natural understanding of semantic relation.
+
+\paragraph{Cosine Similarity}
+
+Cosine similarity is a measurement using the angle between two vectors,
+to capture their directional similarity while normalizing for magnitude:
+
+\begin{equation}
+s_{\cos} = \frac{u \cdot v}{\|u\| \|v\|} = \frac{\displaystyle\sum_{i=1}^n u_i v_i}{\sqrt{\displaystyle\sum_{i=1}^n u_i^2} \sqrt{\displaystyle\sum_{i=1}^n v_i^2}}
+\end{equation}
+
+Thus cosine similarity ranges from -1 which indicates opposite directions to 1
+representing same direction, with 0 indicating orthogonality of the input
+vectors. This metric is widely used in NLP as it is independent of vector
+magnitude, efficient to compute and well-suited for high-dimensional vector
+spaces.
+
+\paragraph{Cosine Distance}
+
+Cosine distance converts similarity into a distance metric:
+
+\begin{equation}
+d_{\cos}(u, v) = 1 - \text{s}_{\cos}(u, v)
+\end{equation}
+
+Thus cosine distance ranges from 0 for identical direction to 2 for opposite
+direction input vectors. This is satisfying properties desirable for distance
+metrics while preserving the angular relationship of the input vectors $u$ and
+$v$.

### 2025-11-23 thesis/evaluation/index.tex
Benchmarking section outline

% Reference: https://github.com/petavue/NL2SQL-Benchmark/blob/main/results/Report/NL2SQL%20Benchmark%20Report.pdf

\section{Evaluation}

The evaluation phase of \textsc{Natural} builds upon the benchmarking
infrastructure outlined in section~\ref{impl:benchmark}. This section is
evaluating benchmarking data gathered while running \textsc{Natural} in
different configurations on two prevalent benchmarking datasets:
\textsc{Spider} and \textsc{Bird}. A representative baseline is introduced,
aiming to mirror the base-model performance. Subsequently an ablation study is
performed to understand the performance impact of individual pipeline components.

\subsection{Experimental Methodology}

% TODO: Introduction paragraph explaining the systematic evaluation approach

\subsubsection{Test Environment}

The tests are performed on a consumer-grade linux system, as research- or
enterprise-grade hardware was not available. Thus only a subset of datasets and
configurations have been measured, as a single benchmark executions on prevalent
datasets takes 12-36 hours.

\paragraph{Hardware Configuration}

The system used for benchmarking has a NVIDIA RTX 3090 GPU with 24GB VRAM
available, an AMD 9990X3D CPU with 12-cores up to 5.5GHz, 64GB of RAM at
6400mt/s and 1TB of SSD storage at a read speed of 7450 MB/s and a write speed
of 6900 MB/s.

\paragraph{Software Stack}

The software stack running on the system used for benchmarking is NixOS (25.05)
with a recent linux kernel version (6.17.7). CUDA 13.0 is being used for
inference on GPUs and a rust compiler version is 1.91 for compiling
\textsc{Natural}. The models used during benchmarking are \textsc{OmniSQL 7B
Q8\_0} and \textsc{Qwen3 Embedding 8B Q4\_K\_M} for generation and embedding
respectively.

\paragraph{System Configuration}

% TODO: Describe configuration parameters
% - Model loading settings
% - Batch sizes
% - Timeout configurations
% - Any other relevant parameters

\subsubsection{Benchmark Datasets}

The benchmarking datasets used are the \textsc{Spider} and \textsc{Bird}
testdatasets that are prevalently used for evaluation of NL2SQL systems.
Comparable research has frequently published performance metrics
(\textsc{Execution Accuracy} and \textsc{Exact Match}) on these benchmarks
which makes the performance of \textsc{Natural} comparable to other works.

\paragraph{Spider Test Set}

The \textsc{Spider}

% TODO: Describe Spider benchmark
% - Dataset size: 10,181 questions, 5,693 SQL queries
% - 200 databases across 138 domains
% - Difficulty distribution (Easy/Medium/Hard/Extra Hard)
% - Why Spider was chosen
% - Train/dev/test split strategy

\paragraph{Bird Benchmark}
% TODO: Describe Bird benchmark
% - Dataset size: 12,751 pairs across 95 databases (33.4 GB)
% - 37 domains
% - Real-world characteristics (dirty data, external knowledge requirements)
% - Why Bird was chosen as complementary to Spider

\paragraph{Dataset Preprocessing}
% TODO: Describe any preprocessing steps
% - Database preparation
% - Schema extraction
% - Gold standard SQL formatting

\subsubsection{Evaluation Metrics}

\paragraph{Execution Accuracy (EA)}
% TODO: Define execution accuracy
% - Definition: whether query results match ground truth
% - Calculation method
% - Why EA is the primary metric
% - Reference to natural-benchmark implementation

\paragraph{Exact Match (EM)}
% TODO: Define exact match
% - Definition: whether SQL query exactly matches ground truth
% - Calculation method
% - Limitations and interpretation

\paragraph{Performance Metrics}
% TODO: Define performance metrics
% - End-to-end latency (time from NL query to SQL result)
% - Per-component latency (σ, φ, π, ρ, ν timings)
% - Throughput (queries per second)
% - Resource utilization (GPU memory, CPU usage)

\paragraph{Statistical Significance}
% TODO: Describe statistical testing approach
% - Test selection (e.g., paired t-test, McNemar's test)
% - Significance threshold (p < 0.05)
% - Confidence intervals

\subsubsection{Baseline Systems and Comparisons}

\paragraph{Internal Baselines}
% TODO: Describe ablated baselines
% - Zero-shot baseline (π only, no σ, φ, ρ, ν)
% - Partial system variants for ablation studies

\paragraph{State-of-the-Art Comparisons}
% TODO: List SOTA systems for comparison
% - GPT-4 results from literature
% - DIN-SQL, DAIL-SQL, C3 results
% - Other relevant open-source approaches
% - Note: Direct comparison constraints (different models, hardware)

\subsection{Benchmark Performance Analysis}

% TODO: Introduction paragraph presenting overall findings

\subsubsection{Spider Results}

\paragraph{Overall Performance}
% TODO: Present overall EA and EM scores
% - Table: Overall Spider dev set results
% - Comparison to baseline
% - Comparison to SOTA from literature

\paragraph{Performance by Difficulty Level}
% TODO: Break down results by Spider difficulty classification
% - Table: EA/EM by Easy/Medium/Hard/Extra Hard
% - Analysis of performance degradation with difficulty
% - Discussion of which difficulty levels benefit most from system components

\paragraph{Performance by Database}
% TODO: Present per-database analysis
% - Identify databases where system excels
% - Identify databases where system struggles
% - Analyze characteristics that correlate with performance

\paragraph{Performance by SQL Complexity}
% TODO: Analyze results by SQL pattern complexity
% - Simple SELECT queries
% - JOIN operations
% - Nested subqueries
% - Aggregations and GROUP BY
% - Set operations (UNION, INTERSECT, EXCEPT)
% - Complex predicates

\paragraph{Learning Curve Analysis}
% TODO: Analyze impact of training example quantity
% - How does performance change with k-shot examples?
% - Diminishing returns analysis
% - Optimal k value determination

\subsubsection{Bird Results}

\paragraph{Overall Performance}
% TODO: Present overall Bird results
% - Table: Overall Bird dev set results
% - Comparison to baseline
% - Comparison to SOTA from literature (GPT-4: 54.89%, human: 92.96%)

\paragraph{Domain-Specific Performance}
% TODO: Break down by Bird's 37 domains
% - Best-performing domains
% - Worst-performing domains
% - Domain characteristics that impact performance

\paragraph{Dirty Data Handling}
% TODO: Analyze performance on Bird's dirty data cases
% - How does system handle real-world data quality issues?
% - Error patterns related to data quality
% - Comparison to clean data performance

\paragraph{External Knowledge Requirements}
% TODO: Analyze cases requiring external knowledge
% - Identify queries needing domain knowledge
% - System performance on these cases
% - Limitations of pure schema-based approaches

\subsubsection{Comparison with State-of-the-Art}

\paragraph{Quantitative Comparison}
% TODO: Create comprehensive comparison table
% - Table: System vs GPT-4, DIN-SQL, DAIL-SQL, C3, etc.
% - Both Spider and Bird results
% - Note differences in model architecture and hardware

\paragraph{Open-Source vs Proprietary Models}
% TODO: Discuss open-source (OmniSQL 7B) vs proprietary (GPT-4) trade-offs
% - Performance gap analysis
% - Cost and accessibility advantages
% - Privacy and control benefits
% - Deployment feasibility differences

\paragraph{Qualitative Comparison}
% TODO: Beyond metrics, discuss systemic differences
% - System architecture differences
% - Component strategies (how others handle schema subsetting, refinement, etc.)
% - Integration and deployment characteristics

\subsection{Performance Characteristics}

% TODO: Introduction paragraph on performance evaluation methodology

\subsubsection{Latency Analysis}

\paragraph{Per-Component Timing Breakdown}
% TODO: Present timing for each pipeline component
% - Table: Average latency for σ, φ, π, ρ, ν
% - Distribution statistics (median, p50, p95, p99)
% - Identify bottlenecks

\paragraph{End-to-End Query Response Time}
% TODO: Present total pipeline latency
% - Distribution of total response times
% - Comparison to acceptable thresholds (what's "interactive"?)
% - Impact of query complexity on latency

\paragraph{Latency vs Accuracy Trade-offs}
% TODO: Analyze speed-accuracy Pareto frontier
% - Can we reduce latency by limiting refinement iterations?
% - Can we reduce latency by using fewer candidates for voting?
% - Configurable performance profiles (fast mode vs accurate mode)

\paragraph{Bottleneck Identification and Optimization Opportunities}
% TODO: Identify where time is spent
% - Model inference time (largest component?)
% - Embedding search time
% - Graph kernel computation time
% - Database execution time (for refinement)
% - Recommendations for optimization

\subsubsection{Resource Utilization}

\paragraph{GPU Memory Usage}
% TODO: Analyze VRAM consumption
% - Peak memory usage during inference
% - Memory required for model loading
% - Impact of schema size on memory (large schemas → CUDA OOM)
% - Memory optimization strategies employed

\paragraph{CPU and System Memory}
% TODO: Analyze CPU usage patterns
% - CPU utilization during pipeline execution
% - RAM consumption for embeddings, candidate storage
% - I/O patterns (database access, vector index reads)

\paragraph{Disk I/O and Vector Index Performance}
% TODO: Analyze storage subsystem impact
% - SQLite-vec index size
% - Read performance for similarity search
% - Impact of index size on query latency

\paragraph{Power Consumption and Thermal Characteristics}
% TODO: If measured, discuss energy efficiency
% - GPU power draw during inference
% - Comparison to CPU-only approaches
% - Sustainability considerations

\subsection{Qualitative Analysis}

% TODO: Introduction paragraph explaining value of qualitative evaluation beyond metrics

\subsubsection{Case Studies}

\paragraph{Success Cases}
% TODO: Present 3-5 exemplar success cases
% - Full trace: NL query → schema → examples → generated SQL → execution results
% - Analysis of what made these queries succeed
% - Demonstrate system capabilities at their best
% - Show component contributions in action

\paragraph{Failure Cases}
% TODO: Present 3-5 exemplar failure cases
% - Full trace showing where system failed
% - Root cause analysis: schema linking error? Semantic error? Syntactic error?
% - What component(s) failed?
% - Could failure be prevented with different design choices?

\paragraph{Edge Cases and Boundary Conditions}
% TODO: Present interesting edge cases
% - Very ambiguous queries
% - Queries requiring domain knowledge
% - Queries with multiple valid interpretations
% - Queries at the limit of SQL expressiveness
% - How system behaves at boundaries

\newpage

### 2025-11-23 thesis/evaluation/index.tex
More evaluation content

@@
 
-\paragraph{System Configuration}
-
-% TODO: Describe configuration parameters
-% - Model loading settings
-% - Batch sizes
-% - Timeout configurations
-% - Any other relevant parameters
-
 \subsubsection{Benchmark Datasets}
@@
 
-\paragraph{Spider Test Set}
+\paragraph{Spider}
 
-The \textsc{Spider}
+\textsc{Spider} is likely the most prevalently used benchmark for NL2SQL
+systems in contemporary research. The test dataset is comprised of 10,181
+questions and 5693 corresponding SQL queries spread across 200 databases
+spanning 138 domains. Questions in \textsc{Spider} are categorized by
+difficulty (easy, medium, hard, extra hard). Evaluations on \textsc{Spider}
+refer to the test dataset.
 
-% TODO: Describe Spider benchmark
-% - Dataset size: 10,181 questions, 5,693 SQL queries
-% - 200 databases across 138 domains
-% - Difficulty distribution (Easy/Medium/Hard/Extra Hard)
-% - Why Spider was chosen
-% - Train/dev/test split strategy
+\paragraph{Bird}
 
-\paragraph{Bird Benchmark}
-% TODO: Describe Bird benchmark
-% - Dataset size: 12,751 pairs across 95 databases (33.4 GB)
-% - 37 domains
-% - Real-world characteristics (dirty data, external knowledge requirements)
-% - Why Bird was chosen as complementary to Spider
-
-\paragraph{Dataset Preprocessing}
-% TODO: Describe any preprocessing steps
-% - Database preparation
-% - Schema extraction
-% - Gold standard SQL formatting
+\textsc{Bird} is another widely used benchmark that is comprised of 12,751
+question answer pairs spread across 95 databases in 37 domains. It aims to be
+more representative of real world scenarios with external knowledge
+requirements and is generally considered harder than \textsc{Spider}.
 
@@
 \paragraph{Execution Accuracy (EA)}
-% TODO: Define execution accuracy
-% - Definition: whether query results match ground truth
-% - Calculation method
-% - Why EA is the primary metric
-% - Reference to natural-benchmark implementation
+
+\textsc{Execution Accuracy} (or EA) is the primary success metric of NL2SQL systems
+as it represents the semantic accuracy of the SQL queries produced NL2SQL
+systems. Execution accuracy is computed by determining whether the rows in the
+candidate results are a permutation of the rows returned by ground truth
+results.
+
+For a query $q_g$ with candidate results $R_c$ and ground truth results $R_g$, 
+where $R_c, R_g \subseteq \mathbb{V}^{n \times m}$ (sets of $n$ rows with $m$ columns 
+of values $\mathbb{V}$), semantically accurate execution is defined as:
+
+\begin{equation}
+    semanticeq(R_c, R_g, q_g) = \mathbbm{1}\left[\exists \pi \in \Pi_m : 
+    \begin{cases}
+        R_c = \pi(R_g) & \text{if } ordered(q_g) \\
+        R_c \equiv_{\text{multiset}} \pi(R_g) & \text{otherwise}
+    \end{cases}
+    \right]
+\end{equation}
+
+where $\Pi_m$ is the set of valid column permutations (those preserving column
+value sets), $\pi(R_g)$ applies permutation $\pi$ to reorder columns in each
+row of $R_g$, $\equiv_{\text{multiset}}$ denotes multiset equality (allowing
+row reordering) and $ordered(q_g)$ is determining if the ground truth query
+$q_g$ contains an \texttt{ORDER BY} clause. Row and value comparisons use
+tolerance-based equality for floating-point values ($|\cdot| < 10^{-5}$). The
+calculation for measuring \textsc{Natural}'s EA was ported from \textsc{Spider}
+\citep{Spider}.
 
 \paragraph{Exact Match (EM)}
+
 % TODO: Define exact match
@@
 
+\textsc{Exact Match}..
+
 \paragraph{Performance Metrics}
-% TODO: Define performance metrics
+
+Measuring real world applicablility, the metric \textsc{Latency} is introduced 
+for measuring the compound pipeline runtime, thus measuring NL query to final
+SQL candidate.
+
+% TODO:
 % - End-to-end latency (time from NL query to SQL result)
@@
 
-\paragraph{Statistical Significance}
-% TODO: Describe statistical testing approach
-% - Test selection (e.g., paired t-test, McNemar's test)
-% - Significance threshold (p < 0.05)
-% - Confidence intervals
-
 \subsubsection{Baseline Systems and Comparisons}
 
+To determine the effect of approaches introduced by \textsc{Natural} two
+a set of baselines is introduced. The baselines are split into two groups,
+internal baselines (measuring the system without crucial components) and
+external baselines (eg, State-of-the-Art system performance).
+
 \paragraph{Internal Baselines}
-% TODO: Describe ablated baselines
-% - Zero-shot baseline (π only, no σ, φ, ρ, ν)
-% - Partial system variants for ablation studies
+
+Two primary internal baselines are introduced:
+
+\begin{enumerate}
+    \item \textbf{Model Baseline} — Using only the inference logic for
+        zero-shot accuracy, aiming to represent \textsc{OmniSQL} performance as
+        closely as possible.
+    \item \textbf{Zero-Shot Baseline} — A full \textsc{Natural} pipeline but
+        without any examples available for the in-context learning inference
+        phase.
+\end{enumerate}
+
+These baselines aim to measure both the actual performance of the NL2SQL model
+(\textsc{OmniSQL}) and the full \textsc{Natural} pipeline without in-context
+learning.
 
 \paragraph{State-of-the-Art Comparisons}
-% TODO: List SOTA systems for comparison
+
+\textcolor{red}{X} external baselines are referenced from previous research:
+
 % - GPT-4 results from literature
@@
 
-\subsection{Benchmark Performance Analysis}
-
-% TODO: Introduction paragraph presenting overall findings
+\subsection{Benchmark Results}
 
@@
 \paragraph{Overall Performance}
+
 % TODO: Present overall EA and EM scores
@@
 \paragraph{Performance by Difficulty Level}
+
 % TODO: Break down results by Spider difficulty classification
@@
 \paragraph{Performance by Database}
+
 % TODO: Present per-database analysis
@@
 \paragraph{Performance by SQL Complexity}
+
 % TODO: Analyze results by SQL pattern complexity
@@
 \paragraph{Learning Curve Analysis}
+
 % TODO: Analyze impact of training example quantity
@@
 \paragraph{Overall Performance}
+
 % TODO: Present overall Bird results
@@
 \paragraph{Domain-Specific Performance}
+
 % TODO: Break down by Bird's 37 domains
@@
 \paragraph{Dirty Data Handling}
+
 % TODO: Analyze performance on Bird's dirty data cases
@@
 \paragraph{External Knowledge Requirements}
+
 % TODO: Analyze cases requiring external knowledge
@@
 \paragraph{Open-Source vs Proprietary Models}
+
 % TODO: Discuss open-source (OmniSQL 7B) vs proprietary (GPT-4) trade-offs
@@
 \paragraph{Qualitative Comparison}
+
 % TODO: Beyond metrics, discuss systemic differences
@@
 \paragraph{Per-Component Timing Breakdown}
+
 % TODO: Present timing for each pipeline component
@@
 \paragraph{End-to-End Query Response Time}
+
 % TODO: Present total pipeline latency
@@
 \paragraph{Latency vs Accuracy Trade-offs}
+
 % TODO: Analyze speed-accuracy Pareto frontier
@@
 \paragraph{Bottleneck Identification and Optimization Opportunities}
+
 % TODO: Identify where time is spent

### 2025-12-29 thesis/assets/visualisations/src/Render.jl
Better julia code organization

@@ -1,0 +1,58 @@
+"""
+Render - Thesis visualization and plotting system
+
+Main module that coordinates benchmark data visualization.
+"""
+module Render
+
+using Plots
+using StatsPlots
+using ColorSchemes
+import PyPlot
+
+export run
+
+# ============================================================================
+# Configuration and Setup
+# ============================================================================
+
+pyplot()
+
+const PLOT_FONT = "Berkeley Mono"
+const BERKELEY_MONO_PATH = "/nix/store/yy92c3cq5lkfbcawckwv733rck0lz04d-berkeley-mono-1.0.0/share/fonts/truetype/berkeley-mono"
+
+PyPlot.matplotlib.font_manager.fontManager.addfont(joinpath(BERKELEY_MONO_PATH, "BerkeleyMono-Regular.ttf"))
+PyPlot.matplotlib.rcParams["svg.fonttype"] = "path"
+PyPlot.matplotlib.rcParams["font.family"] = "monospace"
+PyPlot.matplotlib.rcParams["font.monospace"] = ["Berkeley Mono"]
+
+default(fontfamily=PLOT_FONT, linewidth=2, framestyle=:box, grid=true)
+scalefontsizes()
+scalefontsizes(1.0)
+
+# :lajolla10 or :tofino10 or :vanimo10
+const PALETTE = palette(:tofino10)
+
+# ============================================================================
+# Include submodules
+# ============================================================================
+
+include("data.jl")
+include("plots.jl")
+include("cli.jl")
+
+# ============================================================================
+# Public API
+# ============================================================================
+
+"""
+    run(args::Vector{String})
+
+Main entry point for the visualization system.
+Parses command-line arguments and generates all configured plots.
+"""
+function run(args::Vector{String})
+    return main(args; plot_font=PLOT_FONT, palette=PALETTE)
+end
+
+end # module

### 2025-12-29 thesis/assets/visualisations/src/cli.jl

@@ -1,0 +1,124 @@
+"""
+Command-line interface and configuration parsing.
+"""
+
+using ArgParse
+using TOML
+
+include("data.jl")
+include("plots.jl")
+
+function parse_commandline()
+    s = ArgParseSettings()
+
+    @add_arg_table! s begin
+        "--results"
+            help = "Path to results TOML file"
+            arg_type = String
+            default = "var/conf/benchmark.toml"
+        "--plots"
+            help = "Path to plot configuration TOML file"
+            arg_type = String
+            default = "var/conf/plots.toml"
+    end
+
+    return parse_args(s)
+end
+
+function create_plot_from_config(config::Dict)
+    plot_type = config["type"]
+
+    if plot_type == "benchmark"
+        return BenchmarkPlot(
+            config["benchmark"],
+            config["metric"],
+            config["metric_label"],
+            title=get(config, "title", nothing),
+            output=config["output"]
+        )
+    elseif plot_type == "metric_comparison"
+        return MetricComparisonPlot(
+            config["metric"],
+            metric_label=get(config, "metric_label", nothing),
+            title=get(config, "title", nothing),
+            output=config["output"]
+        )
+    elseif plot_type == "relative_improvement"
+        return RelativeImprovementPlot(
+            baseline_key=config["baseline_key"],
+            comparison_keys=config["comparison_keys"],
+            metric=config["metric"],
+            title=config["title"],
+            output=config["output"]
+        )
+    elseif plot_type == "metric_comparison_bars"
+        return MetricComparisonBarsPlot(
+            config["metric"],
+            metric_label=get(config, "metric_label", nothing),
+            title=get(config, "title", nothing),
+            include_sota=get(config, "include_sota", false),
+            output=config["output"]
+        )
+    elseif plot_type == "sota_comparison"
+        return SotaComparisonPlot(
+            natural_configs=config["natural_configs"],
+            sota_configs=config["sota_configs"],
+            metric=config["metric"],
+            title=config["title"],
+            output=config["output"]
+        )
+    elseif plot_type == "example_source"
+        return ExampleSourcePlot(
+            source_keys=config["source_keys"],
+            metric=config["metric"],
+            title=config["title"],
+            output=config["output"]
+        )
+    else
+        error("Unknown plot type: $plot_type")
+    end
+end
+
+function main(args::Vector{String}; plot_font, palette)
+    parsed_args = parse_commandline()
+
[…]

### 2025-12-29 thesis/evaluation/index.tex
Cleanup

@@
 row reordering) and $ordered(q_g)$ is determining if the ground truth query
-$q_g$ contains an \texttt{ORDER BY} clause. Row and value comparisons use
-tolerance-based equality for floating-point values ($|\cdot| < 10^{-5}$). The
-calculation for measuring \textsc{Natural}'s EA was ported from \textsc{Spider}
-\citep{Spider}.
+$q_g$ contains an \texttt{ORDER BY} clause.
 
@@
 % - Per-component latency (σ, φ, π, ρ, ν timings)
-% - Throughput (queries per second)
 % - Resource utilization (GPU memory, CPU usage)
@@
 
-\textcolor{red}{X} external baselines are referenced from previous research:
+Two primary external baselines are taking into account:
 
-% - GPT-4 results from literature
-% - DIN-SQL, DAIL-SQL, C3 results
-% - Other relevant open-source approaches
-% - Note: Direct comparison constraints (different models, hardware)
+\begin{enumerate}
+    \item \textbf{GPT-4} — A proprietary baseline for advanced model
+        capabilities without finetuning.
+    \item \textbf{OmniSQL} — The open source baseline model series used by
+        \textsc{Natural}.
+\end{enumerate}
 
@@
 \paragraph{Quantitative Comparison}
+
 % TODO: Create comprehensive comparison table

### 2025-12-29 thesis/assets/visualisations/src/plots.jl

@@ -12,6 +12,14 @@ function Label(s::String)
     return "\n$(s)\n"
 end
 
+# Ensure output directory exists before saving
+function ensure_output_dir(output_path::String)
+    output_dir = dirname(output_path)
+    if !isempty(output_dir) && !isdir(output_dir)
+        mkpath(output_dir)
+    end
+end
+
 # ============================================================================
 # BenchmarkPlot - Single benchmark line plot
 # ============================================================================
@@ -69,6 +77,7 @@ function render(plot_config::BenchmarkPlot, data::BenchmarkData; plot_font, pale
           linewidth=2,
           color=palette[1])
 
+    ensure_output_dir(plot_config.output)
     savefig(p, plot_config.output)
     println("plot saved to: $(plot_config.output)")
 
@@ -139,6 +148,7 @@ function render(plot_config::MetricComparisonPlot, data::BenchmarkData; plot_fon
               color=palette[color_idx])
     end
 
+    ensure_output_dir(plot_config.output)
     savefig(p, plot_config.output)
     println("plot saved to: $(plot_config.output)")
 
@@ -209,6 +219,7 @@ function render(plot_config::RelativeImprovementPlot, data::BenchmarkData; plot_
 
     hline!(p, [0], color=:black, linestyle=:dash, label=Label("Baseline"), linewidth=1)
 
+    ensure_output_dir(plot_config.output)
     savefig(p, plot_config.output)
     println("plot saved to: $(plot_config.output)")
     return p
@@ -282,6 +293,7 @@ function render(plot_config::MetricComparisonBarsPlot, data::BenchmarkData; plot
         color_palette=palette
     )
 
+    ensure_output_dir(plot_config.output)
     savefig(p, plot_config.output)
     println("plot saved to: $(plot_config.output)")
     return p
@@ -351,6 +363,7 @@ function render(plot_config::SotaComparisonPlot, data::BenchmarkData; plot_font,
         color_palette=palette
     )
 
+    ensure_output_dir(plot_config.output)
     savefig(p, plot_config.output)
     println("plot saved to: $(plot_config.output)")
     return p
@@ -410,6 +423,7 @@ function render(plot_config::ExampleSourcePlot, data::BenchmarkData; plot_font, 
         color_palette=palette
     )
 
+    ensure_output_dir(plot_config.output)
     savefig(p, plot_config.output)
     println("plot saved to: $(plot_config.output)")
     return p

### 2025-12-29 thesis/assets/visualisations/visualize.jl

@@ -1,612 +1,0 @@
-using Plots
-using StatsPlots
-using TOML
-using ColorSchemes
-using ArgParse
-import PyPlot
-
-# ============================================================================
-# Configuration and Setup
-# ============================================================================
-
-pyplot()
-
-const PLOT_FONT = "Berkeley Mono"
-const BERKELEY_MONO_PATH = "/nix/store/yy92c3cq5lkfbcawckwv733rck0lz04d-berkeley-mono-1.0.0/share/fonts/truetype/berkeley-mono"
-
-PyPlot.matplotlib.font_manager.fontManager.addfont(joinpath(BERKELEY_MONO_PATH, "BerkeleyMono-Regular.ttf"))
-PyPlot.matplotlib.rcParams["svg.fonttype"] = "path"
-PyPlot.matplotlib.rcParams["font.family"] = "monospace"
-PyPlot.matplotlib.rcParams["font.monospace"] = ["Berkeley Mono"]
-
-default(fontfamily=PLOT_FONT, linewidth=2, framestyle=:box, grid=true)
-scalefontsizes()
-scalefontsizes(1.0)
-
-# :lajolla10 or :tofino10 or :vanimo10
-const PALETTE = palette(:tofino10)
-
-# ============================================================================
-# Data Structures
-# ============================================================================
-
-struct Pipeline
-    key::String
-    label::String
-    order::Int
-    external::Bool
-    sota::Bool
-    verified::Bool
-end
-
-function Pipeline(key::String, data::Dict)
-    Pipeline(
-        key,
-        data["label"],
-        data["order"],
-        get(data, "external", false),
-        get(data, "sota", false),
-        get(data, "verified", true)
-    )
-end
-
-struct BenchmarkData
-    file_path::String
-    pipelines::Vector{Pipeline}
-    benchmarks::Dict{String, Dict{String, Dict{String, Float64}}}
-end
-
-function BenchmarkData(file_path::String)
-    data = TOML.parsefile(file_path)
-
-    pipeline_keys = collect(keys(data["pipelines"]))
-    pipelines = [Pipeline(k, data["pipelines"][k]) for k in pipeline_keys]
-    sort!(pipelines, by = p -> p.order)
-
-    benchmarks = get(data, "benchmark", Dict())
-
-    BenchmarkData(file_path, pipelines, benchmarks)
-end
-
[…]
